### 2022-10-19 Manuscript/endo_herbarium.tex
draft of the MS, trying to sort out references in the tex template

\documentclass[11pt]{article}
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,sectionbib,sort]{natbib}
\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{titlesec}
\titleformat{\section}[block]{\Large\bfseries\filcenter}{\thesection}{1em}{}
\titleformat{\subsection}[block]{\Large\itshape\filcenter}{\thesubsection}{1em}{}
\titleformat{\subsubsection}[block]{\large\itshape}{\thesubsubsection}{1em}{}
\titleformat{\paragraph}[runin]{\itshape}{\theparagraph}{1em}{}[. ]\renewcommand{\refname}{Literature Cited}

%%%%%%%%%%%%%%%%%%%%%
% Line numbering
%%%%%%%%%%%%%%%%%%%%%
%
% Please use line numbering with your initial submission and
% subsequent revisions. After acceptance, please turn line numbering
% off by adding percent signs to the lines %\usepackage{lineno} and
% to %\linenumbers{} and %\modulolinenumbers[3] below.
%
% To avoid line numbering being thrown off around math environments,
% the math environments have to be wrapped using
% \begin{linenomath*} and \end{linenomath*}
%
% (Thanks to Vlastimil Krivan for pointing this out to us!)

\title{Idiosyncratic responses of microbial mutualisms over 150 years of climate change}

% This version of the LaTeX template was last updated on
% September 28, 2022.

%%%%%%%%%%%%%%%%%%%%%
% Authorship
%%%%%%%%%%%%%%%%%%%%%
% Please remove authorship information while your paper is under review,
% unless you wish to waive your anonymity under double-blind review. You
% will need to add this information back in to your final files after
% acceptance.

\author{Joshua C. Fowler$^{1,\ast}$ \\
	Generic H. Collaborator$^{2}$ \\
	Tom E. X. Miller$^{1}$}
\date{}

\begin{document}

	\maketitle

	\noindent{} 1. Rice University, Department of BioSciences, Houston, Texas 77006;

	\noindent{} 2. University of;

	\noindent{} $\ast$ Corresponding author; e-mail: jcf3@rice.edu.

	\bigskip

	\textit{Manuscript elements}: Figure~1, figure~2, table~1, appendix~A (for print; including figure~A1, figure~A2, and table~A1), supplemental PDF. Figure~2 is to print in color.

	\bigskip

	\textit{Keywords}: .

	\bigskip

	\textit{Manuscript type}: Article. %Or note, natural history miscellany note, comment, reply, invited symposium, featured topic, or historical perspective.

	\bigskip

	\noindent{\footnotesize Prepared using the suggested \LaTeX{} template for \textit{Am.\ Nat.}}

	%\linenumbers{}
	%\modulolinenumbers[3]

	\newpage{}

	\section*{Abstract}

	Lorem ipsum dolor sit amet, consectetur adipiscing elit. Sed non risus. Suspendisse lectus tortor, dignissim sit amet, adipiscing nec, ultricies sed, dolor. Cras elementum ultrices diam. Praesent quis dolor in dolor molestie cursus et ac nisi. Vestibulum ante purus, semper eget est vitae, vehicula ornare nisl. Morbi efficitur euismod enim, nec feugiat tellus cursus eget.

	\newpage{}

	\section*{Introduction}

	% The journal does not have numbered sections in the main portion of
	% articles. Please refrain from using section references (à la
	% section~\ref{section:CountingOwlEggs}), and refer to sections by name
	% (e.g. section ``Counting Owl Eggs'').
Global climate change involves hetergenous changes in multiple climate drivers \citep{IPCC2012managing}. 
It is of great interest to predict species' responses to these drivers.
Novel conditions .Predicting how species' respond to these drivers, it is important to understand how and under what conditions interactions between species will alter expected responses. 
Incorporating interactions may lead to different predictions than those based on single species responses; for most species, we do not know whether their interactions will facilitate or hinder movement and adaptation to novel conditions \cite{gilman2010framework}. 
Natural history collections, which were originally collected to study and preserve taxonomic diversity, present an exciting opportunity to explore long-term changes in ecological interactions across broad spatial scales \citep{meineke2018unrealized}. 
These collections primarily comprise physical specimens of organisms collected along with information about the time and place of collection. 
These natural history collections, built and maintained by the contributions of thousands of scientists, are invaluable time machines. 
The specimens they contain are samples from dynamically changing environmental conditions through time and space.
These specimens can preserve physical legacies of ecological dynamics and species' interactions. 
Recent work using collections has documented changes in phenology \citep{willis2017old, park2019herbarium, berg2019examination}, as well as pollination \citep{pauw2011reconstruction, duan2019century} and changes in negative interactions such as herbivory \citep{meineke2019herbarium}, but to date, focus has been lacking on long-term changges in reciprocally positive interactions, i.e. mutualisms.

Given that mutualists alter their partner's fitness, they may play an important role in limiting environmental stress, but changing conditions may also disrupt or breakdown the mutualism \citep{aslan2013mutualism}. 
Examples of mutualism disruption as a result of changing climates include phenological mismatches between plants and their pollinators \citep{pauw2011reconstruction}, drought-driven breakdown in seagrass-lucinid mutualisms \citep{de2016drought}, and coral bleaching due to temperature stress \citep{sully2019global}, all of which can have important consequences for habitat and biodiversity maintainence. 
Alternatively, mutualists may facilitate adaptive responses by improving fitness under stressful conditions, such as facultative bacterial symbionts within insects implicated in improving thermal tolerance \citep{truitt2019wolbachia, renoz2019evolutionary}, or fungal symbionts that allow their grass host to persist in more arid edges of its distribution \citep{afkhami2014mutualist}. 
Whether mutualisms breakdown or promote presistence under climate change is unclear, but likely depends on the intimacy and specialization of the interaction, as well as the physiological tolerances of the mutualists\citep{toby2010mutualisms, warren2014mutualism, rafferty2015phenological}. 
Given that climate change is spatially heterogenous, the rate of change may also have a strong influence on when and where we can expect mutualism facilitation or disruption.

Early research on \emph{Epichloë} fungal endophytes, symbionts of cool-season grasses, used herbarium specimens to describe the broad taxonomic diversity of hosts \citep{white1985endophyte}. 
These endophytes are vertically-transmitted, growing throughout the aboveground tissue and into the seeds of their hosts, and consequently are preserved along with the plant on herbarium sheets. 
Vertical transmission links the fitness of host and symbiont together meaning that, over time, endophytes that provide fitness benefits should rise in prevalence within the population \citep{fine1975vectors, douglas1998host, rudgers2009fungus}. 
This leads to the prediction that interaction prevalence should generally be high where there are strong fitness benefits, and that prevalence should track changes in environmental drivers that limit fitness. 
Environmental stress is likely particularly consequential at species' range edges wherer population growth rates are expected to be near replacement levels \cite{}. 
And so understanding the ways that endophytes may mediate stress is important for understanding the factors setting species' range limits. 
The interaction prevalence should respond to the fitness effects of the mutualism, and the prevalence of endophytes across the range determines their actual contribution to setting range limits. 
In addition, the rate of environmental change is variable across space across species ranges \citep{IPCC2012managing}, and local changes in endophyte prevalence likely depend on these rates.

And so, we are taking advantage of the spatial and temporal sampling within herbaria to ask how change in spatial and temporal environmental drivers alter the prevaence of a widespread mutualistic fungal endophyte symbiosis between Epichloë fungi and grasses. 
Specifically,  we ask (1) how variation in temperature and precipitation correlate with endophyte prevalence across space and over time, and (2) does greater rates of envirnmental change drive greater changes in endophyte prevalence.

	\section*{Methods}

I will quantify the spatial and temporal responses of fungal endophyte symbiosis to environmental conditions by assessing the status of endophyte symbiosis in herbarium specimens. 
In particular, I will collect a small sample of seeds from herbarium specimens of \emph{Agrostis hyemalis} and \emph{Elymus virginicus} from across their distributions, and use microscopy to identify the endophyte status of specimens.
	\subsection*{Second-order heading}

	Lorem ipsum nulla facilisi, pace \citet{LemKapEx07}. Etiam semper, orci sit amet facilisis interdum, tellus nunc consequat erat, quis viverra nisi diam ut metus. Pellentesque cursus, sapien malesuada euismod iaculis, mauris purus interdum diam, vel vestibulum justo enim vitae tellus. Nunc interdum lorem sit amet diam volutpat tristique. Quisque pulvinar ac metus commodo lacinia (\citealt{Ing11,Xiao2015}).

	\subsubsection*{Third-order heading}

	Usually two or three levels of heading will be all you need. Journal style even permits a fourth level in case you need it.

	\paragraph*{Fourth-order heading}
	The quick red fox jumps over the lazy brown dog in this paragraph as well. Donec mauris nibh, volutpat vehicula viverra at, iaculis congue sem. Praesent eget erat rhoncus erat sollicitudin volutpat.

	\begin{equation}
	{ \frac{1}{N_k-1} \sum \limits_{t=1}^{N_k} (M_{tjk} - \bar{M}_{jk})^2}
	\end{equation}

	Praesent quis dolor in dolor molestie cursus et ac nisi. Vestibulum ante purus, semper eget est vitae, vehicula ornare nisl. Morbi efficitur euismod enim, nec feugiat tellus cursus eget. Donec mauris nibh, volutpat vehicula viverra at, iaculis congue sem.

	\subsection*{Another second-order heading}

	As \citet{Xiao2015} argued, phasellus porttitor eros et ante condimentum, eget facilisis orci condimentum. Nulla facilisi. Proin placerat elit blandit, euismod dolor nec, dapibus diam. Mauris posuere malesuada lacus, at elementum lacus auctor eu (fig~\ref{Fig:Jumps}A).

	\section*{Results}

	Lorem ipsum dolor sit amet. Aenean pulvinar malesuada commodo (see \citealt{DavisEtAl2011}; table~\ref{Table:Founders}). Sed aliquet mauris odio, in tristique dui egestas a. Etiam eu malesuada quam. Suspendisse tincidunt eu erat sit amet vulputate. Duis at arcu et nisl dictum mattis. Maecenas vel cursus ante. Cras eleifend elit nec velit sollicitudin fermentum in ac mauris. Pellentesque rutrum magna vel elit maximus hendrerit.

	\subsection*{The height of the jump}

	Aenean eu pellentesque quam (fig.~\ref{Fig:OkapiHorn}). Nam pellentesque augue eu finibus lacinia. Nullam nec justo vitae odio imperdiet rhoncus vitae vitae quam. Pellentesque porttitor metus et lectus ornare, ac cursus urna efficitur (fig~\ref{Fig:Jumps}B).

	\subsection*{The laziness of the dog}

	Example paragraph with embedded references (video~\ref{VideoExample}, fig.~\ref{Fig:AnotherFigure}). 
	If you have deposited data to Dryad, it is advisable to cite them somewhere in the main text (usually in the Methods or Results sections). A sentence like the following will do: All data are available in the Dryad Digital Repository (\citealt{CookEtAl2015}).

	\section*{Discussion}

are a potentially rich target for study within musuem specimens\citep{}. Practically all specimens will be preserved alongside their microbiome, and while some evidence of these association can be expected to be lost over time due to specimen handling and storage.

	\section*{Conclusion}

	Duis pharetra enim at libero cursus, eu commodo mi vestibulum. Nullam eget velit nec lectus viverra sodales. Suspendisse egestas, eros at dictum tincidunt, mi orci laoreet libero, eget rutrum sapien arcu blandit odio.

	%%%%%%%%%%%%%%%%%%%%%
	% Acknowledgments
	%%%%%%%%%%%%%%%%%%%%%
	% You may wish to remove the Acknowledgments section while your paper 
	% is under review (unless you wish to waive your anonymity under
	% double-blind review) if the Acknowledgments reveal your identity.
	% If you remove this section, you will need to add it back in to your
	% final files after acceptance.

	\section*{Acknowledgments}

	%%%%%%%%%%%%%%%%%%%%%
	% Statement of Authorship
	%%%%%%%%%%%%%%%%%%%%%
	% This section should also be commented out while your MS is undergoing
	% double-blind review. The specifics should of course be adapted to
	% your paper, but the paragraph below gives some hints of possible
	% contributions.

	\section*{Statement of Authorship}

	\section*{Data and Code Availability}

	On initial submission, you may use this section to provide a URL for editors and reviewers that is `private for peer review'. After acceptance, this section must be updated with correct, working DOIs for data deposits (typically on the Dryad Digital Repository, \citealt{CookEtAl2015}) and code deposits (such as in Zenodo).

	\section*{Appendix A: Additional Methods and Parameters}

	% In most cases, authors should typeset supplementary material in a separate,
	% author-supplied PDF. For author-supplied PDFs, please consult the
	% AmNat_supp_template.tex document, available from
	% https://www.journals.uchicago.edu/journals/an/instruct 
	%
	% By contrast, the Appendix instructions below apply to cases in which
	% a brief appendix is to appear in print after the main body of the article.
	% That notably includes descriptions of methods, tables defining parameters,
	% and other material necessary for reproducing the MS's results.
	%
	% Please reset counters for the appendix (thus normally figure A1, 
	% figure A2, table A1, etc.).
	%
	% Most AmNat articles have no more than one print appendix. If your article
	% has more than one, counters for each appendix should match the letter of
	% that appendix. For example, tables in Appendix B should be numbered table B1, % table C2, etc. This applies to tables, equations, and figures.
	%
	% It's better not to use the \appendix command, because we have some
	% formatting peculiarities that \appendix conflicts with.

	\renewcommand{\theequation}{A\arabic{equation}}
	% redefine the command that creates the equation number.
	\renewcommand{\thetable}{A\arabic{table}}
	\setcounter{equation}{0}  % reset counter 
	\setcounter{figure}{0}
	\setcounter{table}{0}

	\subsection*{Fox--dog encounters through the ages}

	The quick red fox jumps over the lazy brown dog. The quick red fox has always jumped over the lazy brown dog. The quick red fox began jumping over the lazy brown dog in the 19th century and has never ceased from so jumping, as we shall see in figure~\ref{Fig:Jumps}. But there can be surprises (figure~\ref{Fig:JumpsOk}).

	If the order and location of figures is not otherwise clear, feel free to include explanatory dummy text like this:

	[Figure A1 goes here.]

	[Figure A2 goes here.]

	\subsection*{Further insights}

	Tables in the appendices can appear in the appendix text (see table~\ref{Table:Rivers} for an example), unlike appendix figure legends which should be grouped at the end of the document together with the other figure legends.

	\begin{table}[h]
		\caption{Various rivers, cities, and animals}
		\label{Table:Rivers}
		\centering
		\begin{tabular}{lll}\hline
			River        & City        & Animal            \\ \hline
			Chicago      & Chicago     & Raccoon           \\
			Des Plaines  & Joliet      & Coyote            \\
			Illinois     & Peoria      & Cardinal          \\
			Kankakee     & Bourbonnais & White-tailed deer \\
			Mississippi  & Galena      & Bald eagle        \\ \hline
		\end{tabular}
		\bigskip{}
		\\
		{\footnotesize Note: See table~\ref{Table:Founders} below for further table formatting hints.}
	\end{table}

	Lorem ipsum dolor sit amet, as we have seen in figures~\ref{Fig:Jumps} and \ref{Fig:JumpsOk}.

	%%%%%%%%%%%%%%%%%%%%%
	% Bibliography
	%%%%%%%%%%%%%%%%%%%%%
	% You can either type your references following the examples below, or
	% compile your BiBTeX database and paste the contents of your .bbl file
	% here. The amnatnat.bst style file should work for this---but please
	% let us know if you run into any hitches with it!
	%
	% If you upload a .bib file with your submission, please upload the .bbl
	% file as well; this will be required for typesetting.
	%
	% The list below includes sample journal articles, book chapters, and
	% Dryad references.
\bibliographystyle{plain}
\bibliography{endo_herbarium}

	\newpage{}

	\section*{Tables}
	\renewcommand{\thetable}{\arabic{table}}
	\setcounter{table}{0}

	\begin{table}[h]
		\caption{Founders of \textit{The~American Naturalist}}
		\label{Table:Founders}
		\centering
		\begin{tabular}{lll}\hline
			Early editor            & Years with the journal \\ \hline
			Alpheus S. Packard Jr.  & 1867--1886 \\
			Frederick W. Putnam     & 1867--1874 \\ 
			Edward S. Morse         & 1867--1871 \\ 
			Alpheus Hyatt           & 1867--1871 \\
			Edward Drinker Cope$^a$ & 1878--1897 \\
			J.~S. Kingsley          & 1887--1896 \\ \hline 
		\end{tabular}
		\bigskip{}
		\\
		{\footnotesize Note: Table titles should be short. Further details should go in a `notes' area after the tabular environment, like this. $^a$ Published the first description of \textit{Dimetrodon}.}
	\end{table}

	\newpage{}

	\section*{Figure legends}

	\begin{figure}[h!]
		%\includegraphics{horn-of-okapi}
		\caption{Figure legends can be longer than the titles of tables. However, they should not be excessively long---in most cases, they should be no more than 100 words each.}
		\label{Fig:OkapiHorn}
	\end{figure}

	\begin{figure}[h!]
		%\includegraphics{elegance}
		\caption{In this way, figure legends can be listed at the end of the document, with references that work, even though the graphic itself should be included for final files after acceptance. Instead, upload the relevant figure files separately to Editorial Manager; Editorial Manager should insert them at the end of the PDF automatically.}
		\label{Fig:AnotherFigure}
	\end{figure}

	% Figure legends for any print appendices you might have.

	\renewcommand{\thefigure}{A\arabic{figure}}
	\setcounter{figure}{0}

	\begin{figure}[h!]
		%\includegraphics{jumps20m}
		\caption{\textit{A}, the quick red fox proceeding to jump 20~m straight into the air over not one, but several lazy dogs. \textit{B}, the quick red fox landing gracefully despite the skepticism of naysayers.}
		\label{Fig:Jumps}
	\end{figure}

	\begin{figure}[h!]
		%\includegraphics{jumps20m}
		\caption{The quicker the red fox jumps, the likelier it is to land near an okapi. For further details, see \citet{LemKapEx07}.}
		\label{Fig:JumpsOk}
	\end{figure}

	%%%%%%%%%%%%%%%%%%%%%
	% Videos
	%%%%%%%%%%%%%%%%%%%%%
	% If you have videos, journal style for them is generally similar to that for
	% figures.

	%%%%% Include the text below if you have videos

	\renewcommand{\figurename}{Video} 
	\renewcommand{\thefigure}{S\arabic{figure}}
	\setcounter{figure}{0}

	\begin{figure}[h!]
		%\includegraphics{VideoScreengrab.jpg}
		\caption{Video legends can follow the same principles as figure legends. Counters should be set and reset so that videos and figures are enumerated separately.}
		\label{VideoExample}
	\end{figure}

	\renewcommand{\figurename}{Figure}
	\setcounter{figure}{1}

	%%%%% Include the above if you have videos

\end{document}

### 2022-12-15 Manuscript/endo_herbarium.tex
updates to the intro

@@
 
-\title{Idiosyncratic responses of microbial mutualisms over 150 years of climate change}
+\title{Decline for some, resilience for others: changes in the prevalence of microbial symbionts over 150 years of climate change}
 
@@
 Global climate change involves hetergenous changes in multiple climate drivers \citep{IPCC2012managing}. 
-It is of great interest to predict species' responses to these drivers.
-Novel conditions .Predicting how species' respond to these drivers, it is important to understand how and under what conditions interactions between species will alter expected responses. 
-Incorporating interactions may lead to different predictions than those based on single species responses; for most species, we do not know whether their interactions will facilitate or hinder movement and adaptation to novel conditions \cite{gilman2010framework}.
+Predicting species' responses to these drivers, whether species will move, adapt, or potentially go extinct is a major goal of basic ecological research and conservation management. 
+Across the tree of life, species interact with macro- and microscopic symbionts which impact their growth and performance. 
+These biotic interactions likely modulate responses to changing environmental conditions, yet for most species, we do not know whether they will facilitate or hinder movement and adaptation to novel conditions \cite{gilman2010framework}.
+Predictions incorporating biotic interactions are likely different than those based on single species responses, but empirical assements of these predictions have been limited by the lack of long-term and spatially extensive datasets of biotic interactions.
+
 Natural history collections, which were originally collected to study and preserve taxonomic diversity, present an exciting opportunity to explore long-term changes in ecological interactions across broad spatial scales \citep{meineke2018unrealized}.
@@
 This leads to the prediction that interaction prevalence should generally be high where there are strong fitness benefits, and that prevalence should track changes in environmental drivers that limit fitness.
-Environmental stress is likely particularly consequential at species' range edges wherer population growth rates are expected to be near replacement levels \cite{}. 
-And so understanding the ways that endophytes may mediate stress is important for understanding the factors setting species' range limits. 
+Environmental stress is likely particularly consequential at species' range edges where population growth rates are expected to be near replacement levels \cite{}. 
+Understanding the ways that endophytes may mediate stress is important for understanding the factors setting species' range limits. 
 The interaction prevalence should respond to the fitness effects of the mutualism, and the prevalence of endophytes across the range determines their actual contribution to setting range limits. 
 In addition, the rate of environmental change is variable across space across species ranges \citep{IPCC2012managing}, and local changes in endophyte prevalence likely depend on these rates. 
-
-And so, we are taking advantage of the spatial and temporal sampling within herbaria to ask how change in spatial and temporal environmental drivers alter the prevaence of a widespread mutualistic fungal endophyte symbiosis between Epichloë fungi and grasses.
-Specifically,  we ask (1) how variation in temperature and precipitation correlate with endophyte prevalence across space and over time, and (2) does greater rates of envirnmental change drive greater changes in endophyte prevalence.
+We are taking advantage of the spatial and temporal sampling within herbaria to ask how the prevalence of a widespread mutualistic fungal endophyte symbiosis between Epichloë fungi and grasses has changed in response to changing climate drivers.
+Specifically, the objectives of this study are to (i) quantify spatial and temporal change in endophyte prevalence, (ii) identify climate drivers that can explain changes in prevalence, and (iii) assess how the rate of environmental change influences changes in the prevalence of symbionts.
 

### 2022-12-15 endo_herbarium_records_merge.R
merging with the newer scores

@@ -38,7 +38,7 @@ AM_records <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/Digiti
   unite(Institution_specimen_id, CollectionCode, id, sep = "") %>%
   separate(DateCollected, into = c("year", "month", "day"), remove = FALSE) %>% 
   mutate(year = as.numeric(year), month = as.numeric(month), day = as.numeric(day)) %>% 
-  dplyr::select(-contains("X"))
+  dplyr::select(-contains("..."))
 
 
 # BRIT digitized records downloaded from TORCH (includes Vanderbilt and U of Louisiana Monroe) 
@@ -1130,4 +1130,37 @@ endo_herb_georef <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/
   mutate(Spp_code = case_when(grepl("AGHY", Sample_id) ~ "AGHY",
                               grepl("ELVI", Sample_id) ~ "ELVI",
                               grepl("AGPE", Sample_id) ~ "AGPE"))
+
+specimen_counts <- endo_herb_georef %>% 
+  mutate(Spp_code = case_when(grepl("AGHY", Sample_id) ~ "AGHY",
+                              grepl("ELVI", Sample_id) ~ "ELVI",
+                              grepl("AGPE", Sample_id) ~ "AGPE")) %>% 
+  group_by(Spp_code, tissue_type) %>% 
+  summarize(n())
+scored_counts <- endo_herb_georef %>% 
+  filter(!is.na(Endo_status_liberal_1)) %>% 
+  mutate(Spp_code = case_when(grepl("AGHY", Sample_id) ~ "AGHY",
+                              grepl("ELVI", Sample_id) ~ "ELVI",
+                              grepl("AGPE", Sample_id) ~ "AGPE")) %>% 
+  group_by(Spp_code, tissue_type) %>% 
+  summarize(n())
+
+hist(endo_herb_georef$year)
+plot(endo_herb_georef$lon, endo_herb_georef$lat)
+plot(endo_herb_georef$lon, endo_herb_georef$Endo_status_liberal_1)
+plot(endo_herb_georef$lat, endo_herb_georef$Endo_status_liberal_1)
+plot(endo_herb_georef$year, endo_herb_georef$Endo_status_liberal_1)
+
+
[…]

### 2022-12-15 endo_herbarium_spatial_analysis.R
script for spatial SPDE model

@@ -1,0 +1,130 @@
+# Purpose: Fits spatial model of change in endophyte prevalence in herbarium specimens
+# Authors: Joshua Fowler, Mallory Tucker, Ella Segal, Lani Dufresne, and Tom Miller
+# Updated: Noov 11, 2022
+
+library(tidyverse) # for data manipulation and ggplot
+library(INLA) # for fitting integrated nested Laplace approximation models
+library(sf)
+
+
+################################################################################
+############ Read in the herbarium dataset ############################### 
+################################################################################
+
+endo_herb_georef <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/DigitizedHerbariumRecords/endo_herb_georef.csv") %>%
+  filter(Country != "Canada") %>%
+  mutate(Spp_code = case_when(grepl("AGHY", Sample_id) ~ "AGHY",
+                              grepl("ELVI", Sample_id) ~ "ELVI",
+                              grepl("AGPE", Sample_id) ~ "AGPE"))
+
+endo_herb_AGHY <- endo_herb_georef %>% 
+  filter(!is.na(Endo_status_liberal_1)) %>% 
+  filter(Spp_code == "AGHY") %>% 
+  filter(!is.na(lon) & !is.na(year)) %>% 
+  filter(lon>-110)
+
+endo_herb_ELVI <- endo_herb_georef %>% 
+  filter(!is.na(Endo_status_liberal_1)) %>% 
+  filter(Spp_code == "ELVI") %>% 
+  filter(!is.na(lon) & !is.na(year)) 
+
+endo_herb <- endo_herb_georef %>% 
+  filter(!is.na(Endo_status_liberal_1)) %>%
+  filter(!is.na(lon) & !is.na(year)) %>% 
+  filter(lon>-110)
+
+
+################################################################################
+############ Setting up and running INLA model ############################### 
+################################################################################
+# messing aruond with INLA model fitting
+# generate mesh, which is used to fit the spatial effects
+# using a coarser mesh to start out as this is easier computation and sufficient to capture spatial effects
+coords <- cbind(endo_herb$lon, endo_herb$lat)
+
+non_convex_bdry <- inla.nonconvex.hull(coords, -0.03, -0.05, resolution = c(100, 100))
+
+max.edge = diff(range(coords[,2]))/(10)
[…]

### 2022-12-15 endo_herbarium_records_merge.R
pulling out the existing hand georeferenced coordinates

@@ -35,7 +35,7 @@ UTAustin_torch <- rbind(AGHY_UTAustin, ELVI_UTAustin, AGPE_UTAustin)
 
 # Texas A&M digitized records (Includes both AGHY and ELVI)
 AM_records <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/DigitizedHerbariumRecords/TexasA&M_digitized_records/Fowler Data.csv") %>% 
-  unite(Institution_specimen_id, CollectionCode, id, sep = "") %>%
+  unite(Institution_specimen_id, CollectionCode, id, sep = "") %>% 
   separate(DateCollected, into = c("year", "month", "day"), remove = FALSE) %>% 
   mutate(year = as.numeric(year), month = as.numeric(month), day = as.numeric(day)) %>% 
   dplyr::select(-contains("..."))
@@ -135,6 +135,8 @@ AGHY_BRIT <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/Digitiz
   filter(!is.na(county), !is.na(eventDate)) %>% 
   separate(eventDate, into = c("year", "month", "day"), remove = FALSE) %>% 
   mutate_at(c("day", "month", "year"), as.numeric) %>% 
+  mutate(decimalLongitude = case_when(decimalLongitude>0 ~ decimalLongitude*-1,
+                                      TRUE ~ decimalLongitude)) %>% 
   dplyr::select(id, catalogNumber, country, stateProvince, county, municipality, locality, decimalLatitude, decimalLongitude, coordinateUncertaintyInMeters, eventDate, day, month, year) %>% 
   mutate(Spp_code = "AGHY")
 
@@ -841,6 +843,8 @@ specimen_info <- read_csv(file = "~joshuacfowler/Dropbox/Josh&Tom - shared/Endo_
   mutate(year = case_when(Spp_code != "AGPE" ~ year.x,
                           Spp_code == "AGPE" & is.na(year.y) ~ year.x,
                           Spp_code == "AGPE" & is.na(year.x) ~ year.y)) %>% 
+  mutate(hand_georef_lon = case_when(decimalLongitude>0 ~ decimalLongitude*-1,TRUE ~ decimalLongitude),
+         hand_georef_lat = decimalLatitude) %>% 
   dplyr::select(-year.x, - year.y)
 
 # This is the sample info and we will filter for only those that we have scored so far.
[…]

### 2022-12-16 endo_herbarium_records_merge.R
pulling out the collector names

@@ -8,6 +8,7 @@ library(tidyverse) # for data manipulation and ggplot
 library(fuzzyjoin) # add on to tidyverse to merge tables on nearest values
 library(readxl)
 library(lubridate)
+library(ggmap)
 
 
 ################################################################################
@@ -17,26 +18,42 @@ library(lubridate)
 # UT Austin downloaded from TORCH
 AGHY_UTAustin <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/DigitizedHerbariumRecords/UTAustinAGHYrecords/occurrences.csv") %>%
   mutate(new_id = gsub("[a-zA-Z ]", "", catalogNumber)) %>%   #creates a new id that we will use to merge with the
-  mutate(municipality = as.character(municipality)) %>%
-  dplyr::select(id, catalogNumber, country, stateProvince, county, municipality, locality, decimalLatitude, decimalLongitude, coordinateUncertaintyInMeters, new_id, eventDate, day, month, year) %>%
+  mutate(municipality = as.character(municipality)) %>% 
+  dplyr::select(id, catalogNumber, country, stateProvince, county, municipality, locality, decimalLatitude, decimalLongitude, coordinateUncertaintyInMeters, recordedBy, recordNumber, new_id, eventDate, day, month, year) %>%
   mutate(Spp_code = "AGHY")
 ELVI_UTAustin <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/DigitizedHerbariumRecords/UTAustinELVIrecords/occurrences.csv") %>% 
   mutate(new_id = gsub("[a-zA-Z ]", "", catalogNumber)) %>%    #creates a new id that we will use to merge with the 
   mutate(municipality = as.character(municipality)) %>% 
-  dplyr::select(id, catalogNumber, country, stateProvince, county, municipality, locality, decimalLatitude, decimalLongitude, coordinateUncertaintyInMeters, new_id, eventDate, day, month, year) %>% 
+  dplyr::select(id, catalogNumber, country, stateProvince, county, municipality, locality, decimalLatitude, decimalLongitude, coordinateUncertaintyInMeters, recordedBy, recordNumber, new_id, eventDate, day, month, year) %>% 
   mutate(Spp_code = "ELVI")
 AGPE_UTAustin <-   read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/DigitizedHerbariumRecords/UTAustinAGPErecords/occurrences.csv") %>% 
[…]

### 2022-12-16 endo_herbarium_records_merge.R
still figuring out some collector name issues, and then need to fix merging issues for MOBOT

@@ -352,7 +352,8 @@ AGPE_BRIT <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/Digitiz
 BRIT_torch <- rbind(AGHY_BRIT, ELVI_BRIT, AGPE_BRIT) %>% 
   mutate(primary_collector = case_when(str_detect(recordedBy, "Jr.") ~ word(recordedBy, start = 1, end = 2, sep = fixed(",")),
                                        !str_detect(recordedBy, "Jr.") ~ word(recordedBy, start = 1, sep = fixed(","))),
-         primary_collector = word(primary_collector, sep = fixed("and"))) %>% 
+         primary_collector = word(primary_collector, sep = fixed("and")),
+         primary_collector = word(primary_collector, sep = fixed("&"))) %>% 
   mutate(collector_firstname = word(primary_collector),
          collector_lastname = case_when(str_detect(primary_collector,"Jr.") ~ word(str_trim(primary_collector), start = -2, end = -1),
                                         !str_detect(primary_collector,"Jr.") ~ word(str_trim(primary_collector), -1)))
@@ -460,7 +461,7 @@ LSU_records <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/Digit
   mutate(Spp_code = case_when(genus == "Elymus" ~ "ELVI",
                               genus == "Agrostis" & specificEpithet == "hyemalis" | specificEpithet == "hiemalis" ~ "AGHY",
                               genus == "Agrostis" & specificEpithet == "perennans" ~ "AGPE")) %>%  # there are some Agrostis scabra, that may need to be sorted out cause they could be part of hyemalis
-  dplyr::select(id, Spp_code, catalogNumber, country, stateProvince, county, municipality, locality, decimalLatitude, decimalLongitude, coordinateUncertaintyInMeters, recordedBy, recordNumber, eventDate, day, month, year) 
+  dplyr::select(id, Spp_code, catalogNumber, country, stateProvince, county, municipality, locality, decimalLatitude, decimalLongitude, coordinateUncertaintyInMeters, recordedBy, primary_collector, collector_lastname, collector_firstname, recordNumber, eventDate, day, month, year) 
 
 # Reading in the Oklahoma state university digitized records
[…]

### 2023-03-16 endo_herbarium_records_merge.R
calculating sd and cv in climate

@@ -1372,7 +1372,32 @@ tmean_spring_old_norm <- terra::mean(terra::rast(pd_stack(prism_archive_subset(t
 tmean_summer_old_norm <- terra::mean(terra::rast(pd_stack(prism_archive_subset(type = "tmean", temp_period = "monthly", year = 1895:1925, mon = 5:8))))
 tmean_autumn_old_norm <- terra::mean(terra::rast(pd_stack(prism_archive_subset(type = "tmean", temp_period = "monthly", year = 1895:1925, mon = 9:12))))
 
-# calculating the cumulative preciptation for each year and for each season within the year
+# calculating standard deviation in temp
+
+tmean_annual_recent_sd <- terra::stdev(terra::rast(pd_stack(prism_archive_subset(type = "tmean", temp_period = "monthly", year = 1990:2020))))
+tmean_spring_recent_sd <- terra::stdev(terra::rast(pd_stack(prism_archive_subset(type = "tmean", temp_period = "monthly", year = 1990:2020, mon = 1:4))))
+tmean_summer_recent_sd <- terra::stdev(terra::rast(pd_stack(prism_archive_subset(type = "tmean", temp_period = "monthly", year = 1990:2020, mon = 5:8))))
+tmean_autumn_recent_sd <- terra::stdev(terra::rast(pd_stack(prism_archive_subset(type = "tmean", temp_period = "monthly", year = 1990:2020, mon = 9:12))))
+
+tmean_annual_old_sd <- terra::stdev(terra::rast(pd_stack(prism_archive_subset(type = "tmean", temp_period = "monthly", year = 1895:1925))))
+tmean_spring_old_sd <- terra::stdev(terra::rast(pd_stack(prism_archive_subset(type = "tmean", temp_period = "monthly", year = 1895:1925, mon = 1:4))))
+tmean_summer_old_sd <- terra::stdev(terra::rast(pd_stack(prism_archive_subset(type = "tmean", temp_period = "monthly", year = 1895:1925, mon = 5:8))))
+tmean_autumn_old_sd <- terra::stdev(terra::rast(pd_stack(prism_archive_subset(type = "tmean", temp_period = "monthly", year = 1895:1925, mon = 9:12))))
+
+# calculating coefficient of variation in temp
+tmean_annual_recent_cv <- tmean_annual_recent_sd/tmean_annual_recent_norm
+tmean_spring_recent_cv <- tmean_spring_recent_sd/tmean_spring_recent_norm
[…]

### 2023-03-16 endo_herbarium_spatial_analysis.R
adding more loc to the year trend, and updating the climate correlations

@@ -27,23 +27,33 @@ endo_herb_georef <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/
   # filter(Country != "Canada") %>%
   mutate(Spp_code = case_when(grepl("AGHY", Sample_id) ~ "AGHY",
                               grepl("ELVI", Sample_id) ~ "ELVI",
-                              grepl("AGPE", Sample_id) ~ "AGPE"))
+                              grepl("AGPE", Sample_id) ~ "AGPE")) %>% 
+  mutate(species = case_when(Spp_code == "AGHY" ~ "A. hyemalis",
+                             Spp_code == "AGPE" ~ "A. perennans",
+                             Spp_code == "ELVI" ~ "E. virginicus")) 
 
 endo_herb_AGHY <- endo_herb_georef %>% 
   filter(!is.na(Endo_status_liberal)) %>% 
   filter(Spp_code == "AGHY") %>% 
   filter(!is.na(lon) & !is.na(year)) %>% 
-  filter(lon>-110)
+  filter(lon>-110) %>% 
+  filter(Country != "Canada" & !is.na(County)) 
+  
 endo_herb_AGPE <- endo_herb_georef %>% 
   filter(!is.na(Endo_status_liberal)) %>% 
   filter(Spp_code == "AGPE") %>% 
   filter(!is.na(lon) & !is.na(year)) %>% 
-  filter(lon>-110)
+  filter(lon>-110) %>% 
+  filter(Country != "Canada" & !is.na(County))  
+  
 
 endo_herb_ELVI <- endo_herb_georef %>% 
   filter(!is.na(Endo_status_liberal)) %>% 
   filter(Spp_code == "ELVI") %>% 
-  filter(!is.na(lon) & !is.na(year)) 
+  filter(!is.na(lon) & !is.na(year)) %>% 
+  filter(lon>-110) %>% 
+  filter(Country != "Canada" & !is.na(County))
+  
 
 endo_herb <- endo_herb_georef %>% 
   filter(!is.na(Endo_status_liberal)) %>%
@@ -51,13 +61,17 @@ endo_herb <- endo_herb_georef %>%
   filter(!is.na(lon) & !is.na(year)) %>% 
   filter(lon>-110 ) %>% 
   filter(Country != "Canada" & !is.na(County)) %>% 
-  mutate(species = case_when(Spp_code == "AGHY" ~ "A. hyemalis",
[…]

### 2023-03-16 Manuscript/endo_herbarium.tex
adding updates to the manuscript and figures. Will keep updating, but getting feedback from Tom.

--- 2022-10-19 Manuscript/endo_herbarium.tex
@@
 \usepackage{titlesec}
+
+\usepackage{graphicx,float}
+% for setting up equations
+\usepackage{amsmath}
+
 \titleformat{\section}[block]{\Large\bfseries\filcenter}{\thesection}{1em}{}
@@
 
-\title{Idiosyncratic responses of microbial mutualisms over 150 years of climate change}
+\title{Increasing prevalence of plant-fungal symbiosis over 195 years of climate change}
 
@@
 \author{Joshua C. Fowler$^{1,\ast}$ \\
-	Generic H. Collaborator$^{2}$ \\
 	Tom E. X. Miller$^{1}$}
@@
 	\noindent{} 1. Rice University, Department of BioSciences, Houston, Texas 77006;
-
-	\noindent{} 2. University of;
 
@@
 	\section*{Abstract}
-
-	Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Sed non risus.
-Suspendisse lectus tortor, dignissim sit amet, adipiscing nec, ultricies sed, dolor.
-Cras elementum ultrices diam.
-Praesent quis dolor in dolor molestie cursus et ac nisi.
-Vestibulum ante purus, semper eget est vitae, vehicula ornare nisl.
-Morbi efficitur euismod enim, nec feugiat tellus cursus eget.
+Climate change has altered species’ phenologies, has triggered changes in traits, and has led to shifts in abundance and distributions. 
+Yet, we still know little about how biotic interactions alter these environmental responses over long time scales across spatially heterogenous changes in climate. 
+Most species host beneficial microbial symbionts, which may provide resilience to changes in the environment, but at the same time increasing stress may lead these symbioses to breakdown. 
+Microbes can be preserved within the dried plant tissue of herbarium specimens presenting an opportunity to quantify changes across broad temporal and spatial scales. 
+Epichloë fungal endophytes are seed-transmitted symbionts of cool-season grasses which can provide drought tolerance to their hosts. 
+ We ask how endophyte prevalence has responded to climate change. 
+ Specifically, we took samples of seeds from herbarium specimens of three grass host species collected over the last two centuries, quantified the presence or absence of fungal endophytes within the seeds with microscopy, and evaluated spatial and temporal trends in prevalence using spatial modeling.
+ We found strong evidence that fungal endophyte prevalence has increased over the last two centuries across the three host species from 25\% prevalence to 75\% prevalence on average.
+We also found that observed changed in annual  and seasonal climate drivers were associated with temporal trends in endophyte prevalence. 
+In particular, changes in prevalence within each species corresponded to climate drivers during its peak growing season. 
+Thus we provide novel evidence for adaptive biotic responses to ongoing climate change.
 
@@
 	% (e.g. section ``Counting Owl Eggs'').
-Global climate change involves hetergenous changes in multiple climate drivers \citep{IPCC2012managing}. 
-It is of great interest to predict species' responses to these drivers.
-Novel conditions .Predicting how species' respond to these drivers, it is important to understand how and under what conditions interactions between species will alter expected responses. 
-Incorporating interactions may lead to different predictions than those based on single species responses; for most species, we do not know whether their interactions will facilitate or hinder movement and adaptation to novel conditions \cite{gilman2010framework}.
+Predicting how species' interactions will be altered by ongoing climate change is a major goal of basic ecological research and conservation management. 
+Spatially heterogeneous changes in the environment have led species to move, adapt, or face extinction \citep{IPCC2012managing}. 
+Species' responses are likely mediated by the numerous interspecific interactions in which they engage. 
+Yet for most species, empirically evaluating the role that biotic interactions can play in facilitating or hindering adaption to changing environmental conditions has been limited by a lack of data at adequate spatial and temporal scales \cite{gilman2010framework}.
+
 Natural history collections, which were originally collected to study and preserve taxonomic diversity, present an exciting opportunity to explore long-term changes in ecological interactions across broad spatial scales \citep{meineke2018unrealized}.
-These collections primarily comprise physical specimens of organisms collected along with information about the time and place of collection. 
-These natural history collections, built and maintained by the contributions of thousands of scientists, are invaluable time machines. 
-The specimens they contain are samples from dynamically changing environmental conditions through time and space.
-These specimens can preserve physical legacies of ecological dynamics and species' interactions. 
-Recent work using collections has documented changes in phenology \citep{willis2017old, park2019herbarium, berg2019examination}, as well as pollination \citep{pauw2011reconstruction, duan2019century} and changes in negative interactions such as herbivory \citep{meineke2019herbarium}, but to date, focus has been lacking on long-term changges in reciprocally positive interactions, i.e.
-mutualisms.
+Natural history collections, built and maintained by the efforts of thousands of scientists, are invaluable time machines. 
+These collections primarily comprise physical specimens of organisms  along with information about the time and place of their collection. 
+These specimens are samples preserving physical legacies of ecological dynamics and species' interactions from dynamically changing environments across time and space.
+Recent works using plant collections have documented changes in phenology \citep{willis2017old, park2019herbarium,  berg2019examination}, as well as pollination \citep{pauw2011reconstruction, duan2019century}, and changes in antagonistic interactions such as herbivory \citep{meineke2019herbarium}, but to date, focus has been lacking on long-term changes in a particularly common type of biotic interaction, microbial symbioses.
 
-Given that mutualists alter their partner's fitness, they may play an important role in limiting environmental stress, but changing conditions may also disrupt or breakdown the mutualism \citep{aslan2013mutualism}.
-Examples of mutualism disruption as a result of changing climates include phenological mismatches between plants and their pollinators \citep{pauw2011reconstruction}, drought-driven breakdown in seagrass-lucinid mutualisms \citep{de2016drought}, and coral bleaching due to temperature stress \citep{sully2019global}, all of which can have important consequences for habitat and biodiversity maintainence.
-Alternatively, mutualists may facilitate adaptive responses by improving fitness under stressful conditions, such as facultative bacterial symbionts within insects implicated in improving thermal tolerance \citep{truitt2019wolbachia, renoz2019evolutionary}, or fungal symbionts that allow their grass host to persist in more arid edges of its distribution \citep{afkhami2014mutualist}.
-Whether mutualisms breakdown or promote presistence under climate change is unclear, but likely depends on the intimacy and specialization of the interaction, as well as the physiological tolerances of the mutualists\citep{toby2010mutualisms, warren2014mutualism, rafferty2015phenological}.
-Given that climate change is spatially heterogenous, the rate of change may also have a strong influence on when and where we can expect mutualism facilitation or disruption.
+Microbes are hosted by all organisms with documented effects on their hosts' survival, growth and reproduction \cite{rodriguez2009fungal,mcfall2013animals}.
+Many microbial symbionts act as mutualists, engaging in reciprocally beneficial interactions with their hosts. 
+These mutualisms play an important role in limiting the consequences of environmental stress. 
+For example, bacterial symbionts of insects, such as \emph{Wolbachia}, can improve their hosts' thermal tolerance \citep{truitt2019wolbachia, renoz2019evolutionary}.
+Similarly, arbuscular mycorrhizal fungi, documented in 70-90\% of families of land plants \citep{parniske2008arbuscular}, allow their hosts to persist through drought conditions by improving water and nutrient uptake \citep{cheng2021elucidating}.
+At the same time, changes in the mean and variance of environmental conditions may disrupt mutualisms by changing the costs and benefits of the interaction for each partner, leading to the extinction of one or both partner species \citep{aslan2013mutualism}.
+Coral bleaching, the loss of symbiotic algae due to temperature stress \citep{sully2019global} is perhaps the best known example, but this phenomenon is not unique to corals.
+Lichens exposed to elevated temperatures experienced loss of photosynthetic function along with changes in the community composition of their algal symbionts \citep{meyer2022climate}.
+Whether symbiotic mutualisms breakdown or promote resilience under climate change is unclear, but likely depends on the intimacy and specialization of the interaction, as well as the physiological tolerances of the mutualists \citep{toby2010mutualisms, warren2014mutualism, rafferty2015phenological}.
+Given that climate change is spatially heterogenous, the rate of change may have a strong influence on when and where we can expect facilitation or disruption of mutualism.
 
-Early research on \emph{Epichloë} fungal endophytes, symbionts of cool-season grasses, used herbarium specimens to describe the broad taxonomic diversity of hosts \citep{white1985endophyte}. 
-These endophytes are vertically-transmitted, growing throughout the aboveground tissue and into the seeds of their hosts, and consequently are preserved along with the plant on herbarium sheets. 
-Vertical transmission links the fitness of host and symbiont together meaning that, over time, endophytes that provide fitness benefits should rise in prevalence within the population \citep{fine1975vectors, douglas1998host, rudgers2009fungus}.
-This leads to the prediction that interaction prevalence should generally be high where there are strong fitness benefits, and that prevalence should track changes in environmental drivers that limit fitness.
-Environmental stress is likely particularly consequential at species' range edges wherer population growth rates are expected to be near replacement levels \cite{}. 
-And so understanding the ways that endophytes may mediate stress is important for understanding the factors setting species' range limits. 
-The interaction prevalence should respond to the fitness effects of the mutualism, and the prevalence of endophytes across the range determines their actual contribution to setting range limits. 
-In addition, the rate of environmental change is variable across space across species ranges \citep{IPCC2012managing}, and local changes in endophyte prevalence likely depend on these rates.
+Early research on \emph{Epichloë} fungal endophytes, specialized symbionts of cool-season grasses, used herbarium specimens to describe the broad taxonomic diversity of hosts \citep{white1985endophyte}.
+These facultative symbionts, which have been documented in $\sim 30$\% of cool-season grasses \citep{leuchtmann1992systematics}, are transmitted vertically from maternal plants to offspring through the grasses' seeds.
+Vertical transmission creates a feedback between the fitness of host and symbiont \citep{fine1975vectors, douglas1998host, rudgers2009fungus}. 
+Over time, endophytes that act as mutualists for their hosts should rise in prevalence within a population. 
+\emph{Epichloë} are known to improve drought tolerance for their hosts \citep{afkhami2014mutualist}, and are expected to reach high prevalence where these fitness benefits are most important.
+Previous survey studies have documented large-scale spatial patterns in  endophyte prevalence structured by environmental gradients \citep{granath2007variation,bazely2007broad, afkhami2012fungal,sneck2017variation}.
+We predicted that prevalence should also track temporal changes in environmental drivers, but testing this has been impossible with surveys limited to a few years, and to the spatial extent reachable during a single growing season.
+Becauses the fungi are seed-transmitted, they are conveniently preserved along with their host plants on herbarium sheets, creating the opportunity to examine the response of the symbiosis to historical climate change.
 
-And so, we are taking advantage of the spatial and temporal sampling within herbaria to ask how change in spatial and temporal environmental drivers alter the prevaence of a widespread mutualistic fungal endophyte symbiosis between Epichloë fungi and grasses.
-Specifically,  we ask (1) how variation in temperature and precipitation correlate with endophyte prevalence across space and over time, and (2) does greater rates of envirnmental change drive greater changes in endophyte prevalence.
+In this study, we took advantage of the spatial and temporal sampling within herbaria to assess for the first time whether grass-fungal endophyte symbioses have responded adaptively to climate change.
+In particular, we (i) quantified the spatial and temporal change in endophyte prevalence across host ranges over the last two centuries, and (ii) identified climate drivers that can explain the magnitude of these responses.
 
 \section*{Methods}
+        \subsection*{Focal host species}
+We focused on three endophyte hosts, \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}.
+These grass species, which are all cool-season (C3) grasses, are commonly represented in natural history collections with broad distributions covering the Eastern United States.
+\emph{A. hyemalis} is a small short-lived perennial species that germinated in the spring and typically blooms between March and July.
+\emph{A. perennans} of similar stature but blooms in late Summer and early Autumn. 
+This species is relatively more  sparsely distributed, tending to be found in shadier and more moist habitats while \emph{A. hyemalis} is often found in open and recently disturbed ground. 
+Both\emph{Agrostis} species are recorded from throughout the Eastern US, but \emph{A.
+perennans} has a slighty more northern distribution, whereas \emph{A.
+hyemalis} is found rarely as far north as Canada and is listed as a rare plant in Minnesota.
+\emph{E. virginicus} is a larger and relatively longer-lived  species that is more broadly distributed that the \emph{Agrostis} species. 
+It begins flowering as early as March or April but continues throughout the Summer.
 
-I will quantify the spatial and temporal responses of fungal endophyte symbiosis to environmental conditions by assessing the status of endophyte symbiosis in herbarium specimens. 
-In particular, I will collect a small sample of seeds from herbarium specimens of \emph{Agrostis hyemalis} and \emph{Elymus virginicus} from across their distributions, and use microscopy to identify the endophyte status of specimens.
-	\subsection*{Second-order heading}
+		\subsection*{Herbarium surveys}
+We collected 5-10 seeds per specimen during visits to natural history collections between 2019 and 2022 (See Table A1 for a summary of collections) . 
+Grasses are commonly identified and collected based on the presence of their reproductive structures which ensured that specimens we examined contained seeds. 
+Our sampling plan was designed to minimize damage to these specimens.
 
-	Lorem ipsum nulla facilisi, pace \citet{LemKapEx07}.
-Etiam semper, orci sit amet facilisis interdum, tellus nunc consequat erat, quis viverra nisi diam ut metus.
-Pellentesque cursus, sapien malesuada euismod iaculis, mauris purus interdum diam, vel vestibulum justo enim vitae tellus.
-Nunc interdum lorem sit amet diam volutpat tristique.
-Quisque pulvinar ac metus commodo lacinia (\citealt{Ing11,Xiao2015}).
+\begin{figure}[H]
+	\centering
+	\includegraphics[width = \linewidth]{collections_map.png}
+	\caption{\textbf{Collection locations of herbarium specimens sampled for endopnhyte presence absence}.
+Color designates host species ( \emph{A.
+hyemalis} (green), \emph{A.
+perennans} (orange), \emph{E.
+virginicus} (purple))}
+\end{figure}
 
-	\subsubsection*{Third-order heading}
+After collecting seeds, we quantified the presence or absence of \emph{Epichloë} fungal hyphae, which grow intercellularly, using microscopy.
+We first softened seeds with a 10\% NaOH solution, then stained the seeds with aniline blue dye and squashed them under a microscope cover slip.
+We examined the squashed seeds for the presence of fungal hyphae at 100X magnification \cite{bacon2018stains}.
+We designated each specimen as symbiotic or non-symbiotic, a binary response variable, based on the presence of endophytes within any of the scored seeds for that specimen. 
+In some cases, the tissues examined during microscopy came from flowers or otherwise non-viable seeds, which were excluded from the seed counts for that specimen.
+To capture uncertainty in the identification process, we recorded both a "liberal" and a "conservative" endophyte status for each plant.  
+When we identified potential endophytes with unusual morphology, low uptake of stain, or a small amount of fungal hyphae across the scored seeds, we recorded a positive liberal status (more likely to be a true endophyte) and a negative conservative status (less likely to be a true endophyte).
+$89$\% of scored plants had matching liberal and conservative endophyte statuses, reflecting high confidence in the endophyte identification.
+The following analyses presented in the main text used the liberal statuses, but we repeated all analyses with the conservative statuses which yielded qualitatively similar results (Appendix).
 
-	Usually two or three levels of heading will be all you need. Journal style even permits a fourth level in case you need it.
+We excluded specimens that did not include information about the collection location and date from our sampling efforts.
+In total, we quantified endophyte symbiosis for $971$ \emph{A. hyemalis}, $302$ \emph{A. perennans}, and $605$ \emph{E. virginicus} specimens spanning the years $1824$ to $2019$.
+Each specimen was assigned a collection location based on geographic information recorded at the time of collection using the ggmap package in R \citep{kahle2019package}. 
+Collections were georeferenced to the nearest county centroid, or nearest municipality when that information was available.
+For a few of the oldest specimens, only information at the state level was available, and so we used the state centoid (Fig.
+1; Fig A1).
 
-	\paragraph*{Fourth-order heading}
-	The quick red fox jumps over the lazy brown dog in this paragraph as well. Donec mauris nibh, volutpat vehicula viverra at, iaculis congue sem. Praesent eget erat rhoncus erat sollicitudin volutpat.
+\subsection*{Assessing spatial and temporal changes in endophyte prevalence}
+To quantify spatial and temporal trends in endophyte prevalence, we modelled endophyte presence/absence ($P$) as a Bernoulli distribution for specimen $i$ of host species $h$.
 
-	\begin{equation}
-	{ \frac{1}{N_k-1} \sum \limits_{t=1}^{N_k} (M_{tjk} - \bar{M}_{jk})^2}
-	\end{equation}
+\begin{subequations}
+		\label{eq:trends}
+		\begin{align}
+		P_{i,h} \sim Bernoulli(\hat{P_{h}}) \\
+			logit(\hat{P}_{h}) = \beta_{0,h} + 
+		\beta_{1,h}*year_{i} + \\
+		\beta_{2,h}*year_{i} *lat_{i} + 
+		\beta_{3,h}*year_{i} *lon_{i} + 
+		\beta_{4,h}*year_{i} *lat_{i}*lon_{i} +\\
+		\chi + \omega + \phi
+		\end{align}
+\end{subequations}
 
-	Praesent quis dolor in dolor molestie cursus et ac nisi.
-Vestibulum ante purus, semper eget est vitae, vehicula ornare nisl.
-Morbi efficitur euismod enim, nec feugiat tellus cursus eget.
-Donec mauris nibh, volutpat vehicula viverra at, iaculis congue sem.
+The expected endophyte prevalence, $\hat{P}$, was modelled with intercepts specific to each host species ($\beta_{0}$), slopes for changes over time ($\beta_{1}$) as well as the interaction between time and the specimen's latitude and longitude ($\beta_{2}$, $\beta_{3}$, and $\beta_{4}$).
+We accounted for potential biases introduced during the process of collecting specimens as well as in scoring ability by including random effects specific to each collector $\chi$ and to each scorer $\omega$.
+To quantify spatial patterns in prevalence, we included a spatially-dependent random intercept, $\phi$, and fit the model using integrated nested Laplace approximation, an approximate Bayesian method implemented in the rINLA package \citep{lindgren2015bayesian}.
+This method allowed us to model spatial dependency between data points as a continuous process with stochastic partial differential equations (SPDE) that depend on a covariance matrix according to the proximity of each collection location \citep{lindgren2011explicit,bakka2018spatial}.
+The covariance matrix is approximated using a Matérn covariance function, with each data point assigned a location according to the nodes of a mesh of non-overlapping triangles across our study area (Fig A2).
+We fit the model with vague priors, and compared models with different sizes of mesh, which had little effect on the resulting model estimates.
+We assessed model fit with graphical postterior predictive checks (Fig. A3, Fig. A4).
 
-	\subsection*{Another second-order heading}
+We used data on endophyte prevalence from  contemporary surveys of \emph{A. hyemalis} as test data to evaluate model fit. 
+During these surveys, which took place between 2015 and 2020, we collected seeds from living plants from 43 populations across a portion of the species' range and then quantified endophyte status with staining microscopy as described for the herbarium surveys (Fig S2).
+We compared the model's prediction for these locations to the endophyte status of randomly sampled plants from each surveyed location, mimicing the process of collecting a single herbarium specimen from the population.
+The model performed adequately (training data: AUC = 0.77; test data: AUC = 081; Fig. A4).
 
-	As \citet{Xiao2015} argued, phasellus porttitor eros et ante condimentum, eget facilisis orci condimentum.
-Nulla facilisi.
-Proin placerat elit blandit, euismod dolor nec, dapibus diam.
-Mauris posuere malesuada lacus, at elementum lacus auctor eu (fig~\ref{Fig:Jumps}A).
+		\subsection*{Assessing the role of climate drivers}
+To assess how the rate of climate change may have driven changes in endophyte prevalence, we calculated the change in climate between recent (1990 to 2020) and historic (1895 to 1925) periods.
+We first downloaded monthly temperature and precipitation rasters from the PRISM climate group \citep{daly2013prism} covering the time period between 1895 and 2020 using the 'prism' R package \citep{Rprism2015}.
+Prism provides reconstructions of historic climate variables across the United States by spatially-interpolating weather station data \citep{diLuzio2008constructing}. 
+We calculated 30-year climate normals for annual and seasonal mean temperature and cumulative precipitation for the recent and historic periods.
+We used three four-month seasons within the year (Spring: January, February, March, April; Summer: May; June, July, August; Autumn: September, October, November, December). 
+We additionally calculated the coefficient of variation of each climate driver during each period.
+We then took the difference between recent and historic periods for the mean and coefficient of each climate driver (Fig. A5).
+Because initial analyses indicated a high degree of collinearity between seasonal and annual changes in temperature, we used annual temperature only, along with annual and seasonal precipitation, in the subsequent analysis.
+
+We then evaluated how predicted changes in endophyte prevalence correlated with changes in thee environment at an evenly spaced grid of points across the study area. 
+Because regions varied in their predicted starting endophyte prevalence, we calculated the relative change in endophyte prevalence between 1925 and 2020. 
+This time period connects the predicted change in prevalence with the endpoints of the available climate record.
+We then calculated the Pearson correlation coefficient between the relative change in endophyte prevalence and each climate driver.
+
+%We evaluated how predicted changes in endophyte prevalence ($\Delta{P}$) correlated with %changes in seasonal climate at a grid of points across the study area using linear regression.
+
+%\begin{subequations}
+	%\label{eq:climate}
+	%\begin{align}
+		%\Delta{P} \sim Normal(\mu, \sigma)\\
+		%\mu = \beta_{0} + \alpha_{1}*\Delta{Precip_{spring}} + %\alpha_{2}*\Delta{Precip_{summer}} + \alpha_{3}*\Delta{Precip_{autumn} }+ \\ %\alpha_{4}*\Delta{Temp_{spring}} + \alpha_{5}*\Delta{Temp_{summer}}  + %\alpha_{6}*\Delta{Temp_{autumn}}
+%	\end{align}
+%\end{subequations}
+
+ %The change in prevalence is determined by an intercept $\beta_{0}$ slopes specific to the seasonal %changes in preciptation ($\alpha_{1}$, $\alpha_{2}$, and $\alpha_{3}$) and in mean %temperature ($\alpha_{4}$, $\alpha_{5}$, and $\alpha_{6}$).
+ %We fit models for each host species separately with rINLA using default vague priors.
 
@@
 
-	Lorem ipsum dolor sit amet.
-Aenean pulvinar malesuada commodo (see \citealt{DavisEtAl2011}; table~\ref{Table:Founders}).
-Sed aliquet mauris odio, in tristique dui egestas a.
-Etiam eu malesuada quam.
-Suspendisse tincidunt eu erat sit amet vulputate.
-Duis at arcu et nisl dictum mattis.
-Maecenas vel cursus ante.
-Cras eleifend elit nec velit sollicitudin fermentum in ac mauris.
-Pellentesque rutrum magna vel elit maximus hendrerit.
+\subsection*{Temporal and spatial trends}
 
-	\subsection*{The height of the jump}
+We found that endophyte prevalence increased within the examined specimens over the last two centuries for all three host species (Fig. 2). 
+On average, \emph{A.
+hyemalis} and \emph{E.
+virginicus} both increased from ~30 \% to over 50\% prevalence across the study region, and \emph{A.
+perennans} increased from ~ 15\% to over 70\% prevalence.
+Rerunning the analysis excluding specimens collected before 1900 when samples are much sparser (Fig. 2A), lead to qualitatively similar predictions, and so we continued analyses with the full dataset.
 
-	Aenean eu pellentesque quam (fig.~\ref{Fig:OkapiHorn}).
-Nam pellentesque augue eu finibus lacinia.
-Nullam nec justo vitae odio imperdiet rhoncus vitae vitae quam.
-Pellentesque porttitor metus et lectus ornare, ac cursus urna efficitur (fig~\ref{Fig:Jumps}B).
+\begin{figure}[H]
+	\centering
+	\includegraphics[width = \linewidth]{year_plot.png}
+	\caption{\textbf{Temporal trends in endophyte prevalence.} (A) Histograms show the frequency of collection through time for each host species.
+(B) Colored points are binned means of the observed endophyte presence/absence data (black dashes).
+Colors represent each host species and point size is determined by the number of specimens.
+Lines show predicted mean endophyte prevalence over the study period at low (dotted) high (solid) latitudes and along with 95\% CI bands.}
+\end{figure}
 
-	\subsection*{The laziness of the dog}
+Across space, there were clear trends in endophyte prevalence which varied between species (Fig. 3). 
+For example, \emph{Elymus virginicus}, had higher prevalence in the northern and eastern portions of it's range. 
+In contrast, \emph{Agrostis hyemalis} had highest prevalence towards the center of its range with regions of low prevalence towards the north east and towards its western range edge.
+Our model revealed that while there was an overall increase in endophyte prevalence, these changes varied across the host species' ranges.
+In some regions, \emph{A.
+perennans} experienced increases in percent prevalence by as much as 35 percentage points between 1920 and 2020, while \emph{A.
+hyemalis} and \emph{E.
+virginicus} experienced increases up to around 15 percentage points.
+In other regions, there were smaller increases of only five percentage points for \emph{E. virginicus} and fifteen percentage points for \emph{A. perennans}. 
+Notably, the western range edge of \emph{A. hyemalis} did not experience an increase in endophyte prevalence. 
+Collections are sparser to the west as the species reaches its range limit, but the collected specimens were exclusively symbiont-free. (Fig. A1).
 
-	Example paragraph with embedded references (video~\ref{VideoExample}, fig.~\ref{Fig:AnotherFigure}). 
-	If you have deposited data to Dryad, it is advisable to cite them somewhere in the main text (usually in the Methods or Results sections).
-A sentence like the following will do: All data are available in the Dryad Digital Repository (\citealt{CookEtAl2015}).
+\begin{figure}[H]
+	\label{fig:prevalence_map}
+	\centering
+	\includegraphics[width = \linewidth]{prevalence_map.png}
+	\caption{\textbf{Mean predicted endophyte prevalence for each host species (columns) in 1920 (top row) and 2020 (bottom row)}. Color indicates mean predicted rate of endophyte prevalence.}
+\end{figure}
+
+\subsection*{The role of climate drivers}
+We found that the trends in endophyte prevalence were strongly associated with seasonal climate change drivers (Fig. 4). 
+For the majority of the study region, the climate has become wetter and cooler over the last century (Fig. A5), a consequence of regional variation in global climate change. 
+Within the study region, spatially heterogeneous environmental changese were predictive of changes in endophyte prevalence in line with differences in species' reproductive timing. 
+For example, relative change in endophyte prevalence within \emph{E.
+virginicus} were most associated with declines in Summer precipitation  (a negative correlation in Fig.
+4) as well as with increases in the year-to-year variability of summer precipitation (a positive correlation in Fig.
+4).
+These associations are in line with expectations that endophyte symbiosis provides drought tolerance to their hosts, particularly during \emph{E. virginicus}'s summer growing season. 
+For \emph{A.
+perennans}, which typically flowers in the late Summer and Autumn, increases in autumn precipitation along with increases in variability in annual precipitation, were strongly assocated with increasing endophyte prevalence.
+While this result is contrary to the expectation that endophyte provide drought tolerance, this species typically is associated with wetter microhabitats, and it also suggests a role for endophytes to play in buffering the negative consequences of demographic variability \citep{lewontin_population_1969}.
+For \emph{A. hyemalis}, which flowers in the Spring, changes in endophyte prevalence were more weakly associated with seasonal changes in climate drivers than for the other two species.
+It was, however, the only species' for which greater increases in endophyte prevalence were associated with regions that experienced increases in average annual temperatures.
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width = .8\linewidth]{climate_corr_heatmap.png}
+	\caption{\textbf{Changes in climate drivers associated with relative change in endophyte prevalence.} Color denotes the Pearson correlation coefficient between the relative rate of change in endophyte prevalence and the change in annual mean temperature ($^oC$) and cumulative seasonal precipitation (mm), as well as the change in coefficient of variation of each climate driver.}
+\end{figure}
 
 \section*{Discussion}
+Our examination of historic plant specimens revealed a cryptic biotic reponse to climate change. 
+The geographic distributions and phenology of macroorganisms are apparent indicators of climate change.
+For the three host species we examined, there have been clear increases in fungal endophyte prevalence over the last two centuries.
+This increase suggests that \emph{Epichloë} fungal endophytes are playing key adaptive roles in response to climate change and that the mutualism between grasses and their fungal endophytes has contributed to resilience for each partner species under the observed changes in climate.
+Spatial and temporal trends in endophyte prevalence varied between each species in ecologically meaningful ways.
+Our spatially-explicit model predicted regions of both high and low endophyte prevalence, suggesting that symbiotic and non-symbiotic host plants have overlapping, but non-identical niche requirements.
+Endophytes may expand the range of their hosts by allowing them to persist in environments where they otherwise could not \citep{afkhami2014mutualist, kazenel2015mutualistic}.
+Understanding how microbial symbionts respond to climate change is a first step towards incorporating reciprocal host-symbiont interactions into predictions host range shifts.
 
-are a potentially rich target for study within musuem specimens\citep{}.
-Practically all specimens will be preserved alongside their microbiome, and while some evidence of these association can be expected to be lost over time due to specimen handling and storage.
+For \emph{E. virginicus}, we found greater increases in prevalence in regions that have become drier during the Summer, the species' peak growing season. 
+Fitness benefits of the symbiosis under drought conditions could explain this increase and also explain high predicted prevalence towards the species' northern range edge coinciding with a strong latitudinal decline in precipitation.
+Previous population surveys for endophytes in this species found similar latitudinal trends in prevalence \citep{sneck2017variation,rudgers2009benefits}, but at smaller scales. 
+Following previous work demonstrating drought benefits in a greenhouse manipulation with \emph{A.
+hyemalis} \citep{davitt2011understanding}, we had predicted  that endophyte prevalence should similarly increase at a greater rate in regions that have seen reduced precipitation, yet we found the opposite trend in our herbarium surveys and that changes in prevalence for this species had the weakest correlations with climate drivers.
+Given that the study region has experienced relatively moderate changes in precipation and in temperature, potentially the magnitude of climate change has not been great enough to cause larger changes in endophyte prevalence for this species.
+Weak associations with drought could also be explained by climate-driven changes in the rate of imperfect transmission (the generation of non-symbiotic offspring from symbiotic hosts), which could counterbalance endophyte-mediated fitness benefits, and leading to stable intermediate prevalence rates \citep{donald2021context}.
+To our knowledge, the response of the symbiosis in \emph{A.
+perennans} to drought has not been examined experimentally, but in a separate greenhouse experiment, endophytes had a positive effect on reproduction under shaded, low-light conditions \citep{davitt2010costs}.
+\emph{Epichloë} endophytes have been connected to a suite of non-drought related fitness benefits including herbivore protection \citep{brem2001epichloe}, salinity resistence \citep{wang2020effects}, and mediation of the soil microbiome \citep{roberts2015rhizosphere}, potentially mediated by the diverse bioactive alkaloids and other signaling compounds they produce \citep{saikkonen2013chemical}.
+The strong increase in symbiotic \emph{A. perennans} could be explained by these diverse benefits. 
+Differences between the responses of each species underscore that while all of these C3 grasses share similar broad-scale distributions, each engages in unique biotic interactions and has unique niche requirements.
 
-	\section*{Conclusion}
+Our analysis advances the use of herbarium specimens in two ways. 
+First and foremost, this is the first study to link long-term changes in microbial symbioses to changes in climate using specimens from natural history collections.
+The responses of microbial symbioses are a rich target for future studies within museum specimens, particularly those that take advantage of advances in sequencing technology.
+While we used relatively coarse presence/absence data based on fungal morphology, other studies have examined historic plant microbiomes using molecular sequencing and sophisticated bioinformatics techniques, but these studies have so far been limited to relatively few specimens at limited spatial extents \cite{yoshida2015computational, heberling2019utilizing, bieker2020metagenomic, gross2021hidden, bradshaw2021global}.
+Continued advances in capturing historic DNA and in filtering out potential contamination during specimen storage \citep{daru2019novel, bakker2020herbarium, raxworthy2021mining} will be imperative in the effort to scale up these efforts.
+This scaling up will be essential to be able to quantify changes not just in the prevalence of symbionts, but also in symbionts' intraspecific variation and evolutionary responses to climate change, as well as  in changes in the wider microbial community.
+Answering these questions as well as the unknown questions that future researchers may ask also reiterates the value in capturing meta-information during ongoing digitization efforts at herbaria around the world and during the accession of newly collected specimens \citep{lendemer2020extended}.
+Second, we accounted for several potential biases in the data observation process that may be common to many collections-based research questions by using a spatially-explicit random effects model. 
+Spatial autocorrelation \cite{willems2022forest}, potential biases introduced by the sampling habits of collectors \citep{daru2018widespread}, and variation between contemporary researchers during the collection of trait data, if not corrected for could lead to biased inference about the strength and direction of historic change.
+Fitting this model in a Bayesian framework allows for full propagation of uncertainty.
 
-	Duis pharetra enim at libero cursus, eu commodo mi vestibulum.
-Nullam eget velit nec lectus viverra sodales.
-Suspendisse egestas, eros at dictum tincidunt, mi orci laoreet libero, eget rutrum sapien arcu blandit odio.
+Ultimately, a central goal of global change biology is to generate predictive insights into the future of natural systems. 
+While this survey of historic endophyte prevalence is necessarily correlative, it serves as a foundation to develop better predictive models of the response of microbial symbioses to climate change. 
+Combining the insights from this sort of regional-scale survey with field experiments and physiological data could be invaluable. 
+While we found that climate is strongly correlated with endophytes' temporal responses, we do not know why trends in prevalence were weak in some areas or how endophytes would respond to more extreme changes in climate.
+For example, transplanting symbiotic and non-symbiotic plants beyond the range edge of \emph{A.
+hyemalis} could tell us whether persistent lack of endophytes in that area is a result of environmental conditions that lead the symbiosis to negative fitness consequences, or is a result of some historical contingency or dispersal limitation that has thus far limited the presence of symbiotic hosts from a region where they would otherwise flourish and provide resilience.
+While we observed evidence of mutualism resilience, more extreme environmental changes than those observed in our study could potentially push one or both partners beyond their physiological limit, leading to the collapse of the mutualism.
+Our analysis thus far is agnostic to changes in the distributions of hosts. 
+Mechanistic models could connect the responses of both host and symbionts to abiotic climate drivers integrating dispersal processes. 
+Beyond host-microbe symbioses, building these types of models would work towards quantitatively attributing biotic responses to anthropogenically driven climate change, similar to methods in climate science and economics \citep{stott2010detection, carleton2016social}.
+%Elymus genotype
+%Agrostis weedy
 
@@
 	\section*{Acknowledgments}
+	We thank Jessica Budke for help in drafting our initial destructive sampling plan, and to the many members of herbarium staff who facilitated our research visits, as well as to the hundreds of collectors who contributed to the natural history collections.
+	This research was supported by funding from NSF grant () and from funding from the Texas Ecolab Program.
 
@@
 	\section*{Appendix A: Additional Methods and Parameters}
+	\renewcommand{\thefigure}{A\arabic{figure}}
+	\setcounter{figure}{0}
+
+		\renewcommand{\thetable}{A\arabic{table}}
+	\setcounter{equation}{0}  % reset counter 
+	\setcounter{figure}{0}
+	\setcounter{table}{0}
+
+	\begin{figure}[H]
+		\centering
+		\includegraphics[width = \linewidth]{endo_status_map.png}
+		\caption{\textbf{Endophyte presence/absence in specimens of each host species.} Points show collection locations colored according to whether the specimen contained endophytes ( E+; blue points) or did not contain endophytes (E-, tan points) and are faceted based on collection period.}
+	\end{figure}
+
+	\begin{figure}[H]
+		\centering
+		\includegraphics[width = \linewidth]{mesh_plot.png}
+		\caption{\textbf{Delauney triangulation mesh used to estimate spatial dependence between data points}.
+Grey lines indicate edges of triangles used to define distances between observations.
+Red points indicate locations of sampled herbarium specimens, and the blue outlines show the international borders used to define the edge of the mesh along coastlines.}
+	\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width = .8\linewidth]{density_plot.png}
+	\caption{\textbf{Consistency between real data and simulated values indicate that the fitted model accurately describes the data}.
+Graph shows density curves for the observed data (red) along with the mean(solid) and 95\% CI (dashed) of simulated values (black).}
+\end{figure}
+
+	\begin{figure}[H]
+		\centering
+		\includegraphics[width = .5\linewidth]{ROC_plot.png}
+		\caption{\textbf{ROC plot showing model performance classifying observations according to endophyte status.} The curves show adequate model performance for observed (top) and test (bottom) data.
+The AUC for each is 0.77 and 0.81 respectively.}
+	\end{figure}
+
+	\begin{figure}[H]
+	\centering
+	\includegraphics[width = \linewidth]{climate_change_plot.png}
+	\caption{\textbf{Change in 30-year climate normals between the periods ending in 1915 and 2020.} Color represented change in cumulative precipitation (top row) and average seasonal temperature (bottom row) for the study area.}
+\end{figure}
+
+	\begin{table}[h]
+		\caption{Summary of herbarium sampling}
+		\label{Table:herbaria}
+		\centering
+		\begin{tabular}{llll}\hline
+			Collection        & AGHY        & AGPE      &      ELVI\\ \hline
+			Botanical Research Institute of Texas &      &    &        \\
+			Louisiana State University &       &   &          \\
+			Mercer Botanic Garden &       &      &     \\
+			Missouri Botanic Garden&  & & \\
+		    Texas A\&M &  & & \\
+		    University of Kansas &  &&  \\
+		    University of Oklahoma &  &&  \\
+		    University of Texas     &  & & \\		    				 			     			     
+			Oklahoma State University&       &       &   \\ \hline
+		\end{tabular}
+		\bigskip{}
+
+	\end{table}
 
@@
 
-	\renewcommand{\theequation}{A\arabic{equation}}
-	% redefine the command that creates the equation number.
-	\renewcommand{\thetable}{A\arabic{table}}
-	\setcounter{equation}{0}  % reset counter 
-	\setcounter{figure}{0}
-	\setcounter{table}{0}
-
-	\subsection*{Fox--dog encounters through the ages}
-
-	The quick red fox jumps over the lazy brown dog.
-The quick red fox has always jumped over the lazy brown dog.
-The quick red fox began jumping over the lazy brown dog in the 19th century and has never ceased from so jumping, as we shall see in figure~\ref{Fig:Jumps}.
-But there can be surprises (figure~\ref{Fig:JumpsOk}).
-
-	If the order and location of figures is not otherwise clear, feel free to include explanatory dummy text like this:
-
-	[Figure A1 goes here.]
-
-	[Figure A2 goes here.]
-
-	\subsection*{Further insights}
-
-	Tables in the appendices can appear in the appendix text (see table~\ref{Table:Rivers} for an example), unlike appendix figure legends which should be grouped at the end of the document together with the other figure legends.
-
-	\begin{table}[h]
-		\caption{Various rivers, cities, and animals}
-		\label{Table:Rivers}
-		\centering
-		\begin{tabular}{lll}\hline
-			River        & City        & Animal            \\ \hline
-			Chicago      & Chicago     & Raccoon           \\
-			Des Plaines  & Joliet      & Coyote            \\
-			Illinois     & Peoria      & Cardinal          \\
-			Kankakee     & Bourbonnais & White-tailed deer \\
-			Mississippi  & Galena      & Bald eagle        \\ \hline
-		\end{tabular}
-		\bigskip{}
-		\\
-		{\footnotesize Note: See table~\ref{Table:Founders} below for further table formatting hints.}
-	\end{table}
-
-	Lorem ipsum dolor sit amet, as we have seen in figures~\ref{Fig:Jumps} and \ref{Fig:JumpsOk}.
-
 	%%%%%%%%%%%%%%%%%%%%%
@@
 
-	\section*{Tables}
-	\renewcommand{\thetable}{\arabic{table}}
-	\setcounter{table}{0}
-
-	\begin{table}[h]
-		\caption{Founders of \textit{The~American Naturalist}}
-		\label{Table:Founders}
-		\centering
-		\begin{tabular}{lll}\hline
-			Early editor            & Years with the journal \\ \hline
-			Alpheus S. Packard Jr.  & 1867--1886 \\
-			Frederick W. Putnam     & 1867--1874 \\ 
-			Edward S. Morse         & 1867--1871 \\ 
-			Alpheus Hyatt           & 1867--1871 \\
-			Edward Drinker Cope$^a$ & 1878--1897 \\
-			J.~S. Kingsley          & 1887--1896 \\ \hline 
-		\end{tabular}
-		\bigskip{}
-		\\
-		{\footnotesize Note: Table titles should be short. Further details should go in a `notes' area after the tabular environment, like this. $^a$ Published the first description of \textit{Dimetrodon}.}
-	\end{table}
-
-	\newpage{}
-
-	\section*{Figure legends}
-
-	\begin{figure}[h!]
-		%\includegraphics{horn-of-okapi}
-		\caption{Figure legends can be longer than the titles of tables. However, they should not be excessively long---in most cases, they should be no more than 100 words each.}
-		\label{Fig:OkapiHorn}
-	\end{figure}
-
-	\begin{figure}[h!]
-		%\includegraphics{elegance}
-		\caption{In this way, figure legends can be listed at the end of the document, with references that work, even though the graphic itself should be included for final files after acceptance.
-Instead, upload the relevant figure files separately to Editorial Manager; Editorial Manager should insert them at the end of the PDF automatically.}
-		\label{Fig:AnotherFigure}
-	\end{figure}
-
-	% Figure legends for any print appendices you might have.
-
-	\renewcommand{\thefigure}{A\arabic{figure}}
-	\setcounter{figure}{0}
-
-	\begin{figure}[h!]
-		%\includegraphics{jumps20m}
-		\caption{\textit{A}, the quick red fox proceeding to jump 20~m straight into the air over not one, but several lazy dogs.
-\textit{B}, the quick red fox landing gracefully despite the skepticism of naysayers.}
-		\label{Fig:Jumps}
-	\end{figure}
-
-	\begin{figure}[h!]
-		%\includegraphics{jumps20m}
-		\caption{The quicker the red fox jumps, the likelier it is to land near an okapi. For further details, see \citet{LemKapEx07}.}
-		\label{Fig:JumpsOk}
-	\end{figure}
-
-	%%%%%%%%%%%%%%%%%%%%%
-	% Videos
-	%%%%%%%%%%%%%%%%%%%%%
-	% If you have videos, journal style for them is generally similar to that for
-	% figures.
-
-	%%%%% Include the text below if you have videos
-
-	\renewcommand{\figurename}{Video} 
-	\renewcommand{\thefigure}{S\arabic{figure}}
-	\setcounter{figure}{0}
-
-	\begin{figure}[h!]
-		%\includegraphics{VideoScreengrab.jpg}
-		\caption{Video legends can follow the same principles as figure legends. Counters should be set and reset so that videos and figures are enumerated separately.}
-		\label{VideoExample}
-	\end{figure}
-
-	\renewcommand{\figurename}{Figure}
-	\setcounter{figure}{1}
-
-	%%%%% Include the above if you have videos
-
 \end{document}

### 2023-03-17 Manuscript/endo_herbarium.tex
tom's feedback on herbarium draft

@@
 \titleformat{\paragraph}[runin]{\itshape}{\theparagraph}{1em}{}[. ]\renewcommand{\refname}{Literature Cited}
+
+\usepackage{xcolor}
+\newcommand{\tom}[2]{{\color{red}{#1}}\footnote{\textit{\color{red}{#2}}}}
 
@@
 Climate change has altered species’ phenologies, has triggered changes in traits, and has led to shifts in abundance and distributions. 
-Yet, we still know little about how biotic interactions alter these environmental responses over long time scales across spatially heterogenous changes in climate. 
-Most species host beneficial microbial symbionts, which may provide resilience to changes in the environment, but at the same time increasing stress may lead these symbioses to breakdown. 
+Yet, we still know little about how biotic interactions alter these responses over long time scales across spatially heterogenous changes in climate. 
+Most species host beneficial microbial symbionts, \tom{which may provide resilience to changes in the environment, but at the same time increasing stress may lead these symbioses to breakdown}{A little confusing as written.
+Might be better to present as alernative hypotheses}.
 Microbes can be preserved within the dried plant tissue of herbarium specimens presenting an opportunity to quantify changes across broad temporal and spatial scales. 
-Epichloë fungal endophytes are seed-transmitted symbionts of cool-season grasses which can provide drought tolerance to their hosts. 
- We ask how endophyte prevalence has responded to climate change. 
- Specifically, we took samples of seeds from herbarium specimens of three grass host species collected over the last two centuries, quantified the presence or absence of fungal endophytes within the seeds with microscopy, and evaluated spatial and temporal trends in prevalence using spatial modeling.
+\tom{Epichloë fungal endophytes are seed-transmitted symbionts of cool-season grasses which can provide drought tolerance to their hosts.
+}{I would cut this sentence and incorporate need to know info in other sentences.}
+\tom{ We ask how endophyte prevalence has responded to climate change.}{Could be sharpened.} 
+ Specifically, we took samples of seeds from herbarium specimens of three grass host species collected over the last two centuries, quantified the presence or absence of fungal endophytes within the seeds with microscopy, and evaluated spatial and temporal trends in prevalence using \tom{spatial modeling}{Vague}.
  We found strong evidence that fungal endophyte prevalence has increased over the last two centuries across the three host species from 25\% prevalence to 75\% prevalence on average.
-We also found that observed changed in annual  and seasonal climate drivers were associated with temporal trends in endophyte prevalence. 
-In particular, changes in prevalence within each species corresponded to climate drivers during its peak growing season. 
-Thus we provide novel evidence for adaptive biotic responses to ongoing climate change.
+\tom{We also found that observed changed in annual  and seasonal climate drivers were associated with temporal trends in endophyte prevalence. 
+In particular, changes in prevalence within each species corresponded to climate drivers during its peak growing season. }{Could be combined into one sentence.}
+Thus we provide novel evidence for \tom{adaptive biotic responses}{Not sure what you mean by this.} to ongoing climate change.
 
@@
 Predicting how species' interactions will be altered by ongoing climate change is a major goal of basic ecological research and conservation management. 
-Spatially heterogeneous changes in the environment have led species to move, adapt, or face extinction \citep{IPCC2012managing}. 
+\tom{Spatially heterogeneous changes in the environment have led species to move, adapt, or face extinction \citep{IPCC2012managing}.
+}{In theory a spatially homogenous change in temperature -- everywhere increases by 2 degrees -- would have the same effect.
+So significance of spatial heterogeneity here is not clear.}
 Species' responses are likely mediated by the numerous interspecific interactions in which they engage. 
-Yet for most species, empirically evaluating the role that biotic interactions can play in facilitating or hindering adaption to changing environmental conditions has been limited by a lack of data at adequate spatial and temporal scales \cite{gilman2010framework}.
+Yet for most species, empirically evaluating the role that biotic interactions can play in facilitating or hindering \tom{adaption}{I am struck by this word again.
+Are you intentionally bringing in evolutionary ideas?} to changing environmental conditions has been limited by a lack of data at adequate spatial and temporal scales \cite{gilman2010framework}.\tom{}{This paragraph has introduced two related but different ideas: how species interactions per se respond to climate change, and how interactinos mediate responses to climate change.
+This is a little disoroienting.}
 
-Natural history collections, which were originally collected to study and preserve taxonomic diversity, present an exciting opportunity to explore long-term changes in ecological interactions across broad spatial scales \citep{meineke2018unrealized}.
-Natural history collections, built and maintained by the efforts of thousands of scientists, are invaluable time machines. 
+Natural history collections, which were originally collected to study and preserve taxonomic diversity, present a unique opportunity to explore long-term changes in ecological interactions across broad spatial scales \citep{meineke2018unrealized}.
+Natural history collections, built and maintained by the efforts of thousands of scientists, are \tom{invaluable time machines}{Love this!}. 
 These collections primarily comprise physical specimens of organisms  along with information about the time and place of their collection. 
 These specimens are samples preserving physical legacies of ecological dynamics and species' interactions from dynamically changing environments across time and space.
-Recent works using plant collections have documented changes in phenology \citep{willis2017old, park2019herbarium,  berg2019examination}, as well as pollination \citep{pauw2011reconstruction, duan2019century}, and changes in antagonistic interactions such as herbivory \citep{meineke2019herbarium}, but to date, focus has been lacking on long-term changes in a particularly common type of biotic interaction, microbial symbioses.
+\tom{Recent works using plant collections have documented changes in phenology \citep{willis2017old, park2019herbarium,  berg2019examination}, as well as pollination \citep{pauw2011reconstruction, duan2019century}, and changes in antagonistic interactions such as herbivory \citep{meineke2019herbarium}, but to date, focus has been lacking on long-term changes in a particularly common type of biotic interaction, microbial symbioses.}{I like the content of this sentnece but  found the phrasing difficult.}
 
-Microbes are hosted by all organisms with documented effects on their hosts' survival, growth and reproduction \cite{rodriguez2009fungal,mcfall2013animals}.
-Many microbial symbionts act as mutualists, engaging in reciprocally beneficial interactions with their hosts. 
-These mutualisms play an important role in limiting the consequences of environmental stress. 
-For example, bacterial symbionts of insects, such as \emph{Wolbachia}, can improve their hosts' thermal tolerance \citep{truitt2019wolbachia, renoz2019evolutionary}.
-Similarly, arbuscular mycorrhizal fungi, documented in 70-90\% of families of land plants \citep{parniske2008arbuscular}, allow their hosts to persist through drought conditions by improving water and nutrient uptake \citep{cheng2021elucidating}.
-At the same time, changes in the mean and variance of environmental conditions may disrupt mutualisms by changing the costs and benefits of the interaction for each partner, leading to the extinction of one or both partner species \citep{aslan2013mutualism}.
-Coral bleaching, the loss of symbiotic algae due to temperature stress \citep{sully2019global} is perhaps the best known example, but this phenomenon is not unique to corals.
+Microbial symbionts are common to all macroscopic organisms and can have important effects on their hosts' survival, growth and reproduction \cite{rodriguez2009fungal,mcfall2013animals}.
+Many microbial symbionts act as mutualists, engaging in reciprocally beneficial interactions with their hosts and often ameliorating host responses to environmental stress. 
+For example, bacterial symbionts of insects, such as \emph{Wolbachia}, can improve their hosts' thermal tolerance \citep{truitt2019wolbachia, renoz2019evolutionary}, and arbuscular mycorrhizal fungi, documented in 70-90\% of families of land plants \citep{parniske2008arbuscular}, allow their hosts to persist through drought conditions by improving water and nutrient uptake \citep{cheng2021elucidating}.
+On the other hand, changes in the mean and variance of environmental conditions may disrupt microbial  mutualisms by changing the costs and benefits of the interaction for each partner, leading to breakdown of the interaction \citep{aslan2013mutualism}.
+Coral bleaching (the loss of symbiotic algae) due to temperature stress \citep{sully2019global} is perhaps the best known example, but this phenomenon is not unique to corals.
 Lichens exposed to elevated temperatures experienced loss of photosynthetic function along with changes in the community composition of their algal symbionts \citep{meyer2022climate}.
-Whether symbiotic mutualisms breakdown or promote resilience under climate change is unclear, but likely depends on the intimacy and specialization of the interaction, as well as the physiological tolerances of the mutualists \citep{toby2010mutualisms, warren2014mutualism, rafferty2015phenological}.
-Given that climate change is spatially heterogenous, the rate of change may have a strong influence on when and where we can expect facilitation or disruption of mutualism.
+How commonly and under what conditions microbial mutualisms breakdown or strengthem under climate change remain unanswered questions, but previous work suggests that these alternative responses may depend on the intimacy and specialization of the interaction as well as the physiological tolerances of the mutualists \citep{toby2010mutualisms, warren2014mutualism, rafferty2015phenological}.
+\tom{Given that climate change is spatially heterogenous, the rate of change may have a strong influence on when and where we can expect facilitation or disruption of mutualism.
+}{This is an important sentence that needs some elboration and citation.
+Can you explain more about spatial heterogeneity in climate change?
+And are you suggesting that the rate of change itself determines whether mutualism strengthens or breaks down?
+Also, ``facilitation'' is awkward in this context.}
 
-Early research on \emph{Epichloë} fungal endophytes, specialized symbionts of cool-season grasses, used herbarium specimens to describe the broad taxonomic diversity of hosts \citep{white1985endophyte}.
+Early research on \emph{Epichloë} fungal endophytes, specialized symbionts of cool-season grasses, \tom{used herbarium specimens}{I would move this later in the paragraph, where you can make the point that this approach has not been leveraged to its full capacity.} to describe the broad taxonomic diversity of hosts \citep{white1985endophyte}.
 These facultative symbionts, which have been documented in $\sim 30$\% of cool-season grasses \citep{leuchtmann1992systematics}, are transmitted vertically from maternal plants to offspring through the grasses' seeds.
@@
 Over time, endophytes that act as mutualists for their hosts should rise in prevalence within a population. 
-\emph{Epichloë} are known to improve drought tolerance for their hosts \citep{afkhami2014mutualist}, and are expected to reach high prevalence where these fitness benefits are most important.
+\emph{Epichloë} are known to improve \tom{drought}{not just drought - also herbivory and pathogens.
+I would cite more to support the types of endo benefits.} tolerance for their hosts \citep{afkhami2014mutualist}, and are expected to reach high prevalence where these fitness benefits are most important.
 Previous survey studies have documented large-scale spatial patterns in  endophyte prevalence structured by environmental gradients \citep{granath2007variation,bazely2007broad, afkhami2012fungal,sneck2017variation}.
-We predicted that prevalence should also track temporal changes in environmental drivers, but testing this has been impossible with surveys limited to a few years, and to the spatial extent reachable during a single growing season.
-Becauses the fungi are seed-transmitted, they are conveniently preserved along with their host plants on herbarium sheets, creating the opportunity to examine the response of the symbiosis to historical climate change.
+We predicted that \tom{prevalence should also track temporal changes in environmental drivers}{explain and elaborate}, but testing this has been impossible with surveys limited to a few years, and to the \tom{spatial extent reachable during a single growing season}{we have cars and planes so this should not be an obstacle.
+I think the real limitation is the combination spatio-temporal coverage, which is what herbaria uniquely provide}.
+Becauses the fungi are seed-transmitted, they are conveniently preserved \tom{along with their host plants on herbarium sheets}{as long as the specimen has seeds, which is almost always the case for grasses}, creating the opportunity to examine the response of the symbiosis to historical climate change.
 
-In this study, we took advantage of the spatial and temporal sampling within herbaria to assess for the first time whether grass-fungal endophyte symbioses have responded adaptively to climate change.
+In this study, we took advantage of the spatial and temporal sampling within herbaria to assess for the first time whether grass-fungal endophyte symbioses have responded \tom{adaptively}{the adaptive nature of the response needs to be explained} to climate change.
 In particular, we (i) quantified the spatial and temporal change in endophyte prevalence across host ranges over the last two centuries, and (ii) identified climate drivers that can explain the magnitude of these responses.
+\tom{}{This paragraph would benefit from a specific hypothesis or question, which I think you have.
+Your emphasis on space vs time is a little muddled.
+You frame the paper around climate change, but part of your analysis - I think - is asking a purely spatial question.
+This paragraph should also mention the three host species and provide some wow-factor summary stats on the range of years, locations, and sample sizes.}
 
@@
         \subsection*{Focal host species}
-We focused on three endophyte hosts, \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}.
-These grass species, which are all cool-season (C3) grasses, are commonly represented in natural history collections with broad distributions covering the Eastern United States.
-\emph{A. hyemalis} is a small short-lived perennial species that germinated in the spring and typically blooms between March and July.
-\emph{A. perennans} of similar stature but blooms in late Summer and early Autumn. 
+We focused on three endophyte hosts, \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}.\tom{}{Describe the fungal species.}
+These cool-season (C3) grass species are commonly represented in natural history collections with broad distributions covering the Eastern United States.
+\emph{A. hyemalis} is a small short-lived perennial species that germinates in the spring and typically blooms between March and July.
+\emph{A. perennans} is of similar stature but is longer lived than \emph{Agrostis hyemalis} and blooms in late Summer and early Autumn. 
 This species is relatively more  sparsely distributed, tending to be found in shadier and more moist habitats while \emph{A. hyemalis} is often found in open and recently disturbed ground. 
@@
 		\subsection*{Herbarium surveys}
-We collected 5-10 seeds per specimen during visits to natural history collections between 2019 and 2022 (See Table A1 for a summary of collections) . 
-Grasses are commonly identified and collected based on the presence of their reproductive structures which ensured that specimens we examined contained seeds. 
-Our sampling plan was designed to minimize damage to these specimens.
+We collected 5-10 seeds per specimen during visits to natural history collections between 2019 and 2022 \tom{(See Table A1 for a summary of collections)}{It is important to say more in the text here about where you went and what you did.} .
+\tom{Grasses are commonly identified and collected based on the presence of their reproductive structures which ensured that specimens we examined contained seeds.}{Move to intro} 
+\tom{Our sampling plan was designed to minimize damage to these specimens.}{How so?}
 
@@
 We examined the squashed seeds for the presence of fungal hyphae at 100X magnification \cite{bacon2018stains}.
-We designated each specimen as symbiotic or non-symbiotic, a binary response variable, based on the presence of endophytes within any of the scored seeds for that specimen. 
 In some cases, the tissues examined during microscopy came from flowers or otherwise non-viable seeds, which were excluded from the seed counts for that specimen.
+On average we scored \textbf{number} seeds per specimen.
+Due to imperfect vertical transmission (cite), it is possible that symbiotic host-plants produce a mixture of symbiotic and non-symbiotic seeds. 
+We therefore designated each specimen as symbiotic if endophyte hyphae were observed in one or more seeds, or non-symbiotic if hyphae were observed in zero seeds. 
 To capture uncertainty in the identification process, we recorded both a "liberal" and a "conservative" endophyte status for each plant.  
 When we identified potential endophytes with unusual morphology, low uptake of stain, or a small amount of fungal hyphae across the scored seeds, we recorded a positive liberal status (more likely to be a true endophyte) and a negative conservative status (less likely to be a true endophyte).
-$89$\% of scored plants had matching liberal and conservative endophyte statuses, reflecting high confidence in the endophyte identification.
-The following analyses presented in the main text used the liberal statuses, but we repeated all analyses with the conservative statuses which yielded qualitatively similar results (Appendix).
+$89$\% of scored plants had matching liberal and conservative endophyte statuses, reflecting high confidence in endophyte identification.
+\tom{The following analyses presented in the main text used the liberal statuses, but we repeated all analyses with the conservative statuses which yielded qualitatively similar results (Appendix).
+}{Good}
 
-We excluded specimens that did not include information about the collection location and date from our sampling efforts.
-In total, we quantified endophyte symbiosis for $971$ \emph{A. hyemalis}, $302$ \emph{A. perennans}, and $605$ \emph{E. virginicus} specimens spanning the years $1824$ to $2019$.
-Each specimen was assigned a collection location based on geographic information recorded at the time of collection using the ggmap package in R \citep{kahle2019package}. 
-Collections were georeferenced to the nearest county centroid, or nearest municipality when that information was available.
-For a few of the oldest specimens, only information at the state level was available, and so we used the state centoid (Fig.
-1; Fig A1).
+We excluded specimens for which information about the collection location and date were unavailable.
+In total, we quantified endophyte symbiosis for $971$ \emph{A.
+hyemalis}, $302$ \emph{A.
+perennans}, and $605$ \emph{E.
+virginicus} specimens spanning the years \tom{$1824$ to $2019$}{give date range by species}.
+\tom{Each specimen was assigned a collection location based on geographic information recorded at the time of collection using the ggmap package in R \citep{kahle2019package}.}{confusing-not sure what this means}
+Collections were geo-referenced to the nearest county centroid, or nearest municipality when that information was available. 
+For a few of the oldest specimens, only information at the state level was available, and so we used the state centoid \tom{(Fig.
+1; Fig A1).}{I don't see how this is shown in figure 1.
+Also, can you use more distinctive colors in Fig 1?}
 
 \subsection*{Assessing spatial and temporal changes in endophyte prevalence}
-To quantify spatial and temporal trends in endophyte prevalence, we modelled endophyte presence/absence ($P$) as a Bernoulli distribution for specimen $i$ of host species $h$.
+\tom{}{Give a quick overview of the types of models you used and what specific purpose they were built for before getting into details.
+The fact that you estimated climate effects following an initial climate-implicit model needs to come through more clearly.}To quantify spatial and temporal trends in endophyte prevalence, we modeled endophyte presence/absence ($P$) as a Bernoulli response variable for specimen $i$ of host species $h$.
 
@@
 		\begin{align}
-		P_{i,h} \sim Bernoulli(\hat{P_{h}}) \\
+		P_{i,h} \sim Bernoulli(\hat{P_{i,h}}) \\
 			logit(\hat{P}_{h}) = \beta_{0,h} + 
@@
 
-The expected endophyte prevalence, $\hat{P}$, was modelled with intercepts specific to each host species ($\beta_{0}$), slopes for changes over time ($\beta_{1}$) as well as the interaction between time and the specimen's latitude and longitude ($\beta_{2}$, $\beta_{3}$, and $\beta_{4}$).
+The expected endophyte prevalence, $\hat{P}$, was modelled with intercepts specific to each host species ($\beta_{0}$), slopes for changes over time ($\beta_{1}$) \tom{as well as the interaction between time and the specimen's latitude and longitude ($\beta_{2}$, $\beta_{3}$, and $\beta_{4}$).
+}{So you do not have equal emphasis on space and time.
+YOu are really asking a time question and considering how temporal trends may vary spatially.
+I think this needs to come through more clearly in the introduction.}
 We accounted for potential biases introduced during the process of collecting specimens as well as in scoring ability by including random effects specific to each collector $\chi$ and to each scorer $\omega$.
 To quantify spatial patterns in prevalence, we included a spatially-dependent random intercept, $\phi$, and fit the model using integrated nested Laplace approximation, an approximate Bayesian method implemented in the rINLA package \citep{lindgren2015bayesian}.
-This method allowed us to model spatial dependency between data points as a continuous process with stochastic partial differential equations (SPDE) that depend on a covariance matrix according to the proximity of each collection location \citep{lindgren2011explicit,bakka2018spatial}.
+This method allowed us to model spatial dependency between data points (nearer collections may be more similar) as a continuous process with stochastic partial differential equations (SPDE) that depend on a covariance matrix according to the proximity of each collection location \citep{lindgren2011explicit,bakka2018spatial}.
 The covariance matrix is approximated using a Matérn covariance function, with each data point assigned a location according to the nodes of a mesh of non-overlapping triangles across our study area (Fig A2).
 We fit the model with vague priors, and compared models with different sizes of mesh, which had little effect on the resulting model estimates.
-We assessed model fit with graphical postterior predictive checks (Fig. A3, Fig. A4).
+\tom{We assessed model fit with graphical postterior predictive checks (Fig. A3, Fig. A4).}{This is good but also worth mentioning that you assessed convergence.}
 
-We used data on endophyte prevalence from  contemporary surveys of \emph{A. hyemalis} as test data to evaluate model fit. 
+\tom{We used data on endophyte prevalence from  contemporary surveys of \emph{A.
+hyemalis} as test data to evaluate model fit.
+}{This is an extraordinary extra step that should be played up in the Intro and included in the aims at the end of the Intro.}
 During these surveys, which took place between 2015 and 2020, we collected seeds from living plants from 43 populations across a portion of the species' range and then quantified endophyte status with staining microscopy as described for the herbarium surveys (Fig S2).
-We compared the model's prediction for these locations to the endophyte status of randomly sampled plants from each surveyed location, mimicing the process of collecting a single herbarium specimen from the population.
-The model performed adequately (training data: AUC = 0.77; test data: AUC = 081; Fig. A4).
+We compared the model's prediction for these locations to the endophyte status of randomly sampled plants from each surveyed location, \tom{mimicing the process of collecting a single herbarium specimen from the population.}{But isn't the prevalence the better test of the model's predictions?}
+The model performed adequately (\tom{training data: AUC = 0.77; test data: AUC = 081}{Is this 0.81? The model performs better on the test data than the training data? This is odd.}; Fig. A4).
 
 		\subsection*{Assessing the role of climate drivers}
-To assess how the rate of climate change may have driven changes in endophyte prevalence, we calculated the change in climate between recent (1990 to 2020) and historic (1895 to 1925) periods.
+To assess how the rate of climate change may have driven changes in endophyte prevalence,\tom{ we calculated the change in climate between recent (1990 to 2020) and historic (1895 to 1925) periods.}{This is where your hypotheses and predictions in the intro need to come through more clearly.
+Also, this assumes change has been constant during those periods and does not use most years of data.
+Could you try estimating a slope for each location?
+Is this possible?}
 We first downloaded monthly temperature and precipitation rasters from the PRISM climate group \citep{daly2013prism} covering the time period between 1895 and 2020 using the 'prism' R package \citep{Rprism2015}.
@@
 We calculated 30-year climate normals for annual and seasonal mean temperature and cumulative precipitation for the recent and historic periods.
-We used three four-month seasons within the year (Spring: January, February, March, April; Summer: May; June, July, August; Autumn: September, October, November, December). 
-We additionally calculated the coefficient of variation of each climate driver during each period.
-We then took the difference between recent and historic periods for the mean and coefficient of each climate driver (Fig. A5).
+\tom{We used three four-month seasons within the year (Spring: January, February, March, April; Summer: May; June, July, August; Autumn: September, October, November, December).
+}{I think this needs more biological justification.
+I can imagine a simpler divion of cool season and warm season.}
+\tom{We additionally calculated the coefficient of variation of each climate driver during each period.}{why? and how? was this based on monthly means or sums?}
+We then took the difference between recent and historic periods for the mean and coefficient of variation for each climate driver (Fig. A5).
 Because initial analyses indicated a high degree of collinearity between seasonal and annual changes in temperature, we used annual temperature only, along with annual and seasonal precipitation, in the subsequent analysis.
 
-We then evaluated how predicted changes in endophyte prevalence correlated with changes in thee environment at an evenly spaced grid of points across the study area. 
-Because regions varied in their predicted starting endophyte prevalence, we calculated the relative change in endophyte prevalence between 1925 and 2020. 
+We then evaluated how predicted changes in endophyte prevalence correlated with changes in the environment at an evenly spaced grid of points across the study area.\tom{}{Map?} 
+Because regions varied in their predicted starting endophyte prevalence, we calculated the \tom{relative change}{calculated how?} in endophyte prevalence between 1925 and 2020. 
 This time period connects the predicted change in prevalence with the endpoints of the available climate record.
-We then calculated the Pearson correlation coefficient between the relative change in endophyte prevalence and each climate driver.
+We then calculated the Pearson correlation coefficient between the relative change in endophyte prevalence and each climate driver.\tom{}{This paragraph should be clearer that you are using predicted prevalence from the model.}
 
@@
 perennans} increased from ~ 15\% to over 70\% prevalence.
-Rerunning the analysis excluding specimens collected before 1900 when samples are much sparser (Fig. 2A), lead to qualitatively similar predictions, and so we continued analyses with the full dataset.
+Rerunning the analysis excluding specimens collected before 1900 when samples are much sparser (Fig.
+2A), lead to qualitatively similar predictions, and so we continued analyses with the full dataset.
+\tom{}{I really like fig 2 but if you are going to split the predictions by latitude you should explain why, talk about this in the results, and also split the points by high and low latitude.
+Were 35 and 43 the bounds of the data?}
 
@@
 
-Across space, there were clear trends in endophyte prevalence which varied between species (Fig. 3). 
-For example, \emph{Elymus virginicus}, had higher prevalence in the northern and eastern portions of it's range. 
+\tom{Across space, there were clear trends in endophyte prevalence which varied between species (Fig. 3).}{This seems to contradict my earlier comment about how your focus is mainly time.} 
+For example, \emph{Elymus virginicus}, \tom{had higher prevalence in the northern and eastern portions of it's range}{The model predicts low prevalence in Florida but there are basically no records from Florida, so the prediction does not seem very meaningful.
+See also next comment.}.
 In contrast, \emph{Agrostis hyemalis} had highest prevalence towards the center of its range with regions of low prevalence towards the north east and towards its western range edge.
-Our model revealed that while there was an overall increase in endophyte prevalence, these changes varied across the host species' ranges.
+\tom{Our model revealed that while there was an overall increase in endophyte prevalence, these changes varied across the host species' ranges.}{What is problematic here is that the model is not showing predictions relative to each species' range, it is showing the aggregate sampling region.
+I think this will be more effective and meaningful if it is based on species-specific ranges.}
 In some regions, \emph{A.
@@
 In other regions, there were smaller increases of only five percentage points for \emph{E. virginicus} and fifteen percentage points for \emph{A. perennans}. 
-Notably, the western range edge of \emph{A. hyemalis} did not experience an increase in endophyte prevalence. 
+\tom{Notably, the western range edge of \emph{A. hyemalis} did not experience an increase in endophyte prevalence. }{Why is this notable? Also, see prev comment about real spp ranges.}
 Collections are sparser to the west as the species reaches its range limit, but the collected specimens were exclusively symbiont-free. (Fig. A1). 
+\tom{}{For the prevalence map, could you also map the difference between past and present prevalence to show hotspots and coldspots?}
 
@@
 
+\tom{}{The results section describes spatial and temporal trends but not the combination, which I thought was the main point of the paper.}
+\tom{}{There is nothing in this section that addresses statistical significance, however one wants to define that.}
+
 \subsection*{The role of climate drivers}
-We found that the trends in endophyte prevalence were strongly associated with seasonal climate change drivers (Fig. 4). 
+We found that the trends in endophyte prevalence were strongly associated with seasonal climate change drivers \tom{(Fig.
+4).}{How will you respond to the reviewer who asks you which of these correlations are significant?}
 For the majority of the study region, the climate has become wetter and cooler over the last century (Fig. A5), a consequence of regional variation in global climate change. 
-Within the study region, spatially heterogeneous environmental changese were predictive of changes in endophyte prevalence in line with differences in species' reproductive timing. 
-For example, relative change in endophyte prevalence within \emph{E.
+Within the study region, spatially heterogeneous environmental changes were predictive of changes in endophyte prevalence in line with differences in species' reproductive phenology. 
+For example, \tom{relative change in endophyte prevalence}{can you describe this in terms of increasing or decreasing prevalence, to make the interpretation more intuitive?} within \emph{E.
 virginicus} were most associated with declines in Summer precipitation  (a negative correlation in Fig.
@@
 4).
-These associations are in line with expectations that endophyte symbiosis provides drought tolerance to their hosts, particularly during \emph{E. virginicus}'s summer growing season. 
+These associations are in line with expectations that endophyte symbiosis provides drought tolerance to their hosts, particularly during \emph{E.
+virginicus}'s \tom{summer growing season}{It is a stretch to say that this species has a summer growing season.}.
 For \emph{A.
-perennans}, which typically flowers in the late Summer and Autumn, increases in autumn precipitation along with increases in variability in annual precipitation, were strongly assocated with increasing endophyte prevalence.
-While this result is contrary to the expectation that endophyte provide drought tolerance, this species typically is associated with wetter microhabitats, and it also suggests a role for endophytes to play in buffering the negative consequences of demographic variability \citep{lewontin_population_1969}.
+perennans}, which flowers in the late Summer and Autumn, increases in autumn precipitation along with increases in variability in annual precipitation, were strongly assocated with increasing endophyte prevalence.
+While this result is contrary to the expectation that endophyte provide drought tolerance, this species typically is associated with wetter microhabitats, and it also \tom{suggests a role for endophytes to play in buffering the negative consequences of demographic variability \citep{lewontin_population_1969}}{I would save for discussion}.
 For \emph{A. hyemalis}, which flowers in the Spring, changes in endophyte prevalence were more weakly associated with seasonal changes in climate drivers than for the other two species.
-It was, however, the only species' for which greater increases in endophyte prevalence were associated with regions that experienced increases in average annual temperatures.
+It was, however, the only species' for which greater increases in endophyte prevalence were associated with regions that experienced increases in average annual temperatures.\tom{}{Is this a significant association, however you want to define that?
+Or could this result have occurred by chance?}
 
@@
 \section*{Discussion}
-Our examination of historic plant specimens revealed a cryptic biotic reponse to climate change. 
-The geographic distributions and phenology of macroorganisms are apparent indicators of climate change.
+\tom{Our examination of historic plant specimens revealed a cryptic biotic reponse to climate change. }{Great opening sentence.}
+\tom{The geographic distributions and phenology of macroorganisms are apparent indicators of climate change.}{Not sure why you go here next, then go back to endophytes.}
 For the three host species we examined, there have been clear increases in fungal endophyte prevalence over the last two centuries.
-This increase suggests that \emph{Epichloë} fungal endophytes are playing key adaptive roles in response to climate change and that the mutualism between grasses and their fungal endophytes has contributed to resilience for each partner species under the observed changes in climate.
-Spatial and temporal trends in endophyte prevalence varied between each species in ecologically meaningful ways.
-Our spatially-explicit model predicted regions of both high and low endophyte prevalence, suggesting that symbiotic and non-symbiotic host plants have overlapping, but non-identical niche requirements.
-Endophytes may expand the range of their hosts by allowing them to persist in environments where they otherwise could not \citep{afkhami2014mutualist, kazenel2015mutualistic}.
+This increase suggests that \emph{Epichloë} fungal endophytes are playing key \tom{adaptive}{In what way are they adaptive?} roles in response to \tom{climate change}{Climate change is not the only change over this time period, so be careful with attribution.} and that the mutualism between grasses and their fungal endophytes has contributed to resilience for each partner species under the observed changes in climate.
+Spatial and temporal trends in endophyte prevalence varied between each species in \tom{ecologically meaningful ways}{vague}.
+\tom{Our spatially-explicit model predicted regions of both high and low endophyte prevalence, suggesting that symbiotic and non-symbiotic host plants have overlapping, but non-identical niche requirements.}{This is a space-only result and interpretation, which should be better explained and motivated.}
+\tom{Endophytes may expand the range of their hosts by allowing them to persist in environments where they otherwise could not \citep{afkhami2014mutualist, kazenel2015mutualistic}.}{Not sure this is right for the first paragraph.}
 Understanding how microbial symbionts respond to climate change is a first step towards incorporating reciprocal host-symbiont interactions into predictions host range shifts.
+\tom{}{This paragraph should address the core question of whether climate change strengthens or disrupts mutualism.}
 
-For \emph{E. virginicus}, we found greater increases in prevalence in regions that have become drier during the Summer, the species' peak growing season. 
-Fitness benefits of the symbiosis under drought conditions could explain this increase and also explain high predicted prevalence towards the species' northern range edge coinciding with a strong latitudinal decline in precipitation.
+\tom{For \emph{E.
+virginicus}, we found greater increases in prevalence in regions that have become drier during the Summer, the species' peak growing season.}{It is worth paying attention to both absolute and relative change.
+The AGPE results for example are notable because this is notable because this is the species with the greatest absolute change.}
+Fitness benefits of the symbiosis under drought conditions could explain this increase and also explain high predicted prevalence towards the species' northern range edge coinciding with a \tom{strong latitudinal decline in precipitation}{Show this?
+There is also a longitudinal cline in precip - how do you interpret that difference?}.
 Previous population surveys for endophytes in this species found similar latitudinal trends in prevalence \citep{sneck2017variation,rudgers2009benefits}, but at smaller scales. 
@@
 hyemalis} \citep{davitt2011understanding}, we had predicted  that endophyte prevalence should similarly increase at a greater rate in regions that have seen reduced precipitation, yet we found the opposite trend in our herbarium surveys and that changes in prevalence for this species had the weakest correlations with climate drivers.
-Given that the study region has experienced relatively moderate changes in precipation and in temperature, potentially the magnitude of climate change has not been great enough to cause larger changes in endophyte prevalence for this species.
+Given that the study region has experienced \tom{relatively moderate}{Not sure what to make of ``relatively moderate''.} changes in precipation and in temperature, potentially the magnitude of climate change has not been great enough to cause larger changes in endophyte prevalence for this species.
 Weak associations with drought could also be explained by climate-driven changes in the rate of imperfect transmission (the generation of non-symbiotic offspring from symbiotic hosts), which could counterbalance endophyte-mediated fitness benefits, and leading to stable intermediate prevalence rates \citep{donald2021context}.
@@
 The strong increase in symbiotic \emph{A. perennans} could be explained by these diverse benefits. 
-Differences between the responses of each species underscore that while all of these C3 grasses share similar broad-scale distributions, each engages in unique biotic interactions and has unique niche requirements.
+\tom{Differences between the responses of each species underscore that while all of these C3 grasses share similar broad-scale distributions, each engages in unique biotic interactions and has unique niche requirements.}{I think this is an interesting point worth elaboratong on.
+On the one hand the three species are consistent in showing increases in prevalence, yet the mechanisms that explain those increases may be diverse and idiosyncratic.}
 
-Our analysis advances the use of herbarium specimens in two ways. 
+Our analysis advances the use of herbarium specimens in global change biology in two ways. 
 First and foremost, this is the first study to link long-term changes in microbial symbioses to changes in climate using specimens from natural history collections.
@@
 Answering these questions as well as the unknown questions that future researchers may ask also reiterates the value in capturing meta-information during ongoing digitization efforts at herbaria around the world and during the accession of newly collected specimens \citep{lendemer2020extended}.
-Second, we accounted for several potential biases in the data observation process that may be common to many collections-based research questions by using a spatially-explicit random effects model. 
+\tom{Second, we accounted for several potential biases in the data observation process that may be common to many collections-based research questions by using a spatially-explicit random effects model.
+}{I think you should say more about collector and observer effects in the results.}
 Spatial autocorrelation \cite{willems2022forest}, potential biases introduced by the sampling habits of collectors \citep{daru2018widespread}, and variation between contemporary researchers during the collection of trait data, if not corrected for could lead to biased inference about the strength and direction of historic change.
-Fitting this model in a Bayesian framework allows for full propagation of uncertainty.
+Fitting this model in a Bayesian framework allows for full propagation of uncertainty.\tom{}{Excellent paragraph.}
 
@@
 While this survey of historic endophyte prevalence is necessarily correlative, it serves as a foundation to develop better predictive models of the response of microbial symbioses to climate change. 
-Combining the insights from this sort of regional-scale survey with field experiments and physiological data could be invaluable. 
+Combining the insights from this type of regional-scale survey with field experiments and physiological data could be invaluable. 
 While we found that climate is strongly correlated with endophytes' temporal responses, we do not know why trends in prevalence were weak in some areas or how endophytes would respond to more extreme changes in climate.

### 2023-03-22 endo_herbarium_spatial_analysis.R
adding lots of plots and edits to main plots.

@@ -15,7 +15,7 @@ library(ROCR)
 
 
 invlogit<-function(x){exp(x)/(1+exp(x))}
-species_colors <- c("#66c2a5", "#fc8d62", "#8da0cb")
+species_colors <- c("#1b9e77","#d95f02","#7570b3")
 endophyte_colors <- c("#fdedd3","#f3c8a8", "#5a727b", "#4986c7", "#181914",  "#163381")
 
 
@@ -43,7 +43,7 @@ endo_herb_AGPE <- endo_herb_georef %>%
   filter(!is.na(Endo_status_liberal)) %>% 
   filter(Spp_code == "AGPE") %>% 
   filter(!is.na(lon) & !is.na(year)) %>% 
-  filter(lon>-110) %>% 
+  filter(lon>-100) %>% 
   filter(Country != "Canada" & !is.na(County))  
   
 
@@ -60,7 +60,8 @@ endo_herb <- endo_herb_georef %>%
   filter(!is.na(Spp_code)) %>% 
   filter(!is.na(lon) & !is.na(year)) %>% 
   filter(lon>-110 ) %>% 
-  filter(Country != "Canada" & !is.na(County)) %>% 
+  filter(Country != "Canada" ) %>% 
+  filter(!is.na(County) | is.na(Municipality)) %>% 
   mutate(year_bin = case_when(year<1970 ~ "pre-1970",
                               year>=1970 ~ "post-1970")) %>% 
   mutate(scorer_id = case_when(scorer_id == "BellaGuttierez" ~ "BellaGutierrez",
@@ -102,7 +103,8 @@ endo_herb$collector_id <- collector_no[match(as.factor(endo_herb$collector_lastn
 contemp_surveys <- read_csv(file = "contemp_surveys.csv")
 contemp_random_sample <- read_csv(file = "contemp_random_sample.csv")
 
-
+contemp_random_sample <- contemp_random_sample %>% 
+  left_join(contemp_surveys)
 
 
 # register_google(key = "")
@@ -160,6 +162,21 @@ ggsave(endo_status_map, filename = "endo_status_map.png", width = 10, height = 5
 #             n_match = sum(score_match),
 #             percent_match = n_match/n)
 # 
+mode <- function(codes){
+  which.max(tabulate(codes))
+}
+
+summary_endo_herb <- endo_herb %>% 
+  mutate(Sample_id_temp = Sample_id) %>% 
[…]

### 2023-03-22 endo_herbarium_spatial_analysis.R
adding more plots

@@ -103,9 +103,6 @@ endo_herb$collector_id <- collector_no[match(as.factor(endo_herb$collector_lastn
 contemp_surveys <- read_csv(file = "contemp_surveys.csv")
 contemp_random_sample <- read_csv(file = "contemp_random_sample.csv")
 
-contemp_random_sample <- contemp_random_sample %>% 
-  left_join(contemp_surveys)
-
 
 # register_google(key = "")
 # map <- ggmap::get_map(zoom = 3, source = "google", maptype = c("satellite"))
@@ -604,19 +601,17 @@ pred_data_long <- pred_data %>%
                              species == "ELVI" ~ "E. virginicus"))
 
 
-ggplot()+
-  geom_map(data = outline_map, map = outline_map, aes(long, lat, map_id = region), color = "grey", linewidth = .1, fill = "#FAF9F6")+
+prevalence_map_CI <- ggplot()+
+  geom_map(data = outline_map, map = outline_map, aes(map_id = region), color = "grey", linewidth = .1, fill = "#FAF9F6")+
   coord_sf(xlim = c(-109,-68), ylim = c(21,49)) +
-  geom_tile(data = filter(pred_data_long, variable == "pred_mean" & year == 1920 | variable == "pred_mean" & year == 2020), aes(x = lon, y = lat, fill = value), alpha = .8)+
-  facet_wrap(variable~year+species)+
-  scale_fill_gradient(
-    name = "Endophyte Prevalence",
-    low = "blue", high = "orange",
-    breaks = c(0,.25,.5,.75,1),
-    limits = c(0,1)) +
+  geom_tile(data = filter(pred_data_long, variable == "pred_lwr" & year == 1925 | variable == "pred_lwr" & year == 2020 | variable == "pred_upr" & year == 1925 | variable == "pred_upr" & year == 2020), aes(x = lon, y = lat, fill = value), alpha = .8)+
+  facet_wrap(species~year+variable, ncol = 4)+
+  scale_fill_viridis_c(limits = c(0,1))+
   theme_light()+
   # theme(strip.background = element_blank())+
-  labs(x = "Longitude", y = "Latitude")
+  labs(x = "Longitude", y = "Latitude") 
+# prevalence_map_CI
+ggsave(prevalence_map_CI, filename = "prevalence_map_CI.png", width = 15, height = 10)
   
 # making maps for each species for manuscript
 AGHY_prevalence <- ggplot()+
[…]

### 2023-03-22 Manuscript/endo_herbarium.tex
edits to manuscript and addding more plots, still working on discussion

@@
 	\section*{Abstract}
-Climate change has altered species’ phenologies, has triggered changes in traits, and has led to shifts in abundance and distributions. 
-Yet, we still know little about how biotic interactions alter these responses over long time scales across spatially heterogenous changes in climate. 
-Most species host beneficial microbial symbionts, \tom{which may provide resilience to changes in the environment, but at the same time increasing stress may lead these symbioses to breakdown}{A little confusing as written.
-Might be better to present as alernative hypotheses}.
+Many species' distributions and abundances have shifted in response to climate change. 
+Most species harbor microbial symbionts that have the potential to influence these responses.
+Mutualistic microbial symbionts may provide resilience to environmental change by protecting their hosts from increasing stress.
+Alternatively, environmental change that causes declines in host or symbiont may disrupt the interaction.
 Microbes can be preserved within the dried plant tissue of herbarium specimens presenting an opportunity to quantify changes across broad temporal and spatial scales. 
-\tom{Epichloë fungal endophytes are seed-transmitted symbionts of cool-season grasses which can provide drought tolerance to their hosts.
-}{I would cut this sentence and incorporate need to know info in other sentences.}
-\tom{ We ask how endophyte prevalence has responded to climate change.}{Could be sharpened.} 
- Specifically, we took samples of seeds from herbarium specimens of three grass host species collected over the last two centuries, quantified the presence or absence of fungal endophytes within the seeds with microscopy, and evaluated spatial and temporal trends in prevalence using \tom{spatial modeling}{Vague}.
- We found strong evidence that fungal endophyte prevalence has increased over the last two centuries across the three host species from 25\% prevalence to 75\% prevalence on average.
-\tom{We also found that observed changed in annual  and seasonal climate drivers were associated with temporal trends in endophyte prevalence. 
-In particular, changes in prevalence within each species corresponded to climate drivers during its peak growing season. }{Could be combined into one sentence.}
-Thus we provide novel evidence for \tom{adaptive biotic responses}{Not sure what you mean by this.} to ongoing climate change.
+We ask how the prevalence of a symbiont of grasses, \emph{Epichloë} fungal endophytes, which can protect hosts from drought, has changed over time in response to climate change, and how these changes vary across the hosts' ranges.
+Specifically, we took seed samples from herbarium specimens of three grass host species collected over the last two centuries, quantified fungal hyphae presence within the seeds using microscopy, and evaluated spatial and temporal trends.
+Overall, endophytes have increased in prevalence over the last two centuries from 25\% prevalence to 75\% prevalence on average across the three hosts.
+We also found that that the changes in prevalence were associated with observed changes in annual and seasonal climate drivers corresponding with each hosts' peak growing season. 
+Thus we provide novel evidence for a cryptic biological response to climate change that may contribute to the resilience of the symbiosis.
 
@@
 	% (e.g. section ``Counting Owl Eggs'').
-Predicting how species' interactions will be altered by ongoing climate change is a major goal of basic ecological research and conservation management. 
-\tom{Spatially heterogeneous changes in the environment have led species to move, adapt, or face extinction \citep{IPCC2012managing}.
-}{In theory a spatially homogenous change in temperature -- everywhere increases by 2 degrees -- would have the same effect.
-So significance of spatial heterogeneity here is not clear.}
-Species' responses are likely mediated by the numerous interspecific interactions in which they engage. 
-Yet for most species, empirically evaluating the role that biotic interactions can play in facilitating or hindering \tom{adaption}{I am struck by this word again.
-Are you intentionally bringing in evolutionary ideas?} to changing environmental conditions has been limited by a lack of data at adequate spatial and temporal scales \cite{gilman2010framework}.\tom{}{This paragraph has introduced two related but different ideas: how species interactions per se respond to climate change, and how interactinos mediate responses to climate change.
-This is a little disoroienting.}
+
+Understanding how biotic interactions are altered by global climate change is a major goal of basic ecological research and conservation management \cite{gilman2010framework,blois2013climate}.
+Documented responses to environmental change, such as shifts in species' distributions \cite{aitken2008adaptation} and phenology \cite{piao2019plant}, are typically blind to concurrent changes in associated biotic interactions.
+Empirically evaluating these biotic changes, whether interacting species shift in tandem with their partners or not \cite{hillerislambers2013will}, is crucial to predicting future climate change responses, but has been limited because, simply put, data on species interactions are limited, with few datasets extending over the long time scales of contemporary climate change \cite{poisot2021global}.
 
 Natural history collections, which were originally collected to study and preserve taxonomic diversity, present a unique opportunity to explore long-term changes in ecological interactions across broad spatial scales \citep{meineke2018unrealized}.
-Natural history collections, built and maintained by the efforts of thousands of scientists, are \tom{invaluable time machines}{Love this!}. 
-These collections primarily comprise physical specimens of organisms  along with information about the time and place of their collection. 
-These specimens are samples preserving physical legacies of ecological dynamics and species' interactions from dynamically changing environments across time and space.
-\tom{Recent works using plant collections have documented changes in phenology \citep{willis2017old, park2019herbarium,  berg2019examination}, as well as pollination \citep{pauw2011reconstruction, duan2019century}, and changes in antagonistic interactions such as herbivory \citep{meineke2019herbarium}, but to date, focus has been lacking on long-term changes in a particularly common type of biotic interaction, microbial symbioses.}{I like the content of this sentnece but  found the phrasing difficult.}
+Natural history collections, built and maintained by the efforts of thousands of scientists, are invaluable time machines, primarily comprised of physical specimens of organisms  along with information about the time and place of their collection.
+These specimens are samples preserving physical legacies of ecological processes and species' interactions from dynamically changing environments across time and space.
+Researchers have used plant collections to document climate change responses, including shifts in phenology \citep{willis2017old, park2019herbarium,  berg2019examination}, as well as in rates of pollination \citep{pauw2011reconstruction, duan2019century}, and herbivory \citep{meineke2019herbarium}.
+However, focus has been lacking on long-term changes in a particularly common type of biotic interaction, microbial symbioses.
 
 Microbial symbionts are common to all macroscopic organisms and can have important effects on their hosts' survival, growth and reproduction \cite{rodriguez2009fungal,mcfall2013animals}.
-Many microbial symbionts act as mutualists, engaging in reciprocally beneficial interactions with their hosts and often ameliorating host responses to environmental stress. 
+Many microbial symbionts act as mutualists, engaging in reciprocally beneficial interactions with their hosts that ameliorate environmental stress. 
 For example, bacterial symbionts of insects, such as \emph{Wolbachia}, can improve their hosts' thermal tolerance \citep{truitt2019wolbachia, renoz2019evolutionary}, and arbuscular mycorrhizal fungi, documented in 70-90\% of families of land plants \citep{parniske2008arbuscular}, allow their hosts to persist through drought conditions by improving water and nutrient uptake \citep{cheng2021elucidating}.
-On the other hand, changes in the mean and variance of environmental conditions may disrupt microbial  mutualisms by changing the costs and benefits of the interaction for each partner, leading to breakdown of the interaction \citep{aslan2013mutualism}.
+On the other hand, changes in the mean and variance of environmental conditions may disrupt microbial  mutualisms by changing the costs and benefits of the interaction for each partner, leading the interaction to deteriorate \citep{aslan2013mutualism}.
 Coral bleaching (the loss of symbiotic algae) due to temperature stress \citep{sully2019global} is perhaps the best known example, but this phenomenon is not unique to corals.
 Lichens exposed to elevated temperatures experienced loss of photosynthetic function along with changes in the community composition of their algal symbionts \citep{meyer2022climate}.
-How commonly and under what conditions microbial mutualisms breakdown or strengthem under climate change remain unanswered questions, but previous work suggests that these alternative responses may depend on the intimacy and specialization of the interaction as well as the physiological tolerances of the mutualists \citep{toby2010mutualisms, warren2014mutualism, rafferty2015phenological}.
-\tom{Given that climate change is spatially heterogenous, the rate of change may have a strong influence on when and where we can expect facilitation or disruption of mutualism.
-}{This is an important sentence that needs some elboration and citation.
-Can you explain more about spatial heterogeneity in climate change?
-And are you suggesting that the rate of change itself determines whether mutualism strengthens or breaks down?
-Also, ``facilitation'' is awkward in this context.}
+How commonly and under what conditions microbial mutualisms deteriorate or strengthen under climate change remain unanswered questions, but previous work suggests that these alternative responses may depend on the intimacy and specialization of the interaction as well as the physiological tolerances of the mutualists \citep{toby2010mutualisms, warren2014mutualism, rafferty2015phenological}.
 
-Early research on \emph{Epichloë} fungal endophytes, specialized symbionts of cool-season grasses, \tom{used herbarium specimens}{I would move this later in the paragraph, where you can make the point that this approach has not been leveraged to its full capacity.} to describe the broad taxonomic diversity of hosts \citep{white1985endophyte}.
-These facultative symbionts, which have been documented in $\sim 30$\% of cool-season grasses \citep{leuchtmann1992systematics}, are transmitted vertically from maternal plants to offspring through the grasses' seeds.
+Understanding of how microbial symbioses are affected by climate change is additionally complicated by spatial hetereogeneity in the direction and magnitude of environmental change \cite{ipcc_2021}. 
+Beneficial symbionts are likely able to shield their hosts from environmental stress in locations that experience a small degree of change, but symbionts in locations that experience changes of large magnitude may be pushed beyond their physiological limits \cite{webster2008temperature}.
+Additionally, symbionts are often unevenly distributed across their hosts' distribution.
+Facultative symbionts may be absent from portions of the host range \cite{afkhami2014mutualist}, and hosts may engage with a diversity of partners (different species, or different locally-adapted strains of one species) across their environments \cite{frade2008variation, rolshausen2018quantifying}.
+Identifying broader spatial trends in symbiont prevalence is therefore an important step in developing predictions of where to expect changes in the symbiosis in future climates.
+
+\emph{Epichloë} fungal endophytes are specialized symbionts of cool-season grasses, which have been documented in $\sim 30$\% of cool-season grasses \citep{leuchtmann1992systematics}
+There are transmitted vertically from maternal plants to offspring through the grasses' seeds.
 Vertical transmission creates a feedback between the fitness of host and symbiont \citep{fine1975vectors, douglas1998host, rudgers2009fungus}. 
 Over time, endophytes that act as mutualists for their hosts should rise in prevalence within a population. 
-\emph{Epichloë} are known to improve \tom{drought}{not just drought - also herbivory and pathogens.
-I would cite more to support the types of endo benefits.} tolerance for their hosts \citep{afkhami2014mutualist}, and are expected to reach high prevalence where these fitness benefits are most important.
+\emph{Epichloë} are known to improve their hosts' drought tolerance \cite{decunta2021systematic}, and protect their hosts against herbivores \cite{crawford2010fungal} and pathogens \cite{xia2018role} likely through the production of a diverse suite of alkaloids and other secondary metabolites.
+The fitness feedback induced by vertical transmission leads to the prediction that endophyte prevalence should be high in populations where these fitness benefits are most important. 
 Previous survey studies have documented large-scale spatial patterns in  endophyte prevalence structured by environmental gradients \citep{granath2007variation,bazely2007broad, afkhami2012fungal,sneck2017variation}.
-We predicted that \tom{prevalence should also track temporal changes in environmental drivers}{explain and elaborate}, but testing this has been impossible with surveys limited to a few years, and to the \tom{spatial extent reachable during a single growing season}{we have cars and planes so this should not be an obstacle.
-I think the real limitation is the combination spatio-temporal coverage, which is what herbaria uniquely provide}.
-Becauses the fungi are seed-transmitted, they are conveniently preserved \tom{along with their host plants on herbarium sheets}{as long as the specimen has seeds, which is almost always the case for grasses}, creating the opportunity to examine the response of the symbiosis to historical climate change.
+We predicted that prevalence should also track temporal changes in environmental drivers that elicit these fitness benefits.
+For example, endophyte-mediated drought tolerance should lead prevalence to increase in regions where precipitation declines over time.
+Early research on \emph{Epichloë} used herbarium specimens to describe the broad taxonomic diversity of hosts \citep{white1985endophyte}. 
+Grasses are commonly identified and collected based on the presence of their reproductive structures meaning that collected specimens typically contain seeds, conveniently preserving the fungi along with their host plants on herbarium sheets.
+This creates the opportunity to leverage the unique spatio-temporal sampling of herbarium collections to examine the response of the symbiosis to historical climate change.
 
-In this study, we took advantage of the spatial and temporal sampling within herbaria to assess for the first time whether grass-fungal endophyte symbioses have responded \tom{adaptively}{the adaptive nature of the response needs to be explained} to climate change.
-In particular, we (i) quantified the spatial and temporal change in endophyte prevalence across host ranges over the last two centuries, and (ii) identified climate drivers that can explain the magnitude of these responses.
-\tom{}{This paragraph would benefit from a specific hypothesis or question, which I think you have.
-Your emphasis on space vs time is a little muddled.
-You frame the paper around climate change, but part of your analysis - I think - is asking a purely spatial question.
-This paragraph should also mention the three host species and provide some wow-factor summary stats on the range of years, locations, and sample sizes.}
+In this study, we assessed for the first time the long-term responses of endophyte symbiosis to climate change through the use of herbarium specimens of three \emph{Epichloë} endophyte host species, \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}.
+We first address questions describing spatial and temporal trends in endophyte prevalence: (i) How does endophyte prevalence vary across space?, (ii) How has endophyte prevalence changed over time?
+and (iii) How have temporal changes in endophyte prevalence differed across space?
+We evaluated the predictive ability of this climate-implicit model using contemporary endophyte surveys as "out-of-sample test data, an important but rarely used strategy in ecological studies \cite{tredennick2021practical}.
+We then address how climate change may be driving these trends in endophyte prevalence by addressing the question: (iv) What is the relationship between variation in temporal trends in endophyte prevalence and changes in climate drivers?
+We examined 1,865 specimens collected across across eastern North America between 1824 and 2019.
 
 \section*{Methods}
-        \subsection*{Focal host species}
-We focused on three endophyte hosts, \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}.\tom{}{Describe the fungal species.}
-These cool-season (C3) grass species are commonly represented in natural history collections with broad distributions covering the Eastern United States.
-\emph{A. hyemalis} is a small short-lived perennial species that germinates in the spring and typically blooms between March and July.
-\emph{A. perennans} is of similar stature but is longer lived than \emph{Agrostis hyemalis} and blooms in late Summer and early Autumn. 
+        \subsection*{Focal species}
+Our surveys focused on three endophyte hosts, \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}.
+Both \emph{Agrostis} species host \emph{Epichloë amarillans} \cite{craven2001multigene, leuchtmann2014nomenclatural}, while \emph{Elymus virginicus} typically hosts \emph{Epichloë elymi} \cite{clay2002evolutionary}.
+These $C_3$ grass species are commonly represented in natural history collections with broad distributions covering much the Eastern United States.
+\emph{A. hyemalis} is a small short-lived perennial species that germinates in the spring and typically blooms between March and July (most common collection month: May).
+\emph{A. perennans} is of similar stature but is longer lived than \emph{Agrostis hyemalis} and blooms in late Summer and early Autumn (most common collection month: September). 
 This species is relatively more sparsely distributed, tending to be found in shadier and more moist habitats while \emph{A. hyemalis} is often found in open and recently disturbed ground. 
-Both\emph{Agrostis} species are recorded from throughout the Eastern US, but \emph{A.
+Both \emph{Agrostis} species are recorded from throughout the Eastern US, but \emph{A.
 perennans} has a slighty more northern distribution, whereas \emph{A.
@@
 \emph{E. virginicus} is a larger and relatively longer-lived  species that is more broadly distributed that the \emph{Agrostis} species. 
-It begins flowering as early as March or April but continues throughout the Summer.
+It begins flowering as early as March or April but continues throughout the Summer (most common collection month: July).
 
 		\subsection*{Herbarium surveys}
-We collected 5-10 seeds per specimen during visits to natural history collections between 2019 and 2022 \tom{(See Table A1 for a summary of collections)}{It is important to say more in the text here about where you went and what you did.} .
-\tom{Grasses are commonly identified and collected based on the presence of their reproductive structures which ensured that specimens we examined contained seeds.}{Move to intro} 
-\tom{Our sampling plan was designed to minimize damage to these specimens.}{How so?}
+We visited herbaria between 2019 and 2022, including The Botanical Research Institude of Texas (which also houses the Vanderbilt University herbarium and the R.
+Dale Thomas Plant collection from the University of Louisiana at Monroe), the Louisiana State University Herbarium, the Mercer Botanic Garden Herbarium, Missouri Botanic Garden's Herbarium, the S.M.
+Tracy Herbarium at Texas A\&M, the University of Kansas Herbarium, the Robert Bebb Herbarium at the University of Oklahoma, the Lundell Herbarium and the UT Herbarium at the University of Texas, and Oklahoma State University Herbarium (See Table A1 for a summary of specimens includes from each collection).
+Our sampling plan was designed to minimize damage to these specimens.
+We chose our focal species in part because they are commonly represented in herbarium collections, and produce high numbers of seeds, meaning that small samples would not diminish the value of the specimens for future studies.
+We collected 5-10 seeds per specimen after examining the herbarium sheet under a dissecting microscope to ensure that we collected viable seeds fit for our purpose of identifying fungal endophytes.
+
+After collecting seeds, we quantified the presence or absence of \emph{Epichloë} fungal hyphae, which grow intercellularly, using microscopy. 
+We first softened seeds with a 10\% NaOH solution, then stained the seeds with aniline blue dye and squashed them under a microscope cover slip. 
+We examined the squashed seeds for the presence of fungal hyphae at 100X magnification \cite{bacon2018stains}.
+In some cases, the tissues examined during microscopy came from flowers or otherwise non-viable seeds, which were excluded from the seed counts for that specimen.
+On average we scored 4.7 seeds per specimen in \emph{A. hyemalis}, 5 seeds per specimen in \emph{A. perennans}, and 3.8 seeds per specimen in \emph{E. virginicus}.
+Due to imperfect vertical transmission \cite{afkhami2008symbiosis}, it is possible that symbiotic host-plants produce a mixture of symbiotic and non-symbiotic seeds. 
+We therefore designated each specimen as symbiotic if endophyte hyphae were observed in one or more seeds, or non-symbiotic if hyphae were observed in zero seeds. 
+To capture uncertainty in the identification process, we recorded both a "liberal" and a "conservative" endophyte status for each plant.  
+When we identified potential endophytes with unusual morphology, low uptake of stain, or a small amount of fungal hyphae across the scored seeds, we recorded a positive liberal status (more likely to be a true endophyte) and a negative conservative status (less likely to be a true endophyte).
+$89$\% of scored plants had matching liberal and conservative endophyte statuses, reflecting high confidence in endophyte identification.
+The following analyses presented in the main text used the liberal statuses, but we repeated all analyses with the conservative statuses which yielded qualitatively similar results (Figure A5). 
+We excluded specimens for which information about the collection location and date were unavailable.
+In total, we quantified endophyte symbiosis for $964$ \emph{A.
+hyemalis} specimens collected between 1824 and 2019, $299$ \emph{A.
+perennans} specimens collected between 1863 and 2017, and $602$ \emph{E.
+virginicus} specimens collected between 1839 and 2017 (Fig.
+1, Fig 2A).
+Each specimen was assigned geographic coordinates based on collection information recorded on the herbarium sheet using the geocoding functionality of the ggmap R package \citep{kahle2019package}.
+Many specimens had digitized collection information readily available, but for those that did not, we transcribed information from pictures of the specimens. 
+Collections were geo-referenced to the nearest county centroid, or nearest municipality when that information was available. 
+For a few of the oldest specimens, only information at the state level was available, and so we used the state centoid.
 
@@
 	\includegraphics[width = \linewidth]{collections_map.png}
-	\caption{\textbf{Collection locations of herbarium specimens sampled for endopnhyte presence absence}.
+	\caption{\textbf{Collection locations of herbarium specimens sampled for endophyte presence absence}.
 Color designates host species ( \emph{A.
@@
 
-After collecting seeds, we quantified the presence or absence of \emph{Epichloë} fungal hyphae, which grow intercellularly, using microscopy. 
-We first softened seeds with a 10\% NaOH solution, then stained the seeds with aniline blue dye and squashed them under a microscope cover slip. 
-We examined the squashed seeds for the presence of fungal hyphae at 100X magnification \cite{bacon2018stains}.
-In some cases, the tissues examined during microscopy came from flowers or otherwise non-viable seeds, which were excluded from the seed counts for that specimen.
-On average we scored \textbf{number} seeds per specimen.
-Due to imperfect vertical transmission (cite), it is possible that symbiotic host-plants produce a mixture of symbiotic and non-symbiotic seeds. 
-We therefore designated each specimen as symbiotic if endophyte hyphae were observed in one or more seeds, or non-symbiotic if hyphae were observed in zero seeds. 
-To capture uncertainty in the identification process, we recorded both a "liberal" and a "conservative" endophyte status for each plant.  
-When we identified potential endophytes with unusual morphology, low uptake of stain, or a small amount of fungal hyphae across the scored seeds, we recorded a positive liberal status (more likely to be a true endophyte) and a negative conservative status (less likely to be a true endophyte).
-$89$\% of scored plants had matching liberal and conservative endophyte statuses, reflecting high confidence in endophyte identification.
-\tom{The following analyses presented in the main text used the liberal statuses, but we repeated all analyses with the conservative statuses which yielded qualitatively similar results (Appendix).
-}{Good}
+\subsection*{Assessing spatial and temporal changes in endophyte prevalence}
+To quantify spatial and temporal trends in endophyte prevalence, we used an approximate Bayesian method, Integrated Nested Laplace Approximation (INLA), to fit a spatio-temporal model of endophyte prevalence.
+Compared to MCMC Bayesian methods, INLA provides a computationally more efficient method of ascertaining model posteriors for certain models that can be formulated as latent Gausssian Models.
+\cite{rue2009approximate}.
+Many common statistical models, including structured and unstructured mixed-effects models, can be represented as latent Gaussian Models. 
+Fitting models with structured spatial effects is possible with MCMC sampling but can require long computation times, making INLA an effective alternative, which has been used to model spatial patterns in flowering phenology \cite{willems2022forest}, habitat overlap between marine predators and prey \cite{sadykova2017bayesian}, the distribution of temperate trees \cite{engel2022spatial} as well as the population dynamics of endangered amphibians \cite{knapp2016large} and other ecological processes \cite{beguin2012hierarchical}.
 
-We excluded specimens for which information about the collection location and date were unavailable.
-In total, we quantified endophyte symbiosis for $971$ \emph{A.
-hyemalis}, $302$ \emph{A.
-perennans}, and $605$ \emph{E.
-virginicus} specimens spanning the years \tom{$1824$ to $2019$}{give date range by species}.
-\tom{Each specimen was assigned a collection location based on geographic information recorded at the time of collection using the ggmap package in R \citep{kahle2019package}.}{confusing-not sure what this means}
-Collections were geo-referenced to the nearest county centroid, or nearest municipality when that information was available. 
-For a few of the oldest specimens, only information at the state level was available, and so we used the state centoid \tom{(Fig.
-1; Fig A1).}{I don't see how this is shown in figure 1.
-Also, can you use more distinctive colors in Fig 1?}
-
-\subsection*{Assessing spatial and temporal changes in endophyte prevalence}
-\tom{}{Give a quick overview of the types of models you used and what specific purpose they were built for before getting into details.
-The fact that you estimated climate effects following an initial climate-implicit model needs to come through more clearly.}To quantify spatial and temporal trends in endophyte prevalence, we modeled endophyte presence/absence ($P$) as a Bernoulli response variable for specimen $i$ of host species $h$.
+We modeled endophyte presence/absence ($P$) as a Bernoulli response variable for specimen $i$ of host species $h$.
 
@@
 
-The expected endophyte prevalence, $\hat{P}$, was modelled with intercepts specific to each host species ($\beta_{0}$), slopes for changes over time ($\beta_{1}$) \tom{as well as the interaction between time and the specimen's latitude and longitude ($\beta_{2}$, $\beta_{3}$, and $\beta_{4}$).
-}{So you do not have equal emphasis on space and time.
-YOu are really asking a time question and considering how temporal trends may vary spatially.
-I think this needs to come through more clearly in the introduction.}
+The expected endophyte prevalence, $\hat{P}$, was modelled with intercepts specific to each host species ($\beta_{0}$), slopes for changes over time ($\beta_{1}$) as well as the interaction between time and the specimen's latitude and longitude ($\beta_{2}$, $\beta_{3}$, and $\beta_{4}$).
+To quantify spatial patterns in prevalence, we included a spatially-dependent random intercept, $\phi$.
+This random spatial effect was modeled as a stochastic partial differential equation (SPDE) that depended on a covariance matrix according to the proximity of each collection location \citep{lindgren2011explicit,bakka2018spatial}.
+The covariance matrix is approximated using a Matérn covariance function, with each data point assigned a location according to the nodes of a mesh of non-overlapping triangles across our study area (Fig A2).
 We accounted for potential biases introduced during the process of collecting specimens as well as in scoring ability by including random effects specific to each collector $\chi$ and to each scorer $\omega$.
-To quantify spatial patterns in prevalence, we included a spatially-dependent random intercept, $\phi$, and fit the model using integrated nested Laplace approximation, an approximate Bayesian method implemented in the rINLA package \citep{lindgren2015bayesian}.
-This method allowed us to model spatial dependency between data points (nearer collections may be more similar) as a continuous process with stochastic partial differential equations (SPDE) that depend on a covariance matrix according to the proximity of each collection location \citep{lindgren2011explicit,bakka2018spatial}.
-The covariance matrix is approximated using a Matérn covariance function, with each data point assigned a location according to the nodes of a mesh of non-overlapping triangles across our study area (Fig A2).
-We fit the model with vague priors, and compared models with different sizes of mesh, which had little effect on the resulting model estimates.
-\tom{We assessed model fit with graphical postterior predictive checks (Fig. A3, Fig. A4).}{This is good but also worth mentioning that you assessed convergence.}
+We performed modeling fitting using the rINLA package, \citep{lindgren2015bayesian}, with vague priors, and compared models with different sizes of mesh, which had little effect on the resulting model estimates.
+Posterior modes were stable indicating that numeric convergence was successful.
+We assessed model fit with graphical posterior predictive checks (Fig. A3).
+The model performed adequately at classifying the historical data (AUC = 0.77; Fig. A4).
 
-\tom{We used data on endophyte prevalence from  contemporary surveys of \emph{A.
-hyemalis} as test data to evaluate model fit.
-}{This is an extraordinary extra step that should be played up in the Intro and included in the aims at the end of the Intro.}
-During these surveys, which took place between 2015 and 2020, we collected seeds from living plants from 43 populations across a portion of the species' range and then quantified endophyte status with staining microscopy as described for the herbarium surveys (Fig S2).
-We compared the model's prediction for these locations to the endophyte status of randomly sampled plants from each surveyed location, \tom{mimicing the process of collecting a single herbarium specimen from the population.}{But isn't the prevalence the better test of the model's predictions?}
-The model performed adequately (\tom{training data: AUC = 0.77; test data: AUC = 081}{Is this 0.81? The model performs better on the test data than the training data? This is odd.}; Fig. A4).
+We used data from contemporary surveys of on endophyte prevalence  in \emph{A. hyemalis} and \emph{E. virginicus} as test data to evaluate predictions of the model. 
+During these surveys, which took place between 2013 and 2020, we collected seeds from living plants from 63 populations(average number of plants sampled: 22.9) across a portion of each species' range (Fig A4).
+We quantified the endophyte status with staining microscopy as described for the herbarium surveys and calculated prevalence of endophytes within the population (proportion of symbiotic plants divided by the number of sampled plants).
+We compared the model's prediction for these locations to the observed population prevalence.
 
 		\subsection*{Assessing the role of climate drivers}
-To assess how the rate of climate change may have driven changes in endophyte prevalence,\tom{ we calculated the change in climate between recent (1990 to 2020) and historic (1895 to 1925) periods.}{This is where your hypotheses and predictions in the intro need to come through more clearly.
-Also, this assumes change has been constant during those periods and does not use most years of data.
-Could you try estimating a slope for each location?
-Is this possible?}
+
+We assessed how the magnitude of climate change may have driven changes in endophyte prevalence by assessing correlations between changes in climate and changes in endophyte prevalence predicted from our spatial model at evenly spaced pixels across the study area.
 We first downloaded monthly temperature and precipitation rasters from the PRISM climate group \citep{daly2013prism} covering the time period between 1895 and 2020 using the 'prism' R package \citep{Rprism2015}.
 Prism provides reconstructions of historic climate variables across the United States by spatially-interpolating weather station data \citep{diLuzio2008constructing}. 
-We calculated 30-year climate normals for annual and seasonal mean temperature and cumulative precipitation for the recent and historic periods.
-\tom{We used three four-month seasons within the year (Spring: January, February, March, April; Summer: May; June, July, August; Autumn: September, October, November, December).
-}{I think this needs more biological justification.
-I can imagine a simpler divion of cool season and warm season.}
-\tom{We additionally calculated the coefficient of variation of each climate driver during each period.}{why? and how? was this based on monthly means or sums?}
+We calculated 30-year climate normals for annual and seasonal mean temperature and cumulative precipitation for the recent (1990 to 2020) and historic (1895 to 1925) periods.
+We used three four-month seasons within the year (Spring: January, February, March, April; Summer: May; June, July, August; Autumn: September, October, November, December). 
+This division of seasons allowed us to quantify differences in climate associated with the two "cool" seasons  that shoulder summer when we expect our focal species to be most biologically active (\emph{A.
+perennans}: Spring; \emph{E.
+virginicus}: Spring and Summer; \emph{A.
+perennans}: Fall).
+In addition to mean climate conditions, environmental variability  in and of itself can influence population dynamics \cite{tuljapurkar_population_1982} and changes in variability are a key prediction of climate change models \cite{stocker2013technical, ipcc_2021}.
+So we calculated the coefficient of variation during each period for each annual and seasonal climate driver as the interannual standard deviation divided by the mean across each 30-year period.
 We then took the difference between recent and historic periods for the mean and coefficient of variation for each climate driver (Fig. A5).
 Because initial analyses indicated a high degree of collinearity between seasonal and annual changes in temperature, we used annual temperature only, along with annual and seasonal precipitation, in the subsequent analysis.
+All together, this left us with measurements of change in 10 potential climate drivers: the mean and coefficient of variation of annual temperature, as well as the mean and coefficient of variation of cumulative annual precipitation, cumulative spring precipitation, cumulative summer precipitation, and cumulative autumn precipitation (Fig A8-A9).
 
-We then evaluated how predicted changes in endophyte prevalence correlated with changes in the environment at an evenly spaced grid of points across the study area.\tom{}{Map?} 
-Because regions varied in their predicted starting endophyte prevalence, we calculated the \tom{relative change}{calculated how?} in endophyte prevalence between 1925 and 2020. 
+We calculated the relative change in endophyte prevalence across the 1925 to 2020, i.e., the difference between predicted prevalence divided by the endophyte prevalence in 1925, because regions varied in their predicted starting prevalence.
 This time period connects the predicted change in prevalence with the endpoints of the available climate record.
-We then calculated the Pearson correlation coefficient between the relative change in endophyte prevalence and each climate driver.\tom{}{This paragraph should be clearer that you are using predicted prevalence from the model.}
+We then calculated the Pearson correlation coefficient between the relative change in endophyte prevalence and each climate driver.
 
@@
 \section*{Results}
+\subsection*{Temporal and spatial trends}
+We quantified temporal and spatial trends in endophyte prevalence while accounting for potential biases introduced by collectors and by individuals who quantified endophyte presence/absence with the use of random effects.
+We found no evidence that collector biases influenced our results (random effects values centered around 0; Fig 2A), while the identity of individual scorers did contribute to observed patterns in endophyte prevalence (3 of the 16 scorers were more likely than average to assign positive endophyte status, as indicated by 95\% credible intervals that do not overlap 0) (Fig 2B).
+Interpretation of the data without accounting for this source of variation would lead to potentially biased interpretations of temporal and spatial trends.
 
-\subsection*{Temporal and spatial trends}
+\begin{figure}[H]
+	\label{fig:random_fx}
+	\centering
+	\includegraphics[width = .8\linewidth]{random_fx_plot.png}
+	\caption{\textbf{Posterior estimates of (A) collector and (B) scorer random effects.} Points show tthe posterior mean along with 95\% CI for random effects estimate from 532 collectors and 16 scorers.}
+\end{figure}
 
-We found that endophyte prevalence increased within the examined specimens over the last two centuries for all three host species (Fig. 2). 
+\subsubsection*{How does endophyte prevalence vary across space?}
+Across space, there were clear trends in endophyte prevalence which varied between species (Fig. 3).
+\emph{Elymus virginicus} had higher prevalence towards the northern portions of its range. 
+In contrast, \emph{Agrostis hyemalis} had highest prevalence towards the center of its range with regions of low prevalence towards the northeast and towards its western range edge.
+\emph{Agrostis perennans} had  highest prevalence towards the wester portion portion of it's range edge.
+There is considerable uncertainty in the spatial pattern of endophyte prevalence (See Fig.
+A10 for projections of the 95\% credible interval), however these broad spatial patterns are consistent across the low and high probability model predictions.
+
+\begin{figure}[H]
+	\label{fig:prevalence_map}
+	\centering
+	\includegraphics[width = \linewidth]{prevalence_map.png}
+	\caption{\textbf{Mean predicted endophyte prevalence for each host species (columns) in 1925 (top row) and 2020 (bottom row)}.
+Color indicates mean predicted rate of endophyte prevalence across a convex boundary enscribing the collections locations of each species (excluding two disjoint collections of \emph{A.
+perennans} to the west).}
+\end{figure}
+
+\subsubsection*{How has endophyte prevalence changed over time?}
+We found that endophyte prevalence increased within the examined specimens over the last two centuries for all three host species (Fig. 4). 
 On average, \emph{A.
@@
 2A), lead to qualitatively similar predictions, and so we continued analyses with the full dataset.
-\tom{}{I really like fig 2 but if you are going to split the predictions by latitude you should explain why, talk about this in the results, and also split the points by high and low latitude.
-Were 35 and 43 the bounds of the data?}
+Our model indicates a higher certainty that overall temporal trends are positive for \emph{A.
+hyemalis} and \emph{A.
+perennans} than for \emph{E.
+virginicus} (99\% probability of a positive overall year slope in \emph{A.
+hyemalis}, 89\% probability of a positive overall year slope in \emph{A.
+perennans}, and 58\% probability of a positive overall year slope in \emph{E.
+virginicus}) .
 
@@
 
-\tom{Across space, there were clear trends in endophyte prevalence which varied between species (Fig. 3).}{This seems to contradict my earlier comment about how your focus is mainly time.} 
-For example, \emph{Elymus virginicus}, \tom{had higher prevalence in the northern and eastern portions of it's range}{The model predicts low prevalence in Florida but there are basically no records from Florida, so the prediction does not seem very meaningful.
-See also next comment.}.
-In contrast, \emph{Agrostis hyemalis} had highest prevalence towards the center of its range with regions of low prevalence towards the north east and towards its western range edge.
-\tom{Our model revealed that while there was an overall increase in endophyte prevalence, these changes varied across the host species' ranges.}{What is problematic here is that the model is not showing predictions relative to each species' range, it is showing the aggregate sampling region.
-I think this will be more effective and meaningful if it is based on species-specific ranges.}
+\subsubsection*{How have temporal changes in endophyte prevalence differed across space?}
+Our model revealed that while there was an overall increase in endophyte prevalence, these changes varied across the host species' ranges.
+Temporal trends vary with latitude and longitude, as evidenced by variation in the spatial pattern of endophyte prevalence through time in Fig.
+3, and that temporal trends in Fig.
+4 vary at different rates according to their latitude.
 In some regions, \emph{A.
-perennans} experienced increases in percent prevalence by as much as 35 percentage points between 1920 and 2020, while \emph{A.
+perennans} experienced increases in percent prevalence by as much as 35 percentage points between 1925 and 2020, while \emph{A.
 hyemalis} and \emph{E.
@@
 In other regions, there were smaller increases of only five percentage points for \emph{E. virginicus} and fifteen percentage points for \emph{A. perennans}. 
-\tom{Notably, the western range edge of \emph{A. hyemalis} did not experience an increase in endophyte prevalence. }{Why is this notable? Also, see prev comment about real spp ranges.}
-Collections are sparser to the west as the species reaches its range limit, but the collected specimens were exclusively symbiont-free. (Fig. A1). 
-\tom{}{For the prevalence map, could you also map the difference between past and present prevalence to show hotspots and coldspots?}
+We calculated relative change in endophyte prevalence because regions differ in their initial endophyte prevalence.
 
 \begin{figure}[H]
-	\label{fig:prevalence_map}
+	\label{fig:rel_change_plot}
 	\centering
-	\includegraphics[width = \linewidth]{prevalence_map.png}
-	\caption{\textbf{Mean predicted endophyte prevalence for each host species (columns) in 1920 (top row) and 2020 (bottom row)}. Color indicates mean predicted rate of endophyte prevalence.}
+	\includegraphics[width = .8\linewidth]{rel_change_plot.png}
+	\caption{\textbf{Relative change in predicted prevalence for each host species.} Color indicates the relative change in predicted endophyte prevalence between 1925 and 2020 across the collection range.}
 \end{figure}
 
-\tom{}{The results section describes spatial and temporal trends but not the combination, which I thought was the main point of the paper.}
-\tom{}{There is nothing in this section that addresses statistical significance, however one wants to define that.}
+This reveals locations that are hotspots of change in endophyte prevalence (Fig. 5). 
+The overall magnitude of changes were greatest in \emph{A. perennans}, particularly in the northeast and in the south, where endophytes were previously only found in low prevalence (Fig. 3).
+\emph{A.
+hyemalis} and \emph{E.
+virginicus} experience lower overall increases in prevalence, yet we can still identify hotspots of change in the northeast for \emph{A.
+hyemalis} and in the southeast for \emph{E.
+virginicus}.
 
 \subsection*{The role of climate drivers}
-We found that the trends in endophyte prevalence were strongly associated with seasonal climate change drivers \tom{(Fig.
-4).}{How will you respond to the reviewer who asks you which of these correlations are significant?}
-For the majority of the study region, the climate has become wetter and cooler over the last century (Fig. A5), a consequence of regional variation in global climate change. 
-Within the study region, spatially heterogeneous environmental changes were predictive of changes in endophyte prevalence in line with differences in species' reproductive phenology. 
-For example, \tom{relative change in endophyte prevalence}{can you describe this in terms of increasing or decreasing prevalence, to make the interpretation more intuitive?} within \emph{E.
-virginicus} were most associated with declines in Summer precipitation  (a negative correlation in Fig.
-4) as well as with increases in the year-to-year variability of summer precipitation (a positive correlation in Fig.
+\subsubsection*{What is the relationship between variation in temporal trends in endophyte prevalence and changes in climate drivers?}
+
+We found that trends in endophyte prevalence were strongly associated with seasonal climate change drivers (Fig. 6).
+For the majority of the study region, the climate has become wetter and cooler over the last century (Fig. A5), a consequence of regional variation in global climate change \cite{ipcc_2021}. 
+Within the study region, spatially heterogeneous environmental changes were predictive of changes in endophyte prevalence. 
+For example, strong increases in prevalence within \emph{E.
+virginicus} were most associated (as indicated by correlation coefficients $>.3$) with declines in Summer precipitation  (a negative correlation in Fig.
+4) as well as with increases in the year-to-year variability of Summer precipitation (a positive correlation in Fig.
 4).
-These associations are in line with expectations that endophyte symbiosis provides drought tolerance to their hosts, particularly during \emph{E.
+Changes were also associated with reductions in average annual temperatures, and increases in year-to-year temperature variability.
+\emph{A. perennans} endophyte prevalence increased most strongly in regions that experienced reduced spring precipitation and reduced variability in annual temperature.
+Although these correlations were weaker, changes in \emph{A.
+perennans} endophyte prevalence were also associated with increased in increases in annual precipitation and increasing autumn precipitation.
+For \emph{A. hyemalis}, endophyte prevalence increased most strongly in regions that experienced reductions in autumn precipitation variability.
+
+%These associations are in line with expectations that endophyte symbiosis provides drought tolerance to their hosts, particularly during \emph{E.
 virginicus}'s \tom{summer growing season}{It is a stretch to say that this species has a summer growing season.}.
-For \emph{A.
+%For \emph{A.
 perennans}, which flowers in the late Summer and Autumn, increases in autumn precipitation along with increases in variability in annual precipitation, were strongly assocated with increasing endophyte prevalence.
-While this result is contrary to the expectation that endophyte provide drought tolerance, this species typically is associated with wetter microhabitats, and it also \tom{suggests a role for endophytes to play in buffering the negative consequences of demographic variability \citep{lewontin_population_1969}}{I would save for discussion}.
-For \emph{A. hyemalis}, which flowers in the Spring, changes in endophyte prevalence were more weakly associated with seasonal changes in climate drivers than for the other two species.
-It was, however, the only species' for which greater increases in endophyte prevalence were associated with regions that experienced increases in average annual temperatures.\tom{}{Is this a significant association, however you want to define that?
+%While this result is contrary to the expectation that endophyte provide drought tolerance, this species typically is associated with wetter microhabitats, and it also \tom{suggests a role for endophytes to play in buffering the negative consequences of demographic variability \citep{lewontin_population_1969}}{I would save for discussion}.
+%For \emph{A. hyemalis}, which flowers in the Spring, changes in endophyte prevalence were more weakly associated with seasonal changes in climate drivers than for the other two species.
+%It was, however, the only species' for which greater increases in endophyte prevalence were associated with regions that experienced increases in average annual temperatures.\tom{}{Is this a significant association, however you want to define that?
 Or could this result have occurred by chance?}
@@
 	\centering
-	\includegraphics[width = .8\linewidth]{climate_corr_heatmap.png}
-	\caption{\textbf{Changes in climate drivers associated with relative change in endophyte prevalence.} Color denotes the Pearson correlation coefficient between the relative rate of change in endophyte prevalence and the change in annual mean temperature ($^oC$) and cumulative seasonal precipitation (mm), as well as the change in coefficient of variation of each climate driver.}
+	\includegraphics[width = \linewidth]{climate_corr_heatmap.png}
+	\caption{\textbf{Correlations between changes in climate drivers and changes in endophyte prevalence.} Color denotes the Pearson correlation coefficient between the relative rate of change in endophyte prevalence and the change in annual mean temperature ($^oC$) and total annual and seasonal precipitation (mm), as well as the change in the coefficient of variation of each climate driver.
+Positive correlation coefficients indicate that greater increases in a climate driver were associated with larger increases in endophyte prevalence, while negative values indicate that .
+Asteriks denote correlation coefficients $\> .3$ or $< -.3$.}
 \end{figure}
@@
 \section*{Discussion}
-\tom{Our examination of historic plant specimens revealed a cryptic biotic reponse to climate change. }{Great opening sentence.}
-\tom{The geographic distributions and phenology of macroorganisms are apparent indicators of climate change.}{Not sure why you go here next, then go back to endophytes.}
+Our examination of historic plant specimens revealed a cryptic biotic reponse to climate change. 
 For the three host species we examined, there have been clear increases in fungal endophyte prevalence over the last two centuries.
@@
 		\caption{\textbf{ROC plot showing model performance classifying observations according to endophyte status.} The curves show adequate model performance for observed (top) and test (bottom) data.
-The AUC for each is 0.77 and 0.81 respectively.}
+The AUC for each is 0.77.}
 	\end{figure}
@@
 	\centering
-	\includegraphics[width = \linewidth]{climate_change_plot.png}
-	\caption{\textbf{Change in 30-year climate normals between the periods ending in 1915 and 2020.} Color represented change in cumulative precipitation (top row) and average seasonal temperature (bottom row) for the study area.}
+	\includegraphics[width = \linewidth]{fixed_plot.png}
+	\caption{\textbf{Comparison of posterior estimates of fixed effects when using Liberal or Conservative endophyte scores.}}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width = \linewidth]{test_data_map.png}
+	\caption{\textbf{Locations of contemporary surveys of endophytes in \emph{A. hyemalis} used as "test" data (red points), relative to the historical collection data (black crosses).}}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width = \linewidth]{contemp_test_plot.png}
+	\caption{\textbf{Predicted vs observed endophyte prevalence for contemporary test data.} (A) The model, trained on historic herbarium collection data, performed modestly at predicting contemporary endophyte prevalence, as indicated by some overlap of predicted 95\% CI with the 1:1 line, however contemporary test data generally had more variance between populations than model predictions.
+The model did recapitulate broader regional trends across (B) longitude and (C) latitude.
+Colors indicate host species, and point size in panels B and C reflect sample sizes of contemporary endophyte population surveys.}
+\end{figure}
+
+	\begin{figure}[H]
+	\centering
+	\includegraphics[width = .8\linewidth]{ppt_change_maps.png}
+	\caption{\textbf{Change in precipitation between the periods 1895-1925 and 1990-2020.} Color represents change in annual or seasonal total precipitation (A,C,E,G) and in the coeffiecient of variation of annual or seasonal total precipitation (B,D,F,H).
+Maps show the study area of \emph{A.
+hyemalis}.
+Map pixels used in correlation analysis with endophyte change were pulled from studies areas specific to each host species.
+}
+\end{figure}
+
+	\begin{figure}[H]
+		\centering
+		\includegraphics[width = \linewidth]{tmean_change_maps.png}
+		\caption{\textbf{Change in temperature between the periods 1895-1925 and 1990-2020.} Color represents change in annual mean temperature (A) and in the coeffiecient of variation of annual mean temperature (B).
+Maps show the study area of \emph{A.
+hyemalis}.
+Map pixels used in correlation analysis with endophyte change were pulled from studies areas specific to each host species.}
+	\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width = \linewidth]{prevalence_map_CI.png}
+	\caption{\textbf{Uncertainty associated with spatial trends in endophyte prevalence.} Color represents change in predicted endophyte prevalence.
+Panels show upper and lower 95\% posterior probability for each host species between 1925 and 2020.}
 \end{figure}
@@
 	\begin{table}[h]
-		\caption{Summary of herbarium sampling}
+		\caption{Summary of herbarium samples across collections}
 		\label{Table:herbaria}
@@
 		\begin{tabular}{llll}\hline
-			Collection        & AGHY        & AGPE      &      ELVI\\ \hline
-			Botanical Research Institute of Texas &      &    &        \\
-			Louisiana State University &       &   &          \\
-			Mercer Botanic Garden &       &      &     \\
-			Missouri Botanic Garden&  & & \\
-		    Texas A\&M &  & & \\
-		    University of Kansas &  &&  \\
-		    University of Oklahoma &  &&  \\
-		    University of Texas     &  & & \\		    				 			     			     
-			Oklahoma State University&       &       &   \\ \hline
+			Herbarium Collection        & AGHY        & AGPE      &      ELVI\\ \hline
+			Botanical Research Institute of Texas &   341   &    189&    176    \\
+			Louisiana State University &     71  & --  &   61       \\
+			Mercer Botanic Garden &   3    & --     &     6\\
+			Missouri Botanic Garden& 78 & 39& 31\\
+		    Texas A\&M &  73&-- & 49 \\
+		    University of Kansas & 134 & -- &  20\\
+		    University of Oklahoma & 65 &30&  91\\
+		    University of Texas  \& Lundell   &  169& 41& 99\\		    				 			     			     
+			Oklahoma State University&     30  &   --    &  69 \\ \hline
 		\end{tabular}

### 2023-03-23 Manuscript/endo_herbarium.tex
edits to the discussion and lots of stuff

@@
 \usepackage{fullpage}
-\usepackage[authoryear,sectionbib,sort]{natbib}
+\usepackage[authoryear,round,sectionbib,sort]{natbib}   % omit 'round' option if you prefer square brackets
+\let\cite\citep
+
 \linespread{1.7}
@@
 
-\title{Increasing prevalence of plant-fungal symbiosis over 195 years of climate change}
+\title{Increasing prevalence of plant-fungal symbiosis over two centuries of climate change}
 
@@
 Documented responses to environmental change, such as shifts in species' distributions \cite{aitken2008adaptation} and phenology \cite{piao2019plant}, are typically blind to concurrent changes in associated biotic interactions.
-Empirically evaluating these biotic changes, whether interacting species shift in tandem with their partners or not \cite{hillerislambers2013will}, is crucial to predicting future climate change responses, but has been limited because, simply put, data on species interactions are limited, with few datasets extending over the long time scales of contemporary climate change \cite{poisot2021global}.
+Empirically evaluating these biotic changes, whether interacting species shift in tandem with their partners or not \cite{hillerislambers2013will}, is crucial to predicting future climate change responses.
+Such evaluations have been limited because data on species interactions are limited, with few datasets extending over the long time scales of contemporary climate change \cite{poisot2021global}.
 
@@
 \emph{Epichloë} fungal endophytes are specialized symbionts of cool-season grasses, which have been documented in $\sim 30$\% of cool-season grasses \citep{leuchtmann1992systematics}
-There are transmitted vertically from maternal plants to offspring through the grasses' seeds.
+They are transmitted vertically from maternal plants to offspring through the grasses' seeds.
 Vertical transmission creates a feedback between the fitness of host and symbiont \citep{fine1975vectors, douglas1998host, rudgers2009fungus}. 
@@
 and (iii) How have temporal changes in endophyte prevalence differed across space?
-We evaluated the predictive ability of this climate-implicit model using contemporary endophyte surveys as "out-of-sample test data, an important but rarely used strategy in ecological studies \cite{tredennick2021practical}.
+To answer these questions, we fit a spatio-temporal statistical model of endophyte prevalence.
+and evaluated the predictive ability of the model using contemporary endophyte surveys as out-of-sample test data, an important but rarely used strategy in ecological studies \cite{tredennick2021practical}.
 We then address how climate change may be driving these trends in endophyte prevalence by addressing the question: (iv) What is the relationship between variation in temporal trends in endophyte prevalence and changes in climate drivers?
@@
 hyemalis} is found rarely as far north as Canada and is listed as a rare plant in Minnesota.
-\emph{E. virginicus} is a larger and relatively longer-lived  species that is more broadly distributed that the \emph{Agrostis} species. 
+\emph{E. virginicus} is a larger and relatively longer-lived  species that is more broadly distributed than the \emph{Agrostis} species. 
 It begins flowering as early as March or April but continues throughout the Summer (most common collection month: July).
@@
 Dale Thomas Plant collection from the University of Louisiana at Monroe), the Louisiana State University Herbarium, the Mercer Botanic Garden Herbarium, Missouri Botanic Garden's Herbarium, the S.M.
-Tracy Herbarium at Texas A\&M, the University of Kansas Herbarium, the Robert Bebb Herbarium at the University of Oklahoma, the Lundell Herbarium and the UT Herbarium at the University of Texas, and Oklahoma State University Herbarium (See Table A1 for a summary of specimens includes from each collection).
+Tracy Herbarium at Texas A\&M, the University of Kansas Herbarium, the Robert Bebb Herbarium at the University of Oklahoma, the Lundell Herbarium and the UT Herbarium at the University of Texas, and Oklahoma State University Herbarium (See Table A1 for a summary of specimens included from each collection).
+In total, we quantified endophyte symbiosis for $964$ \emph{A.
+hyemalis} specimens collected between 1824 and 2019, $299$ \emph{A.
+perennans} specimens collected between 1863 and 2017, and $602$ \emph{E.
+virginicus} specimens collected between 1839 and 2017 (Fig.
+1, Fig 2A).
 Our sampling plan was designed to minimize damage to these specimens.
 We chose our focal species in part because they are commonly represented in herbarium collections, and produce high numbers of seeds, meaning that small samples would not diminish the value of the specimens for future studies.
-We collected 5-10 seeds per specimen after examining the herbarium sheet under a dissecting microscope to ensure that we collected viable seeds fit for our purpose of identifying fungal endophytes.
+We collected 5-10 seeds per specimen after examining the herbarium sheet under a dissecting microscope to ensure that we collected mature seeds, not florets or unfilled seeds, fit for our purpose of identifying fungal endophytes with microscopy.
+We excluded specimens for which information about the collection location and date were unavailable.
+Each specimen was assigned geographic coordinates based on collection information recorded on the herbarium sheet using the geocoding functionality of the ggmap R package \citep{kahle2019package}.
+Many specimens had digitized collection information readily available, but for those that did not, we transcribed information from pictures of the specimens. 
+Collections were geo-referenced to the nearest county centroid, or nearest municipality when that information was available. 
+For a few of the oldest specimens, only information at the state level was available, and so we used the state centoid.
 
-After collecting seeds, we quantified the presence or absence of \emph{Epichloë} fungal hyphae, which grow intercellularly, using microscopy. 
+After collecting seed samples, we quantified the presence or absence of \emph{Epichloë} fungal hyphae, which grow intercellularly, in each specimen using microscopy. 
 We first softened seeds with a 10\% NaOH solution, then stained the seeds with aniline blue dye and squashed them under a microscope cover slip. 
@@
 The following analyses presented in the main text used the liberal statuses, but we repeated all analyses with the conservative statuses which yielded qualitatively similar results (Figure A5).
-We excluded specimens for which information about the collection location and date were unavailable.
-In total, we quantified endophyte symbiosis for $964$ \emph{A.
-hyemalis} specimens collected between 1824 and 2019, $299$ \emph{A.
-perennans} specimens collected between 1863 and 2017, and $602$ \emph{E.
-virginicus} specimens collected between 1839 and 2017 (Fig.
-1, Fig 2A).
-Each specimen was assigned geographic coordinates based on collection information recorded on the herbarium sheet using the geocoding functionality of the ggmap R package \citep{kahle2019package}.
-Many specimens had digitized collection information readily available, but for those that did not, we transcribed information from pictures of the specimens. 
-Collections were geo-referenced to the nearest county centroid, or nearest municipality when that information was available. 
-For a few of the oldest specimens, only information at the state level was available, and so we used the state centoid.
 
@@
 This random spatial effect was modeled as a stochastic partial differential equation (SPDE) that depended on a covariance matrix according to the proximity of each collection location \citep{lindgren2011explicit,bakka2018spatial}.
-The covariance matrix is approximated using a Matérn covariance function, with each data point assigned a location according to the nodes of a mesh of non-overlapping triangles across our study area (Fig A2).
+The SPDE model allowed us to flexibly estimate endophyte prevalence at locations across the study region, while accounting for potential spatial autocorrelation between data points via the covariance matrix.
+The covariance matrix was approximated using a Matérn covariance function, with each data point assigned a location according to the nodes of a mesh of non-overlapping triangles across our study area (Fig A2).
 We accounted for potential biases introduced during the process of collecting specimens as well as in scoring ability by including random effects specific to each collector $\chi$ and to each scorer $\omega$.
-We performed modeling fitting using the rINLA package, \citep{lindgren2015bayesian}, with vague priors, and compared models with different sizes of mesh, which had little effect on the resulting model estimates.
+Previous work has questioned whether the behavior of historical botanists and uneven sampling may introduce biases into ecological inferences made from historic collections \cite{kozlov2020biases}. 
+Prolific collectors who contribute thousands of specimens may be more or less likely to collect certain species, or specimens with certain traits \cite{daru2018widespread}. 
+Similarly, the process of scoring seeds for hyphae involved many individual researchers, who may vary in their likelihood of positively identifying fungal hyphae. 
+By including a random effect for collectors and for scorers, we can account for the variance across individual researchers that may bias our predictions of changes in endophyte prevalence.
+We performed model fitting using the rINLA package, \citep{lindgren2015bayesian}, with vague priors, and compared models with different sizes of mesh, which had little effect on the resulting model estimates.
 Posterior modes were stable indicating that numeric convergence was successful.
 We assessed model fit with graphical posterior predictive checks (Fig. A3).
-The model performed adequately at classifying the historical data (AUC = 0.77; Fig. A4).
-
-We used data from contemporary surveys of on endophyte prevalence  in \emph{A. hyemalis} and \emph{E. virginicus} as test data to evaluate predictions of the model. 
-During these surveys, which took place between 2013 and 2020, we collected seeds from living plants from 63 populations(average number of plants sampled: 22.9) across a portion of each species' range (Fig A4).
-We quantified the endophyte status with staining microscopy as described for the herbarium surveys and calculated prevalence of endophytes within the population (proportion of symbiotic plants divided by the number of sampled plants).
+The model performed adequately at classifying the historical data, comparing the accuracy of predictions from the model with observed data (AUC = 0.77; Fig. A4). 
+\subsection*{Validating the model with an out-of-sample test}
+We used data from contemporary surveys of endophyte prevalence  in \emph{A. hyemalis} and \emph{E. virginicus} as test data to evaluate predictions of the model. 
+Surveys of \emph{E. virginicus} were conducted in 2013 as described in \citet{sneck2017variation}, and surveys of \emph{A. hyemalis} took place between 2015 and 2020.
+Population surveys of \emph{A.
+hyemalis} were initially designed to cover longitudinal variation in endophyte prevalence towards its range edge, while surveys of \emph{E.
+virginicus} were designed to cover latitudinal variation along its range edge.
+In total, we visited 43 populations of \emph{A. hyemalis} and 20 populations of \emph{E. virginicus} across the central southeastern US, in Texas and neighboring states (Fig A4).
+During surveys, we collected seeds from up to 30 individuals per location  (average number of plants sampled: 22.9).
+We quantified the endophyte status of each individual with staining microscopy as described for the herbarium surveys, and calculated the prevalence of endophytes within the population (proportion of symbiotic plants divided by the number of sampled plants).
+The contempory survey period (2013-2020) is at the most recent edge of the time period encompassed by the historical observations used for model fitting.
 We compared the model's prediction for these locations to the observed population prevalence.
@@
 This time period connects the predicted change in prevalence with the endpoints of the available climate record.
-We then calculated the Pearson correlation coefficient between the relative change in endophyte prevalence and each climate driver.
+We then calculated the Spearman correlation coefficient between the relative change in endophyte prevalence and each climate driver.
+Calculating correlations from many pixels across the study region risks articially inflating confidence in our results due to large sample sizes, and so we repeated this calculation using only a random subsample of 100 pixels across the study region.
 
@@
 We quantified temporal and spatial trends in endophyte prevalence while accounting for potential biases introduced by collectors and by individuals who quantified endophyte presence/absence with the use of random effects.
-We found no evidence that collector biases influenced our results (random effects values centered around 0; Fig 2A), while the identity of individual scorers did contribute to observed patterns in endophyte prevalence (3 of the 16 scorers were more likely than average to assign positive endophyte status, as indicated by 95\% credible intervals that do not overlap 0) (Fig 2B).
+We found no evidence that collector biases influenced our results (collector random effects were consistently small; Fig 2A), while the identity of individual scorers did contribute to observed patterns in endophyte prevalence (3 of the 16 scorers were more likely than average to assign positive endophyte status, as indicated by 95\% credible intervals that do not overlap 0) (Fig 2B).
 Interpretation of the data without accounting for this source of variation would lead to potentially biased interpretations of temporal and spatial trends.
@@
 	\includegraphics[width = .8\linewidth]{random_fx_plot.png}
-	\caption{\textbf{Posterior estimates of (A) collector and (B) scorer random effects.} Points show tthe posterior mean along with 95\% CI for random effects estimate from 532 collectors and 16 scorers.}
+	\caption{\textbf{Posterior estimates of (A) collector and (B) scorer random effects.} Points show the posterior mean along with 95\% CI for random effects estimate from 532 collectors and 16 scorers.}
 \end{figure}
@@
 Across space, there were clear trends in endophyte prevalence which varied between species (Fig. 3).
-\emph{Elymus virginicus} had higher prevalence towards the northern portions of its range. 
-In contrast, \emph{Agrostis hyemalis} had highest prevalence towards the center of its range with regions of low prevalence towards the northeast and towards its western range edge.
-\emph{Agrostis perennans} had  highest prevalence towards the wester portion portion of it's range edge.
+We mapped predictions of endophyte prevalence across the area encompassing the sampled herbarium specimens for each species, and area smaller than the full geographic distribution of each species in nature.
+\emph{Elymus virginicus} had higher prevalence towards the northern portions of the collection range. 
+In contrast, \emph{Agrostis hyemalis} had highest prevalence towards the center of its collection range with regions of low prevalence towards the northeast and towards its western range edge.
+\emph{Agrostis perennans} had  highest prevalence towards the westernmost portion of its collection range.
 There is considerable uncertainty in the spatial pattern of endophyte prevalence (See Fig.
-A10 for projections of the 95\% credible interval), however these broad spatial patterns are consistent across the low and high probability model predictions.
+A10 for projections of the 95\% credible interval), however these broad spatial patterns are consistent across the low and high end of model predictions.
 
@@
 Our model revealed that while there was an overall increase in endophyte prevalence, these changes varied across the host species' ranges.
-Temporal trends vary with latitude and longitude, as evidenced by variation in the spatial pattern of endophyte prevalence through time in Fig.
-3, and that temporal trends in Fig.
-4 vary at different rates according to their latitude.
+The posterior estimates of the parameters describing space by time interactions, $\beta2$, $\beta3$, and $\beta4$ for each species indicate small effect sizes, but high certainty(i.e, a 99\% posterior probability of a negative interaction between year and latitude for \emph{A.
+hyemalis}, or a 67\% postterior probabilty of a postitive interaction between year and latitude for \emph{E.
+virginicus}) (Fig.
+A5).
 In some regions, \emph{A.
@@
 
-This reveals locations that are hotspots of change in endophyte prevalence (Fig. 5). 
-The overall magnitude of changes were greatest in \emph{A. perennans}, particularly in the northeast and in the south, where endophytes were previously only found in low prevalence (Fig. 3).
+This revealed locations that were hotspots of change in endophyte prevalence (Fig. 5). 
+The overall magnitude of changes were greatest in \emph{A. perennans}, particularly in the northeast and in the south, where endophytes were previously only found at low prevalence (Fig. 3).
 \emph{A.
 hyemalis} and \emph{E.
-virginicus} experience lower overall increases in prevalence, yet we can still identify hotspots of change in the northeast for \emph{A.
+virginicus} experience lower absolute increases in prevalence, yet we can still identify hotspots of change in the northeast for \emph{A.
 hyemalis} and in the southeast for \emph{E.
 virginicus}.
+
+\subsubsection*{Performance on test data}
+The model adequately predicted mean endophyte prevalence, but underpredicted variance in the contemporary test observations (Fig. 6 A).
+The model was able to predict large scale spatial trends in endophyte prevalence, such as declining endophyte prevalence towards western longitudes in \emph{A. hyemalis} (Fig. 6 B-C).
+We interpreted this to mean that the model captured regional spatial dynamics, but underpredicts local scale dynamics. We discuss potential model improvements in the Discussion.
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width = \linewidth]{contemp_test_plot.png}
+	\caption{\textbf{Predicted vs observed endophyte prevalence for contemporary test data.} (A) The model, trained on historic herbarium collection data, performed modestly at predicting contemporary endophyte prevalence, as indicated by some overlap of predicted 95\% CI with the 1:1 line, however contemporary test data generally had more variance between populations than model predictions.
+The model did recapitulate broader regional trends across (B) longitude and (C) latitude.
+Colors indicate host species, and point size in panels B and C reflect sample sizes of contemporary endophyte population surveys.}
+\end{figure}
 
@@
 We found that trends in endophyte prevalence were strongly associated with seasonal climate change drivers (Fig. 6).
-For the majority of the study region, the climate has become wetter and cooler over the last century (Fig. A5), a consequence of regional variation in global climate change \cite{ipcc_2021}. 
+For the majority of the study region, the climate has become wetter and cooler over the last century (Fig. A7-A8), a consequence of regional variation in global climate change \cite{ipcc_2021}. 
 Within the study region, spatially heterogeneous environmental changes were predictive of changes in endophyte prevalence. 
 For example, strong increases in prevalence within \emph{E.
-virginicus} were most associated (as indicated by correlation coefficients $>.3$) with declines in Summer precipitation  (a negative correlation in Fig.
-4) as well as with increases in the year-to-year variability of Summer precipitation (a positive correlation in Fig.
-4).
+virginicus} were most associated with declines in Summer precipitation  (a negative correlation in Fig.
+7) as well as with increases in the year-to-year variability of annual temperature (a positive correlation in Fig.
+7).
 Changes were also associated with reductions in average annual temperatures, and increases in year-to-year temperature variability.
@@
 For \emph{A. hyemalis}, endophyte prevalence increased most strongly in regions that experienced reductions in autumn precipitation variability. 
+Correlations using only a subsampling of pixels were qualitatively similar to these results (Fig.
+A11), suggesting that the patterns we find, particulary the strongest correlations are not spurious associations.
 
@@
 	\includegraphics[width = \linewidth]{climate_corr_heatmap.png}
-	\caption{\textbf{Correlations between changes in climate drivers and changes in endophyte prevalence.} Color denotes the Pearson correlation coefficient between the relative rate of change in endophyte prevalence and the change in annual mean temperature ($^oC$) and total annual and seasonal precipitation (mm), as well as the change in the coefficient of variation of each climate driver.
+	\caption{\textbf{Correlations between changes in climate drivers and changes in endophyte prevalence.} Color denotes the Spearman correlation coefficient between the relative rate of change in endophyte prevalence and the change in annual mean temperature ($^oC$) and total annual and seasonal precipitation (mm), as well as the change in the coefficient of variation of each climate driver.
 Positive correlation coefficients indicate that greater increases in a climate driver were associated with larger increases in endophyte prevalence, while negative values indicate that .
-Asteriks denote correlation coefficients $\> .3$ or $< -.3$.}
+Asteriks denote correlation coefficients $> .3$ or $< -.3$.}
 \end{figure}
@@
 For the three host species we examined, there have been clear increases in fungal endophyte prevalence over the last two centuries.
-This increase suggests that \emph{Epichloë} fungal endophytes are playing key \tom{adaptive}{In what way are they adaptive?} roles in response to \tom{climate change}{Climate change is not the only change over this time period, so be careful with attribution.} and that the mutualism between grasses and their fungal endophytes has contributed to resilience for each partner species under the observed changes in climate.
-Spatial and temporal trends in endophyte prevalence varied between each species in \tom{ecologically meaningful ways}{vague}.
-\tom{Our spatially-explicit model predicted regions of both high and low endophyte prevalence, suggesting that symbiotic and non-symbiotic host plants have overlapping, but non-identical niche requirements.}{This is a space-only result and interpretation, which should be better explained and motivated.}
-\tom{Endophytes may expand the range of their hosts by allowing them to persist in environments where they otherwise could not \citep{afkhami2014mutualist, kazenel2015mutualistic}.}{Not sure this is right for the first paragraph.}
-Understanding how microbial symbionts respond to climate change is a first step towards incorporating reciprocal host-symbiont interactions into predictions host range shifts.
-\tom{}{This paragraph should address the core question of whether climate change strengthens or disrupts mutualism.}
+Increases in prevalence of \emph{Epichloë}, which are vertically transmitted, can potentially be interpreted as adaptive changes that improve the fitness of their hosts under stressful conditions.
+This interpretation is in line with theory predicting that the positive fitness feedback caused by vertical transmission leads beneficial symbionts to rise in prevalence within a population \cite{fine1975vectors}.
+We found that trends in endophyte prevalence varied across the collection range of each species in assocation with observed changes in climate drivers, suggesting that the endophytes have contributed to host resilience under environmental change.
+Taken together, this suggests a strengthening of the mutualism over the last two centuries.
 
-\tom{For \emph{E.
-virginicus}, we found greater increases in prevalence in regions that have become drier during the Summer, the species' peak growing season.}{It is worth paying attention to both absolute and relative change.
-The AGPE results for example are notable because this is notable because this is the species with the greatest absolute change.}
-Fitness benefits of the symbiosis under drought conditions could explain this increase and also explain high predicted prevalence towards the species' northern range edge coinciding with a \tom{strong latitudinal decline in precipitation}{Show this?
-There is also a longitudinal cline in precip - how do you interpret that difference?}.
-Previous population surveys for endophytes in this species found similar latitudinal trends in prevalence \citep{sneck2017variation,rudgers2009benefits}, but at smaller scales. 
-Following previous work demonstrating drought benefits in a greenhouse manipulation with \emph{A.
-hyemalis} \citep{davitt2011understanding}, we had predicted  that endophyte prevalence should similarly increase at a greater rate in regions that have seen reduced precipitation, yet we found the opposite trend in our herbarium surveys and that changes in prevalence for this species had the weakest correlations with climate drivers.
-Given that the study region has experienced \tom{relatively moderate}{Not sure what to make of ``relatively moderate''.} changes in precipation and in temperature, potentially the magnitude of climate change has not been great enough to cause larger changes in endophyte prevalence for this species.
-Weak associations with drought could also be explained by climate-driven changes in the rate of imperfect transmission (the generation of non-symbiotic offspring from symbiotic hosts), which could counterbalance endophyte-mediated fitness benefits, and leading to stable intermediate prevalence rates \citep{donald2021context}.
+Differences between the responses of each host species underscore that while all of these $C_3$ grasses share similar broad-scale distributions, each engages in unique biotic interactions and has unique niche requirements.
+We identified hotspots of change in the northeast and south for \emph{A. perennans}, which experienced the strongest absolute changes in endophyte prevalence (Fig. 5). 
+Based on previous work demonstrating that endophytes can shield their hosts from drought stress \cite{decunta2021systematic}, we generally predicted that drought conditions could be a driver of increasing endophyte prevalence.
+In line with this expectation, increasing prevalence for this species was associated with decreasing precipitation, most strongly with spring-season declines (Fig. 7). 
+\emph{A. perennans} typically blooms in the autumn.
+Endophytes could be playing a role helping hosts weather spring-season droughts while the species is dormant. 
+It may be useful to investigate whether lagged climate effects are important predictors of host fitness in this system \cite{evers2021lagged}. 
 To our knowledge, the response of the symbiosis in \emph{A.
-perennans} to drought has not been examined experimentally, but in a separate greenhouse experiment, endophytes had a positive effect on reproduction under shaded, low-light conditions \citep{davitt2010costs}.
-\emph{Epichloë} endophytes have been connected to a suite of non-drought related fitness benefits including herbivore protection \citep{brem2001epichloe}, salinity resistence \citep{wang2020effects}, and mediation of the soil microbiome \citep{roberts2015rhizosphere}, potentially mediated by the diverse bioactive alkaloids and other signaling compounds they produce \citep{saikkonen2013chemical}.
-The strong increase in symbiotic \emph{A. perennans} could be explained by these diverse benefits. 
-\tom{Differences between the responses of each species underscore that while all of these C3 grasses share similar broad-scale distributions, each engages in unique biotic interactions and has unique niche requirements.}{I think this is an interesting point worth elaboratong on.
-On the one hand the three species are consistent in showing increases in prevalence, yet the mechanisms that explain those increases may be diverse and idiosyncratic.}
+perennans} to drought has not been examined experimentally, but in a greenhouse experiment, endophytes had a positive effect on host reproduction under shaded, low-light conditions \citep{davitt2010costs}.
+\emph{Epichloë} endophytes have been connected to a suite of non-drought related fitness benefits including herbivore protection \citep{brem2001epichloe}, salinity resistence \citep{wang2020effects}, and mediation of the soil microbiome \citep{roberts2015rhizosphere}
+These effects are potentially mediated by the diverse bioactive alkaloids and other signaling compounds they produce \citep{saikkonen2013chemical}.
+The strong increase in symbiotic \emph{A. perennans} could be explained, at least in part, by these diverse benefits. 
+\emph{A. hyemalis} experienced smaller overall changes in endophyte prevalence, and these changes were only weakly related to changes in climate drivers. 
+This result was surprising given previous work demonstrating drought benefits in a greenhouse manipulation with this species \citep{davitt2011understanding} that led us to expect  that endophyte prevalence should similarly increase at a greater rate in regions that have experienced increasing drought.
+It is possible that the study region has not experienced a magnitude of climate change great enough to cause larger changes in endophyte prevalence for this species. 
+Weak associations with drought could also be explained by climate-driven changes in the rate of imperfect transmission (the generation of non-symbiotic offspring from symbiotic hosts), which could counterbalance endophyte-mediated fitness benefits \citep{donald2021context}.
+For \emph{E.
+virginicus}, which also experienced more modest changes in endophte prevalence overall, we found a strong relationship between temporal trends and  changes in the mean and variability of temperature, as well as with decreases in summer precipitation.
+Surveys by \citet{sneck2017variation}, used as part of the test data in this study, identified a drought index (SPEI) that integegrates precipitation with estimated evapotranspiration as an important predictor of endophyte prevalence.
+While we show consistent increasing trends in prevalence between the three species, the mechanisms that explain these changes may be diverse and idiosyncratic.
+
+Our spatially-explicit model predicted regions of both high and low endophyte prevalence, suggesting that symbiotic and non-symbiotic host plants have overlapping, but non-identical niche requirements.
+Endophytes fitness benefits potentially explain the spatial distribution of prevalence by allowing their hosts to persist in environments where they otherwise could not \citep{afkhami2014mutualist, kazenel2015mutualistic}.
+For example, fitness benefits of the symbiosis could explain high predicted prevalence in \emph{E.
+virginicus} towards the north or in \emph{A.
+hyemalis} towards its range center coinciding with strong environmental gradients.
+Previous population surveys for endophytes, which were used as test data for our model, found similar latitudinal trends in prevalence in these species \citep{sneck2017variation,rudgers2009benefits}, but at smaller scales.
+While the model recreated these large-scale spatial trends, test data was more variable. 
+Using test data to validate our model predictions allows us to evaluate places to improve the model's ability to perform well at out-of-sample prediction, which will be particularly important for predicting host and symbiont niche-shifts under future climate change.
+Lack of information on local variability may simply be a feature of data derived from herbarium specimens. 
+Even though they are samples from local populations, that are single specimens that are aggregated over in broad-scale model estimates.
+Poor predictive ability at local scales in this grass-endophyte system is not surprising, as previous studies have found that local variation, even to the scale of hundreds of meters can structure endophyte-host niches \cite{kazenel2015mutualistic}.
+\citet{sneck2017variation} also identified host genotype as an important predictor of endophyte prevalence in \emph{E. virginicus}.
+Other studies have found factors including land-use history \cite{vikuk2019infection} and the biotic environment, including herbivory \cite{rudgers2016long}, to be important predictors of endophyte ecology.
+Incorporating available climatic and soil layers as covariates is an obvious first step that could improve predictions.
+Towards the goal of predicting the microbial symbioses dynamics under climate change, models that integrate data from local and regional scales would be an important step to bridge the gap that often exists between large but broad bioclimatic and biodiversity data and small but local data on biotic interactions.
+\cite{miller2019recent, isaac2020data}
 
@@
 Answering these questions as well as the unknown questions that future researchers may ask also reiterates the value in capturing meta-information during ongoing digitization efforts at herbaria around the world and during the accession of newly collected specimens \citep{lendemer2020extended}.
-\tom{Second, we accounted for several potential biases in the data observation process that may be common to many collections-based research questions by using a spatially-explicit random effects model.
-}{I think you should say more about collector and observer effects in the results.}
+Second, we accounted for several potential biases in the data observation process that may be common to many collections-based research questions by using a spatially-explicit random effects model. 
 Spatial autocorrelation \cite{willems2022forest}, potential biases introduced by the sampling habits of collectors \citep{daru2018widespread}, and variation between contemporary researchers during the collection of trait data, if not corrected for could lead to biased inference about the strength and direction of historic change.
-Fitting this model in a Bayesian framework allows for full propagation of uncertainty.\tom{}{Excellent paragraph.}
+Previous studies that have quantified the effects of collector biases typically find them to be small  \cite{davis2015herbarium,meineke2019herbarium}, and we similarly did not find that collector has a strong effect on the results of our analysis.
+Fitting this model in a Bayesian framework allows for full propagation of uncertainty.
 
@@
 For example, transplanting symbiotic and non-symbiotic plants beyond the range edge of \emph{A.
-hyemalis} could tell us whether persistent lack of endophytes in that area is a result of environmental conditions that lead the symbiosis to negative fitness consequences, or is a result of some historical contingency or dispersal limitation that has thus far limited the presence of symbiotic hosts from a region where they would otherwise flourish and provide resilience.
+hyemalis} could tell us whether persistent low endophyte prevalence in that area is a result of environmental conditions that lead the symbiosis to negative fitness consequences, or is a result of some historical contingency or dispersal limitation that has thus far limited the presence of symbiotic hosts from a region where they would otherwise flourish and provide resilience.
 While we observed evidence of mutualism resilience, more extreme environmental changes than those observed in our study could potentially push one or both partners beyond their physiological limit, leading to the collapse of the mutualism.
@@
 	We thank Jessica Budke for help in drafting our initial destructive sampling plan, and to the many members of herbarium staff who facilitated our research visits, as well as to the hundreds of collectors who contributed to the natural history collections.
-	This research was supported by funding from NSF grant () and from funding from the Texas Ecolab Program.
+	Several high schooler and undergraduate researchers contributed to data collection, including A.
+Appio-Riley, P.
+Bilderback, E.
+Chong, K.
+Dickens, L.
+Dufresne, B.
+Gutierrez, A.
+Johnson, S.
+Linder, E.
+Scales, B.
+Scherick, K.
+Schrader, E.
+Segal , G.
+Singla, and M.
+Tucker.
+	This research was supported by funding from National Science Foundation (grants 1754468 and 2208857) and by funding from the Texas Ecolab Program.
 
@@
 	On initial submission, you may use this section to provide a URL for editors and reviewers that is `private for peer review'.
-After acceptance, this section must be updated with correct, working DOIs for data deposits (typically on the Dryad Digital Repository, \citealt{CookEtAl2015}) and code deposits (such as in Zenodo).
+After acceptance, this section must be updated with correct, working DOIs for data deposits (typically on the Dryad Digital Repository, ) and code deposits (such as in Zenodo).
 
-	\section*{Appendix A: Additional Methods and Parameters}
+	\section*{Appendix A}
 	\renewcommand{\thefigure}{A\arabic{figure}}
@@
 	\centering
-	\includegraphics[width = \linewidth]{contemp_test_plot.png}
-	\caption{\textbf{Predicted vs observed endophyte prevalence for contemporary test data.} (A) The model, trained on historic herbarium collection data, performed modestly at predicting contemporary endophyte prevalence, as indicated by some overlap of predicted 95\% CI with the 1:1 line, however contemporary test data generally had more variance between populations than model predictions.
-The model did recapitulate broader regional trends across (B) longitude and (C) latitude.
-Colors indicate host species, and point size in panels B and C reflect sample sizes of contemporary endophyte population surveys.}
-\end{figure}
-
-	\begin{figure}[H]
-	\centering
 	\includegraphics[width = .8\linewidth]{ppt_change_maps.png}
@@
 Panels show upper and lower 95\% posterior probability for each host species between 1925 and 2020.}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width = \linewidth]{climate_corr_heatmap_subsample.png}
+	\caption{\textbf{Correlations between changes in climate drivers and changes in endophyte prevalence from a random sample of 100 pixels across the study region.} Color denotes the Spearman correlation coefficient between the relative rate of change in endophyte prevalence and the change in annual mean temperature ($^oC$) and total annual and seasonal precipitation (mm), as well as the change in the coefficient of variation of each climate driver.
+Positive correlation coefficients indicate that greater increases in a climate driver were associated with larger increases in endophyte prevalence, while negative values indicate that .
+Asteriks denote correlation coefficients $> .3$ or $< -.3$.}
 \end{figure}
@@
 	% Dryad references.
-\bibliographystyle{plain}
+	\newpage{}
+\bibliographystyle{plainnat}
 \bibliography{endo_herbarium}

### 2023-03-23 endo_herbarium_spatial_analysis.R

@@ -793,13 +793,14 @@ ggplot(data = pred.test_data)+
 test_data_map <- ggplot()+
   geom_map(data = outline_map, map = outline_map, aes(map_id = region), color = "grey", linewidth = .1, fill = "#FAF9F6")+
   geom_map(data = states_shape, map = states_shape, aes(map_id = region), color = "grey", linewidth = .1, fill = NA)+
-  geom_point(data = endo_herb, aes(x = lon, y = lat), color = "black",size = 1, pch = 3)+
-  geom_point(data = contemp_surveys, aes(x = lon, y = lat), color = "red", size = 1)+
-  coord_sf(xlim = c(-109,-68), ylim = c(21,49))+
+  geom_point(data = endo_herb, aes(x = lon, y = lat), color = "grey40",size = 1, pch = 3, alpha = .8)+
+  geom_point(data = contemp_surveys, aes(x = lon, y = lat, pch = SpeciesID), color = "red", size = 2, alpha = .8)+
+  coord_sf(xlim = c(-104,-80), ylim = c(25,49))+
+  scale_shape_manual(values = c(19,17))+
   theme_light()+
   theme(strip.background = element_blank(),
         strip.text  = element_text(face = "italic", color = "black"))+
-  labs(x = "Longitude", y = "Latitude", color = "Endophyte Status")
+  labs(x = "Longitude", y = "Latitude", pch = "Species")
 test_data_map
 
 ggsave(test_data_map, filename = "test_data_map.png", width = 10, height = 8)
@@ -821,7 +822,9 @@ endo_herb_binned <- endo_herb %>%
             mean_lat = mean(lat),
             sample = n()) %>% 
   mutate(lat_bin = case_when(mean_lat>=35 ~ paste("43"),
-                             mean_lat<35 ~ paste("35")))
+                             mean_lat<35 ~ paste("35")),
+         lon_bin = case_when(mean_lon<=-94 ~ paste("-90"),
+                             mean_lon>-94 ~ paste("-80") ))
   
 # mean_loc <- endo_herb_binned %>% 
 #   group_by(Spp_code, species) %>% 
@@ -1138,28 +1141,30 @@ tmean_change_maps <- tmean_annual_mean_map + tmean_annual_cv_map +
 ggsave(tmean_change_maps, filename = "tmean_change_maps.png", width = 15, height = 6)
 
[…]

### 2023-12-08 Epichloe Maxent.Rmd
adding sdm code for the host species
I added the raster of the probability of presence of the host species

---
title: 'SDM for Epichloe endophyte host species'
author: "Jacob Moutouama "
date: "`r Sys.Date()`"
output: 
  rmdformats::readthedown:
    code_folding: hide
    self_contained: true
    number_sections: true
    thumbnails: true
    lightbox: true
    gallery: true
    keep_md: true
    highlight: tango
    df_print: kable 
    toc_depth: 3
    fig_width: 8
    fig_height: 8
editor_options: 
  chunk_output_type: console
---

```{r setup,cache=TRUE,include=FALSE}
library(knitr)
opts_chunk$set(external=TRUE,echo=F,warning=FALSE,fig.pos='H',autodep = TRUE, global.device = TRUE)
a4width<- 8.3
a4height<- 11.7
options(knitr.table.format = "latex")
```

```{r eval=FALSE,echo=T}
rm(list = ls())
```

```{r}
# Load packages
library("prism")
library("raster")
library("dismo")
library("usdm")
library("car")
library("FactoMineR")
library("factoextra")
library("corrplot")
library("ENMTools")
library("vip")
library("pdp")
library("fastshap")
library("CalibratR")
library("maptools")
library("rgeos")
library("leaflet")
library("tidyverse")
# install.extras()
library("geodata")
library("terra")
library("rgdal")
library("sp")
library("ENMeval")
library("mecofun")
library("mgcv")
library("randomForest")
library("spThin")
# if( !("rJava" %in% rownames(installed.packages()))  ){
#   install.packages("rJava",repos="http://cran.r-project.org")
# }

# if(Sys.info()["sysname"] != "Windows" ){
#   dyn.load('/Library/Java/JavaVirtualMachines/temurin-17.jdk/Contents/Home/lib/server/libjvm.dylib')
# }
library("rJava")
# if( !file.exists(paste0(system.file("java", package="dismo"),"/maxent.jar"))  )   {
# utils::download.file(url="https://raw.githubusercontent.com/mrmaxent/Maxent/master/ArchivedReleases/3.3.3k/maxent.jar",
#                      destfile=paste0(system.file("java", package="dismo"),"/maxent.jar"),
#                      mode="wb") ## wb for binary file, otherwise maxent.jar can not execute
# }

```

# Climatic data 
## Data from PRISM
```{r}
# making a folder to store prism data
prism_set_dl_dir("/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Clim")

# getting monthly data for mean temp and precipitation
# takes a long time the first time, but can skip when you have raster files saved on your computer.
# get_prism_monthlys(type = "tmean", years = 1895:2020, mon = 1:12, keepZip = FALSE)
# get_prism_monthlys(type = "ppt", years = 1895:2020, mon = 1:12, keepZip = FALSE)
```

```{r}
# pulling out values to get normals for old and new time periods
tmean_annual_recent_norm <- terra::mean(terra::rast(pd_stack(prism_archive_subset(type = "tmean", temp_period = "monthly", year = 1990:2020))))
tmean_spring_recent_norm <- terra::mean(terra::rast(pd_stack(prism_archive_subset(type = "tmean", temp_period = "monthly", year = 1990:2020, mon = 1:4))))
tmean_summer_recent_norm <- terra::mean(terra::rast(pd_stack(prism_archive_subset(type = "tmean", temp_period = "monthly", year = 1990:2020, mon = 5:8))))
tmean_autumn_recent_norm <- terra::mean(terra::rast(pd_stack(prism_archive_subset(type = "tmean", temp_period = "monthly", year = 1990:2020, mon = 9:12))))

tmean_annual_old_norm <- terra::mean(terra::rast(pd_stack(prism_archive_subset(type = "tmean", temp_period = "monthly", year = 1895:1925))))
tmean_spring_old_norm <- terra::mean(terra::rast(pd_stack(prism_archive_subset(type = "tmean", temp_period = "monthly", year = 1895:1925, mon = 1:4))))
tmean_summer_old_norm <- terra::mean(terra::rast(pd_stack(prism_archive_subset(type = "tmean", temp_period = "monthly", year = 1895:1925, mon = 5:8))))
tmean_autumn_old_norm <- terra::mean(terra::rast(pd_stack(prism_archive_subset(type = "tmean", temp_period = "monthly", year = 1895:1925, mon = 9:12))))

```

```{r}
# calculating the cumulative precipitation for each year and for each season within the year
ppt_annual_recent <- ppt_spring_recent <- ppt_summer_recent <- ppt_autumn_recent <- ppt_winter_recent<- list()
for(y in 1990:2020){
ppt_annual_recent[[y]] <- sum(terra::rast(pd_stack(prism_archive_subset(type = "ppt", temp_period = "monthly", year = y))))
ppt_spring_recent[[y]] <- sum(terra::rast(pd_stack(prism_archive_subset(type = "ppt", temp_period = "monthly", year = y, mon = 1:4))))
ppt_summer_recent[[y]] <- sum(terra::rast(pd_stack(prism_archive_subset(type = "ppt", temp_period = "monthly", year = y, mon = 5:8))))
ppt_autumn_recent[[y]] <- sum(terra::rast(pd_stack(prism_archive_subset(type = "ppt", temp_period = "monthly", year = y, mon = 9:12)))) 
}

ppt_annual_old <- ppt_spring_old <- ppt_summer_old <- ppt_autumn_old <- ppt_winter_old<- list()
for(y in 1895:1925){
  ppt_annual_old[[y]] <- sum(terra::rast(pd_stack(prism_archive_subset(type = "ppt", temp_period = "monthly", year = y))))
  ppt_spring_old[[y]] <- sum(terra::rast(pd_stack(prism_archive_subset(type = "ppt", temp_period = "monthly", year = y, mon = 1:4))))
  ppt_summer_old[[y]] <- sum(terra::rast(pd_stack(prism_archive_subset(type = "ppt", temp_period = "monthly", year = y, mon = 5:8))))
  ppt_autumn_old[[y]] <- sum(terra::rast(pd_stack(prism_archive_subset(type = "ppt", temp_period = "monthly", year = y, mon = 9:12)))) # including December here because the year needs to wrap around...
}

# Taking the mean of the cumulative precipation values
ppt_annual_recent_norm <- terra::mean(terra::rast(unlist(ppt_annual_recent)))
ppt_spring_recent_norm <- terra::mean(terra::rast(unlist(ppt_spring_recent)))
ppt_summer_recent_norm <- terra::mean(terra::rast(unlist(ppt_summer_recent)))
ppt_autumn_recent_norm <- terra::mean(terra::rast(unlist(ppt_autumn_recent)))

ppt_annual_old_norm <- terra::mean(terra::rast(unlist(ppt_annual_old)))
ppt_spring_old_norm <- terra::mean(terra::rast(unlist(ppt_spring_old)))
ppt_summer_old_norm <- terra::mean(terra::rast(unlist(ppt_summer_old)))
ppt_autumn_old_norm <- terra::mean(terra::rast(unlist(ppt_autumn_old)))
```

## Variance inflation factor (VIF)  to have  a measure of multicollinearity among the  variables
```{r}
US_worldclim_recent_norm<-terra::rast(list(tmean_spring_recent_norm,tmean_summer_recent_norm,tmean_autumn_recent_norm,ppt_spring_recent_norm,ppt_summer_recent_norm,ppt_autumn_recent_norm))
US_worldclim_recent_norm_stack<-stack(US_worldclim_recent_norm)
# plot(US_worldclim_recent_norm_stack)
```

## Variance inflation factor (VIF)  to have  a measure of multicollinearity among the  variables

$$
VIF =(\frac{1}{1-R^{2}})
$$
```{r}
# (var_vif<-usdm::vifstep(US_worldclim_recent_norm_stack, th=10))
(vif <- vifcor(US_worldclim_recent_norm_stack, th=0.7))

```

## Saved the rasters 
### Recent
```{r}
writeRaster(tmean_spring_recent_norm, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Final variable/recent/tmean_spring_recent_norm.tif')
writeRaster(ppt_spring_recent_norm, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Final variable/recent/ppt_spring_recent_norm.tif')
writeRaster(ppt_summer_recent_norm, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Final variable/recent/ppt_summer_recent_norm.tif')
```
### old 
```{r}
writeRaster(tmean_spring_old_norm, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Final variable/old/tmean_spring_old_norm.tif')
writeRaster(ppt_spring_old_norm, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Final variable/old/ppt_spring_old_norm.tif')
writeRaster(ppt_summer_old_norm, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Final variable/old/ppt_summer_old_norm.tif')
```

## Final climatic variables to use for the SDM

```{r}
US_worldclim_recent_norm_stack_vif<-terra::rast(list(tmean_spring_recent_norm,ppt_spring_recent_norm,ppt_summer_recent_norm))
names(US_worldclim_recent_norm_stack_vif) <- c("tmean_spring", "ppt_spring","ppt_summer")
US_worldclim_recent_norm_stack_final<-stack(US_worldclim_recent_norm_stack_vif)

US_worldclim_old_norm_stack_vif<-terra::rast(list(tmean_spring_old_norm,ppt_spring_old_norm,ppt_summer_old_norm))
names(US_worldclim_old_norm_stack_vif) <- c("tmean_spring", "ppt_spring","ppt_summer")
US_worldclim_old_norm_stack_final<-stack(US_worldclim_old_norm_stack_vif)

```

```{r}
plot(US_worldclim_recent_norm_stack_final)
```

```{r}
plot(US_worldclim_old_norm_stack_final)
```

# *Agrostis hyemalis*
## Import occurence data from Josh database and merge them later to existing online data

```{r}
endo_herb_georef<-read.csv("/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Occurence/endo_herb_georef.csv", header=T)
```

## Download occurrence data from gbif for *Agrostis hyemalis*

```{r} 
# dir.create("/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Occurence")
# aghy_occ_raw <- gbif(genus="Agrostis",species="hyemalis",download=TRUE) 
load("/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Occurence/aghy_occ_raw.rdata")
head(aghy_occ_raw) # to view the first few records the occurrence dataset use
```

## Clean occurrence data from gbif

```{r}
aghy_occ <- subset(aghy_occ_raw,(!is.na(lat))&(!is.na(lon))) # here we remove erroneous coordinates, where either the latitude or longitude is missing
cat(nrow(aghy_occ_raw)-nrow(aghy_occ), "records are removed") # Show the number of records that are removed from the dataset. 
aghy_occ %>% 
  dplyr::select(country,lon, lat, year,basisOfRecord)%>% 
  filter(country=="United States")->aghy_occ_clean # We want only occurrence data for the US

```

```{r}
dups_aghy <- duplicated(aghy_occ_clean[c("lat","lon")]  )
aghy_occ_unique <- aghy_occ_clean[!dups_aghy,] # Remove duplicated data based on latitude and longitude
cat(nrow(aghy_occ_clean)-nrow(aghy_occ_unique), "records are removed")
```

```{r}
table(aghy_occ_unique$basisOfRecord) # show the frequency table of “basisOfRecord”
```

```{r}
hist(aghy_occ_unique$year,main="",xlim=c(1895,2020),ylim=c(0,500),col="orange2",xlab="years")
```

```{r}
aghy_occ_unique %>% 
  dplyr::select(lon, lat, year)%>% 
  filter(year %in% (1990:2020) & as.numeric(lon >=-126.374160))->aghy_occ_1990_2020 #  filter from 1990 to 2020 to match climatic data.

aghy_occ_unique  %>% 
  dplyr::select(lon, lat, year)%>% 
  filter(year %in% (1895:1925) & as.numeric(lon >=-126.374160))->aghy_occ_1895_1925 #  filter from 1895 to 1925 to match climatic data.

summary(aghy_occ_1990_2020$year) # show a quick summary of years in the data
summary(aghy_occ_1895_1925$year) # show a quick summary of years in the data
```

## Merge data from Josh database to GBIF  data

```{r}
endo_herb_georef %>% 
  dplyr::select(lon, lat, year,Spp_code)%>% 
  filter(Spp_code=="AGHY" & year %in% (1990:2020))->aghy_occgeoref_1990_2020

endo_herb_georef %>% 
  dplyr::select(lon, lat, year,Spp_code)%>% 
  filter(Spp_code=="AGHY" & year %in% (1895:1925))->aghy_occgeoref_1895_1925

aghy_occ_final_1990_2020<-rbind(aghy_occ_1990_2020[,-3],aghy_occgeoref_1990_2020[,-c(3,4)])
aghy_occ_final_1895_1925<-rbind(aghy_occ_1895_1925[,-3],aghy_occgeoref_1895_1925[,-c(3,4)])
dupfinal1990_2020 <- duplicated(aghy_occ_final_1990_2020[,c(1,2)])
dupfinal1895_1925 <- duplicated(aghy_occ_final_1895_1925[,c(1,2)])

aghy_occ_unique_final_1990_2020 <- aghy_occ_final_1990_2020[!dupfinal1990_2020,]
aghy_occ_unique_final_1895_1925<-aghy_occ_final_1895_1925[!dupfinal1895_1925,]

```

## Spatial thinning of species occurence records

```{r,eval=FALSE}
ID<-rep("AGHY",nrow(aghy_occ_unique_final_1990_2020))
aghy_sdm_final_1990_2020<-data.frame(ID,aghy_occ_unique_final_1990_2020)
names(aghy_sdm_final_1990_2020)<-c("ID","Long","Lat")
thinned.aghy_recent <- spThin::thin(aghy_sdm_final_1990_2020,lat.col = "Lat",long.col = "Long",spec.col = "ID",
                         thin.par=9, reps=10, locs.thinned.list.return = FALSE,
                         write.files = TRUE, max.files = 5, out.dir="/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Occurence", 
                         out.base = "points_aghy_thinned_recent", write.log.file = TRUE,
                         log.file = "spatial_thin_aghy_recent_log.txt", verbose = TRUE)

```

```{r}
aghy_thinned_recent<-read.csv("/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Occurence/aghy_thinned_recent.csv", header=T)
aghy_thinned<-aghy_thinned_recent[,-1]
```

## Projection on a map to double check if there is any point out of the US

```{r}
library(maptools)
data(wrld_simpl)
plot(US_worldclim_recent_norm_stack_final,1)
plot(wrld_simpl, add=TRUE)
points(aghy_thinned, col='#0072B2',pch=19)
```

## Maxent model for *Agrostis hyemalis*

```{r}
set.seed(13)
group_pres_aghy <- kfold(aghy_thinned, 5)
pres_train_aghy <- aghy_thinned[group_pres_aghy != 1, ] # training data 
pres_test_aghy <- aghy_thinned[group_pres_aghy == 1, ] # testing data
```

```{r}
set.seed(13)
backg_aghy <- randomPoints(US_worldclim_recent_norm_stack_final, n=10000)
colnames(backg_aghy) = c('Long', 'Lat')
group_backg_aghy <- kfold(backg_aghy, 5)
backg_train_aghy <- backg_aghy[group_backg_aghy != 1, ]
backg_test_aghy <- backg_aghy[group_backg_aghy == 1, ]
```

```{r}
r <- raster(US_worldclim_recent_norm_stack_final, 1)
plot(!is.na(r), col=c('white', 'light grey'), legend=FALSE)
points(backg_train_aghy, pch='-', cex=0.5, col='#F0E442')
points(backg_test_aghy, pch='-',  cex=0.5, col='#000000')
points(pres_train_aghy, pch= '+', col='#009E73')
points(pres_test_aghy, pch='+', col='#0072B2')
```

```{r}
mod_aghy <- maxent(US_worldclim_recent_norm_stack_final, pres_train_aghy)
# plot(mod_aghy)
```
## Save the model output 
```{r}
e_aghy <- evaluate(pres_test_aghy, backg_test_aghy, mod_aghy, US_worldclim_recent_norm_stack_final)
map_recent_aghy <- predict(US_worldclim_recent_norm_stack_final, mod_aghy,  progress='')
map_old_aghy <- predict(US_worldclim_old_norm_stack_final, mod_aghy,  progress='')
writeRaster(map_recent_aghy, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Model/raster_recent_aghy.tif')
writeRaster(map_old_aghy, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Model/raster_old_aghy.tif')
```

```{r}
par(mfrow=c(2,2))
plot(map_recent_aghy, main='2020')
plot(wrld_simpl, add=TRUE, border='dark grey')
plot(map_old_aghy, main='1925')
plot(wrld_simpl, add=TRUE, border='dark grey')
```

```{r}
tr_aghy <- threshold(e_aghy, 'spec_sens')
plot(map_recent_aghy > tr_aghy, main='2020')
plot(wrld_simpl, add=TRUE, border='dark grey')
# points(pres_train_aghy, pch='+')

plot(map_old_aghy > tr_aghy, main='1925')
plot(wrld_simpl, add=TRUE, border='dark grey')
# points(pres_train_aghy, pch='+')
dev.off()
```

```{r}
map_recent_aghy_presence_absence<- (map_recent_aghy > tr_aghy)
writeRaster(map_recent_aghy_presence_absence, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Model/raster_recent_aghy_presence_absence.tif')
map_old_aghy_presence_absence<- (map_old_aghy > tr_aghy)
writeRaster(map_old_aghy_presence_absence, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Model/raster_old_aghy_presence_absence.tif')
```

## Download occurrence data from gbif for *Agrostis perennans*

```{r} 
# agpe_occ_raw <- gbif(genus="Agrostis",species="perennans",download=TRUE) 
load("/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Occurence/agpe_occ_raw.rdata")
head(agpe_occ_raw ) # to view the first few records the occurrence dataset use:
```

## Clean occurrence data from gbif

```{r}
agpe_occ <- subset(agpe_occ_raw,(!is.na(lat))&(!is.na(lon))) # here we remove erroneous coordinates, where either the latitude or longitude is missing
cat(nrow(agpe_occ_raw)-nrow(agpe_occ), "records are removed") #Show the number of records that are removed from the dataset. 
agpe_occ %>% 
  dplyr::select(country,lon, lat, year,basisOfRecord)%>% 
  filter(country=="United States")->agpe_occ_clean

```

```{r}
dups_agpe <- duplicated(agpe_occ_clean[c("lat","lon")]  )
agpe_occ_unique <- agpe_occ_clean[!dups_agpe,] # Remove duplicated data based on latitude and longitude
cat(nrow(agpe_occ_clean)-nrow(agpe_occ_unique), "records are removed")
```

```{r}
table(agpe_occ_unique$basisOfRecord) # show the frequency table of “basisOfRecord”
```

```{r}
hist(agpe_occ_unique$year,main="",xlim=c(1895,2020),ylim=c(0,500),col="orange2",xlab="years")
```

```{r}
agpe_occ_unique %>% 
  dplyr::select(lon, lat, year)%>% 
  filter(year %in% (1990:2020) & as.numeric(lon >=-126.374160))->agpe_occ_1990_2020 #  filter from 1990 to 2020 to match climatic data.

agpe_occ_unique  %>% 
  dplyr::select(lon, lat, year)%>% 
  filter(year %in% (1895:1925) & as.numeric(lon >=-126.374160))->agpe_occ_1895_1925

summary(agpe_occ_1990_2020$year) # show a quick summary of years in the data
summary(agpe_occ_1895_1925$year) # show a quick summary of years in the data

```

## Merge endophyte host data from Josh database to GBIF  data

```{r}
endo_herb_georef %>% 
  dplyr::select(lon, lat, year,Spp_code)%>% 
  filter(Spp_code=="AGPE" & year %in% (1990:2020))->agpe_occgeoref_1990_2020

endo_herb_georef %>% 
  dplyr::select(lon, lat, year,Spp_code)%>% 
  filter(Spp_code=="AGPE" & year %in% (1895:1925))->agpe_occgeoref_1895_1925

agpe_occ_final_1990_2020<-rbind(agpe_occ_1990_2020[,-3],agpe_occgeoref_1990_2020[,-c(3,4)])
agpe_occ_final_1895_1925<-rbind(agpe_occ_1895_1925[,-3],agpe_occgeoref_1895_1925[,-c(3,4)])
dupfinal1990_2020 <- duplicated(agpe_occ_final_1990_2020[,c(1,2)])
dupfinal1895_1925 <- duplicated(agpe_occ_final_1895_1925[,c(1,2)])

agpe_occ_unique_final_1990_2020 <- agpe_occ_final_1990_2020[!dupfinal1990_2020,]
agpe_occ_unique_final_1895_1925<-agpe_occ_final_1895_1925[!dupfinal1895_1925,]

```

## Spatial thinning of species occurence records

```{r,eval=FALSE}
ID<-rep("AGPE",nrow(agpe_occ_unique_final_1990_2020))
agpe_sdm_final_1990_2020<-data.frame(ID,agpe_occ_unique_final_1990_2020)
names(agpe_sdm_final_1990_2020)<-c("ID","Long","Lat")
thinned.agpe_recent <- spThin::thin(agpe_sdm_final_1990_2020,lat.col = "Lat",long.col = "Long",spec.col = "ID",
                         thin.par=9, reps=10, locs.thinned.list.return = FALSE,
                         write.files = TRUE, max.files = 5, out.dir="/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Occurence", 
                         out.base = "points_agpe_thinned_recent", write.log.file = TRUE,
                         log.file = "spatial_thin_agpe_recent_log.txt", verbose = TRUE)

```

```{r}
agpe_thinned_recent<-read.csv("/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Occurence/agpe_thinned_recent.csv", header=T)
agpe_thinned<-agpe_thinned_recent[,-1]
```

## Projection on a map to double check if there is any point of the US

```{r}
library(maptools)
data(wrld_simpl)
plot(US_worldclim_recent_norm_stack_final,1)
plot(wrld_simpl, add=TRUE)
points(agpe_thinned, col='#0072B2',pch=19)
```

## Maxent model for *Agrostis perennans*

```{r}
set.seed(13)
group_pres_agpe <- kfold(agpe_thinned, 5)
pres_train_agpe <- agpe_thinned[group_pres_agpe != 1, ] # training data 
pres_test_agpe <- agpe_thinned[group_pres_agpe == 1, ] # testing data
```

```{r}
set.seed(13)
backg_agpe <- randomPoints(US_worldclim_recent_norm_stack_final, n=10000)
colnames(backg_agpe) = c('Long', 'Lat')
group_backg_agpe <- kfold(backg_agpe, 5)
backg_train_agpe <- backg_agpe[group_backg_agpe != 1, ]
backg_test_agpe <- backg_agpe[group_backg_agpe == 1, ]
```

```{r}
r <- raster(US_worldclim_recent_norm_stack_final, 1)
plot(!is.na(r), col=c('white', 'light grey'), legend=FALSE)
points(backg_train_agpe, pch='-', cex=0.5, col='#F0E442')
points(backg_test_agpe, pch='-',  cex=0.5, col='#000000')
points(pres_train_agpe, pch= '+', col='#009E73')
points(pres_test_agpe, pch='+', col='#0072B2')
```

```{r}
mod_agpe <- maxent(US_worldclim_recent_norm_stack_final, pres_train_agpe)
# plot(mod_aghy)
```

```{r}
e_agpe <- evaluate(pres_test_agpe, backg_test_agpe, mod_agpe, US_worldclim_recent_norm_stack_final)
map_recent_agpe <- predict(US_worldclim_recent_norm_stack_final, mod_agpe,  progress='')
map_old_agpe <- predict(US_worldclim_old_norm_stack_final, mod_agpe,  progress='')
writeRaster(map_recent_agpe, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Model/raster_recent_agpe.tif')
writeRaster(map_old_agpe, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Model/raster_old_agpe.tif')

```

```{r}
par(mfrow=c(2,2))
plot(map_recent_agpe, main='2020')
plot(wrld_simpl, add=TRUE, border='dark grey')
plot(map_old_agpe, main='1925')
plot(wrld_simpl, add=TRUE, border='dark grey')

```

```{r}
tr_agpe <- threshold(e_agpe, 'spec_sens')
plot(map_recent_agpe > tr_agpe, main='2020')
plot(wrld_simpl, add=TRUE, border='dark grey')
# points(pres_train_aghy, pch='+')

plot(map_old_agpe > tr_agpe, main='1925')
plot(wrld_simpl, add=TRUE, border='dark grey')
# points(pres_train_aghy, pch='+')
dev.off()
```

```{r}
map_recent_agpe_presence_absence<- (map_recent_agpe > tr_agpe)
writeRaster(map_recent_agpe_presence_absence, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Model/raster_recent_agpe_presence_absence.tif')

map_old_agpe_presence_absence<- (map_old_agpe > tr_agpe)
writeRaster(map_old_agpe_presence_absence, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Model/raster_old_agpe_presence_absence.tif')
```

## Download occurrence data from gbif for *Elymus virginicus*

```{r} 
# elvi_occ_raw <- gbif(genus="Elymus",species="virginicus",download=TRUE) 
load("/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Occurence/elvi_occ_raw.rdata")
head(elvi_occ_raw)
```

## Clean occurrence data from gbif

```{r}
elvi_occ <- subset(elvi_occ_raw,(!is.na(lat))&(!is.na(lon))) # here we remove erroneous coordinates, where either the latitude or longitude is missing
cat(nrow(elvi_occ_raw)-nrow(elvi_occ), "records are removed") #Show the number of records that are removed from the dataset. 
elvi_occ %>% 
  dplyr::select(country,lon, lat, year,basisOfRecord)%>% 
  filter(country=="United States")->elvi_occ_clean

```

```{r}
dups_elvi <- duplicated(elvi_occ_clean[c("lat","lon")]  )
elvi_occ_unique <- elvi_occ_clean[!dups_elvi,] # Remove duplicated data based on latitude and longitude
cat(nrow(elvi_occ_clean)-nrow(elvi_occ_unique), "records are removed")
```

```{r}
table(elvi_occ_unique$basisOfRecord) # show the frequency table of “basisOfRecord”
```

```{r}
hist(elvi_occ_unique$year,main="",xlim=c(1895,2020),ylim=c(0,2000),col="orange2",xlab="years")
```

```{r}
elvi_occ_unique %>% 
  dplyr::select(lon, lat, year)%>% 
  filter(year %in% (1990:2020) & as.numeric(lon >=-126.374160))->elvi_occ_1990_2020 #  filter from 1990 to 2020 to match climatic data.

elvi_occ_unique  %>% 
  dplyr::select(lon, lat, year)%>% 
  filter(year %in% (1895:1925) & as.numeric(lon >=-126.374160))->elvi_occ_1895_1925

summary(elvi_occ_1990_2020$year) # show a quick summary of years in the data
summary(elvi_occ_1895_1925$year) # show a quick summary of years in the data

```

## Import endophyte host data from Josh database and merge them to existing online data

```{r}
endo_herb_georef %>% 
  dplyr::select(lon, lat, year,Spp_code)%>% 
  filter(Spp_code=="ELVI" & year %in% (1990:2020))->elvi_occgeoref_1990_2020

endo_herb_georef %>% 
  dplyr::select(lon, lat, year,Spp_code)%>% 
  filter(Spp_code=="ELVI" & year %in% (1895:1925))->elvi_occgeoref_1895_1925

elvi_occ_final_1990_2020<-rbind(elvi_occ_1990_2020[,-3],elvi_occgeoref_1990_2020[,-c(3,4)])
elvi_occ_final_1895_1925<-rbind(elvi_occ_1895_1925[,-3],elvi_occgeoref_1895_1925[,-c(3,4)])
dupfinal1990_2020 <- duplicated(elvi_occ_final_1990_2020[,c(1,2)])
dupfinal1895_1925 <- duplicated(elvi_occ_final_1895_1925[,c(1,2)])

elvi_occ_unique_final_1990_2020 <- elvi_occ_final_1990_2020[!dupfinal1990_2020,]
elvi_occ_unique_final_1895_1925<-elvi_occ_final_1895_1925[!dupfinal1895_1925,]

```

## Spatial thinning of species occurence records

```{r,eval=FALSE}
ID<-rep("ELVI",nrow(elvi_occ_unique_final_1990_2020))
elvi_sdm_final_1990_2020<-data.frame(ID,elvi_occ_unique_final_1990_2020)
names(elvi_sdm_final_1990_2020)<-c("ID","Long","Lat")
thinned.elvi_recent <- spThin::thin(elvi_sdm_final_1990_2020,lat.col = "Lat",long.col = "Long",spec.col = "ID",
                         thin.par=9, reps=10, locs.thinned.list.return = FALSE,
                         write.files = TRUE, max.files = 5, out.dir="/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Occurence", 
                         out.base = "points_elvi_thinned_recent", write.log.file = TRUE,
                         log.file = "spatial_thin_elvi_recent_log.txt", verbose = TRUE)

```

```{r}
elvi_thinned_recent<-read.csv("/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Occurence/elvi_thinned_recent.csv", header=T)
elvi_thinned<-elvi_thinned_recent[,-1]
```

## Projection on a map to double check if there is any point of the US

```{r}
library(maptools)
data(wrld_simpl)
plot(US_worldclim_recent_norm_stack_final,1)
plot(wrld_simpl, add=TRUE)
points(elvi_thinned, col='#0072B2',pch=19)
```

## Maxent model for *Elymus virginicus*

```{r}
set.seed(13)
group_pres_elvi <- kfold(elvi_thinned, 5)
pres_train_elvi <- elvi_thinned[group_pres_elvi != 1, ] # training data 
pres_test_elvi <- elvi_thinned[group_pres_elvi == 1, ] # testing data
```

```{r}
set.seed(13)
backg_elvi <- randomPoints(US_worldclim_recent_norm_stack_final, n=10000)
colnames(backg_elvi) = c('Long', 'Lat')
group_backg_elvi <- kfold(backg_elvi, 5)
backg_train_elvi <- backg_elvi[group_backg_elvi != 1, ]
backg_test_elvi <- backg_elvi[group_backg_elvi == 1, ]
```

```{r}
r <- raster(US_worldclim_recent_norm_stack_final, 1)
plot(!is.na(r), col=c('white', 'light grey'), legend=FALSE)
points(backg_train_elvi, pch='-', cex=0.5, col='#F0E442')
points(backg_test_elvi, pch='-',  cex=0.5, col='#000000')
points(pres_train_elvi, pch= '+', col='#009E73')
points(pres_test_elvi, pch='+', col='#0072B2')
```

```{r}
mod_elvi <- maxent(US_worldclim_recent_norm_stack_final, pres_train_elvi)
# plot(mod_aghy)
```

```{r}
e_elvi <- evaluate(pres_test_elvi, backg_test_elvi, mod_elvi, US_worldclim_recent_norm_stack_final)
map_recent_elvi <- predict(US_worldclim_recent_norm_stack_final, mod_elvi,  progress='')
map_old_elvi <- predict(US_worldclim_old_norm_stack_final, mod_elvi,  progress='')
writeRaster(map_recent_elvi, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Model/raster_recent_elvi.tif')
writeRaster(map_old_elvi, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Model/raster_old_elvi.tif')
```

```{r}
par(mfrow=c(2,2))
plot(map_recent_elvi, main='2020')
plot(wrld_simpl, add=TRUE, border='dark grey')
plot(map_old_elvi, main='1925')
plot(wrld_simpl, add=TRUE, border='dark grey')
```

```{r}
tr_elvi <- threshold(e_elvi, 'spec_sens')
plot(map_recent_elvi > tr_elvi, main='2020')
plot(wrld_simpl, add=TRUE, border='dark grey')
# points(pres_train_aghy, pch='+')

plot(map_old_elvi > tr_elvi, main='1925')
plot(wrld_simpl, add=TRUE, border='dark grey')
# points(pres_train_aghy, pch='+')
```

```{r}
map_recent_elvi_presence_absence<- (map_recent_elvi > tr_elvi)
writeRaster(map_recent_elvi_presence_absence, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Model/raster_recent_elvi_presence_absence.tif')

map_old_elvi_presence_absence<- (map_old_elvi > tr_elvi)
writeRaster(map_old_elvi_presence_absence, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Data/Model/raster_old_elvi_presence_absence.tif')
```

```{r}
# layout.matrix <- matrix(c(1, 2, 3, 4,5,6), nrow = 2, ncol = 3)
pdf("/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/github/EndoHerbarium/Manuscript/FigSDMMaxent.pdf",height = 5,width = 10)
# layout(mat = layout.matrix)
op<-par(mfrow = c(2,3), mar=c(3,3,2,1),oma = c(5, 5, 1, 3))
plot(map_old_aghy,xaxt='n',main=substitute(paste(italic("A. hyemalis"))))
text(x=-100,y=50,label="1925")
# par(mar=c(2.2,1,3,1))
plot(map_old_agpe,xaxt='n',yaxt='n', main=substitute(paste(italic("A. perennans"))))
text(x=-100,y=50,label="1925")
# par(mar=c(2.2,1,3,1))
plot(map_old_elvi, xaxt='n',yaxt='n',main=substitute(paste(italic("E. virginicus"))))
text(x=-100,y=50,label="1925")
# par(mar=c(5,3,0,1))
plot(map_recent_aghy, main="")
text(x=-100,y=50,label="2020")
# par(mar=c(5,1,0,1))
plot(map_recent_agpe,yaxt='n', main="")
text(x=-100,y=50,label="2020")
# par(mar=c(5,1,0,1))
plot(map_recent_elvi,yaxt='n', main="")
text(x=-100,y=50,label="2020")
title(xlab="Longitude", ylab="Latitude", outer=T, cex.lab=1.2,line = 1.5)

par(op)
dev.off()

```

```{r}
# layout.matrix <- matrix(c(1, 2, 3, 4,5,6), nrow = 2, ncol = 3)
pdf("/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/github/EndoHerbarium/Manuscript/FigSDMMaxent_Presence_absence.pdf",height = 5,width = 10)
# layout(mat = layout.matrix)
op<-par(mfrow = c(2,3), mar=c(3,3,2,1),oma = c(5, 5, 1, 3))
plot(map_old_aghy > tr_aghy,xaxt='n',main=substitute(paste(italic("A. hyemalis"))))
text(x=-100,y=50,label="1925")
# par(mar=c(2.2,1,3,1))
plot(map_old_agpe > tr_agpe,xaxt='n',yaxt='n', main=substitute(paste(italic("A. perennans"))))
text(x=-100,y=50,label="1925")
# par(mar=c(2.2,1,3,1))
plot(map_old_elvi > tr_elvi, xaxt='n',yaxt='n',main=substitute(paste(italic("E. virginicus"))))
text(x=-100,y=50,label="1925")
# par(mar=c(5,3,0,1))
plot(map_recent_aghy > tr_aghy, main="")
text(x=-100,y=50,label="2020")
# par(mar=c(5,1,0,1))
plot(map_recent_agpe > tr_agpe,yaxt='n', main="")
text(x=-100,y=50,label="2020")
# par(mar=c(5,1,0,1))
plot(map_recent_elvi > tr_elvi,yaxt='n', main="")
text(x=-100,y=50,label="2020")
title(xlab="Longitude", ylab="Latitude", outer=T, cex.lab=1.2,line = 1.5)

par(op)
dev.off()

```

```

### 2024-04-24 Manuscript/endo_herbarium.tex
adding updates to manuscript

@@
 \newcommand{\tom}[2]{{\color{red}{#1}}\footnote{\textit{\color{red}{#2}}}}
-
+\newcommand{\josh}[2]{{\color{orange}{#1}}\footnote{\textit{\color{orange}{#2}}}}
 %%%%%%%%%%%%%%%%%%%%%
@@
 % to %\linenumbers{} and %\modulolinenumbers[3] below.
-%
+\linenumbers
 % To avoid line numbering being thrown off around math environments,
@@
 %
-% (Thanks to Vlastimil Krivan for pointing this out to us!)
+% (Thanks to Vltimil Krivan for pointing this out to us!)
 
-\title{Increasing prevalence of plant-fungal symbiosis over two centuries of climate change}
+\title{Increasing prevalence of plant-fungal symbiosis across two centuries of environmental change}
 
@@
 
-\author{Joshua C. Fowler$^{1,\ast}$ \\
+\author{Joshua C. Fowler$^{1,2\ast}$ \\
+	Jacob Moutouama$^{1}$\\
 	Tom E. X. Miller$^{1}$}
@@
 	\noindent{} 1. Rice University, Department of BioSciences, Houston, Texas 77006;
+	\noindent{} 1. University of Miami, Department of Biology, Miami, Florida;
 
-	\noindent{} $\ast$ Corresponding author; e-mail: jcf3@rice.edu.
+	\noindent{} $\ast$ Corresponding author; e-mail: jcf221@miami.edu.
 
@@
 	\section*{Abstract}
-Many species' distributions and abundances have shifted in response to climate change. 
+Species' distributions and abundances are shifting in response to climate change. 
 Most species harbor microbial symbionts that have the potential to influence these responses.
 Mutualistic microbial symbionts may provide resilience to environmental change by protecting their hosts from increasing stress.
-Alternatively, environmental change that causes declines in host or symbiont may disrupt the interaction.
+Alternatively, environmental change disrupts these interactions may lead to declines in hosts or symbionts.
 Microbes can be preserved within the dried plant tissue of herbarium specimens presenting an opportunity to quantify changes across broad temporal and spatial scales. 
-We ask how the prevalence of a symbiont of grasses, \emph{Epichloë} fungal endophytes, which can protect hosts from drought, has changed over time in response to climate change, and how these changes vary across the hosts' ranges.
+We ask how the prevalence of symbionts of grasses, \emph{Epichloë} fungal endophytes, which can protect hosts from drought, have changed over time in response to climate change, and how these changes vary across hosts' ranges.
 Specifically, we took seed samples from herbarium specimens of three grass host species collected over the last two centuries, quantified fungal hyphae presence within the seeds using microscopy, and evaluated spatial and temporal trends.
 Overall, endophytes have increased in prevalence over the last two centuries from 25\% prevalence to 75\% prevalence on average across the three hosts.
-We also found that that the changes in prevalence were associated with observed changes in annual and seasonal climate drivers corresponding with each hosts' peak growing season. 
-Thus we provide novel evidence for a cryptic biological response to climate change that may contribute to the resilience of the symbiosis.
+We also found that changes in prevalence were associated with observed changes in annual and seasonal climate drivers corresponding with each hosts' peak growing season. 
+Thus we provide novel evidence for a cryptic biological response to climate change that may contribute to the resilience of the symbiosis through ongoing environmental change.
 
@@
 Coral bleaching (the loss of symbiotic algae) due to temperature stress \citep{sully2019global} is perhaps the best known example, but this phenomenon is not unique to corals.
-Lichens exposed to elevated temperatures experienced loss of photosynthetic function along with changes in the community composition of their algal symbionts \citep{meyer2022climate}.
-How commonly and under what conditions microbial mutualisms deteriorate or strengthen under climate change remain unanswered questions, but previous work suggests that these alternative responses may depend on the intimacy and specialization of the interaction as well as the physiological tolerances of the mutualists \citep{toby2010mutualisms, warren2014mutualism, rafferty2015phenological}.
+Lichens exposed to elevated temperatures experienced loss of photosynthetic function along with changes in the composition of their algal symbiont community \citep{meyer2022climate}.
+How commonly and under what conditions microbial mutualisms deteriorate or strengthen under climate change remain unanswered questions, but previous work suggests that these alternative responses may depend on the intimacy and specialization of the interaction as well as the physiological tolerances of the mutualist partners \citep{toby2010mutualisms, warren2014mutualism, rafferty2015phenological}.
 
@@
 Facultative symbionts may be absent from portions of the host range \cite{afkhami2014mutualist}, and hosts may engage with a diversity of partners (different species, or different locally-adapted strains of one species) across their environments \cite{frade2008variation, rolshausen2018quantifying}.
-Identifying broader spatial trends in symbiont prevalence is therefore an important step in developing predictions of where to expect changes in the symbiosis in future climates.
+Identifying broader spatial trends in symbiont prevalence is therefore an important step in developing predictions for where to expect changes in the symbiosis in future climates.
 
+\josh{
 \emph{Epichloë} fungal endophytes are specialized symbionts of cool-season grasses, which have been documented in $\sim 30$\% of cool-season grasses \citep{leuchtmann1992systematics}
@@
 Previous survey studies have documented large-scale spatial patterns in  endophyte prevalence structured by environmental gradients \citep{granath2007variation,bazely2007broad, afkhami2012fungal,sneck2017variation}.
-We predicted that prevalence should also track temporal changes in environmental drivers that elicit these fitness benefits.
+We predicted that prevalence should track temporal changes in environmental drivers that elicit these fitness benefits.
 For example, endophyte-mediated drought tolerance should lead prevalence to increase in regions where precipitation declines over time.
@@
 This creates the opportunity to leverage the unique spatio-temporal sampling of herbarium collections to examine the response of the symbiosis to historical climate change.
+}{Volker's suggestions included moving all material specific to study system into methods, but it seems a bit difficult to do.}
 
 In this study, we assessed for the first time the long-term responses of endophyte symbiosis to climate change through the use of herbarium specimens of three \emph{Epichloë} endophyte host species, \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}.
-We first address questions describing spatial and temporal trends in endophyte prevalence: (i) How does endophyte prevalence vary across space?, (ii) How has endophyte prevalence changed over time?
-and (iii) How have temporal changes in endophyte prevalence differed across space?
-To answer these questions, we fit a spatio-temporal statistical model of endophyte prevalence.
-and evaluated the predictive ability of the model using contemporary endophyte surveys as out-of-sample test data, an important but rarely used strategy in ecological studies \cite{tredennick2021practical}.
-We then address how climate change may be driving these trends in endophyte prevalence by addressing the question: (iv) What is the relationship between variation in temporal trends in endophyte prevalence and changes in climate drivers?
-We examined 1,865 specimens collected across across eastern North America between 1824 and 2019.
+We first address questions describing spatial and temporal trends in endophyte prevalence: (i) How has endophyte prevalence changed over time?
+and (ii) How spatially heterogenous are temporal trends in endophyte prevalence?
+We then address how climate change may be driving these trends in endophyte prevalence by asking: (iii) What is the relationship between variation in temporal trends in endophyte prevalence and changes in climate drivers?
+We examined 2,346 specimens collected across across eastern North America between 1824 and 2019.
 
@@
 		\subsection*{Herbarium surveys}
-We visited herbaria between 2019 and 2022, including The Botanical Research Institude of Texas (which also houses the Vanderbilt University herbarium and the R.
-Dale Thomas Plant collection from the University of Louisiana at Monroe), the Louisiana State University Herbarium, the Mercer Botanic Garden Herbarium, Missouri Botanic Garden's Herbarium, the S.M.
-Tracy Herbarium at Texas A\&M, the University of Kansas Herbarium, the Robert Bebb Herbarium at the University of Oklahoma, the Lundell Herbarium and the UT Herbarium at the University of Texas, and Oklahoma State University Herbarium (See Table A1 for a summary of specimens included from each collection).
-In total, we quantified endophyte symbiosis for $964$ \emph{A.
-hyemalis} specimens collected between 1824 and 2019, $299$ \emph{A.
-perennans} specimens collected between 1863 and 2017, and $602$ \emph{E.
-virginicus} specimens collected between 1839 and 2017 (Fig.
+We visited nine herbaria between 2019 and 2022 (see Table A1 for a summary of specimens included from each collection). 
+In total, we quantified endophyte symbiosis for $1135$ \emph{A.
+hyemalis} specimens collected between 1824 and 2019, $357$ \emph{A.
+perennans} specimens collected between 1863 and 2017, and $854$ \emph{E.
+virginicus} specimens collected between 1839 and 2019 (Fig.
 1, Fig 2A).
+
 Our sampling plan was designed to minimize damage to these specimens.
@@
 In some cases, the tissues examined during microscopy came from flowers or otherwise non-viable seeds, which were excluded from the seed counts for that specimen.
-On average we scored 4.7 seeds per specimen in \emph{A. hyemalis}, 5 seeds per specimen in \emph{A. perennans}, and 3.8 seeds per specimen in \emph{E. virginicus}.
+On average we scored $4.7$ seeds per specimen in \emph{A. hyemalis}, $4.2$ seeds per specimen in \emph{A. perennans}, and $3.8$ seeds per specimen in \emph{E. virginicus}.
 Due to imperfect vertical transmission \cite{afkhami2008symbiosis}, it is possible that symbiotic host-plants produce a mixture of symbiotic and non-symbiotic seeds. 
@@
 
-\begin{figure}[H]
+\begin{figure}[h]
 	\centering
@@
 \subsection*{Assessing spatial and temporal changes in endophyte prevalence}
-To quantify spatial and temporal trends in endophyte prevalence, we used an approximate Bayesian method, Integrated Nested Laplace Approximation (INLA), to fit a spatio-temporal model of endophyte prevalence.
-Compared to MCMC Bayesian methods, INLA provides a computationally more efficient method of ascertaining model posteriors for certain models that can be formulated as latent Gausssian Models.
-\cite{rue2009approximate}.
+We assessed spatial and temporal changes in endophyte prevalence across each host distribution, first quantifying the long-term overall trends and then examining how these trends vary across the spatial extent of each host distribution.
+We used an approximate Bayesian method, Integrated Nested Laplace Approximation (INLA), to construct spatio-temporal models of endophyte prevalence.
+INLA provides a computationally more efficient method of ascertaining model posteriors for certain models that can be formulated as latent Gausssian Models \cite{rue2009approximate}. 
 Many common statistical models, including structured and unstructured mixed-effects models, can be represented as latent Gaussian Models. 
-Fitting models with structured spatial effects is possible with MCMC sampling but can require long computation times, making INLA an effective alternative, which has been used to model spatial patterns in flowering phenology \cite{willems2022forest}, habitat overlap between marine predators and prey \cite{sadykova2017bayesian}, the distribution of temperate trees \cite{engel2022spatial} as well as the population dynamics of endangered amphibians \cite{knapp2016large} and other ecological processes \cite{beguin2012hierarchical}.
+Fitting models with structured spatial effects is possible with MCMC sampling but can require long computation times, making INLA an effective alternative, which has been used to model spatial patterns in flowering phenology \cite{willems2022forest}, the abundance of bird species \cite{meehan2019spatial} and butterflies cite{crossley2022opposing}, the distribution of temperate trees \cite{engel2022spatial} as well as the population dynamics of endangered amphibians \cite{knapp2016large} and other ecological processes \cite{beguin2012hierarchical}.
 
-We modeled endophyte presence/absence ($P$) as a Bernoulli response variable for specimen $i$ of host species $h$.
+To quantify overall trends in endophyte prevalence, we modeled endophyte presence/absences ($P$)  as a Bernoulli response variable for each specimen $s$ at a given location $i$.
+In all cases, models for each species were fit separately.
 
@@
 	\begin{align}
-		P_{i,h} \sim Bernoulli(\hat{P_{i,h}}) \\
-			logit(\hat{P}_{h}) = \beta_{0,h} + 
-		\beta_{1,h}*year_{i} + \\
-		\beta_{2,h}*year_{i} *lat_{i} + 
-		\beta_{3,h}*year_{i} *lon_{i} + 
-		\beta_{4,h}*year_{i} *lat_{i}*lon_{i} +\\
-		\chi + \omega + \phi
+		P_{s.i} \sim Bernoulli(\hat{P_{s,i}}) \\
+		logit(\hat{P}_{s,i}) = \alpha_{s,i} + \beta_{s}*year_i+
+		\chi_{i} + \omega_{i} 
 	\end{align}
@@
 
-The expected endophyte prevalence, $\hat{P}$, was modelled with intercepts specific to each host species ($\beta_{0}$), slopes for changes over time ($\beta_{1}$) as well as the interaction between time and the specimen's latitude and longitude ($\beta_{2}$, $\beta_{3}$, and $\beta_{4}$).
-To quantify spatial patterns in prevalence, we included a spatially-dependent random intercept, $\phi$.
-This random spatial effect was modeled as a stochastic partial differential equation (SPDE) that depended on a covariance matrix according to the proximity of each collection location \citep{lindgren2011explicit,bakka2018spatial}.
-The SPDE model allowed us to flexibly estimate endophyte prevalence at locations across the study region, while accounting for potential spatial autocorrelation between data points via the covariance matrix.
-The covariance matrix was approximated using a Matérn covariance function, with each data point assigned a location according to the nodes of a mesh of non-overlapping triangles across our study area (Fig A2).
+The expected endophyte prevalence, $\hat{P}$, was modelled with spatially-indexed random intercepts $\alpha$ that accounts for potential spatial autocorrelation between data points, and with a slope $\beta$ describing the overall change in endophyte prevalence.
 We accounted for potential biases introduced during the process of collecting specimens as well as in scoring ability by including random effects specific to each collector $\chi$ and to each scorer $\omega$.
@@
 Similarly, the process of scoring seeds for hyphae involved many individual researchers, who may vary in their likelihood of positively identifying fungal hyphae. 
-By including a random effect for collectors and for scorers, we can account for the variance across individual researchers that may bias our predictions of changes in endophyte prevalence.
-We performed model fitting using the rINLA package, \citep{lindgren2015bayesian}, with vague priors, and compared models with different sizes of mesh, which had little effect on the resulting model estimates.
+By including a random effect for collectors and for scorers, we accounted for variance across individual researchers that may bias our predictions of changes in endophyte prevalence.
+
+To quantify how temporal trends may vary spatially, we repeated the modelling above, but instead incorporating a spatially-varying coefficient for collection year.
+
+\begin{subequations}
+	\label{eq:trends}
+	\begin{align}
+		P_{s.i} \sim Bernoulli(\hat{P_{s,i}}) \\
+		logit(\hat{P}_{s,i}) = \alpha_{s,i} + \tau_{s,i}*year_i +
+		\chi_{i} + \omega_{i} 
+	\end{align}
+\end{subequations}
+
+Note that the expected endophyte prevalence, $\hat{P}$, was modelled as above, and that $\tau$ describes the spatially-varying random slope for collection year.
+This spatially-varying random effect allowed us to flexibly estimate variation in slope at locations across the study region.
+All spatially-structured random intercepts and slopes were constructed using stochastic partial differential equations (SPDE) that depend on a covariance matrix according to the proximity of each collection location \citep{lindgren2011explicit,bakka2018spatial}.
+The covariance matrix was approximated using a Matérn covariance function, with each data point assigned a location according to the nodes of a mesh of non-overlapping triangles across our study area (Fig A2).
+
+We performed model fitting using the inlabru R package \citep{}, with vague priors, and compared models with different sizes of mesh, which had little effect on the resulting model estimates.
+Each mesh was bounded by the predicted host distribution, described below.
 Posterior modes were stable indicating that numeric convergence was successful.
 We assessed model fit with graphical posterior predictive checks (Fig. A3).
-The model performed adequately at classifying the historical data, comparing the accuracy of predictions from the model with observed data (AUC = 0.77; Fig. A4). 
+The model performed adequately at classifying the historical data, comparing the accuracy of predictions from the model with observed data (avg. AUC = 0.77; Fig. A4).
+
+\subsection*{Modeling distributions of host species}
+
+Jacob needs to write this, but we built SDMs using MaxEnt and the same climate predictors described below.
+
 \subsection*{Validating the model with an out-of-sample test}
-We used data from contemporary surveys of endophyte prevalence  in \emph{A. hyemalis} and \emph{E. virginicus} as test data to evaluate predictions of the model. 
+\josh{
+
+We evaluated the predictive ability of the model using contemporary endophyte surveys as out-of-sample test data, an important but rarely used strategy in ecological studies \cite{tredennick2021practical}.
+We used data from contemporary surveys of endophyte prevalence  in \emph{A. hyemalis} and \emph{E. virginicus}. 
 Surveys of \emph{E. virginicus} were conducted in 2013 as described in \citet{sneck2017variation}, and surveys of \emph{A. hyemalis} took place between 2015 and 2020.
@@
 We compared the model's prediction for these locations to the observed population prevalence.
-
+}{considering dropping this section because the model does a good job at predicting the mean regional prevalenece, but has way less variability that the contemporary surveys}
 		\subsection*{Assessing the role of climate drivers}
@@
 
-We calculated the relative change in endophyte prevalence across the 1925 to 2020, i.e., the difference between predicted prevalence divided by the endophyte prevalence in 1925, because regions varied in their predicted starting prevalence.
-This time period connects the predicted change in prevalence with the endpoints of the available climate record.
-We then calculated the Spearman correlation coefficient between the relative change in endophyte prevalence and each climate driver.
+To evaluate whether areas that have experienced high degrees of change in climate have experienced the greatest changes in endophyte prevalence, we explored correlations between these climate drivers and the predicted spatially-varying trends with a Gaussian model:
+
+\begin{subequations}
+	\label{eq:trends}
+	\begin{align}
+		\tau_{i} \sim Normal(\hat{\mu}_{i}, \sigma^2) \\
+	\hat{\mu}_{i}= \beta *\mathbf{X}_{env}
+	\end{align}
+\end{subequations}
+
+where the trend at a given location ($\tau_{i}$) is predicted by our environmental covariates {$\mathbf{X}_{env}$} according to the vector of parameters $\beta$. 
 Calculating correlations from many pixels across the study region risks articially inflating confidence in our results due to large sample sizes, and so we repeated this calculation using only a random subsample of 100 pixels across the study region.
@@
 \section*{Results}
-\subsection*{Temporal and spatial trends}
+\subsection*{How has endophyte prevalence changed over time?}
+We found that endophyte prevalence increased within the examined specimens over the last two centuries for all three host species (Fig. 4). 
+On average, \emph{A.
+hyemalis} and \emph{E.
+virginicus} both increased from ~30 \% to over 70\% prevalence across the study region, and \emph{A.
+perennans} increased from ~ 15\% to over 70\% prevalence.
+\josh{Our model indicates a higher certainty that overall temporal trends are positive for \emph{A.
+hyemalis} and \emph{A.
+perennans} than for \emph{E.
+virginicus} (99\% probability of a positive overall year slope in \emph{A.
+hyemalis}, 89\% probability of a positive overall year slope in \emph{A.
+perennans}, and 58\% probability of a positive overall year slope in \emph{E.
+virginicus}) .}{These numbers are currently outdated.
+I am making some adjustments to models, and will update with final model}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width = \linewidth]{year_plot.png}
+	\caption[Temporal trends in endophyte prevalence.]{Temporal trends in endophyte prevalence.
+(A) Histograms show the frequency of collection through time for each host species.
+(B) Colored points are binned means of the observed endophyte presence/absence data (black dashes).
+Colors represent each host species and point size is determined by the number of specimens.
+Lines show predicted mean endophyte prevalence over the study period along with the 50\%  and 95\% CI bands.}
+\end{figure}
+
+\subsection*{How spatially heterogenous are temporal trends in endophyte prevalence?}
+Our model revealed hotspots of change in endophyte prevalence . 
+While there was an overall increase in endophyte prevalence, these changes varied across the host species' ranges (Fig. 3).
+In some regions, posterior estimates of our spatially varying temporal trends, $\tau$, indicate that \emph{A.
+hyemalis} and \emph{A.
+perennans} experienced increases in percent prevalence by as much as 4\% per year over the study period, while  \emph{E.
+virginicus} experienced increases up to around 1.5 \% per year.
+In other regions, there were negligible changes.
+Notably, the symbionts of \emph{E.
+virginicus} experienced only slight increases in prevalence, and were less spatially variable than the other two species.
+\josh{Regions that start with low endophyte prevalence, as in the southwestern portion of the range of \emph{A.
+hyemalis}(Fig.
+A1), also experienced negligible change, suggesting that this may be driven more by the absence of the endophyte.}{more discussion material, but putting it here for now.}
+Predicted trends for \emph{A. perennans} show certain areas of both large increase and of large decrease, however this species, for which we have the fewest samples, has the largest uncertainty.
+The posterior estimates of our spatially varying temporal trends, indicate relatively narrow certainty (\josh{need to compute}{}).
+
+\begin{figure}[H]
+	\label{fig:svc_time_map}
+	\centering
+	\includegraphics[width = .6\linewidth]{svc_time_map.png}
+	\caption[Predicted posterior mean of spatially-varying slopes representing change in endophyte prevalence for each host species.]{Predicted posterior mean of spatially-varying slopes representing change in endophyte prevalence for each host species.
+Color indicates the relative change in predicted endophyte prevalence.}
+\end{figure}
+
+\subsection*{Assessing collector and scorer influences on predicted endophyte prevalence}
 We quantified temporal and spatial trends in endophyte prevalence while accounting for potential biases introduced by collectors and by individuals who quantified endophyte presence/absence with the use of random effects.
-We found no evidence that collector biases influenced our results (collector random effects were consistently small; Fig 2A), while the identity of individual scorers did contribute to observed patterns in endophyte prevalence (3 of the 16 scorers were more likely than average to assign positive endophyte status, as indicated by 95\% credible intervals that do not overlap 0) (Fig 2B).
-Interpretation of the data without accounting for this source of variation would lead to potentially biased interpretations of temporal and spatial trends.
+We found no evidence that collector biases influenced our results. 
+Collector random effects were consistently small; Fig 4A, and models fit with and without this random effect provide qualitatively similar results.
+The identity of individual scorers did contribute to observed patterns in endophyte prevalence.
+For example, 3 of the 16 scorers were more likely than average to assign positive endophyte status, as indicated by 95\% credible intervals that do not overlap 0) (Fig 4B). 
+However, this may have been driver by differences in scorers biases during the seed scoring process, or by unintended spatial clustering of the specimens scored by each scorer. 
+Interpreting our models with the inclusion of the scorer effect thus provides conservative estimates of the absolute magnitude of changes in endophyte prevalence.
 
@@
 
-\subsubsection*{How does endophyte prevalence vary across space?}
-Across space, there were clear trends in endophyte prevalence which varied between species (Fig. 3).
-We mapped predictions of endophyte prevalence across the area encompassing the sampled herbarium specimens for each species, and area smaller than the full geographic distribution of each species in nature.
-\emph{Elymus virginicus} had higher prevalence towards the northern portions of the collection range. 
-In contrast, \emph{Agrostis hyemalis} had highest prevalence towards the center of its collection range with regions of low prevalence towards the northeast and towards its western range edge.
-\emph{Agrostis perennans} had  highest prevalence towards the westernmost portion of its collection range.
-There is considerable uncertainty in the spatial pattern of endophyte prevalence (See Fig.
+%\subsubsection*{How does endophyte prevalence vary across space?}
+%Across space, there were clear trends in endophyte prevalence which varied between species (Fig. 5).
+%We mapped predictions of endophyte prevalence across the area encompassing the sampled herbarium specimens for each species, and area smaller than the full geographic distribution of each species in nature.
+%\emph{Elymus virginicus} had higher prevalence towards the northern portions of the collection range. 
+%In contrast, \emph{Agrostis hyemalis} had highest prevalence towards the center of its collection range with regions of low prevalence towards the northeast and towards its western range edge.
+%\emph{Agrostis perennans} had  highest prevalence towards the westernmost portion of its collection range.
+%There is considerable uncertainty in the spatial pattern of endophyte prevalence (See Fig.
 A10 for projections of the 95\% credible interval), however these broad spatial patterns are consistent across the low and high end of model predictions.
@@
 \begin{figure}[H]
-	\label{fig:prevalence_map}
+	\label{fig:svc_space_map}
 	\centering
-	\includegraphics[width = \linewidth]{prevalence_map.png}
+	\includegraphics[width = .6\linewidth]{svc_space_map.png}
 	\caption{\textbf{Mean predicted endophyte prevalence for each host species (columns) in 1925 (top row) and 2020 (bottom row)}.
-Color indicates mean predicted rate of endophyte prevalence across a convex boundary enscribing the collections locations of each species (excluding two disjoint collections of \emph{A.
-perennans} to the west).}
+Color indicates mean predicted rate of endophyte prevalence across the predicted distribution of each species.}
 \end{figure}
 
-\subsubsection*{How has endophyte prevalence changed over time?}
-We found that endophyte prevalence increased within the examined specimens over the last two centuries for all three host species (Fig. 4). 
-On average, \emph{A.
-hyemalis} and \emph{E.
-virginicus} both increased from ~30 \% to over 50\% prevalence across the study region, and \emph{A.
-perennans} increased from ~ 15\% to over 70\% prevalence.
-Rerunning the analysis excluding specimens collected before 1900 when samples are much sparser (Fig.
-2A), lead to qualitatively similar predictions, and so we continued analyses with the full dataset.
-Our model indicates a higher certainty that overall temporal trends are positive for \emph{A.
-hyemalis} and \emph{A.
-perennans} than for \emph{E.
-virginicus} (99\% probability of a positive overall year slope in \emph{A.
-hyemalis}, 89\% probability of a positive overall year slope in \emph{A.
-perennans}, and 58\% probability of a positive overall year slope in \emph{E.
-virginicus}) .
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width = \linewidth]{year_plot.png}
-	\caption{\textbf{Temporal trends in endophyte prevalence.} (A) Histograms show the frequency of collection through time for each host species.
-(B) Colored points are binned means of the observed endophyte presence/absence data (black dashes).
-Colors represent each host species and point size is determined by the number of specimens.
-Lines show predicted mean endophyte prevalence over the study period at low (dotted) high (solid) latitudes and along with 95\% CI bands.}
-\end{figure}
-
-\subsubsection*{How have temporal changes in endophyte prevalence differed across space?}
-Our model revealed that while there was an overall increase in endophyte prevalence, these changes varied across the host species' ranges.
-The posterior estimates of the parameters describing space by time interactions, $\beta2$, $\beta3$, and $\beta4$ for each species indicate small effect sizes, but high certainty(i.e, a 99\% posterior probability of a negative interaction between year and latitude for \emph{A.
-hyemalis}, or a 67\% postterior probabilty of a postitive interaction between year and latitude for \emph{E.
-virginicus}) (Fig.
-A5).
-In some regions, \emph{A.
-perennans} experienced increases in percent prevalence by as much as 35 percentage points between 1925 and 2020, while \emph{A.
-hyemalis} and \emph{E.
-virginicus} experienced increases up to around 15 percentage points.
-In other regions, there were smaller increases of only five percentage points for \emph{E. virginicus} and fifteen percentage points for \emph{A. perennans}. 
-We calculated relative change in endophyte prevalence because regions differ in their initial endophyte prevalence.
-
-\begin{figure}[H]
-	\label{fig:rel_change_plot}
-	\centering
-	\includegraphics[width = .8\linewidth]{rel_change_plot.png}
-	\caption{\textbf{Relative change in predicted prevalence for each host species.} Color indicates the relative change in predicted endophyte prevalence between 1925 and 2020 across the collection range.}
-\end{figure}
-
-This revealed locations that were hotspots of change in endophyte prevalence (Fig. 5). 
-The overall magnitude of changes were greatest in \emph{A. perennans}, particularly in the northeast and in the south, where endophytes were previously only found at low prevalence (Fig. 3).
-\emph{A.
-hyemalis} and \emph{E.
-virginicus} experience lower absolute increases in prevalence, yet we can still identify hotspots of change in the northeast for \emph{A.
-hyemalis} and in the southeast for \emph{E.
-virginicus}.
-
-\subsubsection*{Performance on test data}
-The model adequately predicted mean endophyte prevalence, but underpredicted variance in the contemporary test observations (Fig. 6 A).
-The model was able to predict large scale spatial trends in endophyte prevalence, such as declining endophyte prevalence towards western longitudes in \emph{A. hyemalis} (Fig. 6 B-C).
-We interpreted this to mean that the model captured regional spatial dynamics, but underpredicts local scale dynamics. We discuss potential model improvements in the Discussion.
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width = \linewidth]{contemp_test_plot.png}
-	\caption{\textbf{Predicted vs observed endophyte prevalence for contemporary test data.} (A) The model, trained on historic herbarium collection data, performed modestly at predicting contemporary endophyte prevalence, as indicated by some overlap of predicted 95\% CI with the 1:1 line, however contemporary test data generally had more variance between populations than model predictions.
-The model did recapitulate broader regional trends across (B) longitude and (C) latitude.
-Colors indicate host species, and point size in panels B and C reflect sample sizes of contemporary endophyte population surveys.}
-\end{figure}
-
-\subsection*{The role of climate drivers}
-\subsubsection*{What is the relationship between variation in temporal trends in endophyte prevalence and changes in climate drivers?}
+\subsection*{What is the relationship between variation in temporal trends in endophyte prevalence and changes in climate drivers?}
 
@@
 For \emph{A. hyemalis}, endophyte prevalence increased most strongly in regions that experienced reductions in autumn precipitation variability. 
-Correlations using only a subsampling of pixels were qualitatively similar to these results (Fig.
-A11), suggesting that the patterns we find, particulary the strongest correlations are not spurious associations.
+Correlations using only a subsampling of pixels were qualitatively similar to these results (Fig. A11), suggesting that the patterns we find are not spurious associations.
 
@@
 Or could this result have occurred by chance?}
+\josh{}{Only have plotted results for AGHY right now.}
+\begin{figure}[H]
+	\centering
+	\includegraphics[width = .7\linewidth]{tmean_regression_plot_ESA.png}
+	\caption{\textbf{Correlations between changes in climate drivers and changes in endophyte prevalence.} Color denotes the Spearman correlation coefficient between the relative rate of change in endophyte prevalence and the change in annual mean temperature ($^oC$) and total annual and seasonal precipitation (mm), as well as the change in the coefficient of variation of each climate driver.
+Positive correlation coefficients indicate that greater increases in a climate driver were associated with larger increases in endophyte prevalence, while negative values indicate that .
+Asteriks denote correlation coefficients $> .3$ or $< -.3$.}
+\end{figure}
 
@@
 	\centering
-	\includegraphics[width = \linewidth]{climate_corr_heatmap.png}
+	\includegraphics[width = .7\linewidth]{ppt_regression_plot_ESA.png}
 	\caption{\textbf{Correlations between changes in climate drivers and changes in endophyte prevalence.} Color denotes the Spearman correlation coefficient between the relative rate of change in endophyte prevalence and the change in annual mean temperature ($^oC$) and total annual and seasonal precipitation (mm), as well as the change in the coefficient of variation of each climate driver.
@@
 Asteriks denote correlation coefficients $> .3$ or $< -.3$.}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width = .7\linewidth]{ppt_CV_regression_plot_ESA.png}
+	\caption{\textbf{Correlations between changes in climate drivers and changes in endophyte prevalence.} Color denotes the Spearman correlation coefficient between the relative rate of change in endophyte prevalence and the change in annual mean temperature ($^oC$) and total annual and seasonal precipitation (mm), as well as the change in the coefficient of variation of each climate driver.
+Positive correlation coefficients indicate that greater increases in a climate driver were associated with larger increases in endophyte prevalence, while negative values indicate that .
+Asteriks denote correlation coefficients $> .3$ or $< -.3$.}
+\end{figure}
+
+\subsubsection*{Performance on test data}
+We found that while the model predicts broader regional trends in endophyte prevalence present in the contemporary survey data such as declining endophyte prevalence towards western longitudes in \emph{A.
+hyemalis} (Fig.
+6 B-C), however the contemporary data contains additional variability at smaller scales not captured by our sampling of herbarium specimens.
+We interpreted this to mean that the model captured regional spatial dynamics, but underpredicts local scale dynamics. 
+We discuss potential model improvements in the Discussion.
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width = \linewidth]{contemp_test_plot.png}
+	\caption{\textbf{Predicted vs observed endophyte prevalence for contemporary test data.} (A) The model, trained on historic herbarium collection data, performed modestly at predicting contemporary endophyte prevalence in \emph{A.
+hyemalis}, as indicated by some overlap of predicted 95\% CI with the 1:1 line, however contemporary test data generally had more variance between populations than model predictions.
+The model did recapitulate broader regional trends across (B) longitude and (C) latitude.
+Point size in panels B and C reflect sample sizes of contemporary endophyte population surveys.}
 \end{figure}
@@
 This interpretation is in line with theory predicting that the positive fitness feedback caused by vertical transmission leads beneficial symbionts to rise in prevalence within a population \cite{fine1975vectors}.
-We found that trends in endophyte prevalence varied across the collection range of each species in assocation with observed changes in climate drivers, suggesting that the endophytes have contributed to host resilience under environmental change.
+We found that trends in endophyte prevalence varied across the distribution of each species in assocation with observed changes in climate drivers, suggesting that the endophytes have contributed to host resilience under environmental change.
 Taken together, this suggests a strengthening of the mutualism over the last two centuries.
@@
 Differences between the responses of each host species underscore that while all of these $C_3$ grasses share similar broad-scale distributions, each engages in unique biotic interactions and has unique niche requirements.
-We identified hotspots of change in the northeast and south for \emph{A. perennans}, which experienced the strongest absolute changes in endophyte prevalence (Fig. 5). 
+We identified hotspots of change for \emph{A. perennans}, which experienced the strongest absolute changes in endophyte prevalence (Fig. 5).
+Declines in the southern portion of its range and increases in the north suggest a potential poleward range shift of endophytic plants. 
 Based on previous work demonstrating that endophytes can shield their hosts from drought stress \cite{decunta2021systematic}, we generally predicted that drought conditions could be a driver of increasing endophyte prevalence.
-In line with this expectation, increasing prevalence for this species was associated with decreasing precipitation, most strongly with spring-season declines (Fig. 7). 
+In line with this expectation, increasing prevalence for this species was associated with decreasing precipitation, most strongly with autumn-season declines (Fig. 7). 
 \emph{A. perennans} typically blooms in the autumn.
-Endophytes could be playing a role helping hosts weather spring-season droughts while the species is dormant. 
+Endophytes could be playing a role helping hosts weather autumn-season droughts while the species is dormant. 
 It may be useful to investigate whether lagged climate effects are important predictors of host fitness in this system \cite{evers2021lagged}. 
@@
 The strong increase in symbiotic \emph{A. perennans} could be explained, at least in part, by these diverse benefits. 
-\emph{A. hyemalis} experienced smaller overall changes in endophyte prevalence, and these changes were only weakly related to changes in climate drivers. 
-This result was surprising given previous work demonstrating drought benefits in a greenhouse manipulation with this species \citep{davitt2011understanding} that led us to expect  that endophyte prevalence should similarly increase at a greater rate in regions that have experienced increasing drought.
-It is possible that the study region has not experienced a magnitude of climate change great enough to cause larger changes in endophyte prevalence for this species. 
-Weak associations with drought could also be explained by climate-driven changes in the rate of imperfect transmission (the generation of non-symbiotic offspring from symbiotic hosts), which could counterbalance endophyte-mediated fitness benefits \citep{donald2021context}.
+\emph{A. hyemalis} experienced more consistently positive increases in endophyte prevalence related to changes in spring temperature and precipitation.  
+This result is in line with previous work demonstrating drought benefits in a greenhouse manipulation with this species \citep{davitt2011understanding} that led us to expect  that endophyte prevalence should similarly increase at a greater rate in regions that have experienced increasing drought.
+%It is possible that the study region has not experienced a magnitude of climate change great enough to cause larger changes in endophyte prevalence for this species. 
+%Weak associations with drought could also be explained by climate-driven changes in the rate of imperfect transmission (the generation of non-symbiotic offspring from symbiotic hosts), which could counterbalance endophyte-mediated fitness benefits \citep{donald2021context}.
 For \emph{E.
-virginicus}, which also experienced more modest changes in endophte prevalence overall, we found a strong relationship between temporal trends and  changes in the mean and variability of temperature, as well as with decreases in summer precipitation.
-Surveys by \citet{sneck2017variation}, used as part of the test data in this study, identified a drought index (SPEI) that integegrates precipitation with estimated evapotranspiration as an important predictor of endophyte prevalence.
+virginicus}, which experienced the most modest changes in endophte prevalence overall, we found a strong relationship between temporal trends and  changes in the mean and variability of temperature, as well as with decreases in summer precipitation.
+Surveys by \citet{sneck2017variation}, used as part of the test data in this study, identified a drought index (SPEI) that integrates precipitation with estimated evapotranspiration as an important predictor of endophyte prevalence.
 While we show consistent increasing trends in prevalence between the three species, the mechanisms that explain these changes may be diverse and idiosyncratic.
@@
 Lack of information on local variability may simply be a feature of data derived from herbarium specimens. 
-Even though they are samples from local populations, that are single specimens that are aggregated over in broad-scale model estimates.
+Even though they are samples from local populations, they are single specimens that are aggregated over in broad-scale model estimates.
 Poor predictive ability at local scales in this grass-endophyte system is not surprising, as previous studies have found that local variation, even to the scale of hundreds of meters can structure endophyte-host niches \cite{kazenel2015mutualistic}.
@@
 Incorporating available climatic and soil layers as covariates is an obvious first step that could improve predictions.
-Towards the goal of predicting the microbial symbioses dynamics under climate change, models that integrate data from local and regional scales would be an important step to bridge the gap that often exists between large but broad bioclimatic and biodiversity data and small but local data on biotic interactions.
+Towards the goal of predicting the dynamics of microbial symbioses under climate change, models that integrate data from local and regional scales would be an important step to bridge the gap that often exists between large but broad bioclimatic and biodiversity data and small but local data on biotic interactions.
 \cite{miller2019recent, isaac2020data}
@@
 Second, we accounted for several potential biases in the data observation process that may be common to many collections-based research questions by using a spatially-explicit random effects model. 
-Spatial autocorrelation \cite{willems2022forest}, potential biases introduced by the sampling habits of collectors \citep{daru2018widespread}, and variation between contemporary researchers during the collection of trait data, if not corrected for could lead to biased inference about the strength and direction of historic change.
+Spatial autocorrelation \cite{willems2022forest}, potential biases introduced by the sampling habits of collectors \citep{daru2018widespread}, and variation between contemporary researchers during the collection of trait data, if not corrected for could lead to over-confident inference about the strength and direction of historic change.
 Previous studies that have quantified the effects of collector biases typically find them to be small  \cite{davis2015herbarium,meineke2019herbarium}, and we similarly did not find that collector has a strong effect on the results of our analysis.

### 2024-04-24 endo_herbarium_spatiotemporal_analysis.R
updating model to fit for each species in a loop

@@ -17,7 +17,7 @@ library(tidyterra)
 
 library(patchwork)
 library(ggmap)
-library(ROCR)
+library(pROC)
 
 invlogit<-function(x){exp(x)/(1+exp(x))}
 species_colors <- c("#1b9e77","#d95f02","#7570b3")
@@ -39,8 +39,8 @@ endo_herb_georef <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/
                              spp_code == "AGPE" ~ "A. perennans",
                              spp_code == "ELVI" ~ "E. virginicus")) %>% 
   mutate(decade_bin = floor(year/10)*10) %>% 
-  mutate(std_year = (year-mean(year, na.rm = T))) # I am mean centering but not scaling by standard deviation to preserve units for interpretation of the parameter values
-
+  mutate(std_year = (year-mean(year, na.rm = T))) %>%  # I am mean centering but not scaling by standard deviation to preserve units for interpretation of the parameter values
+  filter(scorer_factor != "Scorer26")
 # mini_dataset <- endo_herb_georef %>% 
 #   filter(year>1970 &year<1990 & species == "A. hyemalis") %>% 
 #   mutate(presence = Endo_status_liberal) %>% 
@@ -125,7 +125,7 @@ collections_map <- ggplot()+
   theme(legend.text = element_text(face = "italic"))+
   labs(x = "Longitude", y = "Latitude", color = "Host Species")
 
-collections_map
+# collections_map
 # ggsave(collections_map, filename = "collections_map.png", width = 7, height = 4)
 
 endo_status_map <- ggplot()+
@@ -139,9 +139,23 @@ endo_status_map <- ggplot()+
   theme(strip.background = element_blank(),
         strip.text  = element_text(face = "italic", color = "black"))+
   labs(x = "Longitude", y = "Latitude", color = "Endophyte Status")
-endo_status_map
+# endo_status_map
 
 # ggsave(endo_status_map, filename = "endo_status_map.png", width = 10, height = 5)
+
+
+
+scorer_map <- ggplot()+
+  geom_map(data = outline_map, map = outline_map, aes(map_id = region), color = "grey", linewidth = .1, fill = "#FAF9F6")+
+  geom_map(data = states_shape, map = states_shape, aes(map_id = region), color = "grey", linewidth = .1, fill = NA)+
[…]

### 2024-04-24 endo_herbarium_records_merge.R
updating to look for new brit records, and trying to get prism data to line up with the new pixels across the map

@@ -106,7 +106,7 @@ AM_merged_records <- AM_symbiota_records %>%
 
 # BRIT digitized records downloaded from TORCH (includes Vanderbilt and U of Louisiana Monroe) 
 # This was downloaded on Jul17 and we can get more specimens transcribed and download again.
-AGHY_BRIT <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/DigitizedHerbariumRecords/BRIT_records/BRIT_AGHY_TORCH_records/SymbOutput_2023-02-03_192156_DwC-A/occurrences.csv",
+AGHY_BRIT <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/DigitizedHerbariumRecords/BRIT_records/BRIT_AGHY_TORCH_records/SymbOutput_2024-04-03_142217_DwC-A/occurrences.csv",
                       col_types = cols(id = col_double(),
                                        institutionCode = col_character(),
                                        collectionCode = col_logical(),
@@ -205,7 +205,7 @@ AGHY_BRIT <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/Digitiz
   dplyr::select(id, catalogNumber, country, stateProvince, county, municipality, locality, decimalLatitude, decimalLongitude, coordinateUncertaintyInMeters, recordedBy, recordNumber, eventDate, day, month, year) %>% 
   mutate(Spp_code = "AGHY")
 
-ELVI_BRIT <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/DigitizedHerbariumRecords/BRIT_records/BRIT_ELVI_TORCH_records/SymbOutput_2023-02-03_182842_DwC-A/occurrences.csv",
+ELVI_BRIT <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/DigitizedHerbariumRecords/BRIT_records/BRIT_ELVI_TORCH_records/SymbOutput_2024-04-03_141739_DwC-A/occurrences.csv",
                       col_types = cols(id = col_double(),
                                        institutionCode = col_character(),
                                        collectionCode = col_logical(),
@@ -301,7 +301,7 @@ ELVI_BRIT <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/Digitiz
   mutate(Spp_code = "ELVI")
 
 
[…]

### 2024-05-02 Manuscript/endo_herbarium.tex
Update endo_herbarium.tex

@@
 Mutualistic microbial symbionts may provide resilience to environmental change by protecting their hosts from increasing stress. 
-Alternatively, environmental change disrupts these interactions may lead to declines in hosts or symbionts.
-Microbes can be preserved within the dried plant tissue of herbarium specimens presenting an opportunity to quantify changes across broad temporal and spatial scales. 
-We ask how the prevalence of symbionts of grasses, \emph{Epichloë} fungal endophytes, which can protect hosts from drought, have changed over time in response to climate change, and how these changes vary across hosts' ranges.
-Specifically, we took seed samples from herbarium specimens of three grass host species collected over the last two centuries, quantified fungal hyphae presence within the seeds using microscopy, and evaluated spatial and temporal trends.
-Overall, endophytes have increased in prevalence over the last two centuries from 25\% prevalence to 75\% prevalence on average across the three hosts.
-We also found that changes in prevalence were associated with observed changes in annual and seasonal climate drivers corresponding with each hosts' peak growing season. 
-Thus we provide novel evidence for a cryptic biological response to climate change that may contribute to the resilience of the symbiosis through ongoing environmental change.
+Alternatively, environmental change that disrupts these interactions may lead to declines in hosts or symbionts. 
+Microbes preserved within herbarium specimens offer a unique opportunity to quantify changes in microbial symbiosis across broad temporal and spatial scales. 
+We asked how the prevalence of seed-transmitted fungal symbionts of grasses (\emph{Epichloë} endophytes), which can protect hosts from abiotic stress, have changed over time in response to climate change, and how these changes vary across host species' ranges.
+Specifically, we analyzed \tom{\#}{Give numbers} herbarium specimens of three grass host species collected over the last two centuries (18\#\# -- 20\#\#) for the presence or absence of endophyte symbiosis, and evaluated spatial and temporal trends in endophyte prevalence.
+We found that endophytes have increased in prevalence over the last two centuries from ca. 25\% prevalence to ca. 75\% prevalence, on average, across the three host species.
+We also found that changes in prevalence were associated with observed \tom{changes in annual and seasonal climate drivers}{Describe ``changes'' -- warming?
+drying?} corresponding to each host species' peak growing season.
+Our results provide novel evidence for a cryptic biological response to climate change that may contribute to the resilience of host-microbe symbiosis through context-dependent benefits that confer a fitness advantage to symbiotic hosts under environmental change.
 
@@
 
-Understanding how biotic interactions are altered by global climate change is a major goal of basic ecological research and conservation management \cite{gilman2010framework,blois2013climate}.
+Understanding how biotic interactions are altered by global change is a major goal of basic and applied ecological research \cite{gilman2010framework,blois2013climate}.
 Documented responses to environmental change, such as shifts in species' distributions \cite{aitken2008adaptation} and phenology \cite{piao2019plant}, are typically blind to concurrent changes in associated biotic interactions.
-Empirically evaluating these biotic changes, whether interacting species shift in tandem with their partners or not \cite{hillerislambers2013will}, is crucial to predicting future climate change responses.
-Such evaluations have been limited because data on species interactions are limited, with few datasets extending over the long time scales of contemporary climate change \cite{poisot2021global}.
+Empirically evaluating these biotic changes -- whether interacting species shift in tandem with their partners or not \cite{hillerislambers2013will} -- is crucial to predicting the reogranization of Earth's biodiversity under global change.
+Such evaluations have been limited because few datasets on species interactions extend over sufficiently long time scales of contemporary climate change \cite{poisot2021global}.
 
-Natural history collections, which were originally collected to study and preserve taxonomic diversity, present a unique opportunity to explore long-term changes in ecological interactions across broad spatial scales \citep{meineke2018unrealized}.
+Natural history specimens, which were originally collected to study and preserve taxonomic diversity, present a unique opportunity to explore long-term changes in ecological interactions across broad spatial and temporal scales \citep{meineke2018unrealized}.
 Natural history collections, built and maintained by the efforts of thousands of scientists, are invaluable time machines, primarily comprised of physical specimens of organisms along with information about the time and place of their collection.
-These specimens are samples preserving physical legacies of ecological processes and species' interactions from dynamically changing environments across time and space.
-Researchers have used plant collections to document climate change responses, including shifts in phenology \citep{willis2017old, park2019herbarium,  berg2019examination}, as well as in rates of pollination \citep{pauw2011reconstruction, duan2019century}, and herbivory \citep{meineke2019herbarium}.
-However, focus has been lacking on long-term changes in a particularly common type of biotic interaction, microbial symbioses.
+These specimens often preserve physical legacies of ecological processes and species' interactions from dynamically changing environments across time and space.
+For example, previous researchers have used plant collections (herbaria) to document shifts in phenology \citep{willis2017old, park2019herbarium,  berg2019examination}, pollination \citep{pauw2011reconstruction, duan2019century}, and herbivory \citep{meineke2019herbarium} related to anthropogenic climate change.
+However, few previous studies have leveraged biological collections to examine climate change-related shifts in a particularly common type of interaction: microbial symbiosis.
 
 Microbial symbionts are common to all macroscopic organisms and can have important effects on their hosts' survival, growth and reproduction \cite{rodriguez2009fungal,mcfall2013animals}.
-Many microbial symbionts act as mutualists, engaging in reciprocally beneficial interactions with their hosts that ameliorate environmental stress. 
+Many microbial symbionts act as mutualists, engaging in reciprocally beneficial interactions with their hosts that can ameliorate environmental stress. 
 For example, bacterial symbionts of insects, such as \emph{Wolbachia}, can improve their hosts' thermal tolerance \citep{truitt2019wolbachia, renoz2019evolutionary}, and arbuscular mycorrhizal fungi, documented in 70-90\% of families of land plants \citep{parniske2008arbuscular}, allow their hosts to persist through drought conditions by improving water and nutrient uptake \citep{cheng2021elucidating}.
@@
 Lichens exposed to elevated temperatures experienced loss of photosynthetic function along with changes in the composition of their algal symbiont community \citep{meyer2022climate}.
-How commonly and under what conditions microbial mutualisms deteriorate or strengthen under climate change remain unanswered questions, but previous work suggests that these alternative responses may depend on the intimacy and specialization of the interaction as well as the physiological tolerances of the mutualist partners \citep{toby2010mutualisms, warren2014mutualism, rafferty2015phenological}.
+How commonly and under what conditions microbial mutualisms deteriorate or strengthen under climate change remain unanswered questions.
+Previous work suggests that these alternative responses may depend on the intimacy and specialization of the interaction as well as the physiological tolerances of the mutualist partners \citep{toby2010mutualisms, warren2014mutualism, rafferty2015phenological}.
 
@@
 Additionally, symbionts are often unevenly distributed across their hosts' distribution.
-Facultative symbionts may be absent from portions of the host range \cite{afkhami2014mutualist}, and hosts may engage with a diversity of partners (different species, or different locally-adapted strains of one species) across their environments \cite{frade2008variation, rolshausen2018quantifying}.
+Facultative symbionts may be absent from portions of the host range \cite{afkhami2014mutualist}, and hosts may engage with a diversity of partners (different symbiont species or locally-adapted strains) across their environments \cite{frade2008variation, rolshausen2018quantifying}.
 Identifying broader spatial trends in symbiont prevalence is therefore an important step in developing predictions for where to expect changes in the symbiosis in future climates.
 
-\josh{
-\emph{Epichloë} fungal endophytes are specialized symbionts of cool-season grasses, which have been documented in $\sim 30$\% of cool-season grasses \citep{leuchtmann1992systematics}
-They are transmitted vertically from maternal plants to offspring through the grasses' seeds.
+\emph{Epichloë} fungal endophytes are specialized symbionts of cool-season grasses, which have been documented in $\sim 30$\% of cool-season grass species \citep{leuchtmann1992systematics}.
+They are transmitted vertically from maternal plants to offspring through seeds.
 Vertical transmission creates a feedback between the fitness of host and symbiont \citep{fine1975vectors, douglas1998host, rudgers2009fungus}. 
-Over time, endophytes that act as mutualists for their hosts should rise in prevalence within a population. 
-\emph{Epichloë} are known to improve their hosts' drought tolerance \cite{decunta2021systematic}, and protect their hosts against herbivores \cite{crawford2010fungal} and pathogens \cite{xia2018role} likely through the production of a diverse suite of alkaloids and other secondary metabolites.
+Over time, endophytes that act as mutualists should rise in prevalence within a host population \citep{donald2021context}. 
+\emph{Epichloë} are known to improve their hosts' drought tolerance \cite{decunta2021systematic} and protect their hosts against herbivores \cite{crawford2010fungal} and pathogens \cite{xia2018role} likely through the production of a diverse suite of alkaloids and other secondary metabolites.
 The fitness feedback induced by vertical transmission leads to the prediction that endophyte prevalence should be high in populations where these fitness benefits are most important. 
@@
 We predicted that prevalence should track temporal changes in environmental drivers that elicit these fitness benefits.
-For example, endophyte-mediated drought tolerance should lead prevalence to increase in regions where precipitation declines over time.
-Early research on \emph{Epichloë} used herbarium specimens to describe the broad taxonomic diversity of hosts \citep{white1985endophyte}. 
-Grasses are commonly identified and collected based on the presence of their reproductive structures meaning that collected specimens typically contain seeds, conveniently preserving the fungi along with their host plants on herbarium sheets.
+%For example, endophyte-mediated drought tolerance should lead endophyte prevalence to increase in regions where precipitation has declined over time.
+
+Early research on \emph{Epichloë} used herbarium specimens to describe the broad taxonomic diversity of host species that harbor these symbionts \citep{white1985endophyte}, \tom{establishing that signatures of endophyte symbiosis can be recovered from long-preserved plant tissue.}{Not sure about this but I am trying to pre-empt skepticism that fungi are less detectable in older samples.}
+However, no subsequent studies, to our knowledge, have used the vast resources of biological collections to quantitatively assess spatio-temporal trends in endophyte prevalence and their environmental correlates.
+Grasses are commonly collected and identified based on the presence of their reproductive structures, meaning that preserved specimens typically contain seeds, conveniently preserving the fungi along with their host plants on herbarium sheets.
 This creates the opportunity to leverage the unique spatio-temporal sampling of herbarium collections to examine the response of the symbiosis to historical climate change.
-}{Volker's suggestions included moving all material specific to study system into methods, but it seems a bit difficult to do.}
 
-In this study, we assessed for the first time the long-term responses of endophyte symbiosis to climate change through the use of herbarium specimens of three \emph{Epichloë} endophyte host species, \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}.
-We first address questions describing spatial and temporal trends in endophyte prevalence: (i) How has endophyte prevalence changed over time?
-and (ii) How spatially heterogenous are temporal trends in endophyte prevalence?
-We then address how climate change may be driving these trends in endophyte prevalence by asking: (iii) What is the relationship between variation in temporal trends in endophyte prevalence and changes in climate drivers?
-We examined 2,346 specimens collected across across eastern North America between 1824 and 2019.
+In this study, we assessed the long-term responses of endophyte symbiosis to climate change through the use of herbarium specimens of three North American host grass species (\emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}).
+We first address questions describing spatial and temporal trends in endophyte prevalence: (i) How has endophyte prevalence changed over the past two centuries?
+and (ii) How spatially heterogenous are temporal trends in endophyte prevalence across eastern North America?
+We then address how climate change may be driving trends in endophyte prevalence by asking: (iii) What is the relationship between variation in temporal trends in endophyte prevalence and changes in climate drivers?
+We predicted that aggregate endophyte prevalence would increase over time in tandem with climate warming, and that hotspots of endophyte change would correspond spatially to hotspots of climate change.
+To answer these questions we examined a total of 2,346 specimens collected across eastern North America between 1824 and 2019.
+\tom{}{I think the consensus was to keep the out-of-sample validation which should absolutely go into the Intro as an important element of novelty. Should go in the Asbtract too.}
 
@@
         \subsection*{Focal species}
-Our surveys focused on three endophyte hosts, \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}.
+Our surveys focused on three native North American grasses: \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}. 
 Both \emph{Agrostis} species host \emph{Epichloë amarillans} \cite{craven2001multigene, leuchtmann2014nomenclatural}, while \emph{Elymus virginicus} typically hosts \emph{Epichloë elymi} \cite{clay2002evolutionary}.
-These $C_3$ grass species are commonly represented in natural history collections with broad distributions covering much the Eastern United States.
-\emph{A. hyemalis} is a small short-lived perennial species that germinates in the spring and typically blooms between March and July (most common collection month: May).
-\emph{A. perennans} is of similar stature but is longer lived than \emph{Agrostis hyemalis} and blooms in late Summer and early Autumn (most common collection month: September). 
-This species is relatively more sparsely distributed, tending to be found in shadier and more moist habitats while \emph{A. hyemalis} is often found in open and recently disturbed ground. 
+These $C_3$ grass species are commonly represented in natural history collections with broad distributions covering much the eastern United States.
+\emph{A. hyemalis} is a small short-lived perennial species that germinates in the spring and typically flowers between March and July (most common collection month: May).
+\emph{A.
+perennans} is of similar stature but is longer lived than \emph{Agrostis hyemalis} and flowers in late \tom{Summer}{Are seasons capitalized?} and early Autumn (most common collection month: September).
+\emph{A. perennans} is more sparsely distributed, tending to be found in shadier and more moist habitats, while \emph{A. hyemalis} is commonly found in open and recently disturbed ground. 
 Both \emph{Agrostis} species are recorded from throughout the Eastern US, but \emph{A.
@@
 We visited nine herbaria between 2019 and 2022 (see Table A1 for a summary of specimens included from each collection). 
-In total, we quantified endophyte symbiosis for $1135$ \emph{A.
+We permission from herbarium staff, we acquired seed samples from $1135$ \emph{A.
 hyemalis} specimens collected between 1824 and 2019, $357$ \emph{A.
@@
 virginicus} specimens collected between 1839 and 2019 (Fig.
-1, Fig 2A).
-
-Our sampling plan was designed to minimize damage to these specimens.
+\ref{fig:map}, Fig \ref{fig:temporal}A).
+%Our sampling plan was designed to minimize damage to these specimens.
 We chose our focal species in part because they are commonly represented in herbarium collections, and produce high numbers of seeds, meaning that small samples would not diminish the value of the specimens for future studies.
@@
 Each specimen was assigned geographic coordinates based on collection information recorded on the herbarium sheet using the geocoding functionality of the ggmap R package \citep{kahle2019package}.
-Many specimens had digitized collection information readily available, but for those that did not, we transcribed information from pictures of the specimens. 
+Many specimens had digitized collection information readily available, but for those that did not, we transcribed information printed on the herbarium sheet. 
 Collections were geo-referenced to the nearest county centroid, or nearest municipality when that information was available. 
@@
 After collecting seed samples, we quantified the presence or absence of \emph{Epichloë} fungal hyphae, which grow intercellularly, in each specimen using microscopy. 
-We first softened seeds with a 10\% NaOH solution, then stained the seeds with aniline blue dye and squashed them under a microscope cover slip. 
-We examined the squashed seeds for the presence of fungal hyphae at 100X magnification \cite{bacon2018stains}.
-In some cases, the tissues examined during microscopy came from flowers or otherwise non-viable seeds, which were excluded from the seed counts for that specimen.
-On average we scored $4.7$ seeds per specimen in \emph{A. hyemalis}, $4.2$ seeds per specimen in \emph{A. perennans}, and $3.8$ seeds per specimen in \emph{E. virginicus}.
+We first softened seeds with a 10\% NaOH solution, then stained the seeds with \tom{aniline blue}{Most of our stains in the lab also have lactic acid.} dye and squashed them under a microscope cover slip.
+We examined the squashed seeds for the presence of fungal hyphae at 200-400X magnification \cite{bacon2018stains}.
+In some cases, the tissues examined during microscopy came from flowers or otherwise non-viable seeds, which were excluded for that specimen.
+On average we scored $4.7$ seeds per specimen of \emph{A.
+hyemalis}, $4.2$ seeds per specimen of \emph{A.
+perennans}, and $3.8$ seeds per specimen of \emph{E.
+virginicus}; we scored \# seeds in total.
+.
 Due to imperfect vertical transmission \cite{afkhami2008symbiosis}, it is possible that symbiotic host-plants produce a mixture of symbiotic and non-symbiotic seeds. 
-We therefore designated each specimen as symbiotic if endophyte hyphae were observed in one or more seeds, or non-symbiotic if hyphae were observed in zero seeds. 
-To capture uncertainty in the identification process, we recorded both a "liberal" and a "conservative" endophyte status for each plant.  
-When we identified potential endophytes with unusual morphology, low uptake of stain, or a small amount of fungal hyphae across the scored seeds, we recorded a positive liberal status (more likely to be a true endophyte) and a negative conservative status (less likely to be a true endophyte).
-$89$\% of scored plants had matching liberal and conservative endophyte statuses, reflecting high confidence in endophyte identification.
-The following analyses presented in the main text used the liberal statuses, but we repeated all analyses with the conservative statuses which yielded qualitatively similar results (Figure A5).
+We therefore designated a specimen as endophyte-symbiotic if \emph{Epichloë} hyphae were observed in one or more of its seeds, or non-symbiotic if hyphae were observed in none of its seeds. 
+To capture uncertainty in the endophyte scoring process, we recorded both a "liberal" and a "conservative" endophyte status for each \tom{plant}{Plant or seed?}.  
+When we identified potential endophytes with unusual morphology, low uptake of stain, or a small amount of fungal hyphae across the scored seeds, we recorded a positive liberal status (more likely to be endophyte-positive) and a negative conservative status (less likely to be endophyte-positive).
+$89$\% of scored plants had matching liberal and conservative scores, reflecting high confidence in endophyte status.
+The following analyses in the main text used the liberal status, but we repeated all analyses with the conservative status which yielded qualitatively similar results \tom{(Figure A5)}{Use \ref{} for citing figures, and stay consistent between Figure and Fig.}.
 
@@
 	\includegraphics[width = \linewidth]{collections_map.png}
-	\caption{\textbf{Collection locations of herbarium specimens sampled for endophyte presence absence}.
-Color designates host species ( \emph{A.
-hyemalis} (green), \emph{A.
-perennans} (orange), \emph{E.
-virginicus} (purple))}
+	\caption{\textbf{Collection locations of herbarium specimens of three grass host species across eastern North America that were sampled for \emph{Epichloë} endophyte presence or absence}.}
+	\label{fig:map}
 \end{figure}
 
-\subsection*{Assessing spatial and temporal changes in endophyte prevalence}
-We assessed spatial and temporal changes in endophyte prevalence across each host distribution, first quantifying the long-term overall trends and then examining how these trends vary across the spatial extent of each host distribution.
-We used an approximate Bayesian method, Integrated Nested Laplace Approximation (INLA), to construct spatio-temporal models of endophyte prevalence.
-INLA provides a computationally more efficient method of ascertaining model posteriors for certain models that can be formulated as latent Gausssian Models \cite{rue2009approximate}. 
+\subsection*{Modeling spatial and temporal changes in endophyte prevalence}
+We assessed spatial and temporal changes in endophyte prevalence across each host distribution, first quantifying the ``global'' trends, aggregating across space, and then examining spatial heterogeneity in the direction and magnitude of endophyte change (hotspots and coldspots) across the spatial extent of each host's distribution.
+\tom{To appropriately account for the spatial non-independence of geo-referenced occurrences we used an approximate Bayesian method, Integrated Nested Laplace Approximation (INLA), to construct spatio-temporal models of endophyte prevalence.
+INLA provides a computationally more efficient method of ascertaining parameter posterior distributions for certain models that can be formulated as latent Gausssian Models \cite{rue2009approximate}. 
 Many common statistical models, including structured and unstructured mixed-effects models, can be represented as latent Gaussian Models. 
-Fitting models with structured spatial effects is possible with MCMC sampling but can require long computation times, making INLA an effective alternative, which has been used to model spatial patterns in flowering phenology \cite{willems2022forest}, the abundance of bird species \cite{meehan2019spatial} and butterflies cite{crossley2022opposing}, the distribution of temperate trees \cite{engel2022spatial} as well as the population dynamics of endangered amphibians \cite{knapp2016large} and other ecological processes \cite{beguin2012hierarchical}.
+Fitting models with structured spatial effects is possible with MCMC sampling but can require long computation times, making INLA an effective alternative, which has been used to model spatial patterns in flowering phenology \cite{willems2022forest}, the abundance of bird species \cite{meehan2019spatial} and butterflies cite{crossley2022opposing}, the distribution of temperate trees \cite{engel2022spatial} as well as the population dynamics of endangered amphibians \cite{knapp2016large} and other ecological processes \cite{beguin2012hierarchical}.}{I think we need another sentence or two in this paragraph that provides an intuitive explanation for why INLA is the right approach for our analysis, written specifically for people who don't know what INLA is and don't want to know.
+I know the model includes a spatial decay term that controls spatial autocorrelation, a feature that I think is worth highlighting.}
 
-To quantify overall trends in endophyte prevalence, we modeled endophyte presence/absences ($P$)  as a Bernoulli response variable for each specimen $s$ at a given location $i$.
-In all cases, models for each species were fit separately.
-
+First, to quantify global trends in endophyte prevalence, we modeled endophyte presence/absences of the $i^{th}$ specimen ($P_i$) as a Bernoulli response variable with expected probability of endophyte occurrence $\hat{P_i}$.
+We modeled $\hat{P_i}$ as a linear function of collection year and accounting for random effects associated with location ($l[i]$, a unique latitude-longitude combination), collector identity ($c[i]$), and scorer identity ($s[i]$) of the $i^{th}$ specimen.
 \begin{subequations}
@@
 	\begin{align}
-		P_{s.i} \sim Bernoulli(\hat{P_{s,i}}) \\
-		logit(\hat{P}_{s,i}) = \alpha_{s,i} + \beta_{s}*year_i+
-		\chi_{i} + \omega_{i} 
+		logit(\hat{P}_{i}) = \alpha_{l[i]} + \beta*year_i + \chi_{c[i]} + \omega_{s[i]} 
 	\end{align}
 \end{subequations}
+Spatially-indexed random intercepts $\alpha_{l[i]}$ account for potential spatial autocorrelation between data points, and year slope $\beta$ describes the overall temporal trend in endophyte prevalence.
+We accounted for potential biases introduced during the process of collecting specimens as well as in scoring ability by including random effects specific to each collector $\chi$ and scorer $\omega$.
+Previous work suggests that behavior of historical botanists and uneven sampling may introduce biases into ecological inferences made from historic collections \cite{kozlov2020biases}. 
+Prolific collectors who contribute thousands of specimens may be more or less likely to collect certain species, or specimens with certain traits \cite{daru2018widespread}. 
+Similarly, the process of scoring seeds for hyphae involved many student researchers who, even with standardized training, may vary in their likelihood of positively identifying \emph{Epichloë} hyphae.
+By including a random effect for collectors and for scorers, we accounted for variance across individual researchers that may bias our predictions of changes in endophyte prevalence.
+\tom{Models for each host species were fit separately.}{It would be great to pull all species into one model and have them share variance terms for the random effects.
+I suspect such a model would give better and more stable estimates.}
+\tom{}{I updated the notation in ways that make more sense to me, but you should check that this is true to the actual model (I think it is).
+Also, a more complete presentation of this model would show the variance terms for $\alpha$, $\chi$, and $\omega$.
+I presume the latter two are Gaussian but I don't know how to represent the distribution of $\alpha$.}
 
-The expected endophyte prevalence, $\hat{P}$, was modelled with spatially-indexed random intercepts $\alpha$ that accounts for potential spatial autocorrelation between data points, and with a slope $\beta$ describing the overall change in endophyte prevalence.
-We accounted for potential biases introduced during the process of collecting specimens as well as in scoring ability by including random effects specific to each collector $\chi$ and to each scorer $\omega$.
-Previous work has questioned whether the behavior of historical botanists and uneven sampling may introduce biases into ecological inferences made from historic collections \cite{kozlov2020biases}. 
-Prolific collectors who contribute thousands of specimens may be more or less likely to collect certain species, or specimens with certain traits \cite{daru2018widespread}. 
-Similarly, the process of scoring seeds for hyphae involved many individual researchers, who may vary in their likelihood of positively identifying fungal hyphae. 
-By including a random effect for collectors and for scorers, we accounted for variance across individual researchers that may bias our predictions of changes in endophyte prevalence.
-
-To quantify how temporal trends may vary spatially, we repeated the modelling above, but instead incorporating a spatially-varying coefficient for collection year.
-
+Second, to quantify how temporal trends may vary spatially, we repeated the modelling above, but incorporated a spatially-varying coefficient for collection year:
 \begin{subequations}
@@
 	\begin{align}
-		P_{s.i} \sim Bernoulli(\hat{P_{s,i}}) \\
-		logit(\hat{P}_{s,i}) = \alpha_{s,i} + \tau_{s,i}*year_i +
-		\chi_{i} + \omega_{i} 
+		logit(\hat{P}_{i}) = \alpha_{l[i]} + \beta_{l[i]}*year_i + \chi_{c[i]} + \omega_{s[i]} 
 	\end{align}
 \end{subequations}
+The spatially-varying year slope $\beta_{l}$ allowed us to flexibly estimate variation in the temporal trajectory of endophyte change at locations across the study region.
 
-Note that the expected endophyte prevalence, $\hat{P}$, was modelled as above, and that $\tau$ describes the spatially-varying random slope for collection year.
-This spatially-varying random effect allowed us to flexibly estimate variation in slope at locations across the study region.
-All spatially-structured random intercepts and slopes were constructed using stochastic partial differential equations (SPDE) that depend on a covariance matrix according to the proximity of each collection location \citep{lindgren2011explicit,bakka2018spatial}.
-The covariance matrix was approximated using a Matérn covariance function, with each data point assigned a location according to the nodes of a mesh of non-overlapping triangles across our study area (Fig A2).
+\tom{For both models, spatially-structured random intercepts ($\alpha_{l}$) and slopes ($\beta_{l}$) were constructed using stochastic partial differential equations (SPDE) that depend on a covariance matrix according to the proximity of each collection location \citep{lindgren2011explicit,bakka2018spatial}.
+The covariance matrix was approximated using a Mat\'{e}rn covariance function, with each data point assigned a location according to the nodes of a mesh of non-overlapping triangles across our study area (Fig A2).}{This paragraph would be a place to describe the variance terms for the other random effects.}
 
-We performed model fitting using the inlabru R package \citep{}, with vague priors, and compared models with different sizes of mesh, which had little effect on the resulting model estimates.
-Each mesh was bounded by the predicted host distribution, described below.
-Posterior modes were stable indicating that numeric convergence was successful.
+We performed model fitting using the inlabru R package \citep{}, \tom{with vague priors}{I thought you needed informative priors on the spatial decay parameters}, and compared models with different sizes of mesh, \tom{which had little effect on the resulting model estimates}{That'sa good but you still need to state what mesh size you used and what that means, biologically.}.
+Each \tom{mesh}{You have not defined what you mean by ``mesh''.} was bounded by the predicted host distribution, described below.
+Posterior modes were \tom{stable}{Assessed how?} indicating that numeric convergence was successful.
 We assessed model fit with graphical posterior predictive checks (Fig. A3).
-The model performed adequately at classifying the historical data, comparing the accuracy of predictions from the model with observed data (avg. AUC = 0.77; Fig. A4).
+\tom{The model performed adequately at classifying the historical data, comparing the accuracy of predictions from the model with observed data (avg.
+AUC = 0.77; Fig.
+A4).
+}{Maybe move this to validation section, and then have both in-sample and out-of-sample approaches.}
 
@@
 \subsection*{Validating the model with an out-of-sample test}
-\josh{
-
-We evaluated the predictive ability of the model using contemporary endophyte surveys as out-of-sample test data, an important but rarely used strategy in ecological studies \cite{tredennick2021practical}.
-We used data from contemporary surveys of endophyte prevalence  in \emph{A. hyemalis} and \emph{E. virginicus}. 
-Surveys of \emph{E. virginicus} were conducted in 2013 as described in \citet{sneck2017variation}, and surveys of \emph{A. hyemalis} took place between 2015 and 2020.
+We evaluated the predictive ability of the model using contemporary endophyte surveys as out-of-sample test data, \tom{an important but rarely used strategy in ecological studies \cite{tredennick2021practical}.}{This is the type of thing to emphasize in the intro?
+Are there any other collections-based papers that have done anything like this??
+None to my knowledge.}
+We used data from contemporary surveys of endophyte prevalence  in \emph{A. hyemalis} and \emph{E. virginicus} in Texas and the southern US. 
+Surveys of \emph{E.
+virginicus} were conducted in 2013 as described in \citet{sneck2017variation}, and \tom{surveys of \emph{A.
+hyemalis} took place between 2015 and 2020}{We have added more recent AGHY survey data.
+I am not sure if you have access to this but you should definitely use it.
+Karl or I can point you to the right file.}.
 Population surveys of \emph{A.
@@
 virginicus} were designed to cover latitudinal variation along its range edge.
-In total, we visited 43 populations of \emph{A. hyemalis} and 20 populations of \emph{E. virginicus} across the central southeastern US, in Texas and neighboring states (Fig A4).
+In total, we visited 43 populations of \emph{A.
+hyemalis} and 20 populations of \emph{E.
+virginicus} across the south-central US, with emphasis on Texas and neighboring states (Fig \tom{A4}{This is now A6.
+Good reminder to use the ref function.}).
 During surveys, we collected seeds from up to 30 individuals per location  (average number of plants sampled: 22.9).
-We quantified the endophyte status of each individual with staining microscopy as described for the herbarium surveys, and calculated the prevalence of endophytes within the population (proportion of symbiotic plants divided by the number of sampled plants).
-The contempory survey period (2013-2020) is at the most recent edge of the time period encompassed by the historical observations used for model fitting.
-We compared the model's prediction for these locations to the observed population prevalence.
-}{considering dropping this section because the model does a good job at predicting the mean regional prevalenece, but has way less variability that the contemporary surveys}
+We quantified the endophyte status of each individual with staining microscopy as described for the herbarium surveys (with 5-10 seeds scored per individual), and calculated the prevalence of endophytes within the population (proportion of symbiotic plants divided by the number of sampled plants).
+For each population, we compared the observed fraction of endophyte-symbiotic hosts to the predicted probability of endophyte occurrence $\hat{P}$ derived from the model based on location and year, with collector and scorer random effects fixed at zero.
+\tom{The contemporary survey period (2013-2020) is at the most recent edge of the time period encompassed by the historical observations used for model fitting.
+We compared the model's prediction for these locations to the observed population prevalence.}{It is not clear if you are testing model 1 (``global trend'') or model 2 (``spatially varying trends'').}
+
 \subsection*{Assessing the role of climate drivers}
-
 We assessed how the magnitude of climate change may have driven changes in endophyte prevalence by assessing correlations between changes in climate and changes in endophyte prevalence predicted from our spatial model at evenly spaced pixels across the study area.
@@
 We used three four-month seasons within the year (Spring: January, February, March, April; Summer: May; June, July, August; Autumn: September, October, November, December). 
-This division of seasons allowed us to quantify differences in climate associated with the two "cool" seasons  that shoulder summer when we expect our focal species to be most biologically active (\emph{A.
-perennans}: Spring; \emph{E.
+This division of seasons allowed us to quantify differences in climate associated with the two ``cool'' seasons, when we expect our focal species to be most biologically active (\emph{A.
+hyemalis} flowering phenology: Spring; \emph{E.
 virginicus}: Spring and Summer; \emph{A.
 perennans}: Fall).
-In addition to mean climate conditions, environmental variability  in and of itself can influence population dynamics \cite{tuljapurkar_population_1982} and changes in variability are a key prediction of climate change models \cite{stocker2013technical, ipcc_2021}.
-So we calculated the coefficient of variation during each period for each annual and seasonal climate driver as the interannual standard deviation divided by the mean across each 30-year period.
-We then took the difference between recent and historic periods for the mean and coefficient of variation for each climate driver (Fig. A5).
+In addition to mean climate conditions, environmental variability itself can influence population dynamics \cite{tuljapurkar_population_1982} and changes in variability are a key prediction of climate change models \cite{stocker2013technical, ipcc_2021}.
+Therefore we calculated the coefficient of variation (CV) during each period for each annual and seasonal climate driver as the interannual standard deviation divided by the mean across each 30-year period.
+We then took the difference between recent and historic periods for the mean and CV for each climate driver \tom{(Fig.
+A5)}{This is Figure A7 -- Can you make the color scale on these diverging at zero?}.
 Because initial analyses indicated a high degree of collinearity between seasonal and annual changes in temperature, we used annual temperature only, along with annual and seasonal precipitation, in the subsequent analysis.
@@
 
-To evaluate whether areas that have experienced high degrees of change in climate have experienced the greatest changes in endophyte prevalence, we explored correlations between these climate drivers and the predicted spatially-varying trends with a Gaussian model:
-
-\begin{subequations}
-	\label{eq:trends}
-	\begin{align}
-		\tau_{i} \sim Normal(\hat{\mu}_{i}, \sigma^2) \\
-	\hat{\mu}_{i}= \beta *\mathbf{X}_{env}
-	\end{align}
-\end{subequations}
-
-where the trend at a given location ($\tau_{i}$) is predicted by our environmental covariates {$\mathbf{X}_{env}$} according to the vector of parameters $\beta$. 
-Calculating correlations from many pixels across the study region risks articially inflating confidence in our results due to large sample sizes, and so we repeated this calculation using only a random subsample of 100 pixels across the study region.
+To evaluate whether areas that have experienced the greatest changes in endophyte prevalence (hotspots of endophyte change) are associated with high degrees of change in climate (hotspots of climate change), \tom{we modeled spatially varying slopes of endophyte change through time ($\beta_{l}$) as a linear function of environmental covariates, with a Gaussian error distribution.
+}{I think we need to account for uncertainty in the slopes.
+They are outputs of a (quasi) Bayesian model so we should be able to propagate all the uncertainty in the posterior disgtribution.}
+Calculating correlations from many pixels across the study region risks articially inflating confidence in our results due to large sample sizes, and \tom{so we repeated this calculation using only a random subsample of 100 pixels across the study region}{100 seems like a low number to me.
+What if we did this for all of the herbarium collection locations?}.
+\tom{}{Are the methods above repeated for each species separately?}
+\tom{}{I cut the notation for the Gaussian model for now because it is a pretty simple model and the notation may be overkill, plus because I changed your tau's to beta's there were betas on both sides of the tilde, which was confusing/annoying.
+Happy have the notation back if you prefer it.
+I am also a little confused because the appendix has spearman correlations but there are no methods here for where those come from.}
 
@@
 Lines show predicted mean endophyte prevalence over the study period along with the 50\%  and 95\% CI bands.}
+	\label{fig:temporal}
 \end{figure}

### 2024-06-19 endo_herbarium_mesh_and_prior_sensitivity_analysis.R
Revert "Merge branch 'master' of https://github.com/joshuacfowler/EndoHerbarium"
This reverts commit 8b1c0b37bcf7c85b7a69a946a74865d3f98a98d5, reversing
changes made to e139f453dd38fc6c0376b65a01535fd934784294.

@@ -1,1558 +1,0 @@
-# Purpose: Fits spatiotemporal model of change in endophyte prevalence in herbarium specimens
-# Authors: Joshua Fowler and Tom Miller
-# Updated: Feb 22, 2024
-
-library(tidyverse) # for data manipulation and ggplot
-library(INLA) # for fitting integrated nested Laplace approximation models
-library(inlabru)
-library(fmesher)
-
-
-library(sf)
-# library(rmapshaper)
-library(terra)
-library(tidyterra)
-# library(ggpolypath)
-
-
-
-library(patchwork)
-library(ggmap)
-library(pROC)
-
-
-invlogit<-function(x){exp(x)/(1+exp(x))}
-species_colors <- c("#1b9e77","#d95f02","#7570b3")
-endophyte_colors <- c("#fdedd3","#f3c8a8", "#5a727b", "#4986c7", "#181914",  "#163381")
-
-
-################################################################################
-############ Read in the herbarium dataset ############################### 
-################################################################################
-
-endo_herb_georef <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/DigitizedHerbariumRecords/endo_herb_georef.csv") %>% 
-  # filter(Country != "Canada") %>%
-  mutate(sample_temp = Sample_id) %>% 
-  separate(sample_temp, into = c("Herb_code", "spp_code", "specimen_code", "tissue_code")) %>% 
-  mutate(species_index = as.factor(case_when(spp_code == "AGHY" ~ "1",
-                                             spp_code == "AGPE" ~ "2",
-                                             spp_code == "ELVI" ~ "3"))) %>% 
-  mutate(species = case_when(spp_code == "AGHY" ~ "A. hyemalis",
-                             spp_code == "AGPE" ~ "A. perennans",
-                             spp_code == "ELVI" ~ "E. virginicus")) %>% 
-  mutate(decade_bin = floor(year/10)*10) %>% 
-  mutate(std_year = (year-mean(year, na.rm = T))) %>%  # I am mean centering but not scaling by standard deviation to preserve units for interpretation of the parameter values
[…]

### 2024-06-19 endo_herbarium_multispecies_spatiotemporal_analysis.R

@@ -41,17 +41,6 @@ endo_herb_georef <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/
   mutate(decade_bin = floor(year/10)*10) %>% 
   mutate(std_year = (year-mean(year, na.rm = T))) %>%  # I am mean centering but not scaling by standard deviation to preserve units for interpretation of the parameter values
   filter(scorer_factor != "Scorer26")
-  
-
-
-
-# Creating herbariumd levels
-herbarium_levels <- levels(as.factor(endo_herb_georef$Herb_code))
-herbarium_no <- paste0("Herbarium",1:nlevels(as.factor(endo_herb_georef$Herb_code)))
-
-endo_herb_georef$herbarium_factor <- herbarium_no[match(as.factor(endo_herb_georef$Herb_code), herbarium_levels)]
-
-
 # mini_dataset <- endo_herb_georef %>% 
 #   filter(year>1970 &year<1990 & species == "A. hyemalis") %>% 
 #   mutate(presence = Endo_status_liberal) %>% 
@@ -71,9 +60,8 @@ endo_herb <- endo_herb_georef %>%
   filter(Country != "Canada" ) %>% 
   mutate(year_bin = case_when(year<1970 ~ "pre-1970",
                               year>=1970 ~ "post-1970")) %>% 
-  mutate(scorer_index = parse_number(scorer_factor),
-         herbarium_index = parse_number(herbarium_factor),
-         collector_index = parse_number(collector_factor)) %>% 
+  mutate(scorer_id = case_when(scorer_id == "BellaGuttierez" ~ "BellaGutierrez",
+                               TRUE ~ scorer_id)) %>% 
   mutate(endo_status_text = case_when(Endo_status_liberal == 0 ~ "E-",
                                       Endo_status_liberal == 1 ~ "E+"))  %>% 
   filter(spp_code %in% c("AGHY", "AGPE", "ELVI")) 
@@ -217,22 +205,7 @@ summary_endo_herb <- endo_herb %>%
             max_year = max(year))
 
 
-endo_herb_AGHY <- endo_herb %>% 
-  filter(!is.na(Endo_status_liberal)) %>% 
-  filter(spp_code == "AGHY") %>% 
-  filter(!is.na(lon) & !is.na(year))
 
-endo_herb_AGPE <- endo_herb %>% 
[…]

### 2024-06-19 Manuscript/endo_herbarium.tex
Added the SDM part

@@
 \newcommand{\tom}[2]{{\color{red}{#1}}\footnote{\textit{\color{red}{#2}}}}
+\newcommand{\jacob}[2]{{\color{blue}{#1}}\footnote{\textit{\color{blue}{#2}}}}
 \newcommand{\josh}[2]{{\color{orange}{#1}}\footnote{\textit{\color{orange}{#2}}}}
@@
 	\noindent{} 1. Rice University, Department of BioSciences, Houston, Texas 77006;
-	\noindent{} 1. University of Miami, Department of Biology, Miami, Florida;
+	\noindent{} \jacob{1}{I think this is should be 2}. University of Miami, Department of Biology, Miami, Florida;
 
@@
 drying?} corresponding to each host species' peak growing season.
-Our results provide novel evidence for a cryptic biological response to climate change that may contribute to the resilience of host-microbe symbiosis through context-dependent benefits that confer a fitness advantage to symbiotic hosts under environmental change.
+\jacob{Our results provide novel evidence for a cryptic biological response to climate change that may contribute to the resilience of host-microbe symbiosis through context-dependent benefits that confer a fitness advantage to symbiotic hosts under environmental change.}{I like this and the abstract in general.
+I agree with Tom and I think we  have some space to add these details.
+Abstract : 300}
 
@@
 \subsection*{Modeling distributions of host species}
-
-Jacob needs to write this, but we built SDMs using MaxEnt and the same climate predictors described below.
+We modeled  epicloe host species distribution to predict their occurrences in space and time as continuous and binary maps of potential presences. 
+These maps were used as a backbone to predict endophyte prevalence on epicloe host species. 
+The species distribution models were built following the ODMAP (overview, data, model, assessment, prediction) protocol \citep{crossley2022opposing}. 
+We used the observed presence of the host species collected from GBIF from 1990 to 2020.
+These occurrences were corrected for spatial autocorrelation due to sampling bias by thinning the occurrences to the spatial scale of the climatic variables.
+The climatic variables were temperature of the spring, precipitation of the spring  and precipitation of the summer.
+We preferred these variables because they were not correlated (Variance Inflation Factor \> 0.7) and also allowed the model to account for the influence of seasonal climate variation on species presence.
+The occurrence data was split into 75\% for model training and 25\% for model testing.
+We fitted the model using maximum entropy (MaxEnt) using the maxent function in the package dismo \citep{hijmans2017package}.
+MaxEnt was preferred because it does not generate response curves that may cause unpredictable behavior when applied to new climates \citep{hijmans2006ability}.
+ We used 10000 pseudo-absences as background points. 
+ To convert the continuous predicted probabilities into binary presence - absence maps, we used the  training sensitivity (true positive rate) and specificity threshold (true negative rate) \citep{liu2005selecting}.
+The performances of the model were evaluated using the AUC \citep{jimenez2012insights}.
 
@@
 Because initial analyses indicated a high degree of collinearity between seasonal and annual changes in temperature, we used annual temperature only, along with annual and seasonal precipitation, in the subsequent analysis.
-All together, this left us with measurements of change in 10 potential climate drivers: the mean and coefficient of variation of annual temperature, as well as the mean and coefficient of variation of cumulative annual precipitation, cumulative spring precipitation, cumulative summer precipitation, and cumulative autumn precipitation (Fig A8-A9).
+All together, this left us with measurements of change in 10 potential climate drivers: the mean and coefficient of variation of annual temperature, as well as the mean and coefficient of variation of cumulative annual precipitation, cumulative spring precipitation, cumulative summer precipitation, and cumulative autumn precipitation \jacob{(Fig A8-A9)}{ The species names are not clear  on Fig A9.
+I suggest increase the font siize }.
 

### 2024-06-20 endo_herbarium_multispecies_spatiotemporal_analysis.R
figuring out how to get polygons as mask for predictions from host SDMs

@@ -830,33 +830,77 @@ ggplot(samps)+
 ################################################################################################################################
 ##########  Plotting the spatially varying trends ###############
 ################################################################################################################################
+# Making a mask for each species based on Jacob's host SDMs
+aghy_raster <-terra::rast("aghy_binary.tif") 
+aghy_raster_distribution <- mask(aghy_raster, mask = aghy_raster == 1, maskvalue = 0)
+
+aghy_distribution_poly<-  terra::as.polygons(aghy_raster_distribution) 
+
+aghy_distribution <- aghy_distribution_poly   %>% 
+  st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE) %>% 
+  st_transform(epsg6703km)  
 
 
-vrt <- inlabru::fm_pixels(mesh, mask = bdry_polygon, format = "sp", dims = c(40,40))# Note this is where we can mask the output according the whatever shape, such as the host distribution
 
-vrt@data <- expand.grid(std_year = rep(1, length.out = length(vrt)),
+
+agpe_raster <- terra::rast("agpe_binary.tif")
+agpe_raster_distribution <- mask(agpe_raster, mask = agpe_raster == 1, maskvalue = 0)
+
+agpe_distribution_poly<-  terra::as.polygons(agpe_raster_distribution) 
+
+agpe_distribution <- agpe_distribution_poly   %>% 
+  st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE) %>% 
+  st_transform(epsg6703km)  
+
+
+elvi_raster <- terra::rast("elvi_binary.tif")
+elvi_raster_distribution <- mask(elvi_raster, mask = elvi_raster == 1, maskvalue = 0)
+
+elvi_distribution_poly<-  terra::as.polygons(elvi_raster_distribution) 
+
+elvi_distribution <- elvi_distribution_poly   %>% 
+  st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE) %>% 
+  st_transform(epsg6703km)  
+
+
[…]

### 2024-06-25 endo_herbarium_multispecies_spatiotemporal_analysis.R
re-adding the commits for multispecies model fit and getting shapes for prism

@@ -41,6 +41,17 @@ endo_herb_georef <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/
   mutate(decade_bin = floor(year/10)*10) %>% 
   mutate(std_year = (year-mean(year, na.rm = T))) %>%  # I am mean centering but not scaling by standard deviation to preserve units for interpretation of the parameter values
   filter(scorer_factor != "Scorer26")
+
+
+
+
+# Creating herbariumd levels
+herbarium_levels <- levels(as.factor(endo_herb_georef$Herb_code))
+herbarium_no <- paste0("Herbarium",1:nlevels(as.factor(endo_herb_georef$Herb_code)))
+
+endo_herb_georef$herbarium_factor <- herbarium_no[match(as.factor(endo_herb_georef$Herb_code), herbarium_levels)]
+
+
 # mini_dataset <- endo_herb_georef %>% 
 #   filter(year>1970 &year<1990 & species == "A. hyemalis") %>% 
 #   mutate(presence = Endo_status_liberal) %>% 
@@ -60,8 +71,9 @@ endo_herb <- endo_herb_georef %>%
   filter(Country != "Canada" ) %>% 
   mutate(year_bin = case_when(year<1970 ~ "pre-1970",
                               year>=1970 ~ "post-1970")) %>% 
-  mutate(scorer_id = case_when(scorer_id == "BellaGuttierez" ~ "BellaGutierrez",
-                               TRUE ~ scorer_id)) %>% 
+  mutate(scorer_index = parse_number(scorer_factor),
+         herbarium_index = parse_number(herbarium_factor),
+         collector_index = parse_number(collector_factor)) %>% 
   mutate(endo_status_text = case_when(Endo_status_liberal == 0 ~ "E-",
                                       Endo_status_liberal == 1 ~ "E+"))  %>% 
   filter(spp_code %in% c("AGHY", "AGPE", "ELVI")) 
@@ -107,7 +119,7 @@ contemp_surveys <- read_csv(file = "contemp_surveys.csv") %>%
     easting = st_coordinates(.)[, 1],
     northing = st_coordinates(.)[, 2]
   ) 
-  
+
 
 contemp_random_sample <- read_csv(file = "contemp_random_sample.csv") %>% 
   st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE) %>% 
[…]

### 2024-06-25 endo_herbarium_records_merge.R
getting the crs formatting right to get prism data at the pixels

@@ -1709,6 +1709,7 @@ ppt_autumn_cv_difference %>% st_transform(epsg6703km)
     coords.x2 = = st_coordinates(.)[, 1],
 )
 
+
 ######### Next extracting these climate values at our points. ####
 # we can extract them at the georeferenced coordinates and also across the grid of points we are using for prediction
 
@@ -1768,6 +1769,53 @@ bdry_polygon_list <- readRDS(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/
 
 
 # THese are the pixels across space for which we will extract climate
+
+vrt_aghy <- readRDS(file = "aghy_distribution_df.rds")
+vrt_agpe <- readRDS(file = "agpe_distribution_df.rds")
+vrt_elvi <- readRDS(file = "elvi_distribution_df.rds")
+
+
+dd_crs <- "GEOGCRS[\"NAD83\",\n    DATUM[\"North American Datum 1983\",\n        ELLIPSOID[\"GRS 1980\",6378137,298.257222101004,\n            LENGTHUNIT[\"metre\",1]]],\n    PRIMEM[\"Greenwich\",0,\n        ANGLEUNIT[\"degree\",0.0174532925199433]],\n    CS[ellipsoidal,2],\n        AXIS[\"geodetic latitude (Lat)\",north,\n            ORDER[1],\n            ANGLEUNIT[\"degree\",0.0174532925199433]],\n        AXIS[\"geodetic longitude (Lon)\",east,\n            ORDER[2],\n            ANGLEUNIT[\"degree\",0.0174532925199433]],\n    ID[\"EPSG\",4269]]"
+
+vrt_aghy_dd <- vrt_aghy  %>% 
+  spTransform(dd_crs)
+
+
+
+
+prism_diff_pred_df <- tibble(lon = vrt_aghy_dd@coords[,1], lat = vrt_aghy_dd@coords[,2],
+                             tmean_annual_diff = unlist(terra::extract(tmean_annual_difference, vrt_aghy_dd@coords)),
+                             tmean_spring_diff = unlist(terra::extract(tmean_spring_difference, vrt_aghy_dd@coords)),
+                             tmean_summer_diff = unlist(terra::extract(tmean_summer_difference, vrt_aghy_dd@coords)),
+                             tmean_autumn_diff = unlist(terra::extract(tmean_autumn_difference, vrt_aghy_dd@coords)),
+                             ppt_annual_diff = unlist(terra::extract(ppt_annual_difference, vrt_aghy_dd@coords)),
[…]

### 2024-06-25 endo_herbarium_multispecies_spatiotemporal_analysis.R
moving files into a separate folder

@@ -1,1460 +1,0 @@
-# Purpose: Fits spatiotemporal model of change in endophyte prevalence in herbarium specimens
-# Authors: Joshua Fowler and Tom Miller
-# Updated: Feb 22, 2024
-
-library(tidyverse) # for data manipulation and ggplot
-library(INLA) # for fitting integrated nested Laplace approximation models
-library(inlabru)
-library(fmesher)
-
-library(sf)
-# library(rmapshaper)
-library(terra)
-library(tidyterra)
-# library(ggpolypath)
-
-
-
-library(patchwork)
-library(ggmap)
-library(pROC)
-
-invlogit<-function(x){exp(x)/(1+exp(x))}
-species_colors <- c("#1b9e77","#d95f02","#7570b3")
-endophyte_colors <- c("#fdedd3","#f3c8a8", "#5a727b", "#4986c7", "#181914",  "#163381")
-
-
-################################################################################
-############ Read in the herbarium dataset ############################### 
-################################################################################
-
-endo_herb_georef <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/DigitizedHerbariumRecords/endo_herb_georef.csv") %>% 
-  # filter(Country != "Canada") %>%
-  mutate(sample_temp = Sample_id) %>% 
-  separate(sample_temp, into = c("Herb_code", "spp_code", "specimen_code", "tissue_code")) %>% 
-  mutate(species_index = as.factor(case_when(spp_code == "AGHY" ~ "1",
-                                             spp_code == "AGPE" ~ "2",
-                                             spp_code == "ELVI" ~ "3"))) %>% 
-  mutate(species = case_when(spp_code == "AGHY" ~ "A. hyemalis",
-                             spp_code == "AGPE" ~ "A. perennans",
-                             spp_code == "ELVI" ~ "E. virginicus")) %>% 
-  mutate(decade_bin = floor(year/10)*10) %>% 
-  mutate(std_year = (year-mean(year, na.rm = T))) %>%  # I am mean centering but not scaling by standard deviation to preserve units for interpretation of the parameter values
-  filter(scorer_factor != "Scorer26")
-
-
-
-
[…]

### 2024-06-25 endo_herbarium_records_merge.R

@@ -1,2042 +1,0 @@
-# Title: Digital Herbarium Records and Endophyte Sample Merging
-# Purpose: Imports and merges downloaded digitized herbarium records with the Endo_Herbarium database
-# Authors: Joshua Fowler, Mallory Tucker, Ella Segal, Lani Dufresne, and Tom Miller
-# Updated: Nov 4, 2022
-
-library(tidyverse) # for data manipulation and ggplot
-# library(slider) # add on to tidyverse to calculate sliding windows for climate data
-library(fuzzyjoin) # add on to tidyverse to merge tables on nearest values
-library(readxl)
-library(lubridate)
-library(ggmap)
-library(prism) # to import prism raster files
-library(terra)
-library(INLA) # using this here to work with the shape boundary
-
-################################################################################
-############ Read in digitized herbarium records ############################### 
-################################################################################
-
-# UT Austin downloaded from TORCH
-AGHY_UTAustin <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/DigitizedHerbariumRecords/UTAustinAGHYrecords/occurrences.csv") %>%
-  mutate(new_id = gsub("[a-zA-Z ]", "", catalogNumber)) %>%   #creates a new id that we will use to merge with the
-  mutate(municipality = as.character(municipality)) %>% 
-  dplyr::select(id, catalogNumber, country, stateProvince, county, municipality, locality, decimalLatitude, decimalLongitude, coordinateUncertaintyInMeters, recordedBy, recordNumber, new_id, eventDate, day, month, year) %>%
-  mutate(Spp_code = "AGHY")
-ELVI_UTAustin <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/DigitizedHerbariumRecords/UTAustinELVIrecords/occurrences.csv") %>% 
-  mutate(new_id = gsub("[a-zA-Z ]", "", catalogNumber)) %>%    #creates a new id that we will use to merge with the 
-  mutate(municipality = as.character(municipality)) %>% 
-  dplyr::select(id, catalogNumber, country, stateProvince, county, municipality, locality, decimalLatitude, decimalLongitude, coordinateUncertaintyInMeters, recordedBy, recordNumber, new_id, eventDate, day, month, year) %>% 
-  mutate(Spp_code = "ELVI")
[…]

### 2024-06-25 Analyses/endo_herbarium_multispecies_spatiotemporal_analysis.R
moving into a new folder, and I think I have the climate analysis working. Not sure what it means, but there ya go

@@ -1,0 +1,1491 @@
+# Purpose: Fits spatiotemporal model of change in endophyte prevalence in herbarium specimens
+# Authors: Joshua Fowler and Tom Miller
+# Updated: Feb 22, 2024
+
+library(tidyverse) # for data manipulation and ggplot
+library(INLA) # for fitting integrated nested Laplace approximation models
+library(inlabru)
+library(fmesher)
+
+library(sf)
+# library(rmapshaper)
+library(terra)
+library(tidyterra)
+# library(ggpolypath)
+
+
+
+library(patchwork)
+library(ggmap)
+library(pROC)
+
+invlogit<-function(x){exp(x)/(1+exp(x))}
+species_colors <- c("#1b9e77","#d95f02","#7570b3")
+endophyte_colors <- c("#fdedd3","#f3c8a8", "#5a727b", "#4986c7", "#181914",  "#163381")
+
+
+################################################################################
+############ Read in the herbarium dataset ############################### 
+################################################################################
+
+endo_herb_georef <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/DigitizedHerbariumRecords/endo_herb_georef.csv") %>% 
+  # filter(Country != "Canada") %>%
+  mutate(sample_temp = Sample_id) %>% 
+  separate(sample_temp, into = c("Herb_code", "spp_code", "specimen_code", "tissue_code")) %>% 
+  mutate(species_index = as.factor(case_when(spp_code == "AGHY" ~ "1",
+                                             spp_code == "AGPE" ~ "2",
+                                             spp_code == "ELVI" ~ "3"))) %>% 
+  mutate(species = case_when(spp_code == "AGHY" ~ "A. hyemalis",
+                             spp_code == "AGPE" ~ "A. perennans",
+                             spp_code == "ELVI" ~ "E. virginicus")) %>% 
+  mutate(decade_bin = floor(year/10)*10) %>% 
+  mutate(std_year = (year-mean(year, na.rm = T))) %>%  # I am mean centering but not scaling by standard deviation to preserve units for interpretation of the parameter values
+  filter(scorer_factor != "Scorer26")
+
+
+
+
[…]

### 2024-06-25 Analyses/endo_herbarium_records_merge.R

@@ -1,0 +1,2133 @@
+# Title: Digital Herbarium Records and Endophyte Sample Merging
+# Purpose: Imports and merges downloaded digitized herbarium records with the Endo_Herbarium database
+# Authors: Joshua Fowler, Mallory Tucker, Ella Segal, Lani Dufresne, and Tom Miller
+# Updated: Nov 4, 2022
+
+library(tidyverse) # for data manipulation and ggplot
+# library(slider) # add on to tidyverse to calculate sliding windows for climate data
+library(fuzzyjoin) # add on to tidyverse to merge tables on nearest values
+library(readxl)
+library(lubridate)
+library(ggmap)
+library(prism) # to import prism raster files
+library(terra)
+library(INLA) # using this here to work with the shape boundary
+
+################################################################################
+############ Read in digitized herbarium records ############################### 
+################################################################################
+
+# UT Austin downloaded from TORCH
+AGHY_UTAustin <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/DigitizedHerbariumRecords/UTAustinAGHYrecords/occurrences.csv") %>%
+  mutate(new_id = gsub("[a-zA-Z ]", "", catalogNumber)) %>%   #creates a new id that we will use to merge with the
+  mutate(municipality = as.character(municipality)) %>% 
+  dplyr::select(id, catalogNumber, country, stateProvince, county, municipality, locality, decimalLatitude, decimalLongitude, coordinateUncertaintyInMeters, recordedBy, recordNumber, new_id, eventDate, day, month, year) %>%
+  mutate(Spp_code = "AGHY")
+ELVI_UTAustin <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/DigitizedHerbariumRecords/UTAustinELVIrecords/occurrences.csv") %>% 
+  mutate(new_id = gsub("[a-zA-Z ]", "", catalogNumber)) %>%    #creates a new id that we will use to merge with the 
+  mutate(municipality = as.character(municipality)) %>% 
+  dplyr::select(id, catalogNumber, country, stateProvince, county, municipality, locality, decimalLatitude, decimalLongitude, coordinateUncertaintyInMeters, recordedBy, recordNumber, new_id, eventDate, day, month, year) %>% 
+  mutate(Spp_code = "ELVI")
[…]

### 2024-06-25 Analyses/posthoc_climate_correlation.R

@@ -1,0 +1,577 @@
+# Purpose: Perform post-hoc correlations between modelled spatially varying trends in endophyte prevalence and observed change in climate
+# Authors: Joshua Fowler
+# Updated: June 25, 2024
+
+library(tidyverse) # for data manipulation and ggplot
+library(INLA) # for fitting integrated nested Laplace approximation models
+library(inlabru)
+library(fmesher)
+
+library(prism) # to import prism raster files
+
+library(sf)
+# library(rmapshaper)
+library(terra)
+library(tidyterra)
+
+
+library(patchwork)
+library(ggmap)
+
+species_colors <- c("#1b9e77","#d95f02","#7570b3")
+species_names <- c("A. hyemalis", "A. perennans", "E. virginicus")
+
+
+################################################################################
+############ Download and read in prism data rasters ###########################
+################################################################################
+
+# making a folder to store prism data
+prism_set_dl_dir(paste0(getwd(),"/prism_download"))
+
+# getting monthly data for mean temp and precipitation
+# takes a long time the first time, but can skip when you have raster files saved on your computer.
+get_prism_monthlys(type = "tmean", years = 1895:2020, mon = 1:12, keepZip = FALSE)
+get_prism_monthlys(type = "ppt", years = 1895:2020, mon = 1:12, keepZip = FALSE)
+
+
+# pulling out values to get normals for old and new time periods
+tmean_annual_recent_norm <- terra::mean(terra::rast(pd_stack(prism_archive_subset(type = "tmean", temp_period = "monthly", year = 1990:2020))))
+tmean_spring_recent_norm <- terra::mean(terra::rast(pd_stack(prism_archive_subset(type = "tmean", temp_period = "monthly", year = 1990:2020, mon = 1:4))))
+tmean_summer_recent_norm <- terra::mean(terra::rast(pd_stack(prism_archive_subset(type = "tmean", temp_period = "monthly", year = 1990:2020, mon = 5:8))))
[…]

### 2024-06-26 Analyses/posthoc_climate_correlation.R
making a plot combining all marginal effects of climate drivers

@@ -325,6 +325,14 @@ elvi_prism_diff_pred_df <- tibble(lon = vrt_elvi_dd@coords[,1], lat = vrt_elvi_d
 
 write_csv(elvi_prism_diff_pred_df, file = "elvi_prism_diff_pred_df.csv")
 
+
+######### reading in and looking at the prism data from above ####
+
+aghy_prism_diff_pred_df <- read_csv("aghy_prism_diff_pred_df.csv")
+agpe_prism_diff_pred_df <- read_csv("agpe_prism_diff_pred_df.csv")
+elvi_prism_diff_pred_df <- read_csv("elvi_prism_diff_pred_df.csv")
+
+
 # dim(elvi_prism_diff_pred_df)
 
 ggplot()+
@@ -342,7 +350,8 @@ svc.pred_climate <- aghy_prism_diff_pred_df %>%
 svc.pred_climate_subsample <- svc.pred_climate %>% 
 group_by(species) %>%
 sample_n(size =200) %>% 
-  ungroup()
+  ungroup() %>% 
+  mutate(Intercept = 1)
 
 # ggplot(svc.pred_climate_subsample)+
 #   geom_point(aes(x = coords.x1, y = coords.x2, color = ppt_spring_sd_diff))+
@@ -351,7 +360,7 @@ sample_n(size =200) %>%
 
 svc.pred_climate_subsample_long <- svc.pred_climate_subsample %>% 
   pivot_longer(cols = contains("_diff")) %>% 
-  filter(!grepl("cv", name))
+  filter(!grepl("cv", name)) 
   # filter(!grepl("tmean_spring", name)) %>% 
   # filter(!grepl("tmean_summer", name)) %>% 
   # filter(!grepl("tmean_autumn", name)) 
@@ -362,7 +371,7 @@ ppt_regression_plot <- ggplot(filter(svc.pred_climate_subsample_long, grepl("ppt
   geom_point(aes(x = value, y = q0.5, color = species), alpha = .2)+
   geom_smooth(aes(x = value, y = q0.5, group = species, fill = species), color = "black", method = "lm")+
   scale_color_manual(values = species_colors) +
-  facet_wrap(~factor(name)+species, scales = "free", nrow = 3) +
+  facet_wrap(~species+ factor(name), scales = "free", nrow = 3) +
   theme(text = element_text(size = 8))+
   theme_light()+
[…]

### 2024-06-26 Analyses/endo_herbarium_multispecies_spatiotemporal_analysis.R
saving plots and cleaning up old stuff in script

@@ -150,7 +150,7 @@ collections_map <- ggplot()+
   labs(x = "Longitude", y = "Latitude", color = "Host Species")
 
 # collections_map
-# ggsave(collections_map, filename = "collections_map.png", width = 7, height = 4)
+ggsave(collections_map, filename = "Plots/collections_map.png", width = 7, height = 4)
 
 endo_status_map <- ggplot()+
   geom_map(data = outline_map, map = outline_map, aes( map_id = region), color = "grey", linewidth = .1, fill = "#FAF9F6")+
@@ -165,7 +165,7 @@ endo_status_map <- ggplot()+
   labs(x = "Longitude", y = "Latitude", color = "Endophyte Status")
 # endo_status_map
 
-# ggsave(endo_status_map, filename = "endo_status_map.png", width = 10, height = 5)
+ggsave(endo_status_map, filename = "Plots/endo_status_map.png", width = 10, height = 5)
 
 
 
@@ -296,13 +296,13 @@ mesh <- fm_mesh_2d_inla(
 # plot(mesh)
 
 
-ggplot() +
+mesh_plot <- ggplot() +
   gg(data = mesh) +
   geom_point(data = data, aes(x = easting, y = northing, col = species), size = 1) +
   coord_sf()+
   theme_bw() +
   labs(x = "", y = "")
-
+ggsave(mesh_plot, filename = "Plots/mesh_plot.png", width = 6, height = 6)
 
 # make spde
 # The priors from online tutorials are :   # P(practic.range < 0.05) = 0.01 # P(sigma > 1) = 0.01
@@ -393,167 +393,6 @@ fit$summary.fixed
 fit$summary.random
 
 
-
-
-
-
-df <- data.frame(year = 0, species_index= 1)
-
-space.int.post <- generate(svc.fit,df, formula = ~space.int, n.samples = 10)
-
-
-inla.posterior.sample(svc.fit)
-
-evaluate_model(svc.fit, state = list(species_index = 1))
-
-
-bc <- bincount(result = year.fit,
-               observations = endo_herb,
-               breaks = seq(0,max(coords[,2]), length = 12),
-               predictor = (Endo_status_liberal ~ invlogit(fixed_effects)))
-
-ipoints()
-
[…]

### 2024-06-26 Analyses/endo_herbarium_multispecies_spatiotemporal_analysis.R
making some formatting changes the overall trend plot and the collector and scorer plots

@@ -414,8 +414,8 @@ validation.pred1 <- predict(
   formula = ~ invlogit(int.species1 + year.species1 + 
                          space +  space.species1 + 
                          time.species1 + 
-                         herbarium_eval(herbarium_index) + scorer_eval(scorer_index) + collector_eval(scorer_index)),
-  n.samples = 500) 
+                         scorer_eval(scorer_index) + collector_eval(scorer_index)),
+  n.samples = 100) 
 
 validation.pred2 <- predict(
   fit,
@@ -423,8 +423,8 @@ validation.pred2 <- predict(
   formula = ~ invlogit(int.species2 + year.species2 + 
                          space + space.species3 + 
                          time.species2 + 
-                         herbarium_eval(herbarium_index) + scorer_eval(scorer_index) + collector_eval(scorer_index)),
-  n.samples = 500) 
+                         scorer_eval(scorer_index) + collector_eval(scorer_index)),
+  n.samples = 100) 
 
 validation.pred3 <- predict(
   fit,
@@ -432,19 +432,20 @@ validation.pred3 <- predict(
   formula = ~ invlogit(int.species3 + year.species3 +
                          space + space.species3 + 
                          time.species3 +
-                         herbarium_eval(herbarium_index) + scorer_eval(scorer_index) + collector_eval(scorer_index)),
-  n.samples = 500) 
+                         scorer_eval(scorer_index) + collector_eval(scorer_index)),
+  n.samples = 100) 
 
 validation.pred <- bind_rows(tibble(validation.pred1), tibble(validation.pred2), tibble(validation.pred3)) %>% 
   arrange(row_number)
 
 rocobj <- pROC::roc(endo_herb$Endo_status_liberal, validation.pred$mean)
 
-ggroc(rocobj) 
-
+ROC_training_plot <- ggroc(rocobj) 
+ggsave(ROC_training_plot, filename = "Plots/ROC_training_plot.png", width = 4, height = 4)
 
 # AUC values
 rocobj$auc
+# 0.7892
 
 
 # Taking alot of material from this blog post: https://inlabru-org.github.io/inlabru/articles/2d_lgcp_covars.html
@@ -502,6 +503,7 @@ year.pred.aghy <- predict(
   fit,
[…]

### 2024-06-26 Analyses/endo_herbarium_multispecies_spatiotemporal_analysis.R
updating the svc maps, and removing some old junk

@@ -150,7 +150,7 @@ collections_map <- ggplot()+
   labs(x = "Longitude", y = "Latitude", color = "Host Species")
 
 # collections_map
-ggsave(collections_map, filename = "Plots/collections_map.png", width = 7, height = 4)
+# ggsave(collections_map, filename = "Plots/collections_map.png", width = 7, height = 4)
 
 endo_status_map <- ggplot()+
   geom_map(data = outline_map, map = outline_map, aes( map_id = region), color = "grey", linewidth = .1, fill = "#FAF9F6")+
@@ -165,7 +165,7 @@ endo_status_map <- ggplot()+
   labs(x = "Longitude", y = "Latitude", color = "Endophyte Status")
 # endo_status_map
 
-ggsave(endo_status_map, filename = "Plots/endo_status_map.png", width = 10, height = 5)
+# ggsave(endo_status_map, filename = "Plots/endo_status_map.png", width = 10, height = 5)
 
 
 
@@ -302,7 +302,7 @@ mesh_plot <- ggplot() +
   coord_sf()+
   theme_bw() +
   labs(x = "", y = "")
-ggsave(mesh_plot, filename = "Plots/mesh_plot.png", width = 6, height = 6)
+# ggsave(mesh_plot, filename = "Plots/mesh_plot.png", width = 6, height = 6)
 
 # make spde
 # The priors from online tutorials are :   # P(practic.range < 0.05) = 0.01 # P(sigma > 1) = 0.01
@@ -757,6 +757,10 @@ saveRDS(svc.pred_aghy, file = "svc.pred_aghy.Rds")
 saveRDS(svc.pred_agpe, file = "svc.pred_agpe.Rds")
 saveRDS(svc.pred_elvi, file = "svc.pred_elvi.Rds")
 
+svc.pred_aghy <- readRDS(file = "svc.pred_aghy.Rds")
+svc.pred_agpe <- readRDS(file = "svc.pred_agpe.Rds")
+svc.pred_elvi <- readRDS(file = "svc.pred_elvi.Rds")
+
 
 
 # make a base map
@@ -789,7 +793,7 @@ svc_time_map_AGHY <- ggplot()+
   scale_fill_viridis_c(option = "turbo", na.value = "transparent", limits = trendrange)+
   labs(title = species_names[1], fill = "% change/year", y = "Latitude", x = "Longitude")+
   theme_light()+
-  theme(plot.title = element_text(face = "italic"))
[…]

### 2024-06-26 Analyses/endo_herbarium_multispecies_spatiotemporal_analysis.R
cleaning up, and saving the contemporary testing data plot

@@ -874,30 +874,6 @@ ggsave(svc_time_map.CI, filename = "Plots/svc_time_map.CI.png", width = 15, heig
 
 
 
-
-
-
-
-
-
-#########
-######### Trying to get weighted spatial posterior #####
-#########
-
-
-LambdaE <- predict(
-  svc.fit,
-  fm_int(mesh, bdry),
-  ~ sum(weight * space.int)
-)
-LambdaE
-
-
-
-icpt <- predict(svc.fit, NULL, ~ c(space_int= space.int_latent))
-plot(icpt)
-
-
 ################################################################################################################################
 ##########  Plotting the spatial Intercepts ###############
 ################################################################################################################################
@@ -1050,8 +1026,10 @@ contemp_lon <- ggplot(contemp.pred)+
   geom_linerange(aes(x = lon, ymin = `q0.025`, ymax = `q0.975`, color = species), alpha = .8)+
   geom_point(aes(x = lon, y = mean), shape = 4) + 
   scale_color_manual(values = c(species_colors[1], species_colors[3]))+
+  guides(color = "none")+
   ylim(0,1) + labs(y = "Endophyte Prevalance", x = "Longitude", color = "Species", size = "Sample Size")+
-  facet_wrap(~species) + theme_classic()
+  facet_wrap(~species) + theme_classic()+ theme(strip.background = element_blank(),
+                                                strip.text = element_text(size = rel(1.2)))
 # contemp_lon
 
 
@@ -1062,8 +1040,10 @@ contemp_lat <- ggplot(contemp.pred)+
   geom_linerange(aes(x = lat, ymin = `q0.025`, ymax = `q0.975`, color = species))+
   geom_point(aes(x = lat, y = mean), shape = 4) +
   scale_color_manual(values = c(species_colors[1], species_colors[3]))+
+  guides(color = "none")+
   ylim(0,1) + labs(y = "Endophyte Prevalance", x = "Latitude",  color = "Species", size = "Sample Size")+
-  facet_wrap(~species) + theme_classic()
+  facet_wrap(~species) + theme_classic() + theme(strip.background = element_blank(),
+                                                 strip.text = element_blank())
 # contemp_lat
 
 contemp_obspred <- ggplot(contemp.pred)+
@@ -1072,27 +1052,29 @@ contemp_obspred <- ggplot(contemp.pred)+
[…]

### 2024-06-26 Analyses/endo_herbarium_multispecies_spatiotemporal_analysis.R
cleaning up, and making map for contemp surveys

@@ -1078,126 +1078,25 @@ rocobj$auc
 
 
 
-##### Post-hoc correlations with climate drivers #####
-
-# Read in the saved rasters which have the calculated change in climate.
-
-# reading in the change in climate normals (1895-1925 and 1990-2020)
-
-
-write_csv(elvi_prism_diff_pred_df, file = "elvi_prism_diff_pred_df.csv")
 
 
 
-aghy_prism_diff_pred_df <- read_csv(file = "aghy_prism_diff_pred_df.csv") 
-agpe_prism_diff_pred_df <- read_csv(file = "agpe_prism_diff_pred_df.csv") 
-elvi_prism_diff_pred_df <- read_csv(file = "elvi_prism_diff_pred_df.csv") 
 
 
-# mutate(across(contains("ppt_"), ~.x*.1)) # changing the units of the change in precip. to be 10ths of millimeters
-# prism_for_plotting <- prism_diff_pred_df %>% 
-#   filter(tmean_annual_cv_diff <25 & tmean_annual_cv_diff>-2)  
 
 
-ggplot()+
-  gg(elvi_prism_diff_pred_df, aes(color = tmean_annual_cv_diff))
-
-ggplot()+
-  gg(svc.pred_elvi, aes(color = mean))
-# Merging the climate date with our predicted temporal slopes
-
-points <- prism_diff_pred_df %>%
-  dplyr::select(lon, lat) 
-
-
-# getting the pixel values of slopes as a dataframe
-svc.pred_aghy <- readRDS( file = "svc.pred_aghy.Rds")
-svc.pred_agpe <- readRDS( file = "svc.pred_agpe.Rds")
-svc.pred_elvi <- readRDS( file = "svc.pred_elvi.Rds")
-
-svc.pred_aghy_df <- svc.pred_aghy %>% 
-  st_as_sf(coords = c("lon", "lat"), crs = epsg6703km, remove = FALSE) %>% 
-  mutate(
-    coords.x1 = st_coordinates(.)[, 1],
-    coords.x2 = st_coordinates(.)[, 2]
-  ) %>% 
-  as.data.frame() %>% 
-  mutate(origin = "svc")
-
-
-svc.pred_aghy_merged <- svc.pred_aghy_df %>% 
-  full_join(aghy_prism_diff_pred_df, by = c("geometry"))
-
-
-trend_df <- svc.pred_aghy_merged
-
[…]

### 2024-06-26 Manuscript/endo_herbarium.tex
updates to the manuscript, incorporating intro comments and equation.

@@
 Mutualistic microbial symbionts may provide resilience to environmental change by protecting their hosts from increasing stress. 
-Alternatively, environmental change that disrupts these interactions may lead to declines in hosts or symbionts. 
+However, environmental change that disrupts these interactions may lead to declines in hosts or symbionts. 
 Microbes preserved within herbarium specimens offer a unique opportunity to quantify changes in microbial symbiosis across broad temporal and spatial scales. 
 We asked how the prevalence of seed-transmitted fungal symbionts of grasses (\emph{Epichloë} endophytes), which can protect hosts from abiotic stress, have changed over time in response to climate change, and how these changes vary across host species' ranges.
-Specifically, we analyzed \tom{\#}{Give numbers} herbarium specimens of three grass host species collected over the last two centuries (18\#\# -- 20\#\#) for the presence or absence of endophyte symbiosis, and evaluated spatial and temporal trends in endophyte prevalence.
+Specifically, we analyzed 2,346 herbarium specimens of three grass host species collected over the last two centuries (1824 -- 2019) for the presence or absence of endophyte symbiosis, and evaluated spatial and temporal trends in endophyte prevalence.
 We found that endophytes have increased in prevalence over the last two centuries from ca. 25\% prevalence to ca. 75\% prevalence, on average, across the three host species.
-We also found that changes in prevalence were associated with observed \tom{changes in annual and seasonal climate drivers}{Describe ``changes'' -- warming?
+We also found that changes in prevalence were associated with observed \tom{changes in seasonal climate drivers}{Describe ``changes'' -- warming?
 drying?} corresponding to each host species' peak growing season.
@@
 For example, bacterial symbionts of insects, such as \emph{Wolbachia}, can improve their hosts' thermal tolerance \citep{truitt2019wolbachia, renoz2019evolutionary}, and arbuscular mycorrhizal fungi, documented in 70-90\% of families of land plants \citep{parniske2008arbuscular}, allow their hosts to persist through drought conditions by improving water and nutrient uptake \citep{cheng2021elucidating}.
-On the other hand, changes in the mean and variance of environmental conditions may disrupt microbial  mutualisms by changing the costs and benefits of the interaction for each partner, leading the interaction to deteriorate \citep{aslan2013mutualism}.
+On the other hand, changes in the mean and variance of environmental conditions may disrupt microbial  mutualisms by changing the costs and benefits of the interaction for each partner, leading the interaction to deteriorate \citep{aslan2013mutualism, fowler2024microbial}.
 Coral bleaching (the loss of symbiotic algae) due to temperature stress \citep{sully2019global} is perhaps the best known example, but this phenomenon is not unique to corals.
@@
 
-Early research on \emph{Epichloë} used herbarium specimens to describe the broad taxonomic diversity of host species that harbor these symbionts \citep{white1985endophyte}, \tom{establishing that signatures of endophyte symbiosis can be recovered from long-preserved plant tissue.}{Not sure about this but I am trying to pre-empt skepticism that fungi are less detectable in older samples.}
+Early research on \emph{Epichloë} used herbarium specimens to describe the broad taxonomic diversity of host species that harbor these symbionts \citep{white1985endophyte}, \josh{establishing that endophyte symbiosis could be identified in plant tissue from as early as 1851.}{Edited this a bit.
+This is the earliest year we have in the database that was part of JFWhites original paper.}
 However, no subsequent studies, to our knowledge, have used the vast resources of biological collections to quantitatively assess spatio-temporal trends in endophyte prevalence and their environmental correlates.
+Biological collections have been used to demonstrate ecological signatures of climate change, however the predictive ability of these historical analyses is rarely tested. 
 Grasses are commonly collected and identified based on the presence of their reproductive structures, meaning that preserved specimens typically contain seeds, conveniently preserving the fungi along with their host plants on herbarium sheets.
@@
 We predicted that aggregate endophyte prevalence would increase over time in tandem with climate warming, and that hotspots of endophyte change would correspond spatially to hotspots of climate change.
+Finally, we evaluated the performance models built on data from historic specimens with an out-of-sample test, data on endophyte prevalence from contemporary surveys of host populations. 
 To answer these questions we examined a total of 2,346 specimens collected across eastern North America between 1824 and 2019.
@@
 \emph{A. hyemalis} is a small short-lived perennial species that germinates in the spring and typically flowers between March and July (most common collection month: May).
-\emph{A.
-perennans} is of similar stature but is longer lived than \emph{Agrostis hyemalis} and flowers in late \tom{Summer}{Are seasons capitalized?} and early Autumn (most common collection month: September).
+\emph{A. perennans} is of similar stature but is longer lived than \emph{Agrostis hyemalis} and flowers in late summer and early autumn (most common collection month: September). 
 \emph{A. perennans} is more sparsely distributed, tending to be found in shadier and more moist habitats, while \emph{A. hyemalis} is commonly found in open and recently disturbed ground. 
@@
 \emph{E. virginicus} is a larger and relatively longer-lived  species that is more broadly distributed than the \emph{Agrostis} species. 
-It begins flowering as early as March or April but continues throughout the Summer (most common collection month: July).
+It begins flowering as early as March or April but continues throughout the summer (most common collection month: July).
 
@@
 We visited nine herbaria between 2019 and 2022 (see Table A1 for a summary of specimens included from each collection). 
-We permission from herbarium staff, we acquired seed samples from $1135$ \emph{A.
+With permission from herbarium staff, we acquired seed samples from $1135$ \emph{A.
 hyemalis} specimens collected between 1824 and 2019, $357$ \emph{A.
@@
 After collecting seed samples, we quantified the presence or absence of \emph{Epichloë} fungal hyphae, which grow intercellularly, in each specimen using microscopy. 
-We first softened seeds with a 10\% NaOH solution, then stained the seeds with \tom{aniline blue}{Most of our stains in the lab also have lactic acid.} dye and squashed them under a microscope cover slip.
+We first softened seeds with a 10\% NaOH solution, then stained the seeds with aniline blue-lactic acid stain and squashed them under a microscope cover slip. 
 We examined the squashed seeds for the presence of fungal hyphae at 200-400X magnification \cite{bacon2018stains}.
@@
 We therefore designated a specimen as endophyte-symbiotic if \emph{Epichloë} hyphae were observed in one or more of its seeds, or non-symbiotic if hyphae were observed in none of its seeds. 
-To capture uncertainty in the endophyte scoring process, we recorded both a "liberal" and a "conservative" endophyte status for each \tom{plant}{Plant or seed?}.  
+To capture uncertainty in the endophyte scoring process, we recorded both a "liberal" and a "conservative" endophyte status for each plant specimen.  
 When we identified potential endophytes with unusual morphology, low uptake of stain, or a small amount of fungal hyphae across the scored seeds, we recorded a positive liberal status (more likely to be endophyte-positive) and a negative conservative status (less likely to be endophyte-positive).
 $89$\% of scored plants had matching liberal and conservative scores, reflecting high confidence in endophyte status.
-The following analyses in the main text used the liberal status, but we repeated all analyses with the conservative status which yielded qualitatively similar results \tom{(Figure A5)}{Use \ref{} for citing figures, and stay consistent between Figure and Fig.}.
+The following analyses in the main text used the liberal status, but we repeated all analyses with the conservative status which yielded qualitatively similar results (Fig. \ref{fig:FigA5})
 
@@
 We assessed spatial and temporal changes in endophyte prevalence across each host distribution, first quantifying the ``global'' trends, aggregating across space, and then examining spatial heterogeneity in the direction and magnitude of endophyte change (hotspots and coldspots) across the spatial extent of each host's distribution.
-\tom{To appropriately account for the spatial non-independence of geo-referenced occurrences we used an approximate Bayesian method, Integrated Nested Laplace Approximation (INLA), to construct spatio-temporal models of endophyte prevalence.
-INLA provides a computationally more efficient method of ascertaining parameter posterior distributions for certain models that can be formulated as latent Gausssian Models \cite{rue2009approximate}. 
+To appropriately account for the spatial non-independence of geo-referenced occurrences, we used an approximate Bayesian method, Integrated Nested Laplace Approximation (INLA), to construct spatio-temporal models of endophyte prevalence.
+INLA provides a computationally efficient method of ascertaining parameter posterior distributions for certain models that can be formulated as latent Gausssian Models \cite{rue2009approximate}.
 Many common statistical models, including structured and unstructured mixed-effects models, can be represented as latent Gaussian Models.
-Fitting models with structured spatial effects is possible with MCMC sampling but can require long computation times, making INLA an effective alternative, which has been used to model spatial patterns in flowering phenology \cite{willems2022forest}, the abundance of bird species \cite{meehan2019spatial} and butterflies cite{crossley2022opposing}, the distribution of temperate trees \cite{engel2022spatial} as well as the population dynamics of endangered amphibians \cite{knapp2016large} and other ecological processes \cite{beguin2012hierarchical}.}{I think we need another sentence or two in this paragraph that provides an intuitive explanation for why INLA is the right approach for our analysis, written specifically for people who don't know what INLA is and don't want to know.
-I know the model includes a spatial decay term that controls spatial autocorrelation, a feature that I think is worth highlighting.}
+We incorporated spatial heterogeneity into this analysis using spatially-structured intercept and slope parameters implemented as stochastic partial differential equation (SPDE) approximations of a continuous spatial Guassian process.
+This SPDE approach is a flexible method of smoothing across space while explicitly accounting for spatial dependence between data-points \citep{lindgren2011explicit,bakka2018spatial}.
+Fitting models with structured spatial effects is possible with MCMC sampling but can require long computation times, making INLA an effective alternative, which has been used to model spatial patterns in flowering phenology \cite{willems2022forest}, the abundance of bird species \cite{meehan2019spatial} and butterflies \cite{crossley2022opposing}, the distribution of temperate trees \cite{engel2022spatial} as well as the population dynamics of endangered amphibians \cite{knapp2016large} and other ecological processes \cite{beguin2012hierarchical}.
+
+We estimated global and spatially-varying trends in endophyte prevalence using a joint-likelihood model. 
+For each host species $h$, endophyte presence/absence of the $i^{th}$ specimen ($P_{[h]i}$) was modeled a Bernoulli response variable with expected probability of endophyte occurrence $\hat{P}_{[h]i}$.
+We modeled
+
+\begin{subequations}
+	\label{eq:trends}
+	\begin{align}
+		logit(\hat{P}_{[h_{1}]i}) =  \mathrm{A}_{[h_{1}]i} + \mathrm{T}_{[h_{1}]}*year_i  + \alpha_{[h_{1}]l[i]} + \tau_{[h_{1}]l[i]}*year_i  + \delta_{l[i]}+ \chi_{c[i]} + \omega_{s[i]} \\
+		logit(\hat{P}_{[h_{2}]i}) = \mathrm{A}_{[h_{2}]i} + \mathrm{T}_{[h_{2}]}*year_i  + \alpha_{[h_{2}]l[i]} + \tau_{[h_{2}]l[i]}*year_i  + \delta_{l[i]}+ \chi_{c[i]} + \omega_{s[i]} \\
+		logit(\hat{P}_{[h_{3}]i}) = \mathrm{A}_{[h_{2}]i} + \mathrm{T}_{[h_{2}]}*year_i  + \alpha_{[h_{2}]l[i]} + \tau_{[h_{3}]l[i]}*year_i  + \delta_{l[i]}+ \chi_{c[i]} + \omega_{s[i]}
+	\end{align}
+\end{subequations}
 
@@
 Prolific collectors who contribute thousands of specimens may be more or less likely to collect certain species, or specimens with certain traits \cite{daru2018widespread}. 
-Similarly, the process of scoring seeds for hyphae involved many student researchers who, even with standardized training, may vary in their likelihood of positively identifying \emph{Epichloë} hyphae.
-By including a random effect for collectors and for scorers, we accounted for variance across individual researchers that may bias our predictions of changes in endophyte prevalence.
+Similarly, the process of scoring seeds for hyphae involved several student researchers who, even with standardized training, may vary in their likelihood of positively identifying \emph{Epichloë} hyphae.
+By including a random effect for collectors and for scorers, we attempted to accound for variance across individual researchers that may bias our predictions of changes in endophyte prevalence.
 \tom{Models for each host species were fit separately.}{It would be great to pull all species into one model and have them share variance terms for the random effects.
@@
 The climatic variables were temperature of the spring, precipitation of the spring  and precipitation of the summer.
-We preferred these variables because they were not correlated (Variance Inflation Factor \> 0.7) and also allowed the model to account for the influence of seasonal climate variation on species presence.
+We preferred these variables because they were not correlated (Variance Inflation Factor $>$ 0.7) and also allowed the model to account for the influence of seasonal climate variation on species presence.
 The occurrence data was split into 75\% for model training and 25\% for model testing.
@@
 For example, 3 of the 16 scorers were more likely than average to assign positive endophyte status, as indicated by 95\% credible intervals that do not overlap 0) (Fig 4B). 
-However, this may have been driver by differences in scorers biases during the seed scoring process, or by unintended spatial clustering of the specimens scored by each scorer. 
+However, this may have been driven by differences in scorers biases during the seed scoring process, or by unintended spatial clustering of the specimens scored by each scorer. 
 Interpreting our models with the inclusion of the scorer effect thus provides conservative estimates of the absolute magnitude of changes in endophyte prevalence.
@@
 Incorporating available climatic and soil layers as covariates is an obvious first step that could improve predictions.
-Towards the goal of predicting the dynamics of microbial symbioses under climate change, models that integrate data from local and regional scales would be an important step to bridge the gap that often exists between large but broad bioclimatic and biodiversity data and small but local data on biotic interactions.
-\cite{miller2019recent, isaac2020data}
+Towards the goal of predicting the dynamics of microbial symbioses under climate change, models that integrate data from local and regional scales would be an important step to bridge the gap that often exists between large but broad bioclimatic and biodiversity data and small but local data on biotic interactions \cite{miller2019recent, isaac2020data}.
 
@@
 Asteriks denote correlation coefficients $> .3$ or $< -.3$.}
+	\label{fig:FigA5}
 \end{figure}

### 2024-06-27 Manuscript/endo_herbarium.tex
updating equations and paths for plots

@@
 \usepackage{amsmath}
+
+\usepackage[T1]{fontenc}
 
@@
 drying?} corresponding to each host species' peak growing season.
+Our analysis performed favorably in an out-of-sample predictive test, however we identified XXX as suggesting the model fusion may be an important step moving forward.
 \jacob{Our results provide novel evidence for a cryptic biological response to climate change that may contribute to the resilience of host-microbe symbiosis through context-dependent benefits that confer a fitness advantage to symbiotic hosts under environmental change.}{I like this and the abstract in general.
@@
 The fitness feedback induced by vertical transmission leads to the prediction that endophyte prevalence should be high in populations where these fitness benefits are most important. 
-Previous survey studies have documented large-scale spatial patterns in  endophyte prevalence structured by environmental gradients \citep{granath2007variation,bazely2007broad, afkhami2012fungal,sneck2017variation}.
+Previous contemporary survey studies have documented large-scale spatial patterns in  endophyte prevalence structured by environmental gradients \citep{granath2007variation,bazely2007broad, afkhami2012fungal,sneck2017variation}.
 We predicted that prevalence should track temporal changes in environmental drivers that elicit these fitness benefits.
@@
 However, no subsequent studies, to our knowledge, have used the vast resources of biological collections to quantitatively assess spatio-temporal trends in endophyte prevalence and their environmental correlates.
-Biological collections have been used to demonstrate ecological signatures of climate change, however the predictive ability of these historical analyses is rarely tested. 
 Grasses are commonly collected and identified based on the presence of their reproductive structures, meaning that preserved specimens typically contain seeds, conveniently preserving the fungi along with their host plants on herbarium sheets.
 This creates the opportunity to leverage the unique spatio-temporal sampling of herbarium collections to examine the response of the symbiosis to historical climate change. 
+\josh{Research using historical collections has clearly demonstrated other ecological signatures of a changing climate. 
+However the predictive ability of these historical analyses is rarely tested against contemporary data \citep{lee2024phenological}. 
+Identifying the ways in which these analyses fall short is a crucial step for the field move from reading signatures in the past to forecasting ecological dynamics into the future.}{what do you think of this?
+trying to presage the out-of-sample test without over promising and without saying outright that our analysis sucks.
+Is this the right place for this?
+I had imagined some of this material will be really developed in the discussion.}
 
@@
 We predicted that aggregate endophyte prevalence would increase over time in tandem with climate warming, and that hotspots of endophyte change would correspond spatially to hotspots of climate change.
-Finally, we evaluated the performance models built on data from historic specimens with an out-of-sample test, data on endophyte prevalence from contemporary surveys of host populations. 
+Finally, we evaluated the performance of models built on data from historic specimens with an out-of-sample test, data on endophyte prevalence from contemporary surveys of host populations. 
 To answer these questions we examined a total of 2,346 specimens collected across eastern North America between 1824 and 2019.
@@
 virginicus} specimens collected between 1839 and 2019 (Fig.
-\ref{fig:map}, Fig \ref{fig:temporal}A).
+\ref{fig:map}, Fig.
+\ref{fig:temporal}A, Fig.
+\ref{fig:endo_status_map}).
 %Our sampling plan was designed to minimize damage to these specimens.
@@
 	\centering
-	\includegraphics[width = \linewidth]{collections_map.png}
+	\includegraphics[width = \linewidth]{../Plots/collections_map.png}
 	\caption{\textbf{Collection locations of herbarium specimens of three grass host species across eastern North America that were sampled for \emph{Epichloë} endophyte presence or absence}.}
@@
 We estimated global and spatially-varying trends in endophyte prevalence using a joint-likelihood model. 
-For each host species $h$, endophyte presence/absence of the $i^{th}$ specimen ($P_{[h]i}$) was modeled a Bernoulli response variable with expected probability of endophyte occurrence $\hat{P}_{[h]i}$.
-We modeled
+For each host species $h$, endophyte presence/absence of the $i^{th}$ specimen ($P_{[h]i}$) was modeled as a Bernoulli response variable with expected probability of endophyte occurrence $\hat{P}_{[h]i}$.
+We modeled $\hat{P}_{[h]i}$ as a linear function of intercept $ \mathrm{A}_{[h]i}$ and slope $\mathrm{T}_{[h]}$ defining the global trend in endophyte prevalence specific to each host species as well as with spatially-varying intercepts $\alpha_{[h_{1}]l[i]}$ and slopes $\tau_{[h_{1}]l[i]}$ associated with location ($l[i]$, a unique latitude-longitude combination).
+The joint-model structure allowed us to share variance terms across focal species to account for dependence associated with the collection of specimens and identification of endophytes. 
+We included the spatial random effect $\delta_{l[i]}$, intended to account for residual spatial variation, and $\chi_{c[i]}$ and $\omega_{s[i]}$ i.i.d.-random effects indexed for each collector identity ($c[i]$), and scorer identity ($s[i]$) of the $i^{th}$ specimen
 
@@
 		logit(\hat{P}_{[h_{2}]i}) = \mathrm{A}_{[h_{2}]i} + \mathrm{T}_{[h_{2}]}*year_i  + \alpha_{[h_{2}]l[i]} + \tau_{[h_{2}]l[i]}*year_i  + \delta_{l[i]}+ \chi_{c[i]} + \omega_{s[i]} \\
-		logit(\hat{P}_{[h_{3}]i}) = \mathrm{A}_{[h_{2}]i} + \mathrm{T}_{[h_{2}]}*year_i  + \alpha_{[h_{2}]l[i]} + \tau_{[h_{3}]l[i]}*year_i  + \delta_{l[i]}+ \chi_{c[i]} + \omega_{s[i]}
+		logit(\hat{P}_{[h_{3}]i}) = \mathrm{A}_{[h_{2}]i} + \mathrm{T}_{[h_{2}]}*year_i  + \alpha_{[h_{2}]l[i]} + \tau_{[h_{3}]l[i]}*year_i  + \delta_{l[i]}+ \chi_{c[i]} + \omega_{s[i]}\\
 	\end{align}
@@
 
-First, to quantify global trends in endophyte prevalence, we modeled endophyte presence/absences of the $i^{th}$ specimen ($P_i$) as a Bernoulli response variable with expected probability of endophyte occurrence $\hat{P_i}$.
-We modeled $\hat{P_i}$ as a linear function of collection year and accounting for random effects associated with location ($l[i]$, a unique latitude-longitude combination), collector identity ($c[i]$), and scorer identity ($s[i]$) of the $i^{th}$ specimen.
-\begin{subequations}
-	\label{eq:trends}
-	\begin{align}
-		logit(\hat{P}_{i}) = \alpha_{l[i]} + \beta*year_i + \chi_{c[i]} + \omega_{s[i]} 
-	\end{align}
-\end{subequations}
-Spatially-indexed random intercepts $\alpha_{l[i]}$ account for potential spatial autocorrelation between data points, and year slope $\beta$ describes the overall temporal trend in endophyte prevalence.
-We accounted for potential biases introduced during the process of collecting specimens as well as in scoring ability by including random effects specific to each collector $\chi$ and scorer $\omega$.
+Each spatially-structured parameter depended on a covariance matrix according to the proximity of each collection location \citep{lindgren2011explicit,bakka2018spatial}. 
+The covariance matrix was approximated using a Mat\'{e}rn covariance function, with each data point assigned a location according to the nodes of a mesh of non-overlapping triangles across our study area (Fig.
+\ref{fig:meshplot}).
+
 Previous work suggests that behavior of historical botanists and uneven sampling may introduce biases into ecological inferences made from historic collections \cite{kozlov2020biases}. 
@@
 	\centering
-	\includegraphics[width = \linewidth]{year_plot.png}
+	\includegraphics[width = \linewidth]{../Plots/year_plot.png}
 	\caption[Temporal trends in endophyte prevalence.]{Temporal trends in endophyte prevalence.
@@
 Colors represent each host species and point size is determined by the number of specimens.
-Lines show predicted mean endophyte prevalence over the study period along with the 50\%  and 95\% CI bands.}
+Lines show predicted mean endophyte prevalence over the study period along with the 50\%  and 95\% CI bands incorporating uncertainty associated with collector and scorer random effects.}
 	\label{fig:temporal}
@@
 		\centering
-		\includegraphics[width = \linewidth]{endo_status_map.png}
+		\includegraphics[width = \linewidth]{../Plots/endo_status_map.png}
 		\caption{\textbf{Endophyte presence/absence in specimens of each host species.} Points show collection locations colored according to whether the specimen contained endophytes ( E+; blue points) or did not contain endophytes (E-, tan points) and are faceted based on collection period.}
+		\label{fig:endo_status_map}
 	\end{figure}
@@
 		\centering
-		\includegraphics[width = \linewidth]{mesh_plot.png}
+		\includegraphics[width = \linewidth]{../Plots/mesh_plot.png}
 		\caption{\textbf{Delauney triangulation mesh used to estimate spatial dependence between data points}.
@@
 Red points indicate locations of sampled herbarium specimens, and the blue outlines show the international borders used to define the edge of the mesh along coastlines.}
+		\label{fig:meshplot}
 	\end{figure}

### 2024-08-15 Manuscript/endo_herbarium.tex
editing the methods

@@
 \subsection*{Modeling spatial and temporal changes in endophyte prevalence}
-We assessed spatial and temporal changes in endophyte prevalence across each host distribution, first quantifying the ``global'' trends, aggregating across space, and then examining spatial heterogeneity in the direction and magnitude of endophyte change (hotspots and coldspots) across the spatial extent of each host's distribution.
-To appropriately account for the spatial non-independence of geo-referenced occurrences, we used an approximate Bayesian method, Integrated Nested Laplace Approximation (INLA), to construct spatio-temporal models of endophyte prevalence.
+We assessed spatial and temporal changes in endophyte prevalence across each host distribution, quantifying the ``global'' temporal trends, aggregating across space, and then examining spatial heterogeneity in the direction and magnitude of endophyte change (hotspots and coldspots) across the spatial extent of each host's distribution.
+To appropriately account for the spatial non-independence of geo-referenced \josh{occurrences}{spelling?}, we used an approximate Bayesian method, Integrated Nested Laplace Approximation (INLA), to construct spatio-temporal models of endophyte prevalence.
 INLA provides a computationally efficient method of ascertaining parameter posterior distributions for certain models that can be formulated as latent Gausssian Models \cite{rue2009approximate}.
@@
 The joint-model structure allowed us to share variance terms across focal species to account for dependence associated with the collection of specimens and identification of endophytes. 
-We included the spatial random effect $\delta_{l[i]}$, intended to account for residual spatial variation, and $\chi_{c[i]}$ and $\omega_{s[i]}$ i.i.d.-random effects indexed for each collector identity ($c[i]$), and scorer identity ($s[i]$) of the $i^{th}$ specimen
+Shared variance terms included the spatially-dependent random effect $\delta_{l[i]}$, intended to account for residual spatial variation, and $\chi_{c[i]}$ and $\omega_{s[i]}$ i.i.d.-random effects indexed for each collector identity ($c[i]$), and scorer identity ($s[i]$) of the $i^{th}$ specimen.
 
@@
 
-Each spatially-structured parameter depended on a covariance matrix according to the proximity of each collection location \citep{lindgren2011explicit,bakka2018spatial}. 
-The covariance matrix was approximated using a Mat\'{e}rn covariance function, with each data point assigned a location according to the nodes of a mesh of non-overlapping triangles across our study area (Fig.
-\ref{fig:meshplot}).
-
 Previous work suggests that behavior of historical botanists and uneven sampling may introduce biases into ecological inferences made from historic collections \cite{kozlov2020biases}. 
@@
 Similarly, the process of scoring seeds for hyphae involved several student researchers who, even with standardized training, may vary in their likelihood of positively identifying \emph{Epichloë} hyphae.
-By including a random effect for collectors and for scorers, we attempted to accound for variance across individual researchers that may bias our predictions of changes in endophyte prevalence.
-\tom{Models for each host species were fit separately.}{It would be great to pull all species into one model and have them share variance terms for the random effects.
-I suspect such a model would give better and more stable estimates.}
-\tom{}{I updated the notation in ways that make more sense to me, but you should check that this is true to the actual model (I think it is).
-Also, a more complete presentation of this model would show the variance terms for $\alpha$, $\chi$, and $\omega$.
-I presume the latter two are Gaussian but I don't know how to represent the distribution of $\alpha$.}
+By including a random effect for collectors and for scorers, we attempted to account for variance across individual researchers that may bias our predictions of changes in endophyte prevalence.
 
-Second, to quantify how temporal trends may vary spatially, we repeated the modelling above, but incorporated a spatially-varying coefficient for collection year:
-\begin{subequations}
-	\label{eq:trends}
-	\begin{align}
-		logit(\hat{P}_{i}) = \alpha_{l[i]} + \beta_{l[i]}*year_i + \chi_{c[i]} + \omega_{s[i]} 
-	\end{align}
-\end{subequations}
-The spatially-varying year slope $\beta_{l}$ allowed us to flexibly estimate variation in the temporal trajectory of endophyte change at locations across the study region.
+We performed model fitting using the inlabru R package \citep{}.
+Global parameters $\mathrm{A}$, and $\mathrm{T}$, and i.i.d. random effects were given penalized complexity priors $\chi$ and $\omega$. 
+Each spatially-structured parameter depended on a covariance matrix according to the proximity of each collection location \citep{lindgren2011explicit,bakka2018spatial}. 
+The covariance matrix was approximated using a Mat\'{e}rn covariance function, with each data point assigned a location according to the nodes of a mesh of non-overlapping triangles encompassing the study area (Fig.
+\ref{fig:meshplot}).
+Priors, termed ''range" and "variance", define the distance of spatial decay described by the Mat\'{e}rn covariance function.
+Priors for results presented in the main text reflect a range of \josh{XX}{} kilometers. 
+We found that model results were sensitive to this choice, and so tested a range of priors (from XX kilometers to XX kilometers) and meshes (Supplemental Material), finding that model results were qualitatively similar, i.e.
+the same direction of effects across space, but that the magnitude and uncertainty varied.
 
-\tom{For both models, spatially-structured random intercepts ($\alpha_{l}$) and slopes ($\beta_{l}$) were constructed using stochastic partial differential equations (SPDE) that depend on a covariance matrix according to the proximity of each collection location \citep{lindgren2011explicit,bakka2018spatial}.
-The covariance matrix was approximated using a Mat\'{e}rn covariance function, with each data point assigned a location according to the nodes of a mesh of non-overlapping triangles across our study area (Fig A2).}{This paragraph would be a place to describe the variance terms for the other random effects.}
+\subsection*{Validating the model with in-sample and out-of-sample tests}
 
-We performed model fitting using the inlabru R package \citep{}, \tom{with vague priors}{I thought you needed informative priors on the spatial decay parameters}, and compared models with different sizes of mesh, \tom{which had little effect on the resulting model estimates}{That'sa good but you still need to state what mesh size you used and what that means, biologically.}.
-Each \tom{mesh}{You have not defined what you mean by ``mesh''.} was bounded by the predicted host distribution, described below.
-Posterior modes were \tom{stable}{Assessed how?} indicating that numeric convergence was successful.
+In all cases, posterior modes were \tom{stable}{Assessed how?} and equal to zero, indicating successful numeric convergence.
 We assessed model fit with graphical posterior predictive checks (Fig. A3).
@@
 
-\subsection*{Modeling distributions of host species}
-We modeled  epicloe host species distribution to predict their occurrences in space and time as continuous and binary maps of potential presences. 
-These maps were used as a backbone to predict endophyte prevalence on epicloe host species. 
-The species distribution models were built following the ODMAP (overview, data, model, assessment, prediction) protocol \citep{crossley2022opposing}. 
-We used the observed presence of the host species collected from GBIF from 1990 to 2020.
-These occurrences were corrected for spatial autocorrelation due to sampling bias by thinning the occurrences to the spatial scale of the climatic variables.
-The climatic variables were temperature of the spring, precipitation of the spring  and precipitation of the summer.
-We preferred these variables because they were not correlated (Variance Inflation Factor $>$ 0.7) and also allowed the model to account for the influence of seasonal climate variation on species presence.
-The occurrence data was split into 75\% for model training and 25\% for model testing.
-We fitted the model using maximum entropy (MaxEnt) using the maxent function in the package dismo \citep{hijmans2017package}.
-MaxEnt was preferred because it does not generate response curves that may cause unpredictable behavior when applied to new climates \citep{hijmans2006ability}.
- We used 10000 pseudo-absences as background points. 
- To convert the continuous predicted probabilities into binary presence - absence maps, we used the  training sensitivity (true positive rate) and specificity threshold (true negative rate) \citep{liu2005selecting}.
-The performances of the model were evaluated using the AUC \citep{jimenez2012insights}.
-
-\subsection*{Validating the model with an out-of-sample test}
 We evaluated the predictive ability of the model using contemporary endophyte surveys as out-of-sample test data, \tom{an important but rarely used strategy in ecological studies \cite{tredennick2021practical}.}{This is the type of thing to emphasize in the intro?
 Are there any other collections-based papers that have done anything like this??
-None to my knowledge.}
+None to my knowledge.} \josh{}{Add Benjamin lee paper, maybe?
+it's not just herbaria, but kind of related}
 We used data from contemporary surveys of endophyte prevalence  in \emph{A. hyemalis} and \emph{E. virginicus} in Texas and the southern US. 
@@
 I am also a little confused because the appendix has spearman correlations but there are no methods here for where those come from.}
+
+\subsection*{Modeling distributions of host species}
+We modeled the distribution of each host species to generate maps on which we predicted the dynamics of \emph{Epichloë} symbionts.
+We followed the ODMAP (overview, data, model, assessment, prediction) protocol \citep{crossley2022opposing}, using \josh{presence-only observations of the host species from GBIF between 1990 to 2020 }{How many data points does this end up being?}.
+To reduce the potential influence of sampling bias and spatial autocorrelation, we thinned the occurrences to the spatial scale \josh{()}{I think this is 4km, but need to check} of our selected climatic predictors.
+We selected climate variables that aligned with our analysis of climatic influences on trends in endophyte prevalence described above.
+We calculated the mean and standard devation of seasonal temperature and precipitation across 1990 to 2020. 
+Among this suite of variables, we chose to include \josh{mean spring temperature, mean spring precipitation, and mean summer temperature}{is this list accurate?
+my re-running my climate analysis means we ought to re-run the sdm's, and use SD instead of CV, so we can update this list based on whatever is appropriate}, which were uncorrelated (Variance Inflation Factor $>$ 0.7) and allowed us to predict the occurrence probabilty of each host species in space and time.
+We fit maximum entropy (MaxEnt) models using the maxent function in the package dismo \citep{hijmans2017package}. 
+\josh{MaxEnt is preferred because it has been shown to generate response curves with less unpredictable behavior when applied to new climates \citep{hijmans2006ability}.}{possibly could remove this sentence?}
+We generated 10,000 pseudo-absences as background points, and split the occurrence data into 75\% for model training and 25\% for model testing.
+The performances of models were evaluated with AUC \josh{AUC}{can we add these values?} \citep{jimenez2012insights}. 
+To convert the continuous predicted probabilities into binary presence - absence maps, we used the  training sensitivity (true positive rate) and specificity threshold (true negative rate) \citep{liu2005selecting}.
+These binary maps serve as boundaries in presented maps of change in endophyte prevalence, and outline the set of pixels used in our analysis of climate correlates with trends in endophyte prevalence
 

### 2024-08-15 Analyses/endo_herbarium_multispecies_spatiotemporal_analysis.R
remaking the mesh plot

@@ -298,10 +298,12 @@ mesh <- fm_mesh_2d_inla(
 
 mesh_plot <- ggplot() +
   gg(data = mesh) +
-  geom_point(data = data, aes(x = easting, y = northing, col = species), size = 1) +
+  geom_point(data = data, aes(x = easting, y = northing, fill = species), color = "grey29", size  = 1, shape = 21) +
   coord_sf()+
   theme_bw() +
-  labs(x = "", y = "")
+  scale_fill_manual(values = species_colors)+
+  labs(x = "", y = "", color = "Species")
+# mesh_plot
 # ggsave(mesh_plot, filename = "Plots/mesh_plot.png", width = 6, height = 6)
 
 # make spde
@@ -339,7 +341,7 @@ pc_prec <- list(prior = "pcprec", param = c(1, 0.1))
 
 # version of the model with multiple likelihoods
 
-cmp <- ~  space(geometry, model = spde) + space.species1(geometry, model = spde) + space.species2(geometry, model = spde) + + space.species3(geometry, model = spde) +
+cmp <- ~  space(geometry, model = spde) + space.species1(geometry, model = spde) + space.species2(geometry, model = spde) + space.species3(geometry, model = spde) +
   time.species1(geometry, weights = std_year, model = spde) + time.species2(geometry, weights = std_year, model = spde) + time.species3(geometry, weights = std_year, model = spde) +
   + int.species1(1) + int.species2(1) + int.species3(1)+
   + year.species1(main = ~0 + std_year, model = "fixed") + year.species2(main = ~0 + std_year, model = "fixed") + year.species3(main = ~0 + std_year, model = "fixed")+

### 2024-08-15 Analyses/endo_herbarium_records_merge.R
fixing a name in climate data at pixels, and going to make new plots for this

@@ -1756,7 +1756,7 @@ prism_diff_df <- tibble(lon = coords[,1], lat = coords[,2],
                         ppt_autumn_cv_diff = unlist(terra::extract(ppt_autumn_cv_difference, coords)))
 write_csv(prism_diff_df, file = "prism_diff_df.csv")
 
-
+prism_diff_df <- read_csv(file = "prism_diff_df.csv")
 # this is the coordinates we are using for the spatial model prediction currently
 # this is the set of data for which we want predictions
 
@@ -1777,7 +1777,7 @@ vrt_elvi <- readRDS(file = "elvi_distribution_df.rds")
 
 dd_crs <- "GEOGCRS[\"NAD83\",\n    DATUM[\"North American Datum 1983\",\n        ELLIPSOID[\"GRS 1980\",6378137,298.257222101004,\n            LENGTHUNIT[\"metre\",1]]],\n    PRIMEM[\"Greenwich\",0,\n        ANGLEUNIT[\"degree\",0.0174532925199433]],\n    CS[ellipsoidal,2],\n        AXIS[\"geodetic latitude (Lat)\",north,\n            ORDER[1],\n            ANGLEUNIT[\"degree\",0.0174532925199433]],\n        AXIS[\"geodetic longitude (Lon)\",east,\n            ORDER[2],\n            ANGLEUNIT[\"degree\",0.0174532925199433]],\n    ID[\"EPSG\",4269]]"
 
-vrt_aghy_dd <- svc.pred_aghy  %>% 
+vrt_aghy_dd <- vrt_aghy  %>% 
   spTransform(dd_crs)
 
 vrt_agpe_dd <- vrt_agpe  %>% 
@@ -1788,7 +1788,7 @@ vrt_elvi_dd <- vrt_elvi  %>%
 
 
 
-aghy_prism_diff_pred_df <- tibble(lon = svc.pred_aghy@coords[,1], lat = svc.pred_aghy@coords[,2],
+aghy_prism_diff_pred_df <- tibble(lon = vrt_aghy_dd@coords[,1], lat = vrt_aghy_dd@coords[,2],
                              tmean_annual_diff = unlist(terra::extract(tmean_annual_difference, vrt_aghy_dd@coords)),
                              tmean_spring_diff = unlist(terra::extract(tmean_spring_difference, vrt_aghy_dd@coords)),
                              tmean_summer_diff = unlist(terra::extract(tmean_summer_difference, vrt_aghy_dd@coords)),

### 2024-08-15 Manuscript/endo_herbarium.tex
more edits

@@
 We performed model fitting using the inlabru R package \citep{}.
-Global parameters $\mathrm{A}$, and $\mathrm{T}$, and i.i.d. random effects were given penalized complexity priors $\chi$ and $\omega$. 
+Global intercept and slope parameters $\mathrm{A}$, and $\mathrm{T}$, were given vague priors.
+Scorer and collector random effects, $\chi$ and $\omega$, were given penalized complexity priors, with distributions approximating a Normal distribution with standard deviation of 5. 
 Each spatially-structured parameter depended on a covariance matrix according to the proximity of each collection location \citep{lindgren2011explicit,bakka2018spatial}. 
@@
 the same direction of effects across space, but that the magnitude and uncertainty varied.
+%In all cases, posterior modes were \tom{stable}{Assessed how?} and equal to zero, indicating successful numeric convergence.
 
-\subsection*{Validating the model with in-sample and out-of-sample tests}
-
-In all cases, posterior modes were \tom{stable}{Assessed how?} and equal to zero, indicating successful numeric convergence.
-We assessed model fit with graphical posterior predictive checks (Fig. A3).
-\tom{The model performed adequately at classifying the historical data, comparing the accuracy of predictions from the model with observed data (avg.
-AUC = 0.77; Fig.
-A4).
-}{Maybe move this to validation section, and then have both in-sample and out-of-sample approaches.}
-
-We evaluated the predictive ability of the model using contemporary endophyte surveys as out-of-sample test data, \tom{an important but rarely used strategy in ecological studies \cite{tredennick2021practical}.}{This is the type of thing to emphasize in the intro?
+\subsection*{Validating model performance with in-sample and out-of-sample tests}
+We evaluated the predictive ability of the model using both in-sample training data from the herbarium surveys, and with out-of-sample test data from contemporary endophyte surveys, \tom{an important but rarely used strategy in ecological studies \cite{tredennick2021practical}.}{This is the type of thing to emphasize in the intro?
 Are there any other collections-based papers that have done anything like this??
@@
 hyemalis} and 20 populations of \emph{E.
-virginicus} across the south-central US, with emphasis on Texas and neighboring states (Fig \tom{A4}{This is now A6.
-Good reminder to use the ref function.}).
-During surveys, we collected seeds from up to 30 individuals per location  (average number of plants sampled: 22.9).
+virginicus} across the south-central US, with emphasis on Texas and neighboring states (Fig \ref{fig:contempsurveysmap}).
+During surveys, we collected seeds from up to 30 individuals per location  (average number of plants sampled: $22.9$).
 We quantified the endophyte status of each individual with staining microscopy as described for the herbarium surveys (with 5-10 seeds scored per individual), and calculated the prevalence of endophytes within the population (proportion of symbiotic plants divided by the number of sampled plants).
-For each population, we compared the observed fraction of endophyte-symbiotic hosts to the predicted probability of endophyte occurrence $\hat{P}$ derived from the model based on location and year, with collector and scorer random effects fixed at zero.
-\tom{The contemporary survey period (2013-2020) is at the most recent edge of the time period encompassed by the historical observations used for model fitting.
-We compared the model's prediction for these locations to the observed population prevalence.}{It is not clear if you are testing model 1 (``global trend'') or model 2 (``spatially varying trends'').}
+For each population, we compared the observed fraction of endophyte-symbiotic hosts to the predicted probability of endophyte occurrence $\hat{P}$ derived from the model based on location and year. 
+The contemporary survey period (2013-2020) is at the most recent edge of the time period encompassed by the historical observations used for model fitting.
+We compared the model's prediction for these locations to the observed population prevalence.
 
@@
 We generated 10,000 pseudo-absences as background points, and split the occurrence data into 75\% for model training and 25\% for model testing.
-The performances of models were evaluated with AUC \josh{AUC}{can we add these values?} \citep{jimenez2012insights}. 
+The performance of models was evaluated with AUC \josh{AUC}{can we add these values?} \citep{jimenez2012insights}. 
 To convert the continuous predicted probabilities into binary presence - absence maps, we used the  training sensitivity (true positive rate) and specificity threshold (true negative rate) \citep{liu2005selecting}.
 These binary maps serve as boundaries in presented maps of change in endophyte prevalence, and outline the set of pixels used in our analysis of climate correlates with trends in endophyte prevalence
-
-%We evaluated how predicted changes in endophyte prevalence ($\Delta{P}$) correlated with %changes in seasonal climate at a grid of points across the study area using linear regression.
-
-%\begin{subequations}
-	%\label{eq:climate}
-	%\begin{align}
-		%\Delta{P} \sim Normal(\mu, \sigma)\\
-		%\mu = \beta_{0} + \alpha_{1}*\Delta{Precip_{spring}} + %\alpha_{2}*\Delta{Precip_{summer}} + \alpha_{3}*\Delta{Precip_{autumn} }+ \\ %\alpha_{4}*\Delta{Temp_{spring}} + \alpha_{5}*\Delta{Temp_{summer}}  + %\alpha_{6}*\Delta{Temp_{autumn}}
-%	\end{align}
-%\end{subequations}
-
- %The change in prevalence is determined by an intercept $\beta_{0}$ slopes specific to the seasonal %changes in preciptation ($\alpha_{1}$, $\alpha_{2}$, and $\alpha_{3}$) and in mean %temperature ($\alpha_{4}$, $\alpha_{5}$, and $\alpha_{6}$).
- %We fit models for each host species separately with rINLA using default vague priors.
 
@@
 hyemalis} (Fig.
-6 B-C), however the contemporary data contains additional variability at smaller scales not captured by our sampling of herbarium specimens.
+\ref{fig:contemptestplot} B-C), however the contemporary data contains additional variability at smaller scales not captured by our sampling of herbarium specimens.
 We interpreted this to mean that the model captured regional spatial dynamics, but underpredicts local scale dynamics. 
@@
 	\centering
-	\includegraphics[width = \linewidth]{contemp_test_plot.png}
+	\includegraphics[width = \linewidth]{../Plots/contemp_test_plot.png}
 	\caption{\textbf{Predicted vs observed endophyte prevalence for contemporary test data.} (A) The model, trained on historic herbarium collection data, performed modestly at predicting contemporary endophyte prevalence in \emph{A.
@@
 Point size in panels B and C reflect sample sizes of contemporary endophyte population surveys.}
+	\label{fig:contemptestplot}
 \end{figure}
@@
 		\includegraphics[width = \linewidth]{../Plots/mesh_plot.png}
-		\caption{\textbf{Delauney triangulation mesh used to estimate spatial dependence between data points}.
+		\caption{\textbf{Triangulation mesh used to estimate spatial dependence between data points}.
 Grey lines indicate edges of triangles used to define distances between observations.
-Red points indicate locations of sampled herbarium specimens, and the blue outlines show the international borders used to define the edge of the mesh along coastlines.}
+Colored points indicate locations of sampled herbarium specimens for each host species, and the blue line shows the convex hull and coastline used to define the edge of the mesh around the data points.
+The thick black line shows the convex hull defining a buffer space around the edge of the mesh to reduce the influence of edge effects on model estimates.}
 		\label{fig:meshplot}
@@
 		\centering
-		\includegraphics[width = .5\linewidth]{ROC_plot.png}
-		\caption{\textbf{ROC plot showing model performance classifying observations according to endophyte status.} The curves show adequate model performance for observed (top) and test (bottom) data.
-The AUC for each is 0.77.}
+		\includegraphics[width = .5\linewidth]{../Plots/ROC_training_plot.png}
+		\caption{\textbf{ROC plot showing model performance classifying observations according to endophyte status within the in-sample data.} The curves show adequate model performance for observed data.
+The AUC value is 0.77.}
+				\label{fig:ROCtraining}
+	\end{figure}
+
+	\begin{figure}[H]
+		\centering
+		\includegraphics[width = .5\linewidth]{../Plots/ROC_test_plot.png}
+		\caption{\textbf{ROC plot showing model performance classifying observations according to endophyte status within the out-of-sample data.} The curves show adequate model performance for test data.
+The AUC value is 0.77.}
+		\label{fig:ROCtest}
 	\end{figure}
@@
 	\centering
-	\includegraphics[width = \linewidth]{test_data_map.png}
+	\includegraphics[width = \linewidth]{../Plots/contemp_surveys_map.png}
 	\caption{\textbf{Locations of contemporary surveys of endophytes in \emph{A. hyemalis} used as "test" data (red points), relative to the historical collection data (black crosses).}}
+	\label{fig:contempsurveysmap}
 \end{figure}

### 2024-08-16 Analyses/posthoc_climate_correlation.R
making prettier plots of the covariate data

@@ -224,7 +224,6 @@ vrt_elvi_dd <- svc.pred_elvi  %>%
   spTransform(dd_crs)
 
 
-
 aghy_prism_diff_pred_df <- tibble(lon = vrt_aghy_dd@coords[,1], lat = vrt_aghy_dd@coords[,2],
                                   coords.x1 = svc.pred_aghy@coords[,1], coords.x2 = svc.pred_aghy@coords[,2],
                                   tmean_annual_diff = as.numeric(unlist(terra::extract(tmean_annual_difference, vrt_aghy_dd@coords))),
@@ -332,8 +331,91 @@ aghy_prism_diff_pred_df <- read_csv("aghy_prism_diff_pred_df.csv")
 agpe_prism_diff_pred_df <- read_csv("agpe_prism_diff_pred_df.csv")
 elvi_prism_diff_pred_df <- read_csv("elvi_prism_diff_pred_df.csv")
 
+# Plotting the change in climate at our pixel values
 
-# dim(elvi_prism_diff_pred_df)
+prism_diff_pred_df <- aghy_prism_diff_pred_df %>% 
+  full_join(agpe_prism_diff_pred_df) %>% 
+  full_join(elvi_prism_diff_pred_df) %>% 
+  pivot_longer(contains("_diff")) %>% 
+  separate(name, into = c("climate", "season"), sep = "_", extra = "drop", remove = FALSE) %>% 
+  filter(!grepl("_cv_", name) & ! grepl("annual", name)) %>% 
+  mutate(season_f = factor(season, level = c("spring", "summer", "autumn")),
+         moment = case_when(grepl("_sd_", name) ~ "sd",
+                            TRUE ~ "mean"))
+  
+
+
+AGHY_tmean_change_plot <- ggplot(filter(prism_diff_pred_df, climate == "tmean" & species == "A. hyemalis"))+
+  geom_tile(aes(x = coords.x1, y = coords.x2, fill = value)) +
+  coord_sf()+
+  facet_wrap(~moment + season_f)+
+  scale_fill_viridis_c(option = "magma") + labs(x = "Lon.", y = "Lat.", fill = "Change in ºC")+
+  theme(strip.text = element_text(size = rel(1)))
+AGHY_tmean_change_plot
+
+AGHY_ppt_change_plot <- ggplot(filter(prism_diff_pred_df, climate == "ppt" & species == "A. hyemalis"))+
[…]

### 2024-08-16 Manuscript/endo_herbarium.tex
editing the climate correlations methods

@@
 Prism provides reconstructions of historic climate variables across the United States by spatially-interpolating weather station data \citep{diLuzio2008constructing}. 
-We calculated 30-year climate normals for annual and seasonal mean temperature and cumulative precipitation for the recent (1990 to 2020) and historic (1895 to 1925) periods.
+We calculated 30-year climate normals for seasonal mean temperature and cumulative precipitation for the recent (1990 to 2020) and historic (1895 to 1925) periods.
 We used three four-month seasons within the year (Spring: January, February, March, April; Summer: May; June, July, August; Autumn: September, October, November, December). 
-This division of seasons allowed us to quantify differences in climate associated with the two ``cool'' seasons, when we expect our focal species to be most biologically active (\emph{A.
+This division of seasons allowed us to quantify differences in climate associated with the two ``cool'' seasons, when we expected our focal species to be most biologically active (\emph{A.
 hyemalis} flowering phenology: Spring; \emph{E.
@@
 In addition to mean climate conditions, environmental variability itself can influence population dynamics \cite{tuljapurkar_population_1982} and changes in variability are a key prediction of climate change models \cite{stocker2013technical, ipcc_2021}.
-Therefore we calculated the coefficient of variation (CV) during each period for each annual and seasonal climate driver as the interannual standard deviation divided by the mean across each 30-year period.
-We then took the difference between recent and historic periods for the mean and CV for each climate driver \tom{(Fig.
-A5)}{This is Figure A7 -- Can you make the color scale on these diverging at zero?}.
-Because initial analyses indicated a high degree of collinearity between seasonal and annual changes in temperature, we used annual temperature only, along with annual and seasonal precipitation, in the subsequent analysis.
-All together, this left us with measurements of change in 10 potential climate drivers: the mean and coefficient of variation of annual temperature, as well as the mean and coefficient of variation of cumulative annual precipitation, cumulative spring precipitation, cumulative summer precipitation, and cumulative autumn precipitation \jacob{(Fig A8-A9)}{ The species names are not clear  on Fig A9.
-I suggest increase the font siize }.
+Therefore, we calculated the standard deviation for each annual and seasonal climate driver across each 30-year period.
+We then took the difference between recent and historic periods for the mean and standard deviation for each climate driver (Figs. \ref{fig:AGHY_climate_covariates}-\ref{fig:ELVI_climate_covariates}).
+%Because initial analyses indicated a high degree of collinearity between seasonal and annual changes in temperature, we used annual temperature only, along with annual and seasonal precipitation, in the subsequent analysis.
+All together, we assessed twelve potential climate drivers: the mean and standardiation of spring, summer, and autumn temperature, as well as the mean and standard deviation of spring, summer, and autumn cumulative precipitation, cumulative precipitation, and cumulative precipitation.
 
@@
 
-We found that trends in endophyte prevalence were strongly associated with seasonal climate change drivers (Fig. 6).
+We found that trends in endophyte prevalence were strongly associated with seasonal climate change drivers (Fig. \ref{fig:climate_correlation_plot}).
 For the majority of the study region, the climate has become wetter and cooler over the last century (Fig. A7-A8), a consequence of regional variation in global climate change \cite{ipcc_2021}. 
@@
 	\centering
-	\includegraphics[width = .7\linewidth]{tmean_regression_plot_ESA.png}
-	\caption{\textbf{Correlations between changes in climate drivers and changes in endophyte prevalence.} Color denotes the Spearman correlation coefficient between the relative rate of change in endophyte prevalence and the change in annual mean temperature ($^oC$) and total annual and seasonal precipitation (mm), as well as the change in the coefficient of variation of each climate driver.
-Positive correlation coefficients indicate that greater increases in a climate driver were associated with larger increases in endophyte prevalence, while negative values indicate that .
-Asteriks denote correlation coefficients $> .3$ or $< -.3$.}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width = .7\linewidth]{ppt_regression_plot_ESA.png}
-	\caption{\textbf{Correlations between changes in climate drivers and changes in endophyte prevalence.} Color denotes the Spearman correlation coefficient between the relative rate of change in endophyte prevalence and the change in annual mean temperature ($^oC$) and total annual and seasonal precipitation (mm), as well as the change in the coefficient of variation of each climate driver.
-Positive correlation coefficients indicate that greater increases in a climate driver were associated with larger increases in endophyte prevalence, while negative values indicate that .
-Asteriks denote correlation coefficients $> .3$ or $< -.3$.}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width = .7\linewidth]{ppt_CV_regression_plot_ESA.png}
-	\caption{\textbf{Correlations between changes in climate drivers and changes in endophyte prevalence.} Color denotes the Spearman correlation coefficient between the relative rate of change in endophyte prevalence and the change in annual mean temperature ($^oC$) and total annual and seasonal precipitation (mm), as well as the change in the coefficient of variation of each climate driver.
-Positive correlation coefficients indicate that greater increases in a climate driver were associated with larger increases in endophyte prevalence, while negative values indicate that .
-Asteriks denote correlation coefficients $> .3$ or $< -.3$.}
+	\includegraphics[width = .7\linewidth]{../Plots/climate_trends_plot.png}
+	\caption{\textbf{Correlations between changes in climate drivers and changes in endophyte prevalence.} Lines show relationships between seasonal climate covariates and spatially-varying trends in endophyte prevalence for each host species, along with 96\% CI.}
+	\label{fig:climate_correlation_plot}
 \end{figure}
@@
 	\centering
-	\includegraphics[width = .8\linewidth]{ppt_change_maps.png}
-	\caption{\textbf{Change in precipitation between the periods 1895-1925 and 1990-2020.} Color represents change in annual or seasonal total precipitation (A,C,E,G) and in the coeffiecient of variation of annual or seasonal total precipitation (B,D,F,H).
-Maps show the study area of \emph{A.
-hyemalis}.
-Map pixels used in correlation analysis with endophyte change were pulled from studies areas specific to each host species.
-}
+	\includegraphics[width = .8\linewidth]{../Plots/AGHY_climate_change_plot.png}
+	\caption{\textbf{Change in seasonal climate variables between the periods 1895-1925 and 1990-2020.} Color represents change in (A) seasonal temperature and (B) seasonal precipitation.
+Maps show pixels covering the modeled distribution of \emph{A.
+hyemalis} used in post-hoc climate correlation analysis.}
+	\label{fig:AGHY_climate_covariates}
 \end{figure}
@@
 	\centering
-		\includegraphics[width = \linewidth]{tmean_change_maps.png}
-		\caption{\textbf{Change in temperature between the periods 1895-1925 and 1990-2020.} Color represents change in annual mean temperature (A) and in the coeffiecient of variation of annual mean temperature (B).
-Maps show the study area of \emph{A.
-hyemalis}.
-Map pixels used in correlation analysis with endophyte change were pulled from studies areas specific to each host species.}
+	\includegraphics[width = .8\linewidth]{../Plots/AGPE_climate_change_plot.png}
+	\caption{\textbf{Change in seasonal climate variables between the periods 1895-1925 and 1990-2020.} Color represents change in (A) seasonal temperature and (B) seasonal precipitation.
+Maps show pixels covering the modeled distribution of \emph{A.
+perennans} used in post-hoc climate correlation analysis.}
+	\label{fig:AGPE_climate_covariates}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width = .8\linewidth]{../Plots/ELVI_climate_change_plot.png}
+	\caption{\textbf{Change in seasonal climate variables between the periods 1895-1925 and 1990-2020.} Color represents change in (A) seasonal temperature and (B) seasonal precipitation.
+Maps show pixels covering the modeled distribution of \emph{E.
+virginicus} used in post-hoc climate correlation analysis.}
+	\label{fig:ELVI_climate_covariates}
 \end{figure}

### 2024-08-22 Analyses/endo_herbarium_multispecies_spatiotemporal_analysis.R
updating plots of posteriors for summary stats

@@ -18,6 +18,7 @@ library(tidyterra)
 library(patchwork)
 library(ggmap)
 library(pROC)
+library(tidybayes)
 
 invlogit<-function(x){exp(x)/(1+exp(x))}
 species_colors <- c("#1b9e77","#d95f02","#7570b3")
@@ -630,7 +631,7 @@ ggsave(posterior_plot, filename = "Plots/posterior_plot.png", width = 6, height 
 ################################################################################################################################
 ##########  Plotting the posteriors of scorer and collector effects ###############
 ################################################################################################################################
-random_collectors <- sample((1:NROW(fit$summary.random$collector)), 12)
+# random_collectors <- sample((1:NROW(fit$summary.random$collector)), 12)
 
 slist <- vector("list", NROW(fit$summary.random$scorer))
 for (i in seq_along(slist)) slist[[i]] <- plot(fit, "scorer", index = i) + lims(x = c(-2.5,2.5))
@@ -660,20 +661,62 @@ plot(fit, "time.species1")
 ################################################################################################################################
 ##########  Generating posterior samples for some summary statistics ###############
 ################################################################################################################################
+n_draws <- 500
+
+posteriors.aghy <- generate(
+  fit,
+  formula = ~ year.species1_latent,
+  n.samples = n_draws) 
+posteriors.agpe <- generate(
+  fit,
+  formula = ~ year.species2_latent,
+  n.samples = n_draws) 
+posteriors.elvi <- generate(
+  fit,
+  formula = ~ year.species3_latent,
+  n.samples = n_draws) 
+
+colnames(posteriors.aghy) <- c( paste0("iter",1:n_draws))
+colnames(posteriors.agpe) <- c( paste0("iter",1:n_draws))
+colnames(posteriors.elvi) <- c( paste0("iter",1:n_draws))
+
+rownames(posteriors.aghy) <- c("year.aghy")
+rownames(posteriors.agpe) <- c("year.agpe")
+rownames(posteriors.elvi) <- c("year.elvi")
 
 
-samps_int.1 <- tibble(value = t(generate(fit, formula = ~ int.species1, n.samples = 1000)), species = species_names[1]) %>% mutate(quantile = case_when(value >=fit$summary.fixed$`0.025quant`[1] | value <=fit$summary.fixed$`0.975quant`[1] ~ 1))
[…]

### 2024-08-22 Analyses/endo_herbarium_records_merge.R
changes to plot the climate data, but I catually do this in the climate clorrelation script now.

@@ -11,7 +11,9 @@ library(lubridate)
 library(ggmap)
 library(prism) # to import prism raster files
 library(terra)
+library(sf)
 library(INLA) # using this here to work with the shape boundary
+library(inlabru) # using this here for some plotting stuff
 
 ################################################################################
 ############ Read in digitized herbarium records ############################### 
@@ -1667,6 +1669,14 @@ ppt_autumn_cv_difference <- terra::diff(terra::rast(list(ppt_autumn_old_cv, ppt_
 
 
 # changing the crs of these rasters to match the pixels of our mesh
+# converting the lat long to epsg 6703km in km
+# define a crs
+epsg6703km <- paste(
+  "+proj=aea +lat_0=23 +lon_0=-96 +lat_1=29.5",
+  "+lat_2=45.5 +x_0=0 +y_0=0 +datum=NAD83",
+  "+units=km +no_defs"
+)
+
 raster::crs(tmean_annual_difference) <- epsg6703km
 raster::crs(tmean_spring_difference) <- epsg6703km
 raster::crs(tmean_summer_difference) <- epsg6703km
@@ -1701,13 +1711,13 @@ raster::crs(ppt_autumn_cv_difference) <- epsg6703km
 
 
 
-ppt_autumn_cv_difference %>% st_transform(epsg6703km)
-  st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE)
-  st_transform(epsg6703km) %>% 
-  mutate(
-    coords.x1 = st_coordinates(.)[, 1],
-    coords.x2 = = st_coordinates(.)[, 1],
-)
+# ppt_autumn_cv_difference %>% st_transform(epsg6703km)
+#   st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE)
+#   st_transform(epsg6703km) %>% 
+#   mutate(
+#     coords.x1 = st_coordinates(.)[, 1],
+#     coords.x2 = = st_coordinates(.)[, 1],
+# )
 
 
 ######### Next extracting these climate values at our points. ####
@@ -1822,7 +1832,9 @@ aghy_prism_diff_pred_df <- tibble(lon = vrt_aghy_dd@coords[,1], lat = vrt_aghy_d
     coords.x2 = st_coordinates(.)[, 2]
   ) %>% 
   as.data.frame() %>% 
-  mutate(origin = "prism") 
+  mutate(origin = "prism") %>% 
[…]

### 2024-08-22 Analyses/posthoc_climate_correlation.R
remaking the climate trend plot

@@ -431,7 +431,7 @@ svc.pred_climate <- aghy_prism_diff_pred_df %>%
 # Making a version of the dataset that subsamples locations
 svc.pred_climate_subsample <- svc.pred_climate %>% 
 group_by(species) %>%
-sample_n(size =200) %>% 
+sample_n(size =250) %>% 
   ungroup() %>% 
   mutate(Intercept = 1)
 
@@ -446,6 +446,24 @@ svc.pred_climate_subsample_long <- svc.pred_climate_subsample %>%
   # filter(!grepl("tmean_spring", name)) %>% 
   # filter(!grepl("tmean_summer", name)) %>% 
   # filter(!grepl("tmean_autumn", name)) 
+
+
+
+
+
+# Making a version of the dataset that takes pixels only at locations with specimens
+# list of unique locations
+# endo_herb_AGHY$lat
+# 
+# svc.pred_climate_specimens <- svc.pred_climate %>% 
+#   group_by(species) %>% 
+#   
+#   sample_n(size =200) %>% 
+#   ungroup() %>% 
+#   mutate(Intercept = 1)
+
+
+
 
 
 # Looking at the data
@@ -659,6 +677,7 @@ climate_labels <- c(
 
 tmean_trend <- ggplot(filter(prediction_df, grepl("tmean", name) & !grepl("sd", name))) +
   # geom_point(data = filter(svc.pred_climate_subsample_long, grepl("tmean", name) & !grepl("annual", name)), aes(x = value, y = mean, color = species), alpha = 0.2)+
+  geom_vline(aes(xintercept = 0), color = "grey80")+geom_hline(aes(yintercept = 0), color = "grey80")+
   geom_line(aes(value, mean)) +
   geom_ribbon(aes(value, ymin = q0.025, ymax = q0.975, fill = species), alpha = 0.2) +
   geom_ribbon(aes(value, ymin = q0.25, ymax = q0.75, fill = species), alpha = 0.2) +
@@ -682,6 +701,7 @@ tmean_trend
 
 tmean_sd_trend <- ggplot(filter(prediction_df, grepl("tmean", name) & grepl("sd", name))) +
   # geom_point(data = filter(svc.pred_climate_subsample_long, grepl("tmean", name) & !grepl("annual", name)), aes(x = value, y = mean, color = species), alpha = 0.2)+
+  geom_vline(aes(xintercept = 0), color = "grey80")+geom_hline(aes(yintercept = 0), color = "grey80")+
   geom_line(aes(value, mean)) +
[…]

### 2024-08-22 Manuscript/endo_herbarium.tex
updates to manuscript to share with tom and jacob

@@
 We first address questions describing spatial and temporal trends in endophyte prevalence: (i) How has endophyte prevalence changed over the past two centuries?
-and (ii) How spatially heterogenous are temporal trends in endophyte prevalence across eastern North America?
+and (ii) How spatially variable are temporal trends in endophyte prevalence across eastern North America?
 We then address how climate change may be driving trends in endophyte prevalence by asking: (iii) What is the relationship between variation in temporal trends in endophyte prevalence and changes in climate drivers?
@@
 This division of seasons allowed us to quantify differences in climate associated with the two ``cool'' seasons, when we expected our focal species to be most biologically active (\emph{A.
-hyemalis} flowering phenology: Spring; \emph{E.
-virginicus}: Spring and Summer; \emph{A.
-perennans}: Fall).
+hyemalis} flowering phenology: spring; \emph{E.
+virginicus}: spring and summer; \emph{A.
+perennans}: autumn).
 In addition to mean climate conditions, environmental variability itself can influence population dynamics \cite{tuljapurkar_population_1982} and changes in variability are a key prediction of climate change models \cite{stocker2013technical, ipcc_2021}.
@@
 
-To evaluate whether areas that have experienced the greatest changes in endophyte prevalence (hotspots of endophyte change) are associated with high degrees of change in climate (hotspots of climate change), \tom{we modeled spatially varying slopes of endophyte change through time ($\beta_{l}$) as a linear function of environmental covariates, with a Gaussian error distribution.
+To evaluate whether areas that have experienced the greatest changes in endophyte prevalence (hotspots of endophyte change) are associated with high degrees of change in climate (hotspots of climate change), \tom{we modeled spatially varying slopes of endophyte change through time ($\tau_{[h]l}$ as a linear function of environmental covariates, with a Gaussian error distribution.
 }{I think we need to account for uncertainty in the slopes.
 They are outputs of a (quasi) Bayesian model so we should be able to propagate all the uncertainty in the posterior disgtribution.}
-Calculating correlations from many pixels across the study region risks articially inflating confidence in our results due to large sample sizes, and \tom{so we repeated this calculation using only a random subsample of 100 pixels across the study region}{100 seems like a low number to me.
-What if we did this for all of the herbarium collection locations?}.
-\tom{}{Are the methods above repeated for each species separately?}
+Data from each host species was analysed separately. 
+Fitting regressions to many pixels across the study region risks articially inflating confidence in our results due to large sample sizes, and \tom{so we performed this analysis using only a random subsample of 250 pixels across the study region, which provided results qualitatively similar to analysis of the full set of pixels}{100 seems like a low number to me.
+What if we did this for all of the herbarium collection locations?}\josh{.}{I upped the number of points.
+I don't think conceptually we need to only do collection locations, but I can work on that as an alternative.}
+
 \tom{}{I cut the notation for the Gaussian model for now because it is a pretty simple model and the notation may be overkill, plus because I changed your tau's to beta's there were betas on both sides of the tilde, which was confusing/annoying.
@@
 \subsection*{How has endophyte prevalence changed over time?}
-We found that endophyte prevalence increased within the examined specimens over the last two centuries for all three host species (Fig. 4). 
-On average, \emph{A.
-hyemalis} and \emph{E.
-virginicus} both increased from ~30 \% to over 70\% prevalence across the study region, and \emph{A.
-perennans} increased from ~ 15\% to over 70\% prevalence.
-\josh{Our model indicates a higher certainty that overall temporal trends are positive for \emph{A.
-hyemalis} and \emph{A.
-perennans} than for \emph{E.
-virginicus} (99\% probability of a positive overall year slope in \emph{A.
-hyemalis}, 89\% probability of a positive overall year slope in \emph{A.
-perennans}, and 58\% probability of a positive overall year slope in \emph{E.
-virginicus}) .}{These numbers are currently outdated.
-I am making some adjustments to models, and will update with final model}
+We found that endophyte prevalence increased within the examined specimens over the last two centuries for all three host species (Fig. \ref{fig:temporal}). 
+On average, modeling indicated that endophytes of \emph{A.
+perennans} and \emph{E.
+virginicus} increased from $\sim 40$ \% to  $70$\% prevalence across the study region, and that of \emph{A.
+hyemalis} increased from $\sim 25$\% to over $50$\% prevalence.
+Our model indicates a high certainty that overall temporal trends are positive across species (99\% probability of a positive overall year slope in \emph{A.
+hyemalis}, 92\% probability of a positive overall year slope in \emph{A.
+perennans}, and 91\% probability of a positive overall year slope in \emph{E.
+virginicus}) (Fig.
+\ref{fig:temporal_posterior})
 
@@
 	\caption[Temporal trends in endophyte prevalence.]{Temporal trends in endophyte prevalence.
-(A) Histograms show the frequency of collection through time for each host species.
-(B) Colored points are binned means of the observed endophyte presence/absence data (black dashes).
-Colors represent each host species and point size is determined by the number of specimens.
-Lines show predicted mean endophyte prevalence over the study period along with the 50\%  and 95\% CI bands incorporating uncertainty associated with collector and scorer random effects.}
+(A) Histograms show the frequency of scored specimens through time for each host species.
+(B) Lines show predicted mean endophyte prevalence over the study period along with the 50\%  and 95\% CI bands incorporating uncertainty associated with collector and scorer random effects.
+Colored points are binned means of the observed endophyte presence/absence data (black dashes).
+Colors represent each host species and point size represents the number of specimens.}
 	\label{fig:temporal}
@@
 
-\subsection*{How spatially heterogenous are temporal trends in endophyte prevalence?}
-Our model revealed hotspots of change in endophyte prevalence . 
-While there was an overall increase in endophyte prevalence, these changes varied across the host species' ranges (Fig. 3).
-In some regions, posterior estimates of our spatially varying temporal trends, $\tau$, indicate that \emph{A.
+\subsection*{How spatially variable are temporal trends in endophyte prevalence?}
+Our model revealed hotspots of change in endophyte prevalence. 
+While there was an overall increase in endophyte prevalence, these changes varied across the host species' ranges (Fig. \ref{fig:svc_time_map}).
+In some regions, posterior estimates of spatially varying temporal trends, $\tau$, indicate that \emph{A.
 hyemalis} and \emph{A.
-perennans} experienced increases in percent prevalence by as much as 4\% per year over the study period, while  \emph{E.
-virginicus} experienced increases up to around 1.5 \% per year.
-In other regions, there were negligible changes.
-Notably, the symbionts of \emph{E.
-virginicus} experienced only slight increases in prevalence, and were less spatially variable than the other two species.
-\josh{Regions that start with low endophyte prevalence, as in the southwestern portion of the range of \emph{A.
-hyemalis}(Fig.
-A1), also experienced negligible change, suggesting that this may be driven more by the absence of the endophyte.}{more discussion material, but putting it here for now.}
-Predicted trends for \emph{A. perennans} show certain areas of both large increase and of large decrease, however this species, for which we have the fewest samples, has the largest uncertainty.
-The posterior estimates of our spatially varying temporal trends, indicate relatively narrow certainty (\josh{need to compute}{}).
+perennans} experienced increases in percent prevalence by as much as $2$\% per year over the study period, while  \emph{E.
+virginicus} experienced increases up to around $1$\% per year.
+Compared to \emph{E.
+virginicus}, which had a weaker overall increase in endophytes and less spatial variability, maps of both \emph{Agrostis} species show areas of strong increase and areas of declining prevalence.
+Notably, endophytes increased towards the western range edge of \emph{A.
+hyemalis} (Fig.
+\ref{fig:svc_time_map}A) and across the northeastern US for \emph{A.
+perennans} (Fig.
+\ref{fig:svc_time_map}B).
+Posterior estimates of uncertainty in spatially varying slopes indicate that these hotspots of change may have experienced increases of up to $5$\% per year while declines in prevalence may be as great as $4$\% per year for \emph{A.
+hyemalis} and \emph{A.
+perennans}.
+For \emph{E. virginicus}, uncertainty ranges between $3.5$\% increases and $2.5$\% decreases (Fig. \ref{fig:svc_time_map_CI}).
 
 \begin{figure}[H]
-	\label{fig:svc_time_map}
 	\centering
-	\includegraphics[width = .6\linewidth]{svc_time_map.png}
+	\includegraphics[width = \linewidth]{../Plots/svc_time_map.png}
 	\caption[Predicted posterior mean of spatially-varying slopes representing change in endophyte prevalence for each host species.]{Predicted posterior mean of spatially-varying slopes representing change in endophyte prevalence for each host species.
 Color indicates the relative change in predicted endophyte prevalence.}
-\end{figure}
-
-\subsection*{Assessing collector and scorer influences on predicted endophyte prevalence}
-We quantified temporal and spatial trends in endophyte prevalence while accounting for potential biases introduced by collectors and by individuals who quantified endophyte presence/absence with the use of random effects.
-We found no evidence that collector biases influenced our results. 
-Collector random effects were consistently small; Fig 4A, and models fit with and without this random effect provide qualitatively similar results.
-The identity of individual scorers did contribute to observed patterns in endophyte prevalence.
-For example, 3 of the 16 scorers were more likely than average to assign positive endophyte status, as indicated by 95\% credible intervals that do not overlap 0) (Fig 4B). 
-However, this may have been driven by differences in scorers biases during the seed scoring process, or by unintended spatial clustering of the specimens scored by each scorer. 
-Interpreting our models with the inclusion of the scorer effect thus provides conservative estimates of the absolute magnitude of changes in endophyte prevalence.
-
-\begin{figure}[H]
-	\label{fig:random_fx}
-	\centering
-	\includegraphics[width = .8\linewidth]{random_fx_plot.png}
-	\caption{\textbf{Posterior estimates of (A) collector and (B) scorer random effects.} Points show the posterior mean along with 95\% CI for random effects estimate from 532 collectors and 16 scorers.}
+	\label{fig:svc_time_map}
 \end{figure}
@@
 
-\begin{figure}[H]
-	\label{fig:svc_space_map}
-	\centering
-	\includegraphics[width = .6\linewidth]{svc_space_map.png}
-	\caption{\textbf{Mean predicted endophyte prevalence for each host species (columns) in 1925 (top row) and 2020 (bottom row)}.
+%\begin{figure}[H]
+%	\label{fig:svc_space_map}
+%	\centering
+%	\includegraphics[width = .6\linewidth]{svc_space_map.png}
+%	\caption{\textbf{Mean predicted endophyte prevalence for each host species (columns) in 1925 (top row) and 2020 (bottom row)}.
 Color indicates mean predicted rate of endophyte prevalence across the predicted distribution of each species.}
-\end{figure}
+%\end{figure}
 
@@
 We found that trends in endophyte prevalence were strongly associated with seasonal climate change drivers (Fig. \ref{fig:climate_correlation_plot}).
-For the majority of the study region, the climate has become wetter and cooler over the last century (Fig. A7-A8), a consequence of regional variation in global climate change \cite{ipcc_2021}. 
-Within the study region, spatially heterogeneous environmental changes were predictive of changes in endophyte prevalence. 
-For example, strong increases in prevalence within \emph{E.
-virginicus} were most associated with declines in Summer precipitation  (a negative correlation in Fig.
-7) as well as with increases in the year-to-year variability of annual temperature (a positive correlation in Fig.
-7).
-Changes were also associated with reductions in average annual temperatures, and increases in year-to-year temperature variability.
-\emph{A. perennans} endophyte prevalence increased most strongly in regions that experienced reduced spring precipitation and reduced variability in annual temperature.
-Although these correlations were weaker, changes in \emph{A.
-perennans} endophyte prevalence were also associated with increased in increases in annual precipitation and increasing autumn precipitation.
-For \emph{A. hyemalis}, endophyte prevalence increased most strongly in regions that experienced reductions in autumn precipitation variability. 
-Correlations using only a subsampling of pixels were qualitatively similar to these results (Fig. A11), suggesting that the patterns we find are not spurious associations.
+For the majority of the study region, the climate has become wetter and cooler over the last century (Fig.
+\ref{fig:AGHY_climate_covariates}-\ref{fig:ELVI_climate_covariates}), a consequence of regional variation in global climate change \cite{ipcc_2021}.
+Within the study region, spatial variation in climate trends were predictive of trends in endophyte prevalence.
+For example, strong increases in prevalence within \emph{A. perennans} were most associated with autumn climate drivers that coincide with its Aug-Sep active growing season. 
+For this species, warmer and wetter autumn climates showed particularly strong relationships, however other seasonal drivers may also contribute to increasing endophyte prevalence (drier springs and cooler summers).
+Trends in endophyte prevalence for \emph{A. hyemalis} were most strongly associated with changes in precipitation variability were the strongest predictors.
+Endophyte prevalence increased the most in regions that experienced greater spring precipitation along with increasing variability in summer and autumn precipitation. 
+While this species actively grows and reproduces in the late spring and early summer, climate effects outside of the growing season may indicate that endophytes play a role in persistence during dormant periods through summer droughts or contribute to the ability to successfully germinate.
+Prevalence of endophytes of \emph{E. virginicus} were least influenced by climate, but decreasing autumn temperature variability and less precipitation in autumn were the strongest predictors.
 
+Correlations assessed using all pixels across each species distribution were qualitatively similar to these results (Fig. A11).
 %These associations are in line with expectations that endophyte symbiosis provides drought tolerance to their hosts, particularly during \emph{E.
@@
 Or could this result have occurred by chance?}
-\josh{}{Only have plotted results for AGHY right now.}
+
 \begin{figure}[H]
 	\centering
-	\includegraphics[width = .7\linewidth]{../Plots/climate_trends_plot.png}
-	\caption{\textbf{Correlations between changes in climate drivers and changes in endophyte prevalence.} Lines show relationships between seasonal climate covariates and spatially-varying trends in endophyte prevalence for each host species, along with 96\% CI.}
+	\includegraphics[width = \linewidth]{../Plots/climate_trends_plot.png}
+	\caption{\textbf{Relationships between changes in seasonal climate drivers and predicted marginal trends in endophyte prevalence.} Lines show relationships between seasonal climate (A: mean temperature, B: cumulative precipitation, C: standard deviation in temperature, D: standard deviation in precipitation) and spatially-varying trends in endophyte prevalence for each host species, along with 50 and 95\% CI.}
 	\label{fig:climate_correlation_plot}
@@
 \subsubsection*{Performance on test data}
-We found that while the model predicts broader regional trends in endophyte prevalence present in the contemporary survey data such as declining endophyte prevalence towards western longitudes in \emph{A.
+We found that model performance, as judged by AUC, was similar between historic herbarium specimens used as training data and the out-of-sample test data from contemporary surveys (0.79 and 0.77 respectively; Fig.
+\ref{fig:ROCtest}-\ref{fig:ROCtraining}).
+The model successfully captured broader regional trends in endophyte prevalence present in the contemporary survey data, such as decline endophyte prevalence towrads western longitudes in \emph{A.
 hyemalis}  (Fig.
-\ref{fig:contemptestplot} B-C), however the contemporary data contains additional variability at smaller scales not captured by our sampling of herbarium specimens.
+\ref{fig:contemptestplot}A).
+However, the contemporary data contains additional variability at smaller scales not captured by our sampling of herbarium specimens. 
 We interpreted this to mean that the model captured regional spatial dynamics, but underpredicts local scale dynamics.
@@
 
+\subsection*{Assessing collector and scorer influences on predicted endophyte prevalence}
+Our modeling effort quantified temporal and spatial trends in endophyte prevalence while accounting for potential biases introduced by collectors and by individual scorers who quantified endophyte presence/absence within specimens with the use of random effects.
+We found no evidence that collector biases influenced our results. 
+Collector random effects were consistently small (Fig. \ref{fig:collector_fx}), and models fit with and without this random effect provide qualitatively similar results.
+The identity of individual scorers did contribute to observed patterns in endophyte prevalence.
+For example, $3$ of the $25$ scorers were more consistently likely than average to assign positive endophyte status, as indicated by 95\% credible intervals greater than zero) (Fig.
+\ref{fig:scorer_fx}).
+This may have been driven by differences in scorers biases during the seed scoring process or by unintended spatial clustering of the specimens scored by each scorer \cite{clayton1993spatial,urdangarin2023evaluating}.
+Interpreting our models with the inclusion of variance associated with the scorer effect thus provides conservative estimates of the absolute magnitude of changes in endophyte prevalence.
+
 \section*{Discussion}
@@
 Differences between the responses of each host species underscore that while all of these $C_3$ grasses share similar broad-scale distributions, each engages in unique biotic interactions and has unique niche requirements.
-We identified hotspots of change for \emph{A. perennans}, which experienced the strongest absolute changes in endophyte prevalence (Fig. 5).
+We identified hotspots of change for \emph{A. perennans}, which experienced the strongest absolute changes in endophyte prevalence (Fig. \ref{fig:svc_time_map}).
 Declines in the southern portion of its range and increases in the north suggest a potential poleward range shift of endophytic plants. 
 Based on previous work demonstrating that endophytes can shield their hosts from drought stress \cite{decunta2021systematic}, we generally predicted that drought conditions could be a driver of increasing endophyte prevalence.
-In line with this expectation, increasing prevalence for this species was associated with decreasing precipitation, most strongly with autumn-season declines (Fig. 7). 
-\emph{A. perennans} typically blooms in the autumn.
-Endophytes could be playing a role helping hosts weather autumn-season droughts while the species is dormant. 
-It may be useful to investigate whether lagged climate effects are important predictors of host fitness in this system \cite{evers2021lagged}. 
+In contrast to this expectation, increasing prevalence for this species was associated with increasing autumn temperature and precipitation (Fig. \ref{fig:climate_correlation_plot}). 
 To our knowledge, the response of the symbiosis in \emph{A.
 perennans} to drought has not been examined experimentally, but in a greenhouse experiment, endophytes had a positive effect on host reproduction under shaded, low-light conditions \citep{davitt2010costs}.
-\emph{Epichloë} endophytes have been connected to a suite of non-drought related fitness benefits including herbivore protection \citep{brem2001epichloe}, salinity resistence \citep{wang2020effects}, and mediation of the soil microbiome \citep{roberts2015rhizosphere}
-These effects are potentially mediated by the diverse bioactive alkaloids and other signaling compounds they produce \citep{saikkonen2013chemical}.
-The strong increase in symbiotic \emph{A. perennans} could be explained, at least in part, by these diverse benefits. 
-\emph{A. hyemalis} experienced more consistently positive increases in endophyte prevalence related to changes in spring temperature and precipitation.  
-This result is in line with previous work demonstrating drought benefits in a greenhouse manipulation with this species \citep{davitt2011understanding} that led us to expect  that endophyte prevalence should similarly increase at a greater rate in regions that have experienced increasing drought.
+Our results also hint that it may be useful to investigate whether lagged climate effects are important predictors of host fitness in this system \cite{evers2021lagged}.
+Endophyte prevalence of the spring-flowering \emph{A. hyemalis} was most strongly linked to increasing variability in precipitation across summer and autumn.  
+Endophytes could be playing a role helping hosts weather autumn-season droughts while the species is dormant.
+Previous work has demonstrating drought benefits in a greenhouse manipulation with this species \citep{davitt2011understanding}, and early life stages may be particularly vulnerable to prolonged droughts.
 %It is possible that the study region has not experienced a magnitude of climate change great enough to cause larger changes in endophyte prevalence for this species. 
 %Weak associations with drought could also be explained by climate-driven changes in the rate of imperfect transmission (the generation of non-symbiotic offspring from symbiotic hosts), which could counterbalance endophyte-mediated fitness benefits \citep{donald2021context}.
-For \emph{E.
-virginicus}, which experienced the most modest changes in endophte prevalence overall, we found a strong relationship between temporal trends and  changes in the mean and variability of temperature, as well as with decreases in summer precipitation.
+For \emph{E. virginicus}, which experienced the most modest changes in endophte prevalence overall, we only modest associations with changes in climate drivers. 
 Surveys by \citet{sneck2017variation}, used as part of the test data in this study, identified a drought index (SPEI) that integrates precipitation with estimated evapotranspiration as an important predictor of endophyte prevalence.
+\emph{Epichloë} endophytes have also been connected to a suite of non-drought related fitness benefits including herbivore protection \citep{brem2001epichloe}, salinity resistence \citep{wang2020effects}, and mediation of the soil microbiome \citep{roberts2015rhizosphere}
+These effects are potentially mediated by the diverse bioactive alkaloids and other signaling compounds they produce \citep{saikkonen2013chemical}.
+Increases in symbionts could be explained, at least in part, by these diverse benefits that may help hosts weather a world made increasingly stressful by changes in climate and other anthropogenically introduced stressors.
 While we show consistent increasing trends in prevalence between the three species, the mechanisms that explain these changes may be diverse and idiosyncratic.
@@
 Endophytes fitness benefits potentially explain the spatial distribution of prevalence by allowing their hosts to persist in environments where they otherwise could not \citep{afkhami2014mutualist, kazenel2015mutualistic}.
-For example, fitness benefits of the symbiosis could explain high predicted prevalence in \emph{E.
-virginicus} towards the north or in \emph{A.
-hyemalis} towards its range center coinciding with strong environmental gradients.
-Previous population surveys for endophytes, which were used as test data for our model, found similar latitudinal trends in prevalence in these species \citep{sneck2017variation,rudgers2009benefits}, but at smaller scales.
-While the model recreated these large-scale spatial trends, test data was more variable. 
-Using test data to validate our model predictions allows us to evaluate places to improve the model's ability to perform well at out-of-sample prediction, which will be particularly important for predicting host and symbiont niche-shifts under future climate change.
+For example, fitness benefits of the symbiosis could explain historically low prevalence in \emph{A. hyemalis} towards its western range edge coinciding with a strong aridity gradient.
+Previous population surveys for endophytes, which were used as test data for our model, found similar regional trends in prevalence for endophyte host species \citep{sneck2017variation,rudgers2009benefits}.
+While the model recreated these large-scale spatial trends, test data contained more population-to-population variability in prevalence. 
+Validating our model predictions in this way allows us to evaluate places to improve the model's out-of-sample predictive ability, which will be particularly important for predicting host and symbiont niche-shifts under future climate change.
 Lack of information on local variability may simply be a feature of data derived from herbarium specimens. 
-Even though they are samples from local populations, they are single specimens that are aggregated over in broad-scale model estimates.
+They are samples from local populations, but they are single specimens that are aggregated over in broad-scale model estimates.
 Poor predictive ability at local scales in this grass-endophyte system is not surprising, as previous studies have found that local variation, even to the scale of hundreds of meters can structure endophyte-host niches \cite{kazenel2015mutualistic}.
-\citet{sneck2017variation} also identified host genotype as an important predictor of endophyte prevalence in \emph{E. virginicus}.
-Other studies have found factors including land-use history \cite{vikuk2019infection} and the biotic environment, including herbivory \cite{rudgers2016long}, to be important predictors of endophyte ecology.
+Other studies have found factors including land-use history \cite{vikuk2019infection} and the biotic environment, including herbivory \cite{rudgers2016long}, and host genotype \citet{sneck2017variation}, to be important predictors of endophyte ecology.
 Incorporating available climatic and soil layers as covariates is an obvious first step that could improve predictions.
-Towards the goal of predicting the dynamics of microbial symbioses under climate change, models that integrate data from local and regional scales would be an important step to bridge the gap that often exists between large but broad bioclimatic and biodiversity data and small but local data on biotic interactions \cite{miller2019recent, isaac2020data}.
+Another important step would be integrating data from local and regional scales  through modeling to constrain estimates of local and regional variation.
+These steps will bridge gaps that often exist between large but broad bioclimatic and biodiversity data and small but local data on biotic interactions, and move towards the goal of predicting the dynamics of microbial symbioses under climate change \cite{miller2019recent, isaac2020data}.
 
@@
 The responses of microbial symbioses are a rich target for future studies within museum specimens, particularly those that take advantage of advances in sequencing technology.
-While we used relatively coarse presence/absence data based on fungal morphology, other studies have examined historic plant microbiomes using molecular sequencing and sophisticated bioinformatics techniques, but these studies have so far been limited to relatively few specimens at limited spatial extents \cite{yoshida2015computational, heberling2019utilizing, bieker2020metagenomic, gross2021hidden, bradshaw2021global}.
-Continued advances in capturing historic DNA and in filtering out potential contamination during specimen storage \citep{daru2019novel, bakker2020herbarium, raxworthy2021mining} will be imperative in the effort to scale up these efforts.
+While we used relatively coarse presence/absence data based on fungal morphology, other studies have examined historic plant microbiomes using molecular sequencing and sophisticated bioinformatics techniques, but these studies have so far been limited to relatively few specimens at limited spatial extents \cite{yoshida2015computational,heberling2019utilizing, bieker2020metagenomic,gross2021hidden,bradshaw2021global}.
+Continued advances in capturing historic DNA and in filtering out potential contamination during specimen storage \citep{daru2019novel,bakker2020herbarium,raxworthy2021mining} will be imperative in the effort to scale up these efforts.
 This scaling up will be essential to be able to quantify changes not just in the prevalence of symbionts, but also in symbionts' intraspecific variation and evolutionary responses to climate change, as well as  in changes in the wider microbial community.
-Answering these questions as well as the unknown questions that future researchers may ask also reiterates the value in capturing meta-information during ongoing digitization efforts at herbaria around the world and during the accession of newly collected specimens \citep{lendemer2020extended}.
+Answering these questions as well as the unknown questions that future researchers may ask also reiterates the value in capturing meta-information during ongoing digitization efforts at herbaria around the world and during the accession of newly collected specimens \citep{lendemer2020extended,edwards2024university}.
 Second, we accounted for several potential biases in the data observation process that may be common to many collections-based research questions by using a spatially-explicit random effects model. 
@@
 Previous studies that have quantified the effects of collector biases typically find them to be small  \cite{davis2015herbarium,meineke2019herbarium}, and we similarly did not find that collector has a strong effect on the results of our analysis.
-Fitting this model in a Bayesian framework allows for full propagation of uncertainty.
+%Fitting this model in a Bayesian framework allows for full propagation of uncertainty.
 
@@
 	\section*{Statement of Authorship}
+J.C.F. contributed to research conception, data collection, data analysis, and led manuscript drafting.
+J.M. contributed to data analysis and manuscript manuscript revisions.
+T.E.X.M. contributed to research conception, data collection, data analysis, and manuscript revisions.
 
@@
 		\caption{\textbf{ROC plot showing model performance classifying observations according to endophyte status within the in-sample data.} The curves show adequate model performance for observed data.
-The AUC value is 0.77.}
+The AUC value is 0.79.}
 				\label{fig:ROCtraining}
@@
 	\centering
+	\includegraphics[width = \linewidth]{../Plots/posterior_plot.png}
+	\caption[Posterior estimates of parameters describing global temporal trends in endophyte prevalence trends.]{Density curves show the probability density along with mean (colored line) and $95$\% CI (black lines) for the (A) intercept and (B) slope terms, \textbf{A} and \textbf{T} respectively.
+Colors represent each host species}
+	\label{fig:temporal_posterior}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width = \linewidth]{../Plots/svc_time_map_CI.png}
+	\caption[Uncertainty in spatially-varying trends representing change in endophyte prevalence for each host species.]{Shading represents the range of the 95\% posterior credible interval for spatially varying slopes, $\tau$.}
+	\label{fig:svc_time_map_CI}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
 	\includegraphics[width = \linewidth]{fixed_plot.png}
 	\caption{\textbf{Comparison of posterior estimates of fixed effects when using Liberal or Conservative endophyte scores.}}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width = .8\linewidth]{../Plots/collector_posterior_plot.pdf}
+	\caption{\textbf{Posterior estimates of collector random effects.} Density curves show the posterior estimate along lines indicating the 95\% CI for 12 randomly selected collectors.}
+		\label{fig:collector_fx}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width = .8\linewidth]{../Plots/scorer_posterior_plot.pdf}
+	\caption{\textbf{Posterior estimates of scorer random effects.} Density curves show the posterior estimate along lines indicating the 95\% CI for 25 scorers.}
+	\label{fig:scorer_fx}
 \end{figure}
@@
 	\label{fig:ELVI_climate_covariates}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width = \linewidth]{prevalence_map_CI.png}
-	\caption{\textbf{Uncertainty associated with spatial trends in endophyte prevalence.} Color represents change in predicted endophyte prevalence.
-Panels show upper and lower 95\% posterior probability for each host species between 1925 and 2020.}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width = \linewidth]{climate_corr_heatmap_subsample.png}
-	\caption{\textbf{Correlations between changes in climate drivers and changes in endophyte prevalence from a random sample of 100 pixels across the study region.} Color denotes the Spearman correlation coefficient between the relative rate of change in endophyte prevalence and the change in annual mean temperature ($^oC$) and total annual and seasonal precipitation (mm), as well as the change in the coefficient of variation of each climate driver.
-Positive correlation coefficients indicate that greater increases in a climate driver were associated with larger increases in endophyte prevalence, while negative values indicate that .
-Asteriks denote correlation coefficients $> .3$ or $< -.3$.}
-	\label{fig:FigA5}
 \end{figure}

### 2024-09-02 Manuscript/endo_herbarium.tex
Updated host species SDM with SD

@@
 \usepackage[T1]{fontenc}
+
+\usepackage[dvipsnames]{xcolor}
 
@@
 	\noindent{} 1. Rice University, Department of BioSciences, Houston, Texas 77006;
-	\noindent{} \jacob{1}{I think this is should be 2}. University of Miami, Department of Biology, Miami, Florida;
+	\noindent{} 2. University of Miami, Department of Biology, Miami, Florida;
 
@@
 We modeled the distribution of each host species to generate maps on which we predicted the dynamics of \emph{Epichloë} symbionts.
-We followed the ODMAP (overview, data, model, assessment, prediction) protocol \citep{crossley2022opposing}, using \josh{presence-only observations of the host species from GBIF between 1990 to 2020 }{How many data points does this end up being?}.
-To reduce the potential influence of sampling bias and spatial autocorrelation, we thinned the occurrences to the spatial scale \josh{()}{I think this is 4km, but need to check} of our selected climatic predictors.
+We followed the ODMAP (overview, data, model, assessment, prediction) protocol \citep{crossley2022opposing}. 
+ A full description of the ODMAP can be found in the  (Supplementary Method \ref{sec:sdm }). 
+ In short, we used presence-only observations of the host species from Global Biodiversity Information Facility (GBIF) between 1990 to 2020 .
+To reduce the potential influence of sampling bias and spatial autocorrelation, we thinned the occurrences to the spatial scale of our selected climatic predictors. 
 We selected climate variables that aligned with our analysis of climatic influences on trends in endophyte prevalence described above.
-We calculated the mean and standard devation of seasonal temperature and precipitation across 1990 to 2020. 
-Among this suite of variables, we chose to include \josh{mean spring temperature, mean spring precipitation, and mean summer temperature}{is this list accurate?
-my re-running my climate analysis means we ought to re-run the sdm's, and use SD instead of CV, so we can update this list based on whatever is appropriate}, which were uncorrelated (Variance Inflation Factor $>$ 0.7) and allowed us to predict the occurrence probabilty of each host species in space and time.
+These climatic variables are: the mean and standard deviation of spring, summer, and autumn temperature, as well as the mean and standard deviation of spring, summer, and autumn cumulative precipitation, cumulative precipitation, and cumulative precipitation.
+%We calculated the mean and standard devation of seasonal temperature and precipitation across 1990 to 2020. 
+Among this suite of variables, we chose to include mean summer temperature, standard deviation of spring precipitation, standard deviation of summer precipitation,  standard deviation of spring  precipitation, standard deviation of summer  precipitation, which were uncorrelated (Variance Inflation Factor $>$ 0.7) and allowed us to predict the occurrence probability of each host species in space and time.
 We fit maximum entropy (MaxEnt) models using the maxent function in the package dismo \citep{hijmans2017package}. 
-\josh{MaxEnt is preferred because it has been shown to generate response curves with less unpredictable behavior when applied to new climates \citep{hijmans2006ability}.}{possibly could remove this sentence?}
+%\josh{MaxEnt is preferred because it has been shown to generate response curves with less unpredictable behavior when applied to new climates \citep{hijmans2006ability}.}{possibly could remove this sentence?}
 We generated 10,000 pseudo-absences as background points, and split the occurrence data into 75\% for model training and 25\% for model testing.
-The performance of models was evaluated with AUC \josh{AUC}{can we add these values?} \citep{jimenez2012insights}. 
+The performance of models was evaluated with AUC  \citep{jimenez2012insights}. 
+We found AUC = 0.862, AUC = 0.838, AUC = 0.821 respectively  for \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}.
+
 To convert the continuous predicted probabilities into binary presence - absence maps, we used the  training sensitivity (true positive rate) and specificity threshold (true negative rate) \citep{liu2005selecting}.
-These binary maps serve as boundaries in presented maps of change in endophyte prevalence, and outline the set of pixels used in our analysis of climate correlates with trends in endophyte prevalence
+These binary maps serve as boundaries in presented maps of change in endophyte prevalence, and outline the set of pixels used in our analysis of climate correlates with trends in endophyte prevalence.
 
@@
 
+	\section*{Supporting Methods}
+	\subsection*{ODMAP Protocol}
+\label{sec:sdm}
+{\color{NavyBlue}{Overview}}\\
+\textbf {Model  purpose}: Mapping current distribution of epichloë host species. \\
+\textbf {Target species}: \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}. \\
+\textbf {Study area}: Eastern North America \\
+%Scale of analysis
+\textbf {Spatial extent}: -125.0208, -66.47917, 24.0625, 49.9375 (xmin, xmax, ymin, ymax).\\
+\textbf {Spatial resolution}: 0.04166667, 0.04166667 (x, y).\\
+\textbf {Temporal extent}: 1990 to 2020.\\
+\textbf {Boundary}: Natural.\\
+{\color{NavyBlue}{Data}}\\
+\textbf {Observation type}: Occurrence records from  Global Biodiversity Information Facility and herbarium collection across eastern North America.
+We used 713 occurrences records for \emph{Agrostis hyemalis}, 656 occurrence records for \emph{Agrostis perennans} and 2338 for \emph{Elymus virginicus}.\\
+\textbf{Response data type}: occurrence record, presence-only.\\
+\textbf{Coordinate reference system}: WGS84 coordinate reference system (EPSG:4326 code)\\
+\textbf{Climatic data}:  raster data extracted from PRISM  \\
+{\color{NavyBlue}{Model }}\\
+\textbf{Model assumption}: We assumed that the target species are at equilibrium with their environment. \\
+\textbf{Algorithms}: Maximum entropy (maxent)\\
+\textbf{Workflow}: We  described the workflow in the method section of the manuscript. \\
+\textbf{Software}:  All statistics were performed using Maxent 3.3.4 and R4.3.1 with packages terra, usdm, spThin and dismo.\\
+\textbf{Code availability}: https://github.com/joshuacfowler/EndoHerbarium\\
+\textbf{Data availability}: https://github.com/joshuacfowler/EndoHerbarium\\
+{\color{NavyBlue}{Assessment}}\\
+We used AUC to test model performance.\\
+{\color{NavyBlue}{Prediction }}\\
+We predicted the probability of presence of the host species as a binary maps (presence or absence)
+
 	% In most cases, authors should typeset supplementary material in a separate,

### 2024-09-02 sdm.R

@@ -216,9 +216,9 @@ test_aghy_a <- predict(mod_aghy, env_aghy_bg)
 e_aghy <- evaluate(p=test_aghy_p, a=test_aghy_a)
 threshold(e_aghy)
 
-par(mfrow=c(1, 3))
+par(mfrow=c(1, 2))
 # density(e_aghy)
-boxplot(e_aghy, col=c('blue', 'red'))
+# boxplot(e_aghy, col=c('blue', 'red'))
 plot(e_aghy, 'ROC')
 # plot(e_aghy, 'TPR')
 plot(e_aghy, 'kappa')
@@ -488,7 +488,7 @@ dev.off()
 
 # pdf("/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Figures/FigSDMMaxent.pdf",height = 5,width = 10)
 # # layout(mat = layout.matrix)
-# op<-par(mfrow = c(2,3), mar=c(3,3,2,1),oma = c(5, 5, 1, 3))
+# op<-par(mfrow = c(1,3), mar=c(3,3,2,1),oma = c(5, 5, 1, 3))
 # plot(map_old_aghy,xaxt='n',main=substitute(paste(italic("A. hyemalis"))))
 # text(x=-80,y=50,label="1925")
 # # par(mar=c(2.2,1,3,1))
@@ -521,7 +521,10 @@ plot(aghy)
 # layout.matrix <- matrix(c(1, 2, 3, 4,5,6), nrow = 2, ncol = 3)
 pdf("/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Figures/FigSDM_Presence_absence.pdf",height = 8,width = 5)
 # layout(mat = layout.matrix)
-op<-par(mfrow = c(3,1), mar=c(4,8,1.5,2))
+op<-par(mfrow = c(1,3), mar=c(4,8,1.5,2))
+plot(e_aghy, 'ROC')
+plot(e_agpe, 'ROC')
+plot(e_elvi, 'ROC')
 # plot(map_old_aghy > tr_aghy,xaxt='n',main=substitute(paste(italic("A. hyemalis"))))
 # text(x=-80,y=50,label="1925")
 # # par(mar=c(2.2,1,3,1))
@@ -559,6 +562,7 @@ op<-par(mfrow = c(3,1), mar=c(4,8,1.5,2))
 # par(mar=c(5,3,0,1))
[…]

### 2024-09-02 sdm.R
Added more comments in sdm code

@@ -572,13 +572,15 @@ mtext("C",side = 3, adj = 0,cex=1.25)
 par(op)
 dev.off()
 
-aghy_binary <- aghy >  0.5102097
-agpe_binary <-agpe > 0.52849
-elvi_binary <-elvi > 0.5422173
+aghy_binary <- aghy >  0.5410866
+agpe_binary <-agpe > 0.452535
+elvi_binary <-elvi > 0.5163273
 
-# writeRaster(aghy_binary, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/github/EndoHerbarium/aghy_binary.tif')
-# writeRaster(agpe_binary, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/github/EndoHerbarium/agpe_binary.tif')
-# writeRaster(elvi_binary, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/github/EndoHerbarium/elvi_binary.tif')
+
+ 
+writeRaster(aghy_binary, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/github/EndoHerbarium/aghy_binary.tif')
+writeRaster(agpe_binary, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/github/EndoHerbarium/agpe_binary.tif')
+writeRaster(elvi_binary, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/github/EndoHerbarium/elvi_binary.tif')
 
 
 graph<-list("A","B","C")

### 2024-09-04 Analyses/posthoc_climate_correlation.R
stashing my updates to the climate analysis after re-running with jacob's maps

@@ -440,13 +440,13 @@ sample_n(size =250) %>%
 #   facet_wrap(~species)
 
 
-svc.pred_climate_subsample_long <- svc.pred_climate_subsample %>% 
-  pivot_longer(cols = contains("_diff")) %>% 
-  filter(!grepl("cv", name)) 
-  # filter(!grepl("tmean_spring", name)) %>% 
-  # filter(!grepl("tmean_summer", name)) %>% 
-  # filter(!grepl("tmean_autumn", name)) 
-
+# svc.pred_climate_subsample_long <- svc.pred_climate_subsample %>% 
+#   pivot_longer(cols = contains("_diff")) %>% 
+#   filter(!grepl("cv", name)) 
+#   # filter(!grepl("tmean_spring", name)) %>% 
+#   # filter(!grepl("tmean_summer", name)) %>% 
+#   # filter(!grepl("tmean_autumn", name)) 
+# 
 
 
 
@@ -512,19 +512,34 @@ tmean_sd_regression_plot
 
 
 ######### perform a regression with INLA to test the relationship between change in prevalence and change in seasonal climate ####
-cmp <- ~ Intercept(main = Intercept, model = "linear", mean.linear=0, prec.linear=0.001) + spring_ppt(main = ppt_spring_diff, model = "linear", mean.linear=0, prec.linear=0.001) + summer_ppt(main = ppt_summer_diff, model = "linear", mean.linear=0, prec.linear=0.001) + autumn_ppt(main = ppt_autumn_diff, model = "linear", mean.linear=0, prec.linear=0.001) +
-         spring_ppt_sd(main = ppt_spring_sd_diff, model = "linear", mean.linear=0, prec.linear=0.001) + summer_ppt_sd(main = ppt_summer_sd_diff, model = "linear", mean.linear=0, prec.linear=0.001) + autumn_ppt_sd(main = ppt_autumn_sd_diff, model = "linear", mean.linear=0, prec.linear=0.001) +
-         spring_tmean(main = tmean_spring_diff, model = "linear", mean.linear=0, prec.linear=0.001) + summer_tmean(main = tmean_summer_diff, model = "linear", mean.linear=0, prec.linear=0.001) + autumn_tmean(main = tmean_autumn_diff, model = "linear", mean.linear=0, prec.linear=0.001) +
[…]

### 2024-09-04 Manuscript/endo_herbarium.tex

@@
 Documented responses to environmental change, such as shifts in species' distributions \cite{aitken2008adaptation} and phenology \cite{piao2019plant}, are typically blind to concurrent changes in associated biotic interactions.
-Empirically evaluating these biotic changes -- whether interacting species shift in tandem with their partners or not \cite{hillerislambers2013will} -- is crucial to predicting the reogranization of Earth's biodiversity under global change.
+Empirically evaluating these biotic changes -- whether interacting species shift in tandem with their partners or not \cite{hillerislambers2013will} -- is crucial to predicting the reorganization of Earth's biodiversity under global change.
 Such evaluations have been limited because few datasets on species interactions extend over sufficiently long time scales of contemporary climate change \cite{poisot2021global}.
@@
 
-Understanding of how microbial symbioses are affected by climate change is additionally complicated by spatial hetereogeneity in the direction and magnitude of environmental change \cite{ipcc_2021}. 
+Understanding of how microbial symbioses are affected by climate change is additionally complicated by spatial heterogeneity in the direction and magnitude of environmental change \cite{ipcc_2021}. 
 Beneficial symbionts are likely able to shield their hosts from environmental stress in locations that experience a small degree of change, but symbionts in locations that experience changes of large magnitude may be pushed beyond their physiological limits \cite{webster2008temperature}.
@@
 Both \emph{Agrostis} species are recorded from throughout the Eastern US, but \emph{A.
-perennans} has a slighty more northern distribution, whereas \emph{A.
+perennans} has a slightly more northern distribution, whereas \emph{A.
 hyemalis} is found rarely as far north as Canada and is listed as a rare plant in Minnesota.
@@
 Collections were geo-referenced to the nearest county centroid, or nearest municipality when that information was available. 
-For a few of the oldest specimens, only information at the state level was available, and so we used the state centoid.
+For a few of the oldest specimens, only information at the state level was available, and so we used the state centroid.
 
@@
 To appropriately account for the spatial non-independence of geo-referenced \josh{occurrences}{spelling?}, we used an approximate Bayesian method, Integrated Nested Laplace Approximation (INLA), to construct spatio-temporal models of endophyte prevalence.
-INLA provides a computationally efficient method of ascertaining parameter posterior distributions for certain models that can be formulated as latent Gausssian Models \cite{rue2009approximate}.
+INLA provides a computationally efficient method of ascertaining parameter posterior distributions for certain models that can be formulated as latent Gaussian Models \cite{rue2009approximate}.
 Many common statistical models, including structured and unstructured mixed-effects models, can be represented as latent Gaussian Models.
-We incorporated spatial heterogeneity into this analysis using spatially-structured intercept and slope parameters implemented as stochastic partial differential equation (SPDE) approximations of a continuous spatial Guassian process.
+We incorporated spatial heterogeneity into this analysis using spatially-structured intercept and slope parameters implemented as stochastic partial differential equation (SPDE) approximations of a continuous spatial Gaussian process.
 This SPDE approach is a flexible method of smoothing across space while explicitly accounting for spatial dependence between data-points \citep{lindgren2011explicit,bakka2018spatial}.
@@
 Shared variance terms included the spatially-dependent random effect $\delta_{l[i]}$, intended to account for residual spatial variation, and $\chi_{c[i]}$ and $\omega_{s[i]}$ i.i.d.-random effects indexed for each collector identity ($c[i]$), and scorer identity ($s[i]$) of the $i^{th}$ specimen.
-
 \begin{subequations}
@@
 		logit(\hat{P}_{[h_{2}]i}) = \mathrm{A}_{[h_{2}]i} + \mathrm{T}_{[h_{2}]}*year_i  + \alpha_{[h_{2}]l[i]} + \tau_{[h_{2}]l[i]}*year_i  + \delta_{l[i]}+ \chi_{c[i]} + \omega_{s[i]} \\
-		logit(\hat{P}_{[h_{3}]i}) = \mathrm{A}_{[h_{2}]i} + \mathrm{T}_{[h_{2}]}*year_i  + \alpha_{[h_{2}]l[i]} + \tau_{[h_{3}]l[i]}*year_i  + \delta_{l[i]}+ \chi_{c[i]} + \omega_{s[i]}\\
+		logit(\hat{P}_{[h_{3}]i}) = \mathrm{A}_{[h_{2}]i} + \mathrm{T}_{[h_{2}]}*year_i  + \alpha_{[h_{2}]l[i]} + \tau_{[h_{3}]l[i]}*year_i  + \delta_{l[i]}+ \chi_{c[i]} + \omega_{s[i]}
 	\end{align}
@@
 
-We performed model fitting using the inlabru R package \citep{}.
+We performed model fitting using the inlabru R package \citep{}\jacob{} {add citation}.
 Global intercept and slope parameters $\mathrm{A}$, and $\mathrm{T}$, were given vague priors.
@@
 %Because initial analyses indicated a high degree of collinearity between seasonal and annual changes in temperature, we used annual temperature only, along with annual and seasonal precipitation, in the subsequent analysis.
-All together, we assessed twelve potential climate drivers: the mean and standardiation of spring, summer, and autumn temperature, as well as the mean and standard deviation of spring, summer, and autumn cumulative precipitation, cumulative precipitation, and cumulative precipitation.
+All together, we assessed twelve potential climate drivers: the mean and standardization of spring, summer, and autumn temperature, as well as the mean and standard deviation of spring, summer, and autumn cumulative precipitation, cumulative precipitation, and cumulative precipitation.
 
@@
 }{I think we need to account for uncertainty in the slopes.
-They are outputs of a (quasi) Bayesian model so we should be able to propagate all the uncertainty in the posterior disgtribution.}
-Data from each host species was analysed separately. 
-Fitting regressions to many pixels across the study region risks articially inflating confidence in our results due to large sample sizes, and \tom{so we performed this analysis using only a random subsample of 250 pixels across the study region, which provided results qualitatively similar to analysis of the full set of pixels}{100 seems like a low number to me.
+They are outputs of a (quasi) Bayesian model so we should be able to propagate all the uncertainty in the posterior distribution.}
+Data from each host species was analyzed separately. 
+Fitting regressions to many pixels across the study region risks artificially inflating confidence in our results due to large sample sizes, and \tom{so we performed this analysis using only a random subsample of 250 pixels across the study region, which provided results qualitatively similar to analysis of the full set of pixels}{100 seems like a low number to me.
 What if we did this for all of the herbarium collection locations?}\josh{.}{I upped the number of points.
@@
  A full description of the ODMAP can be found in the  (Supplementary Method \ref{sec:sdm }). 
- In short, we used presence-only observations of the host species from Global Biodiversity Information Facility (GBIF) between 1990 to 2020 .
+ Here we presented the workflow of the species distribution model. 
+ We used presence-only observations of the host species from Global Biodiversity Information Facility (GBIF) between 1990 to 2020.
 To reduce the potential influence of sampling bias and spatial autocorrelation, we thinned the occurrences to the spatial scale of our selected climatic predictors. 
@@
 These climatic variables are: the mean and standard deviation of spring, summer, and autumn temperature, as well as the mean and standard deviation of spring, summer, and autumn cumulative precipitation, cumulative precipitation, and cumulative precipitation.
-%We calculated the mean and standard devation of seasonal temperature and precipitation across 1990 to 2020. 
-Among this suite of variables, we chose to include mean summer temperature, standard deviation of spring precipitation, standard deviation of summer precipitation,  standard deviation of spring  precipitation, standard deviation of summer  precipitation, which were uncorrelated (Variance Inflation Factor $>$ 0.7) and allowed us to predict the occurrence probability of each host species in space and time.
+%We calculated the mean and standard deviation of seasonal temperature and precipitation across 1990 to 2020. 
+Among this suite of variables, we chose to include mean summer temperature, standard deviation of spring temperature, standard deviation of summer temperature,  standard deviation of spring  precipitation, standard deviation of summer  precipitation, which were uncorrelated (Variance Inflation Factor $>$ 0.7) and allowed us to predict the occurrence probability of each host species in space and time.
 We fit maximum entropy (MaxEnt) models using the maxent function in the package dismo \citep{hijmans2017package}. 
@@
 	\section*{Supporting Methods}
-	\subsection*{ODMAP Protocol}
-\label{sec:sdm}
+	\subsection*{ODMAP Protocol} \label{sec:sdm}
 {\color{NavyBlue}{Overview}}\\
@@
 \textbf {Target species}: \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}. \\
-\textbf {Study area}: Eastern North America \\
+\textbf {Study area}: United States \\
 %Scale of analysis
@@
 {\color{NavyBlue}{Data}}\\
-\textbf {Observation type}: Occurrence records from  Global Biodiversity Information Facility and herbarium collection across eastern North America.
+\textbf {Observation type}: Occurrence records from  Global Biodiversity Information Facility and herbarium collection across the United States.
 We used 713 occurrences records for \emph{Agrostis hyemalis}, 656 occurrence records for \emph{Agrostis perennans} and 2338 for \emph{Elymus virginicus}.\\
@@
 \textbf{Software}:  All statistics were performed using Maxent 3.3.4 and R4.3.1 with packages terra, usdm, spThin and dismo.\\
-\textbf{Code availability}: https://github.com/joshuacfowler/EndoHerbarium\\
-\textbf{Data availability}: https://github.com/joshuacfowler/EndoHerbarium\\
+\textbf{Code availability}: Will  available upon  acceptance\\
+\textbf{Data availability}: Will  available upon  acceptance\\
 {\color{NavyBlue}{Assessment}}\\

### 2024-09-04 sdm.R
trying to figure out merging

@@ -216,9 +216,9 @@ test_aghy_a <- predict(mod_aghy, env_aghy_bg)
 e_aghy <- evaluate(p=test_aghy_p, a=test_aghy_a)
 threshold(e_aghy)
 
-par(mfrow=c(1, 2))
+par(mfrow=c(1, 3))
 # density(e_aghy)
-# boxplot(e_aghy, col=c('blue', 'red'))
+boxplot(e_aghy, col=c('blue', 'red'))
 plot(e_aghy, 'ROC')
 # plot(e_aghy, 'TPR')
 plot(e_aghy, 'kappa')
@@ -488,7 +488,7 @@ dev.off()
 
 # pdf("/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Figures/FigSDMMaxent.pdf",height = 5,width = 10)
 # # layout(mat = layout.matrix)
-# op<-par(mfrow = c(1,3), mar=c(3,3,2,1),oma = c(5, 5, 1, 3))
+# op<-par(mfrow = c(2,3), mar=c(3,3,2,1),oma = c(5, 5, 1, 3))
 # plot(map_old_aghy,xaxt='n',main=substitute(paste(italic("A. hyemalis"))))
 # text(x=-80,y=50,label="1925")
 # # par(mar=c(2.2,1,3,1))
@@ -521,10 +521,7 @@ plot(aghy)
 # layout.matrix <- matrix(c(1, 2, 3, 4,5,6), nrow = 2, ncol = 3)
 pdf("/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Figures/FigSDM_Presence_absence.pdf",height = 8,width = 5)
 # layout(mat = layout.matrix)
-op<-par(mfrow = c(1,3), mar=c(4,8,1.5,2))
-plot(e_aghy, 'ROC')
-plot(e_agpe, 'ROC')
-plot(e_elvi, 'ROC')
+op<-par(mfrow = c(3,1), mar=c(4,8,1.5,2))
 # plot(map_old_aghy > tr_aghy,xaxt='n',main=substitute(paste(italic("A. hyemalis"))))
 # text(x=-80,y=50,label="1925")
 # # par(mar=c(2.2,1,3,1))
@@ -562,7 +559,6 @@ op<-par(mfrow = c(3,1), mar=c(4,8,1.5,2))
 # par(mar=c(5,3,0,1))
[…]

### 2024-09-04 sdm.R
saving jacobs changes to sdm file

@@ -568,14 +568,13 @@ mtext("C",side = 3, adj = 0,cex=1.25)
 par(op)
 dev.off()
 
-aghy_binary <- aghy >  0.5102097
-agpe_binary <-agpe > 0.52849
-elvi_binary <-elvi > 0.5422173
+aghy_binary <- aghy >  0.5410866
+agpe_binary <-agpe > 0.452535
+elvi_binary <-elvi > 0.5163273
 
-writeRaster(aghy_binary, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/github/EndoHerbarium/aghy_binary.tif')
+riteRaster(aghy_binary, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/github/EndoHerbarium/aghy_binary.tif')
 writeRaster(agpe_binary, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/github/EndoHerbarium/agpe_binary.tif')
 writeRaster(elvi_binary, '/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/github/EndoHerbarium/elvi_binary.tif')
-
 
 graph<-list("A","B","C")
 pdf("/Users/jm200/Library/CloudStorage/Dropbox/Miller Lab/Herbaruim Project/Figures/FigSDMMaxent_ROC.pdf",height = 5,width = 5)

### 2024-09-04 Manuscript/endo_herbarium.tex
Updated my edits

@@
 \usepackage[T1]{fontenc}
-
-\usepackage[dvipsnames]{xcolor}
+\usepackage{xcolor}
 
@@
 
-\usepackage{xcolor}
 \newcommand{\tom}[2]{{\color{red}{#1}}\footnote{\textit{\color{red}{#2}}}}
@@
 Documented responses to environmental change, such as shifts in species' distributions \cite{aitken2008adaptation} and phenology \cite{piao2019plant}, are typically blind to concurrent changes in associated biotic interactions.
-Empirically evaluating these biotic changes -- whether interacting species shift in tandem with their partners or not \cite{hillerislambers2013will} -- is crucial to predicting the reorganization of Earth's biodiversity under global change.
+Empirically evaluating these biotic changes -- whether interacting species shift in tandem with their partners or not \cite{hillerislambers2013will} -- is crucial to predicting the reogranization of Earth's biodiversity under global change.
 Such evaluations have been limited because few datasets on species interactions extend over sufficiently long time scales of contemporary climate change \cite{poisot2021global}.
@@
 Both \emph{Agrostis} species are recorded from throughout the Eastern US, but \emph{A.
-perennans} has a slightly more northern distribution, whereas \emph{A.
+perennans} has a slighty more northern distribution, whereas \emph{A.
 hyemalis} is found rarely as far north as Canada and is listed as a rare plant in Minnesota.
@@
 Shared variance terms included the spatially-dependent random effect $\delta_{l[i]}$, intended to account for residual spatial variation, and $\chi_{c[i]}$ and $\omega_{s[i]}$ i.i.d.-random effects indexed for each collector identity ($c[i]$), and scorer identity ($s[i]$) of the $i^{th}$ specimen.
+
 \begin{subequations}
@@
 
-We performed model fitting using the inlabru R package \citep{}\jacob{} {add citation}.
+We performed model fitting using the inlabru R package \citep{}.
 Global intercept and slope parameters $\mathrm{A}$, and $\mathrm{T}$, were given vague priors.
@@
 %Because initial analyses indicated a high degree of collinearity between seasonal and annual changes in temperature, we used annual temperature only, along with annual and seasonal precipitation, in the subsequent analysis.
-All together, we assessed twelve potential climate drivers: the mean and standardization of spring, summer, and autumn temperature, as well as the mean and standard deviation of spring, summer, and autumn cumulative precipitation, cumulative precipitation, and cumulative precipitation.
+All together, we assessed twelve potential climate drivers: the mean and standardiation of spring, summer, and autumn temperature, as well as the mean and standard deviation of spring, summer, and autumn cumulative precipitation, cumulative precipitation, and cumulative precipitation.
 
@@
 }{I think we need to account for uncertainty in the slopes.
-They are outputs of a (quasi) Bayesian model so we should be able to propagate all the uncertainty in the posterior distribution.}
+They are outputs of a (quasi) Bayesian model so we should be able to propagate all the uncertainty in the posterior disgtribution.}
 Data from each host species was analyzed separately. 
-Fitting regressions to many pixels across the study region risks artificially inflating confidence in our results due to large sample sizes, and \tom{so we performed this analysis using only a random subsample of 250 pixels across the study region, which provided results qualitatively similar to analysis of the full set of pixels}{100 seems like a low number to me.
+Fitting regressions to many pixels across the study region risks articially inflating confidence in our results due to large sample sizes, and \tom{so we performed this analysis using only a random subsample of 250 pixels across the study region, which provided results qualitatively similar to analysis of the full set of pixels}{100 seems like a low number to me.
 What if we did this for all of the herbarium collection locations?}\josh{.}{I upped the number of points.
@@
 A full description of the ODMAP can be found in the  (Supplementary Method \ref{sec:sdm }). 
- Here we presented the workflow of the species distribution model. 
- We used presence-only observations of the host species from Global Biodiversity Information Facility (GBIF) between 1990 to 2020.
+ In short, we used presence-only observations of the host species from Global Biodiversity Information Facility (GBIF) between 1990 to 2020.
 To reduce the potential influence of sampling bias and spatial autocorrelation, we thinned the occurrences to the spatial scale  of our selected climatic predictors. 
 We selected climate variables that aligned with our analysis of climatic influences on trends in endophyte prevalence described above.
+We calculated the mean and standard deviation of seasonal temperature and precipitation across 1990 to 2020. 
+Among this suite of variables, we chose to include, which were uncorrelated (Variance Inflation Factor $>$ 0.7) and allowed us to predict the occurrence probability of each host species in space and time.
 These climatic variables are: the mean and standard deviation of spring, summer, and autumn temperature, as well as the mean and standard deviation of spring, summer, and autumn cumulative precipitation, cumulative precipitation, and cumulative precipitation.
-%We calculated the mean and standard deviation of seasonal temperature and precipitation across 1990 to 2020. 
-Among this suite of variables, we chose to include mean summer temperature, standard deviation of spring temperature, standard deviation of summer temperature,  standard deviation of spring  precipitation, standard deviation of summer  precipitation, which were uncorrelated (Variance Inflation Factor $>$ 0.7) and allowed us to predict the occurrence probability of each host species in space and time.
 We fit maximum entropy (MaxEnt) models using the maxent function in the package dismo \citep{hijmans2017package}. 
@@
 We found AUC = 0.862, AUC = 0.838, AUC = 0.821 respectively  for \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}.
-
 To convert the continuous predicted probabilities into binary presence - absence maps, we used the  training sensitivity (true positive rate) and specificity threshold (true negative rate) \citep{liu2005selecting}.
-These binary maps serve as boundaries in presented maps of change in endophyte prevalence, and outline the set of pixels used in our analysis of climate correlates with trends in endophyte prevalence.
+These binary maps serve as boundaries in presented maps of change in endophyte prevalence, and outline the set of pixels used in our analysis of climate correlates with trends in endophyte prevalence
 
@@
 This interpretation is in line with theory predicting that the positive fitness feedback caused by vertical transmission leads beneficial symbionts to rise in prevalence within a population \cite{fine1975vectors}.
-We found that trends in endophyte prevalence varied across the distribution of each species in assocation with observed changes in climate drivers, suggesting that the endophytes have contributed to host resilience under environmental change.
+We found that trends in endophyte prevalence varied across the distribution of each species in association with observed changes in climate drivers, suggesting that the endophytes have contributed to host resilience under environmental change.
 Taken together, this suggests a strengthening of the mutualism over the last two centuries.
@@
 	\end{table}	
-
 	\section*{Supporting Methods}
-	\subsection*{ODMAP Protocol} \label{sec:sdm}
-{\color{NavyBlue}{Overview}}\\
+	\subsection*{ODMAP Protocol}\label{sec:sdm}
+{\color{blue}{Overview}}\\
 \textbf {Model  purpose}: Mapping current distribution of epichloë host species. \\
 \textbf {Target species}: \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}. \\
-\textbf {Study area}: United States \\
+\textbf {Study area}: Eastern North America \\
 %Scale of analysis
@@
 \textbf {Boundary}: Natural.\\
-{\color{NavyBlue}{Data}}\\
-\textbf {Observation type}: Occurrence records from  Global Biodiversity Information Facility and herbarium collection across the United States.
+{\color{blue}{Data}}\\
+\textbf {Observation type}: Occurrence records from  Global Biodiversity Information Facility and herbarium collection across eastern North America.
 We used 713 occurrences records for \emph{Agrostis hyemalis}, 656 occurrence records for \emph{Agrostis perennans} and 2338 for \emph{Elymus virginicus}.\\
@@
 \textbf{Climatic data}:  raster data extracted from PRISM  \\
-{\color{NavyBlue}{Model }}\\
+{\color{blue}{Model }}\\
 \textbf{Model assumption}: We assumed that the target species are at equilibrium with their environment. \\
@@
 \textbf{Software}:  All statistics were performed using Maxent 3.3.4 and R4.3.1 with packages terra, usdm, spThin and dismo.\\
-\textbf{Code availability}: Will  available upon  acceptance\\
-\textbf{Data availability}: Will  available upon  acceptance\\
-{\color{NavyBlue}{Assessment}}\\
+\textbf{Code availability}: Will be available upon acceptance \\
+\textbf{Data availability}: Will be available upon acceptance \\
+{\color{blue}{Assessment}}\\
 We used AUC to test model performance.\\
-{\color{NavyBlue}{Prediction }}\\
+{\color{blue}{Prediction }}\\
 We predicted the probability of presence of the host species as a binary maps (presence or absence)

### 2024-09-05 Manuscript/endo_herbarium.tex
adding in some citations and removing some comments

@@
 Specifically, we analyzed 2,346 herbarium specimens of three grass host species collected over the last two centuries (1824 -- 2019) for the presence or absence of endophyte symbiosis, and evaluated spatial and temporal trends in endophyte prevalence.
-We found that endophytes have increased in prevalence over the last two centuries from ca. 25\% prevalence to ca. 75\% prevalence, on average, across the three host species.
-We also found that changes in prevalence were associated with observed \tom{changes in seasonal climate drivers}{Describe ``changes'' -- warming?
-drying?} corresponding to each host species' peak growing season.
-Our analysis performed favorably in an out-of-sample predictive test, however we identified XXX as suggesting the model fusion may be an important step moving forward.
-\jacob{Our results provide novel evidence for a cryptic biological response to climate change that may contribute to the resilience of host-microbe symbiosis through context-dependent benefits that confer a fitness advantage to symbiotic hosts under environmental change.}{I like this and the abstract in general.
-I agree with Tom and I think we  have some space to add these details.
-Abstract : 300}
+We found that endophytes increased in prevalence over the last two centuries from ca. 25\% prevalence to ca. 75\% prevalence, on average, across three host species.
+We also found that changes in prevalence were associated with observed changes in seasonal climate drivers; noteably increasing precipitation corresponding to each host species' peak growing season and changes in off-peak season variability in precipitation.
+Our analysis performed favorably in an out-of-sample predictive test with contemporary data, however we identified greater local-scale variability in endophyte prevalence in contemporary data compared to historic data, suggesting that model fusion may be an important step moving forward.
+Our results provide novel evidence for a cryptic biological response to climate change that may contribute to the resilience of host-microbe symbiosis through context-dependent benefits that confer a fitness advantage to symbiotic hosts under environmental change.
+
+Abstract : 287 words
 
@@
 
-Early research on \emph{Epichloë} used herbarium specimens to describe the broad taxonomic diversity of host species that harbor these symbionts \citep{white1985endophyte}, \josh{establishing that endophyte symbiosis could be identified in plant tissue from as early as 1851.}{Edited this a bit.
-This is the earliest year we have in the database that was part of JFWhites original paper.}
+Early research on \emph{Epichloë} used herbarium specimens to describe the broad taxonomic diversity of host species that harbor these symbionts \citep{white1985endophyte}, establishing that endophyte symbiosis could be identified in plant tissue from as early as 1851.
 However, no subsequent studies, to our knowledge, have used the vast resources of biological collections to quantitatively assess spatio-temporal trends in endophyte prevalence and their environmental correlates.
@@
 This creates the opportunity to leverage the unique spatio-temporal sampling of herbarium collections to examine the response of the symbiosis to historical climate change. 
-\josh{Research using historical collections has clearly demonstrated other ecological signatures of a changing climate. 
+Research using historical collections has clearly demonstrated other ecological signatures of a changing climate. 
 However the predictive ability of these historical analyses is rarely tested against contemporary data \citep{lee2024phenological}. 
-Identifying the ways in which these analyses fall short is a crucial step for the field move from reading signatures in the past to forecasting ecological dynamics into the future.}{what do you think of this?
-trying to presage the out-of-sample test without over promising and without saying outright that our analysis sucks.
-Is this the right place for this?
-I had imagined some of this material will be really developed in the discussion.}
+Identifying the ways in which these analyses fall short is a crucial step for the field move from reading signatures in the past to forecasting ecological dynamics into the future.
 
@@
 To answer these questions we examined a total of 2,346 specimens collected across eastern North America between 1824 and 2019.
-\tom{}{I think the consensus was to keep the out-of-sample validation which should absolutely go into the Intro as an important element of novelty. Should go in the Asbtract too.}
 
@@
 perennans}, and $3.8$ seeds per specimen of \emph{E.
-virginicus}; we scored \# seeds in total.
+virginicus}; we scored $10,342$ seeds in total.
 .
@@
 We assessed spatial and temporal changes in endophyte prevalence across each host distribution, quantifying the ``global'' temporal trends, aggregating across space, and then examining spatial heterogeneity in the direction and magnitude of endophyte change (hotspots and coldspots) across the spatial extent of each host's distribution.
-To appropriately account for the spatial non-independence of geo-referenced \josh{occurrences}{spelling?}, we used an approximate Bayesian method, Integrated Nested Laplace Approximation (INLA), to construct spatio-temporal models of endophyte prevalence.
+To account for the spatial non-independence of geo-referenced occurrences, we used an approximate Bayesian method, Integrated Nested Laplace Approximation (INLA), to construct spatio-temporal models of endophyte prevalence.
 INLA provides a computationally efficient method of ascertaining parameter posterior distributions for certain models that can be formulated as latent Gaussian Models \cite{rue2009approximate}.
 Many common statistical models, including structured and unstructured mixed-effects models, can be represented as latent Gaussian Models.
-We incorporated spatial heterogeneity into this analysis using spatially-structured intercept and slope parameters implemented as stochastic partial differential equation (SPDE) approximations of a continuous spatial Gaussian process.
+We incorporated spatial heterogeneity into this analysis using spatially-structured intercept and slope parameters implemented as stochastic partial differential equations (SPDE) to approximate a continuous spatial Gaussian process.
 This SPDE approach is a flexible method of smoothing across space while explicitly accounting for spatial dependence between data-points \citep{lindgren2011explicit,bakka2018spatial}.
@@
 
-We performed model fitting using the inlabru R package \citep{}.
+We performed model fitting using the inlabru R package \citep{bachl2019inlabru}.
 Global intercept and slope parameters $\mathrm{A}$, and $\mathrm{T}$, were given vague priors.
@@
 Priors, termed ''range" and "variance", define the distance of spatial decay described by the Mat\'{e}rn covariance function.
-Priors for results presented in the main text reflect a range of \josh{XX}{} kilometers. 
-We found that model results were sensitive to this choice, and so tested a range of priors (from XX kilometers to XX kilometers) and meshes (Supplemental Material), finding that model results were qualitatively similar, i.e.
+Priors for results presented in the main text reflect a range of $342$ kilometers. 
+We found that model results were sensitive to this choice, and so tested a range of priors (from 68 kilometers to 2160 kilometers) and meshes (Supplemental Material), finding that model results were qualitatively similar, i.e.
 the same direction of effects across space, but that the magnitude and uncertainty varied.
@@
 \subsection*{Validating model performance with in-sample and out-of-sample tests}
-We evaluated the predictive ability of the model using both in-sample training data from the herbarium surveys, and with out-of-sample test data from contemporary endophyte surveys, \tom{an important but rarely used strategy in ecological studies \cite{tredennick2021practical}.}{This is the type of thing to emphasize in the intro?
-Are there any other collections-based papers that have done anything like this??
-None to my knowledge.} \josh{}{Add Benjamin lee paper, maybe?
-it's not just herbaria, but kind of related}
+We evaluated the predictive ability of the model using both in-sample training data from the herbarium surveys, and with out-of-sample test data from contemporary endophyte surveys, an important but rarely used strategy in ecological studies \cite{tredennick2021practical, lee2024phenological}.
 We used data from contemporary surveys of endophyte prevalence  in \emph{A. hyemalis} and \emph{E. virginicus} in Texas and the southern US. 
@@
 We then took the difference between recent and historic periods for the mean and standard deviation for each climate driver (Figs. \ref{fig:AGHY_climate_covariates}-\ref{fig:ELVI_climate_covariates}).
-%Because initial analyses indicated a high degree of collinearity between seasonal and annual changes in temperature, we used annual temperature only, along with annual and seasonal precipitation, in the subsequent analysis.
 All together, we assessed twelve potential climate drivers: the mean and standardiation of spring, summer, and autumn temperature, as well as the mean and standard deviation of spring, summer, and autumn cumulative precipitation, cumulative precipitation, and cumulative precipitation.
 
-To evaluate whether areas that have experienced the greatest changes in endophyte prevalence (hotspots of endophyte change) are associated with high degrees of change in climate (hotspots of climate change), \tom{we modeled spatially varying slopes of endophyte change through time ($\tau_{[h]l}$ as a linear function of environmental covariates, with a Gaussian error distribution.
-}{I think we need to account for uncertainty in the slopes.
-They are outputs of a (quasi) Bayesian model so we should be able to propagate all the uncertainty in the posterior disgtribution.}
+To evaluate whether areas that have experienced the greatest changes in endophyte prevalence (hotspots of endophyte change) are associated with high degrees of change in climate (hotspots of climate change), we modeled spatially varying slopes of endophyte change through time ($\tau_{[h]l}$ as a linear function of environmental covariates, with a Gaussian error distribution.
 Data from each host species was analyzed separately. 
-Fitting regressions to many pixels across the study region risks articially inflating confidence in our results due to large sample sizes, and \tom{so we performed this analysis using only a random subsample of 250 pixels across the study region, which provided results qualitatively similar to analysis of the full set of pixels}{100 seems like a low number to me.
-What if we did this for all of the herbarium collection locations?}\josh{.}{I upped the number of points.
-I don't think conceptually we need to only do collection locations, but I can work on that as an alternative.}
-
-\tom{}{I cut the notation for the Gaussian model for now because it is a pretty simple model and the notation may be overkill, plus because I changed your tau's to beta's there were betas on both sides of the tilde, which was confusing/annoying.
-Happy have the notation back if you prefer it.
-I am also a little confused because the appendix has spearman correlations but there are no methods here for where those come from.}
+Fitting regressions to many pixels across the study region risks articially inflating confidence in our results due to large sample sizes, and so we performed this analysis using only a random subsample of 250 pixels across the study region, which provided results qualitatively similar to analysis of the full set of pixels.
 
@@
 	\section*{Data and Code Availability}
-
-	On initial submission, you may use this section to provide a URL for editors and reviewers that is `private for peer review'.
-After acceptance, this section must be updated with correct, working DOIs for data deposits (typically on the Dryad Digital Repository, ) and code deposits (such as in Zenodo).
+	Data from this publication will be made publically available at the time of publication and upon request.
+	Code for analyses can be found through: https://github.com/joshuacfowler/EndoHerbarium
 

### 2024-09-05 Analyses/endo_herbarium_records_merge.R
updating my counts of specimens for the table

@@ -1526,6 +1526,12 @@ need_latlon_summary <- need_latlon %>%
   group_by(collection) %>% 
   summarize(count = n())
 
+# summary of how many scored specimens from each collection
+collections_summary <- endo_herb_georef %>% 
+  mutate(collection = word(Sample_id, 1, sep = "_")) %>% 
+  group_by(collection, Spp_code) %>% 
+  summarize(count = n())
+
 
 
 endo_herb_georef %>% filter(!is.na(Endo_status_liberal) & is.na(year)) %>% view()

### 2024-09-05 Analyses/endo_herbarium_multispecies_spatiotemporal_analysis.R
making a density overlap sim vs obs ppc plot

@@ -452,6 +452,63 @@ rocobj$auc
 
 
 # Taking alot of material from this blog post: https://inlabru-org.github.io/inlabru/articles/2d_lgcp_covars.html
+post.pred1 <- generate(
+  fit,
+  newdata = data1,
+  formula = ~ invlogit(int.species1 + year.species1 + 
+                         space +  space.species1 + 
+                         time.species1 + 
+                         scorer+ collector),
+  n.samples = 100) 
+posterior_samples1 <- bind_cols(data1, post.pred1)
+
+
+post.pred2 <- generate(
+  fit,
+  newdata = data2,
+  formula = ~ invlogit(int.species2 + year.species2 + 
+                         space + space.species3 + 
+                         time.species2 + 
+                         scorer+ collector),
+  n.samples = 100) 
+posterior_samples2 <- bind_cols(data2, post.pred2)
+
+post.pred3 <- generate(
+  fit,
+  newdata = data3,
+  formula = ~ invlogit(int.species3 + year.species3 +
+                         space + space.species3 + 
+                         time.species3 +
+                         scorer+ collector),
+  n.samples = 100) 
+
+posterior_samples3 <- bind_cols(data3, post.pred3)
+
+posterior_samples <- bind_rows(posterior_samples1, posterior_samples2, posterior_samples3) %>% 
+  arrange(row_number) %>% 
+  select(...55:...154) %>% st_drop_geometry %>% as.matrix()
+
+# simulating datasets from the posterior samples
+n_post_draws <- 100
+y_sim <- matrix(NA,n_post_draws,length(data$Endo_status_liberal))
+
+for(i in 1:n_post_draws){
+y_sim[i,] <- rbinom(n = length(endo_herb$Endo_status_liberal), size = 1, prob = posterior_samples[,i])
+}
+
+saveRDS(y_sim, file = "y_sim.rds")
+y_sim_df <- t(y_sim)
+colnames(y_sim_df) <- paste("iter", 1:n_post_draws)
+y_sim_df <- as_tibble(y_sim_df, .name_repair = "minimal") %>% 
+  pivot_longer(cols = everything())
+
+overlay_plot <- ggplot(data)+
+  geom_line(data = y_sim_df, aes(x = value, group = name), stat="density", color = "red", alpha = .3) +
+  geom_line(aes(x = Endo_status_liberal), stat="density") + 
+  labs(x = "Endophyte Status", y = "Density")+
+  theme_minimal()
+overlay_plot
[…]

### 2024-09-05 Manuscript/endo_herbarium.tex
removing comments for submitting to BioRxiv

@@
 
-	\textit{Manuscript elements}: Figure~1, figure~2, table~1, appendix~A (for print; including figure~A1, figure~A2, and table~A1), supplemental PDF. Figure~2 is to print in color.
+	\textit{Manuscript elements}: Figure \ref{fig:map} - Figure \ref{fig:contemptestplot}, appendix~A (including Figure~\ref{fig:endo_status_map} - Figure \ref{fig:ELVI_climate_covariates},  Table \ref{table:herbaria}, and Supplemental Methods).
 
@@
 		\subsection*{Herbarium surveys}
-We visited nine herbaria between 2019 and 2022 (see Table A1 for a summary of specimens included from each collection). 
+We visited nine herbaria between 2019 and 2022 (see Table \ref{table:herbaria} for a summary of specimens included from each collection). 
 With permission from herbarium staff, we acquired seed samples from $1135$ \emph{A.
@@
 $89$\% of scored plants had matching liberal and conservative scores, reflecting high confidence in endophyte status.
-The following analyses in the main text used the liberal status, but we repeated all analyses with the conservative status which yielded qualitatively similar results (Fig. \ref{fig:FigA5})
+The following analyses used the liberal status, but we repeated all analyses with the conservative status which yielded qualitatively similar results (Fig. \ref{fig:FigA5})
 
@@
 \subsection*{Modeling spatial and temporal changes in endophyte prevalence}
-We assessed spatial and temporal changes in endophyte prevalence across each host distribution, quantifying the ``global'' temporal trends, aggregating across space, and then examining spatial heterogeneity in the direction and magnitude of endophyte change (hotspots and coldspots) across the spatial extent of each host's distribution.
+We assessed spatial and temporal changes in endophyte prevalence across each host distribution, quantifying the ``global'' temporal trends aggregated across space, and then examining spatial heterogeneity in the direction and magnitude of endophyte change (hotspots and coldspots) across the spatial extent of each host's distribution.
 To account for the spatial non-independence of geo-referenced occurrences, we used an approximate Bayesian method, Integrated Nested Laplace Approximation (INLA), to construct spatio-temporal models of endophyte prevalence.
@@
 \ref{fig:meshplot}).
+We assessed model fit with visual posterior predictive checks (\ref{fig:overlayplot}) and measurements of AUC (Figs. \ref{fig:ROCtraining}-\ref{fig:ROCtest}).
 Priors, termed ''range" and "variance", define the distance of spatial decay described by the Mat\'{e}rn covariance function.
@@
 We used data from contemporary surveys of endophyte prevalence  in \emph{A. hyemalis} and \emph{E. virginicus} in Texas and the southern US. 
-Surveys of \emph{E.
-virginicus} were conducted in 2013 as described in \citet{sneck2017variation}, and \tom{surveys of \emph{A.
+Surveys of \emph{E. virginicus} were conducted in 2013 as described in \citet{sneck2017variation}, and surveys of \emph{A. hyemalis} took place between 2015 and 2020.
+%\tom{surveys of \emph{A.
 hyemalis} took place between 2015 and 2020}{We have added more recent AGHY survey data.
@@
 We followed the ODMAP (overview, data, model, assessment, prediction) protocol \citep{crossley2022opposing}.
-A full description of the ODMAP can be found in the  (Supplementary Method \ref{sec:sdm }). 
+A full description of the ODMAP can be found in the  (Supporting Methods). 
  In short, we used presence-only observations of the host species from Global Biodiversity Information Facility (GBIF) between 1990 to 2020.
@@
 	\section*{Acknowledgments}
-	We thank Jessica Budke for help in drafting our initial destructive sampling plan, and to the many members of herbarium staff who facilitated our research visits, as well as to the hundreds of collectors who contributed to the natural history collections.
+	We thank Dr.
+Jessica Budke for help in drafting our initial destructive sampling plan, and to the many staff members of herbaria who facilitated our research visits, as well as to the hundreds of collectors who contributed to the natural history collections.
 	Several high schooler and undergraduate researchers contributed to data collection, including A.
@@
 J.C.F. contributed to research conception, data collection, data analysis, and led manuscript drafting.
-J.M. contributed to data analysis and manuscript manuscript revisions.
+J.M. contributed to data analysis and manuscript revisions.
 T.E.X.M. contributed to research conception, data collection, data analysis, and manuscript revisions.
@@
 	\section*{Data and Code Availability}
-	Data from this publication will be made publically available at the time of publication and upon request.
+	Data from this publication will be made publically available upon acceptance and before that upon request.
 	Code for analyses can be found through: https://github.com/joshuacfowler/EndoHerbarium
@@
 	\centering
-	\includegraphics[width = .8\linewidth]{density_plot.png}
+	\includegraphics[width = .5\linewidth]{../Plots/overlay_plot.png}
 	\caption{\textbf{Consistency between real data and simulated values indicate that the fitted model accurately describes the data}.
-Graph shows density curves for the observed data (red) along with the mean(solid) and 95\% CI (dashed) of simulated values (black).}
+Graph shows density curves for the observed data (black) along with with 100 simulated datasets (red).}
+			\label{fig:overlayplot}
 \end{figure}
@@
 		\caption{Summary of herbarium samples across collections}
-		\label{Table:herbaria}
+		\label{table:herbaria}
 		\centering
@@
 			Herbarium Collection        & AGHY        & AGPE      &      ELVI\\ \hline
-			Botanical Research Institute of Texas &   341   &    189&    176    \\
-			Louisiana State University &     71  & --  &   61       \\
+			Botanical Research Institute of Texas &   350   &    190&    198    \\
+			Louisiana State University &     72  & 38  &   62       \\
 			Mercer Botanic Garden &   3    & --     &     6\\
-			Missouri Botanic Garden& 78 & 39& 31\\
-		    Texas A\&M &  73&-- & 49 \\
-		    University of Kansas & 134 & -- &  20\\
-		    University of Oklahoma & 65 &30&  91\\
-		    University of Texas  \& Lundell   &  169& 41& 99\\		    				 			     			     
-			Oklahoma State University&     30  &   --    &  69 \\ \hline
+			Missouri Botanic Garden& 210 & 205 & 122\\
+		    Texas A\&M &  100 &-- & 72 \\
+		    University of Kansas & 134 & 34 &  197\\
+		    University of Oklahoma & 85 &34&  95\\
+		    University of Texas  \& Lundell   &  183& 91& 102\\		    				 			     			     
+			Oklahoma State University&     51  &   10    &  74 \\ \hline
 		\end{tabular}
@@
 {\color{blue}{Overview}}\\
-\textbf {Model  purpose}: Mapping current distribution of epichloë host species. \\
+\textbf {Model  purpose}: Mapping current distribution of \emph{Epichloë} host species. \\
 \textbf {Target species}: \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}. \\
@@
 \textbf{Software}:  All statistics were performed using Maxent 3.3.4 and R4.3.1 with packages terra, usdm, spThin and dismo.\\
-\textbf{Code availability}: Will be available upon acceptance \\
+\textbf{Code availability}: Available through this link: https://github.com/joshuacfowler/EndoHerbarium \\
 \textbf{Data availability}: Will be available upon acceptance \\

### 2024-09-05 Manuscript/endo_herbarium.tex
fixing minor typos

@@
 We found that endophytes increased in prevalence over the last two centuries from ca. 25\% prevalence to ca. 75\% prevalence, on average, across three host species.
-We also found that changes in prevalence were associated with observed changes in seasonal climate drivers; noteably increasing precipitation corresponding to each host species' peak growing season and changes in off-peak season variability in precipitation.
+We also found that changes in prevalence were associated with observed changes in seasonal climate drivers; notably increasing precipitation corresponding to each host species' peak growing season and changes in off-peak season variability in precipitation.
 Our analysis performed favorably in an out-of-sample predictive test with contemporary data, however we identified greater local-scale variability in endophyte prevalence in contemporary data compared to historic data, suggesting that model fusion may be an important step moving forward.

### 2024-09-08 Analyses/endo_herbarium_multispecies_spatiotemporal_analysis_with_conservative_scores.R
okay, got a working script to run the model with conservative scores and make a plot for supplement

@@ -1,0 +1,1157 @@
+# Purpose: Fits spatiotemporal model of change in endophyte prevalence in herbarium specimens
+# Authors: Joshua Fowler and Tom Miller
+# Updated: Feb 22, 2024
+
+library(tidyverse) # for data manipulation and ggplot
+library(INLA) # for fitting integrated nested Laplace approximation models
+library(inlabru)
+library(fmesher)
+
+library(sf)
+# library(rmapshaper)
+library(terra)
+library(tidyterra)
+# library(ggpolypath)
+
+
+
+library(patchwork)
+library(ggmap)
+library(pROC)
+library(tidybayes)
+
+invlogit<-function(x){exp(x)/(1+exp(x))}
+species_colors <- c("#1b9e77","#d95f02","#7570b3")
+endophyte_colors <- c("#fdedd3","#f3c8a8", "#5a727b", "#4986c7", "#181914",  "#163381")
+
+
+################################################################################
+############ Read in the herbarium dataset ############################### 
+################################################################################
+
+endo_herb_georef <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/DigitizedHerbariumRecords/endo_herb_georef.csv") %>% 
+  # filter(Country != "Canada") %>%
+  mutate(sample_temp = Sample_id) %>% 
+  separate(sample_temp, into = c("Herb_code", "spp_code", "specimen_code", "tissue_code")) %>% 
+  mutate(species_index = as.factor(case_when(spp_code == "AGHY" ~ "1",
+                                             spp_code == "AGPE" ~ "2",
+                                             spp_code == "ELVI" ~ "3"))) %>% 
+  mutate(species = case_when(spp_code == "AGHY" ~ "A. hyemalis",
+                             spp_code == "AGPE" ~ "A. perennans",
+                             spp_code == "ELVI" ~ "E. virginicus")) %>% 
+  mutate(decade_bin = floor(year/10)*10) %>% 
+  mutate(std_year = (year-mean(year, na.rm = T))) %>%  # I am mean centering but not scaling by standard deviation to preserve units for interpretation of the parameter values
+  filter(scorer_factor != "Scorer26")
+
+
+
+
[…]

### 2024-09-08 Manuscript/endo_herbarium.tex
adding the new conservative scores plot

@@
 $89$\% of scored plants had matching liberal and conservative scores, reflecting high confidence in endophyte status.
-The following analyses used the liberal status, but we repeated all analyses with the conservative status which yielded qualitatively similar results (Fig. \ref{fig:FigA5})
+The following analyses used the liberal status, but we repeated all analyses with the conservative status which yielded qualitatively similar results (Fig. \ref{fig:conservative_scores_plot})
 
@@
 	\centering
-	\includegraphics[width = \linewidth]{fixed_plot.png}
-	\caption{\textbf{Comparison of posterior estimates of fixed effects when using Liberal or Conservative endophyte scores.}}
+	\includegraphics[width = \linewidth]{../Plots/conservative_comparison_plot.png}
+	\caption{Comparison of liberal versus conservative endophyte scores on modeled outcomes.
+(A) Posterior estimates of global temporal trend for models fit to liberal scores (grey) and to conservative scores (blue).
+Maps show the spatially varying temporal trend estimates from model fit to conservative scores for (B) \emph{A.
+hyemalis}, (C) \emph{A.
+perennans}, and (D) \emph{E.
+virginicus}.
+Note that the color scale differs between this visualization and Fig.
+\ref{fig:svc_time_map}.}
+	\label{fig:conservative_scores_plot}
 \end{figure}

### 2024-09-14 Manuscript/endo_herbarium.tex
tom comments

@@
 	\section*{Abstract}
-Species' distributions and abundances are shifting in response to climate change. 
+\tom{Species' distributions and abundances are shifting in response to climate change. 
 Most species harbor microbial symbionts that have the potential to influence these responses.
 Mutualistic microbial symbionts may provide resilience to environmental change by protecting their hosts from increasing stress. 
-However, environmental change that disrupts these interactions may lead to declines in hosts or symbionts. 
-Microbes preserved within herbarium specimens offer a unique opportunity to quantify changes in microbial symbiosis across broad temporal and spatial scales. 
+However, environmental change that disrupts these interactions may lead to declines in hosts and/or symbionts. 
+Symbionts preserved within herbarium specimens offer a unique opportunity to quantify changes in microbial symbiosis across broad temporal and spatial scales.
+}{If you need to cut words, I think some of this could be condensed.}
 We asked how the prevalence of seed-transmitted fungal symbionts of grasses (\emph{Epichloë} endophytes), which can protect hosts from abiotic stress, have changed over time in response to climate change, and how these changes vary across host species' ranges.
-Specifically, we analyzed 2,346 herbarium specimens of three grass host species collected over the last two centuries (1824 -- 2019) for the presence or absence of endophyte symbiosis, and evaluated spatial and temporal trends in endophyte prevalence.
+Specifically, we analyzed 2,346 herbarium specimens of three grass host species (\emph{Agrostis hyemalis},\emph{Agrostis perennans},\emph{Elymus virginicus}) collected over the past two centuries (1824 -- 2019) for the presence or absence of \emph{Epichloë} symbiosis\tom{, and evaluated spatial and temporal trends in endophyte prevalence}{Could cut, overlaps with previous sentence.}.
 We found that endophytes increased in prevalence over the last two centuries from ca. 25\% prevalence to ca. 75\% prevalence, on average, across three host species.
-We also found that changes in prevalence were associated with observed changes in seasonal climate drivers; notably increasing precipitation corresponding to each host species' peak growing season and changes in off-peak season variability in precipitation.
-Our analysis performed favorably in an out-of-sample predictive test with contemporary data, however we identified greater local-scale variability in endophyte prevalence in contemporary data compared to historic data, suggesting that model fusion may be an important step moving forward.
+We also found that changes in prevalence were associated with observed changes in seasonal climate drivers; notably, \tom{increasing precipitation corresponding to each host species' peak growing season and changes in off-peak season variability in precipitation}{I would rephrase this to better communicate the sign of the association, and thus the inferred causality.}.
+Our analysis performed favorably in an out-of-sample predictive test with contemporary survey data, a rare extra step in collections-based research.
+However, we identified greater local-scale variability in endophyte prevalence in contemporary data compared to what our model could predict based on historic data, suggesting new directions that could improve predictive accuracy.
 Our results provide novel evidence for a cryptic biological response to climate change that may contribute to the resilience of host-microbe symbiosis through context-dependent benefits that confer a fitness advantage to symbiotic hosts under environmental change.
@@
 Additionally, symbionts are often unevenly distributed across their hosts' distribution.
-Facultative symbionts may be absent from portions of the host range \cite{afkhami2014mutualist}, and hosts may engage with a diversity of partners (different symbiont species or locally-adapted strains) across their environments \cite{frade2008variation, rolshausen2018quantifying}.
+\tom{Facultative symbionts may be absent from portions of the host range \cite{afkhami2014mutualist}, and hosts may engage with a diversity of partners (different symbiont species or locally-adapted strains) across their environments \cite{frade2008variation, rolshausen2018quantifying}.}{I would encourage self-citation here!}
 Identifying broader spatial trends in symbiont prevalence is therefore an important step in developing predictions for where to expect changes in the symbiosis in future climates.
@@
 The fitness feedback induced by vertical transmission leads to the prediction that endophyte prevalence should be high in populations where these fitness benefits are most important. 
-Previous contemporary survey studies have documented large-scale spatial patterns in  endophyte prevalence structured by environmental gradients \citep{granath2007variation,bazely2007broad, afkhami2012fungal,sneck2017variation}.
-We predicted that prevalence should track temporal changes in environmental drivers that elicit these fitness benefits.
+Previous survey studies of contemporary populations have documented large-scale spatial patterns in  endophyte prevalence structured by environmental gradients \citep{granath2007variation,bazely2007broad, afkhami2012fungal,sneck2017variation}.
+We predicted that prevalence should track temporal changes in environmental drivers that elicit strong fitness benefits.
 %For example, endophyte-mediated drought tolerance should lead endophyte prevalence to increase in regions where precipitation has declined over time.
@@
 This creates the opportunity to leverage the unique spatio-temporal sampling of herbarium collections to examine the response of the symbiosis to historical climate change. 
-Research using historical collections has clearly demonstrated other ecological signatures of a changing climate. 
-However the predictive ability of these historical analyses is rarely tested against contemporary data \citep{lee2024phenological}. 
-Identifying the ways in which these analyses fall short is a crucial step for the field move from reading signatures in the past to forecasting ecological dynamics into the future.
+%Research using historical collections has clearly demonstrated other ecological signatures of a changing climate. 
+However, the predictive ability derived from historical analyses is rarely tested against contemporary data \citep{lee2024phenological}. 
+Critically evaluating whether insights from historical reconstruction are predictive of variation across contemporary populations is a crucial step for the field to move from reading signatures of the past to forecasting ecological dynamics into the future.
 
@@
 and (ii) How spatially variable are temporal trends in endophyte prevalence across eastern North America?
-We then address how climate change may be driving trends in endophyte prevalence by asking: (iii) What is the relationship between variation in temporal trends in endophyte prevalence and changes in climate drivers?
+We then address how climate change may be driving trends in endophyte prevalence by asking: (iii) What is the relationship between temporal trends in endophyte prevalence and associated changes in climate drivers?
 We predicted that aggregate endophyte prevalence would increase over time in tandem with climate warming, and that hotspots of endophyte change would correspond spatially to hotspots of climate change.
-Finally, we evaluated the performance of models built on data from historic specimens with an out-of-sample test, data on endophyte prevalence from contemporary surveys of host populations. 
-To answer these questions we examined a total of 2,346 specimens collected across eastern North America between 1824 and 2019.
+Finally, we evaluated the performance of models built on data from historic specimens with an out-of-sample test, using data on endophyte prevalence from contemporary surveys of host populations. 
+\tom{To answer these questions we examined a total of 2,346 specimens collected across eastern North America between 1824 and 2019.}{Since the preceding sentence is about the out of sample test, maybe include here the numbers of individuals and populations that the historical predictions were tested against?}
 
@@
 Both \emph{Agrostis} species host \emph{Epichloë amarillans} \cite{craven2001multigene, leuchtmann2014nomenclatural}, while \emph{Elymus virginicus} typically hosts \emph{Epichloë elymi} \cite{clay2002evolutionary}.
-These $C_3$ grass species are commonly represented in natural history collections with broad distributions covering much the eastern United States.
-\emph{A. hyemalis} is a small short-lived perennial species that germinates in the spring and typically flowers between March and July (most common collection month: May).
+These $C_3$ grass species are commonly represented in natural history collections with broad distributions covering much the eastern United States (Fig. \ref{figmap}).
+\emph{A. hyemalis} is a small short-lived perennial species that germinates in spring and typically flowers between March and July (most common collection month: May).
 \emph{A. perennans} is of similar stature but is longer lived than \emph{Agrostis hyemalis} and flowers in late summer and early autumn (most common collection month: September). 
@@
 Collections were geo-referenced to the nearest county centroid, or nearest municipality when that information was available. 
-For a few of the oldest specimens, only information at the state level was available, and so we used the state centroid.
+\tom{For a few of the oldest specimens, only information at the state level was available, and so we used the state centroid.}{Say how many?}
 
-After collecting seed samples, we quantified the presence or absence of \emph{Epichloë} fungal hyphae, which grow intercellularly, in each specimen using microscopy. 
+After collecting seed samples, we quantified the presence or absence of \emph{Epichloë} fungal hyphae in each specimen using microscopy. 
 We first softened seeds with a 10\% NaOH solution, then stained the seeds with aniline blue-lactic acid stain and squashed them under a microscope cover slip. 
 We examined the squashed seeds for the presence of fungal hyphae at 200-400X magnification \cite{bacon2018stains}.
-In some cases, the tissues examined during microscopy came from flowers or otherwise non-viable seeds, which were excluded for that specimen.
+\tom{In some cases, the tissues examined during microscopy came from flowers or otherwise non-viable seeds, which were excluded for that specimen.}{Could cut, since you already say you only took good seeds.}
 On average we scored $4.7$ seeds per specimen of \emph{A.
@@
 Due to imperfect vertical transmission \cite{afkhami2008symbiosis}, it is possible that symbiotic host-plants produce a mixture of symbiotic and non-symbiotic seeds. 
-We therefore designated a specimen as endophyte-symbiotic if \emph{Epichloë} hyphae were observed in one or more of its seeds, or non-symbiotic if hyphae were observed in none of its seeds. 
+We therefore designated a specimen as endophyte-symbiotic if \emph{Epichloë} hyphae were observed in one or more of its seeds, or non-symbiotic if \emph{Epichloë} hyphae were observed in none of its seeds.
 To capture uncertainty in the endophyte scoring process, we recorded both a "liberal" and a "conservative" endophyte status for each plant specimen.  
@@
 $89$\% of scored plants had matching liberal and conservative scores, reflecting high confidence in endophyte status.
-The following analyses used the liberal status, but we repeated all analyses with the conservative status which yielded qualitatively similar results (Fig. \ref{fig:conservative_scores_plot})
+The following analyses used the liberal status, but we repeated all analyses with the conservative status which yielded qualitatively similar results (Fig. \ref{fig:conservative_scores_plot}).
 
@@
 This SPDE approach is a flexible method of smoothing across space while explicitly accounting for spatial dependence between data-points \citep{lindgren2011explicit,bakka2018spatial}.
-Fitting models with structured spatial effects is possible with MCMC sampling but can require long computation times, making INLA an effective alternative, which has been used to model spatial patterns in flowering phenology \cite{willems2022forest}, the abundance of bird species \cite{meehan2019spatial} and butterflies \cite{crossley2022opposing}, the distribution of temperate trees \cite{engel2022spatial} as well as the population dynamics of endangered amphibians \cite{knapp2016large} and other ecological processes \cite{beguin2012hierarchical}.
+Fitting models with structured spatial effects is possible with MCMC sampling but can require long computation times, making INLA an effective alternative. 
+\tom{This approach has been used to model spatial patterns in flowering phenology \cite{willems2022forest}, the abundance of birds \cite{meehan2019spatial} and butterflies \cite{crossley2022opposing}, the distribution of temperate trees \cite{engel2022spatial} as well as the population dynamics of endangered amphibians \cite{knapp2016large} and other ecological processes \cite{beguin2012hierarchical}.}{Probably not necessary if you need space.}
 
 We estimated global and spatially-varying trends in endophyte prevalence using a joint-likelihood model. 
-For each host species $h$, endophyte presence/absence of the $i^{th}$ specimen ($P_{[h]i}$) was modeled as a Bernoulli response variable with expected probability of endophyte occurrence $\hat{P}_{[h]i}$.
-We modeled $\hat{P}_{[h]i}$ as a linear function of intercept $ \mathrm{A}_{[h]i}$ and slope $\mathrm{T}_{[h]}$ defining the global trend in endophyte prevalence specific to each host species as well as with spatially-varying intercepts $\alpha_{[h_{1}]l[i]}$ and slopes $\tau_{[h_{1}]l[i]}$ associated with location ($l[i]$, a unique latitude-longitude combination).
-The joint-model structure allowed us to share variance terms across focal species to account for dependence associated with the collection of specimens and identification of endophytes. 
-Shared variance terms included the spatially-dependent random effect $\delta_{l[i]}$, intended to account for residual spatial variation, and $\chi_{c[i]}$ and $\omega_{s[i]}$ i.i.d.-random effects indexed for each collector identity ($c[i]$), and scorer identity ($s[i]$) of the $i^{th}$ specimen.
+For each host species $h$, endophyte presence/absence of the $i^{th}$ specimen \tom{($P_{[h]i}$)}{Not obvious to me why you use square brackets here.} was modeled as a Bernoulli response variable with expected probability of endophyte occurrence $\hat{P}_{[h]i}$.
+We modeled $\hat{P}_{[h]i}$ as a linear function of intercept \tom{$ \mathrm{A}_{[h]i}$}{I don't think there should be an $i$ subscript here.
+You need $i$ only where there is a covariate unique to the $i$th observation.} and slope $\mathrm{T}_{[h]}$ defining the global trend in endophyte prevalence specific to each host species as well as with spatially-varying intercepts $\alpha_{[h_{1}]l[i]}$ and slopes $\tau_{[h_{1}]l[i]}$ associated with location ($l[i]$, the unique latitude-longitude combination of the $i$th observation).
+The joint-model structure allowed us to ``borrow strength'' across species in the estimation of shared variance terms for the spatially-dependent random effect $\delta_{l[i]}$, intended to account for residual spatial variation, and $\chi_{c[i]}$ and $\omega_{s[i]}$ i.i.d.-random effects indexed for each collector identity ($c[i]$), and scorer identity ($s[i]$) of the $i^{th}$ specimen.\tom{}{I don't think you need three expressions here.
+If you just index host as $h$ it is clear that some pieces are unique to host $h$ and others are shared across hosts.}
 
@@
 
-Previous work suggests that behavior of historical botanists and uneven sampling may introduce biases into ecological inferences made from historic collections \cite{kozlov2020biases}. 
+By including random effects for collectors and scorers, we accounted for ``nuisance'' variance that may bias predictions for changes in endophyte prevalence.
+Previous work suggests that behavior of historical botanists may introduce biases into ecological inferences made from historic collections \cite{kozlov2020biases}. 
 Prolific collectors who contribute thousands of specimens may be more or less likely to collect certain species, or specimens with certain traits \cite{daru2018widespread}. 
-Similarly, the process of scoring seeds for hyphae involved several student researchers who, even with standardized training, may vary in their likelihood of positively identifying \emph{Epichloë} hyphae.
-By including a random effect for collectors and for scorers, we attempted to account for variance across individual researchers that may bias our predictions of changes in endophyte prevalence.
+Similarly, the process of scoring seeds for hyphae involved several student researchers who, even with standardized training, may vary in their likelihood of positively identifying \emph{Epichloë}.
 
@@
 Scorer and collector random effects, $\chi$ and $\omega$, were given penalized complexity priors, with distributions approximating a Normal distribution with standard deviation of 5. 
-Each spatially-structured parameter depended on a covariance matrix according to the proximity of each collection location \citep{lindgren2011explicit,bakka2018spatial}. 
+Each spatially-structured parameter depended on a covariance matrix according to the proximity of each pair of collection locations \citep{lindgren2011explicit,bakka2018spatial}. 
 The covariance matrix was approximated using a Mat\'{e}rn covariance function, with each data point assigned a location according to the nodes of a mesh of non-overlapping triangles encompassing the study area (Fig.
@@
 We assessed model fit with visual posterior predictive checks (\ref{fig:overlayplot}) and measurements of AUC (Figs. \ref{fig:ROCtraining}-\ref{fig:ROCtest}).
-Priors, termed ''range" and "variance", define the distance of spatial decay described by the Mat\'{e}rn covariance function.
-Priors for results presented in the main text reflect a range of $342$ kilometers. 
-We found that model results were sensitive to this choice, and so tested a range of priors (from 68 kilometers to 2160 kilometers) and meshes (Supplemental Material), finding that model results were qualitatively similar, i.e.
+Priors for the Mat\'{e}rn covariance function, termed ''range" and "variance", define how proximity effects decay with distance.
+\tom{Priors for results presented in the main text reflect a range of $342$ kilometers. 
+We found that model results were somewhat sensitive to the choices of priors for these parameters, so we tested a range of values (from 68 kilometers to 2160 kilometers) and meshes (Supplemental Material), finding that model results were qualitatively similar, i.e.
 the same direction of effects across space, but that the magnitude and uncertainty varied.
+}{I struggled here.
+I tried to make some edits but I don't think I understand the methods well enough to edit correctly.
+It is not clear how the distances you quote should be interepreted (locations withing 68 or 2160 km are more likely to have the same endo status just due to proximity?), or whether they should be interpreted at all.
+The text seems to say both the priors do not matter, but also that they do.
+I have not looked at what is in the supplement yet, but if you present results there based on different priors it should be easy to write this in a way that succinctly says ``priors mattered a little bit, and if you care you can look at the supplement for a prior sensitivity analysis.''}
+
 %In all cases, posterior modes were \tom{stable}{Assessed how?} and equal to zero, indicating successful numeric convergence.
@@
 \subsection*{Validating model performance with in-sample and out-of-sample tests}
-We evaluated the predictive ability of the model using both in-sample training data from the herbarium surveys, and with out-of-sample test data from contemporary endophyte surveys, an important but rarely used strategy in ecological studies \cite{tredennick2021practical, lee2024phenological}.
-We used data from contemporary surveys of endophyte prevalence  in \emph{A. hyemalis} and \emph{E. virginicus} in Texas and the southern US. 
+We evaluated the predictive ability of the model using both in-sample training data from the herbarium surveys, and with out-of-sample test data, an important but rarely used strategy in ecological studies \cite{tredennick2021practical, lee2024phenological}.
+We generated out-of-sample test data from contemporary surveys of endophyte prevalence in natural populations of \emph{A. hyemalis} and \emph{E. virginicus} in Texas and the southern US. 
 Surveys of \emph{E. virginicus} were conducted in 2013 as described in \citet{sneck2017variation}, and surveys of \emph{A. hyemalis} took place between 2015 and 2020.
-%\tom{surveys of \emph{A.
-hyemalis} took place between 2015 and 2020}{We have added more recent AGHY survey data.
-I am not sure if you have access to this but you should definitely use it.
-Karl or I can point you to the right file.}.
 Population surveys of \emph{A.
 hyemalis} were initially designed to cover longitudinal variation in endophyte prevalence towards its range edge, while surveys of \emph{E.
-virginicus} were designed to cover latitudinal variation along its range edge.
+virginicus} were designed to cover latitudinal variation.
 In total, we visited 43 populations of \emph{A.
@@
 virginicus} across the south-central US, with emphasis on Texas and neighboring states (Fig \ref{fig:contempsurveysmap}).
-During surveys, we collected seeds from up to 30 individuals per location  (average number of plants sampled: $22.9$).
-We quantified the endophyte status of each individual with staining microscopy as described for the herbarium surveys (with 5-10 seeds scored per individual), and calculated the prevalence of endophytes within the population (proportion of symbiotic plants divided by the number of sampled plants).
-For each population, we compared the observed fraction of endophyte-symbiotic hosts to the predicted probability of endophyte occurrence $\hat{P}$ derived from the model based on location and year. 
+During surveys, we collected seeds from up to 30 individuals per population (average number of plants sampled per population: $22.9$); note that this sampling design provided greater local depth of information than the herbarium records, where only one plant was sampled at each locality.
+We quantified the endophyte status of each individual with staining microscopy as described for the herbarium surveys (with 5-10 seeds scored per individual), and calculated the prevalence of endophytes within the population (proportion of plants that were endophyte-symbiotic).
+For each population, we compared the observed fraction of endophyte-symbiotic hosts to the predicted probability of endophyte occurrence $\hat{P}$ derived from the model for that location and year. 
 The contemporary survey period (2013-2020) is at the most recent edge of the time period encompassed by the historical observations used for model fitting.
-We compared the model's prediction for these locations to the observed population prevalence.
+%We compared the model's prediction for these locations to the observed population prevalence.
 
@@
 We then took the difference between recent and historic periods for the mean and standard deviation for each climate driver (Figs. \ref{fig:AGHY_climate_covariates}-\ref{fig:ELVI_climate_covariates}).
-All together, we assessed twelve potential climate drivers: the mean and standardiation of spring, summer, and autumn temperature, as well as the mean and standard deviation of spring, summer, and autumn cumulative precipitation, cumulative precipitation, and cumulative precipitation.
+All together, we assessed twelve potential climate drivers: the mean and standard deviation of spring, summer, and autumn temperature, as well as the mean and standard deviation of spring, summer, and autumn cumulative precipitation.
 
-To evaluate whether areas that have experienced the greatest changes in endophyte prevalence (hotspots of endophyte change) are associated with high degrees of change in climate (hotspots of climate change), we modeled spatially varying slopes of endophyte change through time ($\tau_{[h]l}$ as a linear function of environmental covariates, with a Gaussian error distribution.
+\tom{To evaluate whether areas that have experienced the greatest changes in endophyte prevalence (hotspots of endophyte change) are associated with high degrees of change in climate (hotspots of climate change), we modeled the fitted, spatially-varying slopes of endophyte change through time ($\tau_{[h]l}$ as a linear function of environmental covariates, with a Gaussian error distribution.}{If I am reading the results figure correctly it looks like you force an intercept of zero, and I am not sure why you would want to do that.}
 Data from each host species was analyzed separately. 
-Fitting regressions to many pixels across the study region risks articially inflating confidence in our results due to large sample sizes, and so we performed this analysis using only a random subsample of 250 pixels across the study region, which provided results qualitatively similar to analysis of the full set of pixels.
+Fitting regressions to many pixels across the study region risks articially inflating confidence in our results due to large sample sizes, and so we performed this analysis using only a random subsample of 250 pixels across the study region; other sizes of subsample yielded similar results.
 
 \subsection*{Modeling distributions of host species}
-We modeled the distribution of each host species to generate maps on which we predicted the dynamics of \emph{Epichloë} symbionts.
-We followed the ODMAP (overview, data, model, assessment, prediction) protocol \citep{crossley2022opposing}.
-A full description of the ODMAP can be found in the  (Supporting Methods). 
- In short, we used presence-only observations of the host species from Global Biodiversity Information Facility (GBIF) between 1990 to 2020.
-To reduce the potential influence of sampling bias and spatial autocorrelation, we thinned the occurrences to the spatial scale  of our selected climatic predictors. 
-We selected climate variables that aligned with our analysis of climatic influences on trends in endophyte prevalence described above.
-We calculated the mean and standard deviation of seasonal temperature and precipitation across 1990 to 2020. 
-Among this suite of variables, we chose to include, which were uncorrelated (Variance Inflation Factor $>$ 0.7) and allowed us to predict the occurrence probability of each host species in space and time.
-These climatic variables are: the mean and standard deviation of spring, summer, and autumn temperature, as well as the mean and standard deviation of spring, summer, and autumn cumulative precipitation, cumulative precipitation, and cumulative precipitation.
-We fit maximum entropy (MaxEnt) models using the maxent function in the package dismo \citep{hijmans2017package}. 
+Because the herbarium records did not encompass the entirety of these host species'ranges, we additionally modeled the geographic distribution of each host species to generate realistic maps on which we could project the predictions of the INLA model.
+We followed the ODMAP (overview, data, model, assessment, prediction) protocol \citep{crossley2022opposing} (see Supporting Methods). 
+In short, we used presence-only observations of each host species from Global Biodiversity Information Facility (GBIF) \tom{between 1990 to 2020}{I am not sure why 1990-2020.
+From our conversations, I thought the SDM was based on all records.
+At this point I would not change anything.}.
+%To reduce the potential influence of sampling bias and spatial autocorrelation, \tom{we thinned the occurrences to the spatial scale of our selected climatic predictors}{I do not follow.}. 
+%We selected climate variables that aligned with our analysis of climatic influences on trends in endophyte prevalence described above.
+%\tom{We calculated the mean and standard deviation of seasonal temperature and precipitation across 1990 to 2020. }{FOr what locations? Is this from the whole raster of NA?}
+%\tom{Among this suite of variables, we chose to include, which were uncorrelated (Variance Inflation Factor $>$ 0.7) and allowed us to predict the occurrence probability of each host species in space and time.}{Do not follow and sounds circular.}
+We fit maximum entropy (MaxEnt) models using the maxent function in the R package dismo \citep{hijmans2017package} using the same set of seasonal climate predictors considered above: \tom{mean and standard deviation of spring, summer, and autumn temperature, and mean and standard deviation of spring, summer, and autumn cumulative precipitation.}{Were these 1990-2020 normals?}
 %\josh{MaxEnt is preferred because it has been shown to generate response curves with less unpredictable behavior when applied to new climates \citep{hijmans2006ability}.}{possibly could remove this sentence?}
@@
 The performance of models was evaluated with AUC \citep{jimenez2012insights}. 
-We found AUC = 0.862, AUC = 0.838, AUC = 0.821 respectively  for \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}.
-To convert the continuous predicted probabilities into binary presence - absence maps, we used the  training sensitivity (true positive rate) and specificity threshold (true negative rate) \citep{liu2005selecting}.
-These binary maps serve as boundaries in presented maps of change in endophyte prevalence, and outline the set of pixels used in our analysis of climate correlates with trends in endophyte prevalence
+\tom{We found AUC values of 0.862, 0.838, 0.821 respectively  for \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}.}{Can we make this sentence more evaluative?
+I think AUC > 0.8 is very good, yes?}
+To convert the continuous predicted probabilities into binary presence - absence maps on which we projected INLA predictions, we used the  training sensitivity (true positive rate) and specificity threshold (true negative rate) \citep{liu2005selecting}.
+%These binary maps serve as boundaries in presented maps of change in endophyte prevalence, and outline the set of pixels used in our analysis of climate correlates with trends in endophyte prevalence
 
@@
 \subsection*{How has endophyte prevalence changed over time?}
-We found that endophyte prevalence increased within the examined specimens over the last two centuries for all three host species (Fig. \ref{fig:temporal}). 
-On average, modeling indicated that endophytes of \emph{A.
+Across >2300 herbarium specimens dating back to 1824, we found that prevalence of \emph{Epichloë} endophytes increased over the last two centuries for all three grass host species (Fig.
+\ref{fig:temporal}).
+On average, endophytes of \emph{A.
 perennans} and \emph{E.
-virginicus} increased from $\sim 40$ \% to  $70$\% prevalence across the study region, and that of \emph{A.
+virginicus} increased from $\sim 40$ \% to  $70$\% prevalence across the study region, and \emph{A.
 hyemalis} increased from $\sim 25$\% to over $50$\% prevalence.
@@
 virginicus}) (Fig.
-\ref{fig:temporal_posterior})
+\ref{fig:temporal_posterior}).
+The model appears to under-predict the observed increase in endophyte prevalence relative to the data, particularly for \emph{A.
+hyemalis} (Fig.
+\ref{fig:temporal}A), but the model is accounting for random effects and spatial non-independence that are not readily seen in the figure.
+By under-weighting endophyte-positive samples that are clustered spatially or by collector or observer, the INLA model is appropriately accounting for nuisance variables and providing a conservative inference of endophyte change relative to the raw data.
 
@@
 \subsection*{How spatially variable are temporal trends in endophyte prevalence?}
-Our model revealed hotspots of change in endophyte prevalence. 
-While there was an overall increase in endophyte prevalence, these changes varied across the host species' ranges (Fig. \ref{fig:svc_time_map}).
-In some regions, posterior estimates of spatially varying temporal trends, $\tau$, indicate that \emph{A.
+While there was an overall increase in endophyte prevalence, our model revealed hotspots and coldspots of change across the host species' ranges, which are mapped in Fig.
+\ref{fig:svc_time_map} across geographic ranges predicted by MaxEnt species distribution models.
+In some regions, posterior mean estimates of spatially varying temporal trends indicate that \emph{A.
 hyemalis} and \emph{A.
-perennans} experienced increases in percent prevalence by as much as $2$\% per year over the study period, while  \emph{E.
+perennans} experienced increases in prevalence by as much as $2$\% per year over the study period, while  \emph{E.
 virginicus} experienced increases up to around $1$\% per year.
-Compared to \emph{E.
-virginicus}, which had a weaker overall increase in endophytes and less spatial variability, maps of both \emph{Agrostis} species show areas of strong increase and areas of declining prevalence.
-Notably, endophytes increased towards the western range edge of \emph{A.
+Both \emph{Agrostis} species show areas of strong increase and areas of declining prevalence, while \emph{E.
+virginicus} had an overall weaker and geographically more consistent increase in endophyte prevalence.
+Notably, endophytes increased most strongly towards the western range edge of \emph{A.
 hyemalis} (Fig.
@@
 \subsection*{What is the relationship between variation in temporal trends in endophyte prevalence and changes in climate drivers?}
-
+\tom{}{I think this section would be strengthed by greater statistical support.
+It is really just eye-balling lines.
+Can we say something more quantitative?
+See my comment about posteriors of slopes -- that is the type of information I have in mind, and maybe we could embed some of this in the text.}
 We found that trends in endophyte prevalence were strongly associated with seasonal climate change drivers (Fig. \ref{fig:climate_correlation_plot}).
-For the majority of the study region, the climate has become wetter and cooler over the last century (Fig.
-\ref{fig:AGHY_climate_covariates}-\ref{fig:ELVI_climate_covariates}), a consequence of regional variation in global climate change \cite{ipcc_2021}.
-Within the study region, spatial variation in climate trends were predictive of trends in endophyte prevalence.
-For example, strong increases in prevalence within \emph{A. perennans} were most associated with autumn climate drivers that coincide with its Aug-Sep active growing season. 
-For this species, warmer and wetter autumn climates showed particularly strong relationships, however other seasonal drivers may also contribute to increasing endophyte prevalence (drier springs and cooler summers).
-Trends in endophyte prevalence for \emph{A. hyemalis} were most strongly associated with changes in precipitation variability were the strongest predictors.
-Endophyte prevalence increased the most in regions that experienced greater spring precipitation along with increasing variability in summer and autumn precipitation. 
-While this species actively grows and reproduces in the late spring and early summer, climate effects outside of the growing season may indicate that endophytes play a role in persistence during dormant periods through summer droughts or contribute to the ability to successfully germinate.
-Prevalence of endophytes of \emph{E. virginicus} were least influenced by climate, but decreasing autumn temperature variability and less precipitation in autumn were the strongest predictors.
-
-Correlations assessed using all pixels across each species distribution were qualitatively similar to these results (Fig. A11).
+For the majority of the study region, \tom{the climate has become wetter and cooler over the last century}{Can you provide the average change in degrees celsius and mm precip across the range?} (Fig.
+\ref{fig:AGHY_climate_covariates}-\ref{fig:ELVI_climate_covariates}), a consequence of regional variation in global climate change \cite{ipcc_2021}.\tom{}{Comment on these supplement figures: perhaps for the revision, I think a color scale that diverges at zero would be helpful here.}
+Spatial variation in climate trends were predictive of trends in endophyte prevalence.
+\tom{For example, strong increases in endophyte prevalence for \emph{A.
+perennans} were most strongly associated with change in autumn climate drivers that coincide with its active growing season.
+}{Here and throughout the results, please reference specific figures and panels for statements like this so readers can see what you see.
+Regarding the figure, since it is just a bunch of regression lines and we only care about the slopes, it might be more effective to just show posteriors of the slopes - probably not worth making that change at this point, and maybe it's ugly.}
+For this species, regions experiencing warmer and wetter autumn climates showed particularly strong relationships with increasing endophyte prevalence, however other seasonal drivers may also contribute (drier springs and cooler summers).
+Trends in endophyte prevalence for \emph{A. hyemalis} were most strongly associated with changes in precipitation and precipitation variability. 
+Endophyte prevalence in \emph{A.
+hyemalis} increased the most in regions that experienced greater increases in spring precipitation, as well as regions with changing variability in spring and autumn precipitation.
+%While this species actively grows and reproduces in the late spring and early summer, climate effects outside of the growing season may indicate that endophytes play a role in persistence during dormant periods through summer droughts or contribute to the ability to successfully germinate.
+\tom{Changes in endophyte prevalence in \emph{E.
+virginicus} were not strongly associated with changes in most climate drivers}{Notice that my edits here importantly change the meaning of the sentence and, as far as I understand, make it more correct.}, but regions of reduced variability in autumn precipitation were associated with weak reductions in prevalence.
+\tom{Correlations}{I am not sure why you describe these as ``correlations''.
+The methods makes these sound like linear regressions.} assessed using all pixels across each species' distribution were qualitatively similar to these results (Fig.
+A11).
 %These associations are in line with expectations that endophyte symbiosis provides drought tolerance to their hosts, particularly during \emph{E.
@@
 \subsubsection*{Performance on test data}
-We found that model performance, as judged by AUC, was similar between historic herbarium specimens used as training data and the out-of-sample test data from contemporary surveys (0.79 and 0.77 respectively; Fig.
+\tom{}{The order of the results does not match the order of the methods.}
+We found that model performance, as quantified by AUC, was similar between historic herbarium specimens used as training data and the out-of-sample test data from contemporary surveys (0.79 and 0.77 respectively; Fig.
 \ref{fig:ROCtest}-\ref{fig:ROCtraining}).
-The model successfully captured broader regional trends in endophyte prevalence present in the contemporary survey data, such as decline endophyte prevalence towrads western longitudes in \emph{A.
-hyemalis}  (Fig.
-\ref{fig:contemptestplot}A).
-However, the contemporary data contains additional variability at smaller scales not captured by our sampling of herbarium specimens. 
-We interpreted this to mean that the model captured regional spatial dynamics, but underpredicts local scale dynamics.
-We discuss potential model improvements in the Discussion.
+The model successfully captured broad regional trends in endophyte prevalence seen in the contemporary survey data, such as decline endophyte prevalence in \emph{A.
+hyemalis} towards western longitudes (Fig.
+\ref{fig:contemptestplot}A) and northern latitudes (Fig.
+\ref{fig:contemptestplot}B).
+However, model predictions for endophyte prevalence exhibited relatively little geographic variation, whereas the out-of-sample survey data were maximally variable with populations spanning 0\% to 100\% endophyte-symbiotic plants (Fig.
+\ref{fig:contemptestplot}C)\tom{}{Love the figure but I think the legend can be improved.
+It does not describe the content very well.}.
+We interpret this to mean that the model captures coarse-scale spatial and temporal trends reasonably well, but is not equipped to capture local-scale nuances that generate population-to-population differences.
+%We discuss potential model improvements in the Discussion.
 
@@
 \subsection*{Assessing collector and scorer influences on predicted endophyte prevalence}
-Our modeling effort quantified temporal and spatial trends in endophyte prevalence while accounting for potential biases introduced by collectors and by individual scorers who quantified endophyte presence/absence within specimens with the use of random effects.
+Our modeling effort quantified temporal and spatial trends in endophyte prevalence while accounting for potential biases introduced by collectors and by individual scorers who quantified endophyte presence/absence within specimens.
 We found no evidence that collector biases influenced our results. 
@@
 \ref{fig:scorer_fx}).
-This may have been driven by differences in scorers biases during the seed scoring process or by unintended spatial clustering of the specimens scored by each scorer \cite{clayton1993spatial,urdangarin2023evaluating}.
+It is difficult to distinguish whether this was driven by true differences in scorers biases during the seed scoring process or by unintended spatial or temporal clustering of the specimens scored by each scorer \cite{clayton1993spatial,urdangarin2023evaluating}.
 Interpreting our models with the inclusion of variance associated with the scorer effect thus provides conservative estimates of the absolute magnitude of changes in endophyte prevalence.
@@
 \section*{Discussion}
-Our examination of historic plant specimens revealed a cryptic biotic reponse to climate change. 
-For the three host species we examined, there have been clear increases in fungal endophyte prevalence over the last two centuries.
-Increases in prevalence of \emph{Epichloë}, which are vertically transmitted, can potentially be interpreted as adaptive changes that improve the fitness of their hosts under stressful conditions.
-This interpretation is in line with theory predicting that the positive fitness feedback caused by vertical transmission leads beneficial symbionts to rise in prevalence within a population \cite{fine1975vectors}.
-We found that trends in endophyte prevalence varied across the distribution of each species in association with observed changes in climate drivers, suggesting that the endophytes have contributed to host resilience under environmental change.
-Taken together, this suggests a strengthening of the mutualism over the last two centuries.
+Our examination of historic plant specimens revealed cryptic shifts in microbial symbiosis over the last two centuries. 
+For the three host species we examined, there have been strong increases in prevalence of fungal endophytes.
+We interpret increases in prevalence of \emph{Epichloë}, which are vertically transmitted, as adaptive changes that improve the fitness of their hosts under increasing environmental stress.
+This interpretation is in line with theory predicting that the positive fitness feedback caused by vertical transmission leads beneficial symbionts to rise in prevalence within a population \cite{fine1975vectors,donald2021context}.
+We further found that trends in endophyte prevalence varied across the distribution of each species in association with changes in climate drivers, suggesting that the increases in endophyte prevalence are driven by context-dependent benefits to hosts that confer resilience under environmental change.
+Taken together, this suggests an overall strengthening of host-symbiont mutualism over the last two centuries.
 
-Differences between the responses of each host species underscore that while all of these $C_3$ grasses share similar broad-scale distributions, each engages in unique biotic interactions and has unique niche requirements.
-We identified hotspots of change for \emph{A. perennans}, which experienced the strongest absolute changes in endophyte prevalence (Fig. \ref{fig:svc_time_map}).
-Declines in the southern portion of its range and increases in the north suggest a potential poleward range shift of endophytic plants. 
-Based on previous work demonstrating that endophytes can shield their hosts from drought stress \cite{decunta2021systematic}, we generally predicted that drought conditions could be a driver of increasing endophyte prevalence.
-In contrast to this expectation, increasing prevalence for this species was associated with increasing autumn temperature and precipitation (Fig. \ref{fig:climate_correlation_plot}). 
+Differences across host species underscore that while all of these $C_3$ grasses share similar broad-scale distributions, each engages in unique biotic interactions and has unique responses to environmental drivers.
+We identified hotspots of change for \emph{A.
+perennans}, which was the species that experienced the \tom{strongest absolute changes}{remind readers what the absolute change was} in endophyte prevalence (Fig.
+\ref{fig:svc_time_map}).
+Declines in the southern portion of its range and increases in the north suggest a potential poleward range shift of endophyte-symbiotic plants (whether the oervall host distribution is shifting in parallel is an exciting next question).
+Based on previous work demonstrating that endophytes can shield their hosts from drought stress \tom{\cite{decunta2021systematic}}{I would cite this as ``reviewed in'' or something like that, so this is not read as a single case study}, we generally predicted that drought conditions would be a driver of increasing endophyte prevalence.
+In contrast to this expectation, increasing prevalence \tom{for this species}{I have now forgotten which species you are talk about, since the previous sentence is about grass-endophyte interactions more generally.
+I think this paragraph can be edited to make it clearer which are general statements and which refer to specific species (and which ones).} was associated with increasing autumn temperature and precipitation (Fig.
+\ref{fig:climate_correlation_plot}).
 To our knowledge, the response of the symbiosis in \emph{A.
@@
 Our results also hint that it may be useful to investigate whether lagged climate effects are important predictors of host fitness in this system \cite{evers2021lagged}.
-Endophyte prevalence of the spring-flowering \emph{A. hyemalis} was most strongly linked to increasing variability in precipitation across summer and autumn.  
-Endophytes could be playing a role helping hosts weather autumn-season droughts while the species is dormant.
+\tom{Endophyte prevalence of the spring-flowering \emph{A.
+hyemalis} was most strongly linked to increasing variability in precipitation across summer and autumn.
+}{I do not see this in the figure.
+I would argue you are over-interpreting the SD effects.
+The ppt amount effects look stronger, and these are in autumn and spring which lines up with this species' growing season.
+Marion's paper showed that wet autumn conditions can neutralize the benefits of the symbiosis, which I think is consistent with your result.}
+Endophytes could be playing a role helping hosts weather autumn-season droughts \tom{while the species is dormant.}{Not dormant in autumn. They probably start germinating in October.}
 Previous work has demonstrating drought benefits in a greenhouse manipulation with this species \citep{davitt2011understanding}, and early life stages may be particularly vulnerable to prolonged droughts.
@@
 Increases in symbionts could be explained, at least in part, by these diverse benefits that may help hosts weather a world made increasingly stressful by changes in climate and other anthropogenically introduced stressors.
-While we show consistent increasing trends in prevalence between the three species, the mechanisms that explain these changes may be diverse and idiosyncratic.
+\tom{While we show consistent increasing trends in prevalence between the three species, the mechanisms that explain these changes may be diverse and idiosyncratic.}{This paragraph was a little hard to read and I think does not really nail the points that I think we should make here, IMO.
+I would open to paragraph setting up the idea that the correlations you detected have some support in previous experimental studies, then elaborate.
+Then I would acknowledge that other factors can elicit benefits of symbiosis, and these factors may ALSO be changing.
+The takeaway, for me and perhaps for readers more generally, is that we are very confident endophytes are increasing, we have some confidence that is associated with (and perhaps due to) climate change, but a lot has changed over the past 200 years (climate, urbanization, nitrogen deposition, invasive species, deer!) and we need follow-up studies and experiments to gain a better mechanistic understanding of what has driven these changes, which will also allow us to build better forecasts about the future.}
 
-Our spatially-explicit model predicted regions of both high and low endophyte prevalence, suggesting that symbiotic and non-symbiotic host plants have overlapping, but non-identical niche requirements.
-Endophytes fitness benefits potentially explain the spatial distribution of prevalence by allowing their hosts to persist in environments where they otherwise could not \citep{afkhami2014mutualist, kazenel2015mutualistic}.
-For example, fitness benefits of the symbiosis could explain historically low prevalence in \emph{A. hyemalis} towards its western range edge coinciding with a strong aridity gradient.
-Previous population surveys for endophytes, which were used as test data for our model, found similar regional trends in prevalence for endophyte host species \citep{sneck2017variation,rudgers2009benefits}.
-While the model recreated these large-scale spatial trends, test data contained more population-to-population variability in prevalence. 
-Validating our model predictions in this way allows us to evaluate places to improve the model's out-of-sample predictive ability, which will be particularly important for predicting host and symbiont niche-shifts under future climate change.
+\tom{Our spatially-explicit model predicted regions of both high and low endophyte prevalence, suggesting that symbiotic and non-symbiotic host plants have overlapping, but non-identical niche requirements.}{I think this paragraph tries to do too much relative to this topic sentence.
+In general I think you could use stronger topics sentences to give these Discussion paragraphs greater focus.}
+Endophytes fitness benefits potentially explain the spatial distribution of prevalence by allowing their hosts to persist in environments where they otherwise could not \tom{\citep{afkhami2014mutualist, kazenel2015mutualistic}}{cite Eco Monograph}.
+\tom{For example, fitness benefits of the symbiosis could explain historically low prevalence in \emph{A.
+hyemalis} towards its western range edge coinciding with a strong aridity gradient.}{But above you say that AGHY gets benefits under drought.}
+Previous population surveys for endophytes found similar regional trends in prevalence for endophyte host species \tom{\citep{sneck2017variation,rudgers2009benefits}}{I would add the semmartin study.}.
+While the model recreated the large-scale spatial trends observed in contemporary population surveys, test data contained more population-to-population variability in prevalence. 
+Validating our model predictions in this way, a rare extra step in collections-based studies, allows us to evaluate places to improve the model's out-of-sample predictive ability, which will be particularly important for predicting host and symbiont niche-shifts under future climate change.
 Lack of information on local variability may simply be a feature of data derived from herbarium specimens. 
-They are samples from local populations, but they are single specimens that are aggregated over in broad-scale model estimates.
+They are samples from local populations, but they are single specimens that are aggregated to derive broad-scale model estimates.
 Poor predictive ability at local scales in this grass-endophyte system is not surprising, as previous studies have found that local variation, even to the scale of hundreds of meters can structure endophyte-host niches \cite{kazenel2015mutualistic}.
 Other studies have found factors including land-use history \cite{vikuk2019infection} and the biotic environment, including herbivory \cite{rudgers2016long}, and host genotype \citet{sneck2017variation}, to be important predictors of endophyte ecology.
-Incorporating available climatic and soil layers as covariates is an obvious first step that could improve predictions.
+\tom{Incorporating available climatic and soil layers as covariates is an obvious first step that could improve predictions.}{This is not where my mind went.
+I was thinking more local replication/redundancy, e.g.
+multiple specimens from the same population.
+Your idea is getting at explaining the local variation, but right now the model is not even seeing the local variation because we (rightly I think) prioritized larger-scale geographic coverage.}
 Another important step would be integrating data from local and regional scales  through modeling to constrain estimates of local and regional variation.
@@
 The responses of microbial symbioses are a rich target for future studies within museum specimens, particularly those that take advantage of advances in sequencing technology.
+\tom{}{This is not the place for it, but this sentence made me think about genetic variation in the endophytes.
+We see presence-absence changing but the endophytes themselves could also be changing (``chemotypes'') which is something we could possibly detect if we could get fungal DNA out of the samples.
+Maybe worth mentioning this somewhere.}
 While we used relatively coarse presence/absence data based on fungal morphology, other studies have examined historic plant microbiomes using molecular sequencing and sophisticated bioinformatics techniques, but these studies have so far been limited to relatively few specimens at limited spatial extents \cite{yoshida2015computational,heberling2019utilizing, bieker2020metagenomic,gross2021hidden,bradshaw2021global}.
 Continued advances in capturing historic DNA and in filtering out potential contamination during specimen storage \citep{daru2019novel,bakker2020herbarium,raxworthy2021mining} will be imperative in the effort to scale up these efforts.
-This scaling up will be essential to be able to quantify changes not just in the prevalence of symbionts, but also in symbionts' intraspecific variation and evolutionary responses to climate change, as well as  in changes in the wider microbial community.
+This scaling up will be essential to be able to quantify changes not just in the prevalence of symbionts, \tom{but also in symbionts' intraspecific variation and evolutionary responses to climate change}{OK maybe this is the place for it.
+I would add a sentence about how endophytes can vary by chemotype and different chemotypes can confer different benefits.}, as well as  in changes in the wider microbial community.
 Answering these questions as well as the unknown questions that future researchers may ask also reiterates the value in capturing meta-information during ongoing digitization efforts at herbaria around the world and during the accession of newly collected specimens \citep{lendemer2020extended,edwards2024university}.
 Second, we accounted for several potential biases in the data observation process that may be common to many collections-based research questions by using a spatially-explicit random effects model. 
-Spatial autocorrelation \cite{willems2022forest}, potential biases introduced by the sampling habits of collectors \citep{daru2018widespread}, and variation between contemporary researchers during the collection of trait data, if not corrected for could lead to over-confident inference about the strength and direction of historic change.
-Previous studies that have quantified the effects of collector biases typically find them to be small  \cite{davis2015herbarium,meineke2019herbarium}, and we similarly did not find that collector has a strong effect on the results of our analysis.
+Spatial autocorrelation \cite{willems2022forest}, potential biases introduced by the sampling habits of collectors \citep{daru2018widespread}, and variation between contemporary researchers during the collection of trait data, if not corrected for could lead to over-confident inference about the strength and direction of historic change (I would cite the AGHY figure as an example).
+\tom{Previous studies that have quantified the effects of collector biases typically find them to be small  \cite{davis2015herbarium,meineke2019herbarium}, and we similarly did not find that collector has a strong effect on the results of our analysis.}{This seems to undercut the previous sentence, but its also a place where you could mention that scorer effects DID matter, so this is worth incorporating when possible.}
 %Fitting this model in a Bayesian framework allows for full propagation of uncertainty.
 
-Ultimately, a central goal of global change biology is to generate predictive insights into the future of natural systems. 
+Ultimately, a central goal of global change biology is to generate predictive insights into the future of natural systems on a rapidly changing planet. 
 While this survey of historic endophyte prevalence is necessarily correlative, it serves as a foundation to develop better predictive models of the response of microbial symbioses to climate change. 
-Combining the insights from this type of regional-scale survey with field experiments and physiological data could be invaluable. 
-While we found that climate is strongly correlated with endophytes' temporal responses, we do not know why trends in prevalence were weak in some areas or how endophytes would respond to more extreme changes in climate.
-For example, transplanting symbiotic and non-symbiotic plants beyond the range edge of \emph{A.
-hyemalis} could tell us whether persistent low endophyte prevalence in that area is a result of environmental conditions that lead the symbiosis to negative fitness consequences, or is a result of some historical contingency or dispersal limitation that has thus far limited the presence of symbiotic hosts from a region where they would otherwise flourish and provide resilience.
-While we observed evidence of mutualism resilience, more extreme environmental changes than those observed in our study could potentially push one or both partners beyond their physiological limit, leading to the collapse of the mutualism.
-Our analysis thus far is agnostic to changes in the distributions of hosts. 
-Mechanistic models could connect the responses of both host and symbionts to abiotic climate drivers integrating dispersal processes. 
-Beyond host-microbe symbioses, building these types of models would work towards quantitatively attributing biotic responses to anthropogenically driven climate change, similar to methods in climate science and economics \citep{stott2010detection, carleton2016social}.
+Combining the insights from this type of regional-scale survey with field experiments and physiological data \tom{could be invaluable. }{Vague.}
+While we found that certain dimensions of climate change were strongly correlated with endophytes' temporal responses, we do not know why trends in prevalence were weak in some areas or how endophytes would respond to more extreme changes in climate.
+\tom{For example, transplanting symbiotic and non-symbiotic plants beyond the range edge of \emph{A.
+hyemalis} could tell us whether persistent low endophyte prevalence in that area is a result of environmental conditions that lead the symbiosis to negative fitness consequences, or is a result of some historical contingency or dispersal limitation that has thus far limited the presence of symbiotic hosts from a region where they would otherwise flourish and provide resilience.}{This does not fit well here.
+I would move earlier where you talk about geographic trends.
+I also want to highlight that there is an important difference between associations of endophyte prevalence and geographic variation in climate (as in this AGHY example), and associations between CHANGES in endophyte prevalence and CHANGES in climate (the lines figure).
+Too often the manuscript blurs this disinction, and it happens here.}
+While the ``time machine'' of natural history collections revealed evidence of mutualism resilience for grass-endophyte symbioses in the face of environmental change, more extreme changes could potentially push one or both partners beyond their physiological limits, leading to the collapse of the mutualism; more research is needed to understand what those limits might be.
+\tom{Our analysis thus far is agnostic to changes in the distributions of hosts. 
+Mechanistic models could connect the responses of both host and symbionts to abiotic climate drivers, integrating dispersal processes. }{Does not feel like it fits well as a concluding point.}
+\tom{Beyond host-microbe symbioses, building these types of models would work towards quantitatively attributing biotic responses to anthropogenically driven climate change, similar to methods in climate science and economics \citep{stott2010detection, carleton2016social}.}{Not knowing these citations, I am not really sure what you are referring to.
+This does not feel like a natural point to end on.}
 %Elymus genotype
@@
 	\section*{Data and Code Availability}
-	Data from this publication will be made publically available upon acceptance and before that upon request.
-	Code for analyses can be found through: https://github.com/joshuacfowler/EndoHerbarium
+	\tom{Data from this publication will be made publically available upon acceptance and before that upon request.}{CHeck journal requirements.
+It is generally better to make data and code available during peer review.}
+	Code for analyses can be found through a publicly available repository (https://github.com/joshuacfowler/EndoHerbarium) that will be permanently archived upon publication.
 

### 2024-09-18 Manuscript/endo_herbarium.tex
adding caterpillar type plot for collector and scorer.

@@
 	\centering
-	\includegraphics[width = .8\linewidth]{../Plots/collector_posterior_plot.pdf}
-	\caption{\textbf{Posterior estimates of collector random effects.} Density curves show the posterior estimate along lines indicating the 95\% CI for 12 randomly selected collectors.}
+	\includegraphics[width = .6\linewidth]{../Plots/collector_posterior.png}
+	\caption{\textbf{Posterior estimates of collector random effects.} Points show posterior median along with 95\% CI for each of $924$ individual collectors.}
 		\label{fig:collector_fx}
@@
 	\centering
-	\includegraphics[width = .8\linewidth]{../Plots/scorer_posterior_plot.pdf}
-	\caption{\textbf{Posterior estimates of scorer random effects.} Density curves show the posterior estimate along lines indicating the 95\% CI for 25 scorers.}
+	\includegraphics[width = .6\linewidth]{../Plots/scorer_posterior.png}
+	\caption{\textbf{Posterior estimates of scorer random effects.} Points show posterior median along with 95\% CI for each of $25$ individual collectors.}
 	\label{fig:scorer_fx}

### 2024-09-18 Analyses/endo_herbarium_multispecies_spatiotemporal_analysis.R

@@ -721,6 +721,59 @@ plot(fit, "time.species1")
 
 
 
+n_draws <- 500
+
+posteriors.collector <- generate(
+  fit,
+  formula = ~ collector_latent,
+  n.samples = n_draws) 
+
+posteriors.scorer <- generate(
+  fit,
+  formula = ~ scorer_latent,
+  n.samples = n_draws) 
+
+
+colnames(posteriors.collector) <- c( paste0("iter",1:n_draws))
+colnames(posteriors.scorer) <- c( paste0("iter",1:n_draws))
+
+rownames(posteriors.collector) <- c(paste0("collector",1:nrow(posteriors.collector)))
+rownames(posteriors.scorer) <- c(paste0("scorer",1:nrow(posteriors.scorer)))
+
+
+posteriors <- rbind(posteriors.collector, posteriors.scorer)
+
+
+
+posteriors_df <- as_tibble(posteriors, rownames = "param") %>% 
+  mutate(param_type = case_when(grepl("collector", param) ~ "collector",
+                                grepl("scorer", param) ~ "scorer"),
+         number = parse_number(param)) %>% 
+  pivot_longer( cols = -c(param, param_type, number), names_to = "iteration")
+
+posteriors_summary <- posteriors_df %>% 
+  group_by(param, number, param_type) %>% 
+  dplyr::summarise(median = median(value),
+                   Q97.5 = quantile(value, .975),
+                   Q2.5 = quantile(value, .025))
+
+collector_plot <- ggplot(data = filter(posteriors_summary, param_type == "collector"))+
+  geom_vline(xintercept = 0)+
+  geom_point(aes(x = median, y = number), alpha = .8)+
+  geom_linerange(aes(xmin = Q2.5, xmax = Q97.5, y = number), alpha = .3)+
+  labs(y = "Collector Number", y = "Posterior Estimate") + theme_minimal()
+collector_plot
+ggsave(collector_plot, filename = "Plots/collector_posterior.png", width = 5, height = 10)
+
+scorer_plot <- ggplot(data = filter(posteriors_summary, param_type == "scorer"))+
+  geom_vline(xintercept = 0)+
+  geom_point(aes(x = median, y = number), alpha = .8)+
+  geom_linerange(aes(xmin = Q2.5, xmax = Q97.5, y = number), alpha = .3)+
+  labs(y = "Scorer Number", x = "Posterior Estimate") + theme_minimal()
+scorer_plot
[…]

### 2024-09-18 Analyses/endo_herbarium_multispecies_spatiotemporal_analysis.R
updating year plot to use the median year

@@ -70,8 +70,8 @@ endo_herb <- endo_herb_georef %>%
   filter(!is.na(lon) & !is.na(year)) %>% 
   filter(lon>-110 ) %>% 
   filter(Country != "Canada" ) %>% 
-  mutate(year_bin = case_when(year<1970 ~ "pre-1970",
-                              year>=1970 ~ "post-1970")) %>% 
+  mutate(year_bin = case_when(year<1969 ~ "pre-1969",
+                              year>=1969 ~ "post-1969")) %>% 
   mutate(scorer_index = parse_number(scorer_factor),
          herbarium_index = parse_number(herbarium_factor),
          collector_index = parse_number(collector_factor)) %>% 
@@ -157,7 +157,7 @@ endo_status_map <- ggplot()+
   geom_map(data = outline_map, map = outline_map, aes( map_id = region), color = "grey", linewidth = .1, fill = "#FAF9F6")+
   geom_map(data = states_shape, map = states_shape, aes(map_id = region), color = "grey", linewidth = .1, fill = NA)+
   geom_point(data = endo_herb, aes(x = lon, y = lat, color = endo_status_text), alpha = .7, size = .8)+
-  facet_wrap(~factor(year_bin, levels = c("pre-1970", "post-1970"))+species)+
+  facet_wrap(~factor(year_bin, levels = c("pre-1969", "post-1969"))+species)+
   coord_sf(xlim = c(-109,-68), ylim = c(21,49))+
   scale_color_manual(values = c(endophyte_colors[2],endophyte_colors[6]))+
   theme_light()+
@@ -165,7 +165,6 @@ endo_status_map <- ggplot()+
         strip.text  = element_text(face = "italic", color = "black"))+
   labs(x = "Longitude", y = "Latitude", color = "Endophyte Status")
 # endo_status_map
-
 # ggsave(endo_status_map, filename = "Plots/endo_status_map.png", width = 10, height = 5)
 
 

### 2024-09-18 Manuscript/endo_herbarium.tex
editing caption

@@
 		\includegraphics[width = \linewidth]{../Plots/endo_status_map.png}
-		\caption{\textbf{Endophyte presence/absence in specimens of each host species.} Points show collection locations colored according to whether the specimen contained endophytes ( E+; blue points) or did not contain endophytes (E-, tan points) and are faceted based on collection period.}
+		\caption{\textbf{Endophyte presence/absence in specimens of each host species.} Points show collection locations colored according to whether the specimen contained endophytes ( E+; blue points) or did not contain endophytes (E-, tan points).
+To visualize temporal change, the data are faceted before and after the median year of collection.}
 		\label{fig:endo_status_map}

### 2024-09-18 Analyses/posthoc_climate_correlation.R
exploring fitting the climate analysis with some interaction terms, and changing to plot regressions with the intercept

@@ -440,12 +440,12 @@ sample_n(size =250) %>%
 #   facet_wrap(~species)
 
 
-# svc.pred_climate_subsample_long <- svc.pred_climate_subsample %>% 
-#   pivot_longer(cols = contains("_diff")) %>% 
-#   filter(!grepl("cv", name)) 
-#   # filter(!grepl("tmean_spring", name)) %>% 
-#   # filter(!grepl("tmean_summer", name)) %>% 
-#   # filter(!grepl("tmean_autumn", name)) 
+svc.pred_climate_subsample_long <- svc.pred_climate_subsample %>%
+  pivot_longer(cols = contains("_diff")) %>%
+  filter(!grepl("cv", name))
+  # filter(!grepl("tmean_spring", name)) %>%
+  # filter(!grepl("tmean_summer", name)) %>%
+  # filter(!grepl("tmean_autumn", name))
 # 
 
 
@@ -520,22 +520,33 @@ data<- svc.pred_climate_subsample %>%
     easting = st_coordinates(.)[, 1],
     northing = st_coordinates(.)[, 2]
   ) 
-# setting up a spatial intercept
 fit_list <- list()
 
 for(s in 1:3){
   data_temp <- data[data$species == species_names[s],]
 
-  cmp <- ~ Intercept(main = Intercept, model = "linear", mean.linear=0, prec.linear=0.001) + 
-  spring_ppt(main = ppt_spring_diff, model = "linear", mean.linear=0, prec.linear=0.001) + summer_ppt(main = ppt_summer_diff, model = "linear", mean.linear=0, prec.linear=0.001) + autumn_ppt(main = ppt_autumn_diff, model = "linear", mean.linear=0, prec.linear=0.001) +
-  spring_ppt_sd(main = ppt_spring_sd_diff, model = "linear", mean.linear=0, prec.linear=0.001) + summer_ppt_sd(main = ppt_summer_sd_diff, model = "linear", mean.linear=0, prec.linear=0.001) + autumn_ppt_sd(main = ppt_autumn_sd_diff, model = "linear", mean.linear=0, prec.linear=0.001) +
-  spring_tmean(main = tmean_spring_diff, model = "linear", mean.linear=0, prec.linear=0.001) + summer_tmean(main = tmean_summer_diff, model = "linear", mean.linear=0, prec.linear=0.001) + autumn_tmean(main = tmean_autumn_diff, model = "linear", mean.linear=0, prec.linear=0.001) +
[…]

### 2024-09-18 Analyses/posthoc_climate_correlation.R
calculating posterior summaries of the climate predictors

@@ -442,10 +442,12 @@ sample_n(size =250) %>%
 
 svc.pred_climate_subsample_long <- svc.pred_climate_subsample %>%
   pivot_longer(cols = contains("_diff")) %>%
-  filter(!grepl("cv", name))
-  # filter(!grepl("tmean_spring", name)) %>%
-  # filter(!grepl("tmean_summer", name)) %>%
-  # filter(!grepl("tmean_autumn", name))
+  filter(!grepl("cv", name)) %>% 
+  filter(!grepl("annual", name)) %>% 
+  mutate(name = fct_relevel(name, c('tmean_spring_diff', 'tmean_summer_diff', "tmean_autumn_diff",
+                                    'tmean_spring_sd_diff', 'tmean_summer_sd_diff', "tmean_autumn_sd_diff",
+                                    'ppt_spring_diff', 'ppt_summer_diff', "ppt_autumn_diff",
+                                    'ppt_spring_sd_diff', 'ppt_summer_sd_diff', "ppt_autumn_sd_diff"))) 
 # 
 
 
@@ -579,6 +581,7 @@ fit_list[[1]]$summary.random
 
 
 prediction_list <- list()
+posterior_list <- list()
 for(s in 1:3){
 data_temp <- data[data$species == species_names[s],]
 
@@ -665,13 +668,37 @@ prediction <- predict(
   probs = c(0.025, 0.25, 0.5, 0.75, 0.975),
   n.samples = 100)
 prediction_list[[species_names[s]]] <- prediction
+
+
+posterior <- generate(
+  fit_list[[s]],
+  newdata = preddata,
+  formula =  ~ c("spring_ppt" =  spring_ppt_latent,
+                      "summer_ppt" =  summer_ppt_latent,
+                      "autumn_ppt" =  autumn_ppt_latent,
+                      # 
+                      "spring_ppt_sd" =  spring_ppt_sd_latent,
+                      "summer_ppt_sd" =  summer_ppt_sd_latent,
+                      "autumn_ppt_sd" =  autumn_ppt_sd_latent,
+                      # 
+                      "spring_tmean" =   spring_tmean_latent,
+                      "summer_tmean" =   summer_tmean_latent,
+                      "autumn_tmean" =   autumn_tmean_latent,
+                      # 
+                      "spring_tmean_sd" = spring_tmean_sd_latent,
[…]

### 2024-10-01 Analyses/endo_herbarium_mesh_and_prior_sensitivity_analysis.R
updating plot for svc maps

@@ -349,7 +349,8 @@ mesh_list <- list(mesh1, mesh2)
 fit_list <- list()
 time_list <- list()
 
-for(m in 1:length(mesh_list)){
+# for(m in 1:length(mesh_list)){
+  for(m in 2){
   mesh <- mesh_list[[m]]
 
 
@@ -381,7 +382,7 @@ spde3 <- inla.spde2.pcmatern(
 )
 
 spde_list <- list(spde1, spde2, spde3)
-
+}
 # inlabru makes making weights spatial effects simpler because we don't have to make projector matrices for each effect. i.e we don't have to make an A-matrix for each spatially varying effect.
 # this means we can go strat to making the components of the model
 
@@ -552,8 +553,12 @@ for(i in c(1,3,5,7,9,11)){
   sd_year <- sd(data$year)
   
 year.pred <- list()
-  
-for(i in 1:n){
+
+# Using these functions to look at just the fits from one set of iid random effects priors
+evens <- function(x) subset(x, x %% 2 = 0)
+odds <- function(x) subset(x, x %% 2 != 0)
+
+for(i in odds(1:n)){
   year.pred.aghy <- predict(
     fit_list[[i]],
     newdata = preddata[preddata$species_index == 1,],
@@ -575,7 +580,12 @@ for(i in 1:n){
   year.pred[[i]] <- bind_rows(year.pred.aghy, year.pred.agpe, year.pred.elvi) %>% mutate(model = names(fit_list)[i])
 } 
 
-year.pred.df <- list_rbind(year.pred)
+year.pred.df <- list_rbind(year.pred) %>% 
+  mutate(mesh = case_when(grepl("m1", model) ~ "standard mesh",
+                          grepl("m2", model) ~ "finer mesh"),
+         range = case_when(grepl("s1", model) ~ "prior range = 342 km",
+                           grepl("s2", model) ~ "prior range = 68 km",
+                           grepl("s3", model) ~ "prior range = 1714 km"))
   # binning the data for plotting
   endo_herb_binned <- endo_herb %>% 
     mutate(binned_year = cut(year, breaks = 10)) %>%
@@ -599,17 +609,18 @@ year_trend <- ggplot(year.pred.df) +
     geom_line(aes(year, mean)) +
     geom_ribbon(aes(year, ymin = q0.025, ymax = q0.975), alpha = 0.2) +
[…]

### 2024-10-03 Manuscript/endo_herbarium.tex
edits to the discussion

@@
 
-	\textit{Manuscript elements}: Figure \ref{fig:map} - Figure \ref{fig:contemptestplot}, appendix~A (including Figure~\ref{fig:endo_status_map} - Figure \ref{fig:ELVI_climate_covariates},  Table \ref{table:herbaria}, and Supplemental Methods).
+	\textit{Manuscript elements}: Figure \ref{fig:map} - Figure \ref{fig:contemptestplot}, Appendix~A (including Figure~\ref{fig:endo_status_map} - Figure \ref{fig:ELVI_climate_covariates},  Table \ref{table:herbaria}, and Supplemental Methods).
 
@@
 
-	\textit{Keywords}: .
+	\textit{Keywords}: climate change, plant-microbe symbiosis, herbarium, museum specimen, spatially-varying coefficients model, \emph{Epichloë}, \emph{Poaceae}.
 
@@
 	\section*{Abstract}
-\tom{Species' distributions and abundances are shifting in response to climate change. 
-Most species harbor microbial symbionts that have the potential to influence these responses.
-Mutualistic microbial symbionts may provide resilience to environmental change by protecting their hosts from increasing stress. 
-However, environmental change that disrupts these interactions may lead to declines in hosts and/or symbionts. 
-Symbionts preserved within herbarium specimens offer a unique opportunity to quantify changes in microbial symbiosis across broad temporal and spatial scales.
-}{If you need to cut words, I think some of this could be condensed.}
+Species' distributions and abundances are shifting in response to ongoing global climate change. 
+Mutualistic microbial symbionts can provide their hosts with protection from environmental stress. 
+This protection may thus contribute towards resilient responses to environmental change, however these changes may also disrupt species interactions and lead to declines in hosts and/or symbionts.
+Symbionts preserved within natual history specimens offer a unique opportunity to quantify changes in microbial symbiosis across broad temporal and spatial scales. 
 We asked how the prevalence of seed-transmitted fungal symbionts of grasses (\emph{Epichloë} endophytes), which can protect hosts from abiotic stress, have changed over time in response to climate change, and how these changes vary across host species' ranges.
-Specifically, we analyzed 2,346 herbarium specimens of three grass host species (\emph{Agrostis hyemalis},\emph{Agrostis perennans},\emph{Elymus virginicus}) collected over the past two centuries (1824 -- 2019) for the presence or absence of \emph{Epichloë} symbiosis\tom{, and evaluated spatial and temporal trends in endophyte prevalence}{Could cut, overlaps with previous sentence.}.
+Specifically, we analyzed 2,346 herbarium specimens of three grass host species (\emph{Agrostis hyemalis},\emph{Agrostis perennans},\emph{Elymus virginicus}) collected over the past two centuries (1824 -- 2019) for the presence or absence of \emph{Epichloë} symbiosis.
 We found that endophytes increased in prevalence over the last two centuries from ca. 25\% prevalence to ca. 75\% prevalence, on average, across three host species.
-We also found that changes in prevalence were associated with observed changes in seasonal climate drivers; notably, \tom{increasing precipitation corresponding to each host species' peak growing season and changes in off-peak season variability in precipitation}{I would rephrase this to better communicate the sign of the association, and thus the inferred causality.}.
+Changes in seasonal climate drivers were associated with increasing endophyte prevalence. 
+Notably, increasing precipitation during the peak growing season for \emph{Agrostis} species and decreasing precipitation for \emph{E. virginicus} were associated with increasing endophyte prevalence.
+We also identified changes in the variability of precipitation and temperature during off-peak seasons as important predictors of increasing endophyte prevalence. 
 Our analysis performed favorably in an out-of-sample predictive test with contemporary survey data, a rare extra step in collections-based research.
@@
 
-Abstract : 287 words
+Abstract : 315 words
 
@@
 Lichens exposed to elevated temperatures experienced loss of photosynthetic function along with changes in the composition of their algal symbiont community \citep{meyer2022climate}.
-How commonly and under what conditions microbial mutualisms deteriorate or strengthen under climate change remain unanswered questions.
+How commonly and under what conditions microbial mutualisms deteriorate or strengthen under climate change remain unanswered questions \citep{frederickson2017mutualisms}.
 Previous work suggests that these alternative responses may depend on the intimacy and specialization of the interaction as well as the physiological tolerances of the mutualist partners \citep{toby2010mutualisms, warren2014mutualism, rafferty2015phenological}.
@@
 Additionally, symbionts are often unevenly distributed across their hosts' distribution.
-\tom{Facultative symbionts may be absent from portions of the host range \cite{afkhami2014mutualist}, and hosts may engage with a diversity of partners (different symbiont species or locally-adapted strains) across their environments \cite{frade2008variation, rolshausen2018quantifying}.}{I would encourage self-citation here!}
+Facultative symbionts may be absent from portions of the host range \cite{afkhami2014mutualist}, and hosts may engage with a diversity of partners (different symbiont species or locally-adapted strains) across their environments \cite{frade2008variation, rolshausen2018quantifying, fowler2023geographic}.
 Identifying broader spatial trends in symbiont prevalence is therefore an important step in developing predictions for where to expect changes in the symbiosis in future climates.
@@
 In this study, we assessed the long-term responses of endophyte symbiosis to climate change through the use of herbarium specimens of three North American host grass species (\emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}).
-We first address questions describing spatial and temporal trends in endophyte prevalence: (i) How has endophyte prevalence changed over the past two centuries?
+We first addressed questions describing spatial and temporal trends in endophyte prevalence: (i) How has endophyte prevalence changed over the past two centuries?
 and (ii) How spatially variable are temporal trends in endophyte prevalence across eastern North America?
-We then address how climate change may be driving trends in endophyte prevalence by asking: (iii) What is the relationship between temporal trends in endophyte prevalence and associated changes in climate drivers?
+We then addressed how climate change may be driving trends in endophyte prevalence by asking: (iii) What is the relationship between temporal trends in endophyte prevalence and associated changes in climate drivers?
 We predicted that aggregate endophyte prevalence would increase over time in tandem with climate warming, and that hotspots of endophyte change would correspond spatially to hotspots of climate change.
 Finally, we evaluated the performance of models built on data from historic specimens with an out-of-sample test, using data on endophyte prevalence from contemporary surveys of host populations. 
-\tom{To answer these questions we examined a total of 2,346 specimens collected across eastern North America between 1824 and 2019.}{Since the preceding sentence is about the out of sample test, maybe include here the numbers of individuals and populations that the historical predictions were tested against?}
+To answer these questions we examined a total of 2,346 historic specimens collected across eastern North America between 1824 and 2019, and evaluated model performance against contemporary surveys comprising 1,442 individuals from 63 populations collected between 2013 and 2020.
 
@@
 Both \emph{Agrostis} species host \emph{Epichloë amarillans} \cite{craven2001multigene, leuchtmann2014nomenclatural}, while \emph{Elymus virginicus} typically hosts \emph{Epichloë elymi} \cite{clay2002evolutionary}.
-These $C_3$ grass species are commonly represented in natural history collections with broad distributions covering much the eastern United States (Fig. \ref{figmap}).
+These $C_3$ grass species are commonly represented in natural history collections with broad distributions covering much the eastern United States (Fig. \ref{fig:map}).
 \emph{A. hyemalis} is a small short-lived perennial species that germinates in spring and typically flowers between March and July (most common collection month: May).
@@
 We chose our focal species in part because they are commonly represented in herbarium collections, and produce high numbers of seeds, meaning that small samples would not diminish the value of the specimens for future studies.
-We collected 5-10 seeds per specimen after examining the herbarium sheet under a dissecting microscope to ensure that we collected mature seeds, not florets or unfilled seeds, fit for our purpose of identifying fungal endophytes with microscopy.
+We collected up to 5-10 seeds per specimen after examining the herbarium sheet under a dissecting microscope to ensure that we collected mature seeds, not florets or unfilled seeds, fit for our purpose of identifying fungal endophytes with microscopy.
 We excluded specimens for which information about the collection location and date were unavailable.
@@
 Collections were geo-referenced to the nearest county centroid, or nearest municipality when that information was available. 
-\tom{For a few of the oldest specimens, only information at the state level was available, and so we used the state centroid.}{Say how many?}
+For fifteen of the oldest specimens, only information at the state level was available, and so we used the state centroid.
 
@@
 We examined the squashed seeds for the presence of fungal hyphae at 200-400X magnification \cite{bacon2018stains}.
-\tom{In some cases, the tissues examined during microscopy came from flowers or otherwise non-viable seeds, which were excluded for that specimen.}{Could cut, since you already say you only took good seeds.}
-On average we scored $4.7$ seeds per specimen of \emph{A.
+%In some cases, the tissues examined during microscopy came from flowers or otherwise non-viable seeds, which were excluded for that specimen.
+On average we scored $4.7$ intact seeds per specimen of \emph{A.
 hyemalis}, $4.2$ seeds per specimen of \emph{A.
@@
 	\includegraphics[width = \linewidth]{../Plots/collections_map.png}
-	\caption{\textbf{Collection locations of herbarium specimens of three grass host species across eastern North America that were sampled for \emph{Epichloë} endophyte presence or absence}.}
+	\caption[Collection locations of herbarium specimens sampled for \emph{Epichloë} endophytes]{\textbf{Collection locations of herbarium specimens sampled for \emph{Epichloë} endophytes.} Specimens span eastern North America from nine herbaria, and are colored by host species.}
 	\label{fig:map}
@@
 Fitting models with structured spatial effects is possible with MCMC sampling but can require long computation times, making INLA an effective alternative. 
-\tom{This approach has been used to model spatial patterns in flowering phenology \cite{willems2022forest}, the abundance of birds \cite{meehan2019spatial} and butterflies \cite{crossley2022opposing}, the distribution of temperate trees \cite{engel2022spatial} as well as the population dynamics of endangered amphibians \cite{knapp2016large} and other ecological processes \cite{beguin2012hierarchical}.}{Probably not necessary if you need space.}
+This approach has been used to model spatial patterns in flowering phenology \cite{willems2022forest}, the abundance of birds \cite{meehan2019spatial} and butterflies \cite{crossley2022opposing}, the distribution of temperate trees \cite{engel2022spatial} as well as the population dynamics of endangered amphibians \cite{knapp2016large} and other ecological processes \cite{beguin2012hierarchical}.
+%\tom{}{Probably not necessary if you need space.}
 
 We estimated global and spatially-varying trends in endophyte prevalence using a joint-likelihood model. 
-For each host species $h$, endophyte presence/absence of the $i^{th}$ specimen \tom{($P_{[h]i}$)}{Not obvious to me why you use square brackets here.} was modeled as a Bernoulli response variable with expected probability of endophyte occurrence $\hat{P}_{[h]i}$.
-We modeled $\hat{P}_{[h]i}$ as a linear function of intercept \tom{$ \mathrm{A}_{[h]i}$}{I don't think there should be an $i$ subscript here.
-You need $i$ only where there is a covariate unique to the $i$th observation.} and slope $\mathrm{T}_{[h]}$ defining the global trend in endophyte prevalence specific to each host species as well as with spatially-varying intercepts $\alpha_{[h_{1}]l[i]}$ and slopes $\tau_{[h_{1}]l[i]}$ associated with location ($l[i]$, the unique latitude-longitude combination of the $i$th observation).
-The joint-model structure allowed us to ``borrow strength'' across species in the estimation of shared variance terms for the spatially-dependent random effect $\delta_{l[i]}$, intended to account for residual spatial variation, and $\chi_{c[i]}$ and $\omega_{s[i]}$ i.i.d.-random effects indexed for each collector identity ($c[i]$), and scorer identity ($s[i]$) of the $i^{th}$ specimen.\tom{}{I don't think you need three expressions here.
-If you just index host as $h$ it is clear that some pieces are unique to host $h$ and others are shared across hosts.}
+For each host species $h$, endophyte presence/absence of the $i^{th}$ specimen ($P_{h,i}$) was modeled as a Bernoulli response variable with expected probability of endophyte occurrence $\hat{P}_{h,i}$.
+We modeled $\hat{P}_{h,i}$ as a linear function of intercept $\boldsymbol{A}_h$ and slope $\boldsymbol{T}_{h}$ defining the global trend in endophyte prevalence specific to each host species as well as with spatially-varying intercepts $\alpha_{h,l_{i}} $ and slopes $\tau_{h,l_i}$ associated with location ($l_i$, the unique latitude-longitude combination of the $i$th observation).
+The joint-model structure allowed us to ``borrow strength'' across species in the estimation of shared variance terms for the spatially-dependent random effect $\delta_{l_i}$, intended to account for residual spatial variation, and $\chi_{c_i}$ and $\omega_{s_i}$ i.i.d.-random effects indexed for each collector identity ($c_i$), and scorer identity ($s_i$) of the $i^{th}$ specimen.
 
-\begin{subequations}
+\begin{equation}
 	\label{eq:trends}
-	\begin{align}
-		logit(\hat{P}_{[h_{1}]i}) =  \mathrm{A}_{[h_{1}]i} + \mathrm{T}_{[h_{1}]}*year_i  + \alpha_{[h_{1}]l[i]} + \tau_{[h_{1}]l[i]}*year_i  + \delta_{l[i]}+ \chi_{c[i]} + \omega_{s[i]} \\
-		logit(\hat{P}_{[h_{2}]i}) = \mathrm{A}_{[h_{2}]i} + \mathrm{T}_{[h_{2}]}*year_i  + \alpha_{[h_{2}]l[i]} + \tau_{[h_{2}]l[i]}*year_i  + \delta_{l[i]}+ \chi_{c[i]} + \omega_{s[i]} \\
-		logit(\hat{P}_{[h_{3}]i}) = \mathrm{A}_{[h_{2}]i} + \mathrm{T}_{[h_{2}]}*year_i  + \alpha_{[h_{2}]l[i]} + \tau_{[h_{3}]l[i]}*year_i  + \delta_{l[i]}+ \chi_{c[i]} + \omega_{s[i]}
-	\end{align}
-\end{subequations}
+		logit(\hat{P}_{h,i}) = \boldsymbol{A}_{h} + \boldsymbol{T}_{h}*year_i  + \alpha_{h,l_{i}} + \tau_{h,l_i}*year_i + \delta_{l_i} + \chi_{c_i} + \omega_{s_i} 
+\end{equation}
 
@@
 We performed model fitting using the inlabru R package \citep{bachl2019inlabru}.
-Global intercept and slope parameters $\mathrm{A}$, and $\mathrm{T}$, were given vague priors.
+Global intercept and slope parameters $\boldsymbol{A}$, and $\boldsymbol{T}$, were given vague priors.
 Scorer and collector random effects, $\chi$ and $\omega$, were given penalized complexity priors, with distributions approximating a Normal distribution with standard deviation of 5. 
@@
 Priors for the Mat\'{e}rn covariance function, termed ''range" and "variance", define how proximity effects decay with distance.
-\tom{Priors for results presented in the main text reflect a range of $342$ kilometers. 
-We found that model results were somewhat sensitive to the choices of priors for these parameters, so we tested a range of values (from 68 kilometers to 2160 kilometers) and meshes (Supplemental Material), finding that model results were qualitatively similar, i.e.
-the same direction of effects across space, but that the magnitude and uncertainty varied.
-}{I struggled here.
-I tried to make some edits but I don't think I understand the methods well enough to edit correctly.
-It is not clear how the distances you quote should be interepreted (locations withing 68 or 2160 km are more likely to have the same endo status just due to proximity?), or whether they should be interpreted at all.
-The text seems to say both the priors do not matter, but also that they do.
-I have not looked at what is in the supplement yet, but if you present results there based on different priors it should be easy to write this in a way that succinctly says ``priors mattered a little bit, and if you care you can look at the supplement for a prior sensitivity analysis.''}
+Results presented in the main text reflect a prior range of $342$ kilometers (i.e. a $50$\% probability of estimating a range less than $342$ kilometers). 
+We tested a range of values (from $68$ kilometers to $1714$ kilometers) and meshes (presented in the Supporting Methods), finding that while the magnitude and uncertainty of effects varied, model results were qualitatively similar, i.e.
+the same direction of effects across space.
 
 %In all cases, posterior modes were \tom{stable}{Assessed how?} and equal to zero, indicating successful numeric convergence.
+
+\subsection*{Modeling distributions of host species}
+Because the herbarium records did not encompass the entirety of these host species'ranges, we additionally modeled the geographic distribution of each host species to generate realistic maps on which we could project the predictions of the INLA model.
+We followed the ODMAP (overview, data, model, assessment, prediction) protocol \citep{crossley2022opposing} (see Supporting Methods). 
+In short, we used presence-only observations of each host species from Global Biodiversity Information Facility (GBIF) between 1990 to 2020.
+%To reduce the potential influence of sampling bias and spatial autocorrelation, \tom{we thinned the occurrences to the spatial scale of our selected climatic predictors}{I do not follow.}. 
+%We selected climate variables that aligned with our analysis of climatic influences on trends in endophyte prevalence described above.
+%\tom{We calculated the mean and standard deviation of seasonal temperature and precipitation across 1990 to 2020. }{FOr what locations? Is this from the whole raster of NA?}
+%\tom{Among this suite of variables, we chose to include, which were uncorrelated (Variance Inflation Factor $>$ 0.7) and allowed us to predict the occurrence probability of each host species in space and time.}{Do not follow and sounds circular.}
+We fit maximum entropy (MaxEnt) models using the maxent function in the R package dismo \citep{hijmans2017package} using the same set of seasonal climate predictors considered above calculated for the 1990-2020 climate normals: mean and standard deviation of spring, summer, and autumn temperature, and mean and standard deviation of spring, summer, and autumn cumulative precipitation.
+%\josh{MaxEnt is preferred because it has been shown to generate response curves with less unpredictable behavior when applied to new climates \citep{hijmans2006ability}.}{possibly could remove this sentence?}
+We generated 10,000 pseudo-absences as background points, and split the occurrence data into 75\% for model training and 25\% for model testing.
+The performance of models was evaluated with AUC \citep{jimenez2012insights}. 
+We found AUC values of 0.862, 0.838, 0.821 respectively  for \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus} indicating good model fit to data.
+To convert the continuous predicted probabilities into binary presence - absence maps on which we projected INLA predictions, we used the  training sensitivity (true positive rate) and specificity threshold (true negative rate) \citep{liu2005selecting}.
+%These binary maps serve as boundaries in presented maps of change in endophyte prevalence, and outline the set of pixels used in our analysis of climate correlates with trends in endophyte prevalence
+
+\subsection*{Assessing the role of climate drivers}
+We assessed how the magnitude of climate change may have driven changes in endophyte prevalence by assessing correlations between changes in climate and changes in endophyte prevalence predicted from our spatial model at evenly spaced pixels across the study area.
+We first downloaded monthly temperature and precipitation rasters from the PRISM climate group \citep{daly2013prism} covering the time period between 1895 and 2020 using the 'prism' R package \citep{Rprism2015}.
+Prism provides reconstructions of historic climate variables across the United States by spatially-interpolating weather station data \citep{diLuzio2008constructing}. 
+We calculated 30-year climate normals for seasonal mean temperature and cumulative precipitation for the recent (1990 to 2020) and historic (1895 to 1925) periods.
+We used three four-month seasons within the year (Spring: January, February, March, April; Summer: May; June, July, August; Autumn: September, October, November, December). 
+This division of seasons allowed us to quantify differences in climate associated with the two ``cool'' seasons, when we expected our focal species to be most biologically active (\emph{A.
+hyemalis} flowering phenology: spring; \emph{E.
+virginicus}: spring and summer; \emph{A.
+perennans}: autumn).
+In addition to mean climate conditions, environmental variability itself can influence population dynamics \cite{tuljapurkar_population_1982} and changes in variability are a key prediction of climate change models \cite{stocker2013technical, ipcc_2021}.
+Therefore, we calculated the standard deviation for each annual and seasonal climate driver across each 30-year period.
+We then took the difference between recent and historic periods for the mean and standard deviation for each climate driver (Figs. \ref{fig:AGHY_climate_covariates}-\ref{fig:ELVI_climate_covariates}).
+All together, we assessed twelve potential climate drivers: the mean and standard deviation of spring, summer, and autumn temperature, as well as the mean and standard deviation of spring, summer, and autumn cumulative precipitation.
+
+To evaluate whether areas that have experienced the greatest changes in endophyte prevalence (hotspots of endophyte change) are associated with high degrees of change in climate (hotspots of climate change), we modeled the fitted, spatially-varying slopes of endophyte change through time ($\tau_{[h]l}$ as a linear function of environmental covariates, with a Gaussian error distribution.
+Data from each host species was analyzed separately. 
+Fitting regressions to many pixels across the study region risks articially inflating confidence in our results due to large sample sizes, and so we performed this analysis using only a random subsample of 250 pixels across the study region; other sizes of subsample yielded similar results.
 
@@
 
-\subsection*{Assessing the role of climate drivers}
-We assessed how the magnitude of climate change may have driven changes in endophyte prevalence by assessing correlations between changes in climate and changes in endophyte prevalence predicted from our spatial model at evenly spaced pixels across the study area.
-We first downloaded monthly temperature and precipitation rasters from the PRISM climate group \citep{daly2013prism} covering the time period between 1895 and 2020 using the 'prism' R package \citep{Rprism2015}.
-Prism provides reconstructions of historic climate variables across the United States by spatially-interpolating weather station data \citep{diLuzio2008constructing}. 
-We calculated 30-year climate normals for seasonal mean temperature and cumulative precipitation for the recent (1990 to 2020) and historic (1895 to 1925) periods.
-We used three four-month seasons within the year (Spring: January, February, March, April; Summer: May; June, July, August; Autumn: September, October, November, December). 
-This division of seasons allowed us to quantify differences in climate associated with the two ``cool'' seasons, when we expected our focal species to be most biologically active (\emph{A.
-hyemalis} flowering phenology: spring; \emph{E.
-virginicus}: spring and summer; \emph{A.
-perennans}: autumn).
-In addition to mean climate conditions, environmental variability itself can influence population dynamics \cite{tuljapurkar_population_1982} and changes in variability are a key prediction of climate change models \cite{stocker2013technical, ipcc_2021}.
-Therefore, we calculated the standard deviation for each annual and seasonal climate driver across each 30-year period.
-We then took the difference between recent and historic periods for the mean and standard deviation for each climate driver (Figs. \ref{fig:AGHY_climate_covariates}-\ref{fig:ELVI_climate_covariates}).
-All together, we assessed twelve potential climate drivers: the mean and standard deviation of spring, summer, and autumn temperature, as well as the mean and standard deviation of spring, summer, and autumn cumulative precipitation.
-
-\tom{To evaluate whether areas that have experienced the greatest changes in endophyte prevalence (hotspots of endophyte change) are associated with high degrees of change in climate (hotspots of climate change), we modeled the fitted, spatially-varying slopes of endophyte change through time ($\tau_{[h]l}$ as a linear function of environmental covariates, with a Gaussian error distribution.}{If I am reading the results figure correctly it looks like you force an intercept of zero, and I am not sure why you would want to do that.}
-Data from each host species was analyzed separately. 
-Fitting regressions to many pixels across the study region risks articially inflating confidence in our results due to large sample sizes, and so we performed this analysis using only a random subsample of 250 pixels across the study region; other sizes of subsample yielded similar results.
-
-\subsection*{Modeling distributions of host species}
-Because the herbarium records did not encompass the entirety of these host species'ranges, we additionally modeled the geographic distribution of each host species to generate realistic maps on which we could project the predictions of the INLA model.
-We followed the ODMAP (overview, data, model, assessment, prediction) protocol \citep{crossley2022opposing} (see Supporting Methods). 
-In short, we used presence-only observations of each host species from Global Biodiversity Information Facility (GBIF) \tom{between 1990 to 2020}{I am not sure why 1990-2020.
-From our conversations, I thought the SDM was based on all records.
-At this point I would not change anything.}.
-%To reduce the potential influence of sampling bias and spatial autocorrelation, \tom{we thinned the occurrences to the spatial scale of our selected climatic predictors}{I do not follow.}. 
-%We selected climate variables that aligned with our analysis of climatic influences on trends in endophyte prevalence described above.
-%\tom{We calculated the mean and standard deviation of seasonal temperature and precipitation across 1990 to 2020. }{FOr what locations? Is this from the whole raster of NA?}
-%\tom{Among this suite of variables, we chose to include, which were uncorrelated (Variance Inflation Factor $>$ 0.7) and allowed us to predict the occurrence probability of each host species in space and time.}{Do not follow and sounds circular.}
-We fit maximum entropy (MaxEnt) models using the maxent function in the R package dismo \citep{hijmans2017package} using the same set of seasonal climate predictors considered above: \tom{mean and standard deviation of spring, summer, and autumn temperature, and mean and standard deviation of spring, summer, and autumn cumulative precipitation.}{Were these 1990-2020 normals?}
-%\josh{MaxEnt is preferred because it has been shown to generate response curves with less unpredictable behavior when applied to new climates \citep{hijmans2006ability}.}{possibly could remove this sentence?}
-We generated 10,000 pseudo-absences as background points, and split the occurrence data into 75\% for model training and 25\% for model testing.
-The performance of models was evaluated with AUC \citep{jimenez2012insights}. 
-\tom{We found AUC values of 0.862, 0.838, 0.821 respectively  for \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}.}{Can we make this sentence more evaluative?
-I think AUC > 0.8 is very good, yes?}
-To convert the continuous predicted probabilities into binary presence - absence maps on which we projected INLA predictions, we used the  training sensitivity (true positive rate) and specificity threshold (true negative rate) \citep{liu2005selecting}.
-%These binary maps serve as boundaries in presented maps of change in endophyte prevalence, and outline the set of pixels used in our analysis of climate correlates with trends in endophyte prevalence
-
 \section*{Results}
@@
 \ref{fig:temporal_posterior}).
-The model appears to under-predict the observed increase in endophyte prevalence relative to the data, particularly for \emph{A.
-hyemalis} (Fig.
-\ref{fig:temporal}A), but the model is accounting for random effects and spatial non-independence that are not readily seen in the figure.
-By under-weighting endophyte-positive samples that are clustered spatially or by collector or observer, the INLA model is appropriately accounting for nuisance variables and providing a conservative inference of endophyte change relative to the raw data.
 
@@
 	\includegraphics[width = \linewidth]{../Plots/year_plot.png}
-	\caption[Temporal trends in endophyte prevalence.]{Temporal trends in endophyte prevalence.
-(A) Histograms show the frequency of scored specimens through time for each host species.
+	\caption[Temporal trends in endophyte prevalence.]{\textbf{Temporal trends in endophyte prevalence.} (A) Histograms show the frequency of scored specimens through time for each host species.
 (B) Lines show predicted mean endophyte prevalence over the study period along with the 50\%  and 95\% CI bands incorporating uncertainty associated with collector and scorer random effects.
@@
 \end{figure}
+
+The model appears to under-predict the observed increase in endophyte prevalence relative to the data, particularly for \emph{A.
+hyemalis} (Fig.
+\ref{fig:temporal}B), but the model is accounting for random effects and spatial non-independence that are not readily seen in the figure.
+We found no evidence that collector biases influenced our results. 
+Collector random effects were consistently small (Fig. \ref{fig:collector_fx}), and models fit with and without this random effect provide qualitatively similar results.
+The identity of individual scorers did contribute to observed patterns in endophyte prevalence.
+For example, $3$ of the $25$ scorers were more consistently likely than average to assign positive endophyte status, as indicated by 95\% credible intervals greater than zero (Fig.
+\ref{fig:scorer_fx}).
+It is difficult to distinguish whether this was driven by true differences in scorers biases during the seed scoring process or by unintended spatial or temporal clustering of the specimens scored by each scorer \cite{clayton1993spatial,urdangarin2023evaluating}.
+By under-weighting endophyte-positive samples that are clustered spatially or by collector or observer, the INLA model is appropriately accounting for nuisance variables and providing a conservative inference of endophyte change relative to the raw data.
 
@@
 Both \emph{Agrostis} species show areas of strong increase and areas of declining prevalence, while \emph{E.
-virginicus} had an overall weaker and geographically more consistent increase in endophyte prevalence.
+virginicus} had an overall weaker and geographically more homogeneous increase in endophyte prevalence.
 Notably, endophytes increased most strongly towards the western range edge of \emph{A.
@@
 \subsection*{What is the relationship between variation in temporal trends in endophyte prevalence and changes in climate drivers?}
-\tom{}{I think this section would be strengthed by greater statistical support.
-It is really just eye-balling lines.
-Can we say something more quantitative?
-See my comment about posteriors of slopes -- that is the type of information I have in mind, and maybe we could embed some of this in the text.}
 We found that trends in endophyte prevalence were strongly associated with seasonal climate change drivers (Fig. \ref{fig:climate_correlation_plot}).
-For the majority of the study region, \tom{the climate has become wetter and cooler over the last century}{Can you provide the average change in degrees celsius and mm precip across the range?} (Fig.
-\ref{fig:AGHY_climate_covariates}-\ref{fig:ELVI_climate_covariates}), a consequence of regional variation in global climate change \cite{ipcc_2021}.\tom{}{Comment on these supplement figures: perhaps for the revision, I think a color scale that diverges at zero would be helpful here.}
-Spatial variation in climate trends were predictive of trends in endophyte prevalence.
-\tom{For example, strong increases in endophyte prevalence for \emph{A.
-perennans} were most strongly associated with change in autumn climate drivers that coincide with its active growing season.
-}{Here and throughout the results, please reference specific figures and panels for statements like this so readers can see what you see.
-Regarding the figure, since it is just a bunch of regression lines and we only care about the slopes, it might be more effective to just show posteriors of the slopes - probably not worth making that change at this point, and maybe it's ugly.}
-For this species, regions experiencing warmer and wetter autumn climates showed particularly strong relationships with increasing endophyte prevalence, however other seasonal drivers may also contribute (drier springs and cooler summers).
-Trends in endophyte prevalence for \emph{A. hyemalis} were most strongly associated with changes in precipitation and precipitation variability. 
-Endophyte prevalence in \emph{A.
-hyemalis} increased the most in regions that experienced greater increases in spring precipitation, as well as regions with changing variability in spring and autumn precipitation.
-%While this species actively grows and reproduces in the late spring and early summer, climate effects outside of the growing season may indicate that endophytes play a role in persistence during dormant periods through summer droughts or contribute to the ability to successfully germinate.
-\tom{Changes in endophyte prevalence in \emph{E.
-virginicus} were not strongly associated with changes in most climate drivers}{Notice that my edits here importantly change the meaning of the sentence and, as far as I understand, make it more correct.}, but regions of reduced variability in autumn precipitation were associated with weak reductions in prevalence.
-\tom{Correlations}{I am not sure why you describe these as ``correlations''.
-The methods makes these sound like linear regressions.} assessed using all pixels across each species' distribution were qualitatively similar to these results (Fig.
-A11).
-%These associations are in line with expectations that endophyte symbiosis provides drought tolerance to their hosts, particularly during \emph{E.
-virginicus}'s \tom{summer growing season}{It is a stretch to say that this species has a summer growing season.}.
-%For \emph{A.
-perennans}, which flowers in the late Summer and Autumn, increases in autumn precipitation along with increases in variability in annual precipitation, were strongly assocated with increasing endophyte prevalence.
-%While this result is contrary to the expectation that endophyte provide drought tolerance, this species typically is associated with wetter microhabitats, and it also \tom{suggests a role for endophytes to play in buffering the negative consequences of demographic variability \citep{lewontin_population_1969}}{I would save for discussion}.
-%For \emph{A. hyemalis}, which flowers in the Spring, changes in endophyte prevalence were more weakly associated with seasonal changes in climate drivers than for the other two species.
-%It was, however, the only species' for which greater increases in endophyte prevalence were associated with regions that experienced increases in average annual temperatures.\tom{}{Is this a significant association, however you want to define that?
-Or could this result have occurred by chance?}
+For the majority of the study region, the climate has become wetter (an average increase in annual precipitation of $60$ mm.) with relatively little temperature warming (an average increase in annual temperature of $0.02$ $^{\circ}$C) over the last century (Fig.
+\ref{fig:AGHY_climate_covariates}-\ref{fig:ELVI_climate_covariates}), a consequence of regional variation in global climate change \cite{ipcc_2021}.
+Within the region, climate changes were spatially variable; certain locations experienced increases in annual precipitation as large as $375$ mm.
+or decreases up to $54$ mm.
+across the last century, while annual temperature changes ranged from warming as great as $1.4$ $^{\circ}$C to cooling by $0.46$ $^{\circ}$C.
+Spatially variable climate trends were predictive of trends in endophyte prevalence.
+For example, strong increases in endophyte prevalence for \emph{A.
+perennans} were most strongly associated with increasing autumn precipitation and with increasing mean and variability in autumn temperature (greater than 97\% posterior probabilities of positive slopes).
+For this species, a $1$ $^{\circ}$C increase in autumn temperature was associated with a $1.07$ \% increase per year in endophyte prevalence (Fig.
+\ref{fig:climate_correlation_plot}A) and a $100$ mm.
+increase in precipitation was associated with a $0.8$\% increase per year in endophyte prevalence (Fig.
+\ref{fig:climate_correlation_plot}B).
+This result aligns with the species' autumn active growing season, however other seasonal climate drivers were also associated with increasing endophyte prevalence. 
+In particular, we found cooler and drier springs and cooler summers to be associated with increasing endophyte prevalence (greater than $99$\% posterior probabilities of negative slopes) however these slopes were generally of smaller magnitude than those for autumn climate drivers.
+
+Changes in endophyte prevalence across the ranges of \emph{A. hyemalis} and \emph{E. virginicus} were less strongly driven by changes in climate. 
+Like \emph{A. perennans}, climate during peak growing season (spring for \emph{A. perennans} and summer for \emph{E. virginicus}) emerged most commonly as drivers of changes in endophyte prevalence.
+Increases in mean spring preciptation were the strongest predictor of increasing trends in endophyte prevalence for \emph{A.
+hyemalis} (Fig.
+\ref{fig:climate_correlation_plot}B) (greater than $99$\% posterior probability of a positive slope).
+For this species, an increase of $100$ mm. in spring precipitation led to an increase of $0.6$\% per year in endophyte prevalence.
+The next greatest slopes were those associated with variability in spring precipitation (greater than $96$\% posterior probability of a negative slope), as well as in the mean and variability of autumn climate (greater than $98$\% probability of negative and positive slopes, respectively).
+Changes in endophyte prevalence in \emph{E.
+virginicus} were not strongly associated with changes in most climate drivers, but regions with reduced variability in autumn precipitation (Fig.
+\ref{fig:climate_correlation_plot}B) and with cooler and more variable summer temperatures (Fig.
+\ref{fig:climate_correlation_plot}A,C) experienced the largest increases in endophyte prevalence.
+While our analysis identified the importance of these drivers with relatively high certainty (greater than $99$\% posterior probability of either negative or positive slopes respectively), they translate to less than $0.2$\% change in endophyte prevalence per year for a change of $100$ mm.
+change in precipitation over the century.
+Repeating this analysis using all pixels across each species' distribution were qualitatively similar to these results.
 
@@
 	\centering
-	\includegraphics[width = \linewidth]{../Plots/climate_trends_plot.png}
-	\caption{\textbf{Relationships between changes in seasonal climate drivers and predicted marginal trends in endophyte prevalence.} Lines show relationships between seasonal climate (A: mean temperature, B: cumulative precipitation, C: standard deviation in temperature, D: standard deviation in precipitation) and spatially-varying trends in endophyte prevalence for each host species, along with 50 and 95\% CI.}
+	\includegraphics[width = \linewidth]{../Plots/climate_trends_plot_intercept.png}
+	\caption[Relationships between predicted trends in endophyte prevalence and changes in seasonal climate drivers.]{\textbf{Relationships between predicted trends in endophyte prevalence and changes in seasonal climate drivers.} Lines show marginal predicted relationship between spatially-varying trends in endophyte prevalence and changes in mean and variability of climate ((A): mean temperature, (B): cumulative precipitation, (C): standard deviation in temperature, (D): standard deviation in precipitation).
+Confidence bands represent the 50 and 95\% CI,  colored by host species.
+Slopes with greater than 95\% probability of being either positive or negative are represented as solid lines while those that have less than 95\% probability are dashed.
+Points show 250 randomly sampled pixels across each host's distribution used in model fitting.}
 	\label{fig:climate_correlation_plot}
@@
 \subsubsection*{Performance on test data}
-\tom{}{The order of the results does not match the order of the methods.}
-We found that model performance, as quantified by AUC, was similar between historic herbarium specimens used as training data and the out-of-sample test data from contemporary surveys (0.79 and 0.77 respectively; Fig.
+Tests of model's predictive performance as quantified by AUC and by visual posterior predictive checks, indicated good predictive ability. 
+Model performance was similar between historic herbarium specimens used as training data and the out-of-sample test data from contemporary surveys (AUC = 0.79 and 0.77 respectively; Fig.
 \ref{fig:ROCtest}-\ref{fig:ROCtraining}).
@@
 \ref{fig:contemptestplot}B).
-However, model predictions for endophyte prevalence exhibited relatively little geographic variation, whereas the out-of-sample survey data were maximally variable with populations spanning 0\% to 100\% endophyte-symbiotic plants (Fig.
-\ref{fig:contemptestplot}C)\tom{}{Love the figure but I think the legend can be improved.
-It does not describe the content very well.}.
+However, model predictions for endophyte prevalence exhibited relatively little local geographic variation, whereas the out-of-sample survey data were maximally variable with populations spanning 0\% to 100\% endophyte-symbiotic plants (Fig.
+\ref{fig:contemptestplot}C).
 We interpret this to mean that the model captures coarse-scale spatial and temporal trends reasonably well, but is not equipped to capture local-scale nuances that generate population-to-population differences.
@@
 	\includegraphics[width = \linewidth]{../Plots/contemp_test_plot.png}
-	\caption{\textbf{Predicted vs observed endophyte prevalence for contemporary test data.} (A) The model, trained on historic herbarium collection data, performed modestly at predicting contemporary endophyte prevalence in \emph{A.
-hyemalis}, as indicated by some overlap of predicted 95\% CI with the 1:1 line, however contemporary test data generally had more variance between populations than model predictions.
-The model did recapitulate broader regional trends across (B) longitude and (C) latitude.
-Point size in panels B and C reflect sample sizes of contemporary endophyte population surveys.}
+	\caption[Predictive performance for contemporary test data.]{\textbf{Predictive performance for contemporary test data.} (A)
+		 The model, trained on historic herbarium collection data, performed modestly at predicting prevalence in contemporary population surveys.
+The model captured regional trends across (A) longitude and (B) latitude.
+Crosses indicate predicted mean prevalence along with the 95\% CI (colored lines) from the herbarium model.
+Contemporary prevalence is represented by grey points (point size reflects sample size) along with trend lines from generalized linear models (black line and shaded 95\% confidence interval).
+(C) Comparison of observed vs.
+predicted endophyte prevalence shows that contemporary test data had more variance between populations than contemporary predictions.}
 	\label{fig:contemptestplot}
 \end{figure}
-
-\subsection*{Assessing collector and scorer influences on predicted endophyte prevalence}
-Our modeling effort quantified temporal and spatial trends in endophyte prevalence while accounting for potential biases introduced by collectors and by individual scorers who quantified endophyte presence/absence within specimens.
-We found no evidence that collector biases influenced our results. 
-Collector random effects were consistently small (Fig. \ref{fig:collector_fx}), and models fit with and without this random effect provide qualitatively similar results.
-The identity of individual scorers did contribute to observed patterns in endophyte prevalence.
-For example, $3$ of the $25$ scorers were more consistently likely than average to assign positive endophyte status, as indicated by 95\% credible intervals greater than zero) (Fig.
-\ref{fig:scorer_fx}).
-It is difficult to distinguish whether this was driven by true differences in scorers biases during the seed scoring process or by unintended spatial or temporal clustering of the specimens scored by each scorer \cite{clayton1993spatial,urdangarin2023evaluating}.
-Interpreting our models with the inclusion of variance associated with the scorer effect thus provides conservative estimates of the absolute magnitude of changes in endophyte prevalence.
 
@@
 Differences across host species underscore that while all of these $C_3$ grasses share similar broad-scale distributions, each engages in unique biotic interactions and has unique responses to environmental drivers.
-We identified hotspots of change for \emph{A.
-perennans}, which was the species that experienced the \tom{strongest absolute changes}{remind readers what the absolute change was} in endophyte prevalence (Fig.
-\ref{fig:svc_time_map}).
-Declines in the southern portion of its range and increases in the north suggest a potential poleward range shift of endophyte-symbiotic plants (whether the oervall host distribution is shifting in parallel is an exciting next question).
-Based on previous work demonstrating that endophytes can shield their hosts from drought stress \tom{\cite{decunta2021systematic}}{I would cite this as ``reviewed in'' or something like that, so this is not read as a single case study}, we generally predicted that drought conditions would be a driver of increasing endophyte prevalence.
-In contrast to this expectation, increasing prevalence \tom{for this species}{I have now forgotten which species you are talk about, since the previous sentence is about grass-endophyte interactions more generally.
-I think this paragraph can be edited to make it clearer which are general statements and which refer to specific species (and which ones).} was associated with increasing autumn temperature and precipitation (Fig.
-\ref{fig:climate_correlation_plot}).
+We identified hotspots of change for \emph{A. perennans}, which was the species that experienced the strongest absolute changes in endophyte prevalence (Fig. \ref{fig:svc_time_map}).
+Declines of $0.9$\% per year in the southern portion of its range and increases of up to $2$\% per year in the north suggest a potential poleward range shift of endophyte-symbiotic plants (whether the overall host distribution is shifting in parallel is an exciting next question).
+Based on previous work demonstrating that endophytes can shield their hosts from drought stress (reviewed in\citet{decunta2021systematic}), we generally predicted that drought conditions would be a driver of increasing endophyte prevalence.
+In contrast to this expectation, increasing prevalence for \emph{A. perennans} were associated with increasing autumn temperature and precipitation (Fig. \ref{fig:climate_correlation_plot}). 
 To our knowledge, the response of the symbiosis in \emph{A.
@@
 Our results also hint that it may be useful to investigate whether lagged climate effects are important predictors of host fitness in this system \cite{evers2021lagged}.
-\tom{Endophyte prevalence of the spring-flowering \emph{A.
-hyemalis} was most strongly linked to increasing variability in precipitation across summer and autumn.
-}{I do not see this in the figure.
-I would argue you are over-interpreting the SD effects.
-The ppt amount effects look stronger, and these are in autumn and spring which lines up with this species' growing season.
-Marion's paper showed that wet autumn conditions can neutralize the benefits of the symbiosis, which I think is consistent with your result.}
-Endophytes could be playing a role helping hosts weather autumn-season droughts \tom{while the species is dormant.}{Not dormant in autumn. They probably start germinating in October.}
-Previous work has demonstrating drought benefits in a greenhouse manipulation with this species \citep{davitt2011understanding}, and early life stages may be particularly vulnerable to prolonged droughts.
+Endophyte prevalence of the autumn-flowering \emph{A.
+perennans} was strongly linked with decreasing spring precipitation, and that of the spring-flowering \emph{A.
+hyemalis} was associated with decreasing autumn precipitation (Fig.
+\ref{fig:climate_correlation_plot}B).
+For \emph{A. hyemalis}, endophytes could be playing a role helping hosts weather autumn-season droughts, which may be an important time for the species' germination.
+Previous work has demonstrated drought benefits in a greenhouse manipulation with this species \citep{davitt2011understanding}, and early life stages may be particularly vulnerable to prolonged droughts.
 %It is possible that the study region has not experienced a magnitude of climate change great enough to cause larger changes in endophyte prevalence for this species. 
 %Weak associations with drought could also be explained by climate-driven changes in the rate of imperfect transmission (the generation of non-symbiotic offspring from symbiotic hosts), which could counterbalance endophyte-mediated fitness benefits \citep{donald2021context}.
-For \emph{E. virginicus}, which experienced the most modest changes in endophte prevalence overall, we only modest associations with changes in climate drivers. 
+For \emph{E.
+virginicus}, which experienced the most modest changes in endophte prevalence overall (ranging between $1.1$\% increases and $0.2$\% decreases), we only found modest associations with changes in climate drivers.
 Surveys by \citet{sneck2017variation}, used as part of the test data in this study, identified a drought index (SPEI) that integrates precipitation with estimated evapotranspiration as an important predictor of endophyte prevalence.
-\emph{Epichloë} endophytes have also been connected to a suite of non-drought related fitness benefits including herbivore protection \citep{brem2001epichloe}, salinity resistence \citep{wang2020effects}, and mediation of the soil microbiome \citep{roberts2015rhizosphere}
+\emph{Epichloë} endophytes have also been connected to a suite of non-drought related fitness benefits including herbivore protection \citep{brem2001epichloe}, salinity resistence \citep{wang2020effects}, and mediation of the soil microbiome \citep{roberts2015rhizosphere}.
 These effects are potentially mediated by the diverse bioactive alkaloids and other signaling compounds they produce \citep{saikkonen2013chemical}.
 Increases in symbionts could be explained, at least in part, by these diverse benefits that may help hosts weather a world made increasingly stressful by changes in climate and other anthropogenically introduced stressors.
-\tom{While we show consistent increasing trends in prevalence between the three species, the mechanisms that explain these changes may be diverse and idiosyncratic.}{This paragraph was a little hard to read and I think does not really nail the points that I think we should make here, IMO.
+While we show consistent increasing trends in prevalence between the three species, the mechanisms that explain these changes may be diverse and idiosyncratic.
+%This paragraph was a little hard to read and I think does not really nail the points that I think we should make here, IMO.
 I would open to paragraph setting up the idea that the correlations you detected have some support in previous experimental studies, then elaborate.
 Then I would acknowledge that other factors can elicit benefits of symbiosis, and these factors may ALSO be changing.
-The takeaway, for me and perhaps for readers more generally, is that we are very confident endophytes are increasing, we have some confidence that is associated with (and perhaps due to) climate change, but a lot has changed over the past 200 years (climate, urbanization, nitrogen deposition, invasive species, deer!) and we need follow-up studies and experiments to gain a better mechanistic understanding of what has driven these changes, which will also allow us to build better forecasts about the future.}
+The takeaway, for me and perhaps for readers more generally, is that we are very confident endophytes are increasing, we have some confidence that is associated with (and perhaps due to) climate change, but a lot has changed over the past 200 years (climate, urbanization, nitrogen deposition, invasive species, deer!) and we need follow-up studies and experiments to gain a better mechanistic understanding of what has driven these changes, which will also allow us to build better forecasts about the future.
 
-\tom{Our spatially-explicit model predicted regions of both high and low endophyte prevalence, suggesting that symbiotic and non-symbiotic host plants have overlapping, but non-identical niche requirements.}{I think this paragraph tries to do too much relative to this topic sentence.
-In general I think you could use stronger topics sentences to give these Discussion paragraphs greater focus.}
-Endophytes fitness benefits potentially explain the spatial distribution of prevalence by allowing their hosts to persist in environments where they otherwise could not \tom{\citep{afkhami2014mutualist, kazenel2015mutualistic}}{cite Eco Monograph}.
-\tom{For example, fitness benefits of the symbiosis could explain historically low prevalence in \emph{A.
-hyemalis} towards its western range edge coinciding with a strong aridity gradient.}{But above you say that AGHY gets benefits under drought.}
-Previous population surveys for endophytes found similar regional trends in prevalence for endophyte host species \tom{\citep{sneck2017variation,rudgers2009benefits}}{I would add the semmartin study.}.
+The combination of a spatially-explicit model and historic herbarium specimens allowed us to identify regions of both increasing and decreasing endophyte prevalence, however we see several next steps towards the goal of predicting host and symbiont niche-shifts in response to future climate change.
 While the model recreated the large-scale spatial trends observed in contemporary population surveys, test data contained more population-to-population variability in prevalence. 
-Validating our model predictions in this way, a rare extra step in collections-based studies, allows us to evaluate places to improve the model's out-of-sample predictive ability, which will be particularly important for predicting host and symbiont niche-shifts under future climate change.
+Validating our model predictions in this way, a rare extra step in collections-based studies, allows us to evaluate places to improve the model's out-of-sample predictive ability.
 Lack of information on local variability may simply be a feature of data derived from herbarium specimens. 
 They are samples from local populations, but they are single specimens that are aggregated to derive broad-scale model estimates.
+This suggests that increasing local replication should be a factor considered in future collection efforts of natural history specimens, balanced with the required time and effort.
 Poor predictive ability at local scales in this grass-endophyte system is not surprising, as previous studies have found that local variation, even to the scale of hundreds of meters can structure endophyte-host niches \cite{kazenel2015mutualistic}.
 Other studies have found factors including land-use history \cite{vikuk2019infection} and the biotic environment, including herbivory \cite{rudgers2016long}, and host genotype \citet{sneck2017variation}, to be important predictors of endophyte ecology.
-\tom{Incorporating available climatic and soil layers as covariates is an obvious first step that could improve predictions.}{This is not where my mind went.
-I was thinking more local replication/redundancy, e.g.
-multiple specimens from the same population.
-Your idea is getting at explaining the local variation, but right now the model is not even seeing the local variation because we (rightly I think) prioritized larger-scale geographic coverage.}
-Another important step would be integrating data from local and regional scales  through modeling to constrain estimates of local and regional variation.
+An important step would be integrating data from local and regional scales  through modeling to constrain estimates of local and regional variation.
+Previous population surveys have found environment-dependent gradients in endophyte prevalence \citep{sneck2017variation,rudgers2009benefits,semmartin2015broad}, that may be caused by symbiont-derived fitness benefits allowing their hosts to persist in environments where they otherwise could not \citep{afkhami2014mutualist,kazenel2015mutualistic,fowler2023geographic}.
+Predicting future niche-shifts of hosts and symbionts will require considering the coupled dynamics of host-symbiont dispersal in addition to fitness benefits.
+For example, transplanting symbiotic and non-symbiotic plants beyond the range edge of \emph{A.
+hyemalis} could tell us whether low endophyte prevalence in that area is a result of environmental conditions that lead the symbiosis to negative fitness consequences, or is a result of some historical contingency or dispersal limitation that has thus far limited the presence of symbiotic hosts from a region where they would otherwise flourish and provide resilience.
+Incorporating available climatic and soil layers as covariates is another obvious step that could improve predictions.
 These steps will bridge gaps that often exist between large but broad bioclimatic and biodiversity data and small but local data on biotic interactions, and move towards the goal of predicting the dynamics of microbial symbioses under climate change \cite{miller2019recent, isaac2020data}.
@@
 The responses of microbial symbioses are a rich target for future studies within museum specimens, particularly those that take advantage of advances in sequencing technology.
-\tom{}{This is not the place for it, but this sentence made me think about genetic variation in the endophytes.
-We see presence-absence changing but the endophytes themselves could also be changing (``chemotypes'') which is something we could possibly detect if we could get fungal DNA out of the samples.
-Maybe worth mentioning this somewhere.}
 While we used relatively coarse presence/absence data based on fungal morphology, other studies have examined historic plant microbiomes using molecular sequencing and sophisticated bioinformatics techniques, but these studies have so far been limited to relatively few specimens at limited spatial extents \cite{yoshida2015computational,heberling2019utilizing, bieker2020metagenomic,gross2021hidden,bradshaw2021global}.
 Continued advances in capturing historic DNA and in filtering out potential contamination during specimen storage \citep{daru2019novel,bakker2020herbarium,raxworthy2021mining} will be imperative in the effort to scale up these efforts.
-This scaling up will be essential to be able to quantify changes not just in the prevalence of symbionts, \tom{but also in symbionts' intraspecific variation and evolutionary responses to climate change}{OK maybe this is the place for it.
-I would add a sentence about how endophytes can vary by chemotype and different chemotypes can confer different benefits.}, as well as  in changes in the wider microbial community.
+This scaling up will be essential to be able to quantify changes not just in the prevalence of symbionts, but also in symbionts' intraspecific variation and evolutionary responses to climate change, as well as  in changes in the wider microbial community.
+Genetic variation in \emph{Epichloë} endophytes, particularly in genes responsible for alkaloid production, produces ``chemotypes" with differing benefits for hosts against insect or mammalian herbivores mediated by environmental conditions \cite{schardl2012chemotypic,saikkonen2013chemical}.
+With improved molecular insights from historic specimens, we could ask whether the broad increases in endophytes that we have identified reflect selection for particular chemotypes and how this selection varies across space.
 Answering these questions as well as the unknown questions that future researchers may ask also reiterates the value in capturing meta-information during ongoing digitization efforts at herbaria around the world and during the accession of newly collected specimens \citep{lendemer2020extended,edwards2024university}.
 Second, we accounted for several potential biases in the data observation process that may be common to many collections-based research questions by using a spatially-explicit random effects model. 
-Spatial autocorrelation \cite{willems2022forest}, potential biases introduced by the sampling habits of collectors \citep{daru2018widespread}, and variation between contemporary researchers during the collection of trait data, if not corrected for could lead to over-confident inference about the strength and direction of historic change (I would cite the AGHY figure as an example).
-\tom{Previous studies that have quantified the effects of collector biases typically find them to be small  \cite{davis2015herbarium,meineke2019herbarium}, and we similarly did not find that collector has a strong effect on the results of our analysis.}{This seems to undercut the previous sentence, but its also a place where you could mention that scorer effects DID matter, so this is worth incorporating when possible.}
+Spatial autocorrelation \cite{willems2022forest}, potential biases introduced by the sampling habits of collectors \citep{daru2018widespread}, and variation between contemporary researchers during the collection of trait data, if not corrected for could lead to over-confident inference about the strength and direction of historic change (Fig.
+\ref{fig:temporal}).
+Previous studies that have quantified the effects of collector biases typically find them to be small  \cite{davis2015herbarium,meineke2019herbarium}, and we similarly did not find that collector has a strong effect on the results of our analysis, but that scorer identity did impact results.
 %Fitting this model in a Bayesian framework allows for full propagation of uncertainty.
@@
 Ultimately, a central goal of global change biology is to generate predictive insights into the future of natural systems on a rapidly changing planet. 
-While this survey of historic endophyte prevalence is necessarily correlative, it serves as a foundation to develop better predictive models of the response of microbial symbioses to climate change. 
-Combining the insights from this type of regional-scale survey with field experiments and physiological data \tom{could be invaluable. }{Vague.}
-While we found that certain dimensions of climate change were strongly correlated with endophytes' temporal responses, we do not know why trends in prevalence were weak in some areas or how endophytes would respond to more extreme changes in climate.
-\tom{For example, transplanting symbiotic and non-symbiotic plants beyond the range edge of \emph{A.
-hyemalis} could tell us whether persistent low endophyte prevalence in that area is a result of environmental conditions that lead the symbiosis to negative fitness consequences, or is a result of some historical contingency or dispersal limitation that has thus far limited the presence of symbiotic hosts from a region where they would otherwise flourish and provide resilience.}{This does not fit well here.
-I would move earlier where you talk about geographic trends.
-I also want to highlight that there is an important difference between associations of endophyte prevalence and geographic variation in climate (as in this AGHY example), and associations between CHANGES in endophyte prevalence and CHANGES in climate (the lines figure).
-Too often the manuscript blurs this disinction, and it happens here.}
-While the ``time machine'' of natural history collections revealed evidence of mutualism resilience for grass-endophyte symbioses in the face of environmental change, more extreme changes could potentially push one or both partners beyond their physiological limits, leading to the collapse of the mutualism; more research is needed to understand what those limits might be.
-\tom{Our analysis thus far is agnostic to changes in the distributions of hosts. 
-Mechanistic models could connect the responses of both host and symbionts to abiotic climate drivers, integrating dispersal processes. }{Does not feel like it fits well as a concluding point.}
-\tom{Beyond host-microbe symbioses, building these types of models would work towards quantitatively attributing biotic responses to anthropogenically driven climate change, similar to methods in climate science and economics \citep{stott2010detection, carleton2016social}.}{Not knowing these citations, I am not really sure what you are referring to.
-This does not feel like a natural point to end on.}
-%Elymus genotype
-%Agrostis weedy
+Beyond host-microbe symbioses, detecting ecological responses to anthropogenic global change and attributing their causes would inform public policy decision-makers and adaptive management strategies.
+This survey of historic endophyte prevalence is necessarily correlative, yet it serves as a foundation to develop better predictive models of the response of microbial symbioses to climate change. 
+By comparing detected ecological responses with alternative mechanistic simulations of the past, we could attribute their cause, in a manner similar to methods from climate science and economics \citep{stott2010detection, carleton2016social,trenberth2015attribution}.
+Combining the insights from this type of regional-scale survey with field experiments and physiological performance data could be invaluable to identify mechanisms driving shifts in host-symbiont dynamics.
+Evidence is strong that certain dimensions of climate change correlated with endophytes' temporal responses, however we do not know why trends in prevalence were weak in some areas or how endophytes would respond to more extreme changes in climate.
+The ``time machine'' of natural history collections revealed evidence of mutualism resilience for grass-endophyte symbioses in the face of environmental change, but more extreme changes could potentially push one or both partners beyond their physiological limits, leading to the collapse of the mutualism; more research is needed to understand what those limits might be.
 
@@
 	\section*{Data and Code Availability}
-	\tom{Data from this publication will be made publically available upon acceptance and before that upon request.}{CHeck journal requirements.
-It is generally better to make data and code available during peer review.}
+	Data from this publication can be found through a publicly available repository \josh{()}{will update with the URL}.
 	Code for analyses can be found through a publicly available repository (https://github.com/joshuacfowler/EndoHerbarium) that will be permanently archived upon publication.
@@
 
+\subsection*{Mesh and Prior Sensitivity analysis}\label{sec:priorsensitivity}
+To test the influence that the triangulation mesh and choice of priors has on results, we compared model results across a range of meshes and priors.
+We re-ran our model for the mesh used in main body of the text (Fig.
+\ref{fig:meshplot}), which we refer to as the "standard mesh", and with a mesh with smaller minimum vertices (finer mesh).
+Finer scale meshes increase computation time.
+For each of these meshes, we ran the model with a range of priors defining the spatial range of our spatial random effects: 342km (the prior used for presented results), as well as ranges five times smaller (68 km) and five times larger (1714 km).
+We found generally that these choices did not alter the direction of model predictions, but did influence the associated uncertainty and magnitude of some effects.
+
+For overall temporal trends, we found that models with differing priors predicted consistently positive relationships over time (Fig. \ref{fig:prior_year_plot}).
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width = \linewidth]{../Plots/prior_comparison_year_plot.png}
+	\caption[Overall trend in endophyte prevalence evaluated for models with different range priors on spatially structured random effects, and for two different meshes]{\textbf{Overall trend in endophyte prevalence evaluated for models with different range priors on spatially structured random effects, and for two different meshes.} Note that these plots, as compared to Fig.
+2 in main text, show mean trends and do not incorporate prediction uncertainty associated with collector and scorer random effects.}
+	\label{fig:prior_year_plot}
+\end{figure}
+
+For spatially-varying temporal trends, we found that models with different priors predicted consistent spatial patterns in temporal trends, although the range of this prediction varied depending on the prior and mesh (Fig.
+\ref{fig:standard_comparison_svc_plot} - \ref{fig:finer_comparison_svc_plot}).
+One noteworthy result of this analysis is that combinations of prior choice and mesh can introduce instability in model fitting.
+This is evident in \ref{fig:standard_comparison_svc_plot} panel B and \ref{fig:finer_comparison_svc_plot} panel B, where the prior range is smaller than the minimum vertex length of the mesh.
+Model fitting takes an extended time period and the model struggles to identify variation across space.
+Results with a set of prior ranges (Fig.
+\ref{fig:standard_comparison_svc_plot} -  A and C; Fig.
+\ref{fig:finer_comparison_svc_plot} - A and C) result in models that estimate trends across space of the same direction and order of magnitude, although the ``smoothness" of these predictions vary.
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width = .8\linewidth]{../Plots/standard_mesh_comparison_svc_plot.png}
+	\caption[Spatially-varying trends in endophyte prevalence evaluated for models with different range priors on spatially structured random effects, and for the "standard" mesh]{\textbf{Spatially-varying trends in endophyte prevalence evaluated for models with different range priors on spatially structured random effects, and for the "standard" mesh.} Shading indicates the magnitude and direction of predicted trends for each of three host species for each of three prior ranges (rows A-C).
+Note that each plot has an individual scale bars.}
+	\label{fig:standard_comparison_svc_plot}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width = .8\linewidth]{../Plots/finer_mesh_comparison_svc_plot.png}
+	\caption[Spatially-varying trends in endophyte prevalence evaluated for models with different range priors on spatially structured random effects, and for the "finer" mesh]{\textbf{Spatially-varying trends in endophyte prevalence evaluated for models with different range priors on spatially structured random effects, and for the "finer" mesh.} Shading indicates the magnitude and direction of predicted trends for each of three host species for each of three prior ranges (rows A-C).
+Note that each plot has an individual scale bars.}
+	\label{fig:finer_comparison_svc_plot}
+\end{figure}
+
 	% In most cases, authors should typeset supplementary material in a separate,

### 2024-10-03 Analyses/posthoc_climate_correlation.R
pulling out some summaries of climate and of the relative change for each species.

@@ -338,11 +338,18 @@ prism_diff_pred_df <- aghy_prism_diff_pred_df %>%
   full_join(elvi_prism_diff_pred_df) %>% 
   pivot_longer(contains("_diff")) %>% 
   separate(name, into = c("climate", "season"), sep = "_", extra = "drop", remove = FALSE) %>% 
-  filter(!grepl("_cv_", name) & ! grepl("annual", name)) %>% 
+  # filter(!grepl("_cv_", name) & ! grepl("annual", name)) %>% 
   mutate(season_f = factor(season, level = c("spring", "summer", "autumn")),
          moment = case_when(grepl("_sd_", name) ~ "sd",
                             TRUE ~ "mean"))
   
+prism_summary <- prism_diff_pred_df %>% 
+  filter(species == "A. hyemalis") %>% 
+  ungroup() %>% 
+  group_by(climate, season, moment) %>% 
+  dplyr::summarise(mean_change = mean(value),
+                   max_change = max(value),
+                   min_change = min(value))
 
 
 AGHY_tmean_change_plot <- ggplot(filter(prism_diff_pred_df, climate == "tmean" & species == "A. hyemalis"))+
@@ -760,7 +767,8 @@ climate_post_df <- bind_rows(aghy_post_df, agpe_post_df, elvi_post_df) %>%
   pivot_longer(cols = -c(species, variable), names_to = "iteration") %>% 
   mutate(positive = value>0) %>% 
   group_by(species, variable) %>% 
-  dplyr::summarise(prob_pos = (sum(positive)/500)*100) %>% 
+  dplyr::summarise(post_median = median(value),
+                   prob_pos = (sum(positive)/500)*100) %>% 
   mutate(overlap = case_when(prob_pos >=95 ~ "significant",
                              prob_pos <95 & prob_pos >5 ~ "flat",
                              prob_pos <=5 ~ "significant"))

### 2024-10-04 Analyses/ecolab_validation_data_prep.R
moving contemporary data prep script to the analysis folder

@@ -114,7 +114,7 @@ ggplot(ecolab_random_sample)+
 
 # getting the county for each survey
 library(ggmap)
-register_google(key = "AIzaSyDuAdpozRqmb8Sms-XfivxLi3tzlifJdMw")
+register_google(key = "")
 contemp_surveys_county <- ecolab_prevalence %>% 
   mutate(coords = as.character(paste(lat, lon))) %>% 
   geocode(coords, output = "more")

### 2024-10-04 Manuscript/endo_herbarium.tex
small edits to the manuscript

@@
 
-	\textit{Manuscript type}: Article. %Or note, natural history miscellany note, comment, reply, invited symposium, featured topic, or historical perspective.
+	\textit{Manuscript type}: Research Article. %Or note, natural history miscellany note, comment, reply, invited symposium, featured topic, or historical perspective.
 
@@
 Species' distributions and abundances are shifting in response to ongoing global climate change. 
-Mutualistic microbial symbionts can provide their hosts with protection from environmental stress. 
-This protection may thus contribute towards resilient responses to environmental change, however these changes may also disrupt species interactions and lead to declines in hosts and/or symbionts.
-Symbionts preserved within natual history specimens offer a unique opportunity to quantify changes in microbial symbiosis across broad temporal and spatial scales. 
-We asked how the prevalence of seed-transmitted fungal symbionts of grasses (\emph{Epichloë} endophytes), which can protect hosts from abiotic stress, have changed over time in response to climate change, and how these changes vary across host species' ranges.
+Mutualistic microbial symbionts can provide their hosts with protection from environmental stress that may contribute towards resilient responses to environmental change, however these changes may also disrupt species interactions and lead to declines in hosts and/or symbionts.
+Symbionts preserved within natural history specimens offer a unique opportunity to quantify changes in microbial symbiosis across broad temporal and spatial scales. 
+We asked how the prevalence of seed-transmitted fungal symbionts of grasses (\emph{Epichloë} endophytes) have changed over time in response to climate change, and how these changes vary across host species' ranges.
 Specifically, we analyzed 2,346 herbarium specimens of three grass host species (\emph{Agrostis hyemalis},\emph{Agrostis perennans},\emph{Elymus virginicus}) collected over the past two centuries (1824 -- 2019) for the presence or absence of \emph{Epichloë} symbiosis.
@@
 Notably, increasing precipitation during the peak growing season for \emph{Agrostis} species and decreasing precipitation for \emph{E. virginicus} were associated with increasing endophyte prevalence.
-We also identified changes in the variability of precipitation and temperature during off-peak seasons as important predictors of increasing endophyte prevalence. 
+Changes in the variability of precipitation and temperature during off-peak seasons were also important predictors of increasing endophyte prevalence. 
 Our analysis performed favorably in an out-of-sample predictive test with contemporary survey data, a rare extra step in collections-based research.
-However, we identified greater local-scale variability in endophyte prevalence in contemporary data compared to what our model could predict based on historic data, suggesting new directions that could improve predictive accuracy.
+However, we identified greater local-scale variability in endophyte prevalence in contemporary data compared to model predictions based on historic data, suggesting new directions that could improve predictive accuracy.
 Our results provide novel evidence for a cryptic biological response to climate change that may contribute to the resilience of host-microbe symbiosis through context-dependent benefits that confer a fitness advantage to symbiotic hosts under environmental change.
 
-Abstract : 315 words
+Abstract : 300 words
 
@@
 
+\newpage{}
+	\bibliographystyle{plainnat}
+	\bibliography{endo_herbarium}
+
+\newpage{}
+
 	\section*{Appendix A}
@@
 	% Dryad references.
-	\newpage{}
-\bibliographystyle{plainnat}
-\bibliography{endo_herbarium}
-
-	\newpage{}
 

### 2024-12-31 Analyses/endo_herbarium_multispecies_spatiotemporal_analysis.R
adding plots of the predicted prevalence across space.

@@ -148,7 +148,7 @@ collections_map <- ggplot()+
   labs(x = "Longitude", y = "Latitude", color = "Host Species")
 
 # collections_map
-# ggsave(collections_map, filename = "Plots/collections_map.png", width = 7, height = 4)
+ggsave(collections_map, filename = "Plots/collections_map.png", width = 7, height = 4)
 
 endo_status_map <- ggplot()+
   geom_map(data = outline_map, map = outline_map, aes( map_id = region), color = "grey", linewidth = .1, fill = "#FAF9F6")+
@@ -162,7 +162,7 @@ endo_status_map <- ggplot()+
         strip.text  = element_text(face = "italic", color = "black"))+
   labs(x = "Longitude", y = "Latitude", color = "Endophyte Status")
 # endo_status_map
-# ggsave(endo_status_map, filename = "Plots/endo_status_map.png", width = 10, height = 5)
+ggsave(endo_status_map, filename = "Plots/endo_status_map.png", width = 10, height = 5)
 
 
 
@@ -1034,106 +1034,196 @@ ggsave(svc_time_map.CI, filename = "Plots/svc_time_map_CI.png", width = 15, heig
 ################################################################################################################################
 ##########  Plotting the spatial Intercepts ###############
 ################################################################################################################################
-svc.pred_space <- list()
-svc.pred_space_year <- list()
+
+
+# generating predicted prevalence in 1895
+vrt_aghy@data <- expand.grid(std_year = rep(1895-mean_year, length.out = length(vrt_aghy)),
+                             species_index = 1,
+                             species = species_names[1])
+
+vrt_agpe@data <- expand.grid(std_year = rep(1895-mean_year, length.out = length(vrt_agpe)),
+                             species_index = 2,
+                             species = species_names[2])
+
+vrt_elvi@data <- expand.grid(std_year = rep(1895-mean_year, length.out = length(vrt_elvi)),
+                             species_index = 3,
+                             species = species_names[3])
 
 
 
[…]

### 2024-12-31 Manuscript/endo_herbarium_R1.tex
starting revisions

--- 2024-10-04 Manuscript/endo_herbarium.tex
@@
 \usepackage[T1]{fontenc}
-\usepackage{xcolor}
+\usepackage[usenames,dvipsnames]{xcolor}
+\usepackage{hyperref}
 
@@
 \newcommand{\josh}[2]{{\color{orange}{#1}}\footnote{\textit{\color{orange}{#2}}}}
+
+\newcommand{\revise}[1]{{\color{Mahogany}{#1}}}
+
 %%%%%%%%%%%%%%%%%%%%%
@@
 
-	\textit{Keywords}: climate change, plant-microbe symbiosis, herbarium, museum specimen, spatially-varying coefficients model, \emph{Epichloë}, \emph{Poaceae}.
+	\textit{Keywords}: climate change, plant-microbe symbiosis, herbarium, museum specimen, INLA, spatially-varying coefficients model, \emph{Epichloë}, \emph{Poaceae}.
 
@@
 	\section*{Abstract}
-Species' distributions and abundances are shifting in response to ongoing global climate change. 
-Mutualistic microbial symbionts can provide their hosts with protection from environmental stress that may contribute towards resilient responses to environmental change, however these changes may also disrupt species interactions and lead to declines in hosts and/or symbionts.
+\linelabel{R2-C15-begin}Species' distributions and abundances are shifting in response to ongoing global climate change. 
+Mutualistic microbial symbionts can provide hosts with protection from environmental stress that may contribute towards \revise{resilience under} environmental change, however \revise{this change} may also disrupt species interactions and lead to declines in hosts and/or symbionts.
 Symbionts preserved within natural history specimens offer a unique opportunity to quantify changes in microbial symbiosis across broad temporal and spatial scales. 
-We asked how the prevalence of seed-transmitted fungal symbionts of grasses (\emph{Epichloë} endophytes) have changed over time in response to climate change, and how these changes vary across host species' ranges.
-Specifically, we analyzed 2,346 herbarium specimens of three grass host species (\emph{Agrostis hyemalis},\emph{Agrostis perennans},\emph{Elymus virginicus}) collected over the past two centuries (1824 -- 2019) for the presence or absence of \emph{Epichloë} symbiosis.
-We found that endophytes increased in prevalence over the last two centuries from ca. 25\% prevalence to ca. 75\% prevalence, on average, across three host species.
+We asked how the prevalence of seed-transmitted fungal symbionts of grasses (\emph{Epichloë} endophytes) have changed over time in response to climate change, and how these changes vary across host species' \revise{distributions}.
+Specifically, we \revise{examined} 2,346 herbarium specimens of three grass host species (\emph{Agrostis hyemalis},\emph{Agrostis perennans},\emph{Elymus virginicus}) collected over the past two centuries (1824 -- 2019) for the presence or absence of \emph{Epichloë} symbiosis.
+\revise{Analysis of an approximate Bayesian spatially-varying coefficients model implemented in INLA revealed} that endophytes increased in prevalence over the last two centuries from ca.
+25\% to ca.
+75\% prevalence, on average, across three host species.
 Changes in seasonal climate drivers were associated with increasing endophyte prevalence. 
@@
 However, we identified greater local-scale variability in endophyte prevalence in contemporary data compared to model predictions based on historic data, suggesting new directions that could improve predictive accuracy.
-Our results provide novel evidence for a cryptic biological response to climate change that may contribute to the resilience of host-microbe symbiosis through context-dependent benefits that confer a fitness advantage to symbiotic hosts under environmental change.
+Our results provide novel evidence for a cryptic biological response to climate change that may contribute to the resilience of host-microbe symbiosis through \revise{fitness benefits to symbiotic hosts}.
+\linelabel{R2-C15-end}
 
@@
         \subsection*{Focal species}
-Our surveys focused on three native North American grasses: \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}. 
-Both \emph{Agrostis} species host \emph{Epichloë amarillans} \cite{craven2001multigene, leuchtmann2014nomenclatural}, while \emph{Elymus virginicus} typically hosts \emph{Epichloë elymi} \cite{clay2002evolutionary}.
-These $C_3$ grass species are commonly represented in natural history collections with broad distributions covering much the eastern United States (Fig. \ref{fig:map}).
+Our surveys focused on three native North American grasses: \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus} that host \emph{Epichloë} symbionts. 
+\linelabel{R2C23-begin}\revise{These cool-season grass species are commonly represented in natural history collections with broad distributions covering much the eastern United States (Fig.
+\ref{fig:map}).
+Cool-season grasses typically grow actively during the cooler temperatures of spring and autumn due to their reliance on $C_3$ photosynthesis.}\linelabel{R2C23-end}
 \emph{A. hyemalis} is a small short-lived perennial species that germinates in spring and typically flowers between March and July (most common collection month: May).
@@
 It begins flowering as early as March or April but continues throughout the summer (most common collection month: July).
+
+\linelabel{R1C3-begin}
+\revise{Both \emph{Agrostis} species host \emph{Epichloë amarillans} \cite{craven2001multigene, leuchtmann2014nomenclatural}, while \emph{Elymus virginicus} typically hosts \emph{Epichloë elymi} \cite{clay2002evolutionary}.
+The fungal symbionts primarily reproduce asexually and are passed from mother to offspring by vertical transmission through seeds. 
+These traits contribute to highly specialized interactions between symbiont and host.
+Some host species have been shown to partner with multiple symbiont species in these genus, and in some cases multiple symbiont species can co-exist within a host population \cite{mc2014species}.
+However, suveys have typically found limited \emph{Epichloë} genotypic diversity within host populations\cite{treindl2023genetic}. 
+Across host populations, concentrations of biologically-active alkaloids and the genes associated with their production vary substantially \cite{schardl2012chemotypic}.}
+\linelabel{R1C3-end}
 
@@
 They are samples from local populations, but they are single specimens that are aggregated to derive broad-scale model estimates.
-This suggests that increasing local replication should be a factor considered in future collection efforts of natural history specimens, balanced with the required time and effort.
+This suggests that increasing local replication should be a factor considered in future collection efforts of natural history specimens, balanced with the required time and effort \linelabel{R1C9-begin} \revise{and with limitations on storage space within collections}.
+\linelabel{R1C9-end}
 Poor predictive ability at local scales in this grass-endophyte system is not surprising, as previous studies have found that local variation, even to the scale of hundreds of meters can structure endophyte-host niches \cite{kazenel2015mutualistic}.
@@
 Our analysis advances the use of herbarium specimens in global change biology in two ways. 
-First and foremost, this is the first study to link long-term changes in microbial symbioses to changes in climate using specimens from natural history collections.
+\linelabel{R1C7-begin}
+First and foremost, this is \revise{one of a growing number of studies to examine microbial symbiosis using specimens from natural history collections, and the first, to our knowledge, to link long-term changes in the symbioses to changes in climate.}\linelabel{R1C7-end}
 The responses of microbial symbioses are a rich target for future studies within museum specimens, particularly those that take advantage of advances in sequencing technology.
@@
 	\section*{Data and Code Availability}
-	Data from this publication can be found through a publicly available repository \josh{()}{will update with the URL}.
+	\sloppy{
+	Data from this publication can be found through a publicly available repository (https://doi.org/10.5061/dryad.rn8pk0pn0).
 	Code for analyses can be found through a publicly available repository (https://github.com/joshuacfowler/EndoHerbarium) that will be permanently archived upon publication.
-
+}
 \newpage{}

### 2024-12-31 Manuscript/Reviewer_response_endo_herbarium.tex
adding draft of response to reviewers

% ======================================================= %
% Document: TEMPLATE FOR RESPONSES TO REVIEWERS
% Author: Andrea Ballatore
% Date: Jan 7, 2013
% Source: https://raw.githubusercontent.com/ucd-spatial/Datasets/master/tex_response_to_reviewers_template/responses_to_reviewers.tex
% Modified by Eduard Szöcs, 10.03.2015
% ======================================================= %
\documentclass[12pt]{article}

% packages

\usepackage{graphicx}
\usepackage{url}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{color}
\definecolor{mygray}{gray}{0.6}
\usepackage[utf8]{inputenc}
\usepackage[onehalfspacing]{setspace}
\usepackage[
	round,	%(defaultage in the main file and \input ) for round parentheses;
	colon,	% (default) to separate multiple citations with colons;
	authoryear,% (default) for author-year citations;
	sort,		% orders multiple citations into the sequence in which they
]{natbib}
\usepackage[%disable
	]{todonotes}

\usepackage{anysize}
\marginsize{2.5cm}{2.5cm}{1.5cm}{2.5cm}

\usepackage{xr}
\externaldocument[ms-]{endo_herbarium_R1}

% macros
% add a counter
\newcounter{cN}
\setcounter{cN}{0}

\newcommand{\comment}[1]{
	\vspace{2em}
	\refstepcounter{cN} % incrment counter
	\noindent \hangindent=0em \textbf{\textcolor{Maroon}{\uline{Comment \thecN}:~}}\emph{``#1''}
	}

\newcommand{\response}[1]{
	\\[0.25em]
	\hangindent=2.3em \textbf{\textcolor{NavyBlue}{\uline{Response}:~}}#1
	}

\newcommand{\revise}[1]{{\color{Mahogany}{#1}}}

\newcommand{\linesref}[2]{
		\\[0.25em]
	\hangindent=2.3em {\color{Mahogany} Line(s): \ref{#1} - \ref{#2}}
}

%\IfValueT{#2}{- \ref{#2}}

\usepackage[normalem]{ulem}
\definecolor{darkred}{rgb}{1,.6,.6}
\DeclareRobustCommand\problemline{\bgroup\markoverwith{\textcolor{darkred}{\rule[-0.9ex]{4pt}{3pt}}}\ULon}
\DeclareRobustCommand{\problem}[1]{\problemline{#1}} % soul
\setcounter{secnumdepth}{-1}

\begin{document}
% ======================================================= %
\title{Manuscript GCB-24-2886 --- Response to reviewers}

\maketitle
% ======================================================= %
\noindent To the editorial board,

Thank you for the opportunity to submit a revision of our manuscript to \emph{Global Change Biology}. 
In response to both reviewers, we have added substantial discussion of the ecological implications for hosts of our finding that symbiosis with \emph{Epichloë} fungal endophytes has increased in prevalence across the last two centuries.
We have also made significant revisions throughout the introduction, methods, results, and discussion to improve the organization and clarity of our ideas in response to Reviewer 2's recommendations.   
In addition, the utility of our supplemental material has been greatly improved in response to Reviewer 2's recommendations for more clearly captioning supplemental figures and their connections to our main results.

All of our changes are denoted in the revised manuscript with \revise{Mahogany font}. 
We think the review process has greatly strengthened our work such that it is now suitable for publication.
We hope you agree.

\vspace{2em}
Sincerely,

Joshua Fowler

Jacob Moutouama

Tom Miller

\newpage

% ======================================================= %
\section{Response to Reviewer 1}
\vspace{-2em}

\comment{I read and reviewed the paper titled “Increasing prevalence of plant-fungal symbiosis across two centuries of environmental change”. The authors of this paper wanted to know if the prevalence of grass-Epichloe fungus symbiosis has changed throughout time, particularly in response to temperature and precipitation. To answer this question, they primarily implemented a study based on herbarium collections for three grass species with the added bonus of evaluating the strength of their findings with a field survey. In general, the authors found strong evidence for changes in the prevalence of this symbiotic relationship along gradients of temperature, precipitation, latitude, and longitude and evidence for general increases in the prevalence of this relationship over time.\\
On the whole, this paper was a pleasure to read and I think it’s in my top two or three papers that I read this year. Not only was it well-written, but the statistics were above and beyond what is typical or expected in this field and just really thorough and well-executed. The authors spend a lot of space describing the novelty of their modeling approach and I think they did it justice for how novel it is in application to biotic interactions. This is one of those papers that (I hope) will be a foundational paper that most other global change ecologists will refer to both in terms of the ecological theory they advance and in terms of the statistical methodology they utilize. Here the authors are breaking new ground in both.}
\response{We are glad for this reviewer's positive response to our manuscript, and appreciate that they see value and novelty in our research for other global change ecologists.  }

\comment{I did have a few key criticisms of the paper or at least points that I think the authors would do well to consider.}
\response{We appreciate Reviewer 1's constructive criticisms, here and in their other line-specific comments. We have made our best effort to diligently revise our paper in response to these criticisms, which we feel has been tangibly improved in the process.}

 \comment{First, on a relatively minor note, I thought the authors could spend a bit more space on providing information about specific interacting pairs and how much variation there is among different fungal species interacting with the three grass species that were investigated. The authors alluded to it being basically a 1:1 relationship with 1 grass interacting with 1 fungus, but I was hoping for a little bit more info here about how specific these relationships are over time (acknowledging that the vertical transmission of fungus from generation to generation would support this notion). Still, is it possible that different Epichloe species interrupt existing relationships? Or even that a single grass species might interact with more than one Epichloe species at the population level or in different parts of their range? Does vertical transmission maintain prevalence through time, or is there regional turnover in relative prevalence of different species through time?}
\response{These are insightful questions that we agree deserve careful consideration. We have added additional description of genotypic diversity within the interaction to provide readers with this context in our methods section. In addition, we return to this topic in expanded discussion material (Lines XXX)}
\linesref{ms-R1C3-begin}{ms-R1C3-end}

\comment{The other main thing that stood out to me was that, out of the 9 herbaria sampled from, four are from Texas, 2 are from Oklahoma, and there is 1 each from Louisiana, Missouri, and Kansas. From Fig. 1 and Fig. A1, it appears that samples in these locations are far over-represented compared to other regions (e.g., Great Lakes Region, Southeast, Rust belt). This seems doubly important because the species that was found to have the largest year-to-year change in interaction prevalence (A. perannans) has substantially lower sample sizes and post-1969 compared to pre-1969 in the northern, cooler part of its range (Fig. A1), whereas the Midwest sample area appears approximately the same throughout. Could the relatively high estimated changes be due to bias associated with geographic distribution of herbarium collections used in the study? To answer this, I have three suggestions: (1) Would it be possible/feasible to add a handful of other herbaria to make the sampling more consistent throughout the study region (as described in the paper: eastern US)? Specifically regions in central and southern Appalachia and in the upper Midwest. (2) If not, would it be possible to employ some sort of sensitivity analysis to determine if the described trends are biased in favor of middle-America (i.e., beyond the INLA spatial autocorrelation already employed)? For example, could the authors assess whether bootstrapping the analysis so that there is relatively even sample size across each species’ range for any given model run affects the overall results? (3) If not, I at least recommend adding a caveat somewhere in the methods that collections were included from across eastern US, but were primarily collected from the four states in which herbaria were visited and that this has the potential to bias your results in specific ways (also see line comment on L504-507).}
\response{}

\comment{Lastly, I have included a handful of line comments below with relatively more minor suggestions to potentially improve the manuscript.\\
Despite these concerns, I still think this paper is excellent and really enjoyed reading it. I commend the authors for producing great work.}
\response{We thank Reviewer 1 for their diligent attention to our manuscript and are pleased that they have given us this positive evaluation.}

\comment{L115-118: Sentence beginning with “Critically evaluating whether insights…” – Just want to say that this is a great sentence and I wholeheartedly agree with the sentiment.}
\response{We thank again Reviewer 1 for sharing this positive feedback.}

\comment{Paragraph beginning L335 and Fig. 3: I think it would help if the authors provided some context about what initial estimated percent prevalence was in each region. Do we see regions with low prevalence increasing at faster rates through the study period? Or does prevalence increase at a relatively consistent rate regardless of “starting” conditions? Maybe adding another figure or another set of panels in Fig 3 with linear correlations between starting prevalence and per year change in prevalence could be helpful?
}
\response{These are an interesting questions. We have added supplemental figures which present the marginal spatial intercepts, along with predicted prevalence across space in 1895 and in 2020. As suggested by the reviewer, we also assessed the relationship between initial prevalence and predicted changes in prevalence. To keep the main text focused on temporal change in prevalence, we placed these figures in the supplemental material, but have added text describing this result.}
\linesref{arg1}{arg2}

\comment{Fig. 4: Text size very difficult to read currently.}
\response{}

\comment{L455-457: Minor suggestion, but authors might consider adding the challenges of space/storage constraints to this list of limitations in herbarium specimen replication.}
\response{This is a great suggestion which we have included in our revision.}
\linesref{ms-R1C9-begin}{ms-R1C9-end}

\comment{L480-481: This statement is false in that there is at least one previous study that attempted to do this with plant-mycorrhizal symbioses (Heberling \& Burke, 2019), a paper that the authors even cite a few sentences later. I recommend they dial this claim back.}
\response{We appreciate Reviewer 1 bringing this to our attention. We would argue that our phrasing is not inaccurate because we have successfully linked changes in symbioses to climate change drivers, whereas previous research including Heberling \& Burke 2019 focus on validating methods for quantifying microbial community composition within historic specimens, but do not connect this to global change drivers. Heberling \& Burke 2019 state, "Our results demonstrate that herbarium specimens have the potential to be utilized to effectively reconstruct historical AMF communities. However, given the scope of the current study, future work is needed to address specific global change hypotheses." However, we do recognize that there is a quickly growing body of research on microbial symbioses within herbarium specimens, especially those using sequencing and bioinformatics techniques. In response to this comment, we have rephrased this statement to clarify that we see novelty particularly in connecting microbial symbiosis to climate change drivers, and to temper our assertion of novelty.}
\linesref{ms-R1C7-begin}{ms-R1C7-end}

\comment{L504-507: See note above about concerns related to herbarium selection bias (i.e., only sampled from herbaria in southwest portion of organismal ranges). The biases listed by the authors here are correct, but I would also like to see them acknowledge how their own analysis contains certain biases that were not necessarily captured by the model and sampling structure they employed.}
\response{}

\comment{Discussion in general: The authors focused the discussion largely on the statistical analysis (which is definitely a huge strength of this paper and holds a lot of novelty in and of itself). However, I also found myself wanting the authors to return to some of the points they introduced earlier in the paper about what this symbiosis means for plant (or even fungal) performance/fitness and then tell me what their findings reflect about how climate change will affect these symbiotes in the future. Not just about the prevalence of the relationship (which they cover thoroughly), but about what this might mean for grass performance at the population or even community scales under ongoing climate change.}
\response{}

\section{Response to Reviewers 2}
\vspace{-2em}

\comment{In this study, Fowler, Moutouama \& Miller used herbarium data to explore the prevalence of Epichloe symbiosis across three plants over two centuries. The authors used spatial and joint-likelihood modelling framework to found out that the prevalence of the symbiosis increase consistently across the three plants across time. Moreover, they precipitation regimes as a key driver explaining the increased prevalence of this symbiosis.While I highly appreciate the quality of the dataset and the methods used, several major and minor concerns need to be addressed before publication. I remain particularly perplex about the structure of the manuscript as a whole. I found that the manuscript was hard to follow at times and that the structure of the sections could be significantly improved. I recommend a substantial revision of the manuscript, with better organization of the discussion (e.g., adding subheadings) to enhance clarity and readability.}
\response{We thank Reviewer 2 for their constructive feedback and attention towards our manuscript.}

\comment{While the research questions were clearly stated and clearly tested and answered by their results (clear and well structured), the hypothesis as well as the importance of the findings remains, in my opinion, “unbaked” at some points. Although the authors legitimately emphasized the importance and robustness of their data and methods throughout the manuscript, by contextualizing their results using recent literature and proposed some future research directions, the broader impact of this study on the field remains insufficiently addressed. A more in-depth discussion about the potential causes (ecological) of the results, and their implications for the ecology of the host plants or plants-endophytes interactions would be highly beneficial.}
\response{We appreciate Reviewer 2's positive evaluation of our work in regards to its success in answering our central research questions. Their review has helped us to thoroughly revise the structure of our work and improve organization. In addition, their desire to see more in-depth discussion of the ecological implications of our findings was echoed by Reviewer 1, and we have added new discussion material addressing this topic, detailed in response to specific comments.}
\linesref{arg1}{arg2} TO BE DONE

\comment{In the abstract, I recommend briefly mentioning the statistical analyses or framework used (e.g., INLA) before presenting the results to enhance readability.}
\response{We have revised the abstract to include information about the statistical analysis.}
\linesref{ms-R2-C15-begin}{ms-R2-C15-end}

\comment{Introduction: While, there is nothing scientifically wrong, I believe that the Introduction can be significantly shortened by going straight to the point. I suggest removing un-necessary examples (i.e, corals or insects’ symbioses), and reorganizing the content for clarity. I found some back-and-forth between discussions on methodological issues (i.,e L49-L60 ; L112-L118)) and the ecological importance of endophytes for host plants (L40-43; L75-82; and vice-versa). I would recommend to group the paragraphs presenting the research gaps of the ecological importance of endophytes, particularly in the context of climate change together and addressing the methodological novelty of this study in a separate, cohesive paragraph.}
\response{}

\comment{Discussion: he Discussion is the most difficult section to follow. Similarly, to the Introduction, I back-and-forth between paragraphs/arguments. I believe that the use of sub-headings (in your rewriting before to remove them) would greatly improve clarity by following the same order as your research questions. For example, you could begin by discussing about the spatial/temporal patterns in prevalence, followed by the relationships with climatic drivers before to conclude with the technical/ecological novelty of your findings.}
\response{}

\comment{Supplementary figures: Fig.A3; A4; A5; A6;A7;A8;A9;A10; A11;Table A1. The captions for these figures do not provide enough information to understand their meaning, making the supplementary figures unexploitable as such. To assist readers (who would otherwise need to juggle between the supplementary figures and the main text), please ensure that each figure’s caption includes: the specific model or analysis from which it was produced; the data used (e.g., whether it is based on all host plants or a particular plant, which period of data used, etc.); and defining terms or acronyms of these figures (examples: A and T from figure A6; $\tau$ from A7; conservative vs liberal A8; AGHY vs ELVI A11…..).}
\response{}

\comment{L54: Which physical legacies or ecological processes? Provide example.}
\response{}

\comment{L64-L74: Please remove example of symbiosis from other systems and focus on your targeted endophytic symbiosis. (Remove corals, insects or lichens.)}
\response{}

\comment{-L104: In which sense? How this hypothesis would be reflected in your results? Would you expect a distribution shifts? Please specify.}
\response{}

\comment{-L131-134: Super nice data!}
\response{}

\comment{-L140: Maybe trivia, so please feel free to ignore if I am wrong, but maybe specify that C3 refers to a metabolic type of plant?}
\response{Thank you for this suggestion. We have added additional context about C3 photosynthesis.}
\linesref{ms-R2C23-begin}{ms-R2C23-end}

\comment{-L167-168: What is the spatial resolution of the different units? What could be the distance between counties or between state centroids. And how big or small county can be?}
\response{}

\comment{-L176: “Partial” would be more adapted than “imperfect”.}
\response{}

\comment{-L180-187: I found that the use of conservative and liberal very confusing. And I am not sure I understood properly what they are. Moreover, I you do not mention these type of status in the results nor discussion. Please reformulate and clarify in the main text and maybe it can be used in the discussion?}
\response{}

\comment{-L190: Spatial resolution of the aggregation?}
\response{}

\comment{-L215: “borrow information” is more appropriate than “borrow strength”.}
\response{}

\comment{-L229: Why 5? Is there a reference?}
\response{}

\comment{-L241-L258: Why this paragraph appears here? I understand that this is for the spatial distribution of the endophytes (within a host) but should it be in the first paragraph? Should it be treated right after the prevalence, and before to mention the spatial models? I feel like this paragraph is coming out of the blue and is making the full material and methods confusing.}
\response{}

\comment{-L281: “which” instead of “whether”?}
\response{}

\comment{-L285-288: Spatial resolution?}
\response{}

\comment{-L316: “Confidence” instead of “certainty”}
\response{}

\comment{-Figure 2: Very nice figure.}
\response{}

\comment{-L323: “We found no evidence that collector biaises influenced our results” contradictory to L325 “The identity of individual scorers did contribute to observed patterns in endophyte prevalence.” I am confused. They are quite contradictory no?}
\response{lines \ref{}}

\comment{L328-333: This section should be in discussion rather than in results no?.}
\response{lines \ref{}}

\comment{-Figure 4: The model that have been used to produce these relationships should be mentioned in the caption. Moreover, in the caption, “points of 250 randomly sampled pixels across each host distribution”. First: Why not taking more points/all points? And what are the spatial resolution of these pixels?}
\response{lines \ref{}}

\comment{- L382-386: This should be in discussion rather than in results.}
\response{lines \ref{}}

\comment{L387-398: I feel like this part is coming quite late in the results. You have already been presented the results of the models, and relationships, I would suggests to move this earlier in the results section. Starting from model performance before to show results produced from this model.}
\response{lines \ref{}}

\comment{-L400: cryptic? This is a little bit vague, rephrase or change the word?}
\response{lines \ref{}}

\comment{- L402: I would replace “fungal endophytes” by Epichloe syombiosis.}
\response{lines \ref{}}

\comment{- L424: More than the lagged effects (which I think is a good point), it would be interested to discuss the potential effect of Epichloe on the host distribution? I would have liked that you linked your conservative/liberal status of symbiosis on this point and what you would expect in term of potential role for the host plant?}
\response{lines \ref{}}

\comment{L453-455: This part is hard to follow, please rephrase?}
\response{lines \ref{}}

\comment{L474: This is a good point that should maybe put in relation with other global changes such as land use changes. This would be nice to emphasize that probably land uses change impacted a lot/or not the natural populations of the three host species and emphasize that could also have impacted the ecological dynamics of your symbiosis at larger scale?}
\response{lines \ref{}}

\comment{L498: This is a good point as well. Moreover, the uses of eDNA metabarcoding or other sequencing methods to characterise the identity of your Epichloe composition would be very interesting to disscuss at this point. Would you expect a turnover of species (from Epichloe genera), that could impact the performance and potentially something about the dynamics of the plant-symbiotic dynamic?}
\response{lines \ref{}}

\comment{L501: “Second”, I did not see the “First”.}
\response{lines \ref{}}

\comment{Supplementary figures:-Page 1; I don’t know if its journal guidelines, but, should there be a title page, with authors information, affiliations and a title being Appendix to “Increasing prevalence of plant-fungal symbiosis across two centuries of environmental change” ?}
\response{lines \ref{}}

\comment{Fig.A3: Which data were used to produce the simulated datasets? Which model is this from?}
\response{lines \ref{}}

\comment{A4 and A5: Which model, which data? (All plants, all years)?}
\response{lines \ref{}}

\comment{A6; From which models this is from? How probabilities are calculated? What is A and T?}
\response{lines \ref{}}

\comment{A7; From which models; what is the legend for the value means, what is $\tau$?}
\response{lines \ref{}}

\comment{A8;redefine what is liberal vs conservative; what AGPE, AGHY and ELVI means if the Panel B-C-F is from which model. What are the value? (This is an average per year, across 200 years?)}
\response{lines \ref{}}

\comment{A9;A10; Specify models}
\response{lines \ref{}}

\comment{A11; Legend need specifications, the data need to be specified as well.}
\response{lines \ref{}}

\comment{Table A1. Which models, what are the columns names? It is for the full temporal I guess? Need to be specified.}
\response{lines \ref{}}

% ======================================================= %
\end{document}
% ======================================================= %

### 2025-01-21 Analyses/endo_herbarium_multispecies_spatiotemporal_analysis.R
adding calculations of bounding box size

@@ -29,7 +29,7 @@ endophyte_colors <- c("#fdedd3","#f3c8a8", "#5a727b", "#4986c7", "#181914",  "#1
 ############ Read in the herbarium dataset ############################### 
 ################################################################################
 
-endo_herb_georef <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/DigitizedHerbariumRecords/endo_herb_georef.csv") %>% 
+endo_herb_georef <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/DigitizedHerbariumRecords/endo_herb_georef.csv") 
   # filter(Country != "Canada") %>%
   mutate(species_index = as.factor(case_when(spp_code == "AGHY" ~ "1",
                                              spp_code == "AGPE" ~ "2",
@@ -94,6 +94,51 @@ endo_herb<- endo_herb %>%
 
 # write_csv(endo_herb, file = "filtered_endo_herb_georef.csv")
 
+# calculating the min and max distance between points
+
+
+# dist_vec <- cbind(endo_herb$easting, endo_herb$northing)
+# dist_mat <- as.matrix(dist(dist_vec, method = "euclidean", diag = TRUE, upper = TRUE))
+# min(dist_mat[dist_mat>0])
+# max(dist_mat[dist_mat>0])
+# median(dist_mat[dist_mat>0])
+# 
+# # calculating the width of georeferencing bounding boxes
+# 
+# endo_herb_boundbox_west <- endo_herb %>% 
+#   select(north, south, east, west, lon, lat, easting, northing) %>% 
+#   as_tibble() %>% 
+#   st_as_sf(coords = c("west", "lat"), crs = 4326, remove = FALSE) %>% 
+#   st_transform(epsg6703km) %>% 
+#   mutate(
+#     west_bb = st_coordinates(.)[, 1],
+#     lat_bb = st_coordinates(.)[, 2]
+#   ) %>% select(west_bb)
+# endo_herb_boundbox_east <- endo_herb %>% 
+#   select(north, south, east, west, lon, lat, easting, northing) %>% 
+#   as_tibble() %>% 
+#   st_as_sf(coords = c("east", "lat"), crs = 4326, remove = FALSE) %>% 
+#   st_transform(epsg6703km) %>% 
+#   mutate(
+#     east_bb = st_coordinates(.)[, 1],
+#     lat_bb = st_coordinates(.)[, 2]
+#   ) %>% select(east_bb)
+# 
[…]

### 2025-01-21 Analyses/endo_herbarium_records_merge.R
pulling out the bounding box info

@@ -1447,7 +1447,7 @@ endo_herb_georef <-endo_herb8 %>%
   unite("location_string" , sep = ", " , Municipality,County_fixed,State,Country, remove = FALSE, na.rm = TRUE) %>%
   # mutate(location_string = case_when(location_string == "NA NA" ~ NA, TRUE ~ location_string)) %>% 
   filter(Endo_status_liberal <= 1) %>% 
-  mutate_geocode(location_string) # Uncomment this to run the geocoding.
+  mutate_geocode(location_string, output = "more") # Uncomment this to run the geocoding.
 
 endo_herb_georef_1 <- endo_herb_georef %>% 
   mutate(sample_temp = Sample_id) %>% 
@@ -1462,6 +1462,7 @@ endo_herb_georef_dryad <- endo_herb_georef_1 %>%
                   Country, State, County, Municipality, Locality, hand_georef_lat, hand_georef_lon, year, month, day, tissue_type, seed_scored, seed_eplus, 
                   Endo_status_liberal, Endo_status_conservative, Date_scored, scorer_id, score_number, scorer_factor, collector_factor, lat, lon))
 write_csv(endo_herb_georef_dryad, file = "~/Desktop/endo_herb_georef.csv") # saving a version of the file used in ensuing scripts
+
 
 
 

### 2025-01-21 Analyses/posthoc_climate_correlation.R
replotting the climate correlations plots with better labels

@@ -366,7 +366,7 @@ AGHY_ppt_change_plot <- ggplot(filter(prism_diff_pred_df, climate == "ppt" & spe
   facet_wrap(~ moment+season)+
   scale_fill_viridis_c(option = "magma") + labs(x = "Lon.", y = "Lat.", fill = "Change in mm.")+
   theme(strip.text = element_text(size = rel(1)))
-# AGHY_ppt_change_plot
+AGHY_ppt_change_plot
 
 
 AGHY_climate_change_plot <- AGHY_tmean_change_plot/AGHY_ppt_change_plot + plot_annotation(tag_levels = "A")
@@ -804,8 +804,11 @@ tmean_trend <- ggplot(filter(prediction_df, grepl("tmean", name) & !grepl("sd", 
   labs(y = "Change in Prevalence (% per Year)", x = "Change in Temperature (ºC)")+
   theme_light()+
   theme(strip.background = element_blank(),
-        strip.text = element_text(color = "black", size = rel(1)),
-        strip.text.y = element_text(face = "italic", angle = 0),
+        strip.text = element_text(color = "black"),
+        strip.text.x = element_text( size = rel(2)),
+        strip.text.y = element_blank(),
+        axis.text = element_text(size = rel(1)),
+        axis.title = element_text(size = rel(1.5)),
         legend.text = element_text(face = "italic"),
         )
   # lims(y = c(0,1))
@@ -829,8 +832,11 @@ tmean_sd_trend <- ggplot(filter(prediction_df, grepl("tmean", name) & grepl("sd"
   labs(y = "Change in Prevalence (% per Year)", x = "Change in SD(Temperature) (ºC)")+
   theme_light()+
   theme(strip.background = element_blank(),
-        strip.text = element_text(color = "black", size = rel(1)),
-        strip.text.y = element_text(face = "italic", angle = 0),
+        strip.text = element_text(color = "black"),
+        strip.text.x = element_text( size = rel(2)),
+        strip.text.y = element_blank(),
+        axis.text = element_text(size = rel(1)),
+        axis.title = element_text(size = rel(1.5)),
         legend.text = element_text(face = "italic"),
   )
 # lims(y = c(0,1))
@@ -857,8 +863,11 @@ ppt_trend <- ggplot(filter(prediction_df, grepl("ppt", name) & !grepl("sd", name
   labs(y = "Change in Prevalence (% per Year)", x = "Change in Precipitation (mm.)")+
   theme_light()+
[…]

### 2025-01-23 Manuscript/endo_herbarium_R1.tex
my first round pass through all of the review comments

@@
 \usepackage{titlesec}
+\usepackage{caption}
 
@@
 
-	\textit{Manuscript elements}: Figure \ref{fig:map} - Figure \ref{fig:contemptestplot}, Appendix~A (including Figure~\ref{fig:endo_status_map} - Figure \ref{fig:ELVI_climate_covariates},  Table \ref{table:herbaria}, and Supplemental Methods).
+	\textit{Manuscript elements}: Figure \ref{fig:map} - Figure \ref{fig:contemptestplot}, Appendix~A (including Figure~\ref{fig:endo_status_map} - Figure \ref{fig:ELVI_climate_covariates},  Table \ref{table:herbaria}, and Supporting Methods).
 
@@
 We asked how the prevalence of seed-transmitted fungal symbionts of grasses (\emph{Epichloë} endophytes) have changed over time in response to climate change, and how these changes vary across host species' \revise{distributions}.
-Specifically, we \revise{examined} 2,346 herbarium specimens of three grass host species (\emph{Agrostis hyemalis},\emph{Agrostis perennans},\emph{Elymus virginicus}) collected over the past two centuries (1824 -- 2019) for the presence or absence of \emph{Epichloë} symbiosis.
+Specifically, we \revise{examined} 2,346 herbarium specimens of three grass host species (\emph{Agrostis hyemalis}, \emph{Agrostis perennans}, \emph{Elymus virginicus}) collected over the past two centuries (1824 -- 2019) for the presence or absence of \emph{Epichloë} symbiosis.
 \revise{Analysis of an approximate Bayesian spatially-varying coefficients model implemented in INLA revealed} that endophytes increased in prevalence over the last two centuries from ca.
@@
 These specimens often preserve physical legacies of ecological processes and species' interactions from dynamically changing environments across time and space.
-For example, previous researchers have used plant collections (herbaria) to document shifts in phenology \citep{willis2017old, park2019herbarium,  berg2019examination}, pollination \citep{pauw2011reconstruction, duan2019century}, and herbivory \citep{meineke2019herbarium} related to anthropogenic climate change.
+For example, previous researchers have used plant collections (herbaria) to document shifts in phenology \linelabel{R2C19-begin}\revise{by examining reproductive structures}\citep{willis2017old, park2019herbarium,  berg2019examination}, pollination \revise{through examination of pollen removal} \citep{pauw2011reconstruction, duan2019century}, and herbivory \revise{by documenting leaf damage} \citep{meineke2019herbarium} related to anthropogenic climate change.
+\linelabel{R2C19-end}
 However, few previous studies have leveraged biological collections to examine climate change-related shifts in a particularly common type of interaction: microbial symbiosis.
@@
 Previous survey studies of contemporary populations have documented large-scale spatial patterns in  endophyte prevalence structured by environmental gradients \citep{granath2007variation,bazely2007broad, afkhami2012fungal,sneck2017variation}.
-We predicted that prevalence should track temporal changes in environmental drivers that elicit strong fitness benefits.
+We predicted that prevalence should track temporal changes in environmental drivers \linelabel{R2C21-begin}\revise{(i.e. drought)} \linelabel{R2C21-end} that elicit strong fitness benefits.
 %For example, endophyte-mediated drought tolerance should lead endophyte prevalence to increase in regions where precipitation has declined over time.
@@
 We then addressed how climate change may be driving trends in endophyte prevalence by asking: (iii) What is the relationship between temporal trends in endophyte prevalence and associated changes in climate drivers?
-We predicted that aggregate endophyte prevalence would increase over time in tandem with climate warming, and that hotspots of endophyte change would correspond spatially to hotspots of climate change.
-Finally, we evaluated the performance of models built on data from historic specimens with an out-of-sample test, using data on endophyte prevalence from contemporary surveys of host populations. 
-To answer these questions we examined a total of 2,346 historic specimens collected across eastern North America between 1824 and 2019, and evaluated model performance against contemporary surveys comprising 1,442 individuals from 63 populations collected between 2013 and 2020.
+\linelabel{R2C21-b-begin}We predicted that aggregate endophyte prevalence would increase over time in tandem with climate warming, and that hotspots of endophyte change would correspond spatially to hotspots of climate change.
+\linelabel{R2C21-b-end}
+Finally, we evaluated \revise{(iv) how our model, built on data from historic specimens, performed in an out-of-sample test - }using data on endophyte prevalence from contemporary surveys of host populations.
+To answer these questions we examined a total of $2,346$ historic specimens collected across eastern North America between 1824 and 2019, and evaluated model performance against contemporary surveys comprising 1,442 individuals from 63 populations collected between 2013 and 2020.
 
@@
 However, suveys have typically found limited \emph{Epichloë} genotypic diversity within host populations\cite{treindl2023genetic}. 
-Across host populations, concentrations of biologically-active alkaloids and the genes associated with their production vary substantially \cite{schardl2012chemotypic}.}
+Across host populations, concentrations of biologically-active alkaloids and the genes associated with their production vary substantially \cite{schardl2012chemotypic}.
+In this analysis, we focus on the presence/absence of \emph{Epichloë} symbionts, and we discuss potential implications of hidden genotypic diversity in the Discussion.}
 \linelabel{R1C3-end}
@@
 For fifteen of the oldest specimens, only information at the state level was available, and so we used the state centroid.
+\linelabel{R2C24-begin}\revise{The median pairwise distance between georeferenced coordinate points was 841 km.
+The median longitudinal width of the bounding boxes generated to geocode municipality, county, or state centroids was 44.7 km. 
+Among those specimens geo-referenced at the state level, the largest bounding box, spanning the state of Texas, was 1233 km wide.
+The smallest bounding boxes were less than 1 km across for small municipalities (while this suggests high precision, we note that some specimens were collected in natural habitat nearby to small munipalities not encompassed by this bounding boxes).}\linelabel{R2C24-end}
+
+\linelabel{R1C11-begin}
+\revise{Our visits focused on herbaria with historic strengths in \emph{Poaceae} collections (e.g.
+Texas A\&M, Missouri Botanic Garden) and other herbaria in the Southern Great Plains region of the United States.
+While these nine herbaria garnered specimens that span the focal species' ranges, our dataset unevenly samples across the study region. 
+Texas, Oklahoma, Louisiana, and Missouri are the most represented states.
+Uneven sampling is most likely to be consequential for \emph{A. perennans}, which has much of its range in the northeastern US. 
+We explore the potential influence of spatial bias in sampling on our results through a simulation analysis (Appendix A - Supporting Methods).}\linelabel{R1C11-end}
 
@@
 virginicus}; we scored $10,342$ seeds in total.
-.
-Due to imperfect vertical transmission \cite{afkhami2008symbiosis}, it is possible that symbiotic host-plants produce a mixture of symbiotic and non-symbiotic seeds. 
+\linelabel{R2C25-begin}Due to imperfect vertical transmission, \revise{the production of symbiont-free offspring from symbiotic hosts} \cite{afkhami2008symbiosis},\linelabel{R2C25-end} it is possible that symbiotic host-plants produce a mixture of symbiotic and non-symbiotic seeds.
 We therefore designated a specimen as endophyte-symbiotic if \emph{Epichloë} hyphae were observed in one or more of its seeds, or non-symbiotic if \emph{Epichloë} hyphae were observed in none of its seeds.
-To capture uncertainty in the endophyte scoring process, we recorded both a "liberal" and a "conservative" endophyte status for each plant specimen.  
-When we identified potential endophytes with unusual morphology, low uptake of stain, or a small amount of fungal hyphae across the scored seeds, we recorded a positive liberal status (more likely to be endophyte-positive) and a negative conservative status (less likely to be endophyte-positive).
+\linelabel{R2C26-begin}
+To capture uncertainty in the endophyte \revise{identification} process, we recorded both a "liberal" and a "conservative" endophyte score for each plant specimen.  
+\revise{When we confidently identified endophytes within a specimen's seeds, we assigned matching liberal and conservative scores.}
+When we identified potential endophytes with unusual morphology, low uptake of stain, or a small amount of fungal hyphae across the scored seeds, we recorded \revise{ a positive identification} for the liberal score
+and a negative \revise{identification for} the conservative score.
+\revise{The liberal status assumed a potential endophyte identification was more likely to be endophyte-positive while the conservative status assumed that the potential endophyte identification was less likely to be endophyte-positive}.
 $89$\% of scored plants had matching liberal and conservative scores, reflecting high confidence in endophyte status.
-The following analyses used the liberal status, but we repeated all analyses with the conservative status which yielded qualitatively similar results (Fig. \ref{fig:conservative_scores_plot}).
+The following analyses used the liberal status, \revise{however} we repeated all analyses with the conservative status which yielded qualitatively similar results (Fig.
+\ref{fig:conservative_scores_plot}).\linelabel{R2C26-end}
 
@@
 	\includegraphics[width = \linewidth]{../Plots/collections_map.png}
-	\caption[Collection locations of herbarium specimens sampled for \emph{Epichloë} endophytes]{\textbf{Collection locations of herbarium specimens sampled for \emph{Epichloë} endophytes.} Specimens span eastern North America from nine herbaria, and are colored by host species.}
+	\caption[Collection locations of herbarium specimens sampled for \emph{Epichloë} endophytes]{\textbf{Collection locations of herbarium specimens sampled for \emph{Epichloë} endophytes.} Specimens span eastern North America from nine herbaria, and are colored by host species (\emph{A.
+hyemalis}: green, \emph{A.
+perennans}: orange, \emph{E.
+virginicus}: purple).
+\revise{Map lines delineate study areas and do not necessarily depict accepted national boundaries.}}
 	\label{fig:map}
@@
 \subsection*{Modeling spatial and temporal changes in endophyte prevalence}
-We assessed spatial and temporal changes in endophyte prevalence across each host distribution, quantifying the ``global'' temporal trends aggregated across space, and then examining spatial heterogeneity in the direction and magnitude of endophyte change (hotspots and coldspots) across the spatial extent of each host's distribution.
+We assessed spatial and temporal changes in endophyte prevalence across each host distribution, quantifying the ``global'' temporal trends \linelabel{R2C27-begin}\revise{averaged}\linelabel{R2C27-end} across space, and then examining spatial heterogeneity in the direction and magnitude of endophyte change (hotspots and coldspots) across the spatial extent of each host's distribution.
 To account for the spatial non-independence of geo-referenced occurrences, we used an approximate Bayesian method, Integrated Nested Laplace Approximation (INLA), to construct spatio-temporal models of endophyte prevalence.
@@
 We modeled $\hat{P}_{h,i}$ as a linear function of intercept $\boldsymbol{A}_h$ and slope $\boldsymbol{T}_{h}$ defining the global trend in endophyte prevalence specific to each host species as well as with spatially-varying intercepts $\alpha_{h,l_{i}} $ and slopes $\tau_{h,l_i}$ associated with location ($l_i$, the unique latitude-longitude combination of the $i$th observation).
-The joint-model structure allowed us to ``borrow strength'' across species in the estimation of shared variance terms for the spatially-dependent random effect $\delta_{l_i}$, intended to account for residual spatial variation, and $\chi_{c_i}$ and $\omega_{s_i}$ i.i.d.-random effects indexed for each collector identity ($c_i$), and scorer identity ($s_i$) of the $i^{th}$ specimen.
+\linelabel{R2C28-begin}The joint-model structure allowed us to ``borrow \revise{information}'' \linelabel{R2C28-end} across species in the estimation of shared variance terms for the spatially-dependent random effect $\delta_{l_i}$, intended to account for residual spatial variation, and $\chi_{c_i}$ and $\omega_{s_i}$ i.i.d.-random effects indexed for each collector identity ($c_i$), and scorer identity ($s_i$) of the $i^{th}$ specimen.
 
@@
 Prolific collectors who contribute thousands of specimens may be more or less likely to collect certain species, or specimens with certain traits \cite{daru2018widespread}. 
-Similarly, the process of scoring seeds for hyphae involved several student researchers who, even with standardized training, may vary in their likelihood of positively identifying \emph{Epichloë}.
+Similarly, the process of scoring seeds for hyphae involved \linelabel{R2C35-begin}\revise{multiple researchers (or "scorers")} who, even with standardized training, may vary in their likelihood of positively identifying \emph{Epichloë}.
+\linelabel{R2C35-end}
 
@@
 Global intercept and slope parameters $\boldsymbol{A}$, and $\boldsymbol{T}$, were given vague priors.
-Scorer and collector random effects, $\chi$ and $\omega$, were given penalized complexity priors, with distributions approximating a Normal distribution with standard deviation of 5. 
+Collector and scorer random effects, $\chi$ and $\omega$ \revise{respectively}, \linelabel{R2C29-begin}\revise{were centered at 0 with precision parameters were assigned penalized complexity (PC) priors with parameter values $U_{PC}$ = 1 and $a_{PC}$ = 0.01 \cite{simpson2017penalising}.}\linelabel{R2C29-end}
 Each spatially-structured parameter depended on a covariance matrix according to the proximity of each pair of collection locations \citep{lindgren2011explicit,bakka2018spatial}. 
@@
 \ref{fig:meshplot}).
-We assessed model fit with visual posterior predictive checks (\ref{fig:overlayplot}) and measurements of AUC (Figs. \ref{fig:ROCtraining}-\ref{fig:ROCtest}).
+We assessed model fit with visual posterior predictive checks (\ref{fig:overlayplot}) and measurements of AUC (Figs.
+\ref{fig:ROCtraining}-\ref{fig:ROCtest}) \linelabel{R2C48-begin} \revise{\cite{gelman2006data}}\linelabel{R2C48-end}.
 Priors for the Mat\'{e}rn covariance function, termed ''range" and "variance", define how proximity effects decay with distance.
@@
 the same direction of effects across space.
+\revise{Through results and discussion that follow, we refer to the model described in this section as the ``endophyte prevalence model''}.
 
@@
 \subsection*{Modeling distributions of host species}
-Because the herbarium records did not encompass the entirety of these host species'ranges, we additionally modeled the geographic distribution of each host species to generate realistic maps on which we could project the predictions of the INLA model.
+\linelabel{R2C30-begin}
+\revise{We modeled the geographic distribution of each host species with two goals: (1) generate realistic maps on which we could project the predictions of the INLA model, and (2) use the geographic distributions to test for relationships between climate change drivers and trends in endophyte prevalence.
+The herbarium records did not encompass the entirety of each host species' range.}\linelabel{R2C30-end}
 We followed the ODMAP (overview, data, model, assessment, prediction) protocol \citep{crossley2022opposing} (see Supporting Methods). 
@@
 %\tom{Among this suite of variables, we chose to include, which were uncorrelated (Variance Inflation Factor $>$ 0.7) and allowed us to predict the occurrence probability of each host species in space and time.}{Do not follow and sounds circular.}
-We fit maximum entropy (MaxEnt) models using the maxent function in the R package dismo \citep{hijmans2017package} using the same set of seasonal climate predictors considered above calculated for the 1990-2020 climate normals: mean and standard deviation of spring, summer, and autumn temperature, and mean and standard deviation of spring, summer, and autumn cumulative precipitation.
+We fit maximum entropy (MaxEnt) models using the maxent function in the R package dismo \citep{hijmans2017package} using the same set of seasonal climate predictors considered \revise{below} calculated for the 1990-2020 climate normals: mean and standard deviation of spring, summer, and autumn temperature, and mean and standard deviation of spring, summer, and autumn cumulative precipitation.
 %\josh{MaxEnt is preferred because it has been shown to generate response curves with less unpredictable behavior when applied to new climates \citep{hijmans2006ability}.}{possibly could remove this sentence?}
@@
 
-To evaluate whether areas that have experienced the greatest changes in endophyte prevalence (hotspots of endophyte change) are associated with high degrees of change in climate (hotspots of climate change), we modeled the fitted, spatially-varying slopes of endophyte change through time ($\tau_{[h]l}$ as a linear function of environmental covariates, with a Gaussian error distribution.
-Data from each host species was analyzed separately. 
+\linelabel{R2C32-begin}\revise{We then evaluated whether areas that have experienced the greatest changes in endophyte prevalence (hotspots of endophyte change) are associated with high degrees of change in climate (hotspots of climate change)
+To do so, we modeled the fitted, spatially-varying slopes of endophyte change through time ($\tau_{[h]l}$) as a linear function of environmental covariates, with a Gaussian error distribution for a set of pixels across each host distribution.
+The continuous SPDE approach taken for our endophyte prevalence model allows us to generate predictions of temporal trends in prevalence at arbitrarily many pixels across each host distribution.
+Balancing computation time with resolution, we generated predicted trends for 546, 645, and 753 pixels across each host distribution for \emph{A.
+perennans}, \emph{A.
+hyemalis}, and \emph{E.
+virginicus} respectively ( pixel dimensions: \emph{A.
+perennans} =  65 km x 36 km; \emph{A.
+hyemalis} = 61km x 45 km; \emph{E.
+virginicus} = 62 km x 40 km )}\linelabel{R2C32-end}
 Fitting regressions to many pixels across the study region risks articially inflating confidence in our results due to large sample sizes, and so we performed this analysis using only a random subsample of 250 pixels across the study region; other sizes of subsample yielded similar results.
+Data from each host species were analyzed separately. 
+\revise{Through the results and discussion that follow, we refer to this analysis as the ``\emph{post hoc} climate regression analysis''.}
 
@@
 \subsection*{How has endophyte prevalence changed over time?}
-Across >2300 herbarium specimens dating back to 1824, we found that prevalence of \emph{Epichloë} endophytes increased over the last two centuries for all three grass host species (Fig.
+Across more than 2300 herbarium specimens dating back to 1824, we found that prevalence of \emph{Epichloë} endophytes increased over the last two centuries for all three grass host species (Fig.
 \ref{fig:temporal}).
@@
 hyemalis} increased from $\sim 25$\% to over $50$\% prevalence.
-Our model indicates a high certainty that overall temporal trends are positive across species (99\% probability of a positive overall year slope in \emph{A.
+\linelabel{R2C33-begin}Our model indicates a high \revise{confidence}\linelabel{R2C33-end} that overall temporal trends are positive across species (99\% probability of a positive overall year slope in \emph{A.
 hyemalis}, 92\% probability of a positive overall year slope in \emph{A.
@@
 (B) Lines show predicted mean endophyte prevalence over the study period along with the 50\%  and 95\% CI bands incorporating uncertainty associated with collector and scorer random effects.
-Colored points are binned means of the observed endophyte presence/absence data (black dashes).
-Colors represent each host species and point size represents the number of specimens.}
+\revise{Trends are estimated from the endophyte prevalence model.} Colored points are binned means of the observed endophyte presence/absence data (black dashes).
+Colors represent each host species (\emph{A.
+hyemalis}: green, \emph{A.
+perennans}: orange, \emph{E.
+virginicus}: purple) and point size represents the number of specimens.}
 	\label{fig:temporal}
@@
 Collector random effects were consistently small (Fig. \ref{fig:collector_fx}), and models fit with and without this random effect provide qualitatively similar results.
-The identity of individual scorers did contribute to observed patterns in endophyte prevalence.
+The identity of individual scorers,\linelabel{R2C35-b-begin} \revise{the researchers who identified endophyte status microscopically},\linelabel{R2C35-b-end} did contribute to observed patterns in endophyte prevalence.
 For example, $3$ of the $25$ scorers were more consistently likely than average to assign positive endophyte status, as indicated by 95\% credible intervals greater than zero (Fig.
 \ref{fig:scorer_fx}).
-It is difficult to distinguish whether this was driven by true differences in scorers biases during the seed scoring process or by unintended spatial or temporal clustering of the specimens scored by each scorer \cite{clayton1993spatial,urdangarin2023evaluating}.
-By under-weighting endophyte-positive samples that are clustered spatially or by collector or observer, the INLA model is appropriately accounting for nuisance variables and providing a conservative inference of endophyte change relative to the raw data.
 
@@
 \ref{fig:svc_time_map}B).
+\linelabel{R1C7-begin}
+\revise{Broad increases in prevalence on average, along with increases towards range edges that had low historic prevalence result in range expansions of the symbiosis for both \emph{Agrostis} species (Fig.
+\ref{fig:svc_space_year_map}).
+Increases in prevalence were strongest in regions with low historic prevalence for the \emph{Agrostis} species (Fig.
+\ref{fig:initialprev_trend_plot} A-B), but for \emph{E.
+virginicus} trends did not differ according to historic prevalence (\ref{fig:initialprev_trend_plot} C).} \linelabel{R1C7-end}
 Posterior estimates of uncertainty in spatially varying slopes indicate that these hotspots of change may have experienced increases of up to $5$\% per year while declines in prevalence may be as great as $4$\% per year for \emph{A.
@@
 For \emph{E. virginicus}, uncertainty ranges between $3.5$\% increases and $2.5$\% decreases (Fig. \ref{fig:svc_time_map_CI}).
-
 \begin{figure}[H]
@@
 	\caption[Predicted posterior mean of spatially-varying slopes representing change in endophyte prevalence for each host species.]{Predicted posterior mean of spatially-varying slopes representing change in endophyte prevalence for each host species.
-Color indicates the relative change in predicted endophyte prevalence.}
+\revise{Spatially-varying trends are estimated from the endophyte prevalence model.} Color indicates the relative change in predicted endophyte prevalence.
+\revise{Map lines delineate study areas and do not necessarily depict accepted national boundaries.}}
 	\label{fig:svc_time_map}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width = \linewidth]{../Plots/svc_space_map_year.png}
+	\caption[Predicted endophyte prevalence for each host species in 1895 and 2020.]{\revise{Predicted endophyte prevalence for each host species in 1895 and 2020.
+\revise{Predictions of prevalence come from the endophyte prevalence model.} Color indicates the posterior mean endophyte prevalence for \emph{A.
+hyemalis} in (A) 1895 and (D) 2020, for \emph{A.
+perennans} in (B) 1895 and (E) 2020, and for \emph{E.
+virginicus} in (C) 1895 and (F) 2020.
+Map lines delineate study areas and do not necessarily depict accepted national boundaries.}}
+	\label{fig:svc_space_year_map}
 \end{figure}
@@
 \ref{fig:climate_correlation_plot}A,C) experienced the largest increases in endophyte prevalence.
-While our analysis identified the importance of these drivers with relatively high certainty (greater than $99$\% posterior probability of either negative or positive slopes respectively), they translate to less than $0.2$\% change in endophyte prevalence per year for a change of $100$ mm.
-change in precipitation over the century.
-Repeating this analysis using all pixels across each species' distribution were qualitatively similar to these results.
+\linelabel{R2C38-begin}\revise{Our analysis indicated relatively high confidence that these climate drivers influence endophyte prevalence shifts in \emph{E.
+virginicus}}(greater than $99$\% posterior probability of either negative or positive slopes respectively), however they translate to less than $0.2$\% change in endophyte prevalence per year for a change of $100$ mm.
+change in precipitation over the century.\linelabel{R2C38-end}
+\linelabel{R2C37-begin}Repeating this analysis using all pixels across each species' distribution were qualitatively similar to these results.\linelabel{R2C37-end}
 
@@
 	\includegraphics[width = \linewidth]{../Plots/climate_trends_plot_intercept.png}
-	\caption[Relationships between predicted trends in endophyte prevalence and changes in seasonal climate drivers.]{\textbf{Relationships between predicted trends in endophyte prevalence and changes in seasonal climate drivers.} Lines show marginal predicted relationship between spatially-varying trends in endophyte prevalence and changes in mean and variability of climate ((A): mean temperature, (B): cumulative precipitation, (C): standard deviation in temperature, (D): standard deviation in precipitation).
-Confidence bands represent the 50 and 95\% CI,  colored by host species.
+	\caption[Relationships between predicted trends in endophyte prevalence and changes in seasonal climate drivers.]{\textbf{Relationships between predicted trends in endophyte prevalence and changes in seasonal climate drivers.} Lines show marginal predicted relationship between spatially-varying trends in endophyte prevalence and changes in mean and variability of climate ((A): mean temperature, (B): cumulative precipitation, (C): standard deviation in temperature, (D): standard deviation in precipitation) \revise{estimated from the \emph{post hoc} climate regression analysis}.
+Confidence bands represent the 50 and 95\% CI,  colored by host species (\emph{A.
+hyemalis}: green, \emph{A.
+perennans}: orange, \emph{E.
+virginicus}: purple).
 Slopes with greater than 95\% probability of being either positive or negative are represented as solid lines while those that have less than 95\% probability are dashed.
-Points show 250 randomly sampled pixels across each host's distribution used in model fitting.}
+Points are the \revise{values of pre-computed SVC trends  and climate drivers at} 250 randomly sampled pixels across each host's distribution used in model fitting \revise{for the \emph{post hoc} climate regression analysis}.}
 	\label{fig:climate_correlation_plot}
@@
 
-\subsubsection*{Performance on test data}
-Tests of model's predictive performance as quantified by AUC and by visual posterior predictive checks, indicated good predictive ability. 
+\subsection*{Evaluation of model performance on an out-of-sample test}
+Tests of models' predictive performance as quantified by AUC and by visual posterior predictive checks, indicated good predictive ability. 
 Model performance was similar between historic herbarium specimens used as training data and the out-of-sample test data from contemporary surveys (AUC = 0.79 and 0.77 respectively; Fig.
@@
 \ref{fig:contemptestplot}B).
-However, model predictions for endophyte prevalence exhibited relatively little local geographic variation, whereas the out-of-sample survey data were maximally variable with populations spanning 0\% to 100\% endophyte-symbiotic plants (Fig.
+\revise{It is noteable that} model predictions for endophyte prevalence exhibited relatively little local geographic variation, whereas the out-of-sample survey data were maximally variable with populations spanning 0\% to 100\% endophyte-symbiotic plants (Fig.
 \ref{fig:contemptestplot}C).
-We interpret this to mean that the model captures coarse-scale spatial and temporal trends reasonably well, but is not equipped to capture local-scale nuances that generate population-to-population differences.
+
 %We discuss potential model improvements in the Discussion.
@@
 	\caption[Predictive performance for contemporary test data.]{\textbf{Predictive performance for contemporary test data.} (A)
-		 The model, trained on historic herbarium collection data, performed modestly at predicting prevalence in contemporary population surveys.
+		 The \revise{endophyte prevalence model}, trained on historic herbarium collection data, performed modestly at predicting prevalence in contemporary population surveys.
 The model captured regional trends across (A) longitude and (B) latitude.
-Crosses indicate predicted mean prevalence along with the 95\% CI (colored lines) from the herbarium model.
+Crosses indicate predicted mean prevalence along with the 95\% CI (colored lines: \emph{A.
+hyemalis}: green, orange, \emph{E.
+virginicus}: purple) from the herbarium model.
 Contemporary prevalence is represented by grey points (point size reflects sample size) along with trend lines from generalized linear models (black line and shaded 95\% confidence interval).
@@
 \section*{Discussion}
-Our examination of historic plant specimens revealed cryptic shifts in microbial symbiosis over the last two centuries. 
-For the three host species we examined, there have been strong increases in prevalence of fungal endophytes.
+
+\linelabel{R2C40-begin}Our examination of historic plant specimens revealed \revise{previously hidden} \linelabel{R2C40-end}shifts in microbial symbiosis over the last two centuries. 
+For the three host species we examined, there have been strong increases in prevalence of\linelabel{R2C41-begin} \revise{\emph{Epichloë} endophyte symbiosis}.\linelabel{R2C41-end}
 We interpret increases in prevalence of \emph{Epichloë}, which are vertically transmitted, as adaptive changes that improve the fitness of their hosts under increasing environmental stress.
@@
 
+\revise{\subsubsection*{Responses of host-microbe symbioses to climate change}}
 Differences across host species underscore that while all of these $C_3$ grasses share similar broad-scale distributions, each engages in unique biotic interactions and has unique responses to environmental drivers.
 We identified hotspots of change for \emph{A. perennans}, which was the species that experienced the strongest absolute changes in endophyte prevalence (Fig. \ref{fig:svc_time_map}).
-Declines of $0.9$\% per year in the southern portion of its range and increases of up to $2$\% per year in the north suggest a potential poleward range shift of endophyte-symbiotic plants (whether the overall host distribution is shifting in parallel is an exciting next question).
-Based on previous work demonstrating that endophytes can shield their hosts from drought stress (reviewed in\citet{decunta2021systematic}), we generally predicted that drought conditions would be a driver of increasing endophyte prevalence.
+Declines of $0.9$\% per year in the southern portion of its range and increases of up to $2$\% per year in the north suggest a potential poleward range shift of endophyte-symbiotic plants (whether the overall host distribution is shifting in parallel is an exciting next question.)
+Based on previous work demonstrating that endophytes can shield their hosts from drought stress (reviewed in \citet{decunta2021systematic}), we generally predicted that drought conditions would be a driver of increasing endophyte prevalence.
 In contrast to this expectation, increasing prevalence for \emph{A. perennans} were associated with increasing autumn temperature and precipitation (Fig. \ref{fig:climate_correlation_plot}). 
@@
 \ref{fig:climate_correlation_plot}B).
-For \emph{A. hyemalis}, endophytes could be playing a role helping hosts weather autumn-season droughts, which may be an important time for the species' germination.
+For \emph{A. hyemalis}, endophytes could be playing a role helping hosts weather autumn-season droughts, which \revise{is likely also} an important time for the species' germination.
 Previous work has demonstrated drought benefits in a greenhouse manipulation with this species \citep{davitt2011understanding}, and early life stages may be particularly vulnerable to prolonged droughts.
@@
 virginicus}, which experienced the most modest changes in endophte prevalence overall (ranging between $1.1$\% increases and $0.2$\% decreases), we only found modest associations with changes in climate drivers.
-Surveys by \citet{sneck2017variation}, used as part of the test data in this study, identified a drought index (SPEI) that integrates precipitation with estimated evapotranspiration as an important predictor of endophyte prevalence.
-\emph{Epichloë} endophytes have also been connected to a suite of non-drought related fitness benefits including herbivore protection \citep{brem2001epichloe}, salinity resistence \citep{wang2020effects}, and mediation of the soil microbiome \citep{roberts2015rhizosphere}.
-These effects are potentially mediated by the diverse bioactive alkaloids and other signaling compounds they produce \citep{saikkonen2013chemical}.
-Increases in symbionts could be explained, at least in part, by these diverse benefits that may help hosts weather a world made increasingly stressful by changes in climate and other anthropogenically introduced stressors.
-While we show consistent increasing trends in prevalence between the three species, the mechanisms that explain these changes may be diverse and idiosyncratic.
-%This paragraph was a little hard to read and I think does not really nail the points that I think we should make here, IMO.
-I would open to paragraph setting up the idea that the correlations you detected have some support in previous experimental studies, then elaborate.
-Then I would acknowledge that other factors can elicit benefits of symbiosis, and these factors may ALSO be changing.
-The takeaway, for me and perhaps for readers more generally, is that we are very confident endophytes are increasing, we have some confidence that is associated with (and perhaps due to) climate change, but a lot has changed over the past 200 years (climate, urbanization, nitrogen deposition, invasive species, deer!) and we need follow-up studies and experiments to gain a better mechanistic understanding of what has driven these changes, which will also allow us to build better forecasts about the future.
+Surveys by \citet{sneck2017variation}, used as part of the test data in this study, identified a drought index (SPEI) that integrates precipitation with estimated evapotranspiration as an important predictor of \revise{contemporary} endophyte prevalence \linelabel{R2C14-begin}\revise{in this species.
+The diverse relationships we detect between trends in endophyte prevalence and climate drivers suggest a more complicated picture than the simple explanation that drought alone, as measured through changes in annual precipitation, causes increasing endophyte prevalence through context-dependent fitness benefits.}
 
+\revise{While we show consistent increasing trends in prevalence between the three species, the mechanisms that explain these changes may be diverse and idiosyncratic.
+\linelabel{R2C45-begin}First, climate change responses may depend on genotype-specific responses that are ignored in our current analysis. 
+While \emph{Epichloë} symbioses are highly specialized, surveys have demonstrated genotypic and chemotypic diversity of the symbionts across populations and even within populations \cite{von2021genetic, treindl2023genetic}.
+Genotypic variation in \emph{Epichloë} endophytes, particularly in genes responsible for alkaloid production, produces ``chemotypes" with differing benefits for hosts against insect or mammalian herbivores mediated by environmental conditions \cite{schardl2012chemotypic,saikkonen2013chemical}.
+Genotypic variation of the hosts themselves can also influence interaction outcomes \cite{gundel2011interaction,parker2017genotype}.
+Whether hotspots of change in endophyte prevalence reflect selection for genotype-pairings with particularly strong fitness benefits is an unanswered question.}\linelabel{R2C45-end}
+\linelabel{R2C44-begin}Additionally, \emph{Epichloë} endophytes have been connected to a suite of non-drought related fitness benefits including herbivory \revise{defense} \citep{brem2001epichloe}, salinity resistence \citep{wang2020effects}, and mediation of pathogens \cite{vikuk2019infection} and the soil microbiome \citep{roberts2015rhizosphere}.
+\revise{Broad changes in the distribution and abundance of natural enemies, along with stresses from anthropogenic changes in landcover and  pollution \cite{sage2020global} likely influence the benefits of symbiosis \cite{rudgers2020climate}.
+The historic trends that we observed result from the combination of these fitness benefits playing out across the heterogenous map of a changing climate.\linelabel{R2C44-end}
+
+Our results indicate that \emph{Epichloë} symbiosis has likely improved host fitness in stressful environments leading to increasing prevalence. 
+\linelabel{R2C42-begin}What is less clear is how this will influence future range shifts.  
+Based on our analysis, it is likely that the symbiosis will facilitate range shifts for hosts by improving population growth at range edges.
+Previous population surveys attributed environment-dependent gradients in endophyte prevalence \citep{sneck2017variation,rudgers2009benefits,semmartin2015broad}to symbiont-derived fitness benefits allowing their hosts to persist in environments where they otherwise could not \citep{afkhami2014mutualist,kazenel2015mutualistic}.
+However, symbiont-facilitated range shifts require that endophytes be present in the populations to be able to support population growth. 
+The arid western range edge of \emph{A.
+hyemalis} has had historically low endophyte prevalence (Fig.
+\ref{fig:svc_space_year_map}), and while prevalence has increased most quickly in the regions with historically low endophyte prevalence (Fig.
+\ref{fig:initialprev_trend_plot}), the complete absence of endophytes range edges would make it impossible for prevalence to increase without dispersal of symbiotic seeds \cite{fowler2023geographic}.
+These factors potentially contribute to the ability of the host species to track its environmental niche. 
+Another interesting question is the degree to which symbiotic and non-symbiotic hosts, which occupy overlapping but distinct niches, are likely to experience distribution shifts in tandem or at different spread rates in future work.\linelabel{R2C42-end}
+More extreme climate stresses, which are expected more frequently in the future \cite{seneviratne2021weather}, have the potential to alter the costs and benefits of the interaction. 
+The past indicates a resilient interaction, but understanding the future climate conditions that may tip this interaction to deteriote will be crucial.}\linelabel{R2C14-end}
+
+\revise{\subsubsection*{Steps towards forecasts of host-microbe symbioses}}
 The combination of a spatially-explicit model and historic herbarium specimens allowed us to identify regions of both increasing and decreasing endophyte prevalence, however we see several next steps towards the goal of predicting host and symbiont niche-shifts in response to future climate change.
 While the model recreated the large-scale spatial trends observed in contemporary population surveys, test data contained more population-to-population variability in prevalence. 
-Validating our model predictions in this way, a rare extra step in collections-based studies, allows us to evaluate places to improve the model's out-of-sample predictive ability.
-Lack of information on local variability may simply be a feature of data derived from herbarium specimens. 
-They are samples from local populations, but they are single specimens that are aggregated to derive broad-scale model estimates.
+\linelabel{R2C38-begin}\revise{We interpret this to mean that the model captures coarse-scale spatial and temporal trends reasonably well, but is not equipped to capture local-scale nuances that generate population-to-population differences.} \linelabel{R2C38-end}
+Validating our model predictions \revise{with this test}, a rare extra step in collections-based studies, allows us to evaluate places to improve the model's out-of-sample predictive ability.
+\linelabel{R2C43-begin}Lack of information on local variability \revise{in symbiont prevalence} may simply be a feature of data derived from herbarium specimens. 
+\revise{Natural history collectors sample one or a few specimens from local populations, which our analyses aggregates across to derive broad-scale model estimates.}\linelabel{R2C43-end}
 This suggests that increasing local replication should be a factor considered in future collection efforts of natural history specimens, balanced with the required time and effort \linelabel{R1C9-begin} \revise{and with limitations on storage space within collections}.
@@
 Poor predictive ability at local scales in this grass-endophyte system is not surprising, as previous studies have found that local variation, even to the scale of hundreds of meters can structure endophyte-host niches \cite{kazenel2015mutualistic}.
-Other studies have found factors including land-use history \cite{vikuk2019infection} and the biotic environment, including herbivory \cite{rudgers2016long}, and host genotype \citet{sneck2017variation}, to be important predictors of endophyte ecology.
 An important step would be integrating data from local and regional scales  through modeling to constrain estimates of local and regional variation.
-Previous population surveys have found environment-dependent gradients in endophyte prevalence \citep{sneck2017variation,rudgers2009benefits,semmartin2015broad}, that may be caused by symbiont-derived fitness benefits allowing their hosts to persist in environments where they otherwise could not \citep{afkhami2014mutualist,kazenel2015mutualistic,fowler2023geographic}.
+
 Predicting future niche-shifts of hosts and symbionts will require considering the coupled dynamics of host-symbiont dispersal in addition to fitness benefits.
 For example, transplanting symbiotic and non-symbiotic plants beyond the range edge of \emph{A.
-hyemalis} could tell us whether low endophyte prevalence in that area is a result of environmental conditions that lead the symbiosis to negative fitness consequences, or is a result of some historical contingency or dispersal limitation that has thus far limited the presence of symbiotic hosts from a region where they would otherwise flourish and provide resilience.
+hyemalis} could tell us whether low endophyte prevalence in that area (Fig.
+\ref{fig:svc_space_year_map}A) is a result of environmental conditions that lead the symbiosis to negative fitness consequences, or is a result of some historical contingency or dispersal limitation that has thus far limited the presence of symbiotic hosts from a region where they would otherwise flourish and provide resilience.
 Incorporating available climatic and soil layers as covariates is another obvious step that could improve predictions.
@@
 
+\revise{\subsubsection*{Herbaria for global change research}}
 Our analysis advances the use of herbarium specimens in global change biology in two ways. 
-\linelabel{R1C7-begin}
-First and foremost, this is \revise{one of a growing number of studies to examine microbial symbiosis using specimens from natural history collections, and the first, to our knowledge, to link long-term changes in the symbioses to changes in climate.}\linelabel{R1C7-end}
-The responses of microbial symbioses are a rich target for future studies within museum specimens, particularly those that take advantage of advances in sequencing technology.
+\linelabel{R1C10-begin}
+First and foremost, this is \revise{one of a growing number of studies to examine microbial symbiosis using specimens from natural history collections, and the first, to our knowledge, to link long-term changes in the symbioses to changes in climate.}\linelabel{R1C10-end}
+The responses of microbial symbioses are a rich target for future studies within historic specimens, particularly those that take advantage of advances in sequencing technology.
 While we used relatively coarse presence/absence data based on fungal morphology, other studies have examined historic plant microbiomes using molecular sequencing and sophisticated bioinformatics techniques, but these studies have so far been limited to relatively few specimens at limited spatial extents \cite{yoshida2015computational,heberling2019utilizing, bieker2020metagenomic,gross2021hidden,bradshaw2021global}.
 Continued advances in capturing historic DNA and in filtering out potential contamination during specimen storage \citep{daru2019novel,bakker2020herbarium,raxworthy2021mining} will be imperative in the effort to scale up these efforts.
-This scaling up will be essential to be able to quantify changes not just in the prevalence of symbionts, but also in symbionts' intraspecific variation and evolutionary responses to climate change, as well as  in changes in the wider microbial community.
-Genetic variation in \emph{Epichloë} endophytes, particularly in genes responsible for alkaloid production, produces ``chemotypes" with differing benefits for hosts against insect or mammalian herbivores mediated by environmental conditions \cite{schardl2012chemotypic,saikkonen2013chemical}.
-With improved molecular insights from historic specimens, we could ask whether the broad increases in endophytes that we have identified reflect selection for particular chemotypes and how this selection varies across space.
+This scaling up will be essential to be able to quantify changes not just in the prevalence of symbionts, but also in symbionts' intraspecific variation and evolutionary responses to climate change, as well as  in changes in the wider \revise{host microbiome}.
+With improved molecular insights from historic specimens, we could ask whether the broad increases in endophytes that we have identified reflect selection for particular \revise{genetic strains or} chemotypes and how this selection varies across space.
 Answering these questions as well as the unknown questions that future researchers may ask also reiterates the value in capturing meta-information during ongoing digitization efforts at herbaria around the world and during the accession of newly collected specimens \citep{lendemer2020extended,edwards2024university}.
-Second, we accounted for several potential biases in the data observation process that may be common to many collections-based research questions by using a spatially-explicit random effects model. 
-Spatial autocorrelation \cite{willems2022forest}, potential biases introduced by the sampling habits of collectors \citep{daru2018widespread}, and variation between contemporary researchers during the collection of trait data, if not corrected for could lead to over-confident inference about the strength and direction of historic change (Fig.
+
+\linelabel{R2C46-begin}
+\revise{The second major advance in this analysis is} \linelabel{R2C46-end} in accounting for several potential biases in the data observation process that may be common to many collections-based research questions by using a spatially-explicit random effects model.
+Potential biases introduced by the sampling habits of collectors \citep{daru2018widespread}, and variation between contemporary researchers during the collection of trait data, if not corrected for could lead to over-confident inference about the strength and direction of historic change (Fig.
 \ref{fig:temporal}).
 Previous studies that have quantified the effects of collector biases typically find them to be small  \cite{davis2015herbarium,meineke2019herbarium}, and we similarly did not find that collector has a strong effect on the results of our analysis, but that scorer identity did impact results.
+\linelabel{R2C36-begin}\revise{It is difficult to distinguish whether the impact of scorers was driven by true differences in scorers' biases during the seed scoring process or by unintended spatial or temporal clustering of the specimens examined by each scorer \cite{clayton1993spatial,urdangarin2023evaluating}.
+By under-weighting endophyte-positive samples that are clustered spatially or by collector or observer, the endophyte prevalence model is appropriately accounting for nuisance variables and providing a conservative inference of endophyte change relative to the raw data.
+}\linelabel{R2C36-end}
+\linelabel{R1C4-dicussion-begin}\revise{
+Spatial autocorrelation is another phenomenom likely common in data derived from herbarium specimens \cite{willems2022forest}, which our spatially-explicit analysis models among samples.
+Beyond spatial autocorrelation of outcomes, systematic differences in sampling across space can result in spatial bias. 
+One strength of herbaria as vehicles for global change research is the relative ease with which specimens from many distinct geographic locations can be examined. 
+We visited just nine institutions in the central southern United States, and we were able to sample seeds from specimens across an area spanning  over 300,000 sq.
+km, including specimens from Mexico and Canada.
+The specimens we examined are concentrated in the South-Central United States,with fewer specimens in the rapidly warming northeastern United States.
+We provide a simulation analysis exploring the potential impact of spatially and temporally biased sampling (Appendix A - Supporting Methods). 
+We found that the spatially-varying coefficient model had a strong ability to re-capitulate temporal trends across space in simulated data, and that this result was robust to relatively high levels of spatial bias ($80$\% of data missing from region).
+Simulation analyses that extend this work to consider the myriad ways herbarium data may be biased (i.e.
+testing different spatial arrangements and scales of spatial bias, or testing different sample sizes) would be extremely valuable \cite{gaul2020data,erickson2021accounting,daru2018widespread,meineke2021bias}.
+}\linelabel{R1C4-discussion-end}
+
 %Fitting this model in a Bayesian framework allows for full propagation of uncertainty.
 
+\revise{\subsubsection*{Conclusion}}
 Ultimately, a central goal of global change biology is to generate predictive insights into the future of natural systems on a rapidly changing planet. 
@@
 	\section*{Appendix A}
-	\renewcommand{\thefigure}{A\arabic{figure}}
-	\setcounter{figure}{0}
 
-		\renewcommand{\thetable}{A\arabic{table}}
 	\setcounter{equation}{0}  % reset counter 
@@
 
+	\renewcommand{\thefigure}{A\arabic{figure}}
+
+	\renewcommand{\thetable}{A\arabic{table}}
+
+	\renewcommand{\theequation}{A\arabic{equation}}
+
+\linelabel{R2C47-begin}\revise{
+
+	\subsection*{Appendix to "Increasing Prevalence of plant-fungal symbiosis across two centuries of environmental change"}
+	   \vspace{2em}
+		\textbf{Authors:}\\
+		Joshua C. Fowler\textsuperscript{1,2*} \\
+		Jacob Moutouama\textsuperscript{1}\\
+		Tom E. X. Miller\textsuperscript{1}\\
+	\begin{flushleft}
+
+		\vspace{1em}
+		\noindent{} 1. Rice University, Department of BioSciences, Houston, Texas 77006\\
+		\noindent{} 2. University of Miami, Department of Biology, Miami, Florida\\
+
+		\noindent{} $\ast$ Corresponding author; e-mail: jcf221@miami.edu.
+
+		\vspace{8em}
+		\textbf{Contents:}\\
+		Appendix~A includes: Figure~\ref{fig:endo_status_map} - Figure \ref{fig:ELVI_climate_covariates},  Table A1, and Supporting Methods).
+
+	\end{flushleft}
+}\linelabel{R2C47-end}
+\newpage
+
+	\subsection*{Supplemental Figures}
 	\begin{figure}[H]
 		\centering
-		\includegraphics[width = \linewidth]{../Plots/endo_status_map.png}
-		\caption{\textbf{Endophyte presence/absence in specimens of each host species.} Points show collection locations colored according to whether the specimen contained endophytes ( E+; blue points) or did not contain endophytes (E-, tan points).
-To visualize temporal change, the data are faceted before and after the median year of collection.}
+		\includegraphics[width =\linewidth]{../Plots/endo_status_map.png}
+		\caption[Endophyte presence/absence in specimens of each host species]{\textbf{Endophyte presence/absence in specimens of each host species.} Points show collection locations colored according to whether the specimen contained endophytes ( E+; blue points) or did not contain endophytes (E-, tan points).
+To visualize temporal change, the data are faceted before and after the median year of collection.
+\revise{Map lines delineate study areas and do not necessarily depict accepted national boundaries.}}
 		\label{fig:endo_status_map}
@@
 		\includegraphics[width = \linewidth]{../Plots/mesh_plot.png}
-		\caption{\textbf{Triangulation mesh used to estimate spatial dependence between data points}.
+		\caption[Triangulation mesh used to estimate spatial dependence between data points]{\textbf{Triangulation mesh used to estimate spatial dependence between data points}.
 Grey lines indicate edges of triangles used to define distances between observations.
@@
 	\includegraphics[width = .5\linewidth]{../Plots/overlay_plot.png}
-	\caption{\textbf{Consistency between real data and simulated values indicate that the fitted model accurately describes the data}.
-Graph shows density curves for the observed data (black) along with with 100 simulated datasets (red).}
+	\caption[Graphical posterior predictive check of the endophyte prevalence model fit]{\revise{\textbf{Graphical posterior predictive check of \revise{the endophyte prevalence}model fit.}} Consistency between \revise{observed} data and \revise{predicted} values indicate that the fitted model accurately describes the data.
+Graph shows density curves for the observed data (black) along with 100 \revise{predicted} datasets (red).}
 			\label{fig:overlayplot}
@@
 		\includegraphics[width = .5\linewidth]{../Plots/ROC_training_plot.png}
-		\caption{\textbf{ROC plot showing model performance classifying observations according to endophyte status within the in-sample data.} The curves show adequate model performance for observed data.
+		\caption[ROC plot showing performance of the endophyte prevalence model in classifying observations according to endophyte status within the in-sample data]{\textbf{ROC plot showing performance \revise{of the endophyte prevalence model in} classifying observations according to endophyte status within the in-sample \revise{training} data \revise{from herbarium collections}.} The curves show adequate model performance for observed data.
 The AUC value is 0.79.}
@@
 		\includegraphics[width = .5\linewidth]{../Plots/ROC_test_plot.png}
-		\caption{\textbf{ROC plot showing model performance classifying observations according to endophyte status within the out-of-sample data.} The curves show adequate model performance for test data.
+		\caption[ROC plot showing performance of the endophyte prevalence model in classifying observations according to endophyte status within the out-of-sample data.]{\textbf{ROC plot showing \revise{performance of the endophyte prevalence model in} classifying observations according to endophyte status within the out-of-sample \revise{test} data \revise{from contemporary surveys}.} The curves show adequate model performance for test data.
 The AUC value is 0.77.}
@@
 	\includegraphics[width = \linewidth]{../Plots/posterior_plot.png}
-	\caption[Posterior estimates of parameters describing global temporal trends in endophyte prevalence trends.]{Density curves show the probability density along with mean (colored line) and $95$\% CI (black lines) for the (A) intercept and (B) slope terms, \textbf{A} and \textbf{T} respectively.
+	\caption[Posterior estimates of parameters describing global temporal trends in endophyte prevalence trends.]{\revise{\textbf{Posterior estimates of parameters describing global intercept and temporal trends from the endophyte prevalence model.}} Density curves show the probability density along with mean (colored line) and $95$\% CI (black lines) for the (A) intercept and (B) slope terms, \textbf{A} and \textbf{T} respectively \revise{from Eqn.
+\ref{eq:trends}}.
 Colors represent each host species}
@@
 	\includegraphics[width = \linewidth]{../Plots/svc_time_map_CI.png}
-	\caption[Uncertainty in spatially-varying trends representing change in endophyte prevalence for each host species.]{Shading represents the range of the 95\% posterior credible interval for spatially varying slopes, $\tau$.}
+	\caption[Credible interval width of temporal trends in endophyte prevalence across the distribution of each host species estimated from the endophyte prevalence model.]{\revise{\textbf{Credible interval width of temporal trends in endophyte prevalence across the distribution of each host species estimated from the endophyte prevalence model.}} Shading represents the range of the 95\% posterior credible interval given in units of \emph{\% change in prevalence/year} for spatially varying slopes, $\tau$ \revise{from Eqn.
+\ref{eq:trends}}.
+\revise{Map lines delineate study areas and do not necessarily depict accepted national boundaries.}}
 	\label{fig:svc_time_map_CI}
@@
 	\includegraphics[width = \linewidth]{../Plots/conservative_comparison_plot.png}
-	\caption{Comparison of liberal versus conservative endophyte scores on modeled outcomes.
-(A) Posterior estimates of global temporal trend for models fit to liberal scores (grey) and to conservative scores (blue).
-Maps show the spatially varying temporal trend estimates from model fit to conservative scores for (B) \emph{A.
+	\caption[Comparison of parameter estimates from endophyte prevalence model fit to data with liberal versus conservative endophyte scores]{\revise{\textbf{Comparison of endophyte prevalence model estimates fit to data with liberal versus conservative endophyte scores.} Liberal and conservative scores document uncertainty in the endophyte identification process.
+Each specimen was given both a liberal and conservative scores.
+In cases of uncertain identification, the liberal status assumed a potential endophyte identification was more likely to be endophyte-positive while the conservative status assumed that the potential endophyte identification was less likely to be endophyte-positive}.
+(A) Posterior estimates of global temporal trend \revise{(\textbf{T} from Eqn.
+\ref{eq:trends}}) for the endophyte prevalence model fit to liberal scores (grey) and to conservative scores (blue).
+Maps show the spatially varying temporal trend estimates \revise{($\tau$ \revise{from Eqn.
+\ref{eq:trends}})} from the endophyte prevalence model fit to conservative scores for (B) \emph{A.
 hyemalis}, (C) \emph{A.
@@
 Note that the color scale differs between this visualization and Fig.
-\ref{fig:svc_time_map}.}
+\ref{fig:svc_time_map} \revise{that shows estimates fit using liberal endophyte scores.}}
 	\label{fig:conservative_scores_plot}
@@
 	\includegraphics[width = .6\linewidth]{../Plots/collector_posterior.png}
-	\caption{\textbf{Posterior estimates of collector random effects.} Points show posterior median along with 95\% CI for each of $924$ individual collectors.}
+	\caption[Posterior estimates of collector random effects from endophyte prevalence model]{\textbf{Posterior estimates of collector random effects \revise{from endophyte prevalence model.}} \revise{Collector random effects are denoted $\chi$ in Eqn.
+\ref{eq:trends} and represent variance associated with researchers who collected historic herbarium specimens.} Points show posterior median along with 95\% CI for each of $924$ individual collectors.}
 		\label{fig:collector_fx}
@@
 	\includegraphics[width = .6\linewidth]{../Plots/scorer_posterior.png}
-	\caption{\textbf{Posterior estimates of scorer random effects.} Points show posterior median along with 95\% CI for each of $25$ individual collectors.}
+	\caption[Posterior estimates of scorer random effects from endophyte prevalence model]{\textbf{Posterior estimates of scorer random effects \revise{from endophyte prevalence model.}} \revise{Scorer random effects are denoted $\omega$ in Eqn.
+\ref{eq:trends} and represent variance associated with researchers who identified \emph{Epichloë} endophytes within herbarium specimen tissue samples.} Points show posterior median along with 95\% CI for each of $25$ individual \revise{scorers}.}
 	\label{fig:scorer_fx}
@@
 	\centering
+	\includegraphics[width = \linewidth]{../Plots/initialprev_trend_plot.png}
+	\caption[Relationship between initial prevalence and temporal trends in prevalence estimated from the endophyte prevalence model]{\textbf{Relationship between initial prevalence and temporal trends in prevalence estimated from the endophyte prevalence model.} Points show predicted posterior mean temporal trend for each species at pixels across each host distribution ((A) \emph{A.
+hyemalis}, (B) \emph{A.
+perennans}, and (C) \emph{E.
+virginicus}).
+along with a linear regression and shaded ribbon showing 95\% confidence interval.}
+	\label{fig:initialprev_trend_plot}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
 	\includegraphics[width = \linewidth]{../Plots/contemp_surveys_map.png}
-	\caption{\textbf{Locations of contemporary surveys of endophytes in \emph{A. hyemalis} used as "test" data (red points), relative to the historical collection data (black crosses).}}
+	\caption[Locations of contemporary surveys of endophytes used as "test" data to evvaluate predictive ability of the endophyte prevalence model]{\textbf{Locations of contemporary surveys of endophytes used as "test" data to \revise{evaluate predictive ability of the endophyte prevalence model.}} \revise{Points are locations of host populations surveyed between 2013 and 2019 for endophytes, colored by species (\emph{A.
+hyemalis}: green, \emph{E.
+virginicus}: purple).
+Black crosses show the historical herbarium collection locations used as "training" data for the endophyte prevalence model.}}
 	\label{fig:contempsurveysmap}
@@
 	\includegraphics[width = .8\linewidth]{../Plots/AGHY_climate_change_plot.png}
-	\caption{\textbf{Change in seasonal climate variables between the periods 1895-1925 and 1990-2020.} Color represents change in (A) seasonal temperature and (B) seasonal precipitation.
+	\caption[Change in seasonal climate variables between the periods 1895-1925 and 1990-2020 across the distribution of A.
+hyemalis]{\textbf{Change in seasonal climate variables between the periods 1895-1925 and 1990-2020 \revise{across the distribution of \emph{A.
+hyemalis}}.} Color represents change in (A) seasonal temperature \revise{($^o$C)} and (B) seasonal precipitation \revise{(mm.)}.
 Maps show pixels covering the modeled distribution of \emph{A.
-hyemalis} used in post-hoc climate correlation analysis.}
+hyemalis} used in \revise{\emph{post hoc} climate regression analysis.}}
 	\label{fig:AGHY_climate_covariates}
@@
 	\includegraphics[width = .8\linewidth]{../Plots/AGPE_climate_change_plot.png}
-	\caption{\textbf{Change in seasonal climate variables between the periods 1895-1925 and 1990-2020.} Color represents change in (A) seasonal temperature and (B) seasonal precipitation.
+	\caption[Change in seasonal climate variables between the periods 1895-1925 and 1990-2020 across the distribution of A.
+perennans]{\textbf{Change in seasonal climate variables between the periods 1895-1925 and 1990-2020 \revise{across the distribution of \emph{A.
+perennans}}.} Color represents change in (A) seasonal temperature \revise{($^o$C)} and (B) seasonal precipitation \revise{(mm.)}.
 Maps show pixels covering the modeled distribution of \emph{A.
-perennans} used in post-hoc climate correlation analysis.}
+perennans} used in \revise{\emph{post hoc} climate regression analysis.}}
 	\label{fig:AGPE_climate_covariates}
@@
 	\includegraphics[width = .8\linewidth]{../Plots/ELVI_climate_change_plot.png}
-	\caption{\textbf{Change in seasonal climate variables between the periods 1895-1925 and 1990-2020.} Color represents change in (A) seasonal temperature and (B) seasonal precipitation.
+	\caption[Change in seasonal climate variables between the periods 1895-1925 and 1990-2020 across distribution of \emph{E.
+virginicus}]{\textbf{Change in seasonal climate variables between the periods 1895-1925 and 1990-2020 \revise{across the distribution of \emph{E.
+virginicus}}.} Color represents change in (A) seasonal temperature \revise{($^o$C)} and (B) seasonal precipitation \revise{(mm.)}.
 Maps show pixels covering the modeled distribution of \emph{E.
-virginicus} used in post-hoc climate correlation analysis.}
+virginicus} used in \revise{\emph{post hoc} climate regression analysis.}}
 	\label{fig:ELVI_climate_covariates}
@@
 
+\newpage
+
 	\begin{table}[h]
-		\caption{Summary of herbarium samples across collections}
+		\caption{Summary of herbarium samples across collections \revise{(no. of specimens)}}
 		\label{table:herbaria}
@@
 		\begin{tabular}{llll}\hline
-			Herbarium Collection        & AGHY        & AGPE      &      ELVI\\ \hline
+			Herbarium Collection        & \emph{A. hyemalis}        & \emph{A. perennans}      &      \emph{E. virginicus}\\ \hline
 			Botanical Research Institute of Texas &   350   &    190&    198    \\
@@
 	\end{table}
+
 	\section*{Supporting Methods}
@@
 
-\subsection*{Mesh and Prior Sensitivity analysis}\label{sec:priorsensitivity}
+\subsection*{Mesh and Prior Sensitivity Analysis}\label{sec:priorsensitivity}
 To test the influence that the triangulation mesh and choice of priors has on results, we compared model results across a range of meshes and priors.
@@
 	\includegraphics[width = \linewidth]{../Plots/prior_comparison_year_plot.png}
-	\caption[Overall trend in endophyte prevalence evaluated for models with different range priors on spatially structured random effects, and for two different meshes]{\textbf{Overall trend in endophyte prevalence evaluated for models with different range priors on spatially structured random effects, and for two different meshes.} Note that these plots, as compared to Fig.
-2 in main text, show mean trends and do not incorporate prediction uncertainty associated with collector and scorer random effects.}
+	\caption[Overall trend in endophyte prevalence evaluated for endophyte prevalence models with different range priors on spatially structured random effects, and for two different meshes]{\textbf{Overall trend in endophyte prevalence evaluated for \revise{endophyte prevalence} models with different range priors on spatially structured random effects, and for two different \revise{triangulation} meshes.} \revise{Data used in model fitting is the same across all panels and as in the main text}.
+Note that these plots, as compared to Fig.
+2 in main text, show mean trends and do not incorporate \revise{variance} associated with collector and scorer random effects.}
 	\label{fig:prior_year_plot}
@@
 	\includegraphics[width = .8\linewidth]{../Plots/standard_mesh_comparison_svc_plot.png}
-	\caption[Spatially-varying trends in endophyte prevalence evaluated for models with different range priors on spatially structured random effects, and for the "standard" mesh]{\textbf{Spatially-varying trends in endophyte prevalence evaluated for models with different range priors on spatially structured random effects, and for the "standard" mesh.} Shading indicates the magnitude and direction of predicted trends for each of three host species for each of three prior ranges (rows A-C).
-Note that each plot has an individual scale bars.}
+	\caption[Spatially-varying trends in endophyte prevalence evaluated for models with different range priors on spatially structured random effects, and for the "standard" mesh]{\textbf{Spatially-varying trends in endophyte prevalence evaluated for \revise{the endophyte prevalence} model with different range priors on spatially structured random effects, and for the "standard" mesh.} \revise{Data used in model fitting is the same across all panels and as in the main text}.
+Shading indicates the magnitude and direction of predicted trends for each of three host species for each of three prior ranges (rows A-C).
+Note that each plot has an individual scale bar.}
 	\label{fig:standard_comparison_svc_plot}
@@
 	\includegraphics[width = .8\linewidth]{../Plots/finer_mesh_comparison_svc_plot.png}
-	\caption[Spatially-varying trends in endophyte prevalence evaluated for models with different range priors on spatially structured random effects, and for the "finer" mesh]{\textbf{Spatially-varying trends in endophyte prevalence evaluated for models with different range priors on spatially structured random effects, and for the "finer" mesh.} Shading indicates the magnitude and direction of predicted trends for each of three host species for each of three prior ranges (rows A-C).
-Note that each plot has an individual scale bars.}
+	\caption[Spatially-varying trends in endophyte prevalence evaluated for the endophyte prevalence model with different range priors on spatially structured random effects, and for the "finer" mesh]{\textbf{Spatially-varying trends in endophyte prevalence evaluated for \revise{the endophyte prevalence} model with different range priors on spatially structured random effects, and for the "finer" mesh.} \revise{Data used in model fitting is the same across all panels and as in the main text}.
+Shading indicates the magnitude and direction of predicted trends for each of three host species for each of three prior ranges (rows A-C).
+Note that each plot has an individual scale bar.}
 	\label{fig:finer_comparison_svc_plot}
 \end{figure}
+\newpage
+
+\subsection*{Spatially-biased Sample Size Simulation Analysis}\label{sec:samplingsensitivity}
+\linelabel{R1C4-begin}
+\revise{To examine how data that is unevenly distributed across host distributions may influence interpretation of spatially-varying coefficients, we performed a simulation analysis. 
+Our focal species, \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}, are widely distributed grasses across the eastern United States that host \emph{Epichloë} fungal endophytes.
+For logistical reasons, our sampling visits to herbaria focused on herbaria in the central southern U.S., which resulted in unevenly distributed data across each host species' range. 
+This is particularly noteable for \emph{Agrostis perennans} which has the most northern distribution and relatively fewer total collected specimens compared to the other focal species. 
+Thus, a significant portion in the northeast of this species' range is relatively sparsely sampled. 
+Our analysis presented in the main text identified this region as having strong increase in endophyte prevalence. 
+Future visits to herbaria with regional focuses in the Northeastern US would certainly garner new specimens that could provide valuable insights into shifting host and symbiont distributions.
+
+\subsubsection*{Simulation of spatially-biased symbiont occurence data}
+We simulated datasets with varying levels of missing-ness to examine how this missing-ness influenced the estimation of spatially-varying trend estimates. 
+We first generated $300$ data points for each of three hypothetical species at random positions across an area approximating the scale of our focal data.
+Each data point was randomly assigned a year of collection across 200 years.
+We then simulated data from a Bernoulli process with trends alternating across nine regions (Fig. \ref{fig:sample_size_fulldata_plot}) in a 3X3 grid pattern.
+This grid pattern was intended to create a complex spatial layout of trends, where trends were either an increase of $1$\% per year or a decrease of $1$\% per year.
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width = \linewidth]{../Plots/sample_size_fulldata_plot.png}
+	\caption[Full simulated dataset of symbiotic association with spatially-varying temporal trends]{\textbf{Full simulated dataset of symbiotic association with spatially-varying temporal trends.} Color indicates the slope parameter used to simulate trends in endophyte status across nine "regions" for three species.
+Data are assigned collection years across a period of 200 years.
+Shape indicates the presence (1) or absence (0) of a symbiont.}
+	\label{fig:sample_size_fulldata_plot}
+\end{figure}
+
+From this full data, we generated six additional datasets with missing-ness in the northeast region of the simulated data for hypothetical species 2.
+The data remained the same for Species 1 and for species 3 across all datasets.
+For these six datasets, we removed data points at random in six ways: 0\% of datapoints in northeast region, 0\% of  recent datapoints, only 20\% of datapoints, only 20\% of recent datapoints, only 50\% of datapoints, and only 50\% of recent datapoints (Fig.
+\ref{fig:sample_size_species2data_plot}).
+We define the datapoints as part of the recent time period if they occur later than the median year.
+The result is 6 scenarios exploring degrees of spatial and temporal bias.
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width = .8\linewidth]{../Plots/sample_size_species2data_plot.png}
+	\caption[Six simulated datasets representing scenarios of spatially-baised missingness for Species 2]{\textbf{Six simulated datasets representing scenarios of spatially-baised missingness for Species 2.} Missingness was imposed in the northeast region for six scenarios: 0\% of recent datapoints available (a,d); only 20\% of recent datapoints (b,e); only 50\% of recent datapoints (c,f); 0\% of datapoints across the full time period available (g,j); only 20\% of datapoints across the full time period (h,k); and only 50\% of datapoints across the full time period(i,l).
+Missingness was imposed only for hypothetical Species 2; Species 1 and 3 remain as in Figure A19.
+Color indicates the slope parameter used to simulate trends in endophyte status across 9 regions in a 3x3 grid.
+Shape indicates the presence (1) or absence (0) of a symbiont.
+}
+	\label{fig:sample_size_species2data_plot}
+\end{figure}
+
+\subsubsection*{Statistical analysis}
+	\setcounter{equation}{0}
+
+\renewcommand{\theequation}{A\arabic{equation}}
+
+We analyzed each dataset with a model given by Eqn. \ref{eq:sim_trends} similar in construction to that used in our central analysis.
+
+\begin{equation}
+	\label{eq:sim_trends}
+	logit(\hat{P}_{h,i}) = \boldsymbol{A}_{h} + \boldsymbol{T}_{h}*year_i  + \alpha_{h,l_{i}} + \tau_{h,l_i}*year_i + \delta_{l_i}
+\end{equation}
+
+Where symbiont presence/absence of the $i^{th}$ specimen ($P_{h,i}$) was modeled as a Bernoulli response variable with expected probability of symbiont occurrence $\hat{P}_{h,i}$ for each host species $h$.
+We modeled $\hat{P}_{h,i}$ as a linear function of intercept $\boldsymbol{A}_h$ and slope $\boldsymbol{T}_{h}$ defining the global trend in endophyte prevalence specific to each host species as well as with spatially-varying intercepts $\alpha_{h,l_{i}} $ and slopes $\tau_{h,l_i}$ associated with location ($l_i$, the unique latitude-longitude combination of the $i$th observation).
+Similar to the SVC model of our central analysis (Eqn.
+\ref{eq:trends}), we estimated a shared variance term with the spatially-dependent random effect $\delta_{l_i}$, intended to account for residual spatial variation.
+However in this analysis we omit i.i.d.-random effects terms associated with collector and scorer identity ($\chi_{c_i}$ and $\omega_{s_i}$ in Eqn. \ref{eq:trends}) for the sake of simplicity.
+
+\subsubsection*{Influence of spatially-biased sampling on model interpretation}
+Our analysis of the full simulated data shows that our model is suitably flexible to capture complex spatial patterns in temporal trends (Fig. \ref{fig:sim_svc_plot} a-c).
+Beyond this,  the model also qualitatively captures the spatial patterns in temporal trends even with large amounts of data missingness (i.e missing up to $80$\% of the datapoints (Fig.
+\ref{fig:sim_svc_plot} p-r)).
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width = .7\linewidth]{../Plots/sim_svc_plot.png}
+	\caption[Mean predicted spatially-varying trend in symbiont prevalence across datasets with different levels of missingness]{\textbf{Mean predicted spatially-varying trend in symbiont prevalence across datasets with different levels of missingness.} Color indicates the estimated mean temporal trend within each pixel across the simulated data.
+Panels show estimates for models fit to different levels of missing data for species 2 in the northeast region ((a-c) the full dataset, (d-f) missing all datapoints across entire temporal period, (g-i) missing all datapoints only during the recent period, (j-l) missing 80\% of the datapoints  only during the recent period, (m-o)  missing 50\% of the datapoints  only during the recent period, (p-r) missing 80\% of the datapoints across the entire temporal period, (s-u) missing 50\% of the datapoints across the entire temporal period).
+The mesh boundary that bounds the "full" simulated dataset is plotted in each panel.}
+	\label{fig:sim_svc_plot}
+\end{figure}
+
+While this analysis is not an exhaustive examination of the influence of sampling bias on our results for several reasons, including not examining how different strengths in temporal trends, different spatial arrangments of missing-ness influence model estimates, or different sample sizes, it demonstrates that the spatially-varying modelling framework implemented in INLA we emply can suitably recover regional trends even with significant spatially-bias within data collection, and further the analysis is likely robust to temporally-structured bias (missing data within recent collection period).
+Future work could more fully explore the scenarios that cause this ability to break down. 
+We expect this simulation reflects what may be a common scenario for research investigating global change using natural history specimens. 
+Collection effort by trained taxonomists and professional collectors peaked in the past, and collections contain relatively fewer modern specimens in many regions. 
+Additionally, most global change research necessarily involves accessing many specimens across collections.
+Research efforts such as ours will be unable to access every specimen from all possible collections. 
+Ongoing digitization efforts will make it possible to more clearly assess how much data is missing from a particular study compared to the actual holdings of natural history collections, but ultimately, the decision of what data and collections to include is a question of sample size and study design.
+ \linelabel{R1C4-end}
+}
 

### 2025-01-23 Manuscript/Reviewer_response_endo_herbarium.tex
first round pass through all of reviewer comments

@@
 In response to both reviewers, we have added substantial discussion of the ecological implications for hosts of our finding that symbiosis with \emph{Epichloë} fungal endophytes has increased in prevalence across the last two centuries.
-We have also made significant revisions throughout the introduction, methods, results, and discussion to improve the organization and clarity of our ideas in response to Reviewer 2's recommendations.
-In addition, the utility of our supplemental material has been greatly improved in response to Reviewer 2's recommendations for more clearly captioning supplemental figures and their connections to our main results.
+Additionally, we have added a completely new simulation analysis evaluating the impact that spatially-biased sampling may have on our analysis, as prompted by Reviewer 1, presented in an expanded Supplemental Methods.
+We have also made significant revisions throughout the manuscript to improve the organization and clarity of our ideas in response to reviewer recommendations.   
+Finally, we have thoroughly revised the Appendix in response to Reviewer 2's recommendations, which has improved the utility of our supplemental materials and connections to our main results.
 
@@
 We have added additional description of genotypic diversity within the interaction to provide readers with this context in our methods section.
-In addition, we return to this topic in expanded discussion material (Lines XXX)}
+In addition, we return to this topic in expanded discussion material}
 \linesref{ms-R1C3-begin}{ms-R1C3-end}
+\linesref{ms-R2C14-begin}{ms-R2C14-end}
 
@@
 (3) If not, I at least recommend adding a caveat somewhere in the methods that collections were included from across eastern US, but were primarily collected from the four states in which herbaria were visited and that this has the potential to bias your results in specific ways (also see line comment on L504-507).}
-\response{}
+\response{The herbarium visits we completed for this analysis do result in fewer specimens from the northeastern portion of each species ranges, and \emph{A.
+perennans} is the least commonly collected species among our focal hosts.
+We are unable to perform any additional sampling. 
+However, we feel our methodology is robust. 
+To demonstrate this, and in response to this comment, we have performed a simulation analysis (\emph{Appendix A - Influence of spatially-biased sampling on model interpretation}) that shows the ability of an equivalent model to recover spatially-varying temporal trends within a region with up to $80$\% of data missing.
+
+Beyond this, we also are reassured when examining the Bayesian posterior credible interval of key results, such as the estimated temporal trend (Fig.
+\ref{ms-fig:svc_time_map_CI}), across the  host distributions.
+For example, we find that for \emph{A.
+perennans}, the width of this credible interval is consistently small (suggesting that estimated trends vary by only $\mp$ 2 percentage points per year) within areas with many samples (the midwest), while within areas with fewer samples (the northeast) this interval is wider but still relative small (trends are estimated to be $\mp$ 3 percentage points per year), suggesting that the trend in this region is highly likely to be in a positive direction.
+Certainly we expect more thorough sampling to be valueable.
+We have added consideration of this and other potential biases to our discussion, along with the methods caveat described by this reviewer.}
+\linesref{ms-R1C11-begin}{ms-R1C11-end} methods
+\linesref{ms-R1C4-dicussion-begin}{ms-R1C4-discussion-end} discussion
+\linesref{ms-R1C4-begin}{ms-R1C4-end} Appendix A - Supporting Methods
 
@@
 }
-\response{These are an interesting questions.
-We have added supplemental figures which present the marginal spatial intercepts, along with predicted prevalence across space in 1895 and in 2020.
-As suggested by the reviewer, we also assessed the relationship between initial prevalence and predicted changes in prevalence.
-To keep the main text focused on temporal change in prevalence, we placed these figures in the supplemental material, but have added text describing this result.}
-\linesref{arg1}{arg2}
+\response{We agree that these are interesting questions.
+We have added additional figures which present the predicted prevalence across host distributions in 1895 and in 2020 (Figure 4).
+% I haven't taken the time to figure out how to get ref to work correctly here.
+They show clear range expansion by the symbiosis towards the western range edge of \emph{A.
+hyemalis} and towards the northern range edge of \emph{A.
+perennans}.
+As suggested by the reviewer, we also assessed the relationship between initial prevalence and predicted changes in prevalence, which shows that for these species the largest increases do occur at locations with low starting prevalence.
+To keep the main text focused on temporal change in prevalence, we placed a figure showing this result in the supplemental material (Figure ), but have added text describing this to our results.}
+\linesref{ms-R1C7-begin}{ms-R1C7-end}
 
 \comment{Fig. 4: Text size very difficult to read currently.}
-\response{}
+\response{We have re-drawn this figure to improve readability. (Now Figure 5)}
 
@@
 In response to this comment, we have rephrased this statement to clarify that we see novelty particularly in connecting microbial symbiosis to climate change drivers, and to temper our assertion of novelty.}
-\linesref{ms-R1C7-begin}{ms-R1C7-end}
+\linesref{ms-R1C10-begin}{ms-R1C10-end}
 
@@
 The biases listed by the authors here are correct, but I would also like to see them acknowledge how their own analysis contains certain biases that were not necessarily captured by the model and sampling structure they employed.}
-\response{}
+\response{In response to this comment and Comment 4, we have added context to our methods describing the uneven sampling of specimens across states, as well as material to our discussion that expands on the potential influence of these spatial biases in interpreting our results.}
+\linesref{ms-R1C11-begin}{ms-R1C11-end}
 
@@
 Not just about the prevalence of the relationship (which they cover thoroughly), but about what this might mean for grass performance at the population or even community scales under ongoing climate change.}
-\response{}
+\response{Thank you for pushing us to connect our work more directly to the biology of our system.
+This was a comment echoed by Reviewer 2 (Comment 14).
+We had added substantial material our discussion that presents the key implications for population and range dynamics under climate change.}
+\linesref{ms-R2C14-begin}{ms-R2C14-end}
 
-\section{Response to Reviewers 2}
+\section{Response to Reviewer 2}
 \vspace{-2em}
@@
 In addition, their desire to see more in-depth discussion of the ecological implications of our findings was echoed by Reviewer 1, and we have added new discussion material addressing this topic, detailed in response to specific comments.}
-\linesref{arg1}{arg2} TO BE DONE
+\linesref{ms-R2C14-begin}{ms-R2C14-end}
 
@@
 
-\comment{Discussion: he Discussion is the most difficult section to follow.
+\comment{Discussion: The Discussion is the most difficult section to follow.
 Similarly, to the Introduction, I back-and-forth between paragraphs/arguments.
@@
 For example, you could begin by discussing about the spatial/temporal patterns in prevalence, followed by the relationships with climatic drivers before to conclude with the technical/ecological novelty of your findings.}
-\response{}
+\response{In response to this comment, we have added sub-headings to our Discussion and revised our discussion section for clarity and flow between ideas.
+Many of these changes are detailed in response to further comments from Reviewer 2}
 
@@
 To assist readers (who would otherwise need to juggle between the supplementary figures and the main text), please ensure that each figure’s caption includes: the specific model or analysis from which it was produced; the data used (e.g., whether it is based on all host plants or a particular plant, which period of data used, etc.); and defining terms or acronyms of these figures (examples: A and T from figure A6; $\tau$ from A7; conservative vs liberal A8; AGHY vs ELVI A11…..).}
-\response{}
+\response{We appreciate Reviewer 2's attention to our supplemental material.
+We have revised through all of the figure captions in the supplement to provide this information.
+These edits are also detailed in our responses to specific comments further down.}
 
 \comment{L54: Which physical legacies or ecological processes? Provide example.}
-\response{}
+\response{In response to this comment, we have added additional information about the examples of ecological processes studied previously using herbarium specimens.}
+\linesref{ms-R2C19-begin}{ms-R2C19-end}
 
 \comment{L64-L74: Please remove example of symbiosis from other systems and focus on your targeted endophytic symbiosis. (Remove corals, insects or lichens.)}
-\response{}
+\response{We strongly feel that these other symbioses provide relevant examples of the potential responses of symbioses to global change.
+Our central point is that this response is known in a few key symbiota, but not in the endophytic symbiosis that is our focus, and that studies across numerous taxa are valueable to understand \emph{why} the response tips from resilience to disruption of the interaction.
+These examples motivate our hypotheses and so we have opted to keep them in the text.}
 
 \comment{-L104: In which sense? How this hypothesis would be reflected in your results? Would you expect a distribution shifts? Please specify.}
-\response{}
+\response{We have edited this sentence to include a specific prediction for how this hypothesis would relate to our study system.
+We additionally have added additional specific predictions two paragraphs further down, where we present our study's questions.}
+\linesref{ms-R2C21-begin}{ms-R2C21-end}
 
 \comment{-L131-134: Super nice data!}
-\response{}
+\response{We thank this Reviewer 2 for acknowledging the value of our dataset which represents significant time and effort.}
 
@@
 \comment{-L167-168: What is the spatial resolution of the different units? What could be the distance between counties or between state centroids. And how big or small county can be?}
-\response{}
+\response{In response to this comment, we have added information about the relative size of the bounding boxes generated during the geocoding process and the pairwise distance between points.
+Because we included coordinates georeferenced to local municipalities, counties, and in a few instances, states, the width of this bounding box varied from less than 1 km.
+to over 1233 km (median 841 km.).}
+\linesref{ms-R2C24-begin}{ms-R2C24-end}
 
 \comment{-L176: “Partial” would be more adapted than “imperfect”.}
-\response{}
+\response{"Imperfect" transmission is a commonly use term within the study of symbiosis therefore we have opted to retain its use.
+We have however also added a definition of the term "imperfect transmission" at the point it is first used.}
+\linesref{ms-R2C25-begin}{ms-R2C25-end}
 
@@
 Please reformulate and clarify in the main text and maybe it can be used in the discussion?}
-\response{}
+\response{We are sorry that this terminology has caused confusion.
+In response to this feedback, we have edited for to provide additional explanation.
+The central point is that we mark two statuses to evaluate uncertainty in the process of assigning an endophyte identification under the microscope for each specimen.
+In cases with clear identifications, the two scores are the same.
+When there is uncertainty in identification, we err on the side of assigning positive for one score and err on the side of assigning negative for the other.
+We present a comparison of results based on this scoring choice in our supplement, which showed minimal difference, and therfore we only present the results of the "liberal" score for the rest of the central results.}
+\linesref{ms-R2C26-begin}{ms-R2C26-end}
 
 \comment{-L190: Spatial resolution of the aggregation?}
-\response{}
+\response{Our global slope estimates a trend averaging across the entire study region.
+We have changed this to "averaged" instead of "aggregated".
+We have added additional information elsewhere about the spatial resolution of our analysis in response to Reviewer 2's subsequent comments.}
+\linesref{ms-R2C27-begin}{ms-R2C27-end}
 
 \comment{-L215: “borrow information” is more appropriate than “borrow strength”.}
-\response{}
+\response{Thank you for pointing this out. We have made this change.}
+\linesref{ms-R2C28-begin}{ms-R2C28-end}
 
 \comment{-L229: Why 5? Is there a reference?}
-\response{}
+\response{We have added a reference supporting the use of penalized complexity priors, and revised terminology to more closely reflect the parameter values used to implement the model in inlabru.}
+\linesref{ms-R2C29-begin}{ms-R2C29-end}
 
@@
 I feel like this paragraph is coming out of the blue and is making the full material and methods confusing.}
-\response{}
+\response{We appreciate the push to improve the organization of our manuscript so that it can be readily understood.
+In this case, we see the current placement of the section "Modeling distributions of host species" as appropriate because the preceeding section describes models of endophyte prevalence that do not rely on these models of host species.
+The predictions from models of host species are used principally for generating pixels with which we test the relationship between climate drivers and trends in endophyte prevalence and for visualizing relationships across study region.
+However, in response to this comment, we have added additional context describing this role that the host distribution models play.}
+\linesref{ms-R2C30-begin}{ms-R2C30-end}
 
 \comment{-L281: “which” instead of “whether”?}
-\response{}
+\response{This change would alter the meaning of the question posed in this sentence, which we feel is unnecessary. We have opted to keep "whether".}
 
 \comment{-L285-288: Spatial resolution?}
-\response{}
+\response{We now provide the spatial resolution of the pixels generated. We have also added added explanation that the choice of pixel number/size is arbitrary, relevant to Reviewer 2 Comment 37.}
+\linesref{ms-R2C32-begin}{ms-R2C32-end}
 
 \comment{-L316: “Confidence” instead of “certainty”}
-\response{}
+\response{We have made this change. Thank you.}
+\linesref{ms-R2C33-begin}{ms-R2C33-end}
 
 \comment{-Figure 2: Very nice figure.}
-\response{}
+\response{We thank reviewer 2 for this supportive comment.}
 
@@
 They are quite contradictory no?}
-\response{lines \ref{}}
+\response{Reviewer 2 confuses 'collector', those researchers who collected herbarium specimens, and who we term 'scorer', those members of our lab who performed the microscope work to identify \emph{Epichloë} fungal endophytes.
+Aside from this confusion, there is no contradiction in our results; we found that collector identify did not influence model predictions, while scorer identity did influence model predictions, as evidenced in Figs.
+A9-A10.
+In response to this comment, we have more clearly defined 'scorer' in our manuscript.}
+\linesref{ms-R2C35-begin}{ms-R2C35-end}
+\linesref{ms-R2C35-b-begin}{ms-R2C35-b-end}
 
 \comment{L328-333: This section should be in discussion rather than in results no?.}
-\response{lines \ref{}}
+\response{We have moved this interpration of the impact of scorer and collector random effects into our discussion section.}
+\linesref{ms-R2C36-begin}{ms-R2C36-end}
 
@@
 And what are the spatial resolution of these pixels?}
-\response{lines \ref{}}
+\response{ Our model uses a continuous spatial process to model variation across space in the estimated trends in endophyte prevalence.
+This permits us to generate predictions from the model at any location across the study region at high resolution, however doing so does require longer computation time.
+Models fit to the full set of pixels for each species show climate regressions in a similar direction as shown here and have narrower credible intervals.
+We could, for example, generate many thousands of points across each distribution, but this would make the credible intervals largely meaningless.
+We have added material describing why the choice of pixel dimensions is arbitrary as described in response to Reviewer 2 Comment 32.
+We also note in our results section that these relationship do not differ when using all points (Line \ref{ms-R2C37-begin}).
+In addition, we have edited the figure caption to include information about the origin of model (Now Figure 5).}
+\linesref{ms-R2C37-begin}{ms-R2C37-end}
+\linesref{ms-R2C32-begin}{ms-R2C32-end}
 
 \comment{- L382-386: This should be in discussion rather than in results.}
-\response{lines \ref{}}
+\response{In response to this comment, we have edited the sentence referenced in this comment to remove interpretation and focus solely on results.
+We feel the sentence referenced in this comment is appropriately placed in that it presents the magnitude of the climate relationship for \emph{E.
+virginicus}, similarly to each of the other two host species.}
+\linesref{ms-R2C38-begin}{ms-R2C38-end}
 
@@
 Starting from model performance before to show results produced from this model.}
-\response{lines \ref{}}
+\response{We disagree that this change would improve the organisation of our manuscript.
+As currently organized, the introduction presents question that explore 1) trends in Epichloë prevalence, 2) the climate drivers of these trends, and 3) evaluating models to improve global change forecasts.
+Following this and other comments from Reviewer 2, we have edited to ensure that this organization of topics is repeated in the presentation of methods, in results, and in discussion, and added subheadings to better organize our manuscript.}
 
 \comment{-L400: cryptic? This is a little bit vague, rephrase or change the word?}
-\response{lines \ref{}}
+\response{We have rephrased this for clarity.}
+\linesref{ms-R2C40-begin}{ms-R2C40-end}
 
 \comment{- L402: I would replace “fungal endophytes” by Epichloe syombiosis.}
-\response{lines \ref{}}
+\response{We have made this change.}
+\linesref{ms-R2C41-begin}{ms-R2C41-end}
 
@@
 I would have liked that you linked your conservative/liberal status of symbiosis on this point and what you would expect in term of potential role for the host plant?}
-\response{lines \ref{}}
+\response{In response to this comment, we have added significant discussion on the potential impacts of \emph{Epichloë} on the host distribution as part of our revised discussion.
+Our intention with the conservative/liberal scores is to capture uncertainty in the endophyte identification process, which we have clarified in our response to Reviewer 2 Comment 26.
+Because we found that the central results using either conservative or liberal scores do not differ qualitatively from each other, we do not expect a different interpretation related to host-symbiont range limits, and we feel bringing it up at this point would distract from the principally ecologically-focused section of the discussion.}
+\linesref{ms-R2C42-begin}{ms-R2C42-end}
 
 \comment{L453-455: This part is hard to follow, please rephrase?}
-\response{lines \ref{}}
+\response{We have edited these sentences for clarity.}
+\linesref{ms-R2C43-begin}{ms-R2C43-end}
 
@@
 This would be nice to emphasize that probably land uses change impacted a lot/or not the natural populations of the three host species and emphasize that could also have impacted the ecological dynamics of your symbiosis at larger scale?}
-\response{lines \ref{}}
+\response{We thank Reviewer 2 for this interesting observation.
+We have in fact been working on a separate paper that analyzes the influence of land use changes and other anthropogenic disturbances on endophyte symbiosis.
+In response to this comment, we have added additional material to our discussion that proposes the potential potential for these other global change drivers to impact the dynamics of the symbiosis.}
+\linesref{ms-R2C44-begin}{ms-R2C44-end}
 
@@
 Would you expect a turnover of species (from Epichloe genera), that could impact the performance and potentially something about the dynamics of the plant-symbiotic dynamic?}
-\response{lines \ref{}}
+\response{In response to this comment, we have expanded our discussion of the potential influence that host-symbiont genotype may have on our focal species' range dynamics in our re-organized discussion.}
+\linesref{ms-R2C45-begin}{ms-R2C45-end}
 
 \comment{L501: “Second”, I did not see the “First”.}
-\response{lines \ref{}}
+\response{We have fixed this.}
+\linesref{ms-R2C46-begin}{ms-R2C46-end}
 
 \comment{Supplementary figures:-Page 1; I don’t know if its journal guidelines, but, should there be a title page, with authors information, affiliations and a title being Appendix to “Increasing prevalence of plant-fungal symbiosis across two centuries of environmental change” ?}
-\response{lines \ref{}}
+\response{We do not believe this is a journal requirement, however in response to this comment, we have added the requested information at the start of our supplementary material to connect clearly its content to our manuscript.}
+\linesref{ms-R2C47-begin}{ms-R2C47-end}
 
 \comment{Fig.A3: Which data were used to produce the simulated datasets? Which model is this from?}
-\response{lines \ref{}}
+\response{The "simulated data" here is generated as a posterior predictive test of our model.
+We predicted new values of endophyte occurence for each row in our observed dataset from model fitted to herbarium data.
+We have edited this for clarity, instead referring to "predicted" vs "observed" data in the figure caption (Figure A3), and have added a relevant citation within our methods.}
+\linesref{ms-R2C48-begin}{ms-R2C48-end}
 
 \comment{A4 and A5: Which model, which data? (All plants, all years)?}
-\response{lines \ref{}}
+\response{We have added description of the model and data for these supplemental figures (A4 and A5).}
 
 \comment{A6; From which models this is from? How probabilities are calculated? What is A and T?}
-\response{lines \ref{}}
+\response{We have added description of the model that generates these parameter estimates in Fig.
+A6.
+We have also provided additional explanation of the parameters A and T, which come from Eqn.
+1 describing the endophyte prevalence model.}
 
 \comment{A7; From which models; what is the legend for the value means, what is $\tau$?}
-\response{lines \ref{}}
+\response{We have added description of the model that generates these parameter estimates in Fig.
+A7.
+We have also provided additional explanation of the parameter  $\tau$, which come from Eqn.
+1 describing the endophyte prevalence model.}
 
 \comment{A8;redefine what is liberal vs conservative; what AGPE, AGHY and ELVI means if the Panel B-C-F is from which model. What are the value? (This is an average per year, across 200 years?)}
-\response{lines \ref{}}
+\response{In response to this comment, we have redefined liberal vs.
+conservative scores in the figure caption of Fig.
+A8, along with adding information that describes the model from which these parameter estimes comes from.
+We have re-plotted the figure with species names in place of the four letter codes noted by Reviewer 2}
 
 \comment{A9;A10; Specify models}
-\response{lines \ref{}}
+\response{We have added description of the models from which these parameter estimates come for Figures A9-A10.}
 
 \comment{A11; Legend need specifications, the data need to be specified as well.}
-\response{lines \ref{}}
+\response{For this figure (now Figure A12), we have re-plotted the figure with species names in place of the four letter codes noted by Reviewer 2.
+We have additionally edited the figure caption to more clearly describe the different sources of data.
+)}
 
 \comment{Table A1. Which models, what are the columns names? It is for the full temporal I guess? Need to be specified.}
-\response{lines \ref{}}
+\response{Table A1 presents counts of our data across the different herbaria included in this study and does not refer to models.
+We have edited the caption of Table A1 for clarity and replaced four letter codes with species names.
+Based on other comments from Reviewer 2, we have also revised the captions for supplemental figures A16-A18 to similarly clarify the origin of the plotted model estimates.}
 

### 2025-01-23 Analyses/endo_herbarium_multispecies_spatiotemporal_analysis_with_conservative_scores.R
replotting the conservative comparison plot

@@ -763,9 +763,10 @@ vrt_elvi <- inlabru::fm_pixels(mesh, mask = elvi_distribution, format = "sp", di
 # saveRDS(vrt_agpe, file = "agpe_distribution_df.rds")
 # saveRDS(vrt_elvi, file = "elvi_distribution_df.rds")
 # 
-# vrt_aghy <- readRDS(file = "aghy_distribution_df.rds")
-# vrt_agpe <- readRDS(file = "agpe_distribution_df.rds")
-# vrt_elvi <- readRDS(file = "elvi_distribution_df.rds")
+vrt_aghy <- readRDS(file = "aghy_distribution_df.rds")
+vrt_agpe <- readRDS(file = "agpe_distribution_df.rds")
+vrt_elvi <- readRDS(file = "elvi_distribution_df.rds")
+
 
 
 # ggplot()+
@@ -819,16 +820,16 @@ states_map <- st_as_sf(maps::map("state", plot = FALSE, fill = TRUE)) %>%
 
 
 
-min_trend <- max(svc.pred_aghy$mean, svc.pred_aghy$mean, svc.pred_aghy$mean)
+# min_trend <- max(svc.pred_aghy$mean, svc.pred_aghy$mean, svc.pred_aghy$mean)
+# 
+# max_trend <- min(svc.pred_aghy$mean, svc.pred_aghy$mean, svc.pred_aghy$mean)
 
-max_trend <- min(svc.pred_aghy$mean, svc.pred_aghy$mean, svc.pred_aghy$mean)
-
-trendrange <- range(svc.pred_aghy$mean, svc.pred_aghy$mean, svc.pred_aghy$mean)
+trendrange <- range(svc.pred_aghy$mean, svc.pred_agpe$mean, svc.pred_elvi$mean)
 
 
 
-space_x <- range(svc.pred_aghy@coords[,1],svc.pred_agpe@coords[,1],svc.pred_agpe@coords[,1])
-space_y <- range(svc.pred_aghy@coords[,2],svc.pred_agpe@coords[,2],svc.pred_agpe@coords[,2])
+space_x <- range(svc.pred_aghy@coords[,1],svc.pred_agpe@coords[,1],svc.pred_elvi@coords[,1])
+space_y <- range(svc.pred_aghy@coords[,2],svc.pred_agpe@coords[,2],svc.pred_elvi@coords[,2])
 
 
 
@@ -841,7 +842,7 @@ svc_time_map_AGHY <- ggplot()+
[…]

### 2025-01-23 Analyses/endo_herbarium_multispecies_spatiotemporal_analysis.R
fixing the wonky longitude axis

@@ -29,7 +29,7 @@ endophyte_colors <- c("#fdedd3","#f3c8a8", "#5a727b", "#4986c7", "#181914",  "#1
 ############ Read in the herbarium dataset ############################### 
 ################################################################################
 
-endo_herb_georef <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/DigitizedHerbariumRecords/endo_herb_georef.csv") 
+endo_herb_georef <- read_csv(file = "~/Dropbox/Josh&Tom - shared/Endo_Herbarium/DigitizedHerbariumRecords/endo_herb_georef.csv") %>% 
   # filter(Country != "Canada") %>%
   mutate(species_index = as.factor(case_when(spp_code == "AGHY" ~ "1",
                                              spp_code == "AGPE" ~ "2",
@@ -186,7 +186,7 @@ collections_map <- ggplot()+
   geom_map(data = outline_map, map = outline_map, aes(map_id = region), color = "grey", linewidth = .1, fill = "#FAF9F6")+
   geom_map(data = states_shape, map = states_shape, aes(map_id = region), color = "grey", linewidth = .1, fill = NA)+
   geom_point(data = endo_herb, aes(x = lon, y = lat, color = species), alpha = .7, size = 1.2)+
-  coord_sf(xlim = c(-109,-68), ylim = c(21,49), crs = 4326)+
+  coord_sf(xlim = c(-109,-68), ylim = c(19,49), crs = 4326)+
   # lims(x = c(-109,-68), y = c(21,49))+
   scale_color_manual(values = species_colors)+
   theme_light()+
@@ -208,13 +208,13 @@ endo_status_map <- ggplot()+
   geom_map(data = states_shape, map = states_shape, aes(map_id = region), color = "grey", linewidth = .1, fill = NA)+
   geom_point(data = endo_herb, aes(x = lon, y = lat, color = endo_status_text), alpha = .7, size = .8)+
   facet_wrap(~factor(year_bin, levels = c("pre-1969", "post-1969"))+species)+
-  coord_sf(xlim = c(-109,-68), ylim = c(21,49), crs = 4326)+
+  coord_sf(xlim = c(-109,-68), ylim = c(19,49), crs = 4326)+
   scale_color_manual(values = c(endophyte_colors[2],endophyte_colors[6]))+
   theme_light()+
   theme(strip.background = element_blank(),
[…]

### 2025-01-30 Manuscript/Reviewer_response_endo_herbarium.tex
Submitted suggestions for R1

@@
 }
-
+\newcommand{\jacob}[2]{{\color{blue}{#1}}\footnote{\textit{\color{blue}{#2}}}}
 %\IfValueT{#2}{- \ref{#2}}
@@
 
-Beyond this, we also are reassured when examining the Bayesian posterior credible interval of key results, such as the estimated temporal trend (Fig.
-\ref{ms-fig:svc_time_map_CI}), across the  host distributions.
+Beyond this, we also are reassured when examining the Bayesian posterior credible interval of key results, such as the estimated temporal trend \jacob{(Fig.
+\ref{ms-fig:svc_time_map_CI}), across the  host distributions.}{For some reason, the PDF is displaying the plot caption, which makes the sentence awkward.}
 For example, we find that for \emph{A.
@@
 perennans}.
-As suggested by the reviewer, we also assessed the relationship between initial prevalence and predicted changes in prevalence, which shows that for these species the largest increases do occur at locations with low starting prevalence.
-To keep the main text focused on temporal change in prevalence, we placed a figure showing this result in the supplemental material (Figure ), but have added text describing this to our results.}
+As suggested by the reviewer, we also assessed the relationship between initial prevalence and predicted changes in prevalence, which shows that for these species, the largest increases do occur at locations with low starting prevalence.
+To keep the main text focused on temporal change in prevalence, we placed a figure showing this result in the supplemental material \jacob{(Figure)}{You forgot to include the figure number.
+I believe it should be Figure A11.}, but have added text describing this to our results.}
 \linesref{ms-R1C7-begin}{ms-R1C7-end}
@@
 I recommend they dial this claim back.}
-\response{We appreciate Reviewer 1 bringing this to our attention.
+\response{\jacob{We appreciate Reviewer 1 bringing this to our attention.
 We would argue that our phrasing is not inaccurate because we have successfully linked changes in symbioses to climate change drivers, whereas previous research including Heberling \& Burke 2019 focus on validating methods for quantifying microbial community composition within historic specimens, but do not connect this to global change drivers.
@@
 However, given the scope of the current study, future work is needed to address specific global change hypotheses." However, we do recognize that there is a quickly growing body of research on microbial symbioses within herbarium specimens, especially those using sequencing and bioinformatics techniques.
-In response to this comment, we have rephrased this statement to clarify that we see novelty particularly in connecting microbial symbiosis to climate change drivers, and to temper our assertion of novelty.}
+In response to this comment, we have rephrased this statement to clarify that we see novelty particularly in connecting microbial symbiosis to climate change drivers, and to temper our assertion of novelty.}{I like your response}}
 \linesref{ms-R1C10-begin}{ms-R1C10-end}
@@
 \comment{In the abstract, I recommend briefly mentioning the statistical analyses or framework used (e.g., INLA) before presenting the results to enhance readability.}
-\response{We have revised the abstract to include information about the statistical analysis.}
+\response{\jacob{We have revised the abstract to include information about the statistical analysis.}{Please see my comment in the abstract}}
 \linesref{ms-R2-C15-begin}{ms-R2-C15-end}
@@
 I would recommend to group the paragraphs presenting the research gaps of the ecological importance of endophytes, particularly in the context of climate change together and addressing the methodological novelty of this study in a separate, cohesive paragraph.}
-\response{}
+\response{\jacob{}{You forgot to answer this comment.
+I think we should push back.
+Here is my proposition: "We appreciate the suggestion to shorten the  Introduction and improve its clarity.
+While we understand the value of streamlining the content, we believe that the examples provided, including those related to coral and insect symbioses, serve to contextualize the study within broader ecological frameworks and may help engage readers who are less familiar with the topic."}}
 
@@
 In cases with clear identifications, the two scores are the same.
-When there is uncertainty in identification, we err on the side of assigning positive for one score and err on the side of assigning negative for the other.
+When there is uncertainty in identification, we \jacob{err}{I don't understand.
+Typo ?} on the side of assigning positive for one score and \jacob{err}{Same here } on the side of assigning negative for the other.
 We present a comparison of results based on this scoring choice in our supplement, which showed minimal difference, and therfore we only present the results of the "liberal" score for the rest of the central results.}

### 2025-02-04 Manuscript/Reviewer_response_endo_herbarium.tex
response to reviewer 1

@@
 %\IfValueT{#2}{- \ref{#2}}
+\newcommand{\tom}[2]{{\color{red}{#1}}\footnote{\textit{\color{red}{#2}}}}
+%\IfValueT{#2}{- \ref{#2}}
 
@@
 Thank you for the opportunity to submit a revision of our manuscript to \emph{Global Change Biology}. 
-In response to both reviewers, we have added substantial discussion of the ecological implications for hosts of our finding that symbiosis with \emph{Epichloë} fungal endophytes has increased in prevalence across the last two centuries.
-Additionally, we have added a completely new simulation analysis evaluating the impact that spatially-biased sampling may have on our analysis, as prompted by Reviewer 1, presented in an expanded Supplemental Methods.
+In response to both reviewers, we have added substantial discussion of the ecological implications of our finding that symbiosis with \emph{Epichloë} fungal endophytes has increased in prevalence across the last two centuries.
+Additionally, we have added a new simulation analysis evaluating the impact that spatially-biased sampling may have on our analysis, as prompted by Reviewer 1, which is now presented in an expanded Supplemental Methods.
 We have also made significant revisions throughout the manuscript to improve the organization and clarity of our ideas in response to reviewer recommendations.   
@@
 \response{We appreciate Reviewer 1's constructive criticisms, here and in their other line-specific comments.
-We have made our best effort to diligently revise our paper in response to these criticisms, which we feel has been tangibly improved in the process.}
+We have made our best effort to diligently revise our paper in response to these criticisms, which we feel has been improved in the process.}
 
@@
 Does vertical transmission maintain prevalence through time, or is there regional turnover in relative prevalence of different species through time?}
-\response{These are insightful questions that we agree deserve careful consideration.
-We have added additional description of genotypic diversity within the interaction to provide readers with this context in our methods section.
-In addition, we return to this topic in expanded discussion material}
+\response{These are great questions that we agree deserve careful consideration.
+Indeed, there can be genetic variation of symbionts within and across host populations, which we now describe in more detail to provide readers with this context in our methods section.
+In addition, we return to this topic in expanded discussion material, \tom{where we...}{In general I would be more thorough in your responses so that you actually answer or address the reviewers' questions here in the letter, in addition to the line numbers for the revised text.
+In this example I would acknowledge here that we can track overall prevalence but not genetic variation or change, and this could be an interesting future direction.
+Anything that could be seen as a weakness, flip it and make it a direction for future research.}
 \linesref{ms-R1C3-begin}{ms-R1C3-end}
 \linesref{ms-R2C14-begin}{ms-R2C14-end}
+}
 
@@
 (3) If not, I at least recommend adding a caveat somewhere in the methods that collections were included from across eastern US, but were primarily collected from the four states in which herbaria were visited and that this has the potential to bias your results in specific ways (also see line comment on L504-507).}
-\response{The herbarium visits we completed for this analysis do result in fewer specimens from the northeastern portion of each species ranges, and \emph{A.
-perennans} is the least commonly collected species among our focal hosts.
-We are unable to perform any additional sampling. 
-However, we feel our methodology is robust. 
-To demonstrate this, and in response to this comment, we have performed a simulation analysis (\emph{Appendix A - Influence of spatially-biased sampling on model interpretation}) that shows the ability of an equivalent model to recover spatially-varying temporal trends within a region with up to $80$\% of data missing.
-
-Beyond this, we also are reassured when examining the Bayesian posterior credible interval of key results, such as the estimated temporal trend \jacob{(Fig.
+\response{Thank you for these insightful comments. 
+While we visited only a handful of herbaria, we were strategic in sampling from collections with broad geographic coverage and were generally pleased with the spatial representation they provided (Fig.
+1).
+That said, we had access to relatively few specimens from the northeastern portion of each species' range, and \emph{A.
+perennans} is overall less commonly collected than our other two focal host species.
+Unfortunately, we are unable to perform any additional sampling for this revision. 
+However, we conducted additional analyses, as recommended by the reviewer, to explore this concern. 
+Specifically, we have performed a new simulation analysis (\emph{Appendix A - Influence of spatially-biased sampling on model interpretation}) that shows \tom{the ability of an equivalent model to recover spatially-varying temporal trends within a region with up to $80$\% of data missing.}{I have not read the revision yet but I don't really know what this sentence means.}
+\\
+\\	
+The Bayesian posterior credible intervals of key results lend confidence to our interpretation, despite the uneven sampling that is inherent to collections-based studies such as ours \jacob{(Fig.
 \ref{ms-fig:svc_time_map_CI}), across the  host distributions.}{For some reason, the PDF is displaying the plot caption, which makes the sentence awkward.}
 For example, we find that for \emph{A.
-perennans}, the width of this credible interval is consistently small (suggesting that estimated trends vary by only $\mp$ 2 percentage points per year) within areas with many samples (the midwest), while within areas with fewer samples (the northeast) this interval is wider but still relative small (trends are estimated to be $\mp$ 3 percentage points per year), suggesting that the trend in this region is highly likely to be in a positive direction.
-Certainly we expect more thorough sampling to be valueable.
-We have added consideration of this and other potential biases to our discussion, along with the methods caveat described by this reviewer.}
+perennans}, the width of the credible interval for temporal change is consistently small (estimated trends in prevalence vary by only $\mp$ 2 percentage points per year) within highly-sampled areas (the midwest), while within areas with fewer samples (the northeast) this interval is only slightly wider (trends are estimated to be $\mp$ 3 percentage points per year).
+These observations suggest high confidence in the positive trend of increasing endophyte prevalence through time, even as the point estimate for that slope is more uncertain in sparsely sampled regions.
+We have added consideration of this and other potential biases to our discussion, along with \tom{the methods caveat described by this reviewer}{Be more specific here?
+I am not sure which caveat you are referring to.}.
 \linesref{ms-R1C11-begin}{ms-R1C11-end} methods
@@
 \linesref{ms-R1C4-begin}{ms-R1C4-end} Appendix A - Supporting Methods
+}
 
@@
 }
-\response{We agree that these are interesting questions.
-We have added additional figures which present the predicted prevalence across host distributions in 1895 and in 2020 (Figure 4).
-% I haven't taken the time to figure out how to get ref to work correctly here.
-They show clear range expansion by the symbiosis towards the western range edge of \emph{A.
-hyemalis} and towards the northern range edge of \emph{A.
-perennans}.
-As suggested by the reviewer, we also assessed the relationship between initial prevalence and predicted changes in prevalence, which shows that for these species, the largest increases do occur at locations with low starting prevalence.
-To keep the main text focused on temporal change in prevalence, we placed a figure showing this result in the supplemental material \jacob{(Figure)}{You forgot to include the figure number.
-I believe it should be Figure A11.}, but have added text describing this to our results.}
+\response{We agree that these are interesting questions. In response to this comment we have added two new figures.
+First, the new Figure 4 shows predicted endophyte prevalence across host distributions in 1895 and in 2020. % I haven't taken the time to figure out how to get ref to work correctly here. 
+Interestingly, this figure shows clear expansion of the symbiosis towards the western range edge of \emph{A. hyemalis} and towards the northern range edge of \emph{A. perennans}. 
+Second, as suggested by the reviewer, we also assessed the relationship between initial prevalence and predicted changes in prevalence (new Figure A11).
+We found that for two species, the largest increases do occur at locations with low starting prevalence.
+To keep the main text focused on temporal change in prevalence, we placed this figure in the supplemental material but have added text describing this to our results.}
 \linesref{ms-R1C7-begin}{ms-R1C7-end}
@@
 However, given the scope of the current study, future work is needed to address specific global change hypotheses." However, we do recognize that there is a quickly growing body of research on microbial symbioses within herbarium specimens, especially those using sequencing and bioinformatics techniques.
-In response to this comment, we have rephrased this statement to clarify that we see novelty particularly in connecting microbial symbiosis to climate change drivers, and to temper our assertion of novelty.}{I like your response}}
+In response to this comment, we have rephrased this statement to clarify that we see novelty particularly in connecting microbial symbiosis to climate change drivers, and to temper our assertion of novelty.}{I like your response}}\tom{}{I agree -- very diplomatic!}
 \linesref{ms-R1C10-begin}{ms-R1C10-end}
@@
 Not just about the prevalence of the relationship (which they cover thoroughly), but about what this might mean for grass performance at the population or even community scales under ongoing climate change.}
-\response{Thank you for pushing us to connect our work more directly to the biology of our system.
+\response{Thank you for pushing us to connect our work more directly to the biology of our system and the fitness consequences of symbiosis under global change.
 This was a comment echoed by Reviewer 2 (Comment 14).
-We had added substantial material our discussion that presents the key implications for population and range dynamics under climate change.}
+In response to both reviewers we have added substantial material to our discussion that presents the key implications of our findings for population and range dynamics under climate change.}
 \linesref{ms-R2C14-begin}{ms-R2C14-end}

### 2025-02-05 Manuscript/Reviewer_response_endo_herbarium.tex

@@
 Here is my proposition: "We appreciate the suggestion to shorten the  Introduction and improve its clarity.
-While we understand the value of streamlining the content, we believe that the examples provided, including those related to coral and insect symbioses, serve to contextualize the study within broader ecological frameworks and may help engage readers who are less familiar with the topic."}}
+While we understand the value of streamlining the content, we believe that the examples provided, including those related to coral and insect symbioses, serve to contextualize the study within broader ecological frameworks and may help engage readers who are less familiar with the topic."}\tom{}{I like Jacob's suggestion here and agree that we should push back on cutting the non-plant examples.
+But maybe the reorganization suggested here could work.
+I have not looked at the ms yet, maybe you have already done this.}}
 
@@
 For example, you could begin by discussing about the spatial/temporal patterns in prevalence, followed by the relationships with climatic drivers before to conclude with the technical/ecological novelty of your findings.}
-\response{In response to this comment, we have added sub-headings to our Discussion and revised our discussion section for clarity and flow between ideas.
-Many of these changes are detailed in response to further comments from Reviewer 2}
+\response{Thank you for the constructive suggestions.
+In response to this comment, we have added sub-headings to our Discussion and revised our discussion section for clarity and flow between ideas.
+Many of these changes are detailed in response to further comments from Reviewer 2.}
 
@@
 \response{We appreciate Reviewer 2's attention to our supplemental material.
-We have revised through all of the figure captions in the supplement to provide this information.
-These edits are also detailed in our responses to specific comments further down.}
+We have revised all of the figure captions in the supplement to provide the information that readers will need to interpret this material.
+These edits are also detailed in our responses to specific comments below.}
 
 \comment{L54: Which physical legacies or ecological processes? Provide example.}
-\response{In response to this comment, we have added additional information about the examples of ecological processes studied previously using herbarium specimens.}
+\response{In response to this comment, we have added additional information about the examples of ecological processes studied previously using herbarium specimens, \tom{such as...}{Again, be sure to answer questions here.
+It's ok (and expected) to have redundancy between the manuscript and response letter.}.}
 \linesref{ms-R2C19-begin}{ms-R2C19-end}
@@
 \comment{L64-L74: Please remove example of symbiosis from other systems and focus on your targeted endophytic symbiosis. (Remove corals, insects or lichens.)}
-\response{We strongly feel that these other symbioses provide relevant examples of the potential responses of symbioses to global change.
-Our central point is that this response is known in a few key symbiota, but not in the endophytic symbiosis that is our focus, and that studies across numerous taxa are valueable to understand \emph{why} the response tips from resilience to disruption of the interaction.
-These examples motivate our hypotheses and so we have opted to keep them in the text.}
+\response{We appreciate the suggestion but, given the broad readership of \textit{Global Change Biology}, we strongly feel that these other symbioses provide relevant examples of the potential responses of symbioses to global change, and that they help place our work into a broader context.
+Our central point is that climate change responses are known in a few key symbiota, but not in the grass-endophyte symbiosis that is our focus, and that studies across diverse taxa are necessary to understand \emph{why} responses can tip from resilience to disruption of the interaction.
+These examples motivated our study and so we have opted to keep them in the text.}
 
@@
 \response{We have edited this sentence to include a specific prediction for how this hypothesis would relate to our study system.
-We additionally have added additional specific predictions two paragraphs further down, where we present our study's questions.}
+We have also added specific predictions \tom{two paragraphs further down}{I would provide line numbers for this change.}, where we present our study's questions.}
 \linesref{ms-R2C21-begin}{ms-R2C21-end}
@@
 \comment{-L131-134: Super nice data!}
-\response{We thank this Reviewer 2 for acknowledging the value of our dataset which represents significant time and effort.}
+\response{We thank this Reviewer 2 for acknowledging the value of our dataset, which represents significant time and effort.}
 
@@
 Because we included coordinates georeferenced to local municipalities, counties, and in a few instances, states, the width of this bounding box varied from less than 1 km.
-to over 1233 km (median 841 km.).}
+to over 1233 km (median 841 km.).\tom{}{Good example of answering the question AND flagging the revised line number.}}
 \linesref{ms-R2C24-begin}{ms-R2C24-end}
@@
 \comment{-L176: “Partial” would be more adapted than “imperfect”.}
-\response{"Imperfect" transmission is a commonly use term within the study of symbiosis therefore we have opted to retain its use.
-We have however also added a definition of the term "imperfect transmission" at the point it is first used.}
+\response{"Imperfect" transmission is a commonly use term within the study of symbiosis, therefore we have opted to retain its use for consistency with previous literature.
+We have, however, also added a definition of the term "imperfect transmission" at the point it is first used.}
 \linesref{ms-R2C25-begin}{ms-R2C25-end}
@@
 Please reformulate and clarify in the main text and maybe it can be used in the discussion?}
-\response{We are sorry that this terminology has caused confusion.
-In response to this feedback, we have edited for to provide additional explanation.
-The central point is that we mark two statuses to evaluate uncertainty in the process of assigning an endophyte identification under the microscope for each specimen.
-In cases with clear identifications, the two scores are the same.
-When there is uncertainty in identification, we \jacob{err}{I don't understand.
-Typo ?} on the side of assigning positive for one score and \jacob{err}{Same here } on the side of assigning negative for the other.
-We present a comparison of results based on this scoring choice in our supplement, which showed minimal difference, and therfore we only present the results of the "liberal" score for the rest of the central results.}
+\response{We are sorry that this terminology has caused confusion, and we appreciate the reviewer bringing this to our attention.
+In response to this feedback, we have made several edits to provide additional explanation.
+We record two scores for each sample to evaluate uncertainty in the process of assigning endophyte status under the microscope.
+In cases with unambiguous identification (presence or absence), the two scores are the same.
+When there is uncertainty in identification, we assign a positive ``liberal'' score and a negative ``conservative'' score.
+\tom{}{I would add here that ``split'' scores are infrequent, and give the frequency.} We present a comparison of results based on this scoring choice in our supplement, which showed minimal difference, and therefore we only present the results of the "liberal" score for the rest of the central results.}
 \linesref{ms-R2C26-begin}{ms-R2C26-end}
@@
 I feel like this paragraph is coming out of the blue and is making the full material and methods confusing.}
-\response{We appreciate the push to improve the organization of our manuscript so that it can be readily understood.
-In this case, we see the current placement of the section "Modeling distributions of host species" as appropriate because the preceeding section describes models of endophyte prevalence that do not rely on these models of host species.
-The predictions from models of host species are used principally for generating pixels with which we test the relationship between climate drivers and trends in endophyte prevalence and for visualizing relationships across study region.
-However, in response to this comment, we have added additional context describing this role that the host distribution models play.}
+\response{We appreciate the push to improve the organization of our manuscript.
+In this case, we see the current placement of the section "Modeling distributions of host species" as appropriate because the preceding section describes models of endophyte prevalence that are independent from these models of host species' distributions.
+The predictions from the host distribution models are used principally for identifying realistic geographic ranges over which we test the relationship between climate drivers and trends in endophyte prevalence, and for visualizing relationships across study region.
+In response to this comment, we have added additional context describing this role that the host distribution models play.\tom{}{Maybe you say this in the text but I would point out that the herbarium records do not reflect the entire species' ranges, so if we want to make range-wide predictions we need some other way to do this.}}
 \linesref{ms-R2C30-begin}{ms-R2C30-end}
@@
 \comment{-L281: “which” instead of “whether”?}
-\response{This change would alter the meaning of the question posed in this sentence, which we feel is unnecessary. We have opted to keep "whether".}
+\response{This change would alter the meaning of the question posed in this sentence, which we feel would not accurately reflect our intention. We have opted to keep "whether".}
 
@@
 They are quite contradictory no?}
-\response{Reviewer 2 confuses 'collector', those researchers who collected herbarium specimens, and who we term 'scorer', those members of our lab who performed the microscope work to identify \emph{Epichloë} fungal endophytes.
+\response{Reviewer 2 confuses 'collector', those researchers who collected herbarium specimens, with the individuals we term 'scorer', those members of our lab who performed the microscope work to identify \emph{Epichloë} fungal endophytes.
+In response to this comment, we have more clearly defined 'collector' and 'scorer' in our manuscript.
 Aside from this confusion, there is no contradiction in our results; we found that collector identify did not influence model predictions, while scorer identity did influence model predictions, as evidenced in Figs.
 A9-A10.
-In response to this comment, we have more clearly defined 'scorer' in our manuscript.}
+}
 \linesref{ms-R2C35-begin}{ms-R2C35-end}
@@
 Models fit to the full set of pixels for each species show climate regressions in a similar direction as shown here and have narrower credible intervals.
-We could, for example, generate many thousands of points across each distribution, but this would make the credible intervals largely meaningless.
+We could, for example, generate many thousands of points across each distribution, but \tom{this would make the credible intervals largely meaningless}{By this logic the CIs are meaningless for any sample size, whether its the full range or an arbitrary subset.}.
 We have added material describing why the choice of pixel dimensions is arbitrary as described in response to Reviewer 2 Comment 32.
@@
 Starting from model performance before to show results produced from this model.}
-\response{We disagree that this change would improve the organisation of our manuscript.
-As currently organized, the introduction presents question that explore 1) trends in Epichloë prevalence, 2) the climate drivers of these trends, and 3) evaluating models to improve global change forecasts.
+\response{We appreciate the suggestion but we respectfully disagree that this change would improve the organisation of our manuscript.
+As currently organized, the introduction presents questions that explore 1) trends in Epichloë prevalence, 2) the climate drivers of these trends, and 3) evaluating models to improve global change forecasts.
 Following this and other comments from Reviewer 2, we have edited to ensure that this organization of topics is repeated in the presentation of methods, in results, and in discussion, and added subheadings to better organize our manuscript.}
@@
 Would you expect a turnover of species (from Epichloe genera), that could impact the performance and potentially something about the dynamics of the plant-symbiotic dynamic?}
-\response{In response to this comment, we have expanded our discussion of the potential influence that host-symbiont genotype may have on our focal species' range dynamics in our re-organized discussion.}
+\response{This is an interesting question.
+\tom{}{I would add a brief response to the question here, even a quick summary of whatever speculation you added to the Discussion.}In response to this comment, we have expanded our discussion of the potential influence that host-symbiont genotype may have on our focal species' range dynamics in our re-organized discussion.}
 \linesref{ms-R2C45-begin}{ms-R2C45-end}
@@
 \comment{A4 and A5: Which model, which data? (All plants, all years)?}
-\response{We have added description of the model and data for these supplemental figures (A4 and A5).}
+\response{We have added description of the model and data for these supplemental figures (A4 and A5).\tom{}{And respond to the question here.}}
 
@@
 conservative scores in the figure caption of Fig.
-A8, along with adding information that describes the model from which these parameter estimes comes from.
-We have re-plotted the figure with species names in place of the four letter codes noted by Reviewer 2}
+A8, along with adding information that describes the model from which these parameter estimates are derived.
+We have re-plotted the figure with species names in place of the four letter codes noted by Reviewer 2.}
 

### 2025-02-05 Manuscript/endo_herbarium_R1.tex

@@
 
-	\textit{Keywords}: climate change, plant-microbe symbiosis, herbarium, museum specimen, INLA, spatially-varying coefficients model, \emph{Epichloë}, \emph{Poaceae}.
+	\textit{Keywords}: \jacob{climate change, plant-microbe symbiosis, herbarium, museum specimen, INLA, spatially-varying coefficients model, \emph{Epichloë}, \emph{Poaceae}.}{alphabetic order ? }
 
@@
 Specifically, we \revise{examined} 2,346 herbarium specimens of three grass host species (\emph{Agrostis hyemalis}, \emph{Agrostis perennans}, \emph{Elymus virginicus}) collected over the past two centuries (1824 -- 2019) for the presence or absence of \emph{Epichloë} symbiosis.
-\revise{Analysis of an approximate Bayesian spatially-varying coefficients model implemented in INLA revealed} that endophytes increased in prevalence over the last two centuries from ca.
+\revise{Analysis of an \jacob{approximate Bayesian spatially-varying coefficients model implemented in INLA}{This is the first time INLA is mentioned.
+It seems to be a suggestion from the reviewer.
+However, I think most readers may not be familiar with it.
+I appreciate the precision regarding the type of Bayesian statistics (approximate), but I wonder if that might be too technical for the abstract.
+I would understand if it were included in the methods section.} revealed} that endophytes increased in prevalence over the last two centuries from ca.
 25\% to ca.
@@
 These specimens often preserve physical legacies of ecological processes and species' interactions from dynamically changing environments across time and space.
-For example, previous researchers have used plant collections (herbaria) to document shifts in phenology \linelabel{R2C19-begin}\revise{by examining reproductive structures}\citep{willis2017old, park2019herbarium,  berg2019examination}, pollination \revise{through examination of pollen removal} \citep{pauw2011reconstruction, duan2019century}, and herbivory \revise{by documenting leaf damage} \citep{meineke2019herbarium} related to anthropogenic climate change.
+For example, previous researchers have used plant collections (herbaria) to document shifts in phenology \linelabel{R2C19-begin}\revise{by examining reproductive structures} \citep{willis2017old, park2019herbarium,  berg2019examination}, pollination \revise{through examination of pollen removal} \citep{pauw2011reconstruction, duan2019century}, and herbivory \revise{by documenting leaf damage} \citep{meineke2019herbarium} related to anthropogenic climate change.
 \linelabel{R2C19-end}
@@
 \linelabel{R1C4-dicussion-begin}\revise{
-Spatial autocorrelation is another phenomenom likely common in data derived from herbarium specimens \cite{willems2022forest}, which our spatially-explicit analysis models among samples.
+Spatial autocorrelation is another \jacob{phenomenom}{phenomenon ?} likely common in data derived from herbarium specimens \cite{willems2022forest}, which our spatially-explicit analysis models among samples.
 Beyond spatial autocorrelation of outcomes, systematic differences in sampling across space can result in spatial bias. 
@@
 km, including specimens from Mexico and Canada.
-The specimens we examined are concentrated in the South-Central United States,with fewer specimens in the rapidly warming northeastern United States.
+The specimens we examined are concentrated in the South-Central United States, with fewer specimens in the rapidly warming northeastern United States.
 We provide a simulation analysis exploring the potential impact of spatially and temporally biased sampling (Appendix A - Supporting Methods). 

### 2025-02-06 Manuscript/endo_herbarium_R1.tex
finished edits and comments

@@
 \documentclass[11pt]{article}
-\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
+%\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
 \usepackage{fullpage}
@@
 
-	\noindent{\footnotesize Prepared using the suggested \LaTeX{} template for \textit{Am.\ Nat.}}
-
 	%\linenumbers{}
@@
 \linelabel{R2-C15-begin}Species' distributions and abundances are shifting in response to ongoing global climate change. 
-Mutualistic microbial symbionts can provide hosts with protection from environmental stress that may contribute towards \revise{resilience under} environmental change, however \revise{this change} may also disrupt species interactions and lead to declines in hosts and/or symbionts.
+Mutualistic microbial symbionts can provide hosts with protection from environmental stress that may \revise{promote resilience under} environmental change, however \revise{this change} may also disrupt species interactions and lead to declines in hosts and/or symbionts.
 Symbionts preserved within natural history specimens offer a unique opportunity to quantify changes in microbial symbiosis across broad temporal and spatial scales. 
-We asked how the prevalence of seed-transmitted fungal symbionts of grasses (\emph{Epichloë} endophytes) have changed over time in response to climate change, and how these changes vary across host species' \revise{distributions}.
+We asked how the prevalence of seed-transmitted fungal symbionts of grasses (\emph{Epichloë} endophytes) has changed over time in response to climate change, and how these changes vary across host species' \revise{distributions}.
 Specifically, we \revise{examined} 2,346 herbarium specimens of three grass host species (\emph{Agrostis hyemalis}, \emph{Agrostis perennans}, \emph{Elymus virginicus}) collected over the past two centuries (1824 -- 2019) for the presence or absence of \emph{Epichloë} symbiosis.
-\revise{Analysis of an \jacob{approximate Bayesian spatially-varying coefficients model implemented in INLA}{This is the first time INLA is mentioned.
+\revise{Analysis of an \jacob{approximate Bayesian spatially-varying coefficients model}{This is the first time INLA is mentioned.
 It seems to be a suggestion from the reviewer.
@@
 I appreciate the precision regarding the type of Bayesian statistics (approximate), but I wonder if that might be too technical for the abstract.
-I would understand if it were included in the methods section.} revealed} that endophytes increased in prevalence over the last two centuries from ca.
+I would understand if it were included in the methods section.}\tom{}{I agree.
+I cut ' implemented in INLA'} revealed} that endophytes increased in prevalence over the last two centuries from ca.
 25\% to ca.
@@
 Changes in the variability of precipitation and temperature during off-peak seasons were also important predictors of increasing endophyte prevalence. 
-Our analysis performed favorably in an out-of-sample predictive test with contemporary survey data, a rare extra step in collections-based research.
-However, we identified greater local-scale variability in endophyte prevalence in contemporary data compared to model predictions based on historic data, suggesting new directions that could improve predictive accuracy.
+Our analysis performed favorably in an out-of-sample predictive test with \tom{contemporary survey data}{Since you drop numbers for the herbarium samples I would drop some numbers here too.}, a rare extra step in collections-based research.
+We identified greater local-scale variability in endophyte prevalence in contemporary data compared to model predictions based on historic data, suggesting new directions that could improve predictive accuracy.
 Our results provide novel evidence for a cryptic biological response to climate change that may contribute to the resilience of host-microbe symbiosis through \revise{fitness benefits to symbiotic hosts}.
@@
 These specimens often preserve physical legacies of ecological processes and species' interactions from dynamically changing environments across time and space.
-For example, previous researchers have used plant collections (herbaria) to document shifts in phenology \linelabel{R2C19-begin}\revise{by examining reproductive structures} \citep{willis2017old, park2019herbarium,  berg2019examination}, pollination \revise{through examination of pollen removal} \citep{pauw2011reconstruction, duan2019century}, and herbivory \revise{by documenting leaf damage} \citep{meineke2019herbarium} related to anthropogenic climate change.
+For example, previous researchers have used plant collections (herbaria) to document shifts in phenology \linelabel{R2C19-begin}\tom{\revise{by examining reproductive structures}}{I don't really understand this revision or what prompted it.
+Phenology studies do not examine reproductive structures, they examine timing.
+I also don't think the sentence works grammatical with each list element having a 'by' clause.} \citep{willis2017old, park2019herbarium,  berg2019examination}, pollination \revise{through examination of pollen removal} \citep{pauw2011reconstruction, duan2019century}, and herbivory \revise{by documenting leaf damage} \citep{meineke2019herbarium} related to anthropogenic climate change.
 \linelabel{R2C19-end}
@@
 For example, bacterial symbionts of insects, such as \emph{Wolbachia}, can improve their hosts' thermal tolerance \citep{truitt2019wolbachia, renoz2019evolutionary}, and arbuscular mycorrhizal fungi, documented in 70-90\% of families of land plants \citep{parniske2008arbuscular}, allow their hosts to persist through drought conditions by improving water and nutrient uptake \citep{cheng2021elucidating}.
-On the other hand, changes in the mean and variance of environmental conditions may disrupt microbial  mutualisms by changing the costs and benefits of the interaction for each partner, leading the interaction to deteriorate \citep{aslan2013mutualism, fowler2024microbial}.
+On the other hand, changes in the mean and variance of environmental conditions may disrupt microbial  mutualisms by changing the costs and benefits of the interaction for each partner in ways that can cause the interaction to deteriorate \citep{aslan2013mutualism, fowler2024microbial}.
 Coral bleaching (the loss of symbiotic algae) due to temperature stress \citep{sully2019global} is perhaps the best known example, but this phenomenon is not unique to corals.
@@
 Beneficial symbionts are likely able to shield their hosts from environmental stress in locations that experience a small degree of change, but symbionts in locations that experience changes of large magnitude may be pushed beyond their physiological limits \cite{webster2008temperature}.
-Additionally, symbionts are often unevenly distributed across their hosts' distribution.
+Additionally, symbionts are often unevenly distributed across their host's distribution.
 Facultative symbionts may be absent from portions of the host range \cite{afkhami2014mutualist}, and hosts may engage with a diversity of partners (different symbiont species or locally-adapted strains) across their environments \cite{frade2008variation, rolshausen2018quantifying, fowler2023geographic}.
@@
 \emph{Epichloë} fungal endophytes are specialized symbionts of cool-season grasses, which have been documented in $\sim 30$\% of cool-season grass species \citep{leuchtmann1992systematics}.
-They are transmitted vertically from maternal plants to offspring through seeds.
+They are transmitted \tom{predominantly vertically}{Maybe two adverbs is awkward but I think we need to acknowledge the possibility of H transmission.} from maternal plants to offspring through seeds.
 Vertical transmission creates a feedback between the fitness of host and symbiont \citep{fine1975vectors, douglas1998host, rudgers2009fungus}. 
-Over time, endophytes that act as mutualists should rise in prevalence within a host population \citep{donald2021context}. 
+Over time, endophytes that act as mutualists should rise in prevalence within a host population, particularly under environmental conditions that elicit protective benefits \citep{donald2021context}. 
 \emph{Epichloë} are known to improve their hosts' drought tolerance \cite{decunta2021systematic} and protect their hosts against herbivores \cite{crawford2010fungal} and pathogens \cite{xia2018role} likely through the production of a diverse suite of alkaloids and other secondary metabolites.
@@
 Grasses are commonly collected and identified based on the presence of their reproductive structures, meaning that preserved specimens typically contain seeds, conveniently preserving the fungi along with their host plants on herbarium sheets.
-This creates the opportunity to leverage the unique spatio-temporal sampling of herbarium collections to examine the response of the symbiosis to historical climate change. 
+This creates the opportunity to leverage the unique spatio-temporal sampling of herbarium collections to examine the response of this symbiosis to historical climate change. 
 %Research using historical collections has clearly demonstrated other ecological signatures of a changing climate. 
@@
 
-In this study, we assessed the long-term responses of endophyte symbiosis to climate change through the use of herbarium specimens of three North American host grass species (\emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}).
+In this study, we assessed the long-term responses of \emph{Epichloë} endophyte symbiosis to climate change through the use of herbarium specimens of three North American host grass species (\emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}).
 We first addressed questions describing spatial and temporal trends in endophyte prevalence: (i) How has endophyte prevalence changed over the past two centuries?
@@
 We then addressed how climate change may be driving trends in endophyte prevalence by asking: (iii) What is the relationship between temporal trends in endophyte prevalence and associated changes in climate drivers?
-\linelabel{R2C21-b-begin}We predicted that aggregate endophyte prevalence would increase over time in tandem with climate warming, and that hotspots of endophyte change would correspond spatially to hotspots of climate change.
+\linelabel{R2C21-b-begin}We predicted that overall endophyte prevalence would increase over time in tandem with climate warming, and that localized hotspots of endophyte change would correspond spatially to hotspots of climate change.
 \linelabel{R2C21-b-end}
-Finally, we evaluated \revise{(iv) how our model, built on data from historic specimens, performed in an out-of-sample test - }using data on endophyte prevalence from contemporary surveys of host populations.
-To answer these questions we examined a total of $2,346$ historic specimens collected across eastern North America between 1824 and 2019, and evaluated model performance against contemporary surveys comprising 1,442 individuals from 63 populations collected between 2013 and 2020.
+Finally, we evaluated \revise{(iv) how our model, built on data from historic specimens, performed in an out-of-sample test }using data on endophyte prevalence from contemporary surveys of host populations.
+To answer these questions we examined a total of $2,346$ historic specimens collected across eastern North America between 1824 and 2019, and evaluated model performance against contemporary surveys comprising 1,442 individuals from 63 populations surveyed between 2013 and 2020.
 
@@
 \ref{fig:map}).
-Cool-season grasses typically grow actively during the cooler temperatures of spring and autumn due to their reliance on $C_3$ photosynthesis.}\linelabel{R2C23-end}
-\emph{A. hyemalis} is a small short-lived perennial species that germinates in spring and typically flowers between March and July (most common collection month: May).
+Cool-season grasses grow during the cooler temperatures of spring and autumn due to their reliance on $C_3$ photosynthesis.}\linelabel{R2C23-end}
+\emph{A.
+hyemalis} is a small short-lived perennial species that germinates in \tom{winter}{Winter, not spring.
+It's 'winter bentgrass'.} and typically flowers between March and July (most common collection month: May).
 \emph{A. perennans} is of similar stature but is longer lived than \emph{Agrostis hyemalis} and flowers in late summer and early autumn (most common collection month: September). 
-\emph{A. perennans} is more sparsely distributed, tending to be found in shadier and more moist habitats, while \emph{A. hyemalis} is commonly found in open and recently disturbed ground. 
+\emph{A. perennans} is more sparsely distributed, tending to be found in shadier and moister habitats, while \emph{A. hyemalis} is commonly found in open and recently disturbed habitats. 
 Both \emph{Agrostis} species are recorded from throughout the Eastern US, but \emph{A.
@@
 These traits contribute to highly specialized interactions between symbiont and host.
-Some host species have been shown to partner with multiple symbiont species in these genus, and in some cases multiple symbiont species can co-exist within a host population \cite{mc2014species}.
-However, suveys have typically found limited \emph{Epichloë} genotypic diversity within host populations\cite{treindl2023genetic}. 
+Some host species have been shown to partner with multiple symbiont species or strains, and in some cases multiple symbiont lineages can co-exist within a host population \cite{mc2014species}.
+However, surveys have typically found limited \emph{Epichloë} genotypic diversity within host populations\cite{treindl2023genetic}. 
 Across host populations, concentrations of biologically-active alkaloids and the genes associated with their production vary substantially \cite{schardl2012chemotypic}.
-In this analysis, we focus on the presence/absence of \emph{Epichloë} symbionts, and we discuss potential implications of hidden genotypic diversity in the Discussion.}
+In this analysis, we focus on the presence/absence of \emph{Epichloë} symbionts, and we discuss potential implications of symbiont genotypic diversity in the Discussion.}
 \linelabel{R1C3-end}
@@
 		\subsection*{Herbarium surveys}
-We visited nine herbaria between 2019 and 2022 (see Table \ref{table:herbaria} for a summary of specimens included from each collection). 
+We visited nine herbaria between 2019 and 2022 (see \tom{Table}{In table, maybe replace -- with 0.} \ref{table:herbaria} for a summary of specimens included from each collection). 
 With permission from herbarium staff, we acquired seed samples from $1135$ \emph{A.
@@
 %Our sampling plan was designed to minimize damage to these specimens.
-We chose our focal species in part because they are commonly represented in herbarium collections, and produce high numbers of seeds, meaning that small samples would not diminish the value of the specimens for future studies.
-We collected up to 5-10 seeds per specimen after examining the herbarium sheet under a dissecting microscope to ensure that we collected mature seeds, not florets or unfilled seeds, fit for our purpose of identifying fungal endophytes with microscopy.
+We chose our focal species in part because they are commonly represented in herbarium collections and produce many seeds, meaning that small samples would not diminish the value of the specimens for future studies.
+We collected 5-10 seeds per specimen after examining the herbarium sheet under a dissecting microscope to ensure that we collected mature seeds, not florets or unfilled seeds, fit for our purpose of identifying fungal endophytes with microscopy.
 We excluded specimens for which information about the collection location and date were unavailable.
+
 Each specimen was assigned geographic coordinates based on collection information recorded on the herbarium sheet using the geocoding functionality of the ggmap R package \citep{kahle2019package}.
@@
 Among those specimens geo-referenced at the state level, the largest bounding box, spanning the state of Texas, was 1233 km wide.
-The smallest bounding boxes were less than 1 km across for small municipalities (while this suggests high precision, we note that some specimens were collected in natural habitat nearby to small munipalities not encompassed by this bounding boxes).}\linelabel{R2C24-end}
+The smallest bounding boxes were less than 1 km across for small municipalities (while this suggests high precision, we note that some specimens were collected in natural habitat nearby to small municipalities not encompassed by this bounding boxes).}\linelabel{R2C24-end}
 
@@
 Texas A\&M, Missouri Botanic Garden) and other herbaria in the Southern Great Plains region of the United States.
-While these nine herbaria garnered specimens that span the focal species' ranges, our dataset unevenly samples across the study region. 
+While these nine herbaria garnered specimens that span the focal species' ranges, our dataset unevenly samples across the study \tom{region}{Reference Figure 1 here?}. 
 Texas, Oklahoma, Louisiana, and Missouri are the most represented states.
-Uneven sampling is most likely to be consequential for \emph{A. perennans}, which has much of its range in the northeastern US. 
+Uneven sampling was most pronounced for \emph{A. perennans}, which has much of its range in the northeastern US. 
 We explore the potential influence of spatial bias in sampling on our results through a simulation analysis (Appendix A - Supporting Methods).}\linelabel{R1C11-end}
@@
 and a negative \revise{identification for} the conservative score.
-\revise{The liberal status assumed a potential endophyte identification was more likely to be endophyte-positive while the conservative status assumed that the potential endophyte identification was less likely to be endophyte-positive}.
+\tom{\revise{The liberal status assumed a potential endophyte identification was more likely to be endophyte-positive while the conservative status assumed that the potential endophyte identification was less likely to be endophyte-positive}.
+}{I don't think this sentence is necessary.}
 $89$\% of scored plants had matching liberal and conservative scores, reflecting high confidence in endophyte status.
-The following analyses used the liberal status, \revise{however} we repeated all analyses with the conservative status which yielded qualitatively similar results (Fig.
+The following analyses used the liberal status, \revise{however} repeating all analyses with the conservative status yielded qualitatively similar results (Fig.
 \ref{fig:conservative_scores_plot}).\linelabel{R2C26-end}
@@
 For each host species $h$, endophyte presence/absence of the $i^{th}$ specimen ($P_{h,i}$) was modeled as a Bernoulli response variable with expected probability of endophyte occurrence $\hat{P}_{h,i}$.
-We modeled $\hat{P}_{h,i}$ as a linear function of intercept $\boldsymbol{A}_h$ and slope $\boldsymbol{T}_{h}$ defining the global trend in endophyte prevalence specific to each host species as well as with spatially-varying intercepts $\alpha_{h,l_{i}} $ and slopes $\tau_{h,l_i}$ associated with location ($l_i$, the unique latitude-longitude combination of the $i$th observation).
-\linelabel{R2C28-begin}The joint-model structure allowed us to ``borrow \revise{information}'' \linelabel{R2C28-end} across species in the estimation of shared variance terms for the spatially-dependent random effect $\delta_{l_i}$, intended to account for residual spatial variation, and $\chi_{c_i}$ and $\omega_{s_i}$ i.i.d.-random effects indexed for each collector identity ($c_i$), and scorer identity ($s_i$) of the $i^{th}$ specimen.
+We modeled $\hat{P}_{h,i}$ as a linear function of \tom{intercept $\boldsymbol{A}_h$ and slope $\boldsymbol{T}_{h}$}{Are these matrices?
+I am concerned that Eq 1 does not make mathematical sense if it's saying scalar = matrix + matrix*scalar + scalar...
+etc.} defining the global trend in endophyte prevalence specific to each host species as well as with spatially-varying intercepts $\alpha_{h,l_{i}} $ and slopes $\tau_{h,l_i}$ associated with location ($l_i$, the unique latitude-longitude combination of the $i$th observation).
+\linelabel{R2C28-begin}The joint-model structure allowed us to ``borrow \revise{information}'' \linelabel{R2C28-end} across species in the estimation of shared variance terms for the spatially-dependent random effect $\delta_{l_i}$, intended to account for residual spatial variation, and $\chi_{c_i}$ and $\omega_{s_i}$, the i.i.d.-random effects indexed for each collector identity ($c_i$) and scorer identity ($s_i$) of the $i^{th}$ specimen.
 
@@
 \linelabel{R2C30-begin}
-\revise{We modeled the geographic distribution of each host species with two goals: (1) generate realistic maps on which we could project the predictions of the INLA model, and (2) use the geographic distributions to test for relationships between climate change drivers and trends in endophyte prevalence.
-The herbarium records did not encompass the entirety of each host species' range.}\linelabel{R2C30-end}
+\revise{The herbarium records did not encompass the entirety of each host species' range.
+Therefore, we used additional data sources to model the geographic distribution of each host species, with two goals: (1) generate realistic maps on which we could project the predictions of the INLA model, and (2) use the geographic distributions to test for relationships between climate change drivers and trends in endophyte prevalence.
+}\linelabel{R2C30-end}
 We followed the ODMAP (overview, data, model, assessment, prediction) protocol \citep{crossley2022opposing} (see Supporting Methods). 
-In short, we used presence-only observations of each host species from Global Biodiversity Information Facility (GBIF) between 1990 to 2020.
+\tom{In short, we used presence-only observations of each host species from Global Biodiversity Information Facility (GBIF) between 1990 to 2020.}{Maybe say how many records there were for each species?}
 %To reduce the potential influence of sampling bias and spatial autocorrelation, \tom{we thinned the occurrences to the spatial scale of our selected climatic predictors}{I do not follow.}. 
@@
 %\tom{Among this suite of variables, we chose to include, which were uncorrelated (Variance Inflation Factor $>$ 0.7) and allowed us to predict the occurrence probability of each host species in space and time.}{Do not follow and sounds circular.}
-We fit maximum entropy (MaxEnt) models using the maxent function in the R package dismo \citep{hijmans2017package} using the same set of seasonal climate predictors considered \revise{below} calculated for the 1990-2020 climate normals: mean and standard deviation of spring, summer, and autumn temperature, and mean and standard deviation of spring, summer, and autumn cumulative precipitation.
+We fit maximum entropy (MaxEnt) models using the maxent function in the R package dismo \citep{hijmans2017package} using following seasonal climate predictors (1990-2020 climate normals): mean and standard deviation of spring, summer, and autumn temperature, and mean and standard deviation of spring, summer, and autumn cumulative precipitation.
 %\josh{MaxEnt is preferred because it has been shown to generate response curves with less unpredictable behavior when applied to new climates \citep{hijmans2006ability}.}{possibly could remove this sentence?}
@@
 We found AUC values of 0.862, 0.838, 0.821 respectively  for \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus} indicating good model fit to data.
-To convert the continuous predicted probabilities into binary presence - absence maps on which we projected INLA predictions, we used the  training sensitivity (true positive rate) and specificity threshold (true negative rate) \citep{liu2005selecting}.
+\tom{To convert the continuous predicted probabilities into binary presence - absence maps on which we projected INLA predictions, we used the training sensitivity (true positive rate) and specificity threshold (true negative rate) \citep{liu2005selecting}.}{I don't really understand this sentence.
+How do the training sensitivity and specificity threshold lead to presence/absence maps?}
 %These binary maps serve as boundaries in presented maps of change in endophyte prevalence, and outline the set of pixels used in our analysis of climate correlates with trends in endophyte prevalence
@@
 We first downloaded monthly temperature and precipitation rasters from the PRISM climate group \citep{daly2013prism} covering the time period between 1895 and 2020 using the 'prism' R package \citep{Rprism2015}.
-Prism provides reconstructions of historic climate variables across the United States by spatially-interpolating weather station data \citep{diLuzio2008constructing}. 
+Prism provides reconstructions of historic climate variables across the United States by spatially interpolating weather station data \citep{diLuzio2008constructing}. 
 We calculated 30-year climate normals for seasonal mean temperature and cumulative precipitation for the recent (1990 to 2020) and historic (1895 to 1925) periods.
 We used three four-month seasons within the year (Spring: January, February, March, April; Summer: May; June, July, August; Autumn: September, October, November, December). 
-This division of seasons allowed us to quantify differences in climate associated with the two ``cool'' seasons, when we expected our focal species to be most biologically active (\emph{A.
+This division of seasons allowed us to quantify differences in climate associated with the two ``cool'' seasons, when we expected our focal species to be most active (\emph{A.
 hyemalis} flowering phenology: spring; \emph{E.
@@
 We then took the difference between recent and historic periods for the mean and standard deviation for each climate driver (Figs. \ref{fig:AGHY_climate_covariates}-\ref{fig:ELVI_climate_covariates}).
-All together, we assessed twelve potential climate drivers: the mean and standard deviation of spring, summer, and autumn temperature, as well as the mean and standard deviation of spring, summer, and autumn cumulative precipitation.
+All together, we assessed twelve potential climate drivers: the mean and standard deviation of spring, summer, and autumn temperature, as well as the mean and standard deviation of spring, summer, and autumn cumulative precipitation (the same climate covariates used in the MaxEnt models).
 
@@
 To do so, we modeled the fitted, spatially-varying slopes of endophyte change through time ($\tau_{[h]l}$) as a linear function of environmental covariates, with a Gaussian error distribution for a set of pixels across each host distribution.
-The continuous SPDE approach taken for our endophyte prevalence model allows us to generate predictions of temporal trends in prevalence at arbitrarily many pixels across each host distribution.
+The continuous SPDE approach taken from our endophyte prevalence model allows us to generate predictions of temporal trends in prevalence at arbitrarily many pixels across each host distribution.
 Balancing computation time with resolution, we generated predicted trends for 546, 645, and 753 pixels across each host distribution for \emph{A.
@@
 hyemalis} = 61km x 45 km; \emph{E.
-virginicus} = 62 km x 40 km )}\linelabel{R2C32-end}
-Fitting regressions to many pixels across the study region risks articially inflating confidence in our results due to large sample sizes, and so we performed this analysis using only a random subsample of 250 pixels across the study region; other sizes of subsample yielded similar results.
+virginicus} = 62 km x 40 km ).}\linelabel{R2C32-end}
+Fitting regressions to many pixels across the study region risks artificially inflating confidence in our results due to large sample sizes, and so we performed this analysis using only a random subsample of 250 pixels across the study region; other sizes of subsample yielded similar results.
 Data from each host species were analyzed separately. 
-\revise{Through the results and discussion that follow, we refer to this analysis as the ``\emph{post hoc} climate regression analysis''.}
+\revise{Throughout the results and discussion that follow, we refer to this analysis as the ``\emph{post hoc} climate regression analysis''.}
 
 \subsection*{Validating model performance with in-sample and out-of-sample tests}
-We evaluated the predictive ability of the model using both in-sample training data from the herbarium surveys, and with out-of-sample test data, an important but rarely used strategy in ecological studies \cite{tredennick2021practical, lee2024phenological}.
+We evaluated the predictive ability of the endophyte prevalence model using both in-sample training data from the herbarium surveys, and with out-of-sample test data, an important but rarely used strategy in ecological studies \cite{tredennick2021practical, lee2024phenological}.
 We generated out-of-sample test data from contemporary surveys of endophyte prevalence in natural populations of \emph{A. hyemalis} and \emph{E. virginicus} in Texas and the southern US. 
@@
 During surveys, we collected seeds from up to 30 individuals per population (average number of plants sampled per population: $22.9$); note that this sampling design provided greater local depth of information than the herbarium records, where only one plant was sampled at each locality.
-We quantified the endophyte status of each individual with staining microscopy as described for the herbarium surveys (with 5-10 seeds scored per individual), and calculated the prevalence of endophytes within the population (proportion of plants that were endophyte-symbiotic).
+We quantified the endophyte status of each individual with microscopy as described for the herbarium surveys (with 5-10 seeds scored per individual), and calculated the prevalence of endophytes within the population (proportion of plants that were endophyte-symbiotic).
 For each population, we compared the observed fraction of endophyte-symbiotic hosts to the predicted probability of endophyte occurrence $\hat{P}$ derived from the model for that location and year. 
-The contemporary survey period (2013-2020) is at the most recent edge of the time period encompassed by the historical observations used for model fitting.
+The contemporary survey period (2013-2020) is at the most recent edge of the time period encompassed by the historical specimens used for model fitting.
 %We compared the model's prediction for these locations to the observed population prevalence.
@@
 hyemalis} increased from $\sim 25$\% to over $50$\% prevalence.
-\linelabel{R2C33-begin}Our model indicates a high \revise{confidence}\linelabel{R2C33-end} that overall temporal trends are positive across species (99\% probability of a positive overall year slope in \emph{A.
+\linelabel{R2C33-begin}Our model indicates high \revise{confidence}\linelabel{R2C33-end} that overall temporal trends are positive across species (99\% probability of a positive overall year slope in \emph{A.
 hyemalis}, 92\% probability of a positive overall year slope in \emph{A.
@@
 	\caption[Temporal trends in endophyte prevalence.]{\textbf{Temporal trends in endophyte prevalence.} (A) Histograms show the frequency of scored specimens through time for each host species.
-(B) Lines show predicted mean endophyte prevalence over the study period along with the 50\%  and 95\% CI bands incorporating uncertainty associated with collector and scorer random effects.
-\revise{Trends are estimated from the endophyte prevalence model.} Colored points are binned means of the observed endophyte presence/absence data (black dashes).
+(B)\revise{ Lines show mean endophyte prevalence predicted by the endophyte prevalence model} over the study period along with the 50\%  and 95\% CI bands incorporating parameter uncertainty and variation associated with collector and scorer random effects.
+Colored points are binned means of the observed endophyte presence/absence data (black dashes).
 Colors represent each host species (\emph{A.
@@
 The identity of individual scorers,\linelabel{R2C35-b-begin} \revise{the researchers who identified endophyte status microscopically},\linelabel{R2C35-b-end} did contribute to observed patterns in endophyte prevalence.
-For example, $3$ of the $25$ scorers were more consistently likely than average to assign positive endophyte status, as indicated by 95\% credible intervals greater than zero (Fig.
+\tom{For example, $3$ of the $25$ scorers were significantly more likely than average to assign positive endophyte status, as indicated by 95\% credible intervals greater than zero (Fig.
 \ref{fig:scorer_fx}).
+}{And I count four that have significantly negative CIs.}
 
@@
 virginicus} had an overall weaker and geographically more homogeneous increase in endophyte prevalence.
-Notably, endophytes increased most strongly towards the western range edge of \emph{A.
+Notably, endophytes are predicted to have increased most strongly towards the western range edge of \emph{A.
 hyemalis} (Fig.
@@
 virginicus} trends did not differ according to historic prevalence (\ref{fig:initialprev_trend_plot} C).} \linelabel{R1C7-end}
-Posterior estimates of uncertainty in spatially varying slopes indicate that these hotspots of change may have experienced increases of up to $5$\% per year while declines in prevalence may be as great as $4$\% per year for \emph{A.
+Posterior estimates of uncertainty in spatially varying slopes indicate that these hotspots of change may have experienced increases of up to $5$\% per year while declines in prevalence may be as great as \tom{$4$\% per year}{Should this be -4\%?} for \emph{A.
 hyemalis} and \emph{A.
@@
 	\includegraphics[width = \linewidth]{../Plots/svc_time_map.png}
-	\caption[Predicted posterior mean of spatially-varying slopes representing change in endophyte prevalence for each host species.]{Predicted posterior mean of spatially-varying slopes representing change in endophyte prevalence for each host species.
+	\caption[Predicted posterior mean of spatially-varying slopes representing change in endophyte prevalence for each host species.]{Predicted posterior mean of spatially-varying slopes representing change in endophyte prevalence for each host species (A,\emph{A.
+hyemalis}; B,\emph{A.
+perennans}; C,\emph{E.
+virginicus}).
 \revise{Spatially-varying trends are estimated from the endophyte prevalence model.} Color indicates the relative change in predicted endophyte prevalence.
@@
 \revise{Predictions of prevalence come from the endophyte prevalence model.} Color indicates the posterior mean endophyte prevalence for \emph{A.
-hyemalis} in (A) 1895 and (D) 2020, for \emph{A.
-perennans} in (B) 1895 and (E) 2020, and for \emph{E.
-virginicus} in (C) 1895 and (F) 2020.
+hyemalis} (A,D), \emph{A.
+perennans} (B,E), and \emph{E.
+virginicus} (C,F).
 Map lines delineate study areas and do not necessarily depict accepted national boundaries.}}
@@
 \subsection*{What is the relationship between variation in temporal trends in endophyte prevalence and changes in climate drivers?}
-We found that trends in endophyte prevalence were strongly associated with seasonal climate change drivers (Fig. \ref{fig:climate_correlation_plot}).
-For the majority of the study region, the climate has become wetter (an average increase in annual precipitation of $60$ mm.) with relatively little temperature warming (an average increase in annual temperature of $0.02$ $^{\circ}$C) over the last century (Fig.
+We found that trends in endophyte prevalence were strongly associated with one or more seasonal climate change drivers (Fig. \ref{fig:climate_correlation_plot}).
+For the majority of the study region, the climate has become wetter (an average increase in annual precipitation of $60$ mm) with relatively little temperature warming (an average increase in annual temperature of $0.02$ $^{\circ}$C) over the last century (Fig.
 \ref{fig:AGHY_climate_covariates}-\ref{fig:ELVI_climate_covariates}), a consequence of regional variation in global climate change \cite{ipcc_2021}.
-Within the region, climate changes were spatially variable; certain locations experienced increases in annual precipitation as large as $375$ mm.
-or decreases up to $54$ mm.
-across the last century, while annual temperature changes ranged from warming as great as $1.4$ $^{\circ}$C to cooling by $0.46$ $^{\circ}$C.
+Within the region, climate changes were spatially variable; certain locations experienced increases in annual precipitation as large as $375$ mm or decreases up to $54$ mm across the last century, while annual temperature changes ranged from warming as great as $1.4$ $^{\circ}$C to cooling by $0.46$ $^{\circ}$C.
 Spatially variable climate trends were predictive of trends in endophyte prevalence.
@@
 For this species, a $1$ $^{\circ}$C increase in autumn temperature was associated with a $1.07$ \% increase per year in endophyte prevalence (Fig.
-\ref{fig:climate_correlation_plot}A) and a $100$ mm.
-increase in precipitation was associated with a $0.8$\% increase per year in endophyte prevalence (Fig.
+\ref{fig:climate_correlation_plot}A) and a $100$ mm increase in precipitation was associated with a $0.8$\% increase per year in endophyte prevalence (Fig.
 \ref{fig:climate_correlation_plot}B).
-This result aligns with the species' autumn active growing season, however other seasonal climate drivers were also associated with increasing endophyte prevalence. 
-In particular, we found cooler and drier springs and cooler summers to be associated with increasing endophyte prevalence (greater than $99$\% posterior probabilities of negative slopes) however these slopes were generally of smaller magnitude than those for autumn climate drivers.
+This result aligns with the species' autumn active growing season, however other seasonal climate drivers were also associated with increasing endophyte prevalence in this host species. 
+In particular, we found cooler and drier springs and cooler summers to be associated with increasing endophyte prevalence (greater than $99$\% posterior probabilities of negative slopes), though these slopes were generally of smaller magnitude than those for autumn climate drivers.
 
@@
 Like \emph{A. perennans}, climate during peak growing season (spring for \emph{A. perennans} and summer for \emph{E. virginicus}) emerged most commonly as drivers of changes in endophyte prevalence.
-Increases in mean spring preciptation were the strongest predictor of increasing trends in endophyte prevalence for \emph{A.
+Increases in mean spring precipitation were the strongest predictor of increasing trends in endophyte prevalence for \emph{A.
 hyemalis} (Fig.
 \ref{fig:climate_correlation_plot}B) (greater than $99$\% posterior probability of a positive slope).
-For this species, an increase of $100$ mm. in spring precipitation led to an increase of $0.6$\% per year in endophyte prevalence.
+For this species, an increase of $100$ mm in spring precipitation led to an increase of $0.6$\% per year in endophyte prevalence.
 The next greatest slopes were those associated with variability in spring precipitation (greater than $96$\% posterior probability of a negative slope), as well as in the mean and variability of autumn climate (greater than $98$\% probability of negative and positive slopes, respectively).
@@
 \linelabel{R2C38-begin}\revise{Our analysis indicated relatively high confidence that these climate drivers influence endophyte prevalence shifts in \emph{E.
-virginicus}}(greater than $99$\% posterior probability of either negative or positive slopes respectively), however they translate to less than $0.2$\% change in endophyte prevalence per year for a change of $100$ mm.
-change in precipitation over the century.\linelabel{R2C38-end}
+virginicus}}(greater than $99$\% posterior probability of either negative or positive slopes respectively), \tom{however they translate to less than $0.2$\% change in endophyte prevalence per year for a change of $100$ mm change in precipitation over the century.\linelabel{R2C38-end}}{I am not sure why you are highlighting the slope with respect to mm of precip, since the preceding sentences emphasize temperature and var(precip).}
 \linelabel{R2C37-begin}Repeating this analysis using all pixels across each species' distribution were qualitatively similar to these results.\linelabel{R2C37-end}
@@
 \subsection*{Evaluation of model performance on an out-of-sample test}
-Tests of models' predictive performance as quantified by AUC and by visual posterior predictive checks, indicated good predictive ability. 
+Tests of \tom{models'}{Which models are you referring to?
+Is this plural because each species had a different model?} predictive performance as quantified by AUC and by visual posterior predictive checks, indicated good predictive ability.
 Model performance was similar between historic herbarium specimens used as training data and the out-of-sample test data from contemporary surveys (AUC = 0.79 and 0.77 respectively; Fig.
@@
 hyemalis} towards western longitudes (Fig.
-\ref{fig:contemptestplot}A) and northern latitudes (Fig.
+\ref{fig:contemptestplot}A) and \tom{northern latitudes}{It's actually an increase toward N latitudes, not decline.} (Fig.
 \ref{fig:contemptestplot}B).
-\revise{It is noteable that} model predictions for endophyte prevalence exhibited relatively little local geographic variation, whereas the out-of-sample survey data were maximally variable with populations spanning 0\% to 100\% endophyte-symbiotic plants (Fig.
-\ref{fig:contemptestplot}C).
+\revise{It is noteable that} model predictions for endophyte prevalence exhibited relatively little local geographic variation, whereas the out-of-sample survey data were highly variable with populations spanning 0\% to 100\% endophyte-symbiotic plants (Fig.
+\ref{fig:contemptestplot}C), \tom{indicating that...}{I would finish this train of thought, even as you leave most of it for the discussion.}.
 
@@
 \linelabel{R2C40-begin}Our examination of historic plant specimens revealed \revise{previously hidden} \linelabel{R2C40-end}shifts in microbial symbiosis over the last two centuries. 
-For the three host species we examined, there have been strong increases in prevalence of\linelabel{R2C41-begin} \revise{\emph{Epichloë} endophyte symbiosis}.\linelabel{R2C41-end}
+For the three grass host species we examined, there have been strong increases in prevalence of\linelabel{R2C41-begin} \revise{\emph{Epichloë} endophyte symbiosis}.\linelabel{R2C41-end}
 We interpret increases in prevalence of \emph{Epichloë}, which are vertically transmitted, as adaptive changes that improve the fitness of their hosts under increasing environmental stress.
-This interpretation is in line with theory predicting that the positive fitness feedback caused by vertical transmission leads beneficial symbionts to rise in prevalence within a population \cite{fine1975vectors,donald2021context}.
+This interpretation is in line with theory predicting that positive fitness feedback caused by vertical transmission leads beneficial symbionts to rise in prevalence within a population \cite{fine1975vectors,donald2021context}.
 We further found that trends in endophyte prevalence varied across the distribution of each species in association with changes in climate drivers, suggesting that the increases in endophyte prevalence are driven by context-dependent benefits to hosts that confer resilience under environmental change.
@@
 Differences across host species underscore that while all of these $C_3$ grasses share similar broad-scale distributions, each engages in unique biotic interactions and has unique responses to environmental drivers.
-We identified hotspots of change for \emph{A. perennans}, which was the species that experienced the strongest absolute changes in endophyte prevalence (Fig. \ref{fig:svc_time_map}).
-Declines of $0.9$\% per year in the southern portion of its range and increases of up to $2$\% per year in the north suggest a potential poleward range shift of endophyte-symbiotic plants (whether the overall host distribution is shifting in parallel is an exciting next question.)
+We identified hotspots of change for \emph{A.
+perennans}, \tom{which was the species that experienced the strongest absolute changes in endophyte prevalence (Fig.
+\ref{fig:svc_time_map}).}{Not sure what this is based on.
+In Fig 2 both AGrostis species look like they have similar increases.}
+Predicted declines of $0.9$\% per year in the southern portion of its range and predicted increases of up to $2$\% per year in the north suggest a potential poleward range shift of endophyte-symbiotic plants; whether the overall host distribution is shifting in parallel is an exciting next question.
+
 Based on previous work demonstrating that endophytes can shield their hosts from drought stress (reviewed in \citet{decunta2021systematic}), we generally predicted that drought conditions would be a driver of increasing endophyte prevalence.
-In contrast to this expectation, increasing prevalence for \emph{A. perennans} were associated with increasing autumn temperature and precipitation (Fig. \ref{fig:climate_correlation_plot}). 
+In contrast to this expectation, increasing prevalence for \emph{A. perennans} was associated with increasing autumn temperature and precipitation (Fig. \ref{fig:climate_correlation_plot}). 
 To our knowledge, the response of the symbiosis in \emph{A.
@@
 For \emph{A. hyemalis}, endophytes could be playing a role helping hosts weather autumn-season droughts, which \revise{is likely also} an important time for the species' germination.
-Previous work has demonstrated drought benefits in a greenhouse manipulation with this species \citep{davitt2011understanding}, and early life stages may be particularly vulnerable to prolonged droughts.
+Previous work demonstrated drought benefits in a greenhouse manipulation with this host-symbiont pair \citep{davitt2011understanding}, and early life stages may be particularly vulnerable to prolonged droughts.
 %It is possible that the study region has not experienced a magnitude of climate change great enough to cause larger changes in endophyte prevalence for this species. 
@@
 For \emph{E.
-virginicus}, which experienced the most modest changes in endophte prevalence overall (ranging between $1.1$\% increases and $0.2$\% decreases), we only found modest associations with changes in climate drivers.
+virginicus}, which experienced the weakest changes in endophyte prevalence overall (ranging between $1.1$\% increases and $0.2$\% decreases), we only found modest associations with changes in climate drivers.
 Surveys by \citet{sneck2017variation}, used as part of the test data in this study, identified a drought index (SPEI) that integrates precipitation with estimated evapotranspiration as an important predictor of \revise{contemporary} endophyte prevalence \linelabel{R2C14-begin}\revise{in this species.
@@
 \revise{While we show consistent increasing trends in prevalence between the three species, the mechanisms that explain these changes may be diverse and idiosyncratic.
-\linelabel{R2C45-begin}First, climate change responses may depend on genotype-specific responses that are ignored in our current analysis. 
-While \emph{Epichloë} symbioses are highly specialized, surveys have demonstrated genotypic and chemotypic diversity of the symbionts across populations and even within populations \cite{von2021genetic, treindl2023genetic}.
+\linelabel{R2C45-begin}First, climate change responses may depend on genotype-specific responses that are not considered in our current analysis. 
+While \emph{Epichloë} symbioses are highly specialized, surveys have demonstrated genotypic and chemotypic diversity of the symbionts among and within populations \cite{von2021genetic, treindl2023genetic}.
 Genotypic variation in \emph{Epichloë} endophytes, particularly in genes responsible for alkaloid production, produces ``chemotypes" with differing benefits for hosts against insect or mammalian herbivores mediated by environmental conditions \cite{schardl2012chemotypic,saikkonen2013chemical}.
@@
 \revise{Broad changes in the distribution and abundance of natural enemies, along with stresses from anthropogenic changes in landcover and pollution \cite{sage2020global} likely influence the benefits of symbiosis \cite{rudgers2020climate}.
-The historic trends that we observed result from the combination of these fitness benefits playing out across the heterogenous map of a changing climate.\linelabel{R2C44-end}
+The historic trends that we observed result from the combination of these fitness benefits playing out across the heterogeneous map of a changing climate and other anthropogenic drivers.\linelabel{R2C44-end}
 
@@
 Based on our analysis, it is likely that the symbiosis will facilitate range shifts for hosts by improving population growth at range edges.
-Previous population surveys attributed environment-dependent gradients in endophyte prevalence \citep{sneck2017variation,rudgers2009benefits,semmartin2015broad}to symbiont-derived fitness benefits allowing their hosts to persist in environments where they otherwise could not \citep{afkhami2014mutualist,kazenel2015mutualistic}.
+Previous population surveys \citep{sneck2017variation,rudgers2009benefits,semmartin2015broad} attributed environment-dependent gradients in endophyte prevalence to symbiont-derived fitness benefit,s allowing hosts to persist in environments where they otherwise could not \citep{afkhami2014mutualist,kazenel2015mutualistic}.
 However, symbiont-facilitated range shifts require that endophytes be present in the populations to be able to support population growth. 
@@
 \ref{fig:svc_space_year_map}), and while prevalence has increased most quickly in the regions with historically low endophyte prevalence (Fig.
-\ref{fig:initialprev_trend_plot}), the complete absence of endophytes range edges would make it impossible for prevalence to increase without dispersal of symbiotic seeds \cite{fowler2023geographic}.
+\ref{fig:initialprev_trend_plot}), the complete absence of endophytes at range edges in this species would make it impossible for prevalence to increase without dispersal of symbiotic seeds \cite{fowler2023geographic}.
 These factors potentially contribute to the ability of the host species to track its environmental niche. 
-Another interesting question is the degree to which symbiotic and non-symbiotic hosts, which occupy overlapping but distinct niches, are likely to experience distribution shifts in tandem or at different spread rates in future work.\linelabel{R2C42-end}
+Another interesting question is the degree to which symbiotic and non-symbiotic hosts, which occupy overlapping but distinct niches, are likely to experience distribution shifts in tandem or at different rates in future work.\linelabel{R2C42-end}
 More extreme climate stresses, which are expected more frequently in the future \cite{seneviratne2021weather}, have the potential to alter the costs and benefits of the interaction. 
-The past indicates a resilient interaction, but understanding the future climate conditions that may tip this interaction to deteriote will be crucial.}\linelabel{R2C14-end}
+The past indicates a resilient symbiosis, but it will be crucial to understand whether more extreme future climate conditions could tip this interaction to deteriote.}\linelabel{R2C14-end}
 
 \revise{\subsubsection*{Steps towards forecasts of host-microbe symbioses}}
-The combination of a spatially-explicit model and historic herbarium specimens allowed us to identify regions of both increasing and decreasing endophyte prevalence, however we see several next steps towards the goal of predicting host and symbiont niche-shifts in response to future climate change.
-While the model recreated the large-scale spatial trends observed in contemporary population surveys, test data contained more population-to-population variability in prevalence. 
+The combination of a spatially-explicit model and historic herbarium specimens allowed us to identify regions of both increasing and decreasing endophyte prevalence, however we see several next steps toward the goal of predicting host and symbiont niche-shifts in response to future climate change.
+While the model recreated the large-scale spatial trends observed in contemporary population surveys, test data contained more population-to-population variability in prevalence than could be explained by the model.
 \linelabel{R2C38-begin}\revise{We interpret this to mean that the model captures coarse-scale spatial and temporal trends reasonably well, but is not equipped to capture local-scale nuances that generate population-to-population differences.} \linelabel{R2C38-end}
-Validating our model predictions \revise{with this test}, a rare extra step in collections-based studies, allows us to evaluate places to improve the model's out-of-sample predictive ability.
+Validating our model predictions \revise{with this test}, a rare extra step in collections-based studies, allows us to identify ways in which the model's out-of-sample predictive ability could be improved.
 \linelabel{R2C43-begin}Lack of information on local variability \revise{in symbiont prevalence} may simply be a feature of data derived from herbarium specimens. 
-\revise{Natural history collectors sample one or a few specimens from local populations, which our analyses aggregates across to derive broad-scale model estimates.}\linelabel{R2C43-end}
+\revise{Natural history collectors sample one or a few specimens from local populations, and these observations are aggregated by the model to derive broad-scale estimates.}\linelabel{R2C43-end}
 This suggests that increasing local replication should be a factor considered in future collection efforts of natural history specimens, balanced with the required time and effort \linelabel{R1C9-begin} \revise{and with limitations on storage space within collections}.
 \linelabel{R1C9-end}
-Poor predictive ability at local scales in this grass-endophyte system is not surprising, as previous studies have found that local variation, even to the scale of hundreds of meters can structure endophyte-host niches \cite{kazenel2015mutualistic}.
+Poor predictive ability at local scales in this grass-endophyte system is not surprising, as previous studies have found that local variation, \tom{even to the scale of hundreds of meters}{Also see new paper in Functional Ecology by Pedro Gundel's group that has very local scale differences (I am a coauthor but did not do much.).} can structure endophyte-host niches \cite{kazenel2015mutualistic}.
 An important step would be integrating data from local and regional scales  through modeling to constrain estimates of local and regional variation.
@@
 hyemalis} could tell us whether low endophyte prevalence in that area (Fig.
-\ref{fig:svc_space_year_map}A) is a result of environmental conditions that lead the symbiosis to negative fitness consequences, or is a result of some historical contingency or dispersal limitation that has thus far limited the presence of symbiotic hosts from a region where they would otherwise flourish and provide resilience.
+\ref{fig:svc_space_year_map}A) is a result of environmental conditions that lead the symbiosis to have negative fitness consequences, or is a result of some historical contingency or dispersal limitation that has thus far limited the presence of symbiotic hosts from a region where they would otherwise flourish and provide resilience.
 Incorporating available climatic and soil layers as covariates is another obvious step that could improve predictions.
-These steps will bridge gaps that often exist between large but broad bioclimatic and biodiversity data and small but local data on biotic interactions, and move towards the goal of predicting the dynamics of microbial symbioses under climate change \cite{miller2019recent, isaac2020data}.
+These steps will bridge gaps that often exist between large but broad bioclimatic and biodiversity data and small but high-resolution data on biotic interactions, and move towards the goal of predicting the dynamics of microbial symbioses under climate change \cite{miller2019recent, isaac2020data}.
 
@@
 Previous studies that have quantified the effects of collector biases typically find them to be small  \cite{davis2015herbarium,meineke2019herbarium}, and we similarly did not find that collector has a strong effect on the results of our analysis, but that scorer identity did impact results.
-\linelabel{R2C36-begin}\revise{It is difficult to distinguish whether the impact of scorers was driven by true differences in scorers' biases during the seed scoring process or by unintended spatial or temporal clustering of the specimens examined by each scorer \cite{clayton1993spatial,urdangarin2023evaluating}.
+\linelabel{R2C36-begin}\revise{It is difficult to distinguish whether the impact of scorers was driven by true differences in scorers' biases or by unintended spatial or temporal clustering of the specimens examined by each scorer \cite{clayton1993spatial,urdangarin2023evaluating}.
 By under-weighting endophyte-positive samples that are clustered spatially or by collector or observer, the endophyte prevalence model is appropriately accounting for nuisance variables and providing a conservative inference of endophyte change relative to the raw data.
@@
 \linelabel{R1C4-dicussion-begin}\revise{
-Spatial autocorrelation is another \jacob{phenomenom}{phenomenon ?} likely common in data derived from herbarium specimens \cite{willems2022forest}, which our spatially-explicit analysis models among samples.
+Spatial autocorrelation is another phenomenon likely common in data derived from herbarium specimens \cite{willems2022forest}, which our spatially-explicit analysis models among samples.
 Beyond spatial autocorrelation of outcomes, systematic differences in sampling across space can result in spatial bias. 
@@
 km, including specimens from Mexico and Canada.
-The specimens we examined are concentrated in the South-Central United States, with fewer specimens in the rapidly warming northeastern United States.
+The specimens we examined are concentrated in the south-central United States, with fewer specimens in the rapidly warming northeastern United States.
 We provide a simulation analysis exploring the potential impact of spatially and temporally biased sampling (Appendix A - Supporting Methods). 
@@
 Jessica Budke for help in drafting our initial destructive sampling plan, and to the many staff members of herbaria who facilitated our research visits, as well as to the hundreds of collectors who contributed to the natural history collections.
-	Several high schooler and undergraduate researchers contributed to data collection, including A.
+	Several high school and undergraduate researchers contributed to data collection, including A.
 Appio-Riley, P.
@@
 	This research was supported by funding from National Science Foundation (grants 1754468 and 2208857) and by funding from the Texas Ecolab Program.
+\tom{Two anonymous reviewers greatly improved an earlier version of this manuscript.}{Always thank the reviewers!}
 
@@
 \newpage{}
+	\tom{}{I am not sure if it's because I messed with some of the font packages, but some of the citations are a litte messy (some lowercase journal names, some uppercase).
+Latin names in titles should always be italicized.
+They will ask you to clean this up sooner or later, so I would do it sooner.}
 	\bibliographystyle{plainnat}

### 2025-02-13 Analyses/posthoc_climate_correlation.R
minor axis change

@@ -883,7 +883,7 @@ ppt_sd_trend <- ggplot(filter(prediction_df, grepl("ppt", name) & grepl("sd", na
   geom_ribbon(aes(value, ymin = q0.25, ymax = q0.75, fill = species), alpha = 0.2) +
   facet_grid(species ~ name,
              labeller = labeller(name =climate_labels),
-             scales = "free")+
+             scales = "free_x")+
   scale_color_manual(values = species_colors)+
   scale_fill_manual(values = species_colors)+
   scale_linetype_manual(values = c("dashed", "solid") )+
@@ -1258,6 +1258,14 @@ elvi_ppt_sd_trend <- ggplot(filter(prediction_df, grepl(species_names[3], specie
 ELVI_climate_trends_plot <- (elvi_tmean_trend + elvi_ppt_trend ) / (elvi_tmean_sd_trend+elvi_ppt_sd_trend)
 
 ggsave(ELVI_climate_trends_plot, filename = "Plots/ELVI_climate_trends_plot.png", width = 12, height = 8)
+
+
+
+
+
+
+###### generating some summaries for the paper ##############
+
 
 
 

### 2025-02-13 Manuscript/Reviewer_response_endo_herbarium.tex
completed response to reviewers

@@
 Indeed, there can be genetic variation of symbionts within and across host populations, which we now describe in more detail to provide readers with this context in our methods section.
-In addition, we return to this topic in expanded discussion material, \tom{where we...}{In general I would be more thorough in your responses so that you actually answer or address the reviewers' questions here in the letter, in addition to the line numbers for the revised text.
-In this example I would acknowledge here that we can track overall prevalence but not genetic variation or change, and this could be an interesting future direction.
-Anything that could be seen as a weakness, flip it and make it a direction for future research.}
+It is worth noting that our assessment of herbarium samples using microscropy allowed us to track overall prevalence but not genetic variation or change, and we agree this could be an interesting future direction.
+We return to the topic of genotypic variation and specialization in the interaction in expanded discussion material.
+In this section,  we posit that the hotspots of changing endophyte prevalence we have identified may reflect rising prevalence following selection for genotypes that are particularly beneficial for enduring global change stressors, which we would expect given some evidence of genotype-specific benefits of the interaction in the literature.}
 \linesref{ms-R1C3-begin}{ms-R1C3-end}
 \linesref{ms-R2C14-begin}{ms-R2C14-end}
-}
 
@@
 \response{Thank you for these insightful comments. 
-While we visited only a handful of herbaria, we were strategic in sampling from collections with broad geographic coverage and were generally pleased with the spatial representation they provided (Fig.
+While we visited only a handful of herbaria, we were strategic in sampling from collections with broad geographic coverage and special strengths in \emph{Poaceae} and were generally pleased with the spatial representation they provided (Fig.
 1).
@@
 However, we conducted additional analyses, as recommended by the reviewer, to explore this concern. 
-Specifically, we have performed a new simulation analysis (\emph{Appendix A - Influence of spatially-biased sampling on model interpretation}) that shows \tom{the ability of an equivalent model to recover spatially-varying temporal trends within a region with up to $80$\% of data missing.}{I have not read the revision yet but I don't really know what this sentence means.}
+Specifically, we have performed a new simulation analysis (\emph{Appendix A - Influence of spatially-biased sampling on model interpretation}) in which we generated datasets with known spatially-varying trends, but with different levels of data-missingness in the northeast region, capturing some of the essential features of Reviewer 1's observation of potential spatial bias in our dataset.
+We then analyzed each of these datasets using a model with equivalent structure to our main text's endophyte prevalence model.
+Comparison of these analyses show that the model has the ability to recover spatially-varying temporal trends even in a scenario with up to $80$\% of datapoints missing from a region. 
 \\
 \\	
-The Bayesian posterior credible intervals of key results lend confidence to our interpretation, despite the uneven sampling that is inherent to collections-based studies such as ours \jacob{(Fig.
-\ref{ms-fig:svc_time_map_CI}), across the  host distributions.}{For some reason, the PDF is displaying the plot caption, which makes the sentence awkward.}
+The Bayesian posterior credible intervals of key results lend confidence to our interpretation, despite the uneven sampling that is inherent to collections-based studies such as ours (Fig. A7). 
+%\jacob{(Fig. \ref{ms-fig:svc_time_map_CI}), across the  host distributions.}{For some reason, the PDF is displaying the plot caption, which makes the sentence awkward.}
 For example, we find that for \emph{A.
-perennans}, the width of the credible interval for temporal change is consistently small (estimated trends in prevalence vary by only $\mp$ 2 percentage points per year) within highly-sampled areas (the midwest), while within areas with fewer samples (the northeast) this interval is only slightly wider (trends are estimated to be $\mp$ 3 percentage points per year).
+perennans}, the width of the credible interval for temporal change is consistently small (estimated trends in prevalence vary by only $\pm$ 2 percentage points per year) within highly-sampled areas (the midwest), while within areas with fewer samples (the northeast) this interval is only slightly wider (trends are estimated to be $\pm$ 3 percentage points per year).
 These observations suggest high confidence in the positive trend of increasing endophyte prevalence through time, even as the point estimate for that slope is more uncertain in sparsely sampled regions.
-We have added consideration of this and other potential biases to our discussion, along with \tom{the methods caveat described by this reviewer}{Be more specific here?
-I am not sure which caveat you are referring to.}.
+We have added consideration of this and other potential biases to our discussion, along with the methods caveat described by this reviewer, acknowledging that our collection efforts focused within four states as a potential bias in our results.
 \linesref{ms-R1C11-begin}{ms-R1C11-end} methods
@@
 Second, as suggested by the reviewer, we also assessed the relationship between initial prevalence and predicted changes in prevalence (new Figure A11).
-We found that for two species, the largest increases do occur at locations with low starting prevalence.
+We found that for the two \emph{Agrostis} species, the largest increases do occur at locations with low starting prevalence.
 To keep the main text focused on temporal change in prevalence, we placed this figure in the supplemental material but have added text describing this to our results.}
@@
 I recommend they dial this claim back.}
-\response{\jacob{We appreciate Reviewer 1 bringing this to our attention.
+\response{We appreciate Reviewer 1 bringing this to our attention.
 We would argue that our phrasing is not inaccurate because we have successfully linked changes in symbioses to climate change drivers, whereas previous research including Heberling \& Burke 2019 focus on validating methods for quantifying microbial community composition within historic specimens, but do not connect this to global change drivers.
-Heberling \& Burke 2019 state, "Our results demonstrate that herbarium specimens have the potential to be utilized to effectively reconstruct historical AMF communities.
+Heberling \& Burke 2019 state, ``Our results demonstrate that herbarium specimens have the potential to be utilized to effectively reconstruct historical AMF communities.
 However, given the scope of the current study, future work is needed to address specific global change hypotheses." However, we do recognize that there is a quickly growing body of research on microbial symbioses within herbarium specimens, especially those using sequencing and bioinformatics techniques.
-In response to this comment, we have rephrased this statement to clarify that we see novelty particularly in connecting microbial symbiosis to climate change drivers, and to temper our assertion of novelty.}{I like your response}}\tom{}{I agree -- very diplomatic!}
+In response to this comment, we have rephrased this statement to clarify that we see novelty particularly in connecting microbial symbiosis to climate change drivers, and to temper our assertion of novelty.}
 \linesref{ms-R1C10-begin}{ms-R1C10-end}
@@
 \linesref{ms-R1C11-begin}{ms-R1C11-end}
+\linesref{ms-R1C11-b-begin}{ms-R1C11-b-end}
 
@@
 This was a comment echoed by Reviewer 2 (Comment 14).
-In response to both reviewers we have added substantial material to our discussion that presents the key implications of our findings for population and range dynamics under climate change.}
+In response to both reviewers we have added substantial material to our discussion that presents the key implications of our findings for population and range dynamics under climate change.
+We do see our result as providing evidence for resilience of the symbiosis under ongoing climate change which we expect comes from benefits to host fitness, that could facilitate range shifts, but confirming these remain open questions, which we suggest in the new discussion.}
 \linesref{ms-R2C14-begin}{ms-R2C14-end}
@@
 \comment{In the abstract, I recommend briefly mentioning the statistical analyses or framework used (e.g., INLA) before presenting the results to enhance readability.}
-\response{\jacob{We have revised the abstract to include information about the statistical analysis.}{Please see my comment in the abstract}}
+\response{We have revised the abstract to include information about the statistical analysis, the use of an ``approximate Bayesian spatially-varying coefficients model".
+Because we felt that the acronym ``INLA" may be unfamiliar to many readers, we have chosen to not include it in the abstract, however we provide this information in the methods.}
 \linesref{ms-R2-C15-begin}{ms-R2-C15-end}
@@
 I would recommend to group the paragraphs presenting the research gaps of the ecological importance of endophytes, particularly in the context of climate change together and addressing the methodological novelty of this study in a separate, cohesive paragraph.}
-\response{\jacob{}{You forgot to answer this comment.
-I think we should push back.
-Here is my proposition: "We appreciate the suggestion to shorten the  Introduction and improve its clarity.
-While we understand the value of streamlining the content, we believe that the examples provided, including those related to coral and insect symbioses, serve to contextualize the study within broader ecological frameworks and may help engage readers who are less familiar with the topic."}\tom{}{I like Jacob's suggestion here and agree that we should push back on cutting the non-plant examples.
-But maybe the reorganization suggested here could work.
-I have not looked at the ms yet, maybe you have already done this.}}
+\response{We appreciate the suggestion to shorten the  Introduction and improve its clarity.
+While we understand the value of streamlining the content, we believe that the examples provided, including those related to coral and insect symbioses (Line \ref{ms-R2C16-begin}- \ref{ms-R2C16-end}), serve to contextualize the study within broader ecological frameworks and may help engage readers who are less familiar with the topic.
+This is particularly important given that symbioses with microbes are present across the tree of life, each faced with global change drivers similar to our focal grass-fungal symbiosis.}
 
@@
 \comment{L54: Which physical legacies or ecological processes? Provide example.}
-\response{In response to this comment, we have added additional information about the examples of ecological processes studied previously using herbarium specimens, \tom{such as...}{Again, be sure to answer questions here.
-It's ok (and expected) to have redundancy between the manuscript and response letter.}.}
+\response{In response to this comment, we have added additional information about the examples of ecological processes studied previously using herbarium specimens, such as the examination of reproductive structures to examine flowering phenology, or the presence of insect damage to study herbivory.}
 \linesref{ms-R2C19-begin}{ms-R2C19-end}
@@
 \response{We have edited this sentence to include a specific prediction for how this hypothesis would relate to our study system.
-We have also added specific predictions \tom{two paragraphs further down}{I would provide line numbers for this change.}, where we present our study's questions.}
+To summarize, theory for vertically-transmitted symbiosis suggests that the interaction should evolve to provide mutualistic benefits, which has been demonstrated for these \emph{Epichloë} symbionts under drought and herbivore stress.
+Under the stressors of climate change (i.e.
+increased drought stress), we expect that the symbionts will provide a fitness advantage and rise in prevalence within the population.
+We have also added specific predictions two paragraphs further down (Line \ref{ms-R2C21-b-begin}), where we present our study's central questions.}
 \linesref{ms-R2C21-begin}{ms-R2C21-end}
+\linesref{ms-R2C21-b-begin}{ms-R2C21-b-end}
 
@@
 Because we included coordinates georeferenced to local municipalities, counties, and in a few instances, states, the width of this bounding box varied from less than 1 km.
-to over 1233 km (median 841 km.).\tom{}{Good example of answering the question AND flagging the revised line number.}}
+to over 1233 km (median 841 km.).}
 \linesref{ms-R2C24-begin}{ms-R2C24-end}
@@
 In cases with unambiguous identification (presence or absence), the two scores are the same.
-When there is uncertainty in identification, we assign a positive ``liberal'' score and a negative ``conservative'' score.
-\tom{}{I would add here that ``split'' scores are infrequent, and give the frequency.} We present a comparison of results based on this scoring choice in our supplement, which showed minimal difference, and therefore we only present the results of the "liberal" score for the rest of the central results.}
+When there is uncertainty in identification, we assign a positive ``liberal'' score and a negative ``conservative'' score for the same sample.
+Samples with liberal and conservative scores that differed were infrequent; only $11$\% of the samples were given differing scores.
+We present a comparison of results based on this scoring choice in our supplement (Fig.
+A8) %\ref{ms-fig:conservative_scores_plot}),
+which did not show a qualitative difference, and therefore we only present the results of the "liberal" score for the rest of the central results.}
 \linesref{ms-R2C26-begin}{ms-R2C26-end}
@@
 \comment{-L229: Why 5? Is there a reference?}
-\response{We have added a reference supporting the use of penalized complexity priors, and revised terminology to more closely reflect the parameter values used to implement the model in inlabru.}
+\response{We have added a reference supporting the use of penalized complexity priors, and revised terminology to more closely reflect the parameter values used to implement the model in the \emph{inlabru} package.}
 \linesref{ms-R2C29-begin}{ms-R2C29-end}
@@
 In this case, we see the current placement of the section "Modeling distributions of host species" as appropriate because the preceding section describes models of endophyte prevalence that are independent from these models of host species' distributions.
-The predictions from the host distribution models are used principally for identifying realistic geographic ranges over which we test the relationship between climate drivers and trends in endophyte prevalence, and for visualizing relationships across study region.
-In response to this comment, we have added additional context describing this role that the host distribution models play.\tom{}{Maybe you say this in the text but I would point out that the herbarium records do not reflect the entire species' ranges, so if we want to make range-wide predictions we need some other way to do this.}}
+The predictions from the host distribution models are used principally for identifying realistic geographic ranges over which we test the relationship between climate drivers and trends in endophyte prevalence, and for visualizing relationships across the study region.
+The herbarium records used for endophyte surveys do not necessarily reflect the complete host species' range in the same way that GBIF records do, and so this is an important step in generating range-wide predictions of endophyte prevalence, but is not required for the mechanics of the endophyte prevalence model, which then bridges to our assessment of climate change relationships.
+In response to this comment, we have added additional context describing this role that the host distribution models play.}
 \linesref{ms-R2C30-begin}{ms-R2C30-end}
@@
 This permits us to generate predictions from the model at any location across the study region at high resolution, however doing so does require longer computation time.
-Models fit to the full set of pixels for each species show climate regressions in a similar direction as shown here and have narrower credible intervals.
-We could, for example, generate many thousands of points across each distribution, but \tom{this would make the credible intervals largely meaningless}{By this logic the CIs are meaningless for any sample size, whether its the full range or an arbitrary subset.}.
+Models fit to the full set of pixels that we chose in the manuscript for each species show climate regressions in a similar direction as shown here and have narrower credible intervals.
+We could, for example, generate many thousands of points across each distribution and generate highly ``significant" trend lines with very narrow credible intervals, but the sample size is relatively artificial in that context.
+Therefore, we have attempted to focus our presentation of results on the effect sizes of the endophyte-climate relationships we identify.
 We have added material describing why the choice of pixel dimensions is arbitrary as described in response to Reviewer 2 Comment 32.
@@
 Following this and other comments from Reviewer 2, we have edited to ensure that this organization of topics is repeated in the presentation of methods, in results, and in discussion, and added subheadings to better organize our manuscript.}
+\linesref{ms-R2C39-begin}{ms-R2C39-end} Discussion
 
@@
 \response{In response to this comment, we have added significant discussion on the potential impacts of \emph{Epichloë} on the host distribution as part of our revised discussion.
+In general, we interpret our results to support the idea that the symbiosis likely extends the environmental niche of the host species, and that the interaction will likely facility future/ongoing range shifts.
+We feel it is a very exciting future research direction to jointly model range shifts of symbiotic and non-symbiotic host plants to explore when and why the two may not overlap.
+	\\
+
 Our intention with the conservative/liberal scores is to capture uncertainty in the endophyte identification process, which we have clarified in our response to Reviewer 2 Comment 26.
@@
 We have in fact been working on a separate paper that analyzes the influence of land use changes and other anthropogenic disturbances on endophyte symbiosis.
-In response to this comment, we have added additional material to our discussion that proposes the potential potential for these other global change drivers to impact the dynamics of the symbiosis.}
+In response to this comment, we have added additional material to our discussion that proposes the potential for these other global change drivers to impact the dynamics of the symbiosis.}
 \linesref{ms-R2C44-begin}{ms-R2C44-end}
@@
 \response{This is an interesting question.
-\tom{}{I would add a brief response to the question here, even a quick summary of whatever speculation you added to the Discussion.}In response to this comment, we have expanded our discussion of the potential influence that host-symbiont genotype may have on our focal species' range dynamics in our re-organized discussion.}
+The literature does provide some evidence of genotype-specific benefits in \emph{Epichloë} symbiosis, and it is possible that the regions that experienced strong increases in prevalence come from the rise of a particular strain of symbiont within the population.
+We expect this may improve host plant populations responses to global change stressors through improved fitness, but it is feasible that this would limit responses to more diverse stressors if the diversity of symbiont strains within a population is reduced.
+In response to this comment, we have expanded our discussion of the potential influence that host-symbiont genotype may have on our focal species' range dynamics in our re-organized discussion.}
 \linesref{ms-R2C45-begin}{ms-R2C45-end}
@@
 \comment{A4 and A5: Which model, which data? (All plants, all years)?}
-\response{We have added description of the model and data for these supplemental figures (A4 and A5).\tom{}{And respond to the question here.}}
+\response{These figures refer to output from the endophyte prevalence model, tested against the full ``training" herbarium survey data.
+We have added description of the model and data for these supplemental figures (A4 and A5).}
 
 \comment{A6; From which models this is from? How probabilities are calculated? What is A and T?}
-\response{We have added description of the model that generates these parameter estimates in Fig.
+\response{This figure refers to output from the endophyte prevalence model fit in INLA.
+These probability density curves are a standard output of the INLA model output to visualize parameter posterior distributions.
+In response to this comment, we have added description of the model that generates these parameter estimates in Fig.
 A6.
@@
 \comment{A7; From which models; what is the legend for the value means, what is $\tau$?}
-\response{We have added description of the model that generates these parameter estimates in Fig.
+\response{This result comes from the endophyte prevalence model.
+Here, the legend value shows the width of the posterior credible interval for the parameter $\tau$ which is the estimate of spatially-varying temporal trends in endophyte prevalence.
+In response to this comment, we have added description of the model that generates these parameter estimates in Fig.
 A7.
@@
 A8, along with adding information that describes the model from which these parameter estimates are derived.
+In this figure, panel A presents the posterior distributions of the global year trend for models with either liberal or conservative scores.
+Panel B shows the spatially-varying temporal trend for the model fit with conservative scores, comparable to Figure 3 in the main text that shows the results for models fit to the liberal score data.
 We have re-plotted the figure with species names in place of the four letter codes noted by Reviewer 2.}
@@
 \comment{A11; Legend need specifications, the data need to be specified as well.}
-\response{For this figure (now Figure A12), we have re-plotted the figure with species names in place of the four letter codes noted by Reviewer 2.
-We have additionally edited the figure caption to more clearly describe the different sources of data.
-)}
+\response{Thank you for pointing this out.
+For this figure (now Figure A12), we have re-plotted the figure with species names in place of the four letter codes noted by Reviewer 2.
+We have additionally edited the figure caption to more clearly describe the different sources of data (Colored points are the contemporary survey data, black crosses are herbarium survey data).}
 

### 2025-02-13 Manuscript/endo_herbarium_R1.tex
completed response to reviewers

@@
 \usepackage[usenames,dvipsnames]{xcolor}
-\usepackage{hyperref}
+\usepackage[hidelinks]{hyperref}
 
@@
 
-	\textit{Keywords}: \jacob{climate change, plant-microbe symbiosis, herbarium, museum specimen, INLA, spatially-varying coefficients model, \emph{Epichloë}, \emph{Poaceae}.}{alphabetic order ? }
+	\textit{Keywords}: climate change, \emph{Epichloë}, herbarium, INLA, museum specimen, plant-microbe symbiosis, \emph{Poaceae}, spatially-varying coefficients model
 
@@
 Specifically, we \revise{examined} 2,346 herbarium specimens of three grass host species (\emph{Agrostis hyemalis}, \emph{Agrostis perennans}, \emph{Elymus virginicus}) collected over the past two centuries (1824 -- 2019) for the presence or absence of \emph{Epichloë} symbiosis.
-\revise{Analysis of an \jacob{approximate Bayesian spatially-varying coefficients model}{This is the first time INLA is mentioned.
-It seems to be a suggestion from the reviewer.
-However, I think most readers may not be familiar with it.
-I appreciate the precision regarding the type of Bayesian statistics (approximate), but I wonder if that might be too technical for the abstract.
-I would understand if it were included in the methods section.}\tom{}{I agree.
-I cut ' implemented in INLA'} revealed} that endophytes increased in prevalence over the last two centuries from ca.
+\revise{Analysis of an approximate Bayesian spatially-varying coefficients model revealed} that endophytes increased in prevalence over the last two centuries from ca.
 25\% to ca.
@@
 Changes in the variability of precipitation and temperature during off-peak seasons were also important predictors of increasing endophyte prevalence. 
-Our analysis performed favorably in an out-of-sample predictive test with \tom{contemporary survey data}{Since you drop numbers for the herbarium samples I would drop some numbers here too.}, a rare extra step in collections-based research.
+Our analysis performed favorably in an out-of-sample predictive test with contemporary survey data \revise{from across 63 populations}, a rare extra step in collections-based research.
 We identified greater local-scale variability in endophyte prevalence in contemporary data compared to model predictions based on historic data, suggesting new directions that could improve predictive accuracy.
@@
 These specimens often preserve physical legacies of ecological processes and species' interactions from dynamically changing environments across time and space.
-For example, previous researchers have used plant collections (herbaria) to document shifts in phenology \linelabel{R2C19-begin}\tom{\revise{by examining reproductive structures}}{I don't really understand this revision or what prompted it.
-Phenology studies do not examine reproductive structures, they examine timing.
-I also don't think the sentence works grammatical with each list element having a 'by' clause.} \citep{willis2017old, park2019herbarium,  berg2019examination}, pollination \revise{through examination of pollen removal} \citep{pauw2011reconstruction, duan2019century}, and herbivory \revise{by documenting leaf damage} \citep{meineke2019herbarium} related to anthropogenic climate change.
+\linelabel{R2C19-begin}For example, previous researchers have \revise{examined the flowers, pollen grains, and leaves of specimens within plant collections (herbaria) collected across time} to document shifts in \revise{reproductive} phenology \citep{willis2017old, park2019herbarium,  berg2019examination}, pollination \citep{pauw2011reconstruction, duan2019century}, and herbivory \citep{meineke2019herbarium} related to anthropogenic climate change.
 \linelabel{R2C19-end}
@@
 On the other hand, changes in the mean and variance of environmental conditions may disrupt microbial  mutualisms by changing the costs and benefits of the interaction for each partner in ways that can cause the interaction to deteriorate \citep{aslan2013mutualism, fowler2024microbial}.
-Coral bleaching (the loss of symbiotic algae) due to temperature stress \citep{sully2019global} is perhaps the best known example, but this phenomenon is not unique to corals.
-Lichens exposed to elevated temperatures experienced loss of photosynthetic function along with changes in the composition of their algal symbiont community \citep{meyer2022climate}.
+\linelabel{R2C16-begin}Coral bleaching (the loss of symbiotic algae) due to temperature stress \citep{sully2019global} is perhaps the best known example, but this phenomenon is not unique to corals.
+Lichens exposed to elevated temperatures experienced loss of photosynthetic function along with changes in the composition of their algal symbiont community \citep{meyer2022climate}.\linelabel{R2C16-end}
 How commonly and under what conditions microbial mutualisms deteriorate or strengthen under climate change remain unanswered questions \citep{frederickson2017mutualisms}.
@@
 \emph{Epichloë} fungal endophytes are specialized symbionts of cool-season grasses, which have been documented in $\sim 30$\% of cool-season grass species \citep{leuchtmann1992systematics}.
-They are transmitted \tom{predominantly vertically}{Maybe two adverbs is awkward but I think we need to acknowledge the possibility of H transmission.} from maternal plants to offspring through seeds.
+They are \revise{predominantly} transmitted vertically from maternal plants to offspring through seeds.
 Vertical transmission creates a feedback between the fitness of host and symbiont \citep{fine1975vectors, douglas1998host, rudgers2009fungus}. 
@@
 We then addressed how climate change may be driving trends in endophyte prevalence by asking: (iii) What is the relationship between temporal trends in endophyte prevalence and associated changes in climate drivers?
-\linelabel{R2C21-b-begin}We predicted that overall endophyte prevalence would increase over time in tandem with climate warming, and that localized hotspots of endophyte change would correspond spatially to hotspots of climate change.
+\linelabel{R2C21-b-begin}We predicted that \revise{overall} endophyte prevalence would increase over time in tandem with climate change, and that localized hotspots of endophyte change would correspond spatially to hotspots of climate \revise{warming and drying}.
 \linelabel{R2C21-b-end}
@@
 Cool-season grasses grow during the cooler temperatures of spring and autumn due to their reliance on $C_3$ photosynthesis.}\linelabel{R2C23-end}
-\emph{A.
-hyemalis} is a small short-lived perennial species that germinates in \tom{winter}{Winter, not spring.
-It's 'winter bentgrass'.} and typically flowers between March and July (most common collection month: May).
+\emph{A. hyemalis} is a small short-lived perennial species that germinates in \revise{late winter} winter and typically flowers between March and July (most common collection month: May).
 \emph{A. perennans} is of similar stature but is longer lived than \emph{Agrostis hyemalis} and flowers in late summer and early autumn (most common collection month: September). 
@@
 Some host species have been shown to partner with multiple symbiont species or strains, and in some cases multiple symbiont lineages can co-exist within a host population \cite{mc2014species}.
-However, surveys have typically found limited \emph{Epichloë} genotypic diversity within host populations\cite{treindl2023genetic}. 
+However, surveys have typically found limited \emph{Epichloë} genotypic diversity within host populations \cite{treindl2023genetic}. 
 Across host populations, concentrations of biologically-active alkaloids and the genes associated with their production vary substantially \cite{schardl2012chemotypic}.
@@
 		\subsection*{Herbarium surveys}
-We visited nine herbaria between 2019 and 2022 (see \tom{Table}{In table, maybe replace -- with 0.} \ref{table:herbaria} for a summary of specimens included from each collection). 
+We visited nine herbaria between 2019 and 2022 (see Table \ref{table:herbaria} for a summary of specimens included from each collection). 
 With permission from herbarium staff, we acquired seed samples from $1135$ \emph{A.
@@
 Texas A\&M, Missouri Botanic Garden) and other herbaria in the Southern Great Plains region of the United States.
-While these nine herbaria garnered specimens that span the focal species' ranges, our dataset unevenly samples across the study \tom{region}{Reference Figure 1 here?}. 
+While these nine herbaria garnered specimens that span the focal species' ranges, our dataset unevenly samples across the study region (Fig. \ref{fig:map}). 
 Texas, Oklahoma, Louisiana, and Missouri are the most represented states.
@@
 \revise{When we confidently identified endophytes within a specimen's seeds, we assigned matching liberal and conservative scores.}
-When we identified potential endophytes with unusual morphology, low uptake of stain, or a small amount of fungal hyphae across the scored seeds, we recorded \revise{ a positive identification} for the liberal score
-and a negative \revise{identification for} the conservative score.
-\tom{\revise{The liberal status assumed a potential endophyte identification was more likely to be endophyte-positive while the conservative status assumed that the potential endophyte identification was less likely to be endophyte-positive}.
-}{I don't think this sentence is necessary.}
+When we identified potential endophytes with unusual morphology, low uptake of stain, or a small amount of fungal hyphae across the scored seeds, we recorded \revise{ a positive identification} for the liberal score and a negative \revise{identification for} the conservative score.
 $89$\% of scored plants had matching liberal and conservative scores, reflecting high confidence in endophyte status.
@@
 For each host species $h$, endophyte presence/absence of the $i^{th}$ specimen ($P_{h,i}$) was modeled as a Bernoulli response variable with expected probability of endophyte occurrence $\hat{P}_{h,i}$.
-We modeled $\hat{P}_{h,i}$ as a linear function of \tom{intercept $\boldsymbol{A}_h$ and slope $\boldsymbol{T}_{h}$}{Are these matrices?
-I am concerned that Eq 1 does not make mathematical sense if it's saying scalar = matrix + matrix*scalar + scalar...
-etc.} defining the global trend in endophyte prevalence specific to each host species as well as with spatially-varying intercepts $\alpha_{h,l_{i}} $ and slopes $\tau_{h,l_i}$ associated with location ($l_i$, the unique latitude-longitude combination of the $i$th observation).
+We modeled $\hat{P}_{h,i}$ as a linear function of intercept $A_{h}$ and slope $T_{h}$ defining the global trend in endophyte prevalence specific to each host species as well as with spatially-varying intercepts $\alpha_{h,l_{i}} $ and slopes $\tau_{h,l_i}$ associated with location ($l_i$, the unique latitude-longitude combination of the $i$th observation).
 \linelabel{R2C28-begin}The joint-model structure allowed us to ``borrow \revise{information}'' \linelabel{R2C28-end} across species in the estimation of shared variance terms for the spatially-dependent random effect $\delta_{l_i}$, intended to account for residual spatial variation, and $\chi_{c_i}$ and $\omega_{s_i}$, the i.i.d.-random effects indexed for each collector identity ($c_i$) and scorer identity ($s_i$) of the $i^{th}$ specimen.
@@
 	\label{eq:trends}
-		logit(\hat{P}_{h,i}) = \boldsymbol{A}_{h} + \boldsymbol{T}_{h}*year_i  + \alpha_{h,l_{i}} + \tau_{h,l_i}*year_i + \delta_{l_i} + \chi_{c_i} + \omega_{s_i} 
+		logit(\hat{P}_{h,i}) = A_{h} + T_{h}*year_i  + \alpha_{h,l_{i}} + \tau_{h,l_i}*year_i + \delta_{l_i} + \chi_{c_i} + \omega_{s_i} 
 \end{equation}
@@
 We performed model fitting using the inlabru R package \citep{bachl2019inlabru}.
-Global intercept and slope parameters $\boldsymbol{A}$, and $\boldsymbol{T}$, were given vague priors.
-Collector and scorer random effects, $\chi$ and $\omega$ \revise{respectively}, \linelabel{R2C29-begin}\revise{were centered at 0 with precision parameters were assigned penalized complexity (PC) priors with parameter values $U_{PC}$ = 1 and $a_{PC}$ = 0.01 \cite{simpson2017penalising}.}\linelabel{R2C29-end}
+Global intercept and slope parameters, $A$ and $T$, were given vague priors.
+Collector and scorer random effects, $\chi$ and $\omega$ \revise{respectively}, \linelabel{R2C29-begin}\revise{were centered at 0 with precision parameters assigned penalized complexity (PC) priors with parameter values $U_{PC}$ = 1 and $a_{PC}$ = 0.01 \cite{simpson2017penalising}.}\linelabel{R2C29-end}
 Each spatially-structured parameter depended on a covariance matrix according to the proximity of each pair of collection locations \citep{lindgren2011explicit,bakka2018spatial}. 
@@
 We followed the ODMAP (overview, data, model, assessment, prediction) protocol \citep{crossley2022opposing} (see Supporting Methods). 
-\tom{In short, we used presence-only observations of each host species from Global Biodiversity Information Facility (GBIF) between 1990 to 2020.}{Maybe say how many records there were for each species?}
+In short, we used presence-only observations of each host species from Global Biodiversity Information Facility (GBIF) between 1990 to 2020 \revise{(713 occurrence records for \emph{A.
+hyemalis} \cite{https://doi.org/10.15468/dl.ryqqcv}, 656 occurrence records for \emph{A.
+perennans} \cite{https://doi.org/10.15468/dl.wtne24}, and 2338 occurrence records for \emph{A.
+virginicus} \cite{https://doi.org/10.15468/dl.yt76uh}}).
 %To reduce the potential influence of sampling bias and spatial autocorrelation, \tom{we thinned the occurrences to the spatial scale of our selected climatic predictors}{I do not follow.}. 
@@
 We found AUC values of 0.862, 0.838, 0.821 respectively  for \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus} indicating good model fit to data.
-\tom{To convert the continuous predicted probabilities into binary presence - absence maps on which we projected INLA predictions, we used the training sensitivity (true positive rate) and specificity threshold (true negative rate) \citep{liu2005selecting}.}{I don't really understand this sentence.
-How do the training sensitivity and specificity threshold lead to presence/absence maps?}
+We used the training sensitivity (true positive rate) and specificity (true negative rate) \revise{to set a threshold for transforming the continuous predicted probabilities into binary presence - absence host distribution maps on which we projected INLA predictions of endophyte prevalence} \citep{liu2005selecting}.
 %These binary maps serve as boundaries in presented maps of change in endophyte prevalence, and outline the set of pixels used in our analysis of climate correlates with trends in endophyte prevalence
@@
 The identity of individual scorers,\linelabel{R2C35-b-begin} \revise{the researchers who identified endophyte status microscopically},\linelabel{R2C35-b-end} did contribute to observed patterns in endophyte prevalence.
-\tom{For example, $3$ of the $25$ scorers were significantly more likely than average to assign positive endophyte status, as indicated by 95\% credible intervals greater than zero (Fig.
+For example, $3$ of the $25$ scorers were significantly more likely than average to assign positive endophyte status, as indicated by 95\% credible intervals greater than zero, \revise{while $4$ of the $25$ had 95\% credible intervals below zero} (Fig.
 \ref{fig:scorer_fx}).
-}{And I count four that have significantly negative CIs.}
 
@@
 virginicus} trends did not differ according to historic prevalence (\ref{fig:initialprev_trend_plot} C).} \linelabel{R1C7-end}
-Posterior estimates of uncertainty in spatially varying slopes indicate that these hotspots of change may have experienced increases of up to $5$\% per year while declines in prevalence may be as great as \tom{$4$\% per year}{Should this be -4\%?} for \emph{A.
+Posterior estimates of uncertainty in spatially varying slopes indicate that these hotspots of change may have experienced increases of up to $5$\% per year while declines in prevalence may be as great as $-4$\% per year for \emph{A.
 hyemalis} and \emph{A.
@@
 perennans} were most strongly associated with increasing autumn precipitation and with increasing mean and variability in autumn temperature (greater than 97\% posterior probabilities of positive slopes).
-For this species, a $1$ $^{\circ}$C increase in autumn temperature was associated with a $1.07$ \% increase per year in endophyte prevalence (Fig.
-\ref{fig:climate_correlation_plot}A) and a $100$ mm increase in precipitation was associated with a $0.8$\% increase per year in endophyte prevalence (Fig.
+For this species, \revise{each} $1$ $^{\circ}$C increase in autumn temperature was associated with a $1.07$ \% greater increase per year in endophyte prevalence (Fig.
+\ref{fig:climate_correlation_plot}A) and a $100$ mm increase in precipitation was associated with a $0.8$\% \revise{greater} increase per year in endophyte prevalence (Fig.
 \ref{fig:climate_correlation_plot}B).
@@
 \ref{fig:climate_correlation_plot}B) (greater than $99$\% posterior probability of a positive slope).
-For this species, an increase of $100$ mm in spring precipitation led to an increase of $0.6$\% per year in endophyte prevalence.
+For this species, an increase of $100$ mm in spring precipitation \revise{was associated with } $0.6$\% per year \revise{stronger} increases in endophyte prevalence \revise{relative to regions with no change in precipitation}.
 The next greatest slopes were those associated with variability in spring precipitation (greater than $96$\% posterior probability of a negative slope), as well as in the mean and variability of autumn climate (greater than $98$\% probability of negative and positive slopes, respectively).
@@
 \linelabel{R2C38-begin}\revise{Our analysis indicated relatively high confidence that these climate drivers influence endophyte prevalence shifts in \emph{E.
-virginicus}}(greater than $99$\% posterior probability of either negative or positive slopes respectively), \tom{however they translate to less than $0.2$\% change in endophyte prevalence per year for a change of $100$ mm change in precipitation over the century.\linelabel{R2C38-end}}{I am not sure why you are highlighting the slope with respect to mm of precip, since the preceding sentences emphasize temperature and var(precip).}
+virginicus}}(greater than $99$\% posterior probability of either negative or positive slopes respectively), however they translate, \revise{for example}, to less than a $0.4$\% decrease in endophyte prevalence per year for \revise{each $1^oC$ of summer warming} over the century.\linelabel{R2C38-end}
 \linelabel{R2C37-begin}Repeating this analysis using all pixels across each species' distribution were qualitatively similar to these results.\linelabel{R2C37-end}
@@
 virginicus}: purple).
-Slopes with greater than 95\% probability of being either positive or negative are represented as solid lines while those that have less than 95\% probability are dashed.
+Slopes with greater than 95\% \revise{posterior} probability of being either positive or negative are represented as solid lines while those that have less than 95\% probability are dashed.
 Points are the \revise{values of pre-computed SVC trends  and climate drivers at} 250 randomly sampled pixels across each host's distribution used in model fitting \revise{for the \emph{post hoc} climate regression analysis}.}
@@
 \subsection*{Evaluation of model performance on an out-of-sample test}
-Tests of \tom{models'}{Which models are you referring to?
-Is this plural because each species had a different model?} predictive performance as quantified by AUC and by visual posterior predictive checks, indicated good predictive ability.
+Tests of the endophyte prevalence model's predictive performance as quantified by AUC and by visual posterior predictive checks, indicated good predictive ability. 
 Model performance was similar between historic herbarium specimens used as training data and the out-of-sample test data from contemporary surveys (AUC = 0.79 and 0.77 respectively; Fig.
@@
 hyemalis} towards western longitudes (Fig.
-\ref{fig:contemptestplot}A) and \tom{northern latitudes}{It's actually an increase toward N latitudes, not decline.} (Fig.
+\ref{fig:contemptestplot}A) and \revise{an increase towards} northern latitudes (Fig.
 \ref{fig:contemptestplot}B).
 \revise{It is noteable that} model predictions for endophyte prevalence exhibited relatively little local geographic variation, whereas the out-of-sample survey data were highly variable with populations spanning 0\% to 100\% endophyte-symbiotic plants (Fig.
-\ref{fig:contemptestplot}C), \tom{indicating that...}{I would finish this train of thought, even as you leave most of it for the discussion.}.
+\ref{fig:contemptestplot}C), \revise{indicating population-to-population variation not captured in the endophyte prevalence model.}
 
@@
 Contemporary prevalence is represented by grey points (point size reflects sample size) along with trend lines from generalized linear models (black line and shaded 95\% confidence interval).
-(C) Comparison of observed vs.
-predicted endophyte prevalence shows that contemporary test data had more variance between populations than contemporary predictions.}
+(C) Comparison of \revise{contemporary} observed \revise{population prevalence} vs.
+predicted endophyte prevalence shows that contemporary test data had more variance between populations than \revise{in model} predictions.}
 	\label{fig:contemptestplot}
@@
 We identified hotspots of change for \emph{A.
-perennans}, \tom{which was the species that experienced the strongest absolute changes in endophyte prevalence (Fig.
-\ref{fig:svc_time_map}).}{Not sure what this is based on.
-In Fig 2 both AGrostis species look like they have similar increases.}
-Predicted declines of $0.9$\% per year in the southern portion of its range and predicted increases of up to $2$\% per year in the north suggest a potential poleward range shift of endophyte-symbiotic plants; whether the overall host distribution is shifting in parallel is an exciting next question.
+perennans}, which was \revise{the species whose endophyte prevalence was most responsive to changes in climate drivers} (Fig.
+\ref{fig:climate_correlation_plot}).
+Predicted declines of $0.9$\% per year in the southern portion of its range and predicted increases of up to $2$\% per year in the north suggest a potential poleward range shift of endophyte-symbiotic plants (Fig.
+\ref{fig:svc_time_map}B); whether the overall host distribution is shifting in parallel is an exciting next question.
 
 Based on previous work demonstrating that endophytes can shield their hosts from drought stress (reviewed in \citet{decunta2021systematic}), we generally predicted that drought conditions would be a driver of increasing endophyte prevalence.
-In contrast to this expectation, increasing prevalence for \emph{A. perennans} was associated with increasing autumn temperature and precipitation (Fig. \ref{fig:climate_correlation_plot}). 
+In contrast to this expectation, increasing prevalence for \emph{A.
+perennans} was associated with \revise{both} increasing autumn temperature and precipitation (Fig.
+\ref{fig:climate_correlation_plot}).
 To our knowledge, the response of the symbiosis in \emph{A.
@@
 \linelabel{R2C44-begin}Additionally, \emph{Epichloë} endophytes have been connected to a suite of non-drought related fitness benefits including herbivory \revise{defense} \citep{brem2001epichloe}, salinity resistence \citep{wang2020effects}, and mediation of pathogens \cite{vikuk2019infection} and the soil microbiome \citep{roberts2015rhizosphere}.
-\revise{Broad changes in the distribution and abundance of natural enemies, along with stresses from anthropogenic changes in landcover and pollution \cite{sage2020global} likely influence the benefits of symbiosis \cite{rudgers2020climate}.
+\revise{Broad changes in the distribution and abundance of natural enemies \cite{cote2004ecological}, along with stresses from anthropogenic changes in landcover and pollution \cite{sage2020global} likely influence the benefits of symbiosis \cite{rudgers2020climate}.
 The historic trends that we observed result from the combination of these fitness benefits playing out across the heterogeneous map of a changing climate and other anthropogenic drivers.\linelabel{R2C44-end}
@@
 Based on our analysis, it is likely that the symbiosis will facilitate range shifts for hosts by improving population growth at range edges.
-Previous population surveys \citep{sneck2017variation,rudgers2009benefits,semmartin2015broad} attributed environment-dependent gradients in endophyte prevalence to symbiont-derived fitness benefit,s allowing hosts to persist in environments where they otherwise could not \citep{afkhami2014mutualist,kazenel2015mutualistic}.
-However, symbiont-facilitated range shifts require that endophytes be present in the populations to be able to support population growth. 
+Previous population surveys \citep{sneck2017variation,rudgers2009benefits,semmartin2015broad} attributed environment-dependent gradients in endophyte prevalence to symbiont-derived fitness benefit's allowing hosts to persist in environments where they otherwise could not \citep{afkhami2014mutualist,kazenel2015mutualistic}.
+\linelabel{R1C11-b-begin}However, symbiont-facilitated range shifts require that endophytes be present in the populations to be able to support population growth. 
 The arid western range edge of \emph{A.
@@
 \ref{fig:svc_space_year_map}), and while prevalence has increased most quickly in the regions with historically low endophyte prevalence (Fig.
-\ref{fig:initialprev_trend_plot}), the complete absence of endophytes at range edges in this species would make it impossible for prevalence to increase without dispersal of symbiotic seeds \cite{fowler2023geographic}.
+\ref{fig:initialprev_trend_plot}), the complete absence of endophytes at range edges in this species would make it impossible for prevalence to increase without dispersal of symbiotic seeds \cite{fowler2023geographic}.\linelabel{R1C11-b-end}
 These factors potentially contribute to the ability of the host species to track its environmental niche. 
-Another interesting question is the degree to which symbiotic and non-symbiotic hosts, which occupy overlapping but distinct niches, are likely to experience distribution shifts in tandem or at different rates in future work.\linelabel{R2C42-end}
+Another interesting question is the degree to which symbiotic and non-symbiotic hosts, which occupy overlapping but distinct niches, are likely to experience distribution shifts in tandem or at different rates in future.\linelabel{R2C42-end}
 More extreme climate stresses, which are expected more frequently in the future \cite{seneviratne2021weather}, have the potential to alter the costs and benefits of the interaction. 
-The past indicates a resilient symbiosis, but it will be crucial to understand whether more extreme future climate conditions could tip this interaction to deteriote.}\linelabel{R2C14-end}
+The past indicates a resilient symbiosis, but it will be crucial to understand whether more extreme future climate conditions could tip this interaction to deteriorate.}\linelabel{R2C14-end}
 
@@
 While the model recreated the large-scale spatial trends observed in contemporary population surveys, test data contained more population-to-population variability in prevalence than could be explained by the model.
-\linelabel{R2C38-begin}\revise{We interpret this to mean that the model captures coarse-scale spatial and temporal trends reasonably well, but is not equipped to capture local-scale nuances that generate population-to-population differences.} \linelabel{R2C38-end}
+\linelabel{R2C39-begin}\revise{We interpret this to mean that the model captures coarse-scale spatial and temporal trends reasonably well, but is not equipped to capture local-scale nuances that generate population-to-population differences.} \linelabel{R2C39-end}
 Validating our model predictions \revise{with this test}, a rare extra step in collections-based studies, allows us to identify ways in which the model's out-of-sample predictive ability could be improved.
@@
 \revise{Natural history collectors sample one or a few specimens from local populations, and these observations are aggregated by the model to derive broad-scale estimates.}\linelabel{R2C43-end}
-This suggests that increasing local replication should be a factor considered in future collection efforts of natural history specimens, balanced with the required time and effort \linelabel{R1C9-begin} \revise{and with limitations on storage space within collections}.
+This suggests that increasing local replication should be a factor considered in future collection efforts of natural history specimens, \revise{balancing} the required time and effort \linelabel{R1C9-begin} \revise{along with limitations on storage space within collections}.
 \linelabel{R1C9-end}
-Poor predictive ability at local scales in this grass-endophyte system is not surprising, as previous studies have found that local variation, \tom{even to the scale of hundreds of meters}{Also see new paper in Functional Ecology by Pedro Gundel's group that has very local scale differences (I am a coauthor but did not do much.).} can structure endophyte-host niches \cite{kazenel2015mutualistic}.
+Poor predictive ability at local scales in this grass-endophyte system is not surprising, as previous studies have found that local variation, even to the scale of hundreds of meters can structure endophyte-host niches \cite{kazenel2015mutualistic,gundel2024temporal}.
 An important step would be integrating data from local and regional scales  through modeling to constrain estimates of local and regional variation.
@@
 km, including specimens from Mexico and Canada.
-The specimens we examined are concentrated in the south-central United States, with fewer specimens in the rapidly warming northeastern United States.
+Despite this advantage, the specimens we examined are concentrated in the south-central United States, with fewer specimens in the rapidly warming northeastern United States reflecting the regional focus of herbaria.
 We provide a simulation analysis exploring the potential impact of spatially and temporally biased sampling (Appendix A - Supporting Methods). 
-We found that the spatially-varying coefficient model had a strong ability to re-capitulate temporal trends across space in simulated data, and that this result was robust to relatively high levels of spatial bias ($80$\% of data missing from region).
+We found that the spatially-varying coefficient model had a strong ability to re-capitulate temporal trends across space in simulated data, and that this result was robust to relatively high levels of spatial bias ($80$\% of data missing from one spatial region).
 Simulation analyses that extend this work to consider the myriad ways herbarium data may be biased (i.e.
@@
 	This research was supported by funding from National Science Foundation (grants 1754468 and 2208857) and by funding from the Texas Ecolab Program.
-\tom{Two anonymous reviewers greatly improved an earlier version of this manuscript.}{Always thank the reviewers!}
+\revise{Two anonymous reviewers greatly improved an earlier version of this manuscript.}
 
@@
 \newpage{}
-	\tom{}{I am not sure if it's because I messed with some of the font packages, but some of the citations are a litte messy (some lowercase journal names, some uppercase).
-Latin names in titles should always be italicized.
-They will ask you to clean this up sooner or later, so I would do it sooner.}
 	\bibliographystyle{plainnat}
@@
 	\includegraphics[width = .5\linewidth]{../Plots/overlay_plot.png}
-	\caption[Graphical posterior predictive check of the endophyte prevalence model fit]{\revise{\textbf{Graphical posterior predictive check of \revise{the endophyte prevalence}model fit.}} Consistency between \revise{observed} data and \revise{predicted} values indicate that the fitted model accurately describes the data.
+	\caption[Graphical posterior predictive check of the endophyte prevalence model fit]{\revise{\textbf{Graphical posterior predictive check of \revise{the endophyte prevalence} model fit.}} Consistency between \revise{observed} data and \revise{predicted} values indicate that the fitted model accurately describes the data.
 Graph shows density curves for the observed data (black) along with 100 \revise{predicted} datasets (red).}
@@
 	\includegraphics[width = \linewidth]{../Plots/posterior_plot.png}
-	\caption[Posterior estimates of parameters describing global temporal trends in endophyte prevalence trends.]{\revise{\textbf{Posterior estimates of parameters describing global intercept and temporal trends from the endophyte prevalence model.}} Density curves show the probability density along with mean (colored line) and $95$\% CI (black lines) for the (A) intercept and (B) slope terms, \textbf{A} and \textbf{T} respectively \revise{from Eqn.
+	\caption[Posterior estimates of parameters describing global temporal trends in endophyte prevalence trends.]{\revise{\textbf{Posterior estimates of parameters describing global intercept and temporal trends from the endophyte prevalence model.}} Density curves show the probability density along with mean (colored line) and $95$\% CI (black lines) for the (A) intercept and (B) slope terms, $A$ and $T$ respectively \revise{from Eqn.
 \ref{eq:trends}}.
@@
 In cases of uncertain identification, the liberal status assumed a potential endophyte identification was more likely to be endophyte-positive while the conservative status assumed that the potential endophyte identification was less likely to be endophyte-positive}.
-(A) Posterior estimates of global temporal trend \revise{(\textbf{T} from Eqn.
+(A) Posterior estimates of global temporal trend \revise{($T$ from Eqn.
 \ref{eq:trends}}) for the endophyte prevalence model fit to liberal scores (grey) and to conservative scores (blue).
@@
 			Louisiana State University &     72  & 38  &   62       \\
-			Mercer Botanic Garden &   3    & --     &     6\\
+			Mercer Botanic Garden &   3    & 0     &     6\\
 			Missouri Botanic Garden& 210 & 205 & 122\\
-		    Texas A\&M &  100 &-- & 72 \\
+		    Texas A\&M &  100 &0 & 72 \\
 		    University of Kansas & 134 & 34 &  197\\
@@
 \textbf{Code availability}: Available through this link: https://github.com/joshuacfowler/EndoHerbarium \\
-\textbf{Data availability}: Will be available upon acceptance \\
+\textbf{Data availability}: Data was accessed through open-source R packages \emph{rgbif}.
+\emph{A.
+hyemalis} \cite{https://doi.org/10.15468/dl.ryqqcv}, \emph{A.
+perennans} \cite{https://doi.org/10.15468/dl.wtne24}, \emph{E.
+virginicus} \cite{https://doi.org/10.15468/dl.yt76uh} \\
 {\color{blue}{Assessment}}\\
@@
 	\label{eq:sim_trends}
-	logit(\hat{P}_{h,i}) = \boldsymbol{A}_{h} + \boldsymbol{T}_{h}*year_i  + \alpha_{h,l_{i}} + \tau_{h,l_i}*year_i + \delta_{l_i}
+	logit(\hat{P}_{h,i}) = A_{h} + T_{h}*year_i  + \alpha_{h,l_{i}} + \tau_{h,l_i}*year_i + \delta_{l_i}
 \end{equation}
@@
 Where symbiont presence/absence of the $i^{th}$ specimen ($P_{h,i}$) was modeled as a Bernoulli response variable with expected probability of symbiont occurrence $\hat{P}_{h,i}$ for each host species $h$.
-We modeled $\hat{P}_{h,i}$ as a linear function of intercept $\boldsymbol{A}_h$ and slope $\boldsymbol{T}_{h}$ defining the global trend in endophyte prevalence specific to each host species as well as with spatially-varying intercepts $\alpha_{h,l_{i}} $ and slopes $\tau_{h,l_i}$ associated with location ($l_i$, the unique latitude-longitude combination of the $i$th observation).
+We modeled $\hat{P}_{h,i}$ as a linear function of intercept $A_h$ and slope $T_{h}$ defining the global trend in endophyte prevalence specific to each host species as well as with spatially-varying intercepts $\alpha_{h,l_{i}} $ and slopes $\tau_{h,l_i}$ associated with location ($l_i$, the unique latitude-longitude combination of the $i$th observation).
 Similar to the SVC model of our central analysis (Eqn.
@@
 
-\begin{figure}[H]
-	\centering
-	\includegraphics[width = .7\linewidth]{../Plots/sim_svc_plot.png}
+\begin{figure}[htbp]
+	\centering
+	\includegraphics[width = .65\linewidth]{../Plots/sim_svc_plot.png}
 	\caption[Mean predicted spatially-varying trend in symbiont prevalence across datasets with different levels of missingness]{\textbf{Mean predicted spatially-varying trend in symbiont prevalence across datasets with different levels of missingness.} Color indicates the estimated mean temporal trend within each pixel across the simulated data.
@@
 
-While this analysis is not an exhaustive examination of the influence of sampling bias on our results for several reasons, including not examining how different strengths in temporal trends, different spatial arrangments of missing-ness influence model estimates, or different sample sizes, it demonstrates that the spatially-varying modelling framework implemented in INLA we emply can suitably recover regional trends even with significant spatially-bias within data collection, and further the analysis is likely robust to temporally-structured bias (missing data within recent collection period).
+While this analysis is not an exhaustive examination of the influence of sampling bias on our results for several reasons (including not examining how different strengths in temporal trends, different spatial arrangments of missing-ness influence model estimates, or different sample sizes influence results), it demonstrates that the spatially-varying modelling framework implemented in INLA we employ can suitably recover regional trends even with significant spatially-bias within data collection, and further the analysis is likely robust to temporally-structured bias (missing data within recent collection period).
 Future work could more fully explore the scenarios that cause this ability to break down. 

### 2025-07-09 Manuscript/Reviewer_response_endo_herbarium_R2.tex
adding second round response to reviewers

% ======================================================= %
% Document: TEMPLATE FOR RESPONSES TO REVIEWERS
% Author: Andrea Ballatore
% Date: Jan 7, 2013
% Source: https://raw.githubusercontent.com/ucd-spatial/Datasets/master/tex_response_to_reviewers_template/responses_to_reviewers.tex
% Modified by Eduard Szöcs, 10.03.2015
% ======================================================= %
\documentclass[12pt]{article}

% packages

\usepackage{graphicx}
\usepackage{url}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{color}
\definecolor{mygray}{gray}{0.6}
\usepackage[utf8]{inputenc}
\usepackage[onehalfspacing]{setspace}
\usepackage[
	round,	%(defaultage in the main file and \input ) for round parentheses;
	colon,	% (default) to separate multiple citations with colons;
	authoryear,% (default) for author-year citations;
	sort,		% orders multiple citations into the sequence in which they
]{natbib}
\usepackage[%disable
	]{todonotes}

\usepackage{anysize}
\marginsize{2.5cm}{2.5cm}{1.5cm}{2.5cm}

\usepackage{xr}
\externaldocument[ms-]{endo_herbarium_R2}

% macros
% add a counter
\newcounter{cN}
\setcounter{cN}{0}

\newcommand{\comment}[1]{
	\vspace{2em}
	\refstepcounter{cN} % incrment counter
	\noindent \hangindent=0em \textbf{\textcolor{Maroon}{\uline{Comment \thecN}:~}}\emph{``#1''}
	}

\newcommand{\response}[1]{
	\\[0.25em]
	\hangindent=2.3em \textbf{\textcolor{NavyBlue}{\uline{Response}:~}}#1
	}

\newcommand{\revise}[1]{{\color{Mahogany}{#1}}}

\newcommand{\linesref}[2]{
		\\[0.25em]
	\hangindent=2.3em {\color{Mahogany} Line(s): \ref{#1} - \ref{#2}}
}
\newcommand{\jacob}[2]{{\color{blue}{#1}}\footnote{\textit{\color{blue}{#2}}}}
%\IfValueT{#2}{- \ref{#2}}
\newcommand{\tom}[2]{{\color{red}{#1}}\footnote{\textit{\color{red}{#2}}}}
%\IfValueT{#2}{- \ref{#2}}

\usepackage[normalem]{ulem}
\definecolor{darkred}{rgb}{1,.6,.6}
\DeclareRobustCommand\problemline{\bgroup\markoverwith{\textcolor{darkred}{\rule[-0.9ex]{4pt}{3pt}}}\ULon}
\DeclareRobustCommand{\problem}[1]{\problemline{#1}} % soul
\setcounter{secnumdepth}{-1}

\begin{document}
% ======================================================= %
\title{Manuscript GCB-25-0529 --- Response to reviewers}

\maketitle
% ======================================================= %
\noindent To the editorial board,

Thank you for the opportunity to submit a second revision of our manuscript to \emph{Global Change Biology}. 
We appreciate that the reviewers acknowledged the positive attributes of our study while providing guidance to clarify our methods and improve our work's connections to the literature. 
In response to their comments, we have revised throughout our manuscript to address key modeling assumptions and choices, as well as editing our manuscript to highlight connections to the quickly growing field of research that that uses herbarium specimens to study global change. We detail these changes in the point-by-point responses below.
All of our changes are denoted in the revised manuscript with \revise{Mahogany font}.

With this revision, we hope you agree that our work is fit for publication in \emph{Global Change Biology}.

\vspace{2em}
Sincerely,

Joshua Fowler

Jacob Moutouama

Tom Miller

\newpage

% ======================================================= %
\section{Response to Reviewer 3}
\vspace{-2em}

%\comment{}
%\response{}
%\linesref{ms-R1C3-begin}{ms-R1C3-end}

\comment{The manuscript presents an intriguing analysis of long-term changes in Epichloë endophyte prevalence in grass species using herbarium specimens. The study's findings, which indicate an increase in endophyte prevalence correlated with climate change, are potentially significant for understanding plant-microbe interactions in the context of environmental change. The use of herbarium specimens to assess these changes over an extended timescale is a novel and valuable approach. However, I have some concerns that require clarification.}
\response{We thank Reviewer 3 for their constructive criticisms. In our point-by-point responses, we detail how we have incorporated them into our paper.}

\comment{ Statistics: (1) The approximate Bayesian spatially-varying coefficients model is a sophisticated approach. However, the manuscript would benefit from a more detailed explanation of the model's assumptions, limitations, and sensitivity to prior specifications. }
	\response{We thank Reviewer 3 for acknowledging the strength of our statistical approach. 
		In response to this comment, we have edited our methods section to provide context on the general INLA approach and it's assumptions, which have been ennumerated by the teams who developed this method. In addition, we point out key areas where we expect this may influence our results.}
	\linesref{ms-R3C1-begin}{ms-R3C1-end}

\comment{(2) The justification for selecting specific prior distributions for the Matérn covariance function should be clarified. A sensitivity analysis exploring the impact of different prior choices on the results is presented in the Appendix, but a more thorough discussion of these results should be included in the main text. }
	\response{We have added additional clarification about the assumptions that the Matérn covariance function rely upon, and about prior choice within our methods section. In addition, we have added material to our Discussion that highlights the important inferences provided by our sensitivity analysis}
	\linesref{ms-R3C2a-begin}{ms-R3C2a-end}
	\linesref{ms-R3C2b-begin}{ms-R3C2b-end}

\comment{  Climate Data: (1) The analysis correlates changes in endophyte prevalence with changes in climate drivers. However, the choice of climate variables (mean and standard deviation of seasonal temperature and precipitation) needs to be better justified. For example, are these variables the most ecologically relevant for the host species and endophytes? (2) The spatial resolution of climate data may not align with the scale of ecological processes influencing endophyte prevalence. Consider discussing the potential implications of this mismatch.}
\response{We have edited our methods section to more clearly explain that our goal in this analysis was to explore temperature and precipitation as the key climate factors of global change, that we expected there to be differences across seasons in the magnitude of climate change in these drivers, and that these seasons connect to the biology of our focal species - that they differ in their growing season. In addition, we have added to our discussion to acknowledge the potential importance of local scale ecological processes that may influence our analysis. We agree that the potential mismatch in scales is a limitation to herbarium studies in general, compounded not just by the scale of the climate data, but also by the spatial scale of georeferencing used to connect herbarium specimens to localities. The INLA modeling approach we employ cannot overcome inherent scale-limitations in the data, however treating space in a continuous way does allow us to make predictions at arbitrary spatial scales. Our climate analysis used predicted trends from the model to assess relationships with climate, meaning that the climate data were actually on a finer scale than the endophyte trend predictions. However this is, as Reviewer 3 notes, a broader scale than many potential ecological processes involved.}
\linesref{ms-R3C3-begin}{ms-R3C3-end}

\comment{ Out-of-Sample Validation: The out-of-sample predictive test with contemporary survey data is a strength of the study. However, the manuscript notes that the model exhibited relatively little local geographic variation, whereas the survey data were highly variable. This discrepancy suggests that local-scale factors not captured by the model may be important drivers of endophyte prevalence. A discussion of these potential factors (e.g., soil conditions, microclimate, local adaptation) and how they could be incorporated into future models would be valuable.}
\response{We agree with Reviewer 3 that there is a wide array of other potential factors that would make valuable future directions, and the existing discussion touched on this already. We have expanded the existing material on this subject.}
\linesref{ms-R3C5-begin}{ms-R3C5-end}

\comment{The contemporary survey period (2013-2020) is at the most recent edge of the time period encompassed by the historical specimens used for model fitting. It would be beneficial to acknowledge and discuss any potential limitations associated with using contemporary data for out-of-sample validation, particularly if climate conditions or other environmental factors have changed significantly during this period.}
\response{In response to this comment, we have revised our Discussion section to include further discussion of the potential implications of the out-of-sample validation test, particularly highlighting that the data do differ in scale and potentially climate conditions, and how this might inform future modeling efforts. Reviewer 3 points to potential non-stationarity as an explanation for mismatch in model validation. Previous investigations have found that stationarity is often a reasonable assumption to make and that there are not meaningul differences in out-of-sample validation tests that use blocks within and without time series (Bergmeir and Benitez, 2012, \emph{Information Sciences}).}
\linesref{ms-R3C6-begin}{ms-R3C6-end}

\comment{The manuscript interprets increases in Epichloë prevalence as adaptive changes that improve host fitness under increasing environmental stress. While this interpretation is reasonable, it is important to acknowledge that other factors (e.g., changes in land management practices, dispersal limitations) could also contribute to the observed trends.}
\response{Thank you for urging us to further explore our interpretation of changes in endophyte prevalence. We also recognize that we can not definitively identify these changes in prevalence as adaptive changes in response to climate. In fact, we have recently submitted an additional manuscript that explore the potential role that other anthropogenic global change drivers, including land use and nitrogen deposition, have on changes in endophyte prevalence. While we touched on these ideas already in the existing Discussion, in response to this comment, we have edited our Discussion to more clearly communicate the potential for land management and other factors to contribute to these trends.}
\linesref{ms-R3C7-begin}{ms-R3C7-end}

\comment{The manuscript mentions that Epichloë endophytes have been connected to a suite of non-drought related fitness benefits, including herbivory defense, salinity resistance, and mediation of pathogens and the soil microbiome. Consider discussing how these factors might interact with climate change to influence endophyte prevalence.
}
\response{This is a great point that different factors may themselves interact with changing climate, and is also connected to Reviewer 3's Comment 7. And so we have edited our discussion material to more clearly communicte that these drivers themselve may interact with changing climate and its potential consequences for endophyte prevalence.}
\linesref{ms-R3C8-begin}{ms-R3C8-end}

\section{Response to Reviewer 4}
\vspace{-2em}

\comment{Fowler et al. examined herbarium specimens of three grass species spanning 200 years to investigate the persistence of Epichloë symbioses. This innovative and well-executed study highlights the underutilized value of herbarium collections and is well-suited for publication in Global Change Biology.}
\response{We thank Reviewer 4 for their positive evaluation of our study, and we agree that the use of herbarium collections is a particuarly valuable part of this research and for studies of global change in particular.}

\comment{The authors have responded thoroughly to previous reviewer comments. My only additional suggestion is to expand the discussion of how herbaria are increasingly being used in contemporary research. While the manuscript includes some general remarks on their importance, it would benefit from a few more references and deeper integration of this theme—particularly in the introduction and conclusion. The authors could note that herbaria have traditionally supported taxonomic studies but are now being used in broader fields, including global change biology, biodiversity monitoring, and disease ecology.}
\response{In response to this comment, we have added references throughout our introduction that touch on these extended uses of herbarium specimens, and have also edited the Discussion and Conclusion sections to highlight the potential role that herbaria can play in global change research.}
\linesref{ms-R4C10-begin}{ms-R4C10-end}
\linesref{ms-R4C10b-begin}{ms-R4C10b-end}
\linesref{ms-R4C10c-begin}{ms-R4C10c-end}

\comment{Recent research by Jean Ristaino (on Phytophthora), Charles Davis (on plant responses to climate change), and Michael Bradshaw (on the spread of fungal pathogens) provides relevant examples (I noticed a couple references from these authors already). Given that plant diseases represent a form of microbial symbiosis, studies like this one contribute to an already growing body of work demonstrating the potential of herbarium specimens to reveal long-term patterns in host-pathogen dynamics and ecosystem change.}
\response{Thank you for pointing us towards this body of work. We have added relevant citations throughout the introduction and discussion and also expanded the discussion to acknowledge the potential role that herbarium specimens can play in understanding plant-pathogen dynamics.}
\linesref{ms-R4C11a-begin}{ms-R4C11a-end}
\linesref{ms-R4C11b-begin}{ms-R4C11b-end}

% ======================================================= %
\end{document}
% ======================================================= %

### 2025-07-09 Analyses/endo_herbarium_multispecies_spatiotemporal_analysis.R
updating some colors on plots

@@ -1174,7 +1174,8 @@ svc_space_map_AGHY.1895 <- ggplot()+
   geom_sf(data = states_map, color = "grey", linewidth = .1, fill = "#FAF9F6") +
   coord_sf(xlim = space_x, ylim = space_y)+
   gg(svc.space.pred_aghy.1895, aes(fill = mean))+
-  scale_fill_viridis_c(option = "turbo", na.value = "transparent", limits = prevrange)+
+  scale_fill_viridis_c(option = "viridis", na.value = "transparent", limits = prevrange)+
+  # scale_fill_viridis_c(option = "turbo", na.value = "transparent", limits = prevrange)+
   labs(title = species_names[1], subtitle = "year: 1895", fill = "% prevalence", y = "Latitude", x = "Longitude")+
   theme_light()+
   theme(plot.title = element_text(face = "italic"))
@@ -1185,7 +1186,8 @@ svc_space_map_AGHY.2020 <- ggplot()+
   geom_sf(data = states_map, color = "grey", linewidth = .1, fill = "#FAF9F6") +
   coord_sf(xlim = space_x, ylim = space_y)+
   gg(svc.space.pred_aghy.2020, aes(fill = mean))+
-  scale_fill_viridis_c(option = "turbo", na.value = "transparent", limits = prevrange)+
+  scale_fill_viridis_c(option = "viridis", na.value = "transparent", limits = prevrange)+
+  # scale_fill_viridis_c(option = "turbo", na.value = "transparent", limits = prevrange)+
   labs(title = species_names[1], subtitle = "year: 2020", fill = "% prevalence", y = "Latitude", x = "Longitude")+
   theme_light()+
   theme(plot.title = element_text(face = "italic"))
@@ -1196,7 +1198,8 @@ svc_space_map_AGPE.1895<- ggplot()+
   geom_sf(data = states_map, color = "grey", linewidth = .1, fill = "#FAF9F6") +
   gg(svc.space.pred_agpe.1895, aes(fill = mean))+
   coord_sf(xlim = space_x, ylim = space_y)+
-  scale_fill_viridis_c(option = "turbo", na.value = "transparent", limits = prevrange)+
+  scale_fill_viridis_c(option = "viridis", na.value = "transparent", limits = prevrange)+
+  # scale_fill_viridis_c(option = "turbo", na.value = "transparent", limits = prevrange)+
[…]

### 2025-07-09 Analyses/posthoc_climate_correlation.R
some updates to plots

@@ -813,7 +813,7 @@ tmean_trend <- ggplot(filter(prediction_df, grepl("tmean", name) & !grepl("sd", 
         )
   # lims(y = c(0,1))
 
-# tmean_trend
+tmean_trend
 
 
 tmean_sd_trend <- ggplot(filter(prediction_df, grepl("tmean", name) & grepl("sd", name))) +
@@ -975,9 +975,9 @@ aghy_tmean_sd_trend <- ggplot(filter(prediction_df, grepl(species_names[1], spec
 # aghy_tmean_sd_trend
 
 
-aghy_ppt_trend <- ggplot(filter(prediction_df, grepl(species_names[1], species)  & grepl("ppt", name) & !grepl("sd", moment))) +
+aghy_ppt_trend <- ggplot(filter(prediction_df, grepl(species_names[1], species)  & grepl("ppt", name) & !grepl("summer", name) & !grepl("sd", moment))) +
   geom_vline(aes(xintercept = 0), color = "grey80")+geom_hline(aes(yintercept = 0), color = "grey80")+
-  geom_point(data = filter(svc.pred_climate_subsample_long, grepl(species_names[1], species)  & grepl("ppt", name) & !grepl("sd", name) & !grepl("annual", name)), aes(x = value, y = mean, color = species), alpha = 0.2)+
+  geom_point(data = filter(svc.pred_climate_subsample_long, grepl(species_names[1], species)  & grepl("ppt", name) & !grepl("summer", name) & !grepl("sd", name) & !grepl("annual", name)), aes(x = value, y = mean, color = species), alpha = 0.2)+
   geom_line(aes(value, mean, linetype = overlap)) +
   geom_ribbon(aes(value, ymin = q0.025, ymax = q0.975, fill = species), alpha = 0.2) +
   geom_ribbon(aes(value, ymin = q0.25, ymax = q0.75, fill = species), alpha = 0.2) +
@@ -997,7 +997,7 @@ aghy_ppt_trend <- ggplot(filter(prediction_df, grepl(species_names[1], species) 
         legend.text = element_text(face = "italic"),
   )
 # lims(y = c(0,1))
-
+ggsave(aghy_ppt_trend, filename = "aghy_ppt_trend.png", width = 9, height = 4)
 # aghy_ppt_trend
 
 aghy_ppt_sd_trend <- ggplot(filter(prediction_df, grepl(species_names[1], species)  & grepl("ppt", name) & grepl("sd", moment))) +
[…]

### 2025-07-09 Manuscript/endo_herbarium_R2.tex

--- 2025-02-13 Manuscript/endo_herbarium_R1.tex
@@
 
+\newcommand{\firstrevise}[1]{{\color{black}{#1}}}
 \newcommand{\revise}[1]{{\color{Mahogany}{#1}}}
@@
 \linelabel{R2-C15-begin}Species' distributions and abundances are shifting in response to ongoing global climate change. 
-Mutualistic microbial symbionts can provide hosts with protection from environmental stress that may \revise{promote resilience under} environmental change, however \revise{this change} may also disrupt species interactions and lead to declines in hosts and/or symbionts.
+Mutualistic microbial symbionts can provide hosts with protection from environmental stress that may \firstrevise{promote resilience under} environmental change, however \firstrevise{this change} may also disrupt species interactions and lead to declines in hosts and/or symbionts.
 Symbionts preserved within natural history specimens offer a unique opportunity to quantify changes in microbial symbiosis across broad temporal and spatial scales. 
-We asked how the prevalence of seed-transmitted fungal symbionts of grasses (\emph{Epichloë} endophytes) has changed over time in response to climate change, and how these changes vary across host species' \revise{distributions}.
-Specifically, we \revise{examined} 2,346 herbarium specimens of three grass host species (\emph{Agrostis hyemalis}, \emph{Agrostis perennans}, \emph{Elymus virginicus}) collected over the past two centuries (1824 -- 2019) for the presence or absence of \emph{Epichloë} symbiosis.
-\revise{Analysis of an approximate Bayesian spatially-varying coefficients model revealed} that endophytes increased in prevalence over the last two centuries from ca.
+We asked how the prevalence of seed-transmitted fungal symbionts of grasses (\emph{Epichloë} endophytes) has changed over time in response to climate change, and how these changes vary across host species' \firstrevise{distributions}.
+Specifically, we \firstrevise{examined} 2,346 herbarium specimens of three grass host species (\emph{Agrostis hyemalis}, \emph{Agrostis perennans}, \emph{Elymus virginicus}) collected over the past two centuries (1824 -- 2019) for the presence or absence of \emph{Epichloë} symbiosis.
+\firstrevise{Analysis of an approximate Bayesian spatially-varying coefficients model revealed} that endophytes increased in prevalence over the last two centuries from ca.
 25\% to ca.
@@
 Changes in the variability of precipitation and temperature during off-peak seasons were also important predictors of increasing endophyte prevalence. 
-Our analysis performed favorably in an out-of-sample predictive test with contemporary survey data \revise{from across 63 populations}, a rare extra step in collections-based research.
+Our analysis performed favorably in an out-of-sample predictive test with contemporary survey data \firstrevise{from across 63 populations}, a rare extra step in collections-based research.
 We identified greater local-scale variability in endophyte prevalence in contemporary data compared to model predictions based on historic data, suggesting new directions that could improve predictive accuracy.
-Our results provide novel evidence for a cryptic biological response to climate change that may contribute to the resilience of host-microbe symbiosis through \revise{fitness benefits to symbiotic hosts}.
+Our results provide novel evidence for a cryptic biological response to climate change that may contribute to the resilience of host-microbe symbiosis through \firstrevise{fitness benefits to symbiotic hosts}.
 \linelabel{R2-C15-end}
-
-Abstract : 300 words
 
@@
 Empirically evaluating these biotic changes -- whether interacting species shift in tandem with their partners or not \cite{hillerislambers2013will} -- is crucial to predicting the reorganization of Earth's biodiversity under global change.
-Such evaluations have been limited because few datasets on species interactions extend over sufficiently long time scales of contemporary climate change \cite{poisot2021global}.
-
-Natural history specimens, which were originally collected to study and preserve taxonomic diversity, present a unique opportunity to explore long-term changes in ecological interactions across broad spatial and temporal scales \citep{meineke2018unrealized}.
+Such evaluations have been limited because few datasets on species interactions extend over sufficiently long time scales of contemporary climate change \citep{poisot2021global}.
+
+\linelabel{R4C10-begin}Natural history specimens, which were originally collected to study and preserve taxonomic diversity, present a unique opportunity to explore long-term changes in \revise{biodiversity} and ecological interactions across broad spatial and temporal scales\revise{ \citep{meineke2018unrealized, davis2023herbarium}.}
 Natural history collections, built and maintained by the efforts of thousands of scientists, are invaluable time machines, primarily comprised of physical specimens of organisms along with information about the time and place of their collection.
-These specimens often preserve physical legacies of ecological processes and species' interactions from dynamically changing environments across time and space.
-\linelabel{R2C19-begin}For example, previous researchers have \revise{examined the flowers, pollen grains, and leaves of specimens within plant collections (herbaria) collected across time} to document shifts in \revise{reproductive} phenology \citep{willis2017old, park2019herbarium,  berg2019examination}, pollination \citep{pauw2011reconstruction, duan2019century}, and herbivory \citep{meineke2019herbarium} related to anthropogenic climate change.
-\linelabel{R2C19-end}
+These specimens often preserve physical legacies of ecological processes and species' interactions from dynamically changing environments across time and space \revise{\citep{lendemer2020extended}.}\linelabel{R4C10-end}
+\linelabel{R2C19-begin}
+For example, previous researchers have \firstrevise{examined the flowers, pollen grains, and leaves of specimens within plant collections (herbaria) collected across time} to document shifts in \firstrevise{reproductive} phenology \citep{willis2017old, park2019herbarium,  berg2019examination}, pollination \citep{pauw2011reconstruction, duan2019century}, and herbivory \citep{meineke2019herbarium} related to anthropogenic climate change.\linelabel{R2C19-end}
 However, few previous studies have leveraged biological collections to examine climate change-related shifts in a particularly common type of interaction: microbial symbiosis.
@@
 \emph{Epichloë} fungal endophytes are specialized symbionts of cool-season grasses, which have been documented in $\sim 30$\% of cool-season grass species \citep{leuchtmann1992systematics}.
-They are \revise{predominantly} transmitted vertically from maternal plants to offspring through seeds.
+They are \firstrevise{predominantly} transmitted vertically from maternal plants to offspring through seeds.
 Vertical transmission creates a feedback between the fitness of host and symbiont \citep{fine1975vectors, douglas1998host, rudgers2009fungus}. 
@@
 Previous survey studies of contemporary populations have documented large-scale spatial patterns in endophyte prevalence structured by environmental gradients \citep{granath2007variation,bazely2007broad, afkhami2012fungal,sneck2017variation}.
-We predicted that prevalence should track temporal changes in environmental drivers \linelabel{R2C21-begin}\revise{(i.e. drought)} \linelabel{R2C21-end} that elicit strong fitness benefits.
+We predicted that prevalence should track temporal changes in environmental drivers \linelabel{R2C21-begin}\firstrevise{(i.e. drought)} \linelabel{R2C21-end} that elicit strong fitness benefits.
 %For example, endophyte-mediated drought tolerance should lead endophyte prevalence to increase in regions where precipitation has declined over time.
@@
 However, no subsequent studies, to our knowledge, have used the vast resources of biological collections to quantitatively assess spatio-temporal trends in endophyte prevalence and their environmental correlates.
+\linelabel{R4C11a-begin}\revise{Previous work has used herbarium specimens to identify the origins and population genomics of  plant diseases such as \emph{Phytophora}, the Irish potato famine pathogen \citep{ristaino2001pcr,ristaino2002tracking,yoshida2013rise}, and have been proposed as vehicles to track other emerging plant pathogens \citep{ristaino2020importance,bradshaw2021global}.}\linelabel{R4C11a-end}
 Grasses are commonly collected and identified based on the presence of their reproductive structures, meaning that preserved specimens typically contain seeds, conveniently preserving the fungi along with their host plants on herbarium sheets.
@@
 We then addressed how climate change may be driving trends in endophyte prevalence by asking: (iii) What is the relationship between temporal trends in endophyte prevalence and associated changes in climate drivers?
-\linelabel{R2C21-b-begin}We predicted that \revise{overall} endophyte prevalence would increase over time in tandem with climate change, and that localized hotspots of endophyte change would correspond spatially to hotspots of climate \revise{warming and drying}.
+\linelabel{R2C21-b-begin}We predicted that \firstrevise{overall} endophyte prevalence would increase over time in tandem with climate change, and that localized hotspots of endophyte change would correspond spatially to hotspots of climate \firstrevise{warming and drying}.
 \linelabel{R2C21-b-end}
-Finally, we evaluated \revise{(iv) how our model, built on data from historic specimens, performed in an out-of-sample test }using data on endophyte prevalence from contemporary surveys of host populations.
+Finally, we evaluated \firstrevise{(iv) how our model, built on data from historic specimens, performed in an out-of-sample test }using data on endophyte prevalence from contemporary surveys of host populations.
 To answer these questions we examined a total of $2,346$ historic specimens collected across eastern North America between 1824 and 2019, and evaluated model performance against contemporary surveys comprising 1,442 individuals from 63 populations surveyed between 2013 and 2020.
@@
 Our surveys focused on three native North American grasses: \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus} that host \emph{Epichloë} symbionts. 
-\linelabel{R2C23-begin}\revise{These cool-season grass species are commonly represented in natural history collections with broad distributions covering much the eastern United States (Fig.
+\linelabel{R2C23-begin}\firstrevise{These cool-season grass species are commonly represented in natural history collections with broad distributions covering much the eastern United States (Fig.
 \ref{fig:map}).
 Cool-season grasses grow during the cooler temperatures of spring and autumn due to their reliance on $C_3$ photosynthesis.}\linelabel{R2C23-end}
-\emph{A. hyemalis} is a small short-lived perennial species that germinates in \revise{late winter} winter and typically flowers between March and July (most common collection month: May).
+\emph{A. hyemalis} is a small short-lived perennial species that germinates in \firstrevise{late winter} winter and typically flowers between March and July (most common collection month: May).
 \emph{A. perennans} is of similar stature but is longer lived than \emph{Agrostis hyemalis} and flowers in late summer and early autumn (most common collection month: September). 
@@
 \linelabel{R1C3-begin}
-\revise{Both \emph{Agrostis} species host \emph{Epichloë amarillans} \cite{craven2001multigene, leuchtmann2014nomenclatural}, while \emph{Elymus virginicus} typically hosts \emph{Epichloë elymi} \cite{clay2002evolutionary}.
+\firstrevise{Both \emph{Agrostis} species host \emph{Epichloë amarillans} \cite{craven2001multigene, leuchtmann2014nomenclatural}, while \emph{Elymus virginicus} typically hosts \emph{Epichloë elymi} \cite{clay2002evolutionary}.
 The fungal symbionts primarily reproduce asexually and are passed from mother to offspring by vertical transmission through seeds. 
@@
 For fifteen of the oldest specimens, only information at the state level was available, and so we used the state centroid.
-\linelabel{R2C24-begin}\revise{The median pairwise distance between georeferenced coordinate points was 841 km.
+\linelabel{R2C24-begin}\firstrevise{The median pairwise distance between georeferenced coordinate points was 841 km.
 The median longitudinal width of the bounding boxes generated to geocode municipality, county, or state centroids was 44.7 km. 
@@
 \linelabel{R1C11-begin}
-\revise{Our visits focused on herbaria with historic strengths in \emph{Poaceae} collections (e.g.
+\firstrevise{Our visits focused on herbaria with historic strengths in \emph{Poaceae} collections (e.g.
 Texas A\&M, Missouri Botanic Garden) and other herbaria in the Southern Great Plains region of the United States.
@@
 virginicus}; we scored $10,342$ seeds in total.
-\linelabel{R2C25-begin}Due to imperfect vertical transmission, \revise{the production of symbiont-free offspring from symbiotic hosts} \cite{afkhami2008symbiosis},\linelabel{R2C25-end} it is possible that symbiotic host-plants produce a mixture of symbiotic and non-symbiotic seeds.
+\linelabel{R2C25-begin}Due to imperfect vertical transmission, \firstrevise{the production of symbiont-free offspring from symbiotic hosts} \cite{afkhami2008symbiosis},\linelabel{R2C25-end} it is possible that symbiotic host-plants produce a mixture of symbiotic and non-symbiotic seeds.
 We therefore designated a specimen as endophyte-symbiotic if \emph{Epichloë} hyphae were observed in one or more of its seeds, or non-symbiotic if \emph{Epichloë} hyphae were observed in none of its seeds.
 \linelabel{R2C26-begin}
-To capture uncertainty in the endophyte \revise{identification} process, we recorded both a "liberal" and a "conservative" endophyte score for each plant specimen.  
-\revise{When we confidently identified endophytes within a specimen's seeds, we assigned matching liberal and conservative scores.}
-When we identified potential endophytes with unusual morphology, low uptake of stain, or a small amount of fungal hyphae across the scored seeds, we recorded \revise{ a positive identification} for the liberal score and a negative \revise{identification for} the conservative score.
+To capture uncertainty in the endophyte \firstrevise{identification} process, we recorded both a "liberal" and a "conservative" endophyte score for each plant specimen.  
+\firstrevise{When we confidently identified endophytes within a specimen's seeds, we assigned matching liberal and conservative scores.}
+When we identified potential endophytes with unusual morphology, low uptake of stain, or a small amount of fungal hyphae across the scored seeds, we recorded \firstrevise{ a positive identification} for the liberal score and a negative \firstrevise{identification for} the conservative score.
 $89$\% of scored plants had matching liberal and conservative scores, reflecting high confidence in endophyte status.
-The following analyses used the liberal status, \revise{however} repeating all analyses with the conservative status yielded qualitatively similar results (Fig.
+The following analyses used the liberal status, \firstrevise{however} repeating all analyses with the conservative status yielded qualitatively similar results (Fig.
 \ref{fig:conservative_scores_plot}).\linelabel{R2C26-end}
@@
 virginicus}: purple).
-\revise{Map lines delineate study areas and do not necessarily depict accepted national boundaries.}}
+\firstrevise{Map lines delineate study areas and do not necessarily depict accepted national boundaries.}}
 	\label{fig:map}
@@
 \subsection*{Modeling spatial and temporal changes in endophyte prevalence}
-We assessed spatial and temporal changes in endophyte prevalence across each host distribution, quantifying the ``global'' temporal trends \linelabel{R2C27-begin}\revise{averaged}\linelabel{R2C27-end} across space, and then examining spatial heterogeneity in the direction and magnitude of endophyte change (hotspots and coldspots) across the spatial extent of each host's distribution.
+We assessed spatial and temporal changes in endophyte prevalence across each host distribution, quantifying the ``global'' temporal trends \linelabel{R2C27-begin}\firstrevise{averaged}\linelabel{R2C27-end} across space, and then examining spatial heterogeneity in the direction and magnitude of endophyte change (hotspots and coldspots) across the spatial extent of each host's distribution.
 To account for the spatial non-independence of geo-referenced occurrences, we used an approximate Bayesian method, Integrated Nested Laplace Approximation (INLA), to construct spatio-temporal models of endophyte prevalence.
@@
 We modeled $\hat{P}_{h,i}$ as a linear function of intercept $A_{h}$ and slope $T_{h}$ defining the global trend in endophyte prevalence specific to each host species as well as with spatially-varying intercepts $\alpha_{h,l_{i}} $ and slopes $\tau_{h,l_i}$ associated with location ($l_i$, the unique latitude-longitude combination of the $i$th observation).
-\linelabel{R2C28-begin}The joint-model structure allowed us to ``borrow \revise{information}'' \linelabel{R2C28-end} across species in the estimation of shared variance terms for the spatially-dependent random effect $\delta_{l_i}$, intended to account for residual spatial variation, and $\chi_{c_i}$ and $\omega_{s_i}$, the i.i.d.-random effects indexed for each collector identity ($c_i$) and scorer identity ($s_i$) of the $i^{th}$ specimen.
+\linelabel{R2C28-begin}The joint-model structure allowed us to ``borrow \firstrevise{information}'' \linelabel{R2C28-end} across species in the estimation of shared variance terms for the spatially-dependent random effect $\delta_{l_i}$, intended to account for residual spatial variation, and $\chi_{c_i}$ and $\omega_{s_i}$, the i.i.d.-random effects indexed for each collector identity ($c_i$) and scorer identity ($s_i$) of the $i^{th}$ specimen.
 
@@
 Prolific collectors who contribute thousands of specimens may be more or less likely to collect certain species, or specimens with certain traits \cite{daru2018widespread}. 
-Similarly, the process of scoring seeds for hyphae involved \linelabel{R2C35-begin}\revise{multiple researchers (or "scorers")} who, even with standardized training, may vary in their likelihood of positively identifying \emph{Epichloë}.
+Similarly, the process of scoring seeds for hyphae involved \linelabel{R2C35-begin}\firstrevise{multiple researchers (or "scorers")} who, even with standardized training, may vary in their likelihood of positively identifying \emph{Epichloë}.
 \linelabel{R2C35-end}
@@
 Global intercept and slope parameters, $A$ and $T$, were given vague priors.
-Collector and scorer random effects, $\chi$ and $\omega$ \revise{respectively}, \linelabel{R2C29-begin}\revise{were centered at 0 with precision parameters assigned penalized complexity (PC) priors with parameter values $U_{PC}$ = 1 and $a_{PC}$ = 0.01 \cite{simpson2017penalising}.}\linelabel{R2C29-end}
+Collector and scorer random effects, $\chi$ and $\omega$ \firstrevise{respectively}, \linelabel{R2C29-begin}\firstrevise{were centered at 0 with precision parameters assigned penalized complexity (PC) priors with parameter values $U_{PC}$ = 1 and $a_{PC}$ = 0.01 \citep{simpson2017penalising}.}\linelabel{R2C29-end}
 Each spatially-structured parameter depended on a covariance matrix according to the proximity of each pair of collection locations \citep{lindgren2011explicit,bakka2018spatial}. 
@@
 \ref{fig:meshplot}).
-We assessed model fit with visual posterior predictive checks (\ref{fig:overlayplot}) and measurements of AUC (Figs.
-\ref{fig:ROCtraining}-\ref{fig:ROCtest}) \linelabel{R2C48-begin} \revise{\cite{gelman2006data}}\linelabel{R2C48-end}.
+\linelabel{R3C2a-begin}\revise{Mat\'{e}rn covariance functions are widely used in spatial-statistical modeling because of their mathematical tractability and flexibility.
+This covariance structure relies on the assumption that the underlying process is stationary and isotropic, such that spatial autocorrelation between data points depends only on their relative positions \citep{bakka2018spatial}.}\linelabel{R3C2a-end}
+
+\linelabel{R3C1-begin}\revise{Implementing spatially-structured parameters in INLA with this SPDE approach is useful particularly because space is treated as a continuous variable, allowing the model to make efficient use of the data and generate predictions across the entire study region.
+The SPDE approach is flexible enough that it can capture smooth trends across space that are informed by the data rather than by spatial regions chosen \emph{a priori} by researchers.
+However this flexibility also invites the risk of overfitting, as with other non-linear modeling approaches \citep{ward2014complexity,lapeyrolerie2023limits,ramampiandra2023comparison}.
+}\linelabel{R3C1-end}
 Priors for the Mat\'{e}rn covariance function, termed ''range" and "variance", define how proximity effects decay with distance.
+\linelabel{R3C2b-begin}\revise{The choice of priors for these types of spatial models is an area of active research \cite{simpson2017penalising,bakka2018spatial}, but another advantage of the INLA approach is that its computational efficiency allows for sensitivity analyses.}\linelabel{R3C2b-end}
 Results presented in the main text reflect a prior range of $342$ kilometers (i.e. a $50$\% probability of estimating a range less than $342$ kilometers). 
@@
 the same direction of effects across space.
-\revise{Through results and discussion that follow, we refer to the model described in this section as the ``endophyte prevalence model''}.
+We assessed model fit with visual posterior predictive checks (\ref{fig:overlayplot}) and measurements of AUC (Figs.
+\ref{fig:ROCtraining}-\ref{fig:ROCtest}) \linelabel{R2C48-begin} \firstrevise{\cite{gelman2006data}}\linelabel{R2C48-end}.
+\firstrevise{Through results and discussion that follow, we refer to the model described in this section as the ``endophyte prevalence model''}.
 
@@
 \linelabel{R2C30-begin}
-\revise{The herbarium records did not encompass the entirety of each host species' range.
+\firstrevise{The herbarium records did not encompass the entirety of each host species' range.
 Therefore, we used additional data sources to model the geographic distribution of each host species, with two goals: (1) generate realistic maps on which we could project the predictions of the INLA model, and (2) use the geographic distributions to test for relationships between climate change drivers and trends in endophyte prevalence.
@@
 We followed the ODMAP (overview, data, model, assessment, prediction) protocol \citep{crossley2022opposing} (see Supporting Methods). 
-In short, we used presence-only observations of each host species from Global Biodiversity Information Facility (GBIF) between 1990 to 2020 \revise{(713 occurrence records for \emph{A.
+In short, we used presence-only observations of each host species from Global Biodiversity Information Facility (GBIF) between 1990 to 2020 \firstrevise{(713 occurrence records for \emph{A.
 hyemalis} \cite{https://doi.org/10.15468/dl.ryqqcv}, 656 occurrence records for \emph{A.
@@
 We found AUC values of 0.862, 0.838, 0.821 respectively  for \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus} indicating good model fit to data.
-We used the training sensitivity (true positive rate) and specificity (true negative rate) \revise{to set a threshold for transforming the continuous predicted probabilities into binary presence - absence host distribution maps on which we projected INLA predictions of endophyte prevalence} \citep{liu2005selecting}.
+We used the training sensitivity (true positive rate) and specificity (true negative rate) \firstrevise{to set a threshold for transforming the continuous predicted probabilities into binary presence - absence host distribution maps on which we projected INLA predictions of endophyte prevalence} \citep{liu2005selecting}.
 %These binary maps serve as boundaries in presented maps of change in endophyte prevalence, and outline the set of pixels used in our analysis of climate correlates with trends in endophyte prevalence
@@
 We assessed how the magnitude of climate change may have driven changes in endophyte prevalence by assessing correlations between changes in climate and changes in endophyte prevalence predicted from our spatial model at evenly spaced pixels across the study area.
+
 We first downloaded monthly temperature and precipitation rasters from the PRISM climate group \citep{daly2013prism} covering the time period between 1895 and 2020 using the 'prism' R package \citep{Rprism2015}.
 Prism provides reconstructions of historic climate variables across the United States by spatially interpolating weather station data \citep{diLuzio2008constructing}. 
-We calculated 30-year climate normals for seasonal mean temperature and cumulative precipitation for the recent (1990 to 2020) and historic (1895 to 1925) periods.
+\linelabel{R3C3-begin}\revise{Because the magnitude of observed climate change differs across seasons, and because different growing seasons is a key feature of the biology of our focal host species, } we calculated 30-year climate normals for seasonal mean temperature and cumulative precipitation for the recent (1990 to 2020) and historic (1895 to 1925) periods.
 We used three four-month seasons within the year (Spring: January, February, March, April; Summer: May; June, July, August; Autumn: September, October, November, December). 
-This division of seasons allowed us to quantify differences in climate associated with the two ``cool'' seasons, when we expected our focal species to be most active (\emph{A.
+This division of seasons allowed us to quantify differences in \revise{the primary climate change drivers, temperature and preciptiation} associated with the two ``cool'' seasons, when we expected our focal species to be most active (\emph{A.
 hyemalis} flowering phenology: spring; \emph{E.
@@
 perennans}: autumn).
+\linelabel{R3C3-end}
 In addition to mean climate conditions, environmental variability itself can influence population dynamics \cite{tuljapurkar_population_1982} and changes in variability are a key prediction of climate change models \cite{stocker2013technical, ipcc_2021}.
@@
 
-\linelabel{R2C32-begin}\revise{We then evaluated whether areas that have experienced the greatest changes in endophyte prevalence (hotspots of endophyte change) are associated with high degrees of change in climate (hotspots of climate change)
+\linelabel{R2C32-begin}\firstrevise{We then evaluated whether areas that have experienced the greatest changes in endophyte prevalence (hotspots of endophyte change) are associated with high degrees of change in climate (hotspots of climate change)
 To do so, we modeled the fitted, spatially-varying slopes of endophyte change through time ($\tau_{[h]l}$) as a linear function of environmental covariates, with a Gaussian error distribution for a set of pixels across each host distribution.
@@
 Data from each host species were analyzed separately. 
-\revise{Throughout the results and discussion that follow, we refer to this analysis as the ``\emph{post hoc} climate regression analysis''.}
+\firstrevise{Throughout the results and discussion that follow, we refer to this analysis as the ``\emph{post hoc} climate regression analysis''.}
 
@@
 hyemalis} increased from $\sim 25$\% to over $50$\% prevalence.
-\linelabel{R2C33-begin}Our model indicates high \revise{confidence}\linelabel{R2C33-end} that overall temporal trends are positive across species (99\% probability of a positive overall year slope in \emph{A.
+\linelabel{R2C33-begin}Our model indicates high \firstrevise{confidence}\linelabel{R2C33-end} that overall temporal trends are positive across species (99\% probability of a positive overall year slope in \emph{A.
 hyemalis}, 92\% probability of a positive overall year slope in \emph{A.
@@
 	\caption[Temporal trends in endophyte prevalence.]{\textbf{Temporal trends in endophyte prevalence.} (A) Histograms show the frequency of scored specimens through time for each host species.
-(B)\revise{ Lines show mean endophyte prevalence predicted by the endophyte prevalence model} over the study period along with the 50\%  and 95\% CI bands incorporating parameter uncertainty and variation associated with collector and scorer random effects.
+(B)\firstrevise{ Lines show mean endophyte prevalence predicted by the endophyte prevalence model} over the study period along with the 50\%  and 95\% CI bands incorporating parameter uncertainty and variation associated with collector and scorer random effects.
 Colored points are binned means of the observed endophyte presence/absence data (black dashes).
@@
 Collector random effects were consistently small (Fig. \ref{fig:collector_fx}), and models fit with and without this random effect provide qualitatively similar results.
-The identity of individual scorers,\linelabel{R2C35-b-begin} \revise{the researchers who identified endophyte status microscopically},\linelabel{R2C35-b-end} did contribute to observed patterns in endophyte prevalence.
-For example, $3$ of the $25$ scorers were significantly more likely than average to assign positive endophyte status, as indicated by 95\% credible intervals greater than zero, \revise{while $4$ of the $25$ had 95\% credible intervals below zero} (Fig.
+The identity of individual scorers,\linelabel{R2C35-b-begin} \firstrevise{the researchers who identified endophyte status microscopically},\linelabel{R2C35-b-end} did contribute to observed patterns in endophyte prevalence.
+For example, $3$ of the $25$ scorers were significantly more likely than average to assign positive endophyte status, as indicated by 95\% credible intervals greater than zero, \firstrevise{while $4$ of the $25$ had 95\% credible intervals below zero} (Fig.
 \ref{fig:scorer_fx}).
@@
 \linelabel{R1C7-begin}
-\revise{Broad increases in prevalence on average, along with increases towards range edges that had low historic prevalence result in range expansions of the symbiosis for both \emph{Agrostis} species (Fig.
+\firstrevise{Broad increases in prevalence on average, along with increases towards range edges that had low historic prevalence result in range expansions of the symbiosis for both \emph{Agrostis} species (Fig.
 \ref{fig:svc_space_year_map}).
@@
 virginicus}).
-\revise{Spatially-varying trends are estimated from the endophyte prevalence model.} Color indicates the relative change in predicted endophyte prevalence.
-\revise{Map lines delineate study areas and do not necessarily depict accepted national boundaries.}}
+\firstrevise{Spatially-varying trends are estimated from the endophyte prevalence model.} Color indicates the relative change in predicted endophyte prevalence.
+\firstrevise{Map lines delineate study areas and do not necessarily depict accepted national boundaries.}}
 	\label{fig:svc_time_map}
@@
 	\includegraphics[width = \linewidth]{../Plots/svc_space_map_year.png}
-	\caption[Predicted endophyte prevalence for each host species in 1895 and 2020.]{\revise{Predicted endophyte prevalence for each host species in 1895 and 2020.
-\revise{Predictions of prevalence come from the endophyte prevalence model.} Color indicates the posterior mean endophyte prevalence for \emph{A.
+	\caption[Predicted endophyte prevalence for each host species in 1895 and 2020.]{\firstrevise{Predicted endophyte prevalence for each host species in 1895 and 2020.
+\firstrevise{Predictions of prevalence come from the endophyte prevalence model.} Color indicates the posterior mean endophyte prevalence for \emph{A.
 hyemalis} (A,D), \emph{A.
@@
 perennans} were most strongly associated with increasing autumn precipitation and with increasing mean and variability in autumn temperature (greater than 97\% posterior probabilities of positive slopes).
-For this species, \revise{each} $1$ $^{\circ}$C increase in autumn temperature was associated with a $1.07$ \% greater increase per year in endophyte prevalence (Fig.
-\ref{fig:climate_correlation_plot}A) and a $100$ mm increase in precipitation was associated with a $0.8$\% \revise{greater} increase per year in endophyte prevalence (Fig.
+For this species, \firstrevise{each} $1$ $^{\circ}$C increase in autumn temperature was associated with a $1.07$ \% greater increase per year in endophyte prevalence (Fig.
+\ref{fig:climate_correlation_plot}A) and a $100$ mm increase in precipitation was associated with a $0.8$\% \firstrevise{greater} increase per year in endophyte prevalence (Fig.
 \ref{fig:climate_correlation_plot}B).
@@
 \ref{fig:climate_correlation_plot}B) (greater than $99$\% posterior probability of a positive slope).
-For this species, an increase of $100$ mm in spring precipitation \revise{was associated with } $0.6$\% per year \revise{stronger} increases in endophyte prevalence \revise{relative to regions with no change in precipitation}.
+For this species, an increase of $100$ mm in spring precipitation \firstrevise{was associated with } $0.6$\% per year \firstrevise{stronger} increases in endophyte prevalence \firstrevise{relative to regions with no change in precipitation}.
 The next greatest slopes were those associated with variability in spring precipitation (greater than $96$\% posterior probability of a negative slope), as well as in the mean and variability of autumn climate (greater than $98$\% probability of negative and positive slopes, respectively).
@@
 \ref{fig:climate_correlation_plot}A,C) experienced the largest increases in endophyte prevalence.
-\linelabel{R2C38-begin}\revise{Our analysis indicated relatively high confidence that these climate drivers influence endophyte prevalence shifts in \emph{E.
-virginicus}}(greater than $99$\% posterior probability of either negative or positive slopes respectively), however they translate, \revise{for example}, to less than a $0.4$\% decrease in endophyte prevalence per year for \revise{each $1^oC$ of summer warming} over the century.\linelabel{R2C38-end}
+\linelabel{R2C38-begin}\firstrevise{Our analysis indicated relatively high confidence that these climate drivers influence endophyte prevalence shifts in \emph{E.
+virginicus}}(greater than $99$\% posterior probability of either negative or positive slopes respectively), however they translate, \firstrevise{for example}, to less than a $0.4$\% decrease in endophyte prevalence per year for \firstrevise{each $1^oC$ of summer warming} over the century.\linelabel{R2C38-end}
 \linelabel{R2C37-begin}Repeating this analysis using all pixels across each species' distribution were qualitatively similar to these results.\linelabel{R2C37-end}
@@
 	\includegraphics[width = \linewidth]{../Plots/climate_trends_plot_intercept.png}
-	\caption[Relationships between predicted trends in endophyte prevalence and changes in seasonal climate drivers.]{\textbf{Relationships between predicted trends in endophyte prevalence and changes in seasonal climate drivers.} Lines show marginal predicted relationship between spatially-varying trends in endophyte prevalence and changes in mean and variability of climate ((A): mean temperature, (B): cumulative precipitation, (C): standard deviation in temperature, (D): standard deviation in precipitation) \revise{estimated from the \emph{post hoc} climate regression analysis}.
+	\caption[Relationships between predicted trends in endophyte prevalence and changes in seasonal climate drivers.]{\textbf{Relationships between predicted trends in endophyte prevalence and changes in seasonal climate drivers.} Lines show marginal predicted relationship between spatially-varying trends in endophyte prevalence and changes in mean and variability of climate ((A): mean temperature, (B): cumulative precipitation, (C): standard deviation in temperature, (D): standard deviation in precipitation) \firstrevise{estimated from the \emph{post hoc} climate regression analysis}.
 Confidence bands represent the 50 and 95\% CI,  colored by host species (\emph{A.
@@
 virginicus}: purple).
-Slopes with greater than 95\% \revise{posterior} probability of being either positive or negative are represented as solid lines while those that have less than 95\% probability are dashed.
-Points are the \revise{values of pre-computed SVC trends  and climate drivers at} 250 randomly sampled pixels across each host's distribution used in model fitting \revise{for the \emph{post hoc} climate regression analysis}.}
+Slopes with greater than 95\% \firstrevise{posterior} probability of being either positive or negative are represented as solid lines while those that have less than 95\% probability are dashed.
+Points are the \firstrevise{values of pre-computed SVC trends  and climate drivers at} 250 randomly sampled pixels across each host's distribution used in model fitting \firstrevise{for the \emph{post hoc} climate regression analysis}.}
 	\label{fig:climate_correlation_plot}
@@
 hyemalis} towards western longitudes (Fig.
-\ref{fig:contemptestplot}A) and \revise{an increase towards} northern latitudes (Fig.
+\ref{fig:contemptestplot}A) and \firstrevise{an increase towards} northern latitudes (Fig.
 \ref{fig:contemptestplot}B).
-\revise{It is noteable that} model predictions for endophyte prevalence exhibited relatively little local geographic variation, whereas the out-of-sample survey data were highly variable with populations spanning 0\% to 100\% endophyte-symbiotic plants (Fig.
-\ref{fig:contemptestplot}C), \revise{indicating population-to-population variation not captured in the endophyte prevalence model.}
+\firstrevise{It is noteable that} model predictions for endophyte prevalence exhibited relatively little local geographic variation, whereas the out-of-sample survey data were highly variable with populations spanning 0\% to 100\% endophyte-symbiotic plants (Fig.
+\ref{fig:contemptestplot}C), \firstrevise{indicating population-to-population variation not captured in the endophyte prevalence model.}
 
@@
 	\caption[Predictive performance for contemporary test data.]{\textbf{Predictive performance for contemporary test data.} (A)
-		 The \revise{endophyte prevalence model}, trained on historic herbarium collection data, performed modestly at predicting prevalence in contemporary population surveys.
+		 The \firstrevise{endophyte prevalence model}, trained on historic herbarium collection data, performed modestly at predicting prevalence in contemporary population surveys.
 The model captured regional trends across (A) longitude and (B) latitude.
@@
 Contemporary prevalence is represented by grey points (point size reflects sample size) along with trend lines from generalized linear models (black line and shaded 95\% confidence interval).
-(C) Comparison of \revise{contemporary} observed \revise{population prevalence} vs.
-predicted endophyte prevalence shows that contemporary test data had more variance between populations than \revise{in model} predictions.}
+(C) Comparison of \firstrevise{contemporary} observed \firstrevise{population prevalence} vs.
+predicted endophyte prevalence shows that contemporary test data had more variance between populations than \firstrevise{in model} predictions.}
 	\label{fig:contemptestplot}
@@
 
-\linelabel{R2C40-begin}Our examination of historic plant specimens revealed \revise{previously hidden} \linelabel{R2C40-end}shifts in microbial symbiosis over the last two centuries. 
-For the three grass host species we examined, there have been strong increases in prevalence of\linelabel{R2C41-begin} \revise{\emph{Epichloë} endophyte symbiosis}.\linelabel{R2C41-end}
+\linelabel{R2C40-begin}Our examination of historic plant specimens revealed \firstrevise{previously hidden} \linelabel{R2C40-end}shifts in microbial symbiosis over the last two centuries. 
+For the three grass host species we examined, there have been strong increases in prevalence of\linelabel{R2C41-begin} \firstrevise{\emph{Epichloë} endophyte symbiosis}.\linelabel{R2C41-end}
 We interpret increases in prevalence of \emph{Epichloë}, which are vertically transmitted, as adaptive changes that improve the fitness of their hosts under increasing environmental stress.
@@
 
-\revise{\subsubsection*{Responses of host-microbe symbioses to climate change}}
+\firstrevise{\subsubsection*{Responses of host-microbe symbioses to climate change}}
 Differences across host species underscore that while all of these $C_3$ grasses share similar broad-scale distributions, each engages in unique biotic interactions and has unique responses to environmental drivers.
 We identified hotspots of change for \emph{A.
-perennans}, which was \revise{the species whose endophyte prevalence was most responsive to changes in climate drivers} (Fig.
+perennans}, which was \firstrevise{the species whose endophyte prevalence was most responsive to changes in climate drivers} (Fig.
 \ref{fig:climate_correlation_plot}).
@@
 In contrast to this expectation, increasing prevalence for \emph{A.
-perennans} was associated with \revise{both} increasing autumn temperature and precipitation (Fig.
+perennans} was associated with \firstrevise{both} increasing autumn temperature and precipitation (Fig.
 \ref{fig:climate_correlation_plot}).
@@
 \ref{fig:climate_correlation_plot}B).
-For \emph{A. hyemalis}, endophytes could be playing a role helping hosts weather autumn-season droughts, which \revise{is likely also} an important time for the species' germination.
+For \emph{A. hyemalis}, endophytes could be playing a role helping hosts weather autumn-season droughts, which \firstrevise{is likely also} an important time for the species' germination.
 Previous work demonstrated drought benefits in a greenhouse manipulation with this host-symbiont pair \citep{davitt2011understanding}, and early life stages may be particularly vulnerable to prolonged droughts.
@@
 virginicus}, which experienced the weakest changes in endophyte prevalence overall (ranging between $1.1$\% increases and $0.2$\% decreases), we only found modest associations with changes in climate drivers.
-Surveys by \citet{sneck2017variation}, used as part of the test data in this study, identified a drought index (SPEI) that integrates precipitation with estimated evapotranspiration as an important predictor of \revise{contemporary} endophyte prevalence \linelabel{R2C14-begin}\revise{in this species.
+Surveys by \citet{sneck2017variation}, used as part of the test data in this study, identified a drought index (SPEI) that integrates precipitation with estimated evapotranspiration as an important predictor of \firstrevise{contemporary} endophyte prevalence \linelabel{R2C14-begin}\firstrevise{in this species.
 The diverse relationships we detect between trends in endophyte prevalence and climate drivers suggest a more complicated picture than the simple explanation that drought alone, as measured through changes in annual precipitation, causes increasing endophyte prevalence through context-dependent fitness benefits.}
 
-\revise{While we show consistent increasing trends in prevalence between the three species, the mechanisms that explain these changes may be diverse and idiosyncratic.
+\firstrevise{While we show consistent increasing trends in prevalence between the three species, the mechanisms that explain these changes may be diverse and idiosyncratic.
 \linelabel{R2C45-begin}First, climate change responses may depend on genotype-specific responses that are not considered in our current analysis. 
@@
 Whether hotspots of change in endophyte prevalence reflect selection for genotype-pairings with particularly strong fitness benefits is an unanswered question.}\linelabel{R2C45-end}
-\linelabel{R2C44-begin}Additionally, \emph{Epichloë} endophytes have been connected to a suite of non-drought related fitness benefits including herbivory \revise{defense} \citep{brem2001epichloe}, salinity resistence \citep{wang2020effects}, and mediation of pathogens \cite{vikuk2019infection} and the soil microbiome \citep{roberts2015rhizosphere}.
-\revise{Broad changes in the distribution and abundance of natural enemies \cite{cote2004ecological}, along with stresses from anthropogenic changes in landcover and pollution \cite{sage2020global} likely influence the benefits of symbiosis \cite{rudgers2020climate}.
-The historic trends that we observed result from the combination of these fitness benefits playing out across the heterogeneous map of a changing climate and other anthropogenic drivers.\linelabel{R2C44-end}
+\linelabel{R2C44-begin}Additionally, \emph{Epichloë} endophytes have been connected to a suite of non-drought related fitness benefits including herbivory \firstrevise{defense} \citep{brem2001epichloe}, salinity resistence \citep{wang2020effects}, and mediation of pathogens \cite{vikuk2019infection} and the soil microbiome \citep{roberts2015rhizosphere}.
+\firstrevise{Broad changes in the distribution and abundance of natural enemies \cite{cote2004ecological}, along with stresses from anthropogenic changes in \linelabel{R3C7-begin}\revise{land management} and pollution \cite{sage2020global} likely influence the benefits of symbiosis \cite{rudgers2020climate}.
+\revise{Host species experience a world that is made increasingly stressful by the combination of these global change drivers, and while historic trends that we observed suggest that symbiotic fitness benefits have helped mitigate this stress, it is likely that at some higher leveler of stress, increasing costs of maintaining the mutualism could lead to declines in endophyte prevalence.}\linelabel{R3C7-end}
+\revise{Changing endophyte prevalence} results from the combination of \revise{net} fitness benefits playing out across the heterogeneous map of a changing climate and \linelabel{R3C8-begin}\revise{and its interactive effects on}other anthropogenic drivers.\linelabel{R2C44-end}\linelabel{R3C8-end}
 
@@
 Previous population surveys \citep{sneck2017variation,rudgers2009benefits,semmartin2015broad} attributed environment-dependent gradients in endophyte prevalence to symbiont-derived fitness benefit's allowing hosts to persist in environments where they otherwise could not \citep{afkhami2014mutualist,kazenel2015mutualistic}.
-\linelabel{R1C11-b-begin}However, symbiont-facilitated range shifts require that endophytes be present in the populations to be able to support population growth. 
+\linelabel{R1C11-b-begin}However, symbiont-facilitated range shifts require that endophytes be present in the populations to be able to \revise{contribute to} population growth. 
 The arid western range edge of \emph{A.
@@
 
-\revise{\subsubsection*{Steps towards forecasts of host-microbe symbioses}}
+\firstrevise{\subsubsection*{Steps towards forecasts of host-microbe symbioses}}
 The combination of a spatially-explicit model and historic herbarium specimens allowed us to identify regions of both increasing and decreasing endophyte prevalence, however we see several next steps toward the goal of predicting host and symbiont niche-shifts in response to future climate change.
 While the model recreated the large-scale spatial trends observed in contemporary population surveys, test data contained more population-to-population variability in prevalence than could be explained by the model.
-\linelabel{R2C39-begin}\revise{We interpret this to mean that the model captures coarse-scale spatial and temporal trends reasonably well, but is not equipped to capture local-scale nuances that generate population-to-population differences.} \linelabel{R2C39-end}
-Validating our model predictions \revise{with this test}, a rare extra step in collections-based studies, allows us to identify ways in which the model's out-of-sample predictive ability could be improved.
-\linelabel{R2C43-begin}Lack of information on local variability \revise{in symbiont prevalence} may simply be a feature of data derived from herbarium specimens. 
-\revise{Natural history collectors sample one or a few specimens from local populations, and these observations are aggregated by the model to derive broad-scale estimates.}\linelabel{R2C43-end}
-This suggests that increasing local replication should be a factor considered in future collection efforts of natural history specimens, \revise{balancing} the required time and effort \linelabel{R1C9-begin} \revise{along with limitations on storage space within collections}.
+\linelabel{R2C39-begin}\firstrevise{We interpret this to mean that the model captures coarse-scale spatial and temporal trends reasonably well, but is not equipped to capture local-scale nuances that generate population-to-population differences.} \linelabel{R2C39-end}
+Validating our model predictions \firstrevise{with this test}, a rare extra step in collections-based studies, allows us to identify ways in which the model's out-of-sample predictive ability could be improved.
+\linelabel{R2C43-begin}Lack of information on local variability \firstrevise{in symbiont prevalence} may simply be a feature of data derived from herbarium specimens. 
+\firstrevise{Natural history collectors sample one or a few specimens from local populations, and these observations are aggregated by the model to derive broad-scale estimates.}\linelabel{R2C43-end}
+This suggests that increasing local replication should be a factor considered in future collection efforts of natural history specimens, \firstrevise{balancing} the required time and effort \linelabel{R1C9-begin} \firstrevise{along with limitations on storage space within collections}.
 \linelabel{R1C9-end}
+\linelabel{R3C6-begin}\revise{An alternative validation test would be to hold-out samples from within the historic data set.
+Such a test would more clearly provide a set of data that matched the conditions of the training data (i.e in spatial scale and climate conditions) missing within the contemporary test data, however a relatively limited number of sampled specimens held us back from exploring this option.
+Splitting datasets can negatively impact model estimates, and the choice of how to split the data for model validation is not trivial \citep{bergmeir2012use,james2013introduction}. 
+\linelabel{R4C10b-begin}From a modeler's perspective, this also urges advocacy for increased collection efforts and expansion of herbarium collections.}\linelabel{R3C6-end}\linelabel{R4C10b-end}
+
+\linelabel{R3C3-being}\revise{Another key consideration in forecasting the dynamics host-microbe symbioses is the spatial scale of both specimen georeferencing and available climate data. 
+For this analysis, most specimen localities were assigned coordinates at county or city centroids, and the climate data examined was on 4 km grid cells.
+Georeferencing of specimens as accurately as possible is a key priority of herbarium specimen digitization efforts \citep{soltis2017digitization,davis2023herbarium}.
+While the INLA modeling approach that we used allows for predictions at arbitrarility small spatial scales, and would simplify connecting model predictions to the scale of a given climate driver, the course scale inherent to our analysis may obscure some local-scale ecological processes} \linelabel{R3C3-end}
 Poor predictive ability at local scales in this grass-endophyte system is not surprising, as previous studies have found that local variation, even to the scale of hundreds of meters can structure endophyte-host niches \cite{kazenel2015mutualistic,gundel2024temporal}.
@@
 \ref{fig:svc_space_year_map}A) is a result of environmental conditions that lead the symbiosis to have negative fitness consequences, or is a result of some historical contingency or dispersal limitation that has thus far limited the presence of symbiotic hosts from a region where they would otherwise flourish and provide resilience.
+\linelabel{R3C5-begin}\revise{Another interesting question is whether local adaption, in either host or symbiont may influence future range shifts.}\linelabel{R3C5-end}
 Incorporating available climatic and soil layers as covariates is another obvious step that could improve predictions.
@@
 
-\revise{\subsubsection*{Herbaria for global change research}}
+\firstrevise{\subsubsection*{Herbaria for global change research}}
 Our analysis advances the use of herbarium specimens in global change biology in two ways. 
 \linelabel{R1C10-begin}
-First and foremost, this is \revise{one of a growing number of studies to examine microbial symbiosis using specimens from natural history collections, and the first, to our knowledge, to link long-term changes in the symbioses to changes in climate.}\linelabel{R1C10-end}
+First and foremost, this is \firstrevise{one of a growing number of studies to examine microbial symbiosis using specimens from natural history collections, and the first, to our knowledge, to link long-term changes in the symbioses to changes in climate.}\linelabel{R1C10-end}
 The responses of microbial symbioses are a rich target for future studies within historic specimens, particularly those that take advantage of advances in sequencing technology.
-While we used relatively coarse presence/absence data based on fungal morphology, other studies have examined historic plant microbiomes using molecular sequencing and sophisticated bioinformatics techniques, but these studies have so far been limited to relatively few specimens at limited spatial extents \cite{yoshida2015computational,heberling2019utilizing, bieker2020metagenomic,gross2021hidden,bradshaw2021global}.
+While we used relatively coarse presence/absence data based on fungal morphology, other studies have examined historic plant microbiomes using molecular sequencing and sophisticated bioinformatics techniques, but these studies have so far been limited to relatively few specimens at limited spatial extents \cite{bearchell2005wheat,yoshida2015computational,heberling2019utilizing, bieker2020metagenomic,gross2021hidden,bradshaw2021global,bradshaw2023genetic}.
+\linelabel{R4C11b-begin}\revise{Much of this work highlights the important role that historic specimens can play in tracking pathogens, a particularly important area as climate change facilitates the spread of new diseases \cite{ristaino2020importance,singh2023climate}}\linelabel{R4C11b-end}
 Continued advances in capturing historic DNA and in filtering out potential contamination during specimen storage \citep{daru2019novel,bakker2020herbarium,raxworthy2021mining} will be imperative in the effort to scale up these efforts.
-This scaling up will be essential to be able to quantify changes not just in the prevalence of symbionts, but also in symbionts' intraspecific variation and evolutionary responses to climate change, as well as in changes in the wider \revise{host microbiome}.
-With improved molecular insights from historic specimens, we could ask whether the broad increases in endophytes that we have identified reflect selection for particular \revise{genetic strains or} chemotypes and how this selection varies across space.
+This scaling up will be essential to be able to quantify changes not just in the prevalence of symbionts, but also in symbionts' intraspecific variation and evolutionary responses to climate change, as well as in changes in the wider \firstrevise{host microbiome}.
+With improved molecular insights from historic specimens, we could ask whether the broad increases in endophytes that we have identified reflect selection for particular \firstrevise{genetic strains or} chemotypes and how this selection varies across space.
 Answering these questions as well as the unknown questions that future researchers may ask also reiterates the value in capturing meta-information during ongoing digitization efforts at herbaria around the world and during the accession of newly collected specimens \citep{lendemer2020extended,edwards2024university}.
@@
 \linelabel{R2C46-begin}
-\revise{The second major advance in this analysis is} \linelabel{R2C46-end} in accounting for several potential biases in the data observation process that may be common to many collections-based research questions by using a spatially-explicit random effects model.
+\firstrevise{The second major advance in this analysis is} \linelabel{R2C46-end} in accounting for several potential biases in the data observation process that may be common to many collections-based research questions by using a spatially-explicit random effects model.
 Potential biases introduced by the sampling habits of collectors \citep{daru2018widespread}, and variation between contemporary researchers during the collection of trait data, if not corrected for could lead to over-confident inference about the strength and direction of historic change (Fig.
@@
 Previous studies that have quantified the effects of collector biases typically find them to be small  \cite{davis2015herbarium,meineke2019herbarium}, and we similarly did not find that collector has a strong effect on the results of our analysis, but that scorer identity did impact results.
-\linelabel{R2C36-begin}\revise{It is difficult to distinguish whether the impact of scorers was driven by true differences in scorers' biases or by unintended spatial or temporal clustering of the specimens examined by each scorer \cite{clayton1993spatial,urdangarin2023evaluating}.
+\linelabel{R2C36-begin}\firstrevise{It is difficult to distinguish whether the impact of scorers was driven by true differences in scorers' biases or by unintended spatial or temporal clustering of the specimens examined by each scorer \cite{clayton1993spatial,urdangarin2023evaluating}.
 By under-weighting endophyte-positive samples that are clustered spatially or by collector or observer, the endophyte prevalence model is appropriately accounting for nuisance variables and providing a conservative inference of endophyte change relative to the raw data.
 }\linelabel{R2C36-end}
-\linelabel{R1C4-dicussion-begin}\revise{
+\linelabel{R1C4-dicussion-begin}\firstrevise{
 Spatial autocorrelation is another phenomenon likely common in data derived from herbarium specimens \cite{willems2022forest}, which our spatially-explicit analysis models among samples.
@@
 Simulation analyses that extend this work to consider the myriad ways herbarium data may be biased (i.e.
-testing different spatial arrangements and scales of spatial bias, or testing different sample sizes) would be extremely valuable \cite{gaul2020data,erickson2021accounting,daru2018widespread,meineke2021bias}.
+testing different spatial arrangements and scales of spatial bias, or testing different sample sizes) would be extremely valuable \cite{gaul2020data,erickson2021accounting,daru2018widespread,meineke2021bias,schmidt2025collector}.
 }\linelabel{R1C4-discussion-end}
@@
 
-\revise{\subsubsection*{Conclusion}}
+\subsubsection*{Conclusion}
 Ultimately, a central goal of global change biology is to generate predictive insights into the future of natural systems on a rapidly changing planet. 
 Beyond host-microbe symbioses, detecting ecological responses to anthropogenic global change and attributing their causes would inform public policy decision-makers and adaptive management strategies.
+\linelabel{R4C10c-begin}\revise{Natural history specimens, such as the plant hosts examined in this study have a clear role to play in informing global change biodiversity science, including building understanding of the dynamics of host-symbiont interactions \cite{davis2023herbarium}.}\linelabel{R4C10c-end}\
 This survey of historic endophyte prevalence is necessarily correlative, yet it serves as a foundation to develop better predictive models of the response of microbial symbioses to climate change. 
@@
 	This research was supported by funding from National Science Foundation (grants 1754468 and 2208857) and by funding from the Texas Ecolab Program.
-\revise{Two anonymous reviewers greatly improved an earlier version of this manuscript.}
+\firstrevise{Two anonymous reviewers greatly improved an earlier version of this manuscript.}
 
@@
 
-\linelabel{R2C47-begin}\revise{
+\linelabel{R2C47-begin}\firstrevise{
 
@@
 To visualize temporal change, the data are faceted before and after the median year of collection.
-\revise{Map lines delineate study areas and do not necessarily depict accepted national boundaries.}}
+\firstrevise{Map lines delineate study areas and do not necessarily depict accepted national boundaries.}}
 		\label{fig:endo_status_map}
@@
 	\includegraphics[width = .5\linewidth]{../Plots/overlay_plot.png}
-	\caption[Graphical posterior predictive check of the endophyte prevalence model fit]{\revise{\textbf{Graphical posterior predictive check of \revise{the endophyte prevalence} model fit.}} Consistency between \revise{observed} data and \revise{predicted} values indicate that the fitted model accurately describes the data.
-Graph shows density curves for the observed data (black) along with 100 \revise{predicted} datasets (red).}
+	\caption[Graphical posterior predictive check of the endophyte prevalence model fit]{\firstrevise{\textbf{Graphical posterior predictive check of \firstrevise{the endophyte prevalence} model fit.}} Consistency between \firstrevise{observed} data and \firstrevise{predicted} values indicate that the fitted model accurately describes the data.
+Graph shows density curves for the observed data (black) along with 100 \firstrevise{predicted} datasets (red).}
 			\label{fig:overlayplot}
@@
 		\includegraphics[width = .5\linewidth]{../Plots/ROC_training_plot.png}
-		\caption[ROC plot showing performance of the endophyte prevalence model in classifying observations according to endophyte status within the in-sample data]{\textbf{ROC plot showing performance \revise{of the endophyte prevalence model in} classifying observations according to endophyte status within the in-sample \revise{training} data \revise{from herbarium collections}.} The curves show adequate model performance for observed data.
+		\caption[ROC plot showing performance of the endophyte prevalence model in classifying observations according to endophyte status within the in-sample data]{\textbf{ROC plot showing performance \firstrevise{of the endophyte prevalence model in} classifying observations according to endophyte status within the in-sample \firstrevise{training} data \firstrevise{from herbarium collections}.} The curves show adequate model performance for observed data.
 The AUC value is 0.79.}
@@
 		\includegraphics[width = .5\linewidth]{../Plots/ROC_test_plot.png}
-		\caption[ROC plot showing performance of the endophyte prevalence model in classifying observations according to endophyte status within the out-of-sample data.]{\textbf{ROC plot showing \revise{performance of the endophyte prevalence model in} classifying observations according to endophyte status within the out-of-sample \revise{test} data \revise{from contemporary surveys}.} The curves show adequate model performance for test data.
+		\caption[ROC plot showing performance of the endophyte prevalence model in classifying observations according to endophyte status within the out-of-sample data.]{\textbf{ROC plot showing \firstrevise{performance of the endophyte prevalence model in} classifying observations according to endophyte status within the out-of-sample \firstrevise{test} data \firstrevise{from contemporary surveys}.} The curves show adequate model performance for test data.
 The AUC value is 0.77.}
@@
 	\includegraphics[width = \linewidth]{../Plots/posterior_plot.png}
-	\caption[Posterior estimates of parameters describing global temporal trends in endophyte prevalence trends.]{\revise{\textbf{Posterior estimates of parameters describing global intercept and temporal trends from the endophyte prevalence model.}} Density curves show the probability density along with mean (colored line) and $95$\% CI (black lines) for the (A) intercept and (B) slope terms, $A$ and $T$ respectively \revise{from Eqn.
+	\caption[Posterior estimates of parameters describing global temporal trends in endophyte prevalence trends.]{\firstrevise{\textbf{Posterior estimates of parameters describing global intercept and temporal trends from the endophyte prevalence model.}} Density curves show the probability density along with mean (colored line) and $95$\% CI (black lines) for the (A) intercept and (B) slope terms, $A$ and $T$ respectively \firstrevise{from Eqn.
 \ref{eq:trends}}.
@@
 	\includegraphics[width = \linewidth]{../Plots/svc_time_map_CI.png}
-	\caption[Credible interval width of temporal trends in endophyte prevalence across the distribution of each host species estimated from the endophyte prevalence model.]{\revise{\textbf{Credible interval width of temporal trends in endophyte prevalence across the distribution of each host species estimated from the endophyte prevalence model.}} Shading represents the range of the 95\% posterior credible interval given in units of \emph{\% change in prevalence/year} for spatially varying slopes, $\tau$ \revise{from Eqn.
+	\caption[Credible interval width of temporal trends in endophyte prevalence across the distribution of each host species estimated from the endophyte prevalence model.]{\firstrevise{\textbf{Credible interval width of temporal trends in endophyte prevalence across the distribution of each host species estimated from the endophyte prevalence model.}} Shading represents the range of the 95\% posterior credible interval given in units of \emph{\% change in prevalence/year} for spatially varying slopes, $\tau$ \firstrevise{from Eqn.
 \ref{eq:trends}}.
-\revise{Map lines delineate study areas and do not necessarily depict accepted national boundaries.}}
+\firstrevise{Map lines delineate study areas and do not necessarily depict accepted national boundaries.}}
 	\label{fig:svc_time_map_CI}
@@
 	\includegraphics[width = \linewidth]{../Plots/conservative_comparison_plot.png}
-	\caption[Comparison of parameter estimates from endophyte prevalence model fit to data with liberal versus conservative endophyte scores]{\revise{\textbf{Comparison of endophyte prevalence model estimates fit to data with liberal versus conservative endophyte scores.} Liberal and conservative scores document uncertainty in the endophyte identification process.
+	\caption[Comparison of parameter estimates from endophyte prevalence model fit to data with liberal versus conservative endophyte scores]{\firstrevise{\textbf{Comparison of endophyte prevalence model estimates fit to data with liberal versus conservative endophyte scores.} Liberal and conservative scores document uncertainty in the endophyte identification process.
 Each specimen was given both a liberal and conservative scores.
 In cases of uncertain identification, the liberal status assumed a potential endophyte identification was more likely to be endophyte-positive while the conservative status assumed that the potential endophyte identification was less likely to be endophyte-positive}.
-(A) Posterior estimates of global temporal trend \revise{($T$ from Eqn.
+(A) Posterior estimates of global temporal trend \firstrevise{($T$ from Eqn.
 \ref{eq:trends}}) for the endophyte prevalence model fit to liberal scores (grey) and to conservative scores (blue).
-Maps show the spatially varying temporal trend estimates \revise{($\tau$ \revise{from Eqn.
+Maps show the spatially varying temporal trend estimates \firstrevise{($\tau$ \firstrevise{from Eqn.
 \ref{eq:trends}})} from the endophyte prevalence model fit to conservative scores for (B) \emph{A.
@@
 Note that the color scale differs between this visualization and Fig.
-\ref{fig:svc_time_map} \revise{that shows estimates fit using liberal endophyte scores.}}
+\ref{fig:svc_time_map} \firstrevise{that shows estimates fit using liberal endophyte scores.}}
 	\label{fig:conservative_scores_plot}
@@
 	\includegraphics[width = .6\linewidth]{../Plots/collector_posterior.png}
-	\caption[Posterior estimates of collector random effects from endophyte prevalence model]{\textbf{Posterior estimates of collector random effects \revise{from endophyte prevalence model.}} \revise{Collector random effects are denoted $\chi$ in Eqn.
+	\caption[Posterior estimates of collector random effects from endophyte prevalence model]{\textbf{Posterior estimates of collector random effects \firstrevise{from endophyte prevalence model.}} \firstrevise{Collector random effects are denoted $\chi$ in Eqn.
 \ref{eq:trends} and represent variance associated with researchers who collected historic herbarium specimens.} Points show posterior median along with 95\% CI for each of $924$ individual collectors.}
@@
 	\includegraphics[width = .6\linewidth]{../Plots/scorer_posterior.png}
-	\caption[Posterior estimates of scorer random effects from endophyte prevalence model]{\textbf{Posterior estimates of scorer random effects \revise{from endophyte prevalence model.}} \revise{Scorer random effects are denoted $\omega$ in Eqn.
-\ref{eq:trends} and represent variance associated with researchers who identified \emph{Epichloë} endophytes within herbarium specimen tissue samples.} Points show posterior median along with 95\% CI for each of $25$ individual \revise{scorers}.}
+	\caption[Posterior estimates of scorer random effects from endophyte prevalence model]{\textbf{Posterior estimates of scorer random effects \firstrevise{from endophyte prevalence model.}} \firstrevise{Scorer random effects are denoted $\omega$ in Eqn.
+\ref{eq:trends} and represent variance associated with researchers who identified \emph{Epichloë} endophytes within herbarium specimen tissue samples.} Points show posterior median along with 95\% CI for each of $25$ individual \firstrevise{scorers}.}
 	\label{fig:scorer_fx}
@@
 	\includegraphics[width = \linewidth]{../Plots/contemp_surveys_map.png}
-	\caption[Locations of contemporary surveys of endophytes used as "test" data to evvaluate predictive ability of the endophyte prevalence model]{\textbf{Locations of contemporary surveys of endophytes used as "test" data to \revise{evaluate predictive ability of the endophyte prevalence model.}} \revise{Points are locations of host populations surveyed between 2013 and 2019 for endophytes, colored by species (\emph{A.
+	\caption[Locations of contemporary surveys of endophytes used as "test" data to evvaluate predictive ability of the endophyte prevalence model]{\textbf{Locations of contemporary surveys of endophytes used as "test" data to \firstrevise{evaluate predictive ability of the endophyte prevalence model.}} \firstrevise{Points are locations of host populations surveyed between 2013 and 2019 for endophytes, colored by species (\emph{A.
 hyemalis}: green, \emph{E.
@@
 	\caption[Change in seasonal climate variables between the periods 1895-1925 and 1990-2020 across the distribution of A.
-hyemalis]{\textbf{Change in seasonal climate variables between the periods 1895-1925 and 1990-2020 \revise{across the distribution of \emph{A.
-hyemalis}}.} Color represents change in (A) seasonal temperature \revise{($^o$C)} and (B) seasonal precipitation \revise{(mm.)}.
+hyemalis]{\textbf{Change in seasonal climate variables between the periods 1895-1925 and 1990-2020 \firstrevise{across the distribution of \emph{A.
+hyemalis}}.} Color represents change in (A) seasonal temperature \firstrevise{($^o$C)} and (B) seasonal precipitation \firstrevise{(mm.)}.
 Maps show pixels covering the modeled distribution of \emph{A.
-hyemalis} used in \revise{\emph{post hoc} climate regression analysis.}}
+hyemalis} used in \firstrevise{\emph{post hoc} climate regression analysis.}}
 	\label{fig:AGHY_climate_covariates}
@@
 	\caption[Change in seasonal climate variables between the periods 1895-1925 and 1990-2020 across the distribution of A.
-perennans]{\textbf{Change in seasonal climate variables between the periods 1895-1925 and 1990-2020 \revise{across the distribution of \emph{A.
-perennans}}.} Color represents change in (A) seasonal temperature \revise{($^o$C)} and (B) seasonal precipitation \revise{(mm.)}.
+perennans]{\textbf{Change in seasonal climate variables between the periods 1895-1925 and 1990-2020 \firstrevise{across the distribution of \emph{A.
+perennans}}.} Color represents change in (A) seasonal temperature \firstrevise{($^o$C)} and (B) seasonal precipitation \firstrevise{(mm.)}.
 Maps show pixels covering the modeled distribution of \emph{A.
-perennans} used in \revise{\emph{post hoc} climate regression analysis.}}
+perennans} used in \firstrevise{\emph{post hoc} climate regression analysis.}}
 	\label{fig:AGPE_climate_covariates}
@@
 	\caption[Change in seasonal climate variables between the periods 1895-1925 and 1990-2020 across distribution of \emph{E.
-virginicus}]{\textbf{Change in seasonal climate variables between the periods 1895-1925 and 1990-2020 \revise{across the distribution of \emph{E.
-virginicus}}.} Color represents change in (A) seasonal temperature \revise{($^o$C)} and (B) seasonal precipitation \revise{(mm.)}.
+virginicus}]{\textbf{Change in seasonal climate variables between the periods 1895-1925 and 1990-2020 \firstrevise{across the distribution of \emph{E.
+virginicus}}.} Color represents change in (A) seasonal temperature \firstrevise{($^o$C)} and (B) seasonal precipitation \firstrevise{(mm.)}.
 Maps show pixels covering the modeled distribution of \emph{E.
-virginicus} used in \revise{\emph{post hoc} climate regression analysis.}}
+virginicus} used in \firstrevise{\emph{post hoc} climate regression analysis.}}
 	\label{fig:ELVI_climate_covariates}
@@
 	\begin{table}[h]
-		\caption{Summary of herbarium samples across collections \revise{(no. of specimens)}}
+		\caption{Summary of herbarium samples across collections \firstrevise{(no. of specimens)}}
 		\label{table:herbaria}
@@
 	\includegraphics[width = \linewidth]{../Plots/prior_comparison_year_plot.png}
-	\caption[Overall trend in endophyte prevalence evaluated for endophyte prevalence models with different range priors on spatially structured random effects, and for two different meshes]{\textbf{Overall trend in endophyte prevalence evaluated for \revise{endophyte prevalence} models with different range priors on spatially structured random effects, and for two different \revise{triangulation} meshes.} \revise{Data used in model fitting is the same across all panels and as in the main text}.
+	\caption[Overall trend in endophyte prevalence evaluated for endophyte prevalence models with different range priors on spatially structured random effects, and for two different meshes]{\textbf{Overall trend in endophyte prevalence evaluated for \firstrevise{endophyte prevalence} models with different range priors on spatially structured random effects, and for two different \firstrevise{triangulation} meshes.} \firstrevise{Data used in model fitting is the same across all panels and as in the main text}.
 Note that these plots, as compared to Fig.
-2 in main text, show mean trends and do not incorporate \revise{variance} associated with collector and scorer random effects.}
+2 in main text, show mean trends and do not incorporate \firstrevise{variance} associated with collector and scorer random effects.}
 	\label{fig:prior_year_plot}
@@
 	\includegraphics[width = .8\linewidth]{../Plots/standard_mesh_comparison_svc_plot.png}
-	\caption[Spatially-varying trends in endophyte prevalence evaluated for models with different range priors on spatially structured random effects, and for the "standard" mesh]{\textbf{Spatially-varying trends in endophyte prevalence evaluated for \revise{the endophyte prevalence} model with different range priors on spatially structured random effects, and for the "standard" mesh.} \revise{Data used in model fitting is the same across all panels and as in the main text}.
+	\caption[Spatially-varying trends in endophyte prevalence evaluated for models with different range priors on spatially structured random effects, and for the "standard" mesh]{\textbf{Spatially-varying trends in endophyte prevalence evaluated for \firstrevise{the endophyte prevalence} model with different range priors on spatially structured random effects, and for the "standard" mesh.} \firstrevise{Data used in model fitting is the same across all panels and as in the main text}.
 Shading indicates the magnitude and direction of predicted trends for each of three host species for each of three prior ranges (rows A-C).
@@
 	\includegraphics[width = .8\linewidth]{../Plots/finer_mesh_comparison_svc_plot.png}
-	\caption[Spatially-varying trends in endophyte prevalence evaluated for the endophyte prevalence model with different range priors on spatially structured random effects, and for the "finer" mesh]{\textbf{Spatially-varying trends in endophyte prevalence evaluated for \revise{the endophyte prevalence} model with different range priors on spatially structured random effects, and for the "finer" mesh.} \revise{Data used in model fitting is the same across all panels and as in the main text}.
+	\caption[Spatially-varying trends in endophyte prevalence evaluated for the endophyte prevalence model with different range priors on spatially structured random effects, and for the "finer" mesh]{\textbf{Spatially-varying trends in endophyte prevalence evaluated for \firstrevise{the endophyte prevalence} model with different range priors on spatially structured random effects, and for the "finer" mesh.} \firstrevise{Data used in model fitting is the same across all panels and as in the main text}.
 Shading indicates the magnitude and direction of predicted trends for each of three host species for each of three prior ranges (rows A-C).
@@
 \linelabel{R1C4-begin}
-\revise{To examine how data that is unevenly distributed across host distributions may influence interpretation of spatially-varying coefficients, we performed a simulation analysis. 
+\firstrevise{To examine how data that is unevenly distributed across host distributions may influence interpretation of spatially-varying coefficients, we performed a simulation analysis. 
 Our focal species, \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}, are widely distributed grasses across the eastern United States that host \emph{Epichloë} fungal endophytes.

### 2025-07-14 Manuscript/Reviewer_response_endo_herbarium_R2.tex
edits to response letter

@@
 Thank you for the opportunity to submit a second revision of our manuscript to \emph{Global Change Biology}. 
-We appreciate that the reviewers acknowledged the positive attributes of our study while providing guidance to clarify our methods and improve our work's connections to the literature. 
-In response to their comments, we have revised throughout our manuscript to address key modeling assumptions and choices, as well as editing our manuscript to highlight connections to the quickly growing field of research that that uses herbarium specimens to study global change.
+The reviewers provided positive feedback on our previous submission while offering some further guidance to clarify our methods and improve our work's connections to the literature. 
+In response to their comments, we have revised our manuscript to address key modeling assumptions and choices, and to highlight connections to the quickly growing field of research that uses herbarium specimens to study biological responses to global change.
 We detail these changes in the point-by-point responses below.
-All of our changes are denoted in the revised manuscript with \revise{Mahogany font}.
+All of our changes are denoted in the revised manuscript with \revise{Mahogany font} and referenced below with line numbers.
 
@@
 	\response{We thank Reviewer 3 for acknowledging the strength of our statistical approach. 
-		In response to this comment, we have edited our methods section to provide context on the general INLA approach and it's assumptions, which have been ennumerated by the teams who developed this method.
+		In response to this comment, we have edited our methods section to provide context on the Integrated Nested Laplace Approximation (INLA) approach and it's assumptions, which have been enumerated by the teams who developed this method.
 In addition, we point out key areas where we expect this may influence our results.}
@@
 }
-	\response{We have added additional clarification about the assumptions that the Matérn covariance function rely upon, and about prior choice within our methods section.
-In addition, we have added material to our Discussion that highlights the important inferences provided by our sensitivity analysis}
+	\response{We have added clarification about the assumptions of the Matérn covariance function, and about our choice of priors within the Methods section.
+In addition, we have added material to our Discussion that highlights the important inferences \tom{provided by our sensitivity analysis}{Is this sensitivity analysis new?
+If not, emphasize that we already had a sensitivity analysis in the previous submission.
+If yes, emphasize that we added a new sensitivity analysis AND added discussion material about it.}.}
 	\linesref{ms-R3C2a-begin}{ms-R3C2a-end}
@@
 Consider discussing the potential implications of this mismatch.}
-\response{We have edited our methods section to more clearly explain that our goal in this analysis was to explore temperature and precipitation as the key climate factors of global change, that we expected there to be differences across seasons in the magnitude of climate change in these drivers, and that these seasons connect to the biology of our focal species - that they differ in their growing season.
+\response{We appreciate these suggestions.
+We have edited our methods section to more clearly explain the rationale behind our analysis of climate drivers.
+Specifically, we expected there to be differences across seasons in the magnitude of climate change in temperature and precipitation, because these seasons connect to the biology and natural history of our focal species.
 In addition, we have added to our discussion to acknowledge the potential importance of local scale ecological processes that may influence our analysis.
 We agree that the potential mismatch in scales is a limitation to herbarium studies in general, compounded not just by the scale of the climate data, but also by the spatial scale of georeferencing used to connect herbarium specimens to localities.
-The INLA modeling approach we employ cannot overcome inherent scale-limitations in the data, however treating space in a continuous way does allow us to make predictions at arbitrary spatial scales.
+The INLA modeling approach we employ cannot overcome inherent scale-limitations in the data, \tom{however treating space in a continuous way does allow us to make predictions at arbitrary spatial scales.
 Our climate analysis used predicted trends from the model to assess relationships with climate, meaning that the climate data were actually on a finer scale than the endophyte trend predictions.
-However this is, as Reviewer 3 notes, a broader scale than many potential ecological processes involved.}
+However this is, as Reviewer 3 notes, a broader scale than many potential ecological processes involved.}{I find this a little hard to follow (and I have not read the changes to the ms yet).
+Can we simplify this response and just say it's a good point and we have added discussion to acknowledge the implications of the scale mismatch?
+That's all the reviewer was asking for.}}
 \linesref{ms-R3C3-begin}{ms-R3C3-end}
@@
 A discussion of these potential factors (e.g., soil conditions, microclimate, local adaptation) and how they could be incorporated into future models would be valuable.}
-\response{We agree with Reviewer 3 that there is a wide array of other potential factors that would make valuable future directions, and the existing discussion touched on this already.
+\response{We agree with Reviewer 3 that there is a wide array of other local-scale factors that would make valuable future directions, and the Discussion section of our previous submission touched on this already.
 We have expanded the existing material on this subject.}
@@
 It would be beneficial to acknowledge and discuss any potential limitations associated with using contemporary data for out-of-sample validation, particularly if climate conditions or other environmental factors have changed significantly during this period.}
-\response{In response to this comment, we have revised our Discussion section to include further discussion of the potential implications of the out-of-sample validation test, particularly highlighting that the data do differ in scale and potentially climate conditions, and how this might inform future modeling efforts.
+\response{This is a great point.
+In response to this comment, we have revised our Discussion section to include further discussion of the potential implications of the out-of-sample validation test, \tom{particularly highlighting that the data do differ in scale and potentially climate conditions, and how this might inform future modeling efforts.
 Reviewer 3 points to potential non-stationarity as an explanation for mismatch in model validation.
-Previous investigations have found that stationarity is often a reasonable assumption to make and that there are not meaningul differences in out-of-sample validation tests that use blocks within and without time series (Bergmeir and Benitez, 2012, \emph{Information Sciences}).}
+Previous investigations have found that stationarity is often a reasonable assumption to make and that there are not meaningul differences in out-of-sample validation tests that use blocks within and without time series (Bergmeir and Benitez, 2012, \emph{Information Sciences}).}{I don't really get the scale and stationarity ideas from the reviewer's comment, so this seems a little tangential to me.
+I think they were simply observing that the census years were at the very end of the herbarium years, so the model predictions are probably extra-uncertain.
+To me, nonstationarity is already built into the model via the year effect.
+}}
 \linesref{ms-R3C6-begin}{ms-R3C6-end}
@@
 \response{Thank you for urging us to further explore our interpretation of changes in endophyte prevalence.
-We also recognize that we can not definitively identify these changes in prevalence as adaptive changes in response to climate.
-In fact, we have recently submitted an additional manuscript that explore the potential role that other anthropogenic global change drivers, including land use and nitrogen deposition, have on changes in endophyte prevalence.
-While we touched on these ideas already in the existing Discussion, in response to this comment, we have edited our Discussion to more clearly communicate the potential for land management and other factors to contribute to these trends.}
+We agree with the reviewer that causal mechanisms underlying these changes in prevalence should be interpreted with caution.
+In fact, \tom{we have recently submitted an additional manuscript}{Some journals ask to see any related manuscripts that are under review, so check GCB guidelines.} that explore the potential role of other anthropogenic global change drivers, including land use and nitrogen deposition.
+While our previous submission already touched on these ideas in the Discussion, in response to this comment, we have revised our Discussion to more clearly communicate the potential for other factors to contribute to these trends.}
 \linesref{ms-R3C7-begin}{ms-R3C7-end}
@@
 }
-\response{This is a great point that different factors may themselves interact with changing climate, and is also connected to Reviewer 3's Comment 7.
-And so we have edited our discussion material to more clearly communicte that these drivers themselve may interact with changing climate and its potential consequences for endophyte prevalence.}
+\response{This is a great point that non-climatic factors may interact with changing climate, and is also connected to Reviewer 3's Comment 7.
+And so we have edited our discussion material to more clearly communicate the potential for interactions between climate drivers and other factors that elicit beneficial effects of endophyte symbiosis.}
 \linesref{ms-R3C8-begin}{ms-R3C8-end}
@@
 This innovative and well-executed study highlights the underutilized value of herbarium collections and is well-suited for publication in Global Change Biology.}
-\response{We thank Reviewer 4 for their positive evaluation of our study, and we agree that the use of herbarium collections is a particuarly valuable part of this research and for studies of global change in particular.}
+\response{We thank Reviewer 4 for their positive evaluation of our study, and we agree that herbarium collections are uniquely valuable yet underused in studies of biological responses to global change.}
 
@@
 The authors could note that herbaria have traditionally supported taxonomic studies but are now being used in broader fields, including global change biology, biodiversity monitoring, and disease ecology.}
-\response{In response to this comment, we have added references throughout our introduction that touch on these extended uses of herbarium specimens, and have also edited the Discussion and Conclusion sections to highlight the potential role that herbaria can play in global change research.}
+\response{This is a great suggestion.
+In response to this comment, we have added references throughout our introduction that touch on these extended uses of herbarium specimens, and have also edited the Discussion and Conclusion sections to highlight the potential role that herbaria can play in global change research.}
 \linesref{ms-R4C10-begin}{ms-R4C10-end}
@@
 \response{Thank you for pointing us towards this body of work.
-We have added relevant citations throughout the introduction and discussion and also expanded the discussion to acknowledge the potential role that herbarium specimens can play in understanding plant-pathogen dynamics.}
+We have added relevant citations throughout the introduction and discussion and also expanded the discussion to acknowledge the potential role that herbarium specimens can play in understanding plant-pathogen interactions as a form of symbiosis.}
 \linesref{ms-R4C11a-begin}{ms-R4C11a-end}

### 2025-07-14 Manuscript/endo_herbarium_R2.tex
Added my comments
intro edits

@@
 \author{Joshua C. Fowler$^{1,2\ast}$ \\
-	Jacob Moutouama$^{1}$\\
+	Jacob Moutouama$^{1,3}$\\
 	Tom E. X. Miller$^{1}$}
@@
 	\noindent{} 2. University of Miami, Department of Biology, Miami, Florida;
+	\noindent{} 3. Department of Botany, University  of British Columbia, 6270 University Blvd, Vancouver, BC V6T 1Z4
 
@@
 
-	\textit{Keywords}: climate change, \emph{Epichloë}, herbarium, INLA, museum specimen, plant-microbe symbiosis, \emph{Poaceae}, spatially-varying coefficients model
+	\textit{Keywords}: climate change, \emph{Epichloë}, herbarium, INLA, museum specimen, plant-microbe symbiosis, Poaceae, spatially-varying coefficients model
 
@@
 Changes in the variability of precipitation and temperature during off-peak seasons were also important predictors of increasing endophyte prevalence. 
-Our analysis performed favorably in an out-of-sample predictive test with contemporary survey data \firstrevise{from across 63 populations}, a rare extra step in collections-based research.
-We identified greater local-scale variability in endophyte prevalence in contemporary data compared to model predictions based on historic data, suggesting new directions that could improve predictive accuracy.
+Our model performed favorably in an out-of-sample predictive test with contemporary survey data \firstrevise{from across 63 populations}, a rare extra step in collections-based research.
+However, there was greater local-scale variability in endophyte prevalence in contemporary data compared to model predictions, suggesting new directions that could improve predictive accuracy.
 Our results provide novel evidence for a cryptic biological response to climate change that may contribute to the resilience of host-microbe symbiosis through \firstrevise{fitness benefits to symbiotic hosts}.
@@
 
-\linelabel{R4C10-begin}Natural history specimens, which were originally collected to study and preserve taxonomic diversity, present a unique opportunity to explore long-term changes in \revise{biodiversity} and ecological interactions across broad spatial and temporal scales\revise{ \citep{meineke2018unrealized, davis2023herbarium}.}
+\linelabel{R4C10-begin}Natural history specimens, which were originally collected to document and preserve taxonomic diversity, present a unique opportunity to explore long-term changes in \revise{biodiversity} and ecological interactions across broad spatial and temporal scales\revise{ \citep{meineke2018unrealized, davis2023herbarium}.}
 Natural history collections, built and maintained by the efforts of thousands of scientists, are invaluable time machines, primarily comprised of physical specimens of organisms along with information about the time and place of their collection.
@@
 \linelabel{R2C19-begin}
-For example, previous researchers have \firstrevise{examined the flowers, pollen grains, and leaves of specimens within plant collections (herbaria) collected across time} to document shifts in \firstrevise{reproductive} phenology \citep{willis2017old, park2019herbarium,  berg2019examination}, pollination \citep{pauw2011reconstruction, duan2019century}, and herbivory \citep{meineke2019herbarium} related to anthropogenic climate change.\linelabel{R2C19-end}
+For example, previous researchers have \firstrevise{examined the flowers, pollen grains, and leaves of specimens within plant collections (herbaria)} to document shifts in \firstrevise{reproductive} phenology \citep{willis2017old, park2019herbarium,  berg2019examination}, pollination \citep{pauw2011reconstruction, duan2019century}, and herbivory \citep{meineke2019herbarium} related to anthropogenic climate change.\linelabel{R2C19-end}
 However, few previous studies have leveraged biological collections to examine climate change-related shifts in a particularly common type of interaction: microbial symbiosis.
@@
 Microbial symbionts are common to all macroscopic organisms and can have important effects on their hosts' survival, growth and reproduction \cite{rodriguez2009fungal,mcfall2013animals}.
-Many microbial symbionts act as mutualists, engaging in reciprocally beneficial interactions with their hosts that can ameliorate environmental stress. 
+Many microbial symbionts act as mutualists, engaging in reciprocally beneficial interactions with their hosts in ways that can ameliorate environmental stress. 
 For example, bacterial symbionts of insects, such as \emph{Wolbachia}, can improve their hosts' thermal tolerance \citep{truitt2019wolbachia, renoz2019evolutionary}, and arbuscular mycorrhizal fungi, documented in 70-90\% of families of land plants \citep{parniske2008arbuscular}, allow their hosts to persist through drought conditions by improving water and nutrient uptake \citep{cheng2021elucidating}.
@@
 
-Understanding of how microbial symbioses are affected by climate change is additionally complicated by spatial heterogeneity in the direction and magnitude of environmental change \cite{ipcc_2021}. 
+Understanding how microbial symbioses are affected by climate change is additionally complicated by spatial heterogeneity in the direction and magnitude of environmental change \cite{ipcc_2021}. 
 Beneficial symbionts are likely able to shield their hosts from environmental stress in locations that experience a small degree of change, but symbionts in locations that experience changes of large magnitude may be pushed beyond their physiological limits \cite{webster2008temperature}.
 Additionally, symbionts are often unevenly distributed across their host's distribution.
-Facultative symbionts may be absent from portions of the host range \cite{afkhami2014mutualist}, and hosts may engage with a diversity of partners (different symbiont species or locally-adapted strains) across their environments \cite{frade2008variation, rolshausen2018quantifying, fowler2023geographic}.
+Facultative symbionts may be absent from portions of the host range \cite{afkhami2014mutualist}, and hosts may engage with a diversity of partners (different symbiont species or locally-adapted strains) across environments \cite{frade2008variation, rolshausen2018quantifying, fowler2023geographic}.
 Identifying broader spatial trends in symbiont prevalence is therefore an important step in developing predictions for where to expect changes in the symbiosis in future climates.
 
-\emph{Epichloë} fungal endophytes are specialized symbionts of cool-season grasses, which have been documented in $\sim 30$\% of cool-season grass species \citep{leuchtmann1992systematics}.
+\emph{Epichloë} fungal endophytes are specialized symbionts of cool-season grasses, documented in $\sim 30$\% of cool-season grass species \citep{leuchtmann1992systematics}.
 They are \firstrevise{predominantly} transmitted vertically from maternal plants to offspring through seeds.
@@
 Previous survey studies of contemporary populations have documented large-scale spatial patterns in endophyte prevalence structured by environmental gradients \citep{granath2007variation,bazely2007broad, afkhami2012fungal,sneck2017variation}.
-We predicted that prevalence should track temporal changes in environmental drivers \linelabel{R2C21-begin}\firstrevise{(i.e. drought)} \linelabel{R2C21-end} that elicit strong fitness benefits.
+We predicted that endophyte prevalence should track temporal changes in environmental drivers \linelabel{R2C21-begin}\firstrevise{(i.e.
+drought)} \linelabel{R2C21-end} that elicit strong fitness benefits.
 %For example, endophyte-mediated drought tolerance should lead endophyte prevalence to increase in regions where precipitation has declined over time.
 
-Early research on \emph{Epichloë} used herbarium specimens to describe the broad taxonomic diversity of host species that harbor these symbionts \citep{white1985endophyte}, establishing that endophyte symbiosis could be identified in plant tissue from as early as 1851.
+Early research on \emph{Epichloë} used herbarium specimens to describe the broad taxonomic diversity of grass host species that harbor these symbionts \citep{white1985endophyte}, establishing that endophyte symbiosis could be identified in plant tissue from as early as 1851.
 However, no subsequent studies, to our knowledge, have used the vast resources of biological collections to quantitatively assess spatio-temporal trends in endophyte prevalence and their environmental correlates.
-\linelabel{R4C11a-begin}\revise{Previous work has used herbarium specimens to identify the origins and population genomics of  plant diseases such as \emph{Phytophora}, the Irish potato famine pathogen \citep{ristaino2001pcr,ristaino2002tracking,yoshida2013rise}, and have been proposed as vehicles to track other emerging plant pathogens \citep{ristaino2020importance,bradshaw2021global}.}\linelabel{R4C11a-end}
-Grasses are commonly collected and identified based on the presence of their reproductive structures, meaning that preserved specimens typically contain seeds, conveniently preserving the fungi along with their host plants on herbarium sheets.
+\tom{\linelabel{R4C11a-begin}\revise{Previous work has used herbarium specimens to identify the origins and population genomics of  plant diseases such as \emph{Phytophora}, the Irish potato famine pathogen \citep{ristaino2001pcr,ristaino2002tracking,yoshida2013rise}, and have been proposed as vehicles to track other emerging plant pathogens \citep{ristaino2020importance,bradshaw2021global}.}\linelabel{R4C11a-end}}{IMO this does not fit well here, because at this point in the Intro we are zoomed in on Epichloe.
+I would move this sentence either to the `Natural history specimens' paragraph or the following one, where you introduce microbial symbioses.}
+Grasses are commonly collected and identified based on the presence of their reproductive structures, meaning that preserved specimens typically contain seeds, conveniently preserving the seed-transmitted fungi along with their host plants on herbarium sheets.
 This creates the opportunity to leverage the unique spatio-temporal sampling of herbarium collections to examine the response of this symbiosis to historical climate change. 

### 2025-07-15 Manuscript/endo_herbarium_R2.tex
methods edits

@@
 Our surveys focused on three native North American grasses: \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus} that host \emph{Epichloë} symbionts. 
-\linelabel{R2C23-begin}\firstrevise{These cool-season grass species are commonly represented in natural history collections with broad distributions covering much the eastern United States (Fig.
-\ref{fig:map}).
+\linelabel{R2C23-begin}\firstrevise{These cool-season grass species have broad distributions covering much the eastern United States (Fig.
+\ref{fig:map}) and are commonly represented in natural history collections.
 Cool-season grasses grow during the cooler temperatures of spring and autumn due to their reliance on $C_3$ photosynthesis.}\linelabel{R2C23-end}
-\emph{A. hyemalis} is a small short-lived perennial species that germinates in \firstrevise{late winter} winter and typically flowers between March and July (most common collection month: May).
+\emph{A.
+hyemalis} is a small short-lived perennial species that germinates in autumn to \firstrevise{late winter} winter and typically flowers between March and July (most common collection month: May).
 \emph{A. perennans} is of similar stature but is longer lived than \emph{Agrostis hyemalis} and flowers in late summer and early autumn (most common collection month: September). 
@@
 hyemalis} is found rarely as far north as Canada and is listed as a rare plant in Minnesota.
-\emph{E. virginicus} is a larger and relatively longer-lived  species that is more broadly distributed than the \emph{Agrostis} species. 
+\emph{E. virginicus} is a larger and longer-lived  species that is more broadly distributed than the \emph{Agrostis} species. 
 It begins flowering as early as March or April but continues throughout the summer (most common collection month: July).
@@
 \linelabel{R1C3-begin}
-\firstrevise{Both \emph{Agrostis} species host \emph{Epichloë amarillans} \cite{craven2001multigene, leuchtmann2014nomenclatural}, while \emph{Elymus virginicus} typically hosts \emph{Epichloë elymi} \cite{clay2002evolutionary}.
-The fungal symbionts primarily reproduce asexually and are passed from mother to offspring by vertical transmission through seeds. 
-These traits contribute to highly specialized interactions between symbiont and host.
+\firstrevise{Both \emph{Agrostis} species host \emph{Epichloë amarillans} \cite{craven2001multigene, leuchtmann2014nomenclatural}, and \emph{Elymus virginicus} typically hosts \emph{Epichloë elymi} \cite{clay2002evolutionary}.
+The fungal symbionts primarily reproduce asexually and are passed from maternal plant to offspring by vertical transmission through seeds. 
+\tom{These traits contribute to highly specialized interactions between symbiont and host.}{Is this sentence necessary?}
 Some host species have been shown to partner with multiple symbiont species or strains, and in some cases multiple symbiont lineages can co-exist within a host population \cite{mc2014species}.
@@
 Among those specimens geo-referenced at the state level, the largest bounding box, spanning the state of Texas, was 1233 km wide.
-The smallest bounding boxes were less than 1 km across for small municipalities (while this suggests high precision, we note that some specimens were collected in natural habitat nearby to small municipalities not encompassed by this bounding boxes).}\linelabel{R2C24-end}
+The smallest bounding boxes were less than 1 km across for small municipalities (while this suggests high precision, we note that some specimens were collected in natural habitat nearby to small municipalities not encompassed by these bounding boxes).}\linelabel{R2C24-end}
 
 \linelabel{R1C11-begin}
-\firstrevise{Our visits focused on herbaria with historic strengths in \emph{Poaceae} collections (e.g.
+\firstrevise{Our visits focused on herbaria with historic strengths in grass collections (e.g.
 Texas A\&M, Missouri Botanic Garden) and other herbaria in the Southern Great Plains region of the United States.
@@
 For each host species $h$, endophyte presence/absence of the $i^{th}$ specimen ($P_{h,i}$) was modeled as a Bernoulli response variable with expected probability of endophyte occurrence $\hat{P}_{h,i}$.
-We modeled $\hat{P}_{h,i}$ as a linear function of intercept $A_{h}$ and slope $T_{h}$ defining the global trend in endophyte prevalence specific to each host species as well as with spatially-varying intercepts $\alpha_{h,l_{i}} $ and slopes $\tau_{h,l_i}$ associated with location ($l_i$, the unique latitude-longitude combination of the $i$th observation).
-\linelabel{R2C28-begin}The joint-model structure allowed us to ``borrow \firstrevise{information}'' \linelabel{R2C28-end} across species in the estimation of shared variance terms for the spatially-dependent random effect $\delta_{l_i}$, intended to account for residual spatial variation, and $\chi_{c_i}$ and $\omega_{s_i}$, the i.i.d.-random effects indexed for each collector identity ($c_i$) and scorer identity ($s_i$) of the $i^{th}$ specimen.
+We modeled $\hat{P}_{h,i}$ as a linear function of collection year, with intercept $A_{h}$ and slope $T_{h}$ defining the global temporal trend in endophyte prevalence specific to each host species as well as with spatially-varying intercepts $\alpha_{h,l_{i}} $ and slopes $\tau_{h,l_i}$ associated with location ($l_i$, the unique latitude-longitude combination of the $i$th observation).
+\linelabel{R2C28-begin}The joint-model structure allowed us to ``borrow \firstrevise{information}'' \linelabel{R2C28-end} across species in the estimation of shared variance terms for the spatially-dependent random effect $\delta_{l_i}$, intended to account for residual spatial variation, and $\chi_{c_i}$ and $\omega_{s_i}$, the i.i.d.-random effects indexed for each \tom{collector identity ($c_i$) and scorer identity ($s_i$) of the $i^{th}$ specimen.}{In the data collection methods I would add that collector was gathered as part of the sample metadata and scorer was recorded as part of the data collection process.
+This will help prepare readers for the statistical methods here.}
 
@@
 \ref{fig:meshplot}).
-\linelabel{R3C2a-begin}\revise{Mat\'{e}rn covariance functions are widely used in spatial-statistical modeling because of their mathematical tractability and flexibility.
+\linelabel{R3C2a-begin}\revise{Mat\'{e}rn covariance functions are widely used in spatially explicit statistical modeling because of their mathematical tractability and flexibility.
 This covariance structure relies on the assumption that the underlying process is stationary and isotropic, such that spatial autocorrelation between data points depends only on their relative positions \citep{bakka2018spatial}.}\linelabel{R3C2a-end}
@@
 }\linelabel{R3C1-end}
-Priors for the Mat\'{e}rn covariance function, termed ''range" and "variance", define how proximity effects decay with distance.
-\linelabel{R3C2b-begin}\revise{The choice of priors for these types of spatial models is an area of active research \cite{simpson2017penalising,bakka2018spatial}, but another advantage of the INLA approach is that its computational efficiency allows for sensitivity analyses.}\linelabel{R3C2b-end}
+Priors for the Mat\'{e}rn covariance function, termed ``range'' and ``variance'', define how proximity effects decay with distance.
+\linelabel{R3C2b-begin}\revise{The choice of priors for these types of spatial models is an area of active research \cite{simpson2017penalising,bakka2018spatial}, but another advantage of the INLA approach is that its computational efficiency allows for prior sensitivity analyses.}\linelabel{R3C2b-end}
 Results presented in the main text reflect a prior range of $342$ kilometers (i.e. a $50$\% probability of estimating a range less than $342$ kilometers). 
@@
 the same direction of effects across space.
-We assessed model fit with visual posterior predictive checks (\ref{fig:overlayplot}) and measurements of AUC (Figs.
+We assessed model fit with visual posterior predictive checks (Fig.
+\ref{fig:overlayplot}) and measurements of AUC (Figs.
 \ref{fig:ROCtraining}-\ref{fig:ROCtest}) \linelabel{R2C48-begin} \firstrevise{\cite{gelman2006data}}\linelabel{R2C48-end}.
@@
 hyemalis} \cite{https://doi.org/10.15468/dl.ryqqcv}, 656 occurrence records for \emph{A.
-perennans} \cite{https://doi.org/10.15468/dl.wtne24}, and 2338 occurrence records for \emph{A.
+perennans} \cite{https://doi.org/10.15468/dl.wtne24}, and 2338 occurrence records for \emph{E.
 virginicus} \cite{https://doi.org/10.15468/dl.yt76uh}}).
@@
 %\tom{Among this suite of variables, we chose to include, which were uncorrelated (Variance Inflation Factor $>$ 0.7) and allowed us to predict the occurrence probability of each host species in space and time.}{Do not follow and sounds circular.}
-We fit maximum entropy (MaxEnt) models using the maxent function in the R package dismo \citep{hijmans2017package} using following seasonal climate predictors (1990-2020 climate normals): mean and standard deviation of spring, summer, and autumn temperature, and mean and standard deviation of spring, summer, and autumn cumulative precipitation.
+We fit maximum entropy (MaxEnt) models using the maxent function in the R package dismo \citep{hijmans2017package} using the following seasonal climate predictors (1990-2020 climate normals): mean and standard deviation of \tom{spring, summer, and autumn}{Did we choose the dates by which these seasons are defined?
+Worth reporting?} temperature, and mean and standard deviation of spring, summer, and autumn cumulative precipitation.
 %\josh{MaxEnt is preferred because it has been shown to generate response curves with less unpredictable behavior when applied to new climates \citep{hijmans2006ability}.}{possibly could remove this sentence?}
@@
 
-We first downloaded monthly temperature and precipitation rasters from the PRISM climate group \citep{daly2013prism} covering the time period between 1895 and 2020 using the 'prism' R package \citep{Rprism2015}.
+We first downloaded monthly temperature and precipitation rasters from the PRISM climate group \citep{daly2013prism} covering the time period between 1895 and 2020 using the \tom{'prism'}{Other package names are not 'quoted'.} R package \citep{Rprism2015}.
 Prism provides reconstructions of historic climate variables across the United States by spatially interpolating weather station data \citep{diLuzio2008constructing}. 
-\linelabel{R3C3-begin}\revise{Because the magnitude of observed climate change differs across seasons, and because different growing seasons is a key feature of the biology of our focal host species, } we calculated 30-year climate normals for seasonal mean temperature and cumulative precipitation for the recent (1990 to 2020) and historic (1895 to 1925) periods.
+\linelabel{R3C3-begin}\revise{Because the magnitude of observed climate change differs across seasons, and because different growing seasons is a key feature of the biology of our focal host species,} we calculated 30-year climate normals for seasonal mean temperature and cumulative precipitation for the recent (1990 to 2020) and historic (1895 to 1925) periods.
 We used three four-month seasons within the year (Spring: January, February, March, April; Summer: May; June, July, August; Autumn: September, October, November, December). 
-This division of seasons allowed us to quantify differences in \revise{the primary climate change drivers, temperature and preciptiation} associated with the two ``cool'' seasons, when we expected our focal species to be most active (\emph{A.
+This division of seasons allowed us to quantify differences in \revise{the primary climate change drivers, temperature and precipitation,} associated with the two ``cool'' seasons, when we expected our focal species to be most active (\emph{A.
 hyemalis} flowering phenology: spring; \emph{E.
@@
 
-\linelabel{R2C32-begin}\firstrevise{We then evaluated whether areas that have experienced the greatest changes in endophyte prevalence (hotspots of endophyte change) are associated with high degrees of change in climate (hotspots of climate change)
+\linelabel{R2C32-begin}\firstrevise{We then evaluated whether areas that have experienced the greatest changes in endophyte prevalence (hotspots of endophyte change) are associated with high degrees of change in climate (hotspots of climate change).
 To do so, we modeled the fitted, spatially-varying slopes of endophyte change through time ($\tau_{[h]l}$) as a linear function of environmental covariates, with a Gaussian error distribution for a set of pixels across each host distribution.
@@
 hyemalis}, and \emph{E.
-virginicus} respectively ( pixel dimensions: \emph{A.
+virginicus} respectively (pixel dimensions: \emph{A.
 perennans} =  65 km x 36 km; \emph{A.

### 2025-07-15 Manuscript/endo_herbarium_R2.tex
results and discussion edits

@@
 virginicus} trends did not differ according to historic prevalence (\ref{fig:initialprev_trend_plot} C).} \linelabel{R1C7-end}
-Posterior estimates of uncertainty in spatially varying slopes indicate that these hotspots of change may have experienced increases of up to $5$\% per year while declines in prevalence may be as great as $-4$\% per year for \emph{A.
+\tom{Posterior estimates of uncertainty in spatially varying slopes indicate that these hotspots of change may have experienced increases of up to $5$\% per year while declines in prevalence may be as great as $-4$\% per year for \emph{A.
 hyemalis} and \emph{A.
 perennans}.
-For \emph{E. virginicus}, uncertainty ranges between $3.5$\% increases and $2.5$\% decreases (Fig. \ref{fig:svc_time_map_CI}).
+For \emph{E.
+virginicus}, uncertainty ranges between $3.5$\% increases and $2.5$\% decreases (Fig.
+\ref{fig:svc_time_map_CI}).}{I wonder if this information would be better placed alongside or integrated with the mean estimates described above (``In some regions...'').}
 \begin{figure}[H]
@@
 Within the region, climate changes were spatially variable; certain locations experienced increases in annual precipitation as large as $375$ mm or decreases up to $54$ mm across the last century, while annual temperature changes ranged from warming as great as $1.4$ $^{\circ}$C to cooling by $0.46$ $^{\circ}$C.
-Spatially variable climate trends were predictive of trends in endophyte prevalence.
+\tom{Spatially}{Could start new paragraph here and combine it with ``Changes in endophyte'' paragraph.
+I also think we might want to hedge a little more in this section to acknowledge that while there are climate signals, there are likely other drivers at play.
+I know this comes up in the Discussion but I think that can be briefly foreshadowed here.} variable climate trends were predictive of trends in endophyte prevalence.
 For example, strong increases in endophyte prevalence for \emph{A.
@@
 \subsection*{Evaluation of model performance on an out-of-sample test}
-Tests of the endophyte prevalence model's predictive performance as quantified by AUC and by visual posterior predictive checks, indicated good predictive ability. 
+Tests of the endophyte prevalence model's predictive performance, as quantified by AUC and by visual posterior predictive checks, indicated good predictive ability. 
 Model performance was similar between historic herbarium specimens used as training data and the out-of-sample test data from contemporary surveys (AUC = 0.79 and 0.77 respectively; Fig.
@@
 This interpretation is in line with theory predicting that positive fitness feedback caused by vertical transmission leads beneficial symbionts to rise in prevalence within a population \cite{fine1975vectors,donald2021context}.
-We further found that trends in endophyte prevalence varied across the distribution of each species in association with changes in climate drivers, suggesting that the increases in endophyte prevalence are driven by context-dependent benefits to hosts that confer resilience under environmental change.
-Taken together, this suggests an overall strengthening of host-symbiont mutualism over the last two centuries.
+We further found that trends in endophyte prevalence often varied across the host distribution in association with changes in climate drivers, consistent with the hypothesis that increases in endophyte prevalence are driven by context-dependent benefits to hosts that confer resilience under environmental change.
+Taken together, our results suggest an overall strengthening of host-symbiont mutualism over the last two centuries.
 
@@
 \firstrevise{Broad changes in the distribution and abundance of natural enemies \cite{cote2004ecological}, along with stresses from anthropogenic changes in \linelabel{R3C7-begin}\revise{land management} and pollution \cite{sage2020global} likely influence the benefits of symbiosis \cite{rudgers2020climate}.
-\revise{Host species experience a world that is made increasingly stressful by the combination of these global change drivers, and while historic trends that we observed suggest that symbiotic fitness benefits have helped mitigate this stress, it is likely that at some higher leveler of stress, increasing costs of maintaining the mutualism could lead to declines in endophyte prevalence.}\linelabel{R3C7-end}
 \revise{Changing endophyte prevalence} results from the combination of \revise{net} fitness benefits playing out across the heterogeneous map of a changing climate and \linelabel{R3C8-begin}\revise{and its interactive effects on}other anthropogenic drivers.\linelabel{R2C44-end}\linelabel{R3C8-end}
+\revise{Host species experience a world that is made increasingly stressful by a combination of global change drivers, and while historic trends that we observed suggest that symbiotic fitness benefits have helped mitigate this stress, it is possible that at yet higher levels of stress, increasing costs of the mutualism could lead to declines in endophyte prevalence.
+Identifying `tipping points' of mutualism breakdown under increasing environmental stress is an important area of future inquiry.}\linelabel{R3C7-end}
 
@@
 \linelabel{R1C11-b-begin}However, symbiont-facilitated range shifts require that endophytes be present in the populations to be able to \revise{contribute to} population growth. 
-The arid western range edge of \emph{A.
+For example, the arid western range edge of \emph{A.
 hyemalis} has had historically low endophyte prevalence (Fig.
-\ref{fig:svc_space_year_map}), and while prevalence has increased most quickly in the regions with historically low endophyte prevalence (Fig.
+\ref{fig:svc_space_year_map}), and while prevalence has increased most quickly in regions with historically low endophyte prevalence (Fig.
 \ref{fig:initialprev_trend_plot}), the complete absence of endophytes at range edges in this species would make it impossible for prevalence to increase without dispersal of symbiotic seeds \cite{fowler2023geographic}.\linelabel{R1C11-b-end}
 These factors potentially contribute to the ability of the host species to track its environmental niche. 
-Another interesting question is the degree to which symbiotic and non-symbiotic hosts, which occupy overlapping but distinct niches, are likely to experience distribution shifts in tandem or at different rates in future.\linelabel{R2C42-end}
+Another interesting question is the degree to which symbiotic and non-symbiotic hosts, which occupy overlapping but distinct niches, are likely to experience distribution shifts in tandem or at different rates in the future.\linelabel{R2C42-end}
 More extreme climate stresses, which are expected more frequently in the future \cite{seneviratne2021weather}, have the potential to alter the costs and benefits of the interaction. 
-The past indicates a resilient symbiosis, but it will be crucial to understand whether more extreme future climate conditions could tip this interaction to deteriorate.}\linelabel{R2C14-end}
+\tom{The past indicates a resilient symbiosis, but it will be crucial to understand whether more extreme future climate conditions could tip this interaction to deteriorate.}{Interesting point but maybe redundant with what I added above about `tipping points', so maybe consolidate these ideas in one place.}}\linelabel{R2C14-end}
 
 \firstrevise{\subsubsection*{Steps towards forecasts of host-microbe symbioses}}
-The combination of a spatially-explicit model and historic herbarium specimens allowed us to identify regions of both increasing and decreasing endophyte prevalence, however we see several next steps toward the goal of predicting host and symbiont niche-shifts in response to future climate change.
-While the model recreated the large-scale spatial trends observed in contemporary population surveys, test data contained more population-to-population variability in prevalence than could be explained by the model.
+The combination of a spatially-explicit model and historic herbarium specimens allowed us to identify regions of both increasing and decreasing endophyte prevalence.
+We see several next steps toward the goal of predicting host and symbiont niche-shifts in response to future climate change.
+While the model successfully predicted large-scale spatial trends observed in the out-of-sample contemporary population surveys, these data contained more population-to-population variability in prevalence than could be explained by the model.
 \linelabel{R2C39-begin}\firstrevise{We interpret this to mean that the model captures coarse-scale spatial and temporal trends reasonably well, but is not equipped to capture local-scale nuances that generate population-to-population differences.} \linelabel{R2C39-end}
@@
 \linelabel{R1C9-end}
-\linelabel{R3C6-begin}\revise{An alternative validation test would be to hold-out samples from within the historic data set.
-Such a test would more clearly provide a set of data that matched the conditions of the training data (i.e in spatial scale and climate conditions) missing within the contemporary test data, however a relatively limited number of sampled specimens held us back from exploring this option.
+\linelabel{R3C6-begin}\revise{An alternative validation test would be to hold-out samples from the historic data set.
+Such a test would more clearly match the conditions of the training data (i.e., in spatial scale and climate conditions), however the trade-off between training and testing the model with a limited number of sampled specimens held us back from exploring this option.
 Splitting datasets can negatively impact model estimates, and the choice of how to split the data for model validation is not trivial \citep{bergmeir2012use,james2013introduction}. 
-\linelabel{R4C10b-begin}From a modeler's perspective, this also urges advocacy for increased collection efforts and expansion of herbarium collections.}\linelabel{R3C6-end}\linelabel{R4C10b-end}
+\linelabel{R4C10b-begin}\tom{From a modeler's perspective, this also urges advocacy for increased collection efforts and expansion of herbarium collections.}{I don't really understand what this is trying to say.
+Not much information in this sentence.}}\linelabel{R3C6-end}\linelabel{R4C10b-end}
 
@@
 Poor predictive ability at local scales in this grass-endophyte system is not surprising, as previous studies have found that local variation, even to the scale of hundreds of meters can structure endophyte-host niches \cite{kazenel2015mutualistic,gundel2024temporal}.
-An important step would be integrating data from local and regional scales  through modeling to constrain estimates of local and regional variation.
+An important next step would be integrating data from local and regional scales  through modeling to constrain estimates of local and regional variation.
 
@@
 \ref{fig:svc_space_year_map}A) is a result of environmental conditions that lead the symbiosis to have negative fitness consequences, or is a result of some historical contingency or dispersal limitation that has thus far limited the presence of symbiotic hosts from a region where they would otherwise flourish and provide resilience.
-\linelabel{R3C5-begin}\revise{Another interesting question is whether local adaption, in either host or symbiont may influence future range shifts.}\linelabel{R3C5-end}
+\linelabel{R3C5-begin}\revise{\tom{Another interesting question is whether local adaption, in either host or symbiont, may influence future range shifts.}{This dangles here a little awkwardly, and I don't think it came up in the reviews...?
+Maybe cut?}}\linelabel{R3C5-end}
 Incorporating available climatic and soil layers as covariates is another obvious step that could improve predictions.
@@
 Beyond spatial autocorrelation of outcomes, systematic differences in sampling across space can result in spatial bias.
+
 One strength of herbaria as vehicles for global change research is the relative ease with which specimens from many distinct geographic locations can be examined. 
@@
 Beyond host-microbe symbioses, detecting ecological responses to anthropogenic global change and attributing their causes would inform public policy decision-makers and adaptive management strategies.
-\linelabel{R4C10c-begin}\revise{Natural history specimens, such as the plant hosts examined in this study have a clear role to play in informing global change biodiversity science, including building understanding of the dynamics of host-symbiont interactions \cite{davis2023herbarium}.}\linelabel{R4C10c-end}\
+\linelabel{R4C10c-begin}\revise{Natural history specimens, such as the plant hosts examined in this study, have a clear role to play in informing global change biodiversity science, including building understanding of the dynamics of host-symbiont interactions \cite{davis2023herbarium}.}\linelabel{R4C10c-end}\
 This survey of historic endophyte prevalence is necessarily correlative, yet it serves as a foundation to develop better predictive models of the response of microbial symbioses to climate change. 
-By comparing detected ecological responses with alternative mechanistic simulations of the past, we could attribute their cause, in a manner similar to methods from climate science and economics \citep{stott2010detection, carleton2016social,trenberth2015attribution}.
+By comparing detected ecological responses with \tom{alternative mechanistic simulations of the past}{I don't really know what this means or refers to.}, we could attribute their cause, in a manner similar to methods from climate science and economics \citep{stott2010detection, carleton2016social,trenberth2015attribution}.
 Combining the insights from this type of regional-scale survey with field experiments and physiological performance data could be invaluable to identify mechanisms driving shifts in host-symbiont dynamics.
@@
 	This research was supported by funding from National Science Foundation (grants 1754468 and 2208857) and by funding from the Texas Ecolab Program.
-\firstrevise{Two anonymous reviewers greatly improved an earlier version of this manuscript.}
+\firstrevise{\tom{Two}{Three? Four?} anonymous reviewers greatly improved an earlier version of this manuscript.}
 

### 2025-10-01 Manuscript/Reviewer_response_endo_herbarium_R2.tex
adding response to reviewers round 3

@@
 	\response{We have added clarification about the assumptions of the Matérn covariance function, and about our choice of priors within the Methods section.
-In addition, we have added material to our Discussion that highlights the important inferences \tom{provided by our sensitivity analysis}{Is this sensitivity analysis new?
-If not, emphasize that we already had a sensitivity analysis in the previous submission.
-If yes, emphasize that we added a new sensitivity analysis AND added discussion material about it.}.}
+In addition, we have added material to our Discussion that highlights the important inferences provided by the existing sensitivity analysis we present in our supplemental materials.}
 	\linesref{ms-R3C2a-begin}{ms-R3C2a-end}
 	\linesref{ms-R3C2b-begin}{ms-R3C2b-end}
+\linesref{ms-R3C2c-begin}{ms-R3C3c-end}
 
@@
 We have edited our methods section to more clearly explain the rationale behind our analysis of climate drivers.
-Specifically, we expected there to be differences across seasons in the magnitude of climate change in temperature and precipitation, because these seasons connect to the biology and natural history of our focal species.
+Specifically, we expected there to be differences across seasons in the magnitude of climate change in temperature and precipitation because these seasons connect to the biology and natural history of our focal species.
 In addition, we have added to our discussion to acknowledge the potential importance of local scale ecological processes that may influence our analysis.
 We agree that the potential mismatch in scales is a limitation to herbarium studies in general, compounded not just by the scale of the climate data, but also by the spatial scale of georeferencing used to connect herbarium specimens to localities.
-The INLA modeling approach we employ cannot overcome inherent scale-limitations in the data, \tom{however treating space in a continuous way does allow us to make predictions at arbitrary spatial scales.
-Our climate analysis used predicted trends from the model to assess relationships with climate, meaning that the climate data were actually on a finer scale than the endophyte trend predictions.
-However this is, as Reviewer 3 notes, a broader scale than many potential ecological processes involved.}{I find this a little hard to follow (and I have not read the changes to the ms yet).
-Can we simplify this response and just say it's a good point and we have added discussion to acknowledge the implications of the scale mismatch?
-That's all the reviewer was asking for.}}
+We have added discussion material that acknowledges the implications of this scale mismatch.}
 \linesref{ms-R3C3-begin}{ms-R3C3-end}
+\linesref{ms-R3C3b-begin}{ms-R3C3b-end}
 
@@
 \response{This is a great point.
-In response to this comment, we have revised our Discussion section to include further discussion of the potential implications of the out-of-sample validation test, \tom{particularly highlighting that the data do differ in scale and potentially climate conditions, and how this might inform future modeling efforts.
-Reviewer 3 points to potential non-stationarity as an explanation for mismatch in model validation.
-Previous investigations have found that stationarity is often a reasonable assumption to make and that there are not meaningul differences in out-of-sample validation tests that use blocks within and without time series (Bergmeir and Benitez, 2012, \emph{Information Sciences}).}{I don't really get the scale and stationarity ideas from the reviewer's comment, so this seems a little tangential to me.
-I think they were simply observing that the census years were at the very end of the herbarium years, so the model predictions are probably extra-uncertain.
-To me, nonstationarity is already built into the model via the year effect.
-}}
+In response to this comment, we have revised our Discussion section to include further discussion of the potential implications of the out-of-sample validation test, particularly highlighting that the data do differ in scale and potentially climate conditions, and how this might inform future modeling efforts.
+We discuss the main potential alternative validation test -- leaving out test data from within the historic dataset.
+In addition, we now cite research that evaluates the importance of alternative choices of test-data time-period across datasets.}
 \linesref{ms-R3C6-begin}{ms-R3C6-end}
@@
 We agree with the reviewer that causal mechanisms underlying these changes in prevalence should be interpreted with caution.
-In fact, \tom{we have recently submitted an additional manuscript}{Some journals ask to see any related manuscripts that are under review, so check GCB guidelines.} that explore the potential role of other anthropogenic global change drivers, including land use and nitrogen deposition.
+In fact, we have recently submitted an additional manuscript that explores the potential role of other anthropogenic global change drivers, including land use and nitrogen deposition.
 While our previous submission already touched on these ideas in the Discussion, in response to this comment, we have revised our Discussion to more clearly communicate the potential for other factors to contribute to these trends.}

### 2025-10-01 Manuscript/Reviewer_response_endo_herbarium_R3.tex
adding response to reviewers round 3

% ======================================================= %
% Document: TEMPLATE FOR RESPONSES TO REVIEWERS
% Author: Andrea Ballatore
% Date: Jan 7, 2013
% Source: https://raw.githubusercontent.com/ucd-spatial/Datasets/master/tex_response_to_reviewers_template/responses_to_reviewers.tex
% Modified by Eduard Szöcs, 10.03.2015
% ======================================================= %
\documentclass[12pt]{article}

% packages

\usepackage{graphicx}
\usepackage{url}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{color}
\definecolor{mygray}{gray}{0.6}
\usepackage[utf8]{inputenc}
\usepackage[onehalfspacing]{setspace}
\usepackage[
	round,	%(defaultage in the main file and \input ) for round parentheses;
	colon,	% (default) to separate multiple citations with colons;
	authoryear,% (default) for author-year citations;
	sort,		% orders multiple citations into the sequence in which they
]{natbib}
\usepackage[%disable
	]{todonotes}

\usepackage{anysize}
\marginsize{2.5cm}{2.5cm}{1.5cm}{2.5cm}

\usepackage{xr}
\externaldocument[ms-]{endo_herbarium_R3}

% macros
% add a counter
\newcounter{cN}
\setcounter{cN}{0}

\newcommand{\comment}[1]{
	\vspace{2em}
	\refstepcounter{cN} % incrment counter
	\noindent \hangindent=0em \textbf{\textcolor{Maroon}{\uline{Comment \thecN}:~}}\emph{``#1''}
	}

\newcommand{\response}[1]{
	\\[0.25em]
	\hangindent=2.3em \textbf{\textcolor{NavyBlue}{\uline{Response}:~}}#1
	}

\newcommand{\revise}[1]{{\color{Mahogany}{#1}}}

\newcommand{\linesref}[2]{
		\\[0.25em]
	\hangindent=2.3em {\color{Mahogany} Line(s): \ref{#1} - \ref{#2}}
}
\newcommand{\jacob}[2]{{\color{blue}{#1}}\footnote{\textit{\color{blue}{#2}}}}
%\IfValueT{#2}{- \ref{#2}}
\newcommand{\tom}[2]{{\color{red}{#1}}\footnote{\textit{\color{red}{#2}}}}
%\IfValueT{#2}{- \ref{#2}}

\usepackage[normalem]{ulem}
\definecolor{darkred}{rgb}{1,.6,.6}
\DeclareRobustCommand\problemline{\bgroup\markoverwith{\textcolor{darkred}{\rule[-0.9ex]{4pt}{3pt}}}\ULon}
\DeclareRobustCommand{\problem}[1]{\problemline{#1}} % soul
\setcounter{secnumdepth}{-1}

\begin{document}
% ======================================================= %
\title{Manuscript GCB-25-0529.R1 --- Response to reviewers}

\maketitle
% ======================================================= %
\noindent To the editorial board,

Following the positive feedback from reviewers on our last submission, we have revised our manuscript to clarify the description of the biology of \emph{Epichloë} symbiosis, including description of the biochemical mechanisms underlying the interactions and the taxonomic diversity. 
We have also added additional discussion of the potential for horizontal and imperfect transmission to influence host responses to climate change.

We appreciate the opportunity improve our work with these revisions, and we believe that our manuscript is now suitable for publication in \emph{Global Change Biology}.
We detail these changes in the point-by-point responses below.
All of our changes are denoted in the revised manuscript with \revise{Mahogany font} and referenced below with line numbers.

\vspace{2em}
Sincerely,

Joshua Fowler

Jacob Moutouama

Tom Miller

\newpage

% ======================================================= %
\section{Response to Reviewer 3}
\vspace{-2em}

%\comment{}
%\response{}
%\linesref{ms-R1C3-begin}{ms-R1C3-end}

\comment{The authors did an excellent job; all my concerns have been addressed.}
\response{We are glad to have addressed the concerns of Reviewer 3 through our previous revision, whose Review helped us to more clearly present the underlying assumptions in our modeling approach.}

\section{Response to Reviewer 5}
\vspace{-2em}

\setcounter{cN}{0}

\comment{I appreciate the opportunity to review your paper, which I think is a very important contribution linking a little known but important and emblematic symbiotic system to global climate change. I find your paper very interesting and very well written and organized. I have only minor suggestions.}
\response{We appreciate this positive evaluation of our work, and we are glad that Reviewer 5 sees its potential importance to the broader audience of global climate change researchers.}

\comment{I seem to have been invited in as an additional reviewer. I'm no expert on the statistics or modeling, so I assume I'm invited to comment on the biology and wet-lab techniques. I think the technique is fine for the analysis, and innovative in that it utilizes a rich archival resource.}
\response{Reviewer 5's suggestions to improve our communication of the biology of our study system were especially valuable, and we appreciate that they acknowledge the unique and innovative data presented.}

\comment{You chose to analyze seeds, which I think is the best choice. You take a cautious approach with both conservative and liberal judgments made for each sample (which concurred in 89\% of cases), and recognize that imperfect transmission can lead to mixtures of infected and uninfected seeds in the same sample. Your lab technique reflects good understanding of the biology of the system, as does most of the relevant text. So, I just have some subtle points to raise that, if taken into account would (I think) improve the manuscript a bit.}
\response{We are glad that Reviewer 5 agrees with our choice to analyze seed tissue from the historic herbarium specimens. We are also glad that they recognize our efforts to quantify potential uncertainty in our results through analysis of both ``conservative'' and ``liberal'' scores.}

\comment{Results are discussed with a very host-centric perspective. In particular, I didn't find consideration of the effect of stress on transmissibility. A relevant paper on the subject is: Gundel PE, Garibaldi LA, Martínez-Ghersa MA, Ghersa CM (2011) Neotyphodium endophyte transmission to Lolium multiflorum seeds depends on the host plant fitness. Environmental and Experimental Botany 71: 359-366. doi https://doi.org/10.1016/j.envexpbot.2011.02.002 (Note to the editor, "Neotyphodium" is a now disfavored name for asexual species of Epichloë.) It might be interesting to look into whether the data on proportions of positive seeds in positive samples suggest stress-induced reductions in vertical transmission, though that would probably be something for future research. But I think that the potential for such an effect should receive some discussion here.}
\response{We agree that evaluating potential effects of stress on imperfect transmission would be very interesting! We feel, based on previous work from our lab group and others (Sneck et al., {Microbial Ecology}, 2017; Cavazos et al., \emph{New Phytologist}, 2018; Gagic et al., \emph{Frontiers in Plant Science}, 2018), that doing so properly would require evaluating larger numbers of seeds per individual to quantify with greater precision the rate of imperfect transmission. Therefore, we have opted to incorporate this material within the discussion, suggesting future targeted sampling that could be done to get these data from historic specimens.}
\linesref{ms-R5C5-begin}{ms-R5C5-end}

\comment{I also didn't see mention of the potential for horizontal transmission. Some \emph{Epichloë} species on some host "choke" the hosts, meaning that they produce their fruiting structures (stromata) and consequently can spread via airborne ascospores. The three associations investigated here (two \emph{Agrostis} spp. probably with \emph{Epichloe amarillans} and \emph{Elymus virginicus} probably with \emph{Epichloe elymi}) all exhibit this capability even though seed transmission does appear to be the more common route for them.}
\response{Reviewer 5 is correct that there is a possibility of horizontal transmission in general for \emph{Epichloë}, and indeed for these host species. In response to this comment, we have mentioned the potential for horizontal transmission in our description of the study system, and where relevant within the the discussion section.}
\linesref{ms-R5C3-begin}{ms-R5C3-end}
\linesref{ms-R5C5-discussion-begin}{ms-R5C5-discussion-end}

\comment{Lines 534–536 seems to imply that transmission is impossible outside of seed transmission, which isn't the case. I haven't checked the rationale for that conclusion in the Fowler et al. 2023 paper cited there, but the phrasing here seems to me to be a bit misleading. Maybe "impossible" is an overly strong word.}
\response{Lines 534-536 discuss the potential that an isolated, range-edge population starting with no symbionts will experience an increase in symbiosis. Our wording does indeed suggest that this could happen primarily through the introduction of symbiotic seeds via dispersal into the population, without mentioning the other possiblity pointed out by Reviewer 5 -- that the symbiont could disperse without its host as a horizontally-transmitted propagule. We have edited the wording of this sentence to acknowledge this possibility, and to highlight the point we intended to make, which is that dispersal of symbionts must occur before such an isolated population might experience a signal of increasing endophyte prevalence.}
\linesref{ms-R5C7-begin}{ms-R5C7-end}

\comment{Lines 94–95: This is a dubious statement. Considering that the grass subfamily Pooideae is the most speciose of the subfamilies in an extremely speciose plant family, I don't think adequate sampling has ever been done to determine the proportion of species that have these endophytes. Perhaps you mean 30\% of surveyed species at that time? Unfortunately, 1992 is a long time ago, and there has been a lot of surveying done since then, together with improved techniques, but I don't have a recent reference that springs to mind to update the count. Perhaps this sentence can simply be modified to the effect that in 1992 Leuchtmann estimated that ca. 30 \% of surveyed species had Epichloe?}
\response{In response to this comment, we have edited the wording of this sentence to acknowledge that these numbers are an estimate based on a subset of surveyed species, and updated with additional references to more recent surveys (Card et al., \emph{FEMS microbiology ecology}, 2014; Iannone et al., \emph{Mycological Progress}, 2011), that find estimates similar to that by Leuchtmann. We agree a more comprehensive compilation of these surveys would be valuable.}
\linesref{ms-R5C7-end-dubious}{ms-R5C7-end-dubious}

\comment{Finally, line 102 mentions "other secondary metabolites", which I take to be a grab bag for unknown anti-herbivore factors. Yet, there is one Epichloe protein known (MCF = "makes caterpillars floppy"), and possibly others, that have anti-herbivore effects. Proteins aren't "secondary metabolites". Maybe a better word here would be "factors"?}
\response{We thank Reviewer 5 for pointing us towards this literature. The biochemical nature of \emph{Epichloë} symbioses is not our particular area of expertise, so we appreciate this feedback. In response to this comment, we have edited this sentence and others in our discussion section, to acknowledge the potentially diverse biologically active factores involved, and have added references to research uncovering the role of these chemicals.}
\linesref{ms-R5-C9a-begin}{ms-R5-C9a-end}
\linesref{ms-R5C8-discussion-begin}{ms-R5C8-discussion-end}

% ======================================================= %
\end{document}
% ======================================================= %

### 2025-10-01 Manuscript/endo_herbarium_R2.tex
stashing some old stuff for last revision

@@
 \newcommand{\firstrevise}[1]{{\color{black}{#1}}}
-\newcommand{\revise}[1]{{\color{Mahogany}{#1}}}
+\newcommand{\revise}[1]{{\color{black}{#1}}}
 
@@
 For example, previous researchers have \firstrevise{examined the flowers, pollen grains, and leaves of specimens within plant collections (herbaria)} to document shifts in \firstrevise{reproductive} phenology \citep{willis2017old, park2019herbarium,  berg2019examination}, pollination \citep{pauw2011reconstruction, duan2019century}, and herbivory \citep{meineke2019herbarium} related to anthropogenic climate change.\linelabel{R2C19-end}
-However, few previous studies have leveraged biological collections to examine climate change-related shifts in a particularly common type of interaction: microbial symbiosis.
+\linelabel{R4C11a-begin}\revise{Herbarium specimens have also been used to identify the origins and population genomics of  plant diseases such as \emph{Phytophora}, the Irish potato famine pathogen \citep{ristaino2001pcr,ristaino2002tracking,yoshida2013rise}, and have been proposed as vehicles to track other emerging plant pathogens \citep{ristaino2020importance,bradshaw2021global}.}\linelabel{R4C11a-end}
+However, few previous studies have leveraged biological collections to examine climate change-related shifts in a particularly common type of interaction: \revise{mutualistic} microbial symbiosis.
 
@@
 However, no subsequent studies, to our knowledge, have used the vast resources of biological collections to quantitatively assess spatio-temporal trends in endophyte prevalence and their environmental correlates.
-\tom{\linelabel{R4C11a-begin}\revise{Previous work has used herbarium specimens to identify the origins and population genomics of  plant diseases such as \emph{Phytophora}, the Irish potato famine pathogen \citep{ristaino2001pcr,ristaino2002tracking,yoshida2013rise}, and have been proposed as vehicles to track other emerging plant pathogens \citep{ristaino2020importance,bradshaw2021global}.}\linelabel{R4C11a-end}}{IMO this does not fit well here, because at this point in the Intro we are zoomed in on Epichloe.
-I would move this sentence either to the `Natural history specimens' paragraph or the following one, where you introduce microbial symbioses.}
 Grasses are commonly collected and identified based on the presence of their reproductive structures, meaning that preserved specimens typically contain seeds, conveniently preserving the seed-transmitted fungi along with their host plants on herbarium sheets.
@@
 The fungal symbionts primarily reproduce asexually and are passed from maternal plant to offspring by vertical transmission through seeds. 
-\tom{These traits contribute to highly specialized interactions between symbiont and host.}{Is this sentence necessary?}
 Some host species have been shown to partner with multiple symbiont species or strains, and in some cases multiple symbiont lineages can co-exist within a host population \cite{mc2014species}.
@@
 %Our sampling plan was designed to minimize damage to these specimens.
-We chose our focal species in part because they are commonly represented in herbarium collections and produce many seeds, meaning that small samples would not diminish the value of the specimens for future studies.
+We chose our focal species in part because they are commonly represented in herbarium collections and produce many seeds, meaning that small samples would not diminish the value of specimens for future studies.
 We collected 5-10 seeds per specimen after examining the herbarium sheet under a dissecting microscope to ensure that we collected mature seeds, not florets or unfilled seeds, fit for our purpose of identifying fungal endophytes with microscopy.
@@
 
-Each specimen was assigned geographic coordinates based on collection information recorded on the herbarium sheet using the geocoding functionality of the ggmap R package \citep{kahle2019package}.
+Each specimen was assigned geographic coordinates based on collection information recorded on the herbarium sheet using the geocoding functionality of the \emph{ggmap} R package \citep{kahle2019package}.
 Many specimens had digitized collection information readily available, but for those that did not, we transcribed information printed on the herbarium sheet. 
+\revise{The identity of each specimen collector was gathered as part of the sample's metadata.}
 Collections were geo-referenced to the nearest county centroid, or nearest municipality when that information was available. 
@@
 When we identified potential endophytes with unusual morphology, low uptake of stain, or a small amount of fungal hyphae across the scored seeds, we recorded \firstrevise{ a positive identification} for the liberal score and a negative \firstrevise{identification for} the conservative score.
+\revise{We recorded the identity of ech scorer as part of the data collection process.}
 $89$\% of scored plants had matching liberal and conservative scores, reflecting high confidence in endophyte status.
@@
 We modeled $\hat{P}_{h,i}$ as a linear function of collection year, with intercept $A_{h}$ and slope $T_{h}$ defining the global temporal trend in endophyte prevalence specific to each host species as well as with spatially-varying intercepts $\alpha_{h,l_{i}} $ and slopes $\tau_{h,l_i}$ associated with location ($l_i$, the unique latitude-longitude combination of the $i$th observation).
-\linelabel{R2C28-begin}The joint-model structure allowed us to ``borrow \firstrevise{information}'' \linelabel{R2C28-end} across species in the estimation of shared variance terms for the spatially-dependent random effect $\delta_{l_i}$, intended to account for residual spatial variation, and $\chi_{c_i}$ and $\omega_{s_i}$, the i.i.d.-random effects indexed for each \tom{collector identity ($c_i$) and scorer identity ($s_i$) of the $i^{th}$ specimen.}{In the data collection methods I would add that collector was gathered as part of the sample metadata and scorer was recorded as part of the data collection process.
-This will help prepare readers for the statistical methods here.}
+\linelabel{R2C28-begin}The joint-model structure allowed us to ``borrow \firstrevise{information}'' \linelabel{R2C28-end} across species in the estimation of shared variance terms for the spatially-dependent random effect $\delta_{l_i}$, intended to account for residual spatial variation, and $\chi_{c_i}$ and $\omega_{s_i}$, the i.i.d.-random effects indexed for each collector identity ($c_i$) and scorer identity ($s_i$) of the $i^{th}$ specimen.
 
@@
 
-We performed model fitting using the inlabru R package \citep{bachl2019inlabru}.
+We performed model fitting using the \emph{inlabru} R package \citep{bachl2019inlabru}.
 Global intercept and slope parameters, $A$ and $T$, were given vague priors.
@@
 Results presented in the main text reflect a prior range of $342$ kilometers (i.e. a $50$\% probability of estimating a range less than $342$ kilometers). 
-We tested a range of values (from $68$ kilometers to $1714$ kilometers) and meshes (presented in the Supporting Methods), finding that while the magnitude and uncertainty of effects varied, model results were qualitatively similar, i.e.
+We tested a range of values (from $68$ kilometers to $1714$ kilometers) and meshes (presented in the Supporting Methods \revise{-- \emph{Mesh and Prior Sensitivity Analysis}}), finding that while the magnitude and uncertainty of effects varied, model results were qualitatively similar, i.e.
 the same direction of effects across space.
@@
 perennans} \cite{https://doi.org/10.15468/dl.wtne24}, and 2338 occurrence records for \emph{E.
-virginicus} \cite{https://doi.org/10.15468/dl.yt76uh}}).
+virginicus} \cite{https://doi.org/10.15468/dl.yt76uh})}.
 %To reduce the potential influence of sampling bias and spatial autocorrelation, \tom{we thinned the occurrences to the spatial scale of our selected climatic predictors}{I do not follow.}. 
@@
 %\tom{Among this suite of variables, we chose to include, which were uncorrelated (Variance Inflation Factor $>$ 0.7) and allowed us to predict the occurrence probability of each host species in space and time.}{Do not follow and sounds circular.}
-We fit maximum entropy (MaxEnt) models using the maxent function in the R package dismo \citep{hijmans2017package} using the following seasonal climate predictors (1990-2020 climate normals): mean and standard deviation of \tom{spring, summer, and autumn}{Did we choose the dates by which these seasons are defined?
-Worth reporting?} temperature, and mean and standard deviation of spring, summer, and autumn cumulative precipitation.
+We fit maximum entropy (MaxEnt) models using the maxent function in the R package \emph{dismo} \citep{hijmans2017package} using the following seasonal climate predictors (1990-2020 climate normals): mean and standard deviation of spring, summer, and autumn temperature, and mean and standard deviation of spring, summer, and autumn cumulative precipitation.
 %\josh{MaxEnt is preferred because it has been shown to generate response curves with less unpredictable behavior when applied to new climates \citep{hijmans2006ability}.}{possibly could remove this sentence?}
+
 We generated 10,000 pseudo-absences as background points, and split the occurrence data into 75\% for model training and 25\% for model testing.
@@
 
-We first downloaded monthly temperature and precipitation rasters from the PRISM climate group \citep{daly2013prism} covering the time period between 1895 and 2020 using the \tom{'prism'}{Other package names are not 'quoted'.} R package \citep{Rprism2015}.
-Prism provides reconstructions of historic climate variables across the United States by spatially interpolating weather station data \citep{diLuzio2008constructing}. 
+We first downloaded monthly temperature and precipitation rasters from the PRISM climate group \citep{daly2013prism} covering the time period between 1895 and 2020 using the \emph{prism} R package \citep{Rprism2015}.
+PRISM provides reconstructions of historic climate variables across the United States by spatially interpolating weather station data \citep{diLuzio2008constructing}. 
 \linelabel{R3C3-begin}\revise{Because the magnitude of observed climate change differs across seasons, and because different growing seasons is a key feature of the biology of our focal host species,} we calculated 30-year climate normals for seasonal mean temperature and cumulative precipitation for the recent (1990 to 2020) and historic (1895 to 1925) periods.
@@
 \linelabel{R2C32-begin}\firstrevise{We then evaluated whether areas that have experienced the greatest changes in endophyte prevalence (hotspots of endophyte change) are associated with high degrees of change in climate (hotspots of climate change).
-To do so, we modeled the fitted, spatially-varying slopes of endophyte change through time ($\tau_{[h]l}$) as a linear function of environmental covariates, with a Gaussian error distribution for a set of pixels across each host distribution.
+To do so, we modeled the fitted, spatially-varying slopes of endophyte change through time ($\tau_{h,l}$) as a linear function of environmental covariates, with a Gaussian error distribution for a set of pixels across each host distribution.
 The continuous SPDE approach taken from our endophyte prevalence model allows us to generate predictions of temporal trends in prevalence at arbitrarily many pixels across each host distribution.
@@
 virginicus} across the south-central US, with emphasis on Texas and neighboring states (Fig \ref{fig:contempsurveysmap}).
-During surveys, we collected seeds from up to 30 individuals per population (average number of plants sampled per population: $22.9$); note that this sampling design provided greater local depth of information than the herbarium records, where only one plant was sampled at each locality.
+number of plants sampled per population: $22.9$); note that this sampling design provided greater local depth of information than the herbarium records, where only one plant was sampled at each locality.
 We quantified the endophyte status of each individual with microscopy as described for the herbarium surveys (with 5-10 seeds scored per individual), and calculated the prevalence of endophytes within the population (proportion of plants that were endophyte-symbiotic).
@@
 hyemalis} and \emph{A.
-perennans} experienced increases in prevalence by as much as $2$\% per year over the study period, while  \emph{E.
-virginicus} experienced increases up to around $1$\% per year.
-Both \emph{Agrostis} species show areas of strong increase and areas of declining prevalence, while \emph{E.
+perennans} experienced increases in prevalence by as much as $2$\% per year over the study period.
+\revise{Posterior estimates of uncertainty in spatially varying slopes indicate that these hotspots of change may have experienced increases of up to $5$\% per year while declines in prevalence may be as great as $-4$\% per year for the \emph{Agrostis} species.
+(Fig.
+\ref{fig:svc_time_map_CI})}
+\revise{In contrast}, \emph{E.
+virginicus} experienced increases up to around $1$\% per year, \revise{with uncertainty ranging between $3.5$\% increases and $2.5$\% decreases (Fig.
+\ref{fig:svc_time_map_CI})}
+\revise{Taken together, both} \emph{Agrostis} species show areas of both strong increasing and declining prevalence, while \emph{E.
 virginicus} had an overall weaker and geographically more homogeneous increase in endophyte prevalence.
@@
 virginicus} trends did not differ according to historic prevalence (\ref{fig:initialprev_trend_plot} C).} \linelabel{R1C7-end}
-\tom{Posterior estimates of uncertainty in spatially varying slopes indicate that these hotspots of change may have experienced increases of up to $5$\% per year while declines in prevalence may be as great as $-4$\% per year for \emph{A.
-hyemalis} and \emph{A.
-perennans}.
-For \emph{E.
-virginicus}, uncertainty ranges between $3.5$\% increases and $2.5$\% decreases (Fig.
-\ref{fig:svc_time_map_CI}).}{I wonder if this information would be better placed alongside or integrated with the mean estimates described above (``In some regions...'').}
 \begin{figure}[H]
@@
 	\includegraphics[width = \linewidth]{../Plots/svc_time_map.png}
-	\caption[Predicted posterior mean of spatially-varying slopes representing change in endophyte prevalence for each host species.]{Predicted posterior mean of spatially-varying slopes representing change in endophyte prevalence for each host species (A,\emph{A.
-hyemalis}; B,\emph{A.
-perennans}; C,\emph{E.
+	\caption[Predicted posterior mean of spatially-varying slopes representing change in endophyte prevalence for each host species.]{Predicted posterior mean of spatially-varying slopes representing change in endophyte prevalence for each host species ((A)\emph{A.
+hyemalis}; (B)\emph{A.
+perennans}; (C)\emph{E.
 virginicus}).
@@
 	\caption[Predicted endophyte prevalence for each host species in 1895 and 2020.]{\firstrevise{Predicted endophyte prevalence for each host species in 1895 and 2020.
-\firstrevise{Predictions of prevalence come from the endophyte prevalence model.} Color indicates the posterior mean endophyte prevalence for \emph{A.
-hyemalis} (A,D), \emph{A.
-perennans} (B,E), and \emph{E.
-virginicus} (C,F).
+\firstrevise{Predictions of prevalence come from the endophyte prevalence model.} Color indicates the posterior mean endophyte prevalence for (A, D) \emph{A.
+hyemalis}, (B, E) \emph{A.
+perennans}, and (C, F) \emph{E.
+virginicus}.
 Map lines delineate study areas and do not necessarily depict accepted national boundaries.}}
@@
 We found that trends in endophyte prevalence were strongly associated with one or more seasonal climate change drivers (Fig. \ref{fig:climate_correlation_plot}).
-For the majority of the study region, the climate has become wetter (an average increase in annual precipitation of $60$ mm) with relatively little temperature warming (an average increase in annual temperature of $0.02$ $^{\circ}$C) over the last century (Fig.
+For the majority of the study region, the climate has become wetter (an average increase in annual precipitation of $60$ mm) with relatively minimal temperature warming (an average increase in annual temperature of $0.02$ $^{\circ}$C) over the last century (Fig.
 \ref{fig:AGHY_climate_covariates}-\ref{fig:ELVI_climate_covariates}), a consequence of regional variation in global climate change \cite{ipcc_2021}.
 Within the region, climate changes were spatially variable; certain locations experienced increases in annual precipitation as large as $375$ mm or decreases up to $54$ mm across the last century, while annual temperature changes ranged from warming as great as $1.4$ $^{\circ}$C to cooling by $0.46$ $^{\circ}$C.
-\tom{Spatially}{Could start new paragraph here and combine it with ``Changes in endophyte'' paragraph.
-I also think we might want to hedge a little more in this section to acknowledge that while there are climate signals, there are likely other drivers at play.
-I know this comes up in the Discussion but I think that can be briefly foreshadowed here.} variable climate trends were predictive of trends in endophyte prevalence.
-For example, strong increases in endophyte prevalence for \emph{A.
+
+Spatially variable climate trends were predictive of trends in endophyte prevalence.
+For example, \revise{among the tested climate drivers}, strong increases in endophyte prevalence for \emph{A.
 perennans} were most strongly associated with increasing autumn precipitation and with increasing mean and variability in autumn temperature (greater than 97\% posterior probabilities of positive slopes).
@@
 \ref{fig:climate_correlation_plot}B).
-This result aligns with the species' autumn active growing season, however other seasonal climate drivers were also associated with increasing endophyte prevalence in this host species. 
+This result aligns with the species' autumn active growing season, however other seasonal climate drivers were also \revise{positively} associated with increasing endophyte prevalence in this host species.
 In particular, we found cooler and drier springs and cooler summers to be associated with increasing endophyte prevalence (greater than $99$\% posterior probabilities of negative slopes), though these slopes were generally of smaller magnitude than those for autumn climate drivers.
-
 Changes in endophyte prevalence across the ranges of \emph{A. hyemalis} and \emph{E. virginicus} were less strongly driven by changes in climate. 
 Like \emph{A. perennans}, climate during peak growing season (spring for \emph{A. perennans} and summer for \emph{E. virginicus}) emerged most commonly as drivers of changes in endophyte prevalence.
-Increases in mean spring precipitation were the strongest predictor of increasing trends in endophyte prevalence for \emph{A.
+\revise{Across the tested climate drivers}, increases in mean spring precipitation were the strongest predictor of increasing trends in endophyte prevalence for \emph{A.
 hyemalis} (Fig.
@@
 
-\firstrevise{While we show consistent increasing trends in prevalence between the three species, the mechanisms that explain these changes may be diverse and idiosyncratic.
+\firstrevise{While we show consistent increasing trends in prevalence between the three species, the mechanisms that explain these changes may be diverse and idiosyncratic.}
 \linelabel{R2C45-begin}First, climate change responses may depend on genotype-specific responses that are not considered in our current analysis. 
@@
 Genotypic variation of the hosts themselves can also influence interaction outcomes \cite{gundel2011interaction,parker2017genotype}.
-Whether hotspots of change in endophyte prevalence reflect selection for genotype-pairings with particularly strong fitness benefits is an unanswered question.}\linelabel{R2C45-end}
+Whether hotspots of change in endophyte prevalence reflect selection for genotype-pairings with particularly strong fitness benefits is an unanswered question.\linelabel{R2C45-end}
 \linelabel{R2C44-begin}Additionally, \emph{Epichloë} endophytes have been connected to a suite of non-drought related fitness benefits including herbivory \firstrevise{defense} \citep{brem2001epichloe}, salinity resistence \citep{wang2020effects}, and mediation of pathogens \cite{vikuk2019infection} and the soil microbiome \citep{roberts2015rhizosphere}.
-\firstrevise{Broad changes in the distribution and abundance of natural enemies \cite{cote2004ecological}, along with stresses from anthropogenic changes in \linelabel{R3C7-begin}\revise{land management} and pollution \cite{sage2020global} likely influence the benefits of symbiosis \cite{rudgers2020climate}.
-\revise{Changing endophyte prevalence} results from the combination of \revise{net} fitness benefits playing out across the heterogeneous map of a changing climate and \linelabel{R3C8-begin}\revise{and its interactive effects on}other anthropogenic drivers.\linelabel{R2C44-end}\linelabel{R3C8-end}
+\firstrevise{Broad changes in the distribution and abundance of natural enemies \cite{cote2004ecological}, along with stresses from anthropogenic changes in \linelabel{R3C7-begin}\revise{land management} and pollution \cite{sage2020global} likely influence the benefits of symbiosis \cite{rudgers2020climate}.}
+\revise{Changing endophyte prevalence} results from the combination of \revise{net} fitness benefits playing out across the heterogeneous map of a changing climate and \linelabel{R3C8-begin}\revise{and its interactive effects on} other anthropogenic drivers.\linelabel{R2C44-end}\linelabel{R3C8-end}
 \revise{Host species experience a world that is made increasingly stressful by a combination of global change drivers, and while historic trends that we observed suggest that symbiotic fitness benefits have helped mitigate this stress, it is possible that at yet higher levels of stress, increasing costs of the mutualism could lead to declines in endophyte prevalence.
+More extreme climate stresses, which are expected more frequently in the future \cite{seneviratne2021weather} could shift the balance of interactions costs and benefits.
 Identifying `tipping points' of mutualism breakdown under increasing environmental stress is an important area of future inquiry.}\linelabel{R3C7-end}
@@
 Another interesting question is the degree to which symbiotic and non-symbiotic hosts, which occupy overlapping but distinct niches, are likely to experience distribution shifts in tandem or at different rates in the future.\linelabel{R2C42-end}
-More extreme climate stresses, which are expected more frequently in the future \cite{seneviratne2021weather}, have the potential to alter the costs and benefits of the interaction. 
-\tom{The past indicates a resilient symbiosis, but it will be crucial to understand whether more extreme future climate conditions could tip this interaction to deteriorate.}{Interesting point but maybe redundant with what I added above about `tipping points', so maybe consolidate these ideas in one place.}}\linelabel{R2C14-end}
 
@@
 \linelabel{R1C9-end}
+\linelabel{R4C10b-begin} \revise{Herbarium collections were predominately used for taxonomic research in the past, but use of specimens to understand ongoing global change would benefit from increased collection efforts and expansion of herbarium collections.} \linelabel{R4C10b-end}
 \linelabel{R3C6-begin}\revise{An alternative validation test would be to hold-out samples from the historic data set.
 Such a test would more clearly match the conditions of the training data (i.e., in spatial scale and climate conditions), however the trade-off between training and testing the model with a limited number of sampled specimens held us back from exploring this option.
-Splitting datasets can negatively impact model estimates, and the choice of how to split the data for model validation is not trivial \citep{bergmeir2012use,james2013introduction}. 
-\linelabel{R4C10b-begin}\tom{From a modeler's perspective, this also urges advocacy for increased collection efforts and expansion of herbarium collections.}{I don't really understand what this is trying to say.
-Not much information in this sentence.}}\linelabel{R3C6-end}\linelabel{R4C10b-end}
-
-\linelabel{R3C3-being}\revise{Another key consideration in forecasting the dynamics host-microbe symbioses is the spatial scale of both specimen georeferencing and available climate data. 
+Splitting datasets can negatively impact model estimates, and the choice of how to split the data for model validation is not trivial \citep{bergmeir2012use,james2013introduction}.} \linelabel{R3C6-end}
+
+\linelabel{R3C3b-begin}\revise{Another key consideration in forecasting the dynamics of host-microbe symbioses is the spatial scale of both specimen georeferencing and available climate data. 
 For this analysis, most specimen localities were assigned coordinates at county or city centroids, and the climate data examined was on 4 km grid cells.
 Georeferencing of specimens as accurately as possible is a key priority of herbarium specimen digitization efforts \citep{soltis2017digitization,davis2023herbarium}.
-While the INLA modeling approach that we used allows for predictions at arbitrarility small spatial scales, and would simplify connecting model predictions to the scale of a given climate driver, the course scale inherent to our analysis may obscure some local-scale ecological processes} \linelabel{R3C3-end}
-Poor predictive ability at local scales in this grass-endophyte system is not surprising, as previous studies have found that local variation, even to the scale of hundreds of meters can structure endophyte-host niches \cite{kazenel2015mutualistic,gundel2024temporal}.
+While the INLA modeling approach that we used allows for predictions at arbitrarility small spatial scales, and would simplify connecting model predictions to the scale of a given climate driver, the course scale inherent to our analysis may obscure some local-scale ecological processes} \linelabel{R3C3b-end}
+Poor predictive ability at local scales in this grass-endophyte system is not surprising, as previous studies have found that local variation (e.g., in soil conditions, in microclimate), even to the scale of hundreds of meters can structure endophyte-host niches \cite{kazenel2015mutualistic,gundel2024temporal}.
+\linelabel{R3C5-begin}\revise{Local adaptation in either the host or symbiont to microclimate or soil conditions could cause populations to differ from broad regional trends.}\linelabel{R3C5-end}
+\linelabel{R3C2c-begin}\revise{The choice of prior distributions for spatially-varying random effects also impacts the model's flexibility to capture spatial trends. 
+Our exploration of model sensitivity to prior choice (presented in the \emph{Supplemental Methods}) reveals qualitatively similar results across a broad range of priors.}
+\linelabel{R3C3c-end}
 An important next step would be integrating data from local and regional scales  through modeling to constrain estimates of local and regional variation.
@@
 \ref{fig:svc_space_year_map}A) is a result of environmental conditions that lead the symbiosis to have negative fitness consequences, or is a result of some historical contingency or dispersal limitation that has thus far limited the presence of symbiotic hosts from a region where they would otherwise flourish and provide resilience.
-\linelabel{R3C5-begin}\revise{\tom{Another interesting question is whether local adaption, in either host or symbiont, may influence future range shifts.}{This dangles here a little awkwardly, and I don't think it came up in the reviews...?
-Maybe cut?}}\linelabel{R3C5-end}
 Incorporating available climatic and soil layers as covariates is another obvious step that could improve predictions.
 These steps will bridge gaps that often exist between large but broad bioclimatic and biodiversity data and small but high-resolution data on biotic interactions, and move towards the goal of predicting the dynamics of microbial symbioses under climate change \cite{miller2019recent, isaac2020data}.
+
+\newpage
 
@@
 \linelabel{R1C10-begin}
-First and foremost, this is \firstrevise{one of a growing number of studies to examine microbial symbiosis using specimens from natural history collections, and the first, to our knowledge, to link long-term changes in the symbioses to changes in climate.}\linelabel{R1C10-end}
+First and foremost, this is \firstrevise{one of a growing number of studies to examine microbial symbiosis using specimens from natural history collections, and the first, to our knowledge, to link long-term changes in symbioses to changes in climate.}\linelabel{R1C10-end}
 The responses of microbial symbioses are a rich target for future studies within historic specimens, particularly those that take advantage of advances in sequencing technology.
@@
 This survey of historic endophyte prevalence is necessarily correlative, yet it serves as a foundation to develop better predictive models of the response of microbial symbioses to climate change. 
-By comparing detected ecological responses with \tom{alternative mechanistic simulations of the past}{I don't really know what this means or refers to.}, we could attribute their cause, in a manner similar to methods from climate science and economics \citep{stott2010detection, carleton2016social,trenberth2015attribution}.
 Combining the insights from this type of regional-scale survey with field experiments and physiological performance data could be invaluable to identify mechanisms driving shifts in host-symbiont dynamics.
@@
 	This research was supported by funding from National Science Foundation (grants 1754468 and 2208857) and by funding from the Texas Ecolab Program.
-\firstrevise{\tom{Two}{Three? Four?} anonymous reviewers greatly improved an earlier version of this manuscript.}
+	\revise{Four} anonymous reviewers greatly improved earlier versions of this manuscript.
 
@@
 \textbf{Coordinate reference system}: WGS84 coordinate reference system (EPSG:4326 code)\\
-\textbf{Climatic data}:  raster data extracted from PRISM  \\
+\textbf{Climatic data}:  raster data extracted from PRISM;  30-year normal mean and standard deviation of temperature and of precipitation for three four-month seasons within the year (Spring: January, February, March, April; Summer: May; June, July, August; Autumn: September, October, November, December).
+\\
 {\color{blue}{Model }}\\
@@
 \textbf{Algorithms}: Maximum entropy (maxent)\\
-\textbf{Workflow}: We  described the workflow in the method section of the manuscript. \\
+\textbf{Workflow}: We  described the workflow in the methods section of the manuscript. \\
 \textbf{Software}:  All statistics were performed using Maxent 3.3.4 and R4.3.1 with packages terra, usdm, spThin and dismo.\\

### 2025-10-01 Manuscript/endo_herbarium_R3.tex
adding round 3 revisions

--- 2025-10-01 Manuscript/endo_herbarium_R2.tex
@@
 \newcommand{\firstrevise}[1]{{\color{black}{#1}}}
-\newcommand{\revise}[1]{{\color{black}{#1}}}
+\newcommand{\secondrevise}[1]{{\color{black}{#1}}}
+\newcommand{\revise}[1]{{\color{Mahogany}{#1}}}
 
@@
 
-\linelabel{R4C10-begin}Natural history specimens, which were originally collected to document and preserve taxonomic diversity, present a unique opportunity to explore long-term changes in \revise{biodiversity} and ecological interactions across broad spatial and temporal scales\revise{ \citep{meineke2018unrealized, davis2023herbarium}.}
+\linelabel{R4C10-begin}Natural history specimens, which were originally collected to document and preserve taxonomic diversity, present a unique opportunity to explore long-term changes in \secondrevise{biodiversity} and ecological interactions across broad spatial and temporal scales\secondrevise{ \citep{meineke2018unrealized, davis2023herbarium}.}
 Natural history collections, built and maintained by the efforts of thousands of scientists, are invaluable time machines, primarily comprised of physical specimens of organisms along with information about the time and place of their collection.
-These specimens often preserve physical legacies of ecological processes and species' interactions from dynamically changing environments across time and space \revise{\citep{lendemer2020extended}.}\linelabel{R4C10-end}
+These specimens often preserve physical legacies of ecological processes and species' interactions from dynamically changing environments across time and space \secondrevise{\citep{lendemer2020extended}.}\linelabel{R4C10-end}
 \linelabel{R2C19-begin}
 For example, previous researchers have \firstrevise{examined the flowers, pollen grains, and leaves of specimens within plant collections (herbaria)} to document shifts in \firstrevise{reproductive} phenology \citep{willis2017old, park2019herbarium,  berg2019examination}, pollination \citep{pauw2011reconstruction, duan2019century}, and herbivory \citep{meineke2019herbarium} related to anthropogenic climate change.\linelabel{R2C19-end}
-\linelabel{R4C11a-begin}\revise{Herbarium specimens have also been used to identify the origins and population genomics of  plant diseases such as \emph{Phytophora}, the Irish potato famine pathogen \citep{ristaino2001pcr,ristaino2002tracking,yoshida2013rise}, and have been proposed as vehicles to track other emerging plant pathogens \citep{ristaino2020importance,bradshaw2021global}.}\linelabel{R4C11a-end}
-However, few previous studies have leveraged biological collections to examine climate change-related shifts in a particularly common type of interaction: \revise{mutualistic} microbial symbiosis.
+\linelabel{R4C11a-begin}\secondrevise{Herbarium specimens have also been used to identify the origins and population genomics of  plant diseases such as \emph{Phytophora}, the Irish potato famine pathogen \citep{ristaino2001pcr,ristaino2002tracking,yoshida2013rise}, and have been proposed as vehicles to track other emerging plant pathogens \citep{ristaino2020importance,bradshaw2021global}.}\linelabel{R4C11a-end}
+However, few previous studies have leveraged biological collections to examine climate change-related shifts in a particularly common type of interaction: \secondrevise{mutualistic} microbial symbiosis.
 
@@
 
-\emph{Epichloë} fungal endophytes are specialized symbionts of cool-season grasses, documented in $\sim 30$\% of cool-season grass species \citep{leuchtmann1992systematics}.
+\emph{Epichloë} fungal endophytes are specialized symbionts of cool-season grasses, \linelabel{R5C7-begin-dubious}\revise{estimated from surveys to associate with $\sim 20 - 30$\%} of species \revise{in the diverse Poaceae family} \citep{leuchtmann1992systematics}.
+\revise{Surveys within the cool-season grass subfamily (Pooideae) have found that associations occur for between ~17\% to ~40\% of sampled species \cite{iannone2011diversity,card2014mutualistic}}.
+\linelabel{R5C7-end-dubious}
 They are \firstrevise{predominantly} transmitted vertically from maternal plants to offspring through seeds.
@@
 Over time, endophytes that act as mutualists should rise in prevalence within a host population, particularly under environmental conditions that elicit protective benefits \citep{donald2021context}. 
-\emph{Epichloë} are known to improve their hosts' drought tolerance \cite{decunta2021systematic} and protect their hosts against herbivores \cite{crawford2010fungal} and pathogens \cite{xia2018role} likely through the production of a diverse suite of alkaloids and other secondary metabolites.
+\emph{Epichloë} are known to improve their hosts' drought tolerance \cite{decunta2021systematic} and protect their hosts against herbivores \linelabel{R5C8-discussion-begin}\revise{\cite{crawford2010fungal,ambrose2014horizontal}} and pathogens \revise{\cite{xia2018role,tian2017epichloe}} likely through the production of a suite of  \linelabel{R5-C9a-begin}\revise{biologically active molecules, including diverse} alkaloids, \revise{proteins,} and other secondary metabolites.\linelabel{R5-C9a-end}\linelabel{R5C8-discussion-end}
 The fitness feedback induced by vertical transmission leads to the prediction that endophyte prevalence should be high in populations where these fitness benefits are most important. 
@@
 The fungal symbionts primarily reproduce asexually and are passed from maternal plant to offspring by vertical transmission through seeds. 
-Some host species have been shown to partner with multiple symbiont species or strains, and in some cases multiple symbiont lineages can co-exist within a host population \cite{mc2014species}.
+\linelabel{R5C3-begin}\revise{These symbionts are also capable of horizontal transmission between hosts via the production of external reproductive structures, including sexual spore-bearing stromata that grow over host inflorescences (known as 'choke disease') and epiphyllous conidia on leaf surfaces that produce asexual spores \cite{tadych2014epichloe}.
+Evidence suggests that production of horizontal transmission structures by \emph{Epichloë} occurs at low levels, and may be influenced by environmental and genotypic factors \cite{brem1999high,meijer2000effects,tintjer2008variation}.
+In line with this, monitoring of long-term plots of \emph{A.
+perennans} and \emph{E.
+virginicus} showed no production of stromata on \emph{A.
+perennans} and on only less than 1\% of \emph{E.
+virginicus} plants over 14 years \cite{fowler2024microbial}}.\linelabel{R5C3-end}
+Some host species have shown \revise{the capacity} to partner with multiple symbiont species or strains, and in some cases multiple symbiont lineages can co-exist within a host population \cite{mc2014species}.
 However, surveys have typically found limited \emph{Epichloë} genotypic diversity within host populations \cite{treindl2023genetic}. 
-Across host populations, concentrations of biologically-active alkaloids and the genes associated with their production vary substantially \cite{schardl2012chemotypic}.
+Across host populations, concentrations of biologically-active \revise{biomolecules} and the genes associated with their production vary substantially \cite{schardl2012chemotypic}.
 In this analysis, we focus on the presence/absence of \emph{Epichloë} symbionts, and we discuss potential implications of symbiont genotypic diversity in the Discussion.}
@@
 Many specimens had digitized collection information readily available, but for those that did not, we transcribed information printed on the herbarium sheet. 
-\revise{The identity of each specimen collector was gathered as part of the sample's metadata.}
+\secondrevise{The identity of each specimen collector was gathered as part of the sample's metadata.}
 Collections were geo-referenced to the nearest county centroid, or nearest municipality when that information was available. 
@@
 When we identified potential endophytes with unusual morphology, low uptake of stain, or a small amount of fungal hyphae across the scored seeds, we recorded \firstrevise{ a positive identification} for the liberal score and a negative \firstrevise{identification for} the conservative score.
-\revise{We recorded the identity of ech scorer as part of the data collection process.}
+\secondrevise{We recorded the identity of each scorer as part of the data collection process.}
 $89$\% of scored plants had matching liberal and conservative scores, reflecting high confidence in endophyte status.
@@
 \ref{fig:meshplot}).
-\linelabel{R3C2a-begin}\revise{Mat\'{e}rn covariance functions are widely used in spatially explicit statistical modeling because of their mathematical tractability and flexibility.
+\linelabel{R3C2a-begin}\secondrevise{Mat\'{e}rn covariance functions are widely used in spatially explicit statistical modeling because of their mathematical tractability and flexibility.
 This covariance structure relies on the assumption that the underlying process is stationary and isotropic, such that spatial autocorrelation between data points depends only on their relative positions \citep{bakka2018spatial}.}\linelabel{R3C2a-end}
 
-\linelabel{R3C1-begin}\revise{Implementing spatially-structured parameters in INLA with this SPDE approach is useful particularly because space is treated as a continuous variable, allowing the model to make efficient use of the data and generate predictions across the entire study region.
+\linelabel{R3C1-begin}\secondrevise{Implementing spatially-structured parameters in INLA with this SPDE approach is useful particularly because space is treated as a continuous variable, allowing the model to make efficient use of the data and generate predictions across the entire study region.
 The SPDE approach is flexible enough that it can capture smooth trends across space that are informed by the data rather than by spatial regions chosen \emph{a priori} by researchers.
@@
 Priors for the Mat\'{e}rn covariance function, termed ``range'' and ``variance'', define how proximity effects decay with distance.
-\linelabel{R3C2b-begin}\revise{The choice of priors for these types of spatial models is an area of active research \cite{simpson2017penalising,bakka2018spatial}, but another advantage of the INLA approach is that its computational efficiency allows for prior sensitivity analyses.}\linelabel{R3C2b-end}
+\linelabel{R3C2b-begin}\secondrevise{The choice of priors for these types of spatial models is an area of active research \cite{simpson2017penalising,bakka2018spatial}, but another advantage of the INLA approach is that its computational efficiency allows for prior sensitivity analyses.}\linelabel{R3C2b-end}
 Results presented in the main text reflect a prior range of $342$ kilometers (i.e. a $50$\% probability of estimating a range less than $342$ kilometers). 
-We tested a range of values (from $68$ kilometers to $1714$ kilometers) and meshes (presented in the Supporting Methods \revise{-- \emph{Mesh and Prior Sensitivity Analysis}}), finding that while the magnitude and uncertainty of effects varied, model results were qualitatively similar, i.e.
+We tested a range of values (from $68$ kilometers to $1714$ kilometers) and meshes (presented in the Supporting Methods \secondrevise{-- \emph{Mesh and Prior Sensitivity Analysis}}), finding that while the magnitude and uncertainty of effects varied, model results were qualitatively similar, i.e.
 the same direction of effects across space.
@@
 PRISM provides reconstructions of historic climate variables across the United States by spatially interpolating weather station data \citep{diLuzio2008constructing}. 
-\linelabel{R3C3-begin}\revise{Because the magnitude of observed climate change differs across seasons, and because different growing seasons is a key feature of the biology of our focal host species,} we calculated 30-year climate normals for seasonal mean temperature and cumulative precipitation for the recent (1990 to 2020) and historic (1895 to 1925) periods.
+\linelabel{R3C3-begin}\secondrevise{Because the magnitude of observed climate change differs across seasons, and because different growing seasons is a key feature of the biology of our focal host species,} we calculated 30-year climate normals for seasonal mean temperature and cumulative precipitation for the recent (1990 to 2020) and historic (1895 to 1925) periods.
 We used three four-month seasons within the year (Spring: January, February, March, April; Summer: May; June, July, August; Autumn: September, October, November, December). 
-This division of seasons allowed us to quantify differences in \revise{the primary climate change drivers, temperature and precipitation,} associated with the two ``cool'' seasons, when we expected our focal species to be most active (\emph{A.
+This division of seasons allowed us to quantify differences in \secondrevise{the primary climate change drivers, temperature and precipitation,} associated with the two ``cool'' seasons, when we expected our focal species to be most active (\emph{A.
 hyemalis} flowering phenology: spring; \emph{E.
@@
 perennans} experienced increases in prevalence by as much as $2$\% per year over the study period.
-\revise{Posterior estimates of uncertainty in spatially varying slopes indicate that these hotspots of change may have experienced increases of up to $5$\% per year while declines in prevalence may be as great as $-4$\% per year for the \emph{Agrostis} species.
+\secondrevise{Posterior estimates of uncertainty in spatially varying slopes indicate that these hotspots of change may have experienced increases of up to $5$\% per year while declines in prevalence may be as great as $-4$\% per year for the \emph{Agrostis} species.
 (Fig.
 \ref{fig:svc_time_map_CI})}
-\revise{In contrast}, \emph{E.
-virginicus} experienced increases up to around $1$\% per year, \revise{with uncertainty ranging between $3.5$\% increases and $2.5$\% decreases (Fig.
+\secondrevise{In contrast}, \emph{E.
+virginicus} experienced increases up to around $1$\% per year, \secondrevise{with uncertainty ranging between $3.5$\% increases and $2.5$\% decreases (Fig.
 \ref{fig:svc_time_map_CI})}
-\revise{Taken together, both} \emph{Agrostis} species show areas of both strong increasing and declining prevalence, while \emph{E.
+\secondrevise{Taken together, both} \emph{Agrostis} species show areas of both strong increasing and declining prevalence, while \emph{E.
 virginicus} had an overall weaker and geographically more homogeneous increase in endophyte prevalence.
@@
 Spatially variable climate trends were predictive of trends in endophyte prevalence.
-For example, \revise{among the tested climate drivers}, strong increases in endophyte prevalence for \emph{A.
+For example, \secondrevise{among the tested climate drivers}, strong increases in endophyte prevalence for \emph{A.
 perennans} were most strongly associated with increasing autumn precipitation and with increasing mean and variability in autumn temperature (greater than 97\% posterior probabilities of positive slopes).
@@
 \ref{fig:climate_correlation_plot}B).
-This result aligns with the species' autumn active growing season, however other seasonal climate drivers were also \revise{positively} associated with increasing endophyte prevalence in this host species.
+This result aligns with the species' autumn active growing season, however other seasonal climate drivers were also \secondrevise{positively} associated with increasing endophyte prevalence in this host species.
 In particular, we found cooler and drier springs and cooler summers to be associated with increasing endophyte prevalence (greater than $99$\% posterior probabilities of negative slopes), though these slopes were generally of smaller magnitude than those for autumn climate drivers.
@@
 Like \emph{A. perennans}, climate during peak growing season (spring for \emph{A. perennans} and summer for \emph{E. virginicus}) emerged most commonly as drivers of changes in endophyte prevalence.
-\revise{Across the tested climate drivers}, increases in mean spring precipitation were the strongest predictor of increasing trends in endophyte prevalence for \emph{A.
+\secondrevise{Across the tested climate drivers}, increases in mean spring precipitation were the strongest predictor of increasing trends in endophyte prevalence for \emph{A.
 hyemalis} (Fig.
@@
 While \emph{Epichloë} symbioses are highly specialized, surveys have demonstrated genotypic and chemotypic diversity of the symbionts among and within populations \cite{von2021genetic, treindl2023genetic}.
-Genotypic variation in \emph{Epichloë} endophytes, particularly in genes responsible for alkaloid production, produces ``chemotypes" with differing benefits for hosts against insect or mammalian herbivores mediated by environmental conditions \cite{schardl2012chemotypic,saikkonen2013chemical}.
+Genotypic variation in \emph{Epichloë} endophytes, particularly in genes responsible for alkaloid production, produces ``chemotypes" with differing benefits for hosts against insect or mammalian herbivores mediated by environmental conditions \cite{ambrose2014horizontal,schardl2012chemotypic,saikkonen2013chemical}.
 Genotypic variation of the hosts themselves can also influence interaction outcomes \cite{gundel2011interaction,parker2017genotype}.
@@
 \linelabel{R2C44-begin}Additionally, \emph{Epichloë} endophytes have been connected to a suite of non-drought related fitness benefits including herbivory \firstrevise{defense} \citep{brem2001epichloe}, salinity resistence \citep{wang2020effects}, and mediation of pathogens \cite{vikuk2019infection} and the soil microbiome \citep{roberts2015rhizosphere}.
-\firstrevise{Broad changes in the distribution and abundance of natural enemies \cite{cote2004ecological}, along with stresses from anthropogenic changes in \linelabel{R3C7-begin}\revise{land management} and pollution \cite{sage2020global} likely influence the benefits of symbiosis \cite{rudgers2020climate}.}
-\revise{Changing endophyte prevalence} results from the combination of \revise{net} fitness benefits playing out across the heterogeneous map of a changing climate and \linelabel{R3C8-begin}\revise{and its interactive effects on} other anthropogenic drivers.\linelabel{R2C44-end}\linelabel{R3C8-end}
-\revise{Host species experience a world that is made increasingly stressful by a combination of global change drivers, and while historic trends that we observed suggest that symbiotic fitness benefits have helped mitigate this stress, it is possible that at yet higher levels of stress, increasing costs of the mutualism could lead to declines in endophyte prevalence.
+\firstrevise{Broad changes in the distribution and abundance of natural enemies \cite{cote2004ecological}, along with stresses from anthropogenic changes in \linelabel{R3C7-begin}\secondrevise{land management} and pollution \cite{sage2020global} likely influence the benefits of symbiosis \cite{rudgers2020climate}.}
+\secondrevise{Changing endophyte prevalence} results from the combination of \secondrevise{net} fitness benefits playing out across the heterogeneous map of a changing climate and \linelabel{R3C8-begin}\secondrevise{and its interactive effects on} other anthropogenic drivers.\linelabel{R2C44-end}\linelabel{R3C8-end}
+\secondrevise{Host species experience a world that is made increasingly stressful by a combination of global change drivers, and while historic trends that we observed suggest that symbiotic fitness benefits have helped mitigate this stress, it is possible that at yet higher levels of stress, increasing costs of the mutualism could lead to declines in endophyte prevalence.
+\linelabel{R5C5-begin}\revise{It is also know that stressful conditions can reduce the rate of successful transmission of \emph{Epichloë} endophytes from mother plant to offspring \cite{gundel2011neotyphodium}, that may result from host sanctions on costly symbionts or symbiont failure to successfully colonize seeds \cite{afkhami2008symbiosis,gundel2011incorporating}.
+While we did not investigate differences in the rate of transmission across these historic specimens due to low numbers of sampled seeds per individual, it would be extremely valuable to conduct deeper sampling for a subset of specimens with known endophyte status and investigate how transmission itself may respond to environmental change.}\linelabel{R5C5-end}
 More extreme climate stresses, which are expected more frequently in the future \cite{seneviratne2021weather} could shift the balance of interactions costs and benefits.
@@
 Previous population surveys \citep{sneck2017variation,rudgers2009benefits,semmartin2015broad} attributed environment-dependent gradients in endophyte prevalence to symbiont-derived fitness benefit's allowing hosts to persist in environments where they otherwise could not \citep{afkhami2014mutualist,kazenel2015mutualistic}.
-\linelabel{R1C11-b-begin}However, symbiont-facilitated range shifts require that endophytes be present in the populations to be able to \revise{contribute to} population growth. 
-For example, the arid western range edge of \emph{A.
+\linelabel{R1C11-b-begin}However, symbiont-facilitated range shifts require that endophytes be present in the populations to be able to \secondrevise{contribute to} population growth. 
+\linelabel{R5C7-begin}For example, the arid western range edge of \emph{A.
 hyemalis} has had historically low endophyte prevalence (Fig.
-\ref{fig:svc_space_year_map}), and while prevalence has increased most quickly in regions with historically low endophyte prevalence (Fig.
-\ref{fig:initialprev_trend_plot}), the complete absence of endophytes at range edges in this species would make it impossible for prevalence to increase without dispersal of symbiotic seeds \cite{fowler2023geographic}.\linelabel{R1C11-b-end}
+\ref{fig:svc_space_year_map})\revise{, and dispersal of symbionts may limit the capacity for range shifts.
+}
+\revise{A range edge population with no endophytes would require dispersal of symbionts to occur, either via symbiotic seeds or \linelabel{R5C5-discussion-begin} horizontally-transmitted fungal spores, before the symbiosis could result in an expanding distibution \cite{fowler2023geographic}}\linelabel{R5C5-discussion-end}
+\revise{At the same time, we found that} endophyte prevalence has increased most quickly in regions with historically low endophyte prevalence (Fig.
+\ref{fig:initialprev_trend_plot}), suggesting strong selection for symbiotic hosts when they are present.\linelabel{R1C11-b-end}\linelabel{R5C7-end}
 These factors potentially contribute to the ability of the host species to track its environmental niche. 
@@
 \linelabel{R1C9-end}
-\linelabel{R4C10b-begin} \revise{Herbarium collections were predominately used for taxonomic research in the past, but use of specimens to understand ongoing global change would benefit from increased collection efforts and expansion of herbarium collections.} \linelabel{R4C10b-end}
-\linelabel{R3C6-begin}\revise{An alternative validation test would be to hold-out samples from the historic data set.
+\linelabel{R4C10b-begin} \secondrevise{Herbarium collections were predominately used for taxonomic research in the past, but use of specimens to understand ongoing global change would benefit from increased collection efforts and expansion of herbarium collections.} \linelabel{R4C10b-end}
+\linelabel{R3C6-begin}\secondrevise{An alternative validation test would be to hold-out samples from the historic data set.
 Such a test would more clearly match the conditions of the training data (i.e., in spatial scale and climate conditions), however the trade-off between training and testing the model with a limited number of sampled specimens held us back from exploring this option.
@@
 
-\linelabel{R3C3b-begin}\revise{Another key consideration in forecasting the dynamics of host-microbe symbioses is the spatial scale of both specimen georeferencing and available climate data. 
+\linelabel{R3C3b-begin}\secondrevise{Another key consideration in forecasting the dynamics of host-microbe symbioses is the spatial scale of both specimen georeferencing and available climate data. 
 For this analysis, most specimen localities were assigned coordinates at county or city centroids, and the climate data examined was on 4 km grid cells.
@@
 Poor predictive ability at local scales in this grass-endophyte system is not surprising, as previous studies have found that local variation (e.g., in soil conditions, in microclimate), even to the scale of hundreds of meters can structure endophyte-host niches \cite{kazenel2015mutualistic,gundel2024temporal}.
-\linelabel{R3C5-begin}\revise{Local adaptation in either the host or symbiont to microclimate or soil conditions could cause populations to differ from broad regional trends.}\linelabel{R3C5-end}
-\linelabel{R3C2c-begin}\revise{The choice of prior distributions for spatially-varying random effects also impacts the model's flexibility to capture spatial trends. 
+\linelabel{R3C5-begin}\secondrevise{Local adaptation in either the host or symbiont to microclimate or soil conditions could cause populations to differ from broad regional trends.}\linelabel{R3C5-end}
+\linelabel{R3C2c-begin}\secondrevise{The choice of prior distributions for spatially-varying random effects also impacts the model's flexibility to capture spatial trends. 
 Our exploration of model sensitivity to prior choice (presented in the \emph{Supplemental Methods}) reveals qualitatively similar results across a broad range of priors.}
@@
 While we used relatively coarse presence/absence data based on fungal morphology, other studies have examined historic plant microbiomes using molecular sequencing and sophisticated bioinformatics techniques, but these studies have so far been limited to relatively few specimens at limited spatial extents \cite{bearchell2005wheat,yoshida2015computational,heberling2019utilizing, bieker2020metagenomic,gross2021hidden,bradshaw2021global,bradshaw2023genetic}.
-\linelabel{R4C11b-begin}\revise{Much of this work highlights the important role that historic specimens can play in tracking pathogens, a particularly important area as climate change facilitates the spread of new diseases \cite{ristaino2020importance,singh2023climate}}\linelabel{R4C11b-end}
+\linelabel{R4C11b-begin}\secondrevise{Much of this work highlights the important role that historic specimens can play in tracking pathogens, a particularly important area as climate change facilitates the spread of new diseases \cite{ristaino2020importance,singh2023climate}}\linelabel{R4C11b-end}
 Continued advances in capturing historic DNA and in filtering out potential contamination during specimen storage \citep{daru2019novel,bakker2020herbarium,raxworthy2021mining} will be imperative in the effort to scale up these efforts.
@@
 Beyond host-microbe symbioses, detecting ecological responses to anthropogenic global change and attributing their causes would inform public policy decision-makers and adaptive management strategies.
-\linelabel{R4C10c-begin}\revise{Natural history specimens, such as the plant hosts examined in this study, have a clear role to play in informing global change biodiversity science, including building understanding of the dynamics of host-symbiont interactions \cite{davis2023herbarium}.}\linelabel{R4C10c-end}\
+\linelabel{R4C10c-begin}\secondrevise{Natural history specimens, such as the plant hosts examined in this study, have a clear role to play in informing global change biodiversity science, including building understanding of the dynamics of host-symbiont interactions \cite{davis2023herbarium}.}\linelabel{R4C10c-end}\
 This survey of historic endophyte prevalence is necessarily correlative, yet it serves as a foundation to develop better predictive models of the response of microbial symbioses to climate change. 
@@
 	This research was supported by funding from National Science Foundation (grants 1754468 and 2208857) and by funding from the Texas Ecolab Program.
-	\revise{Four} anonymous reviewers greatly improved earlier versions of this manuscript.
+	\secondrevise{Four} anonymous reviewers greatly improved earlier versions of this manuscript.
 

### 2025-10-02 Manuscript/Reviewer_response_endo_herbarium_R3.tex
feedback on R3

@@
 % ======================================================= %
-\title{Manuscript GCB-25-0529.R1 --- Response to reviewers}
+\title{Manuscript GCB-25-0529.R3 --- Response to reviewers}
 
@@
 Following the positive feedback from reviewers on our last submission, we have revised our manuscript to clarify the description of the biology of \emph{Epichloë} symbiosis, including description of the biochemical mechanisms underlying the interactions and the taxonomic diversity.
-We have also added additional discussion of the potential for horizontal and imperfect transmission to influence host responses to climate change.
+We have also added discussion of the potential for horizontal and imperfect transmission to influence host responses to climate change, as suggested.
 
@@
 \comment{The authors did an excellent job; all my concerns have been addressed.}
-\response{We are glad to have addressed the concerns of Reviewer 3 through our previous revision, whose Review helped us to more clearly present the underlying assumptions in our modeling approach.}
+\response{We are glad to have addressed the concerns of Reviewer 3 through our previous revision. Their review helped us to more clearly present the underlying assumptions in our modeling approach.}
 
@@
 I have only minor suggestions.}
-\response{We appreciate this positive evaluation of our work, and we are glad that Reviewer 5 sees its potential importance to the broader audience of global climate change researchers.}
+\response{We appreciate this positive evaluation of our work, and we are glad that Reviewer 5 sees its potential importance to the broader audience of global change biologists.}
 
@@
 I think the technique is fine for the analysis, and innovative in that it utilizes a rich archival resource.}
-\response{Reviewer 5's suggestions to improve our communication of the biology of our study system were especially valuable, and we appreciate that they acknowledge the unique and innovative data presented.}
+\response{Reviewer 5's suggestions to improve our communication of the biology of grass-endophyte symbioses were especially valuable, and we appreciate that they acknowledge the unique and innovative data presented.}
 
@@
 We feel, based on previous work from our lab group and others (Sneck et al., {Microbial Ecology}, 2017; Cavazos et al., \emph{New Phytologist}, 2018; Gagic et al., \emph{Frontiers in Plant Science}, 2018), that doing so properly would require evaluating larger numbers of seeds per individual to quantify with greater precision the rate of imperfect transmission.
+We scored up to five seeds per specimen, which was adequate for assessing symbiont status but probably too low to provide robust estimates for the seed transmission rate of symbiotic specimens. 
 Therefore, we have opted to incorporate this material within the discussion, suggesting future targeted sampling that could be done to get these data from historic specimens.}
@@
 \response{Reviewer 5 is correct that there is a possibility of horizontal transmission in general for \emph{Epichloë}, and indeed for these host species.
-In response to this comment, we have mentioned the potential for horizontal transmission in our description of the study system, and where relevant within the the discussion section.}
+\tom{In response to this comment, we acknowledge the potential for horizontal transmission in our description of the study system, and where relevant within the the discussion section.}{I have not read your changes to the ms yet, but it might be worth going further here to acknowledge how H transmission would or would not influence the results and interpretation.
+As far as I can see, it basically changes nothing about the overall pattern that we detect, but it could influence our interpretation.
+If H transmission were very common (which I am quite sure it is not) then the rise in prevalence may not be driven by strengthening host-symbiont mutualism, but the opposite: symbionts become more parasitic and getting around more commonly through infectious spread.
+It might be worth quoting some numbers from our previous work so show that H transmission in these species is very very low (your Eco Lett paper and see Marion's AGHY paper).}}
 \linesref{ms-R5C3-begin}{ms-R5C3-end}
@@
 Maybe "impossible" is an overly strong word.}
-\response{Lines 534-536 discuss the potential that an isolated, range-edge population starting with no symbionts will experience an increase in symbiosis.
+\response{Thanks you for this comment.
+These lines of text discuss the potential that an isolated, range-edge population starting with no symbionts will experience an increase in symbiosis.
 Our wording does indeed suggest that this could happen primarily through the introduction of symbiotic seeds via dispersal into the population, without mentioning the other possiblity pointed out by Reviewer 5 -- that the symbiont could disperse without its host as a horizontally-transmitted propagule.
@@
 \response{We thank Reviewer 5 for pointing us towards this literature.
-The biochemical nature of \emph{Epichloë} symbioses is not our particular area of expertise, so we appreciate this feedback.
-In response to this comment, we have edited this sentence and others in our discussion section, to acknowledge the potentially diverse biologically active factores involved, and have added references to research uncovering the role of these chemicals.}
+The biochemistry of \emph{Epichloë} symbioses is not our particular area of expertise, so we appreciate this feedback.
+In response to this comment, we have edited this sentence and others in our discussion section, to acknowledge the potentially diverse biologically active factors involved, and have added references to research uncovering the role of these chemicals.}
 \linesref{ms-R5-C9a-begin}{ms-R5-C9a-end}

### 2025-10-02 Manuscript/endo_herbarium_R3.tex
feedback on R3

@@
 \linelabel{R5C7-end-dubious}
+\tom{}{I have not made any edits here but I find it a little confusing to provide to sets of estimates for endophyte prevalence, one from Poaceae and one from Pooideae.
+Maybe that's just a me thing and not worth changing.}
 They are \firstrevise{predominantly} transmitted vertically from maternal plants to offspring through seeds.
@@
 \secondrevise{Host species experience a world that is made increasingly stressful by a combination of global change drivers, and while historic trends that we observed suggest that symbiotic fitness benefits have helped mitigate this stress, it is possible that at yet higher levels of stress, increasing costs of the mutualism could lead to declines in endophyte prevalence.
-\linelabel{R5C5-begin}\revise{It is also know that stressful conditions can reduce the rate of successful transmission of \emph{Epichloë} endophytes from mother plant to offspring \cite{gundel2011neotyphodium}, that may result from host sanctions on costly symbionts or symbiont failure to successfully colonize seeds \cite{afkhami2008symbiosis,gundel2011incorporating}.
-While we did not investigate differences in the rate of transmission across these historic specimens due to low numbers of sampled seeds per individual, it would be extremely valuable to conduct deeper sampling for a subset of specimens with known endophyte status and investigate how transmission itself may respond to environmental change.}\linelabel{R5C5-end}
+\linelabel{R5C5-begin}\revise{\tom{It is also know that stressful conditions can reduce the rate of successful transmission of \emph{Epichloë}}{I think this statement is too strong given that there is also evidence that stress promotes transmission -- see Pedro Gundel's paper on Poa autumnalis.} endophytes from mother plant to offspring \cite{gundel2011neotyphodium}, a likely consequences of host sanctions on costly symbionts or symbiont failure to successfully colonize seeds \cite{afkhami2008symbiosis,gundel2011incorporating}.
+While we did not investigate differences in the rate of transmission across these historic specimens due to low numbers of sampled seeds per individual, it would be valuable to conduct deeper sampling for a subset of specimens with known endophyte status and investigate how transmission itself may respond to environmental change.}\linelabel{R5C5-end}
 More extreme climate stresses, which are expected more frequently in the future \cite{seneviratne2021weather} could shift the balance of interactions costs and benefits.
@@
 Tucker.
-	This research was supported by funding from National Science Foundation (grants 1754468 and 2208857) and by funding from the Texas Ecolab Program.
-	\secondrevise{Four} anonymous reviewers greatly improved earlier versions of this manuscript.
+	This research was supported by funding from National Science Foundation Division of Environmental Biology (grants 1754468 and 2208857) and by funding from the Texas Ecolab Program.
+	\secondrevise{Five} anonymous reviewers greatly improved earlier versions of this manuscript.
 

### 2025-10-06 Manuscript/Reviewer_response_endo_herbarium_R3.tex
committing manuscript and response updates

@@
 \response{Reviewer 5 is correct that there is a possibility of horizontal transmission in general for \emph{Epichloë}, and indeed for these host species.
-\tom{In response to this comment, we acknowledge the potential for horizontal transmission in our description of the study system, and where relevant within the the discussion section.}{I have not read your changes to the ms yet, but it might be worth going further here to acknowledge how H transmission would or would not influence the results and interpretation.
-As far as I can see, it basically changes nothing about the overall pattern that we detect, but it could influence our interpretation.
-If H transmission were very common (which I am quite sure it is not) then the rise in prevalence may not be driven by strengthening host-symbiont mutualism, but the opposite: symbionts become more parasitic and getting around more commonly through infectious spread.
-It might be worth quoting some numbers from our previous work so show that H transmission in these species is very very low (your Eco Lett paper and see Marion's AGHY paper).}}
+In response to this comment, we acknowledge the potential for horizontal transmission in our description of the study system, and where relevant within the the discussion section.
+In particular, we note that stromata production has been rare when it has been observed for these host species (less than 1\% of individuals), and also we acknowledge how horizontal transmission would or would not influence our interpretation -- If horizontal transmission were very common, the rising symbiont prevalence we observed in our study could be a signature of symbionts becoming more parasitic and dispersing more commonly through infectious spread, rather than a strengthening of mutualism.}
 \linesref{ms-R5C3-begin}{ms-R5C3-end}
+\linesref{ms-R5C5-b-begin}{ms-R5C5-b-end}
 \linesref{ms-R5C5-discussion-begin}{ms-R5C5-discussion-end}
@@
 In response to this comment, we have edited this sentence and others in our discussion section, to acknowledge the potentially diverse biologically active factors involved, and have added references to research uncovering the role of these chemicals.}
-\linesref{ms-R5-C9a-begin}{ms-R5-C9a-end}
-\linesref{ms-R5C8-discussion-begin}{ms-R5C8-discussion-end}
+\linesref{ms-R5C8-begin}{ms-R5C8-end}
 

### 2025-10-06 Manuscript/endo_herbarium_R3.tex
committing manuscript and response updates

--- 2025-10-01 Manuscript/endo_herbarium_R3.tex
@@
 \newcommand{\secondrevise}[1]{{\color{black}{#1}}}
-\newcommand{\revise}[1]{{\color{Mahogany}{#1}}}
+\newcommand{\revise}[1]{{\color{black}{#1}}}
 
@@
 
-\emph{Epichloë} fungal endophytes are specialized symbionts of cool-season grasses, \linelabel{R5C7-begin-dubious}\revise{estimated from surveys to associate with $\sim 20 - 30$\%} of species \revise{in the diverse Poaceae family} \citep{leuchtmann1992systematics}.
-\revise{Surveys within the cool-season grass subfamily (Pooideae) have found that associations occur for between ~17\% to ~40\% of sampled species \cite{iannone2011diversity,card2014mutualistic}}.
+\emph{Epichloë} fungal endophytes are specialized symbionts of cool-season grasses, \linelabel{R5C7-begin-dubious}\revise{estimated from surveys to associate with $\sim 20 - 30$\%} of species \revise{across the diverse Poaceae family} \citep{leuchtmann1992systematics}.
+\revise{Within the cool-season grass subfamily (Pooideae), it has been estimated that between 17\% to 40\% of sampled species act as \emph{Epichloë} hosts \cite{iannone2011diversity,card2014mutualistic}}.
 \linelabel{R5C7-end-dubious}
@@
 Over time, endophytes that act as mutualists should rise in prevalence within a host population, particularly under environmental conditions that elicit protective benefits \citep{donald2021context}. 
-\emph{Epichloë} are known to improve their hosts' drought tolerance \cite{decunta2021systematic} and protect their hosts against herbivores \linelabel{R5C8-discussion-begin}\revise{\cite{crawford2010fungal,ambrose2014horizontal}} and pathogens \revise{\cite{xia2018role,tian2017epichloe}} likely through the production of a suite of  \linelabel{R5-C9a-begin}\revise{biologically active molecules, including diverse} alkaloids, \revise{proteins,} and other secondary metabolites.\linelabel{R5-C9a-end}\linelabel{R5C8-discussion-end}
+\emph{Epichloë} are known to improve their hosts' drought tolerance \cite{decunta2021systematic} and protect their hosts against herbivores \linelabel{R5C8-begin}\revise{\cite{crawford2010fungal,ambrose2014horizontal}} and pathogens \revise{\cite{xia2018role,tian2017epichloe}} likely through the production of a suite of  \linelabel{R5-C9a-begin}\revise{biologically active molecules, including diverse} alkaloids, \revise{proteins,} and other secondary metabolites.\linelabel{R5-C9a-end}\linelabel{R5C8-end}
 The fitness feedback induced by vertical transmission leads to the prediction that endophyte prevalence should be high in populations where these fitness benefits are most important. 
@@
 perennans} and on only less than 1\% of \emph{E.
-virginicus} plants over 14 years \cite{fowler2024microbial}}.\linelabel{R5C3-end}
+virginicus} plants over 14 years \cite{fowler2024microbial}}.
+A similar low frequency of stromata formation (only 0.37\% of recorded inflorescences) was observed for \emph{A. hyemalis} in a separate field experiment \cite{donald2021context}.\linelabel{R5C3-end}
 Some host species have shown \revise{the capacity} to partner with multiple symbiont species or strains, and in some cases multiple symbiont lineages can co-exist within a host population \cite{mc2014species}.
@@
 For the three grass host species we examined, there have been strong increases in prevalence of\linelabel{R2C41-begin} \firstrevise{\emph{Epichloë} endophyte symbiosis}.\linelabel{R2C41-end}
-We interpret increases in prevalence of \emph{Epichloë}, which are vertically transmitted, as adaptive changes that improve the fitness of their hosts under increasing environmental stress.
+We interpret increases in prevalence of \emph{Epichloë}, which are \revise{predominantly} vertically transmitted, as adaptive changes that improve the fitness of their hosts under increasing environmental stress.
 This interpretation is in line with theory predicting that positive fitness feedback caused by vertical transmission leads beneficial symbionts to rise in prevalence within a population \cite{fine1975vectors,donald2021context}.
@@
 \secondrevise{Host species experience a world that is made increasingly stressful by a combination of global change drivers, and while historic trends that we observed suggest that symbiotic fitness benefits have helped mitigate this stress, it is possible that at yet higher levels of stress, increasing costs of the mutualism could lead to declines in endophyte prevalence.
-\linelabel{R5C5-begin}\revise{It is also know that stressful conditions can reduce the rate of successful transmission of \emph{Epichloë} endophytes from mother plant to offspring \cite{gundel2011neotyphodium}, that may result from host sanctions on costly symbionts or symbiont failure to successfully colonize seeds \cite{afkhami2008symbiosis,gundel2011incorporating}.
-While we did not investigate differences in the rate of transmission across these historic specimens due to low numbers of sampled seeds per individual, it would be extremely valuable to conduct deeper sampling for a subset of specimens with known endophyte status and investigate how transmission itself may respond to environmental change.}\linelabel{R5C5-end}
+\linelabel{R5C5-begin}\revise{It is also known that stressful conditions can both reduce \cite{gundel2011neotyphodium} and increase \cite{gundel2020simulated} the rate of successful transmission of \emph{Epichloë} endophytes from mother plant to offspring.
+These responses likely reflect both consequences of host sanctions on costly symbionts and the ability of symbionts to successfully colonize seeds \cite{afkhami2008symbiosis,gundel2011incorporating}. 
+While we did not investigate differences in the rate of transmission across these historic specimens due to low numbers of sampled seeds per individual, it would be valuable to conduct deeper sampling for a subset of specimens with known endophyte status and investigate how transmission itself may respond to environmental change.}\linelabel{R5C5-end}
+\linelabel{R5C5-b-begin}\revise{Additionally, we have interpreted increasing prevalence as a signature of increased host fitness driven by vertical transmission, which we expect is the predominant mode of tranmission in this system.
+Horizontal transmission, while rare, could also explain increasing prevalence if global change stressors lead symbionts to engage in increased parasitic behavior, resulting in increased disperal and transmission at the cost of host fitness.}\linelabel{R5C5-b-end}
 More extreme climate stresses, which are expected more frequently in the future \cite{seneviratne2021weather} could shift the balance of interactions costs and benefits.
@@
 }
-\revise{A range edge population with no endophytes would require dispersal of symbionts to occur, either via symbiotic seeds or \linelabel{R5C5-discussion-begin} horizontally-transmitted fungal spores, before the symbiosis could result in an expanding distibution \cite{fowler2023geographic}}\linelabel{R5C5-discussion-end}
+\revise{A range edge population with no endophytes would require dispersal of symbionts to occur, either via symbiotic seeds or \linelabel{R5C5-discussion-begin} horizontally-transmitted fungal spores, before the symbiosis could result in an expanding distribution \cite{fowler2023geographic}.}\linelabel{R5C5-discussion-end}
 \revise{At the same time, we found that} endophyte prevalence has increased most quickly in regions with historically low endophyte prevalence (Fig.

### 2025-10-08 Analyses/endo_herbarium_multispecies_spatiotemporal_analysis.R
updating plots for facet labeling style

@@ -535,7 +535,7 @@ posterior_samples3 <- bind_cols(data3, post.pred3)
 
 posterior_samples <- bind_rows(posterior_samples1, posterior_samples2, posterior_samples3) %>% 
   arrange(row_number) %>% 
-  select(...55:...154) %>% st_drop_geometry %>% as.matrix()
+  select(...64:...163) %>% st_drop_geometry %>% as.matrix()
 
 # simulating datasets from the posterior samples
 n_post_draws <- 100
@@ -546,6 +546,7 @@ y_sim[i,] <- rbinom(n = length(endo_herb$Endo_status_liberal), size = 1, prob = 
 }
 
 saveRDS(y_sim, file = "y_sim.rds")
+y_sim <- readRDS("y_sim.rds")
 y_sim_df <- t(y_sim)
 colnames(y_sim_df) <- paste("iter", 1:n_post_draws)
 y_sim_df <- as_tibble(y_sim_df, .name_repair = "minimal") %>% 
@@ -673,7 +674,7 @@ year_trend <- ggplot(year.pred) +
   lims(y = c(0,1))
 
 # year_trend
-
+# ggsave(year_trend, filename = "Plots/year_trend.png", width = 11, height = 3.5)
 
 # Making a histogram of the data
 year_hist <- ggplot()+
@@ -699,7 +700,7 @@ year_hist <- ggplot()+
 
 
 year_plot <- year_hist/year_trend +
- plot_layout(ncol = 1, heights = c(1, 2))+ plot_annotation(tag_levels = "A") &   theme(plot.tag.position = c(-0, .96))
+ plot_layout(ncol = 1, heights = c(1, 2))+ plot_annotation(tag_levels = "a",   tag_prefix = '(',tag_suffix = ')') &   theme(plot.tag.position = c(-0, .96))
 
 year_plot
 ggsave(year_plot, filename = "Plots/year_plot.png", width = 7, height = 5)
@@ -731,9 +732,9 @@ year_posterior <- year_aghy + year_agpe + year_elvi + plot_layout(ncol = 1, axis
 
 
 
-int_posterior_wrap <- int_posterior + plot_annotation(tag_levels = list(c("A", "", ""))) & theme(plot.tag = element_text(size = rel(2)))
+int_posterior_wrap <- int_posterior + plot_annotation(tag_levels = list(c("(a)", "", ""))) & theme(plot.tag = element_text(size = rel(2)))
 
[…]

### 2025-10-08 Analyses/posthoc_climate_correlation.R
updating the climate regression plots facet labels

@@ -369,7 +369,7 @@ AGHY_ppt_change_plot <- ggplot(filter(prism_diff_pred_df, climate == "ppt" & spe
 AGHY_ppt_change_plot
 
 
-AGHY_climate_change_plot <- AGHY_tmean_change_plot/AGHY_ppt_change_plot + plot_annotation(tag_levels = "A")
+AGHY_climate_change_plot <- AGHY_tmean_change_plot/AGHY_ppt_change_plot + plot_annotation(tag_levels = "a", tag_prefix = "(", tag_suffix = ")")
 
 ggsave(AGHY_climate_change_plot, filename = "Plots/AGHY_climate_change_plot.png", width = 8, height = 10)
 
@@ -392,7 +392,7 @@ AGPE_ppt_change_plot <- ggplot(filter(prism_diff_pred_df, climate == "ppt" & spe
 # AGPE_ppt_change_plot
 
 
-AGPE_climate_change_plot <- AGPE_tmean_change_plot/AGPE_ppt_change_plot + plot_annotation( tag_levels = "A")
+AGPE_climate_change_plot <- AGPE_tmean_change_plot/AGPE_ppt_change_plot + plot_annotation( tag_levels = "a", tag_prefix = "(", tag_suffix = ")")
 
 ggsave(AGPE_climate_change_plot, filename = "Plots/AGPE_climate_change_plot.png", width = 8, height = 10)
 
@@ -415,7 +415,7 @@ ELVI_ppt_change_plot <- ggplot(filter(prism_diff_pred_df, climate == "ppt" & spe
 # ELVI_ppt_change_plot
 
 
-ELVI_climate_change_plot <- ELVI_tmean_change_plot/ELVI_ppt_change_plot + plot_annotation( tag_levels = "A")
+ELVI_climate_change_plot <- ELVI_tmean_change_plot/ELVI_ppt_change_plot + plot_annotation( tag_levels = "a", tag_prefix = "(", tag_suffix = ")")
 
 ggsave(ELVI_climate_change_plot, filename = "Plots/ELVI_climate_change_plot.png", width = 8, height = 10)
 
@@ -912,7 +912,7 @@ ppt_sd_trend <- ggplot(filter(prediction_df, grepl("ppt", name) & grepl("sd", na
 # ppt_sd_trend$layers[4:6] <- NULL; ppt_sd_trend$layers[[3]]$aes_params$alpha  <-0
 
   
[…]

### 2025-10-08 Manuscript/endo_herbarium_GCB.tex
commiting accepted version of manuscipt, but still need to organize figure files
organized figure captions and updated data statement

--- 2025-10-06 Manuscript/endo_herbarium_R3.tex
@@
 
-	\noindent{} $\ast$ Corresponding author; e-mail: jcf221@miami.edu.
+	\noindent{} $\ast$ Corresponding author; e-mail:jcf221@miami.edu
 
@@
 	\textit{Keywords}: climate change, \emph{Epichloë}, herbarium, INLA, museum specimen, plant-microbe symbiosis, Poaceae, spatially-varying coefficients model
+
+	\bigskip
+
+	\textit{Running Title:} Increasing prevalence of \emph{Epichloë} endophytes
 
@@
 
-\begin{figure}[h]
-	\centering
-	\includegraphics[width = \linewidth]{../Plots/collections_map.png}
-	\caption[Collection locations of herbarium specimens sampled for \emph{Epichloë} endophytes]{\textbf{Collection locations of herbarium specimens sampled for \emph{Epichloë} endophytes.} Specimens span eastern North America from nine herbaria, and are colored by host species (\emph{A.
-hyemalis}: green, \emph{A.
-perennans}: orange, \emph{E.
-virginicus}: purple).
-\firstrevise{Map lines delineate study areas and do not necessarily depict accepted national boundaries.}}
-	\label{fig:map}
-\end{figure}
-
 \subsection*{Modeling spatial and temporal changes in endophyte prevalence}
@@
 \ref{fig:temporal_posterior}).
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width = \linewidth]{../Plots/year_plot.png}
-	\caption[Temporal trends in endophyte prevalence.]{\textbf{Temporal trends in endophyte prevalence.} (A) Histograms show the frequency of scored specimens through time for each host species.
-(B)\firstrevise{ Lines show mean endophyte prevalence predicted by the endophyte prevalence model} over the study period along with the 50\%  and 95\% CI bands incorporating parameter uncertainty and variation associated with collector and scorer random effects.
-Colored points are binned means of the observed endophyte presence/absence data (black dashes).
-Colors represent each host species (\emph{A.
-hyemalis}: green, \emph{A.
-perennans}: orange, \emph{E.
-virginicus}: purple) and point size represents the number of specimens.}
-	\label{fig:temporal}
-\end{figure}
 
@@
 virginicus} trends did not differ according to historic prevalence (\ref{fig:initialprev_trend_plot} C).} \linelabel{R1C7-end}
-\begin{figure}[H]
-	\centering
-	\includegraphics[width = \linewidth]{../Plots/svc_time_map.png}
-	\caption[Predicted posterior mean of spatially-varying slopes representing change in endophyte prevalence for each host species.]{Predicted posterior mean of spatially-varying slopes representing change in endophyte prevalence for each host species ((A)\emph{A.
-hyemalis}; (B)\emph{A.
-perennans}; (C)\emph{E.
-virginicus}).
-\firstrevise{Spatially-varying trends are estimated from the endophyte prevalence model.} Color indicates the relative change in predicted endophyte prevalence.
-\firstrevise{Map lines delineate study areas and do not necessarily depict accepted national boundaries.}}
-	\label{fig:svc_time_map}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width = \linewidth]{../Plots/svc_space_map_year.png}
-	\caption[Predicted endophyte prevalence for each host species in 1895 and 2020.]{\firstrevise{Predicted endophyte prevalence for each host species in 1895 and 2020.
-\firstrevise{Predictions of prevalence come from the endophyte prevalence model.} Color indicates the posterior mean endophyte prevalence for (A, D) \emph{A.
-hyemalis}, (B, E) \emph{A.
-perennans}, and (C, F) \emph{E.
-virginicus}.
-Map lines delineate study areas and do not necessarily depict accepted national boundaries.}}
-	\label{fig:svc_space_year_map}
-\end{figure}
 
@@
 
-\begin{figure}[H]
-	\centering
-	\includegraphics[width = \linewidth]{../Plots/climate_trends_plot_intercept.png}
-	\caption[Relationships between predicted trends in endophyte prevalence and changes in seasonal climate drivers.]{\textbf{Relationships between predicted trends in endophyte prevalence and changes in seasonal climate drivers.} Lines show marginal predicted relationship between spatially-varying trends in endophyte prevalence and changes in mean and variability of climate ((A): mean temperature, (B): cumulative precipitation, (C): standard deviation in temperature, (D): standard deviation in precipitation) \firstrevise{estimated from the \emph{post hoc} climate regression analysis}.
-Confidence bands represent the 50 and 95\% CI,  colored by host species (\emph{A.
-hyemalis}: green, \emph{A.
-perennans}: orange, \emph{E.
-virginicus}: purple).
-Slopes with greater than 95\% \firstrevise{posterior} probability of being either positive or negative are represented as solid lines while those that have less than 95\% probability are dashed.
-Points are the \firstrevise{values of pre-computed SVC trends  and climate drivers at} 250 randomly sampled pixels across each host's distribution used in model fitting \firstrevise{for the \emph{post hoc} climate regression analysis}.}
-	\label{fig:climate_correlation_plot}
-\end{figure}
-
 \subsection*{Evaluation of model performance on an out-of-sample test}
@@
 %We discuss potential model improvements in the Discussion.
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width = \linewidth]{../Plots/contemp_test_plot.png}
-	\caption[Predictive performance for contemporary test data.]{\textbf{Predictive performance for contemporary test data.} (A)
-		 The \firstrevise{endophyte prevalence model}, trained on historic herbarium collection data, performed modestly at predicting prevalence in contemporary population surveys.
-The model captured regional trends across (A) longitude and (B) latitude.
-Crosses indicate predicted mean prevalence along with the 95\% CI (colored lines: \emph{A.
-hyemalis}: green, orange, \emph{E.
-virginicus}: purple) from the herbarium model.
-Contemporary prevalence is represented by grey points (point size reflects sample size) along with trend lines from generalized linear models (black line and shaded 95\% confidence interval).
-(C) Comparison of \firstrevise{contemporary} observed \firstrevise{population prevalence} vs.
-predicted endophyte prevalence shows that contemporary test data had more variance between populations than \firstrevise{in model} predictions.}
-	\label{fig:contemptestplot}
-\end{figure}
 
@@
 
-	\section*{Data and Code Availability}
-	\sloppy{
-	Data from this publication can be found through a publicly available repository (https://doi.org/10.5061/dryad.rn8pk0pn0).
-	Code for analyses can be found through a publicly available repository (https://github.com/joshuacfowler/EndoHerbarium) that will be permanently archived upon publication.
-}
+	%\section*{Data and Code Availability}
+	%\sloppy{
+	%Data from this publication can be found through a publicly available repository (https://doi.org/10.5061/dryad.rn8pk0pn0).
+	%Code for analyses can be found through a publicly available repository (https://github.com/joshuacfowler/EndoHerbarium) that is %permanently archived on Zenodo: DOI: 10.5281/zenodo.17298101.
+Code includes methods for downloading publically available %datasets, including temperature and precipitation data from the PRISM climate group (https://prism.oregonstate.edu) and GBIF %occurrence records for \emph{Epichloë} host species (\emph{A.
+hyemalis} \cite{https://doi.org/10.15468/dl.ryqqcv}, \emph{A.
+%perennans} \cite{https://doi.org/10.15468/dl.wtne24}, \emph{E.
+virginicus} \cite{https://doi.org/10.15468/dl.yt76uh}).
+%}
 \newpage{}
@@
 	\bibliography{endo_herbarium}
+
+\newpage{}
+\section*{Figure Captions}
+
+\begin{figure}[h]
+	\centering
+	%\includegraphics[width = \linewidth]{../Plots/collections_map.png}
+	\caption[Collection locations of herbarium specimens sampled for \emph{Epichloë} endophytes]{\textbf{Collection locations of herbarium specimens sampled for \emph{Epichloë} endophytes.} Specimens span eastern North America from nine herbaria, and are colored by host species (\emph{A.
+hyemalis}: green, \emph{A.
+perennans}: orange, \emph{E.
+virginicus}: purple).
+\firstrevise{Map lines delineate study areas and do not necessarily depict accepted national boundaries.}}
+	\label{fig:map}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	%\includegraphics[width = \linewidth]{../Plots/year_plot.png}
+	\caption[Temporal trends in endophyte prevalence.]{\textbf{Temporal trends in endophyte prevalence.} (A) Histograms show the frequency of scored specimens through time for each host species.
+(B)\firstrevise{ Lines show mean endophyte prevalence predicted by the endophyte prevalence model} over the study period along with the 50\%  and 95\% CI bands incorporating parameter uncertainty and variation associated with collector and scorer random effects.
+Colored points are binned means of the observed endophyte presence/absence data (black dashes).
+Colors represent each host species (\emph{A.
+hyemalis}: green, \emph{A.
+perennans}: orange, \emph{E.
+virginicus}: purple) and point size represents the number of specimens.}
+	\label{fig:temporal}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	%\includegraphics[width = \linewidth]{../Plots/svc_time_map.png}
+	\caption[Predicted posterior mean of spatially-varying slopes representing change in endophyte prevalence for each host species.]{Predicted posterior mean of spatially-varying slopes representing change in endophyte prevalence for each host species ((A)\emph{A.
+hyemalis}; (B)\emph{A.
+perennans}; (C)\emph{E.
+virginicus}).
+\firstrevise{Spatially-varying trends are estimated from the endophyte prevalence model.} Color indicates the relative change in predicted endophyte prevalence.
+\firstrevise{Map lines delineate study areas and do not necessarily depict accepted national boundaries.}}
+	\label{fig:svc_time_map}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	%\includegraphics[width = \linewidth]{../Plots/svc_space_map_year.png}
+	\caption[Predicted endophyte prevalence for each host species in 1895 and 2020.]{\firstrevise{Predicted endophyte prevalence for each host species in 1895 and 2020.
+\firstrevise{Predictions of prevalence come from the endophyte prevalence model.} Color indicates the posterior mean endophyte prevalence for (A, D) \emph{A.
+hyemalis}, (B, E) \emph{A.
+perennans}, and (C, F) \emph{E.
+virginicus}.
+Map lines delineate study areas and do not necessarily depict accepted national boundaries.}}
+	\label{fig:svc_space_year_map}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	%\includegraphics[width = \linewidth]{../Plots/climate_trends_plot_intercept.png}
+	\caption[Relationships between predicted trends in endophyte prevalence and changes in seasonal climate drivers.]{\textbf{Relationships between predicted trends in endophyte prevalence and changes in seasonal climate drivers.} Lines show marginal predicted relationship between spatially-varying trends in endophyte prevalence and changes in mean and variability of climate ((A): mean temperature, (B): cumulative precipitation, (C): standard deviation in temperature, (D): standard deviation in precipitation) \firstrevise{estimated from the \emph{post hoc} climate regression analysis}.
+Confidence bands represent the 50 and 95\% CI,  colored by host species (\emph{A.
+hyemalis}: green, \emph{A.
+perennans}: orange, \emph{E.
+virginicus}: purple).
+Slopes with greater than 95\% \firstrevise{posterior} probability of being either positive or negative are represented as solid lines while those that have less than 95\% probability are dashed.
+Points are the \firstrevise{values of pre-computed SVC trends  and climate drivers at} 250 randomly sampled pixels across each host's distribution used in model fitting \firstrevise{for the \emph{post hoc} climate regression analysis}.}
+	\label{fig:climate_correlation_plot}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	%\includegraphics[width = \linewidth]{../Plots/contemp_test_plot.png}
+	\caption[Predictive performance for contemporary test data.]{\textbf{Predictive performance for contemporary test data.} (A)
+		The \firstrevise{endophyte prevalence model}, trained on historic herbarium collection data, performed modestly at predicting prevalence in contemporary population surveys.
+The model captured regional trends across (A) longitude and (B) latitude.
+Crosses indicate predicted mean prevalence along with the 95\% CI (colored lines: \emph{A.
+hyemalis}: green, orange, \emph{E.
+virginicus}: purple) from the herbarium model.
+Contemporary prevalence is represented by grey points (point size reflects sample size) along with trend lines from generalized linear models (black line and shaded 95\% confidence interval).
+(C) Comparison of \firstrevise{contemporary} observed \firstrevise{population prevalence} vs.
+predicted endophyte prevalence shows that contemporary test data had more variance between populations than \firstrevise{in model} predictions.}
+	\label{fig:contemptestplot}
+\end{figure}
 
@@
 \textbf{Software}:  All statistics were performed using Maxent 3.3.4 and R4.3.1 with packages terra, usdm, spThin and dismo.\\
-\textbf{Code availability}: Available through this link: https://github.com/joshuacfowler/EndoHerbarium \\
+\textbf{Code availability}: Available through this link: https://github.com/joshuacfowler/EndoHerbarium and archived on Zenodo (DOI: 10.5281/zenodo.17298101)\\
 \textbf{Data availability}: Data was accessed through open-source R packages \emph{rgbif}.
